### 2021-04-21 about/about.Rmd
update

---
layout: default
title: "About"
nav_order: 1
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Rigorous, Reproducible Research in Marine Genomics

Welcome! We are a working group dedicated to promoting reproducible and robust genomic research in marine science and evolutionary biology.

## Organizers:

Katherine Silliman and Sam Bogan

## Faculty Mentors

Jon Puritz and Andrew Whitehead

## Working group members

``````{r, echo=FALSE,eval=FALSE}
library(knitr)
my.jekyll.site <- "marineomics.github.io/Template_Site_1/"
KnitPost <- function(input, base.url = my.jekyll.site) {
  opts_knit$set(base.url = base.url)
  fig.path <- paste0("images/", sub(".Rmd$", "", basename(input)), "/")
  opts_chunk$set(fig.path = fig.path)
  opts_chunk$set(fig.cap = "center")
  render_jekyll()
  knit(input, envir = parent.frame())
}

setwd("C:/Users/jmcgirr/Documents/GitHub/MarineOmics/Template_Site_1.github.io/about/")
KnitPost("C:/Users/jmcgirr/Documents/GitHub/MarineOmics/Template_Site_1.github.io/about/about.Rmd")

```

### 2021-04-21 functional_genomics/functional_genomics.Rmd
update

--- 2021-04-21 about/about.Rmd
@@
 layout: default
-title: "About"
-nav_order: 1
+title: "Functional Genomics"
+nav_order: 2
+has_children: true
 ---
@@
 
-# Rigorous, Reproducible Research in Marine Genomics
-
-Welcome! We are a working group dedicated to promoting reproducible and robust genomic research in marine science and evolutionary biology.
-
-## Organizers:
-
-Katherine Silliman and Sam Bogan
-
-## Faculty Mentors
-
-Jon Puritz and Andrew Whitehead
-
-## Working group members
+# Functional Genomics
 
@@
 
-setwd("C:/Users/jmcgirr/Documents/GitHub/MarineOmics/Template_Site_1.github.io/about/")
-KnitPost("C:/Users/jmcgirr/Documents/GitHub/MarineOmics/Template_Site_1.github.io/about/about.Rmd")
+setwd("C:/Users/jmcgirr/Documents/GitHub/MarineOmics/Template_Site_1.github.io/functional_genomics/")
+KnitPost("C:/Users/jmcgirr/Documents/GitHub/MarineOmics/Template_Site_1.github.io/functional_genomics/functional_genomics.Rmd")
 

### 2021-04-21 genome_assembly/genome_assembly.Rmd
update

--- 2021-04-21 functional_genomics/functional_genomics.Rmd
@@
 layout: default
-title: "Functional Genomics"
-nav_order: 2
+title: "Genome Assembly"
+nav_order: 3
 has_children: true
@@
 
-# Functional Genomics
+# Genome Assembly
 
@@
 
-setwd("C:/Users/jmcgirr/Documents/GitHub/MarineOmics/Template_Site_1.github.io/functional_genomics/")
-KnitPost("C:/Users/jmcgirr/Documents/GitHub/MarineOmics/Template_Site_1.github.io/functional_genomics/functional_genomics.Rmd")
+setwd("C:/Users/jmcgirr/Documents/GitHub/MarineOmics/Template_Site_1.github.io/genome_assembly/")
+KnitPost("C:/Users/jmcgirr/Documents/GitHub/MarineOmics/Template_Site_1.github.io/genome_assembly/genome_assembly.Rmd")
 

### 2021-04-21 population_genetics/high_coverage.Rmd
update

--- 2021-04-21 genome_assembly/genome_assembly.Rmd
@@
 layout: default
-title: "Genome Assembly"
-nav_order: 3
+title: "High Coverage"
+parent: Whole Genome Sequencing
 has_children: true
@@
 
-# Genome Assembly
+# High Coverage WGS
 
@@
 
-setwd("C:/Users/jmcgirr/Documents/GitHub/MarineOmics/Template_Site_1.github.io/genome_assembly/")
-KnitPost("C:/Users/jmcgirr/Documents/GitHub/MarineOmics/Template_Site_1.github.io/genome_assembly/genome_assembly.Rmd")
+KnitPost("C:/Users/jmcgirr/Documents/GitHub/MarineOmics/Template_Site_1.github.io/population_genetics/high_coverage.Rmd")
 

### 2021-04-21 population_genetics/low_coverage.Rmd
update

--- 2021-04-21 population_genetics/high_coverage.Rmd
@@
 layout: default
-title: "High Coverage"
+title: "Low Coverage"
 parent: Whole Genome Sequencing
+grand_parent: Population Genetics
 has_children: true
@@
 
-# High Coverage WGS
+# Low Coverage
 
@@
 
-KnitPost("C:/Users/jmcgirr/Documents/GitHub/MarineOmics/Template_Site_1.github.io/population_genetics/high_coverage.Rmd")
+KnitPost("C:/Users/jmcgirr/Documents/GitHub/MarineOmics/Template_Site_1.github.io/population_genetics/low_coverage.Rmd")
 

### 2021-04-21 population_genetics/population_genetics.Rmd
update

--- 2021-04-21 population_genetics/high_coverage.Rmd
@@
 layout: default
-title: "High Coverage"
-parent: Whole Genome Sequencing
+title: "Population Genetics"
+nav_order: 4
 has_children: true
@@
 
-# High Coverage WGS
+# Population Genetics
+
+Check out [Quality Control and Filtering](https://marineomics.github.io/Template_Site_1/Quality_Control_and_Filtering.html)
+
+General introduction to whole genome sequencing and reduced representation sequencing. Pros and Cons for each.
 
@@
 
-KnitPost("C:/Users/jmcgirr/Documents/GitHub/MarineOmics/Template_Site_1.github.io/population_genetics/high_coverage.Rmd")
+setwd("C:/Users/jmcgirr/Documents/GitHub/MarineOmics/Template_Site_1.github.io/population_genetics/")
+KnitPost("C:/Users/jmcgirr/Documents/GitHub/MarineOmics/Template_Site_1.github.io/population_genetics/population_genetics.Rmd")
 

### 2021-04-21 population_genetics/population_structure.Rmd
update

---
layout: default
title: "Population Structure"
parent: Low Coverage
grand_parent: Whole Genome Sequencing
has_children: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Population Structure

Intro to methods

# Vignette

Perform population structure analyses with [PCAngsd](http://www.popgen.dk/software/index.php/PCAngsd),
[PLINK PCA](https://www.cog-genomics.org/plink/1.9/strat), [NGSadmix](http://www.popgen.dk/software/index.php/NgsAdmix), and [popVAE](https://github.com/kr-colab/popvae).

[Example with Pacific herring](http://htmlpreview.github.com/?https://github.com/MarineOmics/Template_Site_1/blob/main/population_genetics/vignettes/Pacific_Herring_population_strucutre.html)

```{r, echo=FALSE,eval=FALSE}
library(knitr)
my.jekyll.site <- "marineomics.github.io/Template_Site_1/"
KnitPost <- function(input, base.url = my.jekyll.site) {
  opts_knit$set(base.url = base.url)
  fig.path <- paste0("images/", sub(".Rmd$", "", basename(input)), "/")
  opts_chunk$set(fig.path = fig.path)
  opts_chunk$set(fig.cap = "center")
  render_jekyll()
  knit(input, envir = parent.frame())
}

KnitPost("C:/Users/jmcgirr/Documents/GitHub/MarineOmics/Template_Site_1.github.io/population_genetics/population_structure.Rmd")

```

### 2021-04-21 population_genetics/reduced_representation.Rmd
update

--- 2021-04-21 population_genetics/high_coverage.Rmd
@@
 layout: default
-title: "High Coverage"
-parent: Whole Genome Sequencing
+title: "Reduced Representation"
+parent: Population Genetics
 has_children: true
@@
 
-# High Coverage WGS
+# Reduced Representation Sequencing
 
@@
 
-KnitPost("C:/Users/jmcgirr/Documents/GitHub/MarineOmics/Template_Site_1.github.io/population_genetics/high_coverage.Rmd")
+KnitPost("C:/Users/jmcgirr/Documents/GitHub/MarineOmics/Template_Site_1.github.io/population_genetics/reduced_representation.Rmd")
 

### 2021-04-21 population_genetics/whole_genome_seq.Rmd
update

--- 2021-04-21 population_genetics/low_coverage.Rmd
@@
 layout: default
-title: "Low Coverage"
-parent: Whole Genome Sequencing
-grand_parent: Population Genetics
+title: "Whole Genome Sequencing"
+parent: Population Genetics
 has_children: true
@@
 
-# Low Coverage
+# Whole Genome Sequencing
+
+Description of low vs. high coverage
 
@@
 
-KnitPost("C:/Users/jmcgirr/Documents/GitHub/MarineOmics/Template_Site_1.github.io/population_genetics/low_coverage.Rmd")
+KnitPost("C:/Users/jmcgirr/Documents/GitHub/MarineOmics/Template_Site_1.github.io/population_genetics/whole_genome_seq.Rmd")
 

### 2021-04-21 quality_control/high_coverage.Rmd
update

--- 2021-04-21 population_genetics/reduced_representation.Rmd
@@
 layout: default
-title: "Reduced Representation"
-parent: Population Genetics
+title: "High Coverage"
+parent: Quality Control and Filtering
 has_children: true
@@
 
-# Reduced Representation Sequencing
+# High Coverage WGS Sequencing
 
@@
 
-KnitPost("C:/Users/jmcgirr/Documents/GitHub/MarineOmics/Template_Site_1.github.io/population_genetics/reduced_representation.Rmd")
+KnitPost("C:/Users/jmcgirr/Documents/GitHub/MarineOmics/Template_Site_1.github.io/quality_control/high_coverage.Rmd")
 

### 2021-04-21 quality_control/low_coverage.Rmd
update

--- 2021-04-21 population_genetics/low_coverage.Rmd
@@
 title: "Low Coverage"
-parent: Whole Genome Sequencing
-grand_parent: Population Genetics
+parent: Quality Control and Filtering
 has_children: true
@@
 
-KnitPost("C:/Users/jmcgirr/Documents/GitHub/MarineOmics/Template_Site_1.github.io/population_genetics/low_coverage.Rmd")
+KnitPost("C:/Users/jmcgirr/Documents/GitHub/MarineOmics/Template_Site_1.github.io/quality_control/low_coverage.Rmd")
 

### 2021-04-21 quality_control/quality_control.Rmd
update

--- 2021-04-21 quality_control/low_coverage.Rmd
@@
 layout: default
-title: "Low Coverage"
-parent: Quality Control and Filtering
-has_children: true
+title: "Quality Control and Filtering"
+nav_order: 5
 ---
@@
 
-# Low Coverage
+# Quality Control and Filtering
+
+Describe general considerations and link to `.fastq` to `.vcf` pipelines for specific -omics data types.
+
+Describe common file formats?
 
@@
 
-KnitPost("C:/Users/jmcgirr/Documents/GitHub/MarineOmics/Template_Site_1.github.io/quality_control/low_coverage.Rmd")
+KnitPost("C:/Users/jmcgirr/Documents/GitHub/MarineOmics/Template_Site_1.github.io/quality_control/quality_control.Rmd")
 

### 2021-04-21 about/about.Rmd
update

@@
 nav_order: 1
+editor_options: 
+  chunk_output_type: console
 ---
@@
 library(knitr)
-my.jekyll.site <- "marineomics.github.io/Template_Site_1/"
+my.jekyll.site <- "marineomics.github.io"
 KnitPost <- function(input, base.url = my.jekyll.site) {
@@
 
-setwd("C:/Users/jmcgirr/Documents/GitHub/MarineOmics/Template_Site_1.github.io/about/")
-KnitPost("C:/Users/jmcgirr/Documents/GitHub/MarineOmics/Template_Site_1.github.io/about/about.Rmd")
+setwd("C:/Users/jmcgirr/Documents/GitHub/MarineOmics/website/about/")
+KnitPost("C:/Users/jmcgirr/Documents/GitHub/MarineOmics/website/about/about.Rmd")
 

### 2021-04-21 panels.Rmd
update

--- 2021-04-21 about/about.Rmd
@@
 layout: default
-title: "About"
+title: "Panel Seminars"
 nav_order: 1
@@
 
-# Rigorous, Reproducible Research in Marine Genomics
+# About
 
@@
 
-## Organizers:
+# Panel Seminar Series
 
-Katherine Silliman and Sam Bogan
+Lorem ipsum dolor sit amet, graece quidam duo ea, ut nominavi expetenda mea.
+Habeo nominavi repudiandae at his, et eos lorem nusquam suavitate, noster equidem his no.
+Commodo facilisis elaboraret eu sea.
+An eam eleifend evertitur.
+Ad eum delicata definitionem.
 
-## Faculty Mentors
+## Seminar Schedule
+```{r, echo=FALSE,eval=FALSE}
+library(knitr)
 
-Jon Puritz and Andrew Whitehead
+schedule_df <- data.frame(date = c("5/7", "5/14", "5/21"),
+                          speaker = c("John","Jane","Jake"),
+                          )
+dates <- 
+dates <- c("5/7", "5/14", "5/21")
 
-## Working group members
+```
+
+| Date        | Topic       | Panelists     |
+| :---        | :---        | :---          |
+| 5/7         | topic1      | John          |
+| 5/8         | topic2      | Jane          |
+
+# Contact
 
@@
 setwd("C:/Users/jmcgirr/Documents/GitHub/MarineOmics/website/about/")
-KnitPost("C:/Users/jmcgirr/Documents/GitHub/MarineOmics/website/about/about.Rmd")
+KnitPost("C:/Users/jmcgirr/Documents/GitHub/MarineOmics/website/panels.Rmd")
 

### 2021-04-21 index.Rmd
update

---
layout: default
title: "Panel Seminars"
nav_order: 1
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Welcome

Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.

# Panel Seminar Series

Sed ut perspiciatis unde omnis iste natus error sit voluptatem accusantium doloremque laudantium, totam rem aperiam, eaque ipsa quae ab illo inventore veritatis et quasi architecto beatae vitae dicta sunt explicabo. Nemo enim ipsam voluptatem quia voluptas sit aspernatur aut odit aut fugit, sed quia consequuntur magni dolores eos qui ratione voluptatem sequi nesciunt. Neque porro quisquam est, qui dolorem ipsum quia dolor sit amet, consectetur, adipisci velit, sed quia non numquam eius modi tempora incidunt ut labore et dolore magnam aliquam quaerat voluptatem.

## Seminar Schedule

| Date        | Topic       | Panelists     |
| :---        | :---        | :---          |
| 5/7         | topic1      | [Katherine Silliman](https://ksilliman.weebly.com/)|
| 5/8         | topic2      | [Joe McGirr](https://joemcgirr.github.io./)|

# Contact

Katherine Silliman (email)
Sam Bogan (email)

``````{r, echo=FALSE,eval=FALSE}
library(knitr)
my.jekyll.site <- "marineomics.github.io"
KnitPost <- function(input, base.url = my.jekyll.site) {
  opts_knit$set(base.url = base.url)
  fig.path <- paste0("images/", sub(".Rmd$", "", basename(input)), "/")
  opts_chunk$set(fig.path = fig.path)
  opts_chunk$set(fig.cap = "center")
  render_jekyll()
  knit(input, envir = parent.frame())
}

setwd("C:/Users/jmcgirr/Documents/GitHub/MarineOmics/website/")
KnitPost("C:/Users/jmcgirr/Documents/GitHub/MarineOmics/website/index.Rmd")

```

### 2021-05-03 about/about.Rmd
update
Updated Rmd files

@@
 title: "About"
-nav_order: 1
+nav_order: 2
 editor_options: 
@@
 
-# Rigorous, Reproducible Research in Marine Genomics
+# About
 
-Welcome! We are a working group dedicated to promoting reproducible and robust genomic research in marine science and evolutionary biology.
+Welcome!
+The MarineOmics working group is part of the RCN for Evolution in Changing Seas, and was formed to promote reproducible and robust genomic research in evolutionary biology and marine science.
+Methods in next generation sequencing and bioinformatics advance at a pace that quickly render any static set of best practices as obsolete.
+To support robust ‘omics research now and moving forward, the working group will disseminate guidelines through an easily-interfaced and dynamic website that will be deployed in August 2021.
+To help develop content for the website, we are hosting Q & A panel seminars with experts on various topics relating to ‘omics analyses in non-model systems.
 
-## Organizers:
-
-Katherine Silliman and Sam Bogan
-
-## Faculty Mentors
-
-Jon Puritz and Andrew Whitehead
-
-## Working group members
+The working group is also currently hosting three undergraduate research interns with funding from the RCN-ECS.
+These interns are part of a subgroup conducting a literature survey to characterize trends in data availability, code accessibility, and methods reproducibility for a peer-reviewed article.
 
@@
 
-setwd("C:/Users/jmcgirr/Documents/GitHub/MarineOmics/website/about/")
-KnitPost("C:/Users/jmcgirr/Documents/GitHub/MarineOmics/website/about/about.Rmd")
+setwd("~/Postdoc_Projects/RCN_MarineOmics/Website/marineomics.github.io/about/")
+KnitPost("~/Postdoc_Projects/RCN_MarineOmics/Website/marineomics.github.io/about/about.Rmd")
 

### 2021-05-03 index.Rmd
update
Updated Rmd files

@@
 
-# Welcome
+# About
 
-Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
-Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
-Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
-Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
+The MarineOmics working group is part of the [RCN for Evolution in Changing Seas](https://rcn-ecs.github.io/), and was formed to promote reproducible and robust genomic research in marine science.
+Methods in next generation sequencing and bioinformatics advance at a pace that quickly render any static set of best practices as obsolete.
+To support robust 'omics research now and moving forward, the working group will disseminate guidelines through an easily-interfaced and dynamic website that will be deployed in August 2021.
+*Keep an eye on this URL for updates!*
 
@@
 
-Sed ut perspiciatis unde omnis iste natus error sit voluptatem accusantium doloremque laudantium, totam rem aperiam, eaque ipsa quae ab illo inventore veritatis et quasi architecto beatae vitae dicta sunt explicabo.
-Nemo enim ipsam voluptatem quia voluptas sit aspernatur aut odit aut fugit, sed quia consequuntur magni dolores eos qui ratione voluptatem sequi nesciunt.
-Neque porro quisquam est, qui dolorem ipsum quia dolor sit amet, consectetur, adipisci velit, sed quia non numquam eius modi tempora incidunt ut labore et dolore magnam aliquam quaerat voluptatem.
+To facilitate discussions about practices and pitfalls in applying 'omics to non-model systems, we are hosting Q & A panel seminars with experts on topics ranging from RADseq, population genomics with WGS, and transcriptomics.
+Seminars are every two weeks, on **Fridays from 11:00-12:15 US Central Time**.
+3-5 expert panelists will discuss each topic, with questions coming from the working group and the audience.
+If you would like to submit a question for the panelists ahead of time, please send an email to rcnecs.marineomics[at]gmail.com .
+
+Everyone is welcome to join!
+
+**To receive a Zoom link, please fill out the registration form [here](
+https://auburn.zoom.us/meeting/register/tZ0ode2rpz4sG9Il_TuYexWfqd6HyjdQ58Zi).**
 
@@
 | :---        | :---        | :---          |
-| 5/7         | topic1      | [Katherine Silliman](https://ksilliman.weebly.com/)|
-| 5/8         | topic2      | [Joe McGirr](https://joemcgirr.github.io./)|
+| May 7         | GBS/RADseq genotyping pipelines     | [Melanie LaCava](https://melanielacava.weebly.com/), [Alicia Mastretta-Yanes](http://mastrettayanes-lab.org/), [Isaac Overcast](https://isaacovercast.github.io/), [Jon Puritz](https://www.marineevoeco.com/)   |
+| May 21       |  WGS genotyping (low and high coverage)     | [Mikhail "Misha" Matz](https://matzlab.weebly.com/), [Tim Sackton](https://scholar.harvard.edu/tsackton), [Nina Overgaard Therkildsen](https://www.therkildsenlab.com/), TBD |
+| June 4       |  Population genomic analyses and filtering data     | TBD|
+| June 18       |  Functional Genomics (RNAseq)     | TBD|
+
+# Reproducibility in Marine Genomics  
+The working group is also currently hosting three undergraduate research interns with funding from the [RCN-ECS](https://rcn-ecs.github.io/).
+These interns are part of a sub-committee conducting a literature survey to characterize trends in data availability, code accessibility, and methods reproducibility in papers that use genomics to study adaptation in marine systems.
+The results of this survey will be shared in a peer-reviewed article, along with guidance to improve the state of reproducibility in the field.
+# Working Group Members
+
+#### [Katherine Silliman, PhD](https://ksilliman.weebly.com/), Auburn University - kes0132[at]auburn.edu
+#### [Sam Bogan](http://www.hofmannlab.com/sam-bogan.html), UC-Santa Barbara -snbogan[at]ucsb.edu
+#### [Alan Downey-Wall](https://cos.northeastern.edu/people/alan-downey-wall/), Northeastern University 
+#### Chris Mantegna, University of Washington  
+#### [Danielle Davenport](https://www.researchgate.net/profile/Danielle-Davenport-2), University of Queensland  
+#### [Hanny Rivera, PhD](https://www2.whoi.edu/staff/hrivera/), Woods Hole Oceanographic Institute  
+#### Jecy Klinkam, College of William and Mary  
+#### [Joe McGirr, PhD](https://joemcgirr.github.io/), UC-Davis 
+#### Kaitlin Macaranas, CSU-Bakersfield 
+#### Megan Guidry, University of Rhode Island
+#### Natalie Ameral, University of Rhode Island & RI Dept. of Environmental Management 
+#### [Sara Schaal](https://schaals.wixsite.com/smschaal), Northeastern University 
+#### [Sara Wuitchik, PhD](https://sjswuitchik.weebly.com/), Harvard University
+
+### Faculty Advisors
+#### [Andrew Whitehead, PhD](https://whiteheadresearch.wordpress.com/), UC-Davis 
+#### [Jon Puritz, PhD](https://www.marineevoeco.com/), University of Rhode Island 
+#### [Steven Roberts, PhD](http://faculty.washington.edu/sr320/?page_id=10850), University of Washington
 
 # Contact 
-
-Katherine Silliman (email)
-Sam Bogan (email)
+For more information about the working group or panel seminars, contact Katherine Silliman: kes0132[at]auburn.edu  
+For more information about the reproducibility review and undergraduate internship, contact Sam Bogan: snbogan[at]ucsb.edu
 
@@
 
-setwd("C:/Users/jmcgirr/Documents/GitHub/MarineOmics/website/")
-KnitPost("C:/Users/jmcgirr/Documents/GitHub/MarineOmics/website/index.Rmd")
+setwd("~/Postdoc_Projects/RCN_MarineOmics/Website/marineomics.github.io/")
+KnitPost("~/Postdoc_Projects/RCN_MarineOmics/Website/marineomics.github.io/index.Rmd")
 

### 2021-05-17 index.Rmd
Updated panelists and members

@@
 
-Everyone is welcome to join!
+Everyone is welcome to join! All seminars will be recorded and available on this website.
 
@@
 | May 7         | GBS/RADseq genotyping pipelines     | [Melanie LaCava](https://melanielacava.weebly.com/), [Alicia Mastretta-Yanes](http://mastrettayanes-lab.org/), [Isaac Overcast](https://isaacovercast.github.io/), [Jon Puritz](https://www.marineevoeco.com/)   |
-| May 21       |  WGS genotyping (low and high coverage)     | [Mikhail "Misha" Matz](https://matzlab.weebly.com/), [Tim Sackton](https://scholar.harvard.edu/tsackton), [Nina Overgaard Therkildsen](https://www.therkildsenlab.com/), TBD |
-| June 4       |  Population genomic analyses and filtering data     | TBD|
-| June 18       |  Functional Genomics (RNAseq)     | TBD|
+| May 21       |  WGS genotyping (low and high coverage)     | [Jeffrey Good](http://www.thegoodlab.org/), [Mikhail "Misha" Matz](https://matzlab.weebly.com/), [Noah Reid](http://nreid.github.io/), [Tim Sackton](https://scholar.harvard.edu/tsackton), [Nina Overgaard Therkildsen](https://www.therkildsenlab.com/) |
+| June 4       |  Population genomic analyses and filtering data     | [Zach Gompert](https://gompertlab.com/), [Morgan Kelly](http://www.morgankelly.biology.lsu.edu/), [Katie Lotterhos](https://sites.google.com/site/katielotterhos/home), [Jeff Ross-Ibarra](https://rilab.ucdavis.edu/) |
+| June 18       |  Functional Genomics (RNAseq)     | [Ana Conesa](http://conesalab.org/), [Joanna Kelley](https://labs.wsu.edu/genomes/), TBD|
 
@@
 The results of this survey will be shared in a peer-reviewed article, along with guidance to improve the state of reproducibility in the field.
+
 # Working Group Members
@@
 #### Jecy Klinkam, College of William and Mary  
+#### [Joanna Griffiths, PhD](https://joannasgriffiths.wordpress.com/), UC-Davis
 #### [Joe McGirr, PhD](https://joemcgirr.github.io/), UC-Davis 
@@
 #### Natalie Ameral, University of Rhode Island & RI Dept. of Environmental Management  
+#### Nicolas Lou, Cornell University
 #### [Sara Schaal](https://schaals.wixsite.com/smschaal), Northeastern University 

### 2021-05-19 index.Rmd
Update index.Rmd

@@
 | June 4       |  Population genomic analyses and filtering data     | [Zach Gompert](https://gompertlab.com/), [Morgan Kelly](http://www.morgankelly.biology.lsu.edu/), [Katie Lotterhos](https://sites.google.com/site/katielotterhos/home), [Jeff Ross-Ibarra](https://rilab.ucdavis.edu/) |
-| June 18       |  Functional Genomics (RNAseq)     | [Ana Conesa](http://conesalab.org/), [Joanna Kelley](https://labs.wsu.edu/genomes/), TBD|
+| June 18       |  Functional Genomics (RNAseq)     | [Ana Conesa](http://conesalab.org/), [Joanna Kelley](https://labs.wsu.edu/genomes/), [Martin Hölzer](https://hoelzer-lab.github.io/), TBD|
 

### 2021-05-19 index.Rmd
Updated RNAseq panel info

--- 2021-05-17 index.Rmd
@@
 | June 4       |  Population genomic analyses and filtering data     | [Zach Gompert](https://gompertlab.com/), [Morgan Kelly](http://www.morgankelly.biology.lsu.edu/), [Katie Lotterhos](https://sites.google.com/site/katielotterhos/home), [Jeff Ross-Ibarra](https://rilab.ucdavis.edu/) |
-| June 18       |  Functional Genomics (RNAseq)     | [Ana Conesa](http://conesalab.org/), [Joanna Kelley](https://labs.wsu.edu/genomes/), TBD|
+| June 18       |  RNAseq (experimental design through differential expression)     | [Ana Conesa](http://conesalab.org/), [Joanna Kelley](https://labs.wsu.edu/genomes/), [Martin Hölzer](https://hoelzer-lab.github.io/), TBD|
 

### 2021-05-19 index.Rmd
Add link to RAD panel video

--- 2021-05-17 index.Rmd
@@
 
-| Date        | Topic       | Panelists     |
-| :---        | :---        | :---          |
-| May 7         | GBS/RADseq genotyping pipelines     | [Melanie LaCava](https://melanielacava.weebly.com/), [Alicia Mastretta-Yanes](http://mastrettayanes-lab.org/), [Isaac Overcast](https://isaacovercast.github.io/), [Jon Puritz](https://www.marineevoeco.com/)   |
-| May 21       |  WGS genotyping (low and high coverage)     | [Jeffrey Good](http://www.thegoodlab.org/), [Mikhail "Misha" Matz](https://matzlab.weebly.com/), [Noah Reid](http://nreid.github.io/), [Tim Sackton](https://scholar.harvard.edu/tsackton), [Nina Overgaard Therkildsen](https://www.therkildsenlab.com/) |
-| June 4       |  Population genomic analyses and filtering data     | [Zach Gompert](https://gompertlab.com/), [Morgan Kelly](http://www.morgankelly.biology.lsu.edu/), [Katie Lotterhos](https://sites.google.com/site/katielotterhos/home), [Jeff Ross-Ibarra](https://rilab.ucdavis.edu/) |
-| June 18       |  Functional Genomics (RNAseq)     | [Ana Conesa](http://conesalab.org/), [Joanna Kelley](https://labs.wsu.edu/genomes/), TBD|
+| Date        | Topic       | Panelists     | Video     |
+| :---        | :---        | :---          | :---      |
+| May 7         | GBS/RADseq genotyping pipelines     | [Melanie LaCava](https://melanielacava.weebly.com/), [Alicia Mastretta-Yanes](http://mastrettayanes-lab.org/), [Isaac Overcast](https://isaacovercast.github.io/), [Jon Puritz](https://www.marineevoeco.com/)   | [View video](https://www.youtube.com/watch?v=C74GBESeIq4)     |
+| May 21       |  WGS genotyping (low and high coverage)     | [Jeffrey Good](http://www.thegoodlab.org/), [Mikhail "Misha" Matz](https://matzlab.weebly.com/), [Noah Reid](http://nreid.github.io/), [Tim Sackton](https://scholar.harvard.edu/tsackton), [Nina Overgaard Therkildsen](https://www.therkildsenlab.com/) |     |
+| June 4       |  Population genomic analyses and filtering data     | [Zach Gompert](https://gompertlab.com/), [Morgan Kelly](http://www.morgankelly.biology.lsu.edu/), [Katie Lotterhos](https://sites.google.com/site/katielotterhos/home), [Jeff Ross-Ibarra](https://rilab.ucdavis.edu/) |     |
+| June 18       |  RNAseq (experimental design through differential expression)     | [Ana Conesa](http://conesalab.org/), [Joanna Kelley](https://labs.wsu.edu/genomes/), [Martin Hölzer](https://hoelzer-lab.github.io/), TBD     |     |
 

### 2021-05-24 index.Rmd
Update RNA-seq panelists

--- 2021-05-19 index.Rmd
@@
 
-| Date        | Topic       | Panelists     |
-| :---        | :---        | :---          |
-| May 7         | GBS/RADseq genotyping pipelines     | [Melanie LaCava](https://melanielacava.weebly.com/), [Alicia Mastretta-Yanes](http://mastrettayanes-lab.org/), [Isaac Overcast](https://isaacovercast.github.io/), [Jon Puritz](https://www.marineevoeco.com/)   |
-| May 21       |  WGS genotyping (low and high coverage)     | [Jeffrey Good](http://www.thegoodlab.org/), [Mikhail "Misha" Matz](https://matzlab.weebly.com/), [Noah Reid](http://nreid.github.io/), [Tim Sackton](https://scholar.harvard.edu/tsackton), [Nina Overgaard Therkildsen](https://www.therkildsenlab.com/) |
-| June 4       |  Population genomic analyses and filtering data     | [Zach Gompert](https://gompertlab.com/), [Morgan Kelly](http://www.morgankelly.biology.lsu.edu/), [Katie Lotterhos](https://sites.google.com/site/katielotterhos/home), [Jeff Ross-Ibarra](https://rilab.ucdavis.edu/) |
-| June 18       |  Functional Genomics (RNAseq)     | [Ana Conesa](http://conesalab.org/), [Joanna Kelley](https://labs.wsu.edu/genomes/), [Martin Hölzer](https://hoelzer-lab.github.io/), TBD|
+| Date        | Topic       | Panelists     | Video     |
+| :---        | :---        | :---          | :---      |
+| May 7         | GBS/RADseq genotyping pipelines     | [Melanie LaCava](https://melanielacava.weebly.com/), [Alicia Mastretta-Yanes](http://mastrettayanes-lab.org/), [Isaac Overcast](https://isaacovercast.github.io/), [Jon Puritz](https://www.marineevoeco.com/)   | [View video](https://www.youtube.com/watch?v=C74GBESeIq4)     |
+| May 21       |  WGS genotyping (low and high coverage)     | [Jeffrey Good](http://www.thegoodlab.org/), [Mikhail "Misha" Matz](https://matzlab.weebly.com/), [Noah Reid](http://nreid.github.io/), [Tim Sackton](https://scholar.harvard.edu/tsackton), [Nina Overgaard Therkildsen](https://www.therkildsenlab.com/) |     |
+| June 4       |  Population genomic analyses and filtering data     | [Zach Gompert](https://gompertlab.com/), [Morgan Kelly](http://www.morgankelly.biology.lsu.edu/), [Katie Lotterhos](https://sites.google.com/site/katielotterhos/home), [Jeff Ross-Ibarra](https://rilab.ucdavis.edu/) |     |
+| June 18       |  RNAseq (experimental design through differential expression)     | [Ana Conesa](http://conesalab.org/), [Sarah Davies](http://sites.bu.edu/davieslab/), [Martin Hölzer](https://hoelzer-lab.github.io/), [Joanna Kelley](https://labs.wsu.edu/genomes/), [Harold Pimentel](https://pimentellab.com/)     |     |
 

### 2021-06-04 index.Rmd
Add WGS video link

--- 2021-05-19 index.Rmd
@@
 
-| Date        | Topic       | Panelists     |
-| :---        | :---        | :---          |
-| May 7         | GBS/RADseq genotyping pipelines     | [Melanie LaCava](https://melanielacava.weebly.com/), [Alicia Mastretta-Yanes](http://mastrettayanes-lab.org/), [Isaac Overcast](https://isaacovercast.github.io/), [Jon Puritz](https://www.marineevoeco.com/)   |
-| May 21       |  WGS genotyping (low and high coverage)     | [Jeffrey Good](http://www.thegoodlab.org/), [Mikhail "Misha" Matz](https://matzlab.weebly.com/), [Noah Reid](http://nreid.github.io/), [Tim Sackton](https://scholar.harvard.edu/tsackton), [Nina Overgaard Therkildsen](https://www.therkildsenlab.com/) |
-| June 4       |  Population genomic analyses and filtering data     | [Zach Gompert](https://gompertlab.com/), [Morgan Kelly](http://www.morgankelly.biology.lsu.edu/), [Katie Lotterhos](https://sites.google.com/site/katielotterhos/home), [Jeff Ross-Ibarra](https://rilab.ucdavis.edu/) |
-| June 18       |  RNAseq (experimental design through differential expression)     | [Ana Conesa](http://conesalab.org/), [Joanna Kelley](https://labs.wsu.edu/genomes/), [Martin Hölzer](https://hoelzer-lab.github.io/), TBD|
+| Date        | Topic       | Panelists     | Video     |
+| :---        | :---        | :---          | :---      |
+| May 7         | GBS/RADseq genotyping pipelines     | [Melanie LaCava](https://melanielacava.weebly.com/), [Alicia Mastretta-Yanes](http://mastrettayanes-lab.org/), [Isaac Overcast](https://isaacovercast.github.io/), [Jon Puritz](https://www.marineevoeco.com/)   | [View video](https://www.youtube.com/watch?v=C74GBESeIq4)     |
+| May 21       |  WGS genotyping (low and high coverage)     | [Jeffrey Good](http://www.thegoodlab.org/), [Mikhail "Misha" Matz](https://matzlab.weebly.com/), [Noah Reid](http://nreid.github.io/), [Tim Sackton](https://scholar.harvard.edu/tsackton), [Nina Overgaard Therkildsen](https://www.therkildsenlab.com/) | [View Video](https://www.youtube.com/watch?v=ObJ3OT3dwSw)    |
+| June 4       |  Population genomic analyses and filtering data     | [Zach Gompert](https://gompertlab.com/), [Morgan Kelly](http://www.morgankelly.biology.lsu.edu/), [Katie Lotterhos](https://sites.google.com/site/katielotterhos/home), [Jeff Ross-Ibarra](https://rilab.ucdavis.edu/) |     |
+| June 18       |  RNAseq (experimental design through differential expression)     | [Ana Conesa](http://conesalab.org/), [Sarah Davies](http://sites.bu.edu/davieslab/), [Martin Hölzer](https://hoelzer-lab.github.io/), [Joanna Kelley](https://labs.wsu.edu/genomes/), [Harold Pimentel](https://pimentellab.com/)     |     |
 

### 2021-06-11 index.Rmd
added video link for panel 3

--- 2021-05-19 index.Rmd
@@
 | May 7         | GBS/RADseq genotyping pipelines     | [Melanie LaCava](https://melanielacava.weebly.com/), [Alicia Mastretta-Yanes](http://mastrettayanes-lab.org/), [Isaac Overcast](https://isaacovercast.github.io/), [Jon Puritz](https://www.marineevoeco.com/)   | [View video](https://www.youtube.com/watch?v=C74GBESeIq4)     |
-| May 21       |  WGS genotyping (low and high coverage)     | [Jeffrey Good](http://www.thegoodlab.org/), [Mikhail "Misha" Matz](https://matzlab.weebly.com/), [Noah Reid](http://nreid.github.io/), [Tim Sackton](https://scholar.harvard.edu/tsackton), [Nina Overgaard Therkildsen](https://www.therkildsenlab.com/) |     |
-| June 4       |  Population genomic analyses and filtering data     | [Zach Gompert](https://gompertlab.com/), [Morgan Kelly](http://www.morgankelly.biology.lsu.edu/), [Katie Lotterhos](https://sites.google.com/site/katielotterhos/home), [Jeff Ross-Ibarra](https://rilab.ucdavis.edu/) |     |
-| June 18       |  RNAseq (experimental design through differential expression)     | [Ana Conesa](http://conesalab.org/), [Joanna Kelley](https://labs.wsu.edu/genomes/), [Martin Hölzer](https://hoelzer-lab.github.io/), TBD     |     |
+| May 21       |  WGS genotyping (low and high coverage)     | [Jeffrey Good](http://www.thegoodlab.org/), [Mikhail "Misha" Matz](https://matzlab.weebly.com/), [Noah Reid](http://nreid.github.io/), [Tim Sackton](https://scholar.harvard.edu/tsackton), [Nina Overgaard Therkildsen](https://www.therkildsenlab.com/) | [View Video](https://www.youtube.com/watch?v=ObJ3OT3dwSw)    |
+| June 4       |  Population genomic analyses and filtering data     | [Zach Gompert](https://gompertlab.com/), [Morgan Kelly](http://www.morgankelly.biology.lsu.edu/), [Katie Lotterhos](https://sites.google.com/site/katielotterhos/home), [Jeff Ross-Ibarra](https://rilab.ucdavis.edu/) |  [View Video](https://youtu.be/dZyuqVEGPMo)   |
+| June 18       |  RNAseq (experimental design through differential expression)     | [Ana Conesa](http://conesalab.org/), [Sarah Davies](http://sites.bu.edu/davieslab/), [Martin Hölzer](https://hoelzer-lab.github.io/), [Joanna Kelley](https://labs.wsu.edu/genomes/), [Harold Pimentel](https://pimentellab.com/)     |     |
 

### 2021-06-28 index.Rmd
Add RNAseq video

--- 2021-05-24 index.Rmd
@@
 | May 7         | GBS/RADseq genotyping pipelines     | [Melanie LaCava](https://melanielacava.weebly.com/), [Alicia Mastretta-Yanes](http://mastrettayanes-lab.org/), [Isaac Overcast](https://isaacovercast.github.io/), [Jon Puritz](https://www.marineevoeco.com/)   | [View video](https://www.youtube.com/watch?v=C74GBESeIq4)     |
-| May 21       |  WGS genotyping (low and high coverage)     | [Jeffrey Good](http://www.thegoodlab.org/), [Mikhail "Misha" Matz](https://matzlab.weebly.com/), [Noah Reid](http://nreid.github.io/), [Tim Sackton](https://scholar.harvard.edu/tsackton), [Nina Overgaard Therkildsen](https://www.therkildsenlab.com/) |     |
-| June 4       |  Population genomic analyses and filtering data     | [Zach Gompert](https://gompertlab.com/), [Morgan Kelly](http://www.morgankelly.biology.lsu.edu/), [Katie Lotterhos](https://sites.google.com/site/katielotterhos/home), [Jeff Ross-Ibarra](https://rilab.ucdavis.edu/) |     |
-| June 18       |  RNAseq (experimental design through differential expression)     | [Ana Conesa](http://conesalab.org/), [Sarah Davies](http://sites.bu.edu/davieslab/), [Martin Hölzer](https://hoelzer-lab.github.io/), [Joanna Kelley](https://labs.wsu.edu/genomes/), [Harold Pimentel](https://pimentellab.com/)     |     |
+| May 21       |  WGS genotyping (low and high coverage)     | [Jeffrey Good](http://www.thegoodlab.org/), [Mikhail "Misha" Matz](https://matzlab.weebly.com/), [Noah Reid](http://nreid.github.io/), [Tim Sackton](https://scholar.harvard.edu/tsackton), [Nina Overgaard Therkildsen](https://www.therkildsenlab.com/) | [View Video](https://www.youtube.com/watch?v=ObJ3OT3dwSw)    |
+| June 4       |  Population genomic analyses and filtering data     | [Zach Gompert](https://gompertlab.com/), [Morgan Kelly](http://www.morgankelly.biology.lsu.edu/), [Katie Lotterhos](https://sites.google.com/site/katielotterhos/home), [Jeff Ross-Ibarra](https://rilab.ucdavis.edu/) |  [View Video](https://youtu.be/dZyuqVEGPMo)   |
+| June 18       |  RNAseq (experimental design through differential expression)     | [Ana Conesa](http://conesalab.org/), [Sarah Davies](http://sites.bu.edu/davieslab/), [Martin Hölzer](https://hoelzer-lab.github.io/), [Joanna Kelley](https://labs.wsu.edu/genomes/), [Harold Pimentel](https://pimentellab.com/)     |  [View Video](https://youtu.be/-gQ59m2cck0)   |
 

### 2021-10-26 RAD_code/RADpage_text.Rmd
First test of website layout, organizing RAD code to generate figures in site

---
layout: default
title: RADseq
parent: Choosing a PopGenomics Approach
grand_parent: Best Principles
#editor_options: 
#  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#set bash code chunks to use bash_profile
knitr::opts_chunk$set(engine.opts = list(bash = "-l"))

if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

#BiocManager::install("SeqArray")
#BiocManager::install("SNPRelate")
#install.packages("rlang")
```

# RADseq  
{: .no_toc }

<details open markdown="block">
  <summary>
    Table of contents
  </summary>
  {: .text-delta }
1. TOC
{:toc}
</details>

If you want to run the acompanying code chunks, install Docker container.
```{r}
library(ggplot2)
library(SeqArray)
library(tidyverse)
```

<!-- `r colorize("Dani to do: make a dockerfile with all the packages used in this Rscript for reproducibility", "red")` -->

<!-- `r colorize("Example files needed: example strata file 'data/example.metadata.csv', example vcf for PCA './data/test.vcf' ", "red")` -->

# Reduced Representation Sequencing (RADseq/GBS)  
<!-- `r colorize("What kinds of protocols is this advice relevant to?", "red")`   -->
"Restriction-site   Associated   DNA   sequencing"  - RADseq - combines  restriction  enzymes  with next-gen, massively parallel,  short-read  sequencing. RADseq involves the use of restriction enzymes which are used to shear DNA at selected sites. RADseq comes in different flavors. Double-digest RADseq (ddRAD; Peterson et al.,2012)  selects  markers  with  two  restriction enzymes  with  different  cut  frequencies. Genotype-by-Sequencing  (GBS;  Elshire  et  al.,  2011)  uses  a  frequent-cutting  restriction  enzyme  with PCR size selection. See this review for a comparison of some of the more popular protocols (link). In this guide we use the term "RADseq" to refer to any of these protocols, including those that don't involve random shearing of data. When recommendations are specific to a certain type of RAD/GBS, we will explicitly say so.

<!-- `r colorize("characteristics of radseq that are different/ important to know compared to other types of sequencing ", "red")`   -->

Using RADseq to generate single nucleotide polymorphisms (SNPs) involves 1) library preparation in the lab, 2) bioinformatic processing through assembly and/or mapping to a reference, then 3) filtering of SNPs and individuals for quality. All of these steps can (and will) introduce some error, so the goal is to minimize this error through mitigation steps at all three parts of the process. Every dataset is different [add more about philosophy over best practices].   
Much of this guide is directly inspired by this [excellent review paper by O'Leary et al. (2018)](link), especially the section on minimizing errors due to library prep. We recommend reading and cross-referencing with this paper, and citing it if you follow its suggestions. Table 1 from this paper summarizes the various potential issues that can arise from RAD datasets and some mitigation steps. The goal of this guide is to expand on the O'Leary paper and provide some example code to help implement quality control and mitigation steps.

![Table 1 from O'Leary et al. 2018.](../RAD_code/images/oleary_table1.png)

## Considerations During Lab Work -- Setting up for Success {#labwork}

There are steps you can take before you even sequence RAD libraries that can help minimize issues downstream. Here, we use "library" to refer to a set of RADseq fragments from a group of individuals that are barcoded and sequenced together on a single lane or group of lanes. While specific RAD library prep methods have their own nuances for minimizing error, there are some steps you can take that are common across methods.

* If this RAD/GBS method has not been done in your species or in your molecular lab setup before, spend some time optimizing the protocol using a representative subset of individuals. Then try to keep everything about the library prep as consistent as possible across samples (eg, DNA extraction kit, PCR cycles, sequencing platform). This isn't always possible, especially if you need to optimize the protocol for certain tricky samples. Just make sure to keep track of everything!   
* Randomize samples across library prep batches and sequencing lanes! For example, if you are sequencing two different groups of samples on two different sequencing lanes, make sure they are randomized in respect to sample location or whatever your groups of interest are. If you are preparing groups of individuals in different batches to be pooled later, randomize across these batches. Alicia MY even recommends randomizing across DNA extractions, especially if the person doing the extractions is new to molecular work. This is to allow you to control for potential [batch effects](#batch) that are often observed with RAD data.  
```{r}
# This code reads in a comma-delimited STRATA/meta-data file, randomises samples, and returns a file with wells and plates and duplicates which may be used to assist with plating libraries
# ** TODO** will need to consider where to duplicate samples - need to consider STARTA, DOC, ect..?
# maybe have user write in a list of samples to replicate

wells = 96 # how many wells can you use on your plate?
data.in = read.table("./data/example.metadata.csv", sep = ",", header = T) # each sample should be in its own ROW
l = length(data.in[,1])
plates = rep(1:length(ceiling(101/ wells)), wells)
data.out = cbind(
   data.in[sample(1:l, l, replace = F),], # randomise samples
   PLATEID = plates[c(1:l)],  # add plate info
   WELLID = rep(1:wells, length(unique(plates)))[1:l] # add well info
) # add duplicates **TODO**

# print example
 print(head(data.out))

#save to file

```
* Keep track of all potential batch effect sources in a Metadata file (eg, storage conditions of tissue/DNA, date/method of DNA extraction, date/method library prep batch, sequencing lane).   
* Have a core set of 2-4 technical sample replicates across all libraries. Ideally these are true technical replicates from library preparation to sequencing, but even just sequencing replicates can be useful for downstream quality control.

## Principles for Analyzing Your Data

### Steps for a robust RAD analysis  
This is just one approach for working through your data. Some people will prefer to run just a subset of samples through a pipeline at first and evaluate parameters, then run all the samples through. Either way, be prepared to make MULTIPLE assemblies and go through this process iteratively, especially if this is a new-to-you study system.

1) [Look at your raw data.](#fastqc)
2) [Run an assembly pipeline](link assembly) through to a SNP dataset, using either all samples or a representative subset. Parameters can come from those used in a similar study, or default parameters.
3) Filter your data minimally, then [evaluate for potential sources of error.](#error)
4) Subset or remove individuals based on initial evaluation.  
5) Using a representative subset of samples, [test key parameters](#test) to optimize your assembly.
6) Run your optimized assembly on all non-removed samples.
7) Evaluate the difference between multiple filtering schemes for your analyses of interest. *Popgen analysis guide coming soon*
8) Repeat as needed.

### First, look at the raw data! {#fastqc}  
Always look at your data with fastqc(link) before starting an assembly. First, this is a good check to just make sure the sequencing worked. If you have demultiplexed data, you can use multiqc(link) to generate fastqc plots for all individuals and quickly identify ones that did not sequence well.  
Check out [this informative tutorial](https://datacarpentry.org/wrangling-genomics/02-quality-control/index.html) on running fastqc and interpreting the output.  
Do you have a lot of adapter sequences? Are the ends of your reads very low quality? If so, you should expect a fair amount of trimming and read filtering to occur prior to assembly. If that doesn't occur or if too many reads are being filtered so as to only recover a small number of SNPs, something might need to be tweaked with your trimming and filtering step. [See how to evaluate for sequencing error](#seq)

You should also look at the top few reads in the Terminal.  
```
## zcat: unZip and conCATenate the file to the screen
## head -n 20: Just take the first 8 lines of input

$ zcat raw-fastqs/BC2_10_C5.fastq.gz | head -n 20
@BC2_10_C5.1 1 length=96
CAGCGTGTAATAGTCACCGGCGGCTCCCTCTGGAGAATAGCACAAGTGATCATTTTGCTCATCTTCCGTCCACTGGTGATTGTGGACCAGCCTCAC
+BC2_10_C5.1 1 length=96
<GGGGGA<GGGGIIIGIGGIGGGIIIIIGGGGGGGGGGGGGGGGGIIIIIIIIGGIIIGGGGIGIIGIGGGIIIIIAGGGGIIIGIIGGGGGAAGG
@BC2_10_C5.2 2 length=96
CTGCTACATGCAGTGTTCTGTATTACTTTTATTGTACGTTGATATGAATGAATGAGTGTTTTGTATACTTAGAGTACAAGTTTGTCAGTCATATCG
+BC2_10_C5.2 2 length=96
GIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIGGIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIG
@BC2_10_C5.3 3 length=96
CAGCACATGTTCCTGTGTAGAAAGCTTGTTAGTAGAATAAATAACACATGGCTGGTCAAACACAACACATGAAGAAACAACTTTCTGAACAGTTTT
+BC2_10_C5.3 3 length=96
GIIGIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIGIIIIIIIIIIIIIIIIIIIIIIIIGGGGIIGIIIGGGII
@BC2_10_C5.4 4 length=96
CAGCGATTCGGCCCAAATTTGCACCACATCAGGCCCTTGACAGGGCGCTTCGATGGTGCAAATTTGGTGCGATTCGCTGCGCACCTAGCATATATG
+BC2_10_C5.4 4 length=96
AGGGIGIIIGIGGGIIGIGGGIIIGGGGAGGGGGAGGIGIGGIGGGGAGGGGGGGGGGGGGGGIIIII.G<GGIGGGIIIGGGGGIGGIIIGAGGI
@BC2_10_C5.5 5 length=96
CAGCAGTTTGGTGGAGTTCTGCAACCTTCCATTTCCAAAGAATTACCCAGGAGCTCTTCCCAGTGAATTTCTTCGGCACTTTTCATTGACCTTTTA
+BC2_10_C5.5 5 length=96
GGAGAAA.<AGAA.G.GGAGA<.GGA.<GAAGGAAGGIA<...<GA<..<G.<.<<.<AAAGGGG..<GGGGG<G.A.<GGGII.AG..<.GGGGG

```

If the sequencing center gave you one fastq file with all your samples, you should expect to see a barcode sequence, followed by the cutsite at the start of the read. If the Data are already demultiplexed (as the example is), you should only see the cutsite overhang (in this case, CWGC). Sometimes you can look at your fastq data files and see that there was a problem with the sequencing such that the cut site is either offset by one or more bases, or contains many errors. If this is not being addressed by the default filtering steps in your assembly pipeline, you can trim off N bases from the beginning or end of R1 and R2 reads in ipyrad (`trim_reads` param), or with cutadapt before using stacks, or customize the Trimmoatic step for dDocent.

### Evaluate potential confounding factors {#confound}

#### The power of PCA

One of the most powerful methods for exploring your data for issues is a Principle Components Analysis (see [here](https://michaelaalcorn.medium.com/yet-another-pca-explanation-727dff6ce26a) for a linear algebr-based explanation of PCA). From [this genetics tutorial](https://comppopgenworkshop2019.readthedocs.io/en/latest/contents/03_pca/pca.html): "To understand how PCA works, consider a single individual and its representation by its 593,124 markers. Formally, each individual is a point in a 593,124-dimensional space, where each dimension can take only the three possible genotypes indicated above, or have missing data. To visualise this high-dimensional dataset, we would like to project it down to two dimensions. But as there are many ways to project the shadow of a three-dimensional object on a two dimensional plane, there are many (and even more) ways to project a 593,124-dimensional cloud of points to two dimensions. What PCA does is figuring out the “best” way to do this project in order to visualise the major components of variance in the data.

As soon as you have a file with SNPs for all samples, before extensive filtering or playing with parameters, do a PCA on it and look at all the axes with the most variation. Be careful that you understand how missing data is imputed for PCA.

For the code examples in this section, we mostly make use of the R package SeqArray, which can read in and manipulate VCFs (*notes on Docker install*). We like this program because .... For those you are less familiar with R or more familar with Plink, check out [this tutorial]() which uses Plink and minimal R plotting.  a tutorial that uses We also include some examples using VCFtools and minimal R plotting, for those who are less familiar with R.

(?) dani asks: you want me to make this take multiple different vcfs from pipeline outputs for plotting? should we implement pcANGSD which can handle missing data better  (?)  
K: Making it take multiple vcfs is a secondary goal, and can be added later if we have time/interest. Adding a code example for pcANGSD would be useful though.

## ill get onto this sorry

Doing a PCA with SeqArray and SNPRelate in R. Filtering for SNPs with a MSF > 2.5%
```{r, eval =T}
## read in VCF and convert to GDS format
#vcf.file = "./data/test.vcf"
#seqVCF2GDS(vcf.file, "test.gds.rad", storage.option="ZIP_RA")

# read in existing gds file  
filename = "data/test.gds.rad"
gdsin= SeqArray::seqOpen(filename,)
# summarize the data
print(paste0("The number of SAMPLES in data: ", length(c(SeqArray::seqGetData(gdsin, "sample.id")))))
print(paste0("The number of SNPs in data: ",  length(c(SeqArray::seqGetData(gdsin, "variant.id")))))
#read in metadata
sample.ids = seqGetData(gdsin, "sample.id")
sample.strata =  read.table("data/test.metadata.csv", header = T, sep = "\t") %>%
                  dplyr::select(TARGET_ID, STRATA, LONG, LAT, PLATE)
id.order = sapply(sample.ids, function(x,df){which(df$TARGET_ID == x)}, df=sample.strata) #incase your strata file is not in the same order as your vcf
sample.strata.order = sample.strata[id.order,]

# get pruned SNPs
# NOT WORKING YET, issue with chromosome names
# Try this with vcf https://github.com/zhengxwen/SNPRelate/issues/10  
#snpset <- SNPRelate::snpgdsLDpruning(gdsin, ld.threshold=0.2, autosome.only = F, start.pos="random", num.thread=1)  
#snpset.id <- unlist(unname(snpset))

# PCA only on SNPs with a minor allele freq greater than 2.5%
pca.out = SNPRelate::snpgdsPCA(autosome.only = F, gdsin, num.thread=2, remove.monosnp = T, maf = 0.025)
#                               snp.id=snpset.id) # filtering for pruned SNPs
eig = pca.out$eigenval[!is.na(pca.out$eigenval)]
barplot(100*eig/sum(eig), main="PCA Eigenvalues")
```

* If you used sample replicates, first see if replicates cluster very close together. If not, then there may be batch effects or bioinformatic artifacts leading to genotyping error that must be addressed. Then remove replicates for a subsequent PCA evaluation.

```{r}
#look at PC values for reps
#remove reps
```

First color/shape the individuals by the factor you expect to matter (eg, sampling site or region, family, ecotype).  
```{r, eval=T}
#PLOT PCA
#PC1 v PC2 colored by collection location
print(
  as.data.frame(pca.out$eigenvect) %>%
      #tibble::add_column(., ID =  sample.ids) %>%
      tibble::add_column(., STRATA =  sample.strata.order$STRATA) %>%
      ggplot(., aes(x=V1, y=V2, color = STRATA))  + #label = ID
      geom_point(size=2) +
      #geom_text() +
      stat_ellipse(level = 0.95, size = 1) +
      #scale_color_manual(values = gds_colors) +
      geom_hline(yintercept = 0) +
      geom_vline(xintercept = 0) +
      theme_bw() +
      xlab(paste0("PC1 [",paste0(round(eig[1], 2)), "%]")) +
      ylab(paste0("PC2 [",paste0(round(eig[2], 2)), "%]")) +
      ggtitle("PCA Colored by Collection Location")
)
```

* Is there any clustering? If so, is it different than your expectation?
  + If different than expectation, suggests either [batch effects](#batch), [cryptic variation](#cryptic) among individuals, or systematic biological issues affecting assembly/mapping (eg, inversions (Katies inversion, also paper ref))  
  <!-- + `r colorize("figure example, weird clustering", "red")` -->
* Are there outlier samples driving a large amount of variation?
  + If so, they may be [cryptic species](#cryptic), clones/sample replicates, or they [sequenced/genotyped poorly](#bad)
  + In our example we see 2 potential outlier samples in the bottom left corner. We can check which samples they are by plotting the PCA with sample ID and/or looking at the actual PC scores.
```{r, eval=T}
#PLOT PCA
#PC1 v PC2 colored by collection location
print(
  as.data.frame(pca.out$eigenvect) %>%
      tibble::add_column(., ID =  sample.ids) %>%
      ggplot(., aes(x=V1, y=V2, label = ID)) +
      geom_text(size =1.5) +
      geom_hline(yintercept = 0) +
      geom_vline(xintercept = 0) +
      theme_bw() +
      xlab(paste0("PC1 [",paste0(round(eig[1], 2)), "%]")) +
      ylab(paste0("PC2 [",paste0(round(eig[2], 2)), "%]")) +
      ggtitle("PCA with Labels")
)
```
```{r}
# which samples have PC1 < -0.1 and PC2 < -0.6?
as.data.frame(pca.out$eigenvect) %>%
      tibble::add_column(., ID =  sample.ids) %>%
  filter(V1 < -0.1 & V2 < -0.6) %>%
  select(ID,V1,V2)
```

* Is there no clustering at all? If so, it may be due to:  
  + bioinformatic artifacts leading to noise, [see Testing Parameters](#test)
  + issues with how you are [filtering the data](#filter) (eg, too strict or too lax)
  + issues with how your PCA is treating missing data
  + or you actually have no structure in your dataset!
  <!-- + `r colorize("figure example, no clustering", "red")` -->

Next, color the individuals by potential sources of batch effects (sequencing lane, library prep batch, age of tissue sample, person doing the DNA extraction...).    
```{r, eval=T}
# again, but colored by batch
print(
  as.data.frame(pca.out$eigenvect) %>%
      tibble::add_column(., PLATE =  as.factor(sample.strata.order$PLATE)) %>%
      ggplot(., aes(x=V1, y=V2, color = PLATE))  + #label = ID
      geom_point(size=2) +
      stat_ellipse(level = 0.95, size = 1) +
      geom_hline(yintercept = 0) +
      geom_vline(xintercept = 0) +
      theme_bw() +
      xlab(paste0("PC1 [",paste0(round(eig[1], 2)), "%]")) +
      ylab(paste0("PC2 [",paste0(round(eig[2], 2)), "%]")) +
      ggtitle("PCA colored by Batch/Sequencing Plate")
)
```

* Do any observed clusters match a source of batch effect? [how to remove batch effects](#batch)

#### Visually inspect alignments {#align}  
Another useful approach is to visually inspect alignments for SNPs or loci that are identified as "weird" in the subsequent analyses. In ipyrad this can be done by searching the .loci file, in dDocent... If the alignment looks wonky, you can feel better about removing the whole locus (?).  
<!-- `r colorize("code example .loci", "red")` -->
<!-- `r colorize("code dDocent", "red")`   -->
<!-- `r colorize("code stacks", "red")`     -->

#### Batch effects {#batch}
As discussed in [Setting up for success](#labwork), batch effects can arise by minor (or not so minor) differences during the library prep and sequencing stage. Randomizing samples across libraries and sequencing lanes can help mitigate the influence of batch effects on your downstream analyses, but sometimes this isn't possible, esp. when using pre-existing data. Even if you randomize your samples, it is a good idea to check for batch effects with a PCA and then try to mitigate them.

<!-- `r colorize("PCA figure of batch effect", "red")` -->

You can identify which loci are driving the batch effect by 1) identifying the top PC loadings (SNPs) and/or 2) do an Fst outlier analysis with the batches specified as populations (eg, Bayescan). Both of these methods are most appropriate if you've randomized your samples between batches. If you have not randomized samples but are certain there is a batch effect driving variation in your dataset, you can still remove loci this way but you may also be removing biologically informative loci.

Nicolas Lou and Nina Therkildsen have a [great preprint](https://www.authorea.com/users/380682/articles/532568-batch-effects-in-population-genomic-studies-with-low-coverage-whole-genome-sequencing-data-causes-detection-and-mitigation) and [Github repo](https://github.com/therkildsen-lab/batch-effect) discussing causes, detection, and mitigation of batch effects in low coverage WGS, where batch effects can have particularly large effects. While some of their recommendations are specific to reference-based analyses, others are relevant to RAD methods too (eg, batch effects due to different sequencing platforms, DNA degradation levels, and sequencing depth).

Once you've identified these batch effect SNPs, it can be useful to [visualize the alignments](#align) before removing them. If the entire alignment for that locus looks wonky, you should remove all SNPs from that RAD locus from your final dataset before conducting analyses.     
<!-- `r colorize("code to identify top loading snps", "red")` -->

```{r top loading SNPS}
```

#### Cryptic species/contamination/clones {#cryptic}  
One issue that is particularly common, especially for marine invertebrates, are cryptic, yet highly diverged lineages occurring in the same geographic location. If you are trying to do a landscape/seascape genetics study, cryptic species can dramatically throw off population genetic analyses. A PCA using all the samples can help identify cryptic species if samples are forming clusters that do not align with any other possible factor (eg, geography, sequencing batch).   
<!-- `r colorize("PCA figure of cryptic species", "red")`   -->

```{r}
```

Note: Notice the huge variation explained along the first PC-axis.

You can also check heterozygosity, and missingness!

```{r}
```

 Note: Two outliers are obvious. Large differences in heterozygosity between samples are clear, with one location showing two distinct clusters of samples. These samples were later determined using mtDNA and morphological analysis to be cryptic species.
*Unpublished data from a marine species in Australian waters.*

You can then estimate genetic divergence between these clusters using pairwise Fst and/or phylogenetic inference to determine how diverged the groups are. If they are very diverged, it may be better to assemble each cryptic species separately, especially for denovo assemblies with dDocent and Stacks, especially since often times restriction cut-sites are not always conserved; see [Wagner et al., 2013](link), [Rubin et al., 2012](link)  These two pipelines used a set of given samples to create a "reference" or "catalog" to which all other samples are mapped. If highly diverged samples are included in creating this reference, it can lead to [oversplitting](#cluster).

Cryptic clones can also dramatically throw off population genetic analyses. In a PCA, clones will often appear as their own very tight cluster, similar to sample replicates. Below the PCA shows x and a simple calculate of pairwise-distance (i.e. here manhattan distance) shows x known duplicates, and x unknown clones. *other ways to tell they are clones*  
<!-- `r colorize("PCA figure of cryptic clones", "red")` -->

#### "Bad" samples {#bad}  
Sometimes a sample doesn't sequence well (few sequencing reads, higher error rate). This can be due to DNA quality, an issue during library prep, or not enough sequencing depth (average # of reads per sample). Generally, it will lead to a sample with fewer sequencing reads, higher missing data in a SNP dataset, and fewer shared loci with other samples. Identifying and then removing these samples *prior* to the final RADseq assembly analysis can help minimize mis-assembled loci, genotyping errors, and excessive filtering of acceptable loci.

The process of identifying low quality individuals is usually iterative, as the way you initially filter your SNPs will influence the amount of missing data and locus sharing among samples. This is why we recommend minimally filtering your SNPs for sample coverage (the # of individuals a locus is called in) when initially exploring your data. Some ways to identify bad samples:

* For every SNP dataset you generate, it is a good idea is always evaluate the missingness per sample (and report this distribution in your manuscript!). Identify samples with way more missingness than the rest, and observe how they look in a PCA and locus sharing plot. If they stick out, then try removing them from the assembly and seeing if it changes downstream analyses. If so, you may want to specify a missingness cutoff for including samples in the final analysis.
**vcftools example, on the command line**  
```{bash, eval = F}
vcftools --vcf data/stickleback_ferchaud_2015.vcf --missing-indv --out analysis/stickleback
# sort the file by most missing data and print the top 10  
cat analysis/stickleback.imiss | (read h; echo "$h"; sort -k5 -r) | head -n 10
```
##### Plot missingness {.tabset .tabset-fade .tabset-pills}
###### With R
```{r}
miss <- read.table("analysis/stickleback.imiss",header = T,sep="\t")
hist(miss$F_MISS,breaks=50)
```

###### With bash (requires gnuplot installed)
```{bash}
#code from Jon Puritz
mawk '!/IN/' analysis/stickleback.imiss | cut -f5 > totalmissing
gnuplot << \EOF
set terminal dumb size 120, 30
set autoscale
unset label
set title "Histogram of % missing data per individual"
set ylabel "Number of Occurrences"
set xlabel "% of missing data"
#set yr [0:100000]
binwidth=0.01
bin(x,width)=width*floor(x/width) + binwidth/2.0
plot 'totalmissing' using (bin($1,binwidth)):(1.0) smooth freq with boxes
pause -1
EOF
```
##### {-}

* Construct a locus sharing heatmap between pairs of samples, which shows the proportion of loci that are genotyped in both samples. Generally, samples that are more closely related will share more loci with each other due to cutsite dropout. Deviations from that pattern can pinpoint bad samples as well as batch effects.
  + [ipyrad API tutorial](https://ipyrad.readthedocs.io/en/latest/API-analysis/cookbook-sharing.html) for making a locus sharing heatmap, requires ipyrad output or ipyrad's vcf2hdf5 conversion  
  + [code to generate locus sharing/missingness heatmap from a VCF](https://github.com/atcg/clustOpt)

### Test a range of key parameters {#test}

#### Clustering threshold (de novo assembly) {#cluster}  
During de novo RADseq assembly, reads are assembled into contigs so that each contig represents a single locus. This can occur separately for each individual (ipyrad) or jointly for all or a subset of individuals (dDocent, Stacks). To call SNPs, in the case of ipyrad, contigs are then clustered and aligned across individuals, while in dDocent and Stacks, reads for each individual are mapped to the "reference" contigs. All of these pipelines use sequence similarity thresholds to determine if fragments are homologous (from the same place in the genome) and therefore should be clustered together. If the sequence similarity threshold is too low, *undersplitting* can occur so that multiple loci are combined into a single contig.  
<!-- `r colorize("figure example, undersplit", "red")`   -->
Undersplitting is common with repetitive regions and paralogs. It can inflate mean observed heterozygosity.

If the threshold is too high, it can lead to *oversplitting*, where alleles of the same locus are split into two or more contigs. This can lead to reduced mean observed heterozygosity.  
<!-- `r colorize("figure example, oversplit", "red")`   -->
If severely off, both of these issues can bias the models used to identify SNPs. Undersplitting a little is better than oversplitting, as it is easier to identify lumped paralogs downstream and remove them.  
**What influences the "optimal" clustering threshold of your data?**  
Clustering threshold is usually determined by how much genetic diversity you have in your dataset. If your dataset has high genetic diversity (eg., a single species with high heterozygosity, multiple highly diverged populations, multiple species), then you will need a lower clustering threshold. If your dataset has low genetic diversity (a single species with lower heterozygosity, a single family), then you will need a higher clustering threshold. Polyploid or highly repetitive (ie large) genomes may also have different requirements than diplod species with smaller genomes. Other factors, such as sequencing error rate, may play a role as well so it is usually a good idea to always test a few clustering thresholds with a new dataset. However, as long as your clustering threshold is in the ballpark it shouldn't affect downstream analyses (according to Isaac).

There are a few strategies to identify a good clustering threshold for your data, all of which require assembling your data with a range of clustering thresholds.  
1) Metrics developed by [McCartney-Melstad 2017](link), based on popgen theory and landscape genetic expectations  
2) Utilize sample replicates (either exact replicates or individuals from the same locality) to quantify genotyping error and minimize error while maximizing the number of informative loci [Mastretta-Yanes 2015](link)
3) Evaluating the number of haplotypes across loci, with many 3+ haplotypes indicating undersplitting [Harvey 2015](link), [Ilut 2014](link)  
4) Make assemblies at a range of thresholds and parameters, plot the # of SNPs or contigs per assembly, and make a judgement call [example in dDocent tutorial](link). If most parameters give you about the same # of SNPs or contigs, you can feel confident that this parameter doesn't matter much for your data and just choose a reasonable value.

#### Mapping parameters (reference-based and some de novo assemblers)   
When a sequencing read maps to the wrong assembled contig or place in a reference genome (ie, a non-homologous locus), this can result in erroneous SNPs. In many cases, these SNPs can be filtered out later (link to filtering). However, data should be checked to make sure there aren't systematic, wide-scale mapping issues that need to be addressed.

For referenced-based assembly, both dDocent and ipyrad make use of bwa(link) to map reads, while Stacks allows the user to choose their aligner. dDocent also utilizes bwa for de novo assemblies.   *

* how many reads aligned for each sample? If many reads didn't aligned, this could be an issue with the reference, read trimming/filtering, or the mapping parameters.
* What is the distribution of mapping scores?  
* ratio of mean (across inds) mapping quality scores for reference and alternate allele (O'Leary cite). 0.25-1.75 cut off implemented in dDocent.  
  + this info might not be coded in all VCFs

(link to WGS mapping quality section)

Mapping defaults:
dDocent: bwa mem -L 20,5 -t 8 -a -M -T 10 -A1 -B 3 -O 5 -R
L: clipping penalty for 5' and 3'
A: matching score
O gap open penalty
T: mapping quality filter
ipyrad: (defaults) -L 5,5 -B 4 -A1 -O 6 -T 30

#### Read depth threshold {#seq}

Mistakes that occur during sequencing can lead to a base-calling error in a read, which can then lead to an erroneous SNP. This type of error most commonly manifests as rare alleles (eg, only observed in one individual) or singletons (only one copy of the minor allele across all individuals). A conservative approach to deal with these errors is to filter out all loci with rare alleles (link filtering), but for some demographic analyses this filtering is inappropriate as they require the site frequency spectrum.  
The best ways to minimize the impact of sequencing error during bioinformatic processing is to 1) properly trim and filter your data, and 2) require higher read depth thresholds for assembly and genotyping.

2) Require higher read depth thresholds for assembly and genotyping.  
For RADseq, ideally you are generating an average of *at least* 10-20 sequence reads (10x-20x coverage) for each RAD locus in your library preparation (but this doesn't always happen...). For diploid species, this would be sequencing each RAD haplotype at least 5x-10x. The SNP calling step of RAD pipelines all require setting a minimum number of reads to determine whether a site is homozygous or heterozygous **add image example**. If this threshold is too low, sequencing errors are more likely to be included and the SNP caller will have a harder time accurately identifying SNPs ([but see our section on low coverage WGS methods!](link). If it's too high, then real polymorphic loci will be excluded. All of the three major pipelines allow users to implement read depth thresholds

Generating multiple sequencing reads per locus is helpful when calling SNPs in order to  distinguishing between reads with sequencing errors and

<!-- `r colorize("Fountain et al 2016 found error rates > 0.06 with minimum 20 read depth and Stacks v1. Can this be due to not optimizing other Stacks parameters? Hard to interpret/ know how relevant it is now, even to Stacks2", "red")` -->

#### Filtering reads

FASTQ files that typically come from the sequencing center include a [Phred scale quality score](link) for each base, which indicates the probability of that base call being correct. This information can be used prior to assembly to trim low-quality sections from either end of reads or remove reads altogether. Some SNP callers, including FreeBayes which is part of dDocent, also use Phred scores when calling SNPs (ipyrad and Stacks do not). All three of the most popular RAD pipelines include steps to filter reads due to low quality bases low quality bases or adapter sequences, but vary slightly on what kinds of read trimming are allowed. Always make sure to filter out adapter sequences (this is not default with stacks), otherwise default parameters are usually adequate. If your initial [fastqc](#fastqc) indicates that you have many low quality bases in your reads, you may lose too much of your data during filtering. In this case, you may want to make your read filtering a little more permissible (keeping in mind that this may increase SNP error rate).  
**Always make sure to check with your sequencing center to see if they have done anything to the raw reads before you received them!**
<!---turn these into html tabs--->

**dDocent: **

* does not allow you to use reads that have already been trimmed and filtered for de novo assembly, only referenced-based.
* Uses Trimmomatic to trim Illumina TruSeq adaters and low quality bases. Make sure to change the defaults if you used a different adapter/sequencing system.

**Stacks: **

* requires your reads to be all the same length, so if they were previously trimmed of low quality bases or adapter sequence this must be specified and they will be truncated to the same length.

### Filter your data many ways (and embrace the missingness!)
* Don't over filter for missingness
 + papers showing how too much filtering can skew results  
  + Empirical: Diaz-Arce 2019, Tripp 2017, Silliman 2021, Eaton 2017, Martin-Hernanz (2019)
  + Simulations: Huang & Knowles 2016, Leaché et al., 2015, Gautier et al., 2012, Rivera-Colón et al., 2021

* some ways to account for missing data in downstream analyses
 + impute (PCA), subsample (PCA, structure, pretty much anything with SNPs), try a few different missing filters and see what impact it has on results
 + use [pcANGSD](link)

## Assembly and Genotyping Pipelines  {#assembly}

There are a number of freely available pipelines for processing RADseq data, with the most popular being dDocent, Stacks, and ipyrad. These pipelines vary slightly in their underlying methodologies, customization options, and additional included analyses.

|  | dDocent | ipyrad | Stacks2 |
|---|---|---|---|
| Supported datatypes | Paired-end: ddRAD, ezRAD, RAD (random shearing), data with large overlap  between forward and reverse reads Single-end: any RADseq method  de novo and reference-based  If doing de novo assembly, reads cannot be trimmed outside of dDocent |  |  |
| Unique aspects | Novel data reduction approach used to inform coverage cutoffs |  |  |
| Documentation | Very good (esp. tutorials), active community support on Google Groups | Excellent (esp. installation, parameter explanations, and tutorials), active community support on Gitter | Very good, active community support on Google Groups |
| Speed/Accuracy | fastest and most accurate | close to dDocent accuracy, more over-splitting is possible  if parameters are not tuned | produces some untrue genome fragments (esp. with higher levels of indel polymorphism), but can be addressed with downstream filtering |
| Open source/development | Open source | Open source | no |
| Filtering options | Minimal default filtering as implemented in VCFtools. Ideal for those who want full freedom in filtering their SNPs. |  |  |
| Output options | Only produces a VCF, fasta of the de novo assembly, and individual BAM files  from mapping to the reference  |  |  |
| Popgen analyses | No additional popgen analyses included  |  |  |
|  |  |  |  |

Some groups have also developed pipelines for specific flavors of lab (e.g., Matz lab and 2bRAD) or proprietary software (eg, DARTseq), which we touch on below. **In most cases, it is recommended that you use a pipeline developed and tested for RAD data, especially if you are making a de novo assembly.** LaCava et al, 2019 have an excellent study where they review various de novo assemblers that are used in these pipelines. Assemblers that were not explicitly developed for short reads (eg, Velvet, ABySS) performed very poorly, while CD-HIT (the assembler in dDocent) performed the best. Stacks/Stacks2 and vsearch (the assembler in ipyrad) both performed worse when analyzing simulated data with indels. It should be noted that all of these RAD pipelines have optimized the specific parameters of these de novo assemblers to work with RAD data of various flavors, which may not be fully reflected in the LaCava study. Based on the MarineOmics RAD panel, the general consensus is that all the popular pipelines can produce adequately accurate datasets (with appropriate parameter optimization and data filtering).     
![Table 3 from LaCava et al 2019. CD-HIT is used in dDocent, VSEARCH is used in ipyrad.](../RAD_code/images/lacavaTable3.png)

### iPyrad walkthrough

Link to annotated ipyrad notebook

## Special Cases  
### DARTseq  
### 2bRAD

### Reference-based vs. de novo RADseq
<!---brief outline on general pipeline--->

## Good tutorials  
* [Marine Genomics open course](https://baylab.github.io/MarineGenomics/)
* [UNIX tutorial](http://evomics.org/learning/unix-tutorial/)
* [what is a VCF file and how can I filter it?](http://evomics.org/learning/population-and-speciation-genomics/2020-population-and-speciation-genomics/first-steps-in-genomic-data-analysis/)

``````{r, echo=FALSE,eval=FALSE}
library(knitr)
my.jekyll.site <- "marineomics.github.io"
KnitPost <- function(input, output) {
  base.dir <- "~/Postdoc_Projects/RCN_MarineOmics/Website/marineomics.github.io/"
  base.url <- "/"
  fig.path <- "RAD_code/images/"
  #opts_knit$set(base.url = base.url)
  #fig.path <- paste0("img/", sub(".Rmd$", "", basename(input)), "/")
  opts_knit$set(base.dir = base.dir, base.url = base.url)
  opts_chunk$set(fig.path = fig.path) 
  #opts_chunk$set(fig.path = "../RAD_code/images/")
  opts_chunk$set(fig.cap = "center")
  render_jekyll()
  knit(input, envir = parent.frame(), output = output)
}
setwd("~/Postdoc_Projects/RCN_MarineOmics/Website/marineomics.github.io/RAD_code")
KnitPost("~/Postdoc_Projects/RCN_MarineOmics/Website/marineomics.github.io/RAD_code/RADpage_text.Rmd",output = "~/Postdoc_Projects/RCN_MarineOmics/Website/marineomics.github.io/guidelines/RADpage_text.md")
```

### 2021-10-26 guidelines/WGS_intro.Rmd
First test of website layout, organizing RAD code to generate figures in site

---
layout: default
title: Whole Genome Resequencing
parent: Choosing a PopGenomics Approach
grand_parent: Best Principles
has_children: true
#editor_options: 
#  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Choosing a PopGenomics Approach  
{: .no_toc }

<details open markdown="block">
  <summary>
    Table of contents
  </summary>
  {: .text-delta }
1. TOC
{:toc}
</details>

## Whole Genome Resequencing for Population Genomics  
For the purposes of this guide, we use "whole genome (re)sequencing (WGS)) to refer to  methods where a reference genome exists (whether for the focal species or a related species), and samples are sequenced and then mapped to most or all of the reference genome. This method can provide high density SNPs, structural information...for population genomic analyses. This is separate from whole-genome de novo sequencing, which aims to produce a reference genome by sequencing and assembling a complete genome for the first time.   
WGS can be divided into three main approaches, and choosing between these or RADseq(link to ) mostly depends on your questions of interest.   
1) high coverage individual WGS (hcWGS)  
+ description  
2) low coverage individual WGS (lcWGS)  
+ description
3) pooled individuals (Poolseq)  
RADseq is... and m

 also a factor of your funds and how you choose to allocate them. There is a financial trade-off between the number of samples you can genotype individually and the amount of sequencing coverage per sample. If you have 1000s of samples, it may not be financially feasible to sequence most of them to high depth.  Large genome sizes can also make hcWGS cost-prohibited. This topic was discussed during the MarineOmics WGS panel seminar (link).    
An excellent and thorough review of the pros  between these approaches, esp. as they apply to nonmodel organisms, is [Fuentes-Pardo and Ruzzante 2017)(link).

### Tutorials

Here we provide a couple different tutorials by our working group members, as well as links to other detailed tutorials on the web. The goal of the MarineOmics tutorials are to provide extensive detail on "why" certain parameters are chosen, and some guidance on how to evaluate different parameter options to fit your data.

* Fastq-to-VCF pipeline(link): In-depth explanation of automated short-read mapping and variant calling pipeline maintained by [Harvard Informatics](https://github.com/harvardinformatics/shortRead_mapping_variantCalling). Useful for moderate to high coverage data, where *add description of why one would use pipeline*  
*

## Considerations Before Lab Work

```{r, echo=FALSE,eval=FALSE}
library(knitr)
my.jekyll.site <- "marineomics.github.io"
KnitPost <- function(input, base.url = my.jekyll.site) {
  opts_knit$set(base.url = base.url)
  fig.path <- paste0("images/", sub(".Rmd$", "", basename(input)), "/")
  opts_chunk$set(fig.path = fig.path)
  opts_chunk$set(fig.cap = "center")
  render_jekyll()
  knit(input, envir = parent.frame())
}

setwd("~/Postdoc_Projects/RCN_MarineOmics/Website/marineomics.github.io/guidelines/")
KnitPost("~/Postdoc_Projects/RCN_MarineOmics/Website/marineomics.github.io/guidelines/WGS_intro.Rmd")

```

### 2021-10-26 guidelines/guideline-intro.Rmd
First test of website layout, organizing RAD code to generate figures in site

---
layout: default
title: Best Principles
editor_options: 
  chunk_output_type: console
output:
  html_document
has_children: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# This content is still under construction!

# Best Principles for 'Omics in Nonmodel Systems

What are 'omics methods?

Best principles vs Best Practices.

Here are a collection of guidelines, tutorials, recommended readings, and method evaluations for various 'omics analyses that are commonly used in studies of nonmodel systems (particularly marine species).

## Omics Approaches {.tabset .tabset-fade .tabset-pills}
### Population Genomics

* [Choosing a population genomics sequencing approach](WGS_intro.md)  
* Whole Genome Resequencing  
* Reduced Representation Sequencing

### Functional Genomics

* Transcriptomics with RNA-Seq  
* Differential gene expression

## {-}

```{r, echo=FALSE,eval=FALSE}
library(knitr)
my.jekyll.site <- "marineomics.github.io"
KnitPost <- function(input, base.url = my.jekyll.site) {
  opts_knit$set(base.url = base.url)
  fig.path <- paste0("images/", sub(".Rmd$", "", basename(input)), "/")
  opts_chunk$set(fig.path = fig.path)
  opts_chunk$set(fig.cap = "center")
  render_jekyll()
  knit(input, envir = parent.frame())
}

setwd("~/Postdoc_Projects/RCN_MarineOmics/Website/marineomics.github.io/guidelines")
KnitPost("~/Postdoc_Projects/RCN_MarineOmics/Website/marineomics.github.io/guidelines/guideline-intro.Rmd")

```

### 2021-10-26 guidelines/popgen_intro.Rmd
First test of website layout, organizing RAD code to generate figures in site

---
layout: default
title: Choosing a PopGenomics Approach
parent: Best Principles
has_children: true
#editor_options: 
#  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Choosing a PopGenomics Approach  
{: .no_toc }

<details open markdown="block">
  <summary>
    Table of contents
  </summary>
  {: .text-delta }
1. TOC
{:toc}
</details>

## Choosing a Population Genomics Approach  
Deciding on a sequencing approach for a new project can be challenging, and depends primarily on the questions you are interested in, but also constraints set by funding, existing genomic resources, sample size, genome characteristics, and expertise in the group. Below we list a number of great review articles and empirical comparison studies that can help when making the decision. This topic was also discussed during the MarineOmics seminars (link).  
Presently, our site covers four approaches (WGS=whole genome (re)sequencing):

1) high coverage individual WGS (hcWGS)  
 i) description  
2) low coverage individual WGS (lcWGS)  
 i) description
3) pooled individuals (Poolseq)  
4) reduced-representation sequencing 
 i) can be split into restriction-based methods (eg RADseq) and targeted capture methods (eg exome capture)

For each approach, we provide guiding principles and tutorials for bioinformatic processing from raw data to genotypes. While we currently don't discuss library preparation methods in depth, we do include advice on how to organize your sequencing runs in order to improve quality control downstream.

It should be noted that you don't need to choose only one approach for all samples! If you are starting a research program on a species without any genomic resources, you may want to start with de novo RADseq to understand basic population structure and genetic diversity, then once a reference genome is assembled move to high/low coverage WGS. In the [MarineOmics seminar](link), Misha Matz recommended choosing one individual to use for genome assembly, doing high coverage WGS on a few other individuals that represent most of the variation in your taxa, and then WGS many other individuals at low coverage. The high coverage individuals can be useful for imputing missing genotypes in the low coverage samples.

References:

* Matz 2018: great intro to ecological genomics methods in species with few to no prior genomic resources, accessible to those just getting started.  
* Fuentes-Pardo and Ruzzante 2017: a VERY thorough review of the different approaches to generating population genomics data, with non-model species in mind. Some of the content about relative cost and available software is a little dated, but otherwise a very valuable reference.   
* Benjelloun et al 2019: empirical comparison of high coverage WGS, low coverage WGS, random variants (eg. from RADseq), exome capture, and commercial SNP chip for mammals (genome size=2.6Gb). For neutral genomic diversity, found 5k-10k random variants were enough to accurately estimate, but comercial panales/exome capture had acertainment bias. To detect selection and accurate estimates of LD, at least 1M variants were required. For the studied species, 5x coverage was just as accurate as 12x coverage. Low coverage WGS (< 5x) should use genotype likelihood methods, however the rate of false positives for heterozygotes is higher than classic genotyping methods at higher coverage.  
* Dorant et al. 2019: Case study comparing Poolseq vs. reduced representation seq (GBS) vs Rapture (targeted capture of RAD loci) for inferring population structure with weak genetic differentiation. All three detected same weak pop structure pattern, but Poolseq gave Fst values 3-5x higher than Rapture/GBS.

This table briefly summarizes the pros and cons of five sequencing approaches and their appropriateness for answering specific questions. If you are interested in answering multiple questions (eg, neutral population structure and adaptive variation) then you would often want a method that can do both (eg, WGS).

|  | hcWGS | lcWGS | RADseq | Poolseq | Targeted capture (eg, exome capture) |
|---|---|---|---|---|---|
| Pros | provides high quality, high density genotypes; can be used to improve development of a reference genome | provides high density genotypes at a reduced cost | generally cheaper per sample, density of genotypes can be tuned to fit question, does not require a reference | many individuals can be mixed for a low library prep cost, only option for some larval studies | allows consistent capture of loci between batches, good for sequencing targeted regions across many samples  |
| Cons | most expensive per sample, requires a reference genome, data requires large computational resources | requires a reference, can produce false positive heterozygotes, sensitive to batch effects, may be inappropriate to use individual SNP calls for some analyses | only 1-5% of genome covered which limits studying adaptive variation, de novo assembly can result in paralogs if not tuned or filtered well, can be hard to capture the same loci between batches | need higher sequencing depth so it can be more expensive than RAD (but less than other options), no info on individual genotypes (duh), requires a reference, requires some replicates to account for batch effects | expensive to start a new project and requires reference of some kind to design baits (unless using method like eecSeq), only gives info on targeted region so can result in ascertainment bias for some analyses |
| Population structure, genetic diversity summary stats | great, but if only interested in population structure then this may be $$$$ overkill | good, but may be limited to popgen analyses based on genotype likelihoods | great, esp. for many samples | good for methods that only use allele frequencies, won't work for individual based methods (eg Admixture) | ascertainment bias is likely to skew results, esp. for measures of genetic diversity |
| Demography (migration rates, population size through time) | great, may be overkill unless you incorporate haplotype-based methods | good, esp. with some high coverage samples to help with imputation | good for methods based on site frequency spectrum (eg moments), but not ideal for methods using extended haplotype or phasing info | active area of method development, but still not common except in cases of multiple temporal samples (cite) | likely not appropriate due to ascertainment bias |
| Signatures of selection, specific genomic regions of interest (GWAS) | great | good | Hotly debated, but unless your genome is way too big, WGS is the way to go | good if covering majority of the genome | only works for targeted regions |
| Phylogenetic inference | computationally challenging, but good at all divergence levels | good for tree shapes but not branch lengths, good for extracting organelle sequences and developing cost-effective tools (eg primers) for species and hybrid ID studies | good at shallow to medium divergence levels, but need to play with filtering parameters | not possible | an improvement over RAD as there will be less missing data among samples |
| Genetic crosses or mapping panels | likely $$$$ overkill | great if you have a genome | great as it allows many individuals | maybe | poor |
| molecular evolution (where accurate low-frequency alleles are required) | best | poor | poor | poor | good if only interested in targeted regions |
| Evolution of protein coding genes | great but requires decent annotations on genome | ? | poor, as you will miss many genes | ? | will miss introns and regulatory regions, but otherwise good |

```{r, echo=FALSE,eval=FALSE}
library(knitr)
my.jekyll.site <- "marineomics.github.io"
KnitPost <- function(input, base.url = my.jekyll.site) {
  opts_knit$set(base.url = base.url)
  fig.path <- paste0("images/", sub(".Rmd$", "", basename(input)), "/")
  opts_chunk$set(fig.path = fig.path)
  opts_chunk$set(fig.cap = "center")
  render_jekyll()
  knit(input, envir = parent.frame())
}

setwd("~/Postdoc_Projects/RCN_MarineOmics/Website/marineomics.github.io/guidelines/")
KnitPost("~/Postdoc_Projects/RCN_MarineOmics/Website/marineomics.github.io/guidelines/popgen_intro.Rmd")

```

### 2021-10-26 RAD_code/analysis_tools/clustOpt/ibdSlope.R

@@ -1,0 +1,45 @@
+suppressPackageStartupMessages(library(geosphere))
+suppressPackageStartupMessages(library(SNPRelate))
+
+### Make sure that there are exactly two flags--the VCF file input list [1] and the latlong file [2]
+args = commandArgs(trailingOnly=TRUE)
+if (length(args)!=2) {
+  stop("Exactly two arguments must be supplied (<vcf file> and <latLongFile>).n", call.=FALSE)
+}
+
+# Pull in the VCF filenames
+vcfFiles <- scan(args[1], what = character(), quiet=T)
+
+# Set up the geographic distance matrix:
+latLongFile <- read.csv(args[2], sep="\t", header=T)
+geoDistMatrix <- matrix(nrow=length(latLongFile$Sample), ncol=length(latLongFile$Sample))
+rownames(geoDistMatrix) <- latLongFile$Sample
+colnames(geoDistMatrix) <- latLongFile$Sample
+
+# Calculate the pairwise geographic distance between each pair of points, in km:
+for (i in 1:length(latLongFile$Sample)) {
+    for (j in 1:length(latLongFile$Sample)) {
+        geoDistMatrix[i,j] <- distm(c(latLongFile$Long[i], latLongFile$Lat[i]), c(latLongFile$Long[j], latLongFile$Lat[j]), fun = distHaversine) / 1000
+    }
+}
+
+for (i in 1:length(vcfFiles)) {
+    gdsOut <- paste(gsub(".vcf","", vcfFiles[i], fixed=T), '.pcaVar.gds', sep="")
+    snpgdsVCF2GDS(vcfFiles[i], gdsOut, verbose=F)
+    genofile <- snpgdsOpen(gdsOut)
+    ibsMat <- snpgdsIBS(genofile, num.thread=2, verbose=F, autosome.only=FALSE)
+
+#    cat(head(rownames(geoDistMatrix)))
+#    cat(head(ibsMat$sample.id))
+#    cat(match(rownames(geoDistMatrix), ibsMat$sample.id), "\n")
+
+
+    # Rearrange IBS matrix to the same order as geoDistMatrix
+    ibsMatRearranged <- ibsMat$ibs[match(rownames(geoDistMatrix), ibsMat$sample.id), match(rownames(geoDistMatrix), ibsMat$sample.id)]
+
[…]

### 2021-10-26 RAD_code/analysis_tools/clustOpt/resources/missingnessHeatMaps.R

@@ -1,0 +1,155 @@
+#!/usr/bin/env Rscript
+
+suppressPackageStartupMessages(library(SNPRelate))
+library(pheatmap)
+library(grImport2)
+suppressPackageStartupMessages(library(dendextend))
+
+### Make sure that there is exactly one flag--the VCF file input list
+args = commandArgs(trailingOnly=TRUE)
+if (length(args)!=1) {
+  stop("Exactly one argument must be supplied (<vcf file list>).n", call.=FALSE)
+}
+
+### Get and print a random seed for SNPRelate clustering
+randomSeed <- sample(1:100000000, 1)
+set.seed(randomSeed)
+paste("Random seed for snpgdsHCluster = ", randomSeed, sep="")
+
+### Core functions
+makeMatrix <- function(vcfFilename, printStats = 0) {
+    intermediateFileName <- paste(gsub(".vcf","", vcfFilename, fixed=T), '_missHM012.missingness', sep="")
+    missInter <- read.csv(intermediateFileName, sep="\t", header=T)
+
+    interMatrix <- matrix(nrow=length(unique(missInter$Sample1)), ncol=length(unique(missInter$Sample1)))
+    interMatrix[lower.tri(interMatrix, diag=T)] <- (missInter$oneMiss / (missInter$oneMiss + missInter$neitherMiss))
+    colnames(interMatrix) <- unique(missInter$Sample1)
+    rownames(interMatrix) <- unique(missInter$Sample2)
+    interMatrix[upper.tri(interMatrix, diag=F)] <- t(interMatrix)[upper.tri(interMatrix, diag=F)]
+    
+    if (printStats == 1) {
+        minMissingness <- min(interMatrix, na.rm=T)
+        maxMissingness <- max(interMatrix, na.rm=T)
+        cat(paste("Minimum pairwise missingness for ", vcfFilename, ": ", minMissingness, sep=""), "\n")
+        cat(paste("Maximum pairwise missingness for ", vcfFilename, ": ", maxMissingness, sep=""), "\n")
+    }
+    return(interMatrix)
+}
+
+makeVectorHeatmap <- function(missingnessMatrix, dendrogramz, name) {
+    # Make the heat map SVG
+    heatmapFile <- paste(name, ".heatmap.svg", sep = "")
+    svg(heatmapFile, height=4, width=4.44)
[…]

### 2021-10-26 RAD_code/analysis_tools/clustOpt/resources/pairwiseMissingnessFrom012.pl

@@ -1,0 +1,97 @@
+#!/usr/bin/perl
+
+use strict;
+use warnings;
+use autodie;
+use Getopt::Long;
+
+my $genotypesFile;
+my $indFile;
+my $outFile;
+my $help;
+
+GetOptions  ("genotypes=s"        => \$genotypesFile,
+             "individuals=s"      => \$indFile,
+             "out=s"              => \$outFile,
+             "help|man"     => \$help) || die "Couldn't get options with GetOpt::Long: $!\n";
+
+if (!$genotypesFile or !$indFile or !$outFile or $help) {
+    die "Usage: perl pairwiseMissingnessFromVCF.pl --genotypes <file.012> --individuals <file.indv> --out <file>";
+}
+
+# The indv file has one sample per line, in the same order as the rows in the 012 file
+my @samples;
+open(my $indFH, "<", $indFile);
+while (my $line = <$indFH>) {
+    chomp($line);
+    push(@samples, $line);
+}
+
+my %genotypesHash;
+open(my $genotypesFH, "<", $genotypesFile);
+while (my $line = <$genotypesFH>) {
+    chomp($line);
+    my @fields = split(/\t/, $line);
+    my $sampleIndex = shift(@fields);
+    my $sample = $samples[$sampleIndex];
+    foreach my $genotype (@fields) {
+        push(@{$genotypesHash{$sample}}, $genotype);
+    } 
+}
+
+# Get the number of loci from the first sample genotype array (this should be safe because there is always at least 1 
+# sample, and there should be the same number of loci for each sample:
+my $numLoci = scalar(@{$genotypesHash{$samples[0]}});
+foreach my $sample (@samples) {
+    if (scalar(@{$genotypesHash{$sample}}) != $numLoci) {
+        die "$sample does not have the same number of loci as $samples[0]. Check to make sure each sample has the same number of genotypes\n";
+    }
+}
+
+# Now do locus comparisons
+my %missingness;
+foreach my $locus (0..$numLoci-1) {
+    if ($locus % 10000 == 0) {
+        print "Processed $locus loci from $genotypesFile\n";
+    }
+    foreach my $sampleIndex1 (0..scalar(@samples)-1) {
+        foreach my $sampleIndex2 ($sampleIndex1..scalar(@samples)-1) {
[…]

### 2021-10-26 RAD_code/analysis_tools/clustOpt/vcfMissingness.pl

@@ -1,0 +1,68 @@
+#!/usr/bin/env perl
+
+use strict;
+use warnings;
+use Getopt::Long;
+use File::Basename;
+
+my $vcflist;
+my $help;
+my $clustOptDir = dirname($0);
+
+GetOptions  ("vcflist=s"   => \$vcflist,
+             "help|man"    => \$help) || die "Couldn't get options with GetOpt::Long: $!\n";
+
+if (!$vcflist or $help) {
+    die "Must supply --vcflist.
+    --vcflist is a file that contains one VCF filename per line, each one corresponding to a different clustering threshold\n";
+}
+
+open(my $vcfFilesFH, "<", $vcflist);
+# Get the VCF filenames from the input file
+my @vcfFiles;
+while (my $line = <$vcfFilesFH>) {
+    chomp($line);
+    push(@vcfFiles, $line);
+}
+
+
+################################################
+### Translate the VCF files into 012 format: ###
+################################################
+print "#### Translating VCF files into 012 format ####\n";
+
+foreach my $vcf (@vcfFiles) {
+    my $outPrefix = $1 . "_missHM012" if ($vcf =~ /(.*)\.vcf/);
+    if (-e $outPrefix . '.012') {
+        die "$outPrefix.012 already exists. Exiting now to prevent overwriting files\n";
+    } elsif (-e $outPrefix . '.012.indv') {
+        die "$outPrefix.012.indv already exists. Exiting now to prevent overwriting files\n";
+    } elsif (-e $outPrefix . '.012.pos') {
+        die "$outPrefix.012.pos already exists. Exiting now to prevent overwriting files\n";
+    }
+    system("vcftools --vcf $vcf --012 --out $outPrefix");
+}
+print "\n\n\n";
+print "#### Finished translating VCF files into 012 format ####\n";
+
+########################################################################
+### Now use the 012 files to generate pairwise missingness matrices: ###
+########################################################################
+
+print "\n\n\n";
+print "#### Generating pairwise missingness matrices from 012 files ####\n";
+foreach my $vcf (@vcfFiles) {
+    my $inPrefix = $1 . "_missHM012" if ($vcf =~ /(.*)\.vcf/);
+    my $genotypes = $inPrefix . ".012";
+    my $individuals = $inPrefix . ".012.indv";
+    my $out = $inPrefix . ".missingness";
+    system("perl $clustOptDir/resources/pairwiseMissingnessFrom012.pl --genotypes $genotypes --individuals $individuals --out $out");
+}
+print "\n\n\n";
[…]

### 2021-10-26 RAD_code/analysis_tools/clustOpt/vcfToPCAvarExplained.R

@@ -1,0 +1,24 @@
+#!/usr/bin/env Rscript
+
+# Load SNPRelate
+suppressPackageStartupMessages(library(SNPRelate))
+
+### Make sure that there is exactly one flag--the VCF file input list
+args = commandArgs(trailingOnly=TRUE)
+if (length(args)!=2) {
+  stop("Exactly two arguments must be supplied (<vcf file list> and <numPCsToSum>).n", call.=FALSE)
+}
+
+### Put the VCF filenames into a vector
+vcfFiles <- scan(args[1], what = character(), quiet=T)
+
+for (i in 1:length(vcfFiles)) {
+    gdsOut <- paste(gsub(".vcf","", vcfFiles[i], fixed=T), '.pcaVar.gds', sep="")
+    snpgdsVCF2GDS(vcfFiles[i], gdsOut, verbose=F)
+    genofile <- snpgdsOpen(gdsOut)
+    pcaz <- snpgdsPCA(genofile, num.thread=2, verbose=F, autosome.only=FALSE)
+    varExplained <- sum(pcaz$varprop[1:args[2]])
+    cat("Variance explained by first ", args[2], " PCs for ", vcfFiles[i], ": ", varExplained, "\n", sep="")
+    snpgdsClose(genofile)
+    file.remove(gdsOut)
+}

### 2021-11-04 RADseq.Rmd
Changed format of site to Bootswatch

--- 2021-10-26 RAD_code/RADpage_text.Rmd
@@
+```{r,message = FALSE,echo = FALSE}
+  library("knitcitations")
+  library("knitr")
+  cite_options(citation_format = "pandoc", max.names = 3, style = "html", hyperlink = "to.doc")
+  bib <- read.bibtex("bibtexlib.bib")
+  opts_chunk$set(tidy = FALSE, message = FALSE, warning = FALSE,
+                 fig.width = 10, fig.height = 6, cache = TRUE)
+  # use this to set knitr options:
+  # http://yihui.name/knitr/options #chunk_options
+```
 ---
-layout: default
-title: RADseq
-parent: Choosing a PopGenomics Approach
-grand_parent: Best Principles
-#editor_options: 
-#  chunk_output_type: console
+title: 'RADseq'
+subtitle: '*NJ Gr&uuml;nwald, ZN Kamvar, and SE Everhart*'
 ---
@@
 
+colorize <- function(x, color) {
+  if (knitr::is_latex_output()) {
+    sprintf("\\textcolor{%s}{%s}", color, x)
+  } else if (knitr::is_html_output()) {
+    sprintf("<span style='color: %s;'>%s</span>", color,
+      x)
+  } else x
+}
 #set bash code chunks to use bash_profile
@@
 
+```
+
+If you would like to run the R code examples that are scattered throughout the guide (recommended but not required!), you will need to install some R packages:  
+```{r, eval=F}
+install.packages("tidyverse")
 if (!requireNamespace("BiocManager", quietly = TRUE))
@@
 
-#BiocManager::install("SeqArray")
-#BiocManager::install("SNPRelate")
-#install.packages("rlang")
+BiocManager::install("SeqArray")
+BiocManager::install("SNPRelate")
 ```
 
-# RADseq  
-{: .no_toc }
-
-<details open markdown="block">
-  <summary>
-    Table of contents
-  </summary>
-  {: .text-delta }
-1. TOC
-{:toc}
-</details>
-
-If you want to run the acompanying code chunks, install Docker container.
+Now load those packages, if using:  
 ```{r}
@@
 library(tidyverse)
+library(SNPRelate)
 ```
 
-<!-- `r colorize("Dani to do: make a dockerfile with all the packages used in this Rscript for reproducibility", "red")` -->
+We also provide alternative coding examples that are based only on the command line, requiring these software packages:
 
-<!-- `r colorize("Example files needed: example strata file 'data/example.metadata.csv', example vcf for PCA './data/test.vcf' ", "red")` -->
+* vcftools(link)
+* gnuplot
 
 # Reduced Representation Sequencing (RADseq/GBS)
-<!-- `r colorize("What kinds of protocols is this advice relevant to?", "red")`   -->
+
 "Restriction-site   Associated   DNA   sequencing"  - RADseq - combines  restriction  enzymes  with next-gen, massively parallel,  short-read  sequencing.
-RADseq involves the use of restriction enzymes which are used to shear DNA at selected sites.
+RADseq involves the use of restriction enzymes which are used to shear DNA at restriction enzyme cutsite.
 RADseq comes in different flavors.
@@
 
-<!-- `r colorize("characteristics of radseq that are different/ important to know compared to other types of sequencing ", "red")`   -->
+Some aspects of RADseq data that are unique compared to whole genome resequencing (more):
 
@@
 
-![Table 1 from O'Leary et al. 2018.](../RAD_code/images/oleary_table1.png)
+![Table 1 from O'Leary et al. 2018.](images/oleary_table1.png)
 
@@
 This is to allow you to control for potential [batch effects](#batch) that are often observed with RAD data.
-```{r}
+```{r, eval = F, include=FALSE}
 # This code reads in a comma-delimited STRATA/meta-data file, randomises samples, and returns a file with wells and plates and duplicates which may be used to assist with plating libraries
@@
 If that doesn't occur or if too many reads are being filtered so as to only recover a small number of SNPs, something might need to be tweaked with your trimming and filtering step.
-[See how to evaluate for sequencing error](#seq)
 
@@
 
-### Evaluate potential confounding factors {#confound}
-
-#### The power of PCA
-
-One of the most powerful methods for exploring your data for issues is a Principle Components Analysis (see [here](https://michaelaalcorn.medium.com/yet-another-pca-explanation-727dff6ce26a) for a linear algebr-based explanation of PCA).
-From [this genetics tutorial](https://comppopgenworkshop2019.readthedocs.io/en/latest/contents/03_pca/pca.html): "To understand how PCA works, consider a single individual and its representation by its 593,124 markers.
-Formally, each individual is a point in a 593,124-dimensional space, where each dimension can take only the three possible genotypes indicated above, or have missing data.
-To visualise this high-dimensional dataset, we would like to project it down to two dimensions.
-But as there are many ways to project the shadow of a three-dimensional object on a two dimensional plane, there are many (and even more) ways to project a 593,124-dimensional cloud of points to two dimensions.
-What PCA does is figuring out the “best” way to do this project in order to visualise the major components of variance in the data.
-
-As soon as you have a file with SNPs for all samples, before extensive filtering or playing with parameters, do a PCA on it and look at all the axes with the most variation.
-Be careful that you understand how missing data is imputed for PCA.
-
-For the code examples in this section, we mostly make use of the R package SeqArray, which can read in and manipulate VCFs (*notes on Docker install*).
-We like this program because ....
-For those you are less familiar with R or more familar with Plink, check out [this tutorial]() which uses Plink and minimal R plotting.
-a tutorial that uses We also include some examples using VCFtools and minimal R plotting, for those who are less familiar with R.
-
-(?) dani asks: you want me to make this take multiple different vcfs from pipeline outputs for plotting? should we implement pcANGSD which can handle missing data better  (?)  
-K: Making it take multiple vcfs is a secondary goal, and can be added later if we have time/interest. Adding a code example for pcANGSD would be useful though.
-
-## ill get onto this sorry
-
-Doing a PCA with SeqArray and SNPRelate in R. Filtering for SNPs with a MSF > 2.5%
-```{r, eval =T}
-## read in VCF and convert to GDS format
-#vcf.file = "./data/test.vcf"
-#seqVCF2GDS(vcf.file, "test.gds.rad", storage.option="ZIP_RA")
-
-# read in existing gds file  
-filename = "data/test.gds.rad"
-gdsin= SeqArray::seqOpen(filename,)
-# summarize the data
-print(paste0("The number of SAMPLES in data: ", length(c(SeqArray::seqGetData(gdsin, "sample.id")))))
-print(paste0("The number of SNPs in data: ",  length(c(SeqArray::seqGetData(gdsin, "variant.id")))))
-#read in metadata
-sample.ids = seqGetData(gdsin, "sample.id")
-sample.strata =  read.table("data/test.metadata.csv", header = T, sep = "\t") %>%
-                  dplyr::select(TARGET_ID, STRATA, LONG, LAT, PLATE)
-id.order = sapply(sample.ids, function(x,df){which(df$TARGET_ID == x)}, df=sample.strata) #incase your strata file is not in the same order as your vcf
-sample.strata.order = sample.strata[id.order,]
-
-# get pruned SNPs
-# NOT WORKING YET, issue with chromosome names
-# Try this with vcf https://github.com/zhengxwen/SNPRelate/issues/10  
-#snpset <- SNPRelate::snpgdsLDpruning(gdsin, ld.threshold=0.2, autosome.only = F, start.pos="random", num.thread=1)  
-#snpset.id <- unlist(unname(snpset))
-
-# PCA only on SNPs with a minor allele freq greater than 2.5%
-pca.out = SNPRelate::snpgdsPCA(autosome.only = F, gdsin, num.thread=2, remove.monosnp = T, maf = 0.025)
-#                               snp.id=snpset.id) # filtering for pruned SNPs
-eig = pca.out$eigenval[!is.na(pca.out$eigenval)]
-barplot(100*eig/sum(eig), main="PCA Eigenvalues")
-```
-
-* If you used sample replicates, first see if replicates cluster very close together.
-If not, then there may be batch effects or bioinformatic artifacts leading to genotyping error that must be addressed.
-Then remove replicates for a subsequent PCA evaluation.
-
-```{r}
-#look at PC values for reps
-#remove reps
-```
-
-First color/shape the individuals by the factor you expect to matter (eg, sampling site or region, family, ecotype).  
-```{r, eval=T}
-#PLOT PCA
-#PC1 v PC2 colored by collection location
-print(
-  as.data.frame(pca.out$eigenvect) %>%
-      #tibble::add_column(., ID =  sample.ids) %>%
-      tibble::add_column(., STRATA =  sample.strata.order$STRATA) %>%
-      ggplot(., aes(x=V1, y=V2, color = STRATA))  + #label = ID
-      geom_point(size=2) +
-      #geom_text() +
-      stat_ellipse(level = 0.95, size = 1) +
-      #scale_color_manual(values = gds_colors) +
-      geom_hline(yintercept = 0) +
-      geom_vline(xintercept = 0) +
-      theme_bw() +
-      xlab(paste0("PC1 [",paste0(round(eig[1], 2)), "%]")) +
-      ylab(paste0("PC2 [",paste0(round(eig[2], 2)), "%]")) +
-      ggtitle("PCA Colored by Collection Location")
-)
-```
-
-* Is there any clustering? If so, is it different than your expectation?
-  + If different than expectation, suggests either [batch effects](#batch), [cryptic variation](#cryptic) among individuals, or systematic biological issues affecting assembly/mapping (eg, inversions (Katies inversion, also paper ref))
-  <!-- + `r colorize("figure example, weird clustering", "red")` -->
-* Are there outlier samples driving a large amount of variation?
-  + If so, they may be [cryptic species](#cryptic), clones/sample replicates, or they [sequenced/genotyped poorly](#bad)
-  + In our example we see 2 potential outlier samples in the bottom left corner. We can check which samples they are by plotting the PCA with sample ID and/or looking at the actual PC scores.
-```{r, eval=T}
-#PLOT PCA
-#PC1 v PC2 colored by collection location
-print(
-  as.data.frame(pca.out$eigenvect) %>%
-      tibble::add_column(., ID =  sample.ids) %>%
-      ggplot(., aes(x=V1, y=V2, label = ID)) +
-      geom_text(size =1.5) +
-      geom_hline(yintercept = 0) +
-      geom_vline(xintercept = 0) +
-      theme_bw() +
-      xlab(paste0("PC1 [",paste0(round(eig[1], 2)), "%]")) +
-      ylab(paste0("PC2 [",paste0(round(eig[2], 2)), "%]")) +
-      ggtitle("PCA with Labels")
-)
-```
-```{r}
-# which samples have PC1 < -0.1 and PC2 < -0.6?
-as.data.frame(pca.out$eigenvect) %>%
-      tibble::add_column(., ID =  sample.ids) %>%
-  filter(V1 < -0.1 & V2 < -0.6) %>%
-  select(ID,V1,V2)
-```
-
-* Is there no clustering at all? If so, it may be due to:  
-  + bioinformatic artifacts leading to noise, [see Testing Parameters](#test)
-  + issues with how you are [filtering the data](#filter) (eg, too strict or too lax)
-  + issues with how your PCA is treating missing data
-  + or you actually have no structure in your dataset!
-  <!-- + `r colorize("figure example, no clustering", "red")` -->
-
-Next, color the individuals by potential sources of batch effects (sequencing lane, library prep batch, age of tissue sample, person doing the DNA extraction...).    
-```{r, eval=T}
-# again, but colored by batch
-print(
-  as.data.frame(pca.out$eigenvect) %>%
-      tibble::add_column(., PLATE =  as.factor(sample.strata.order$PLATE)) %>%
-      ggplot(., aes(x=V1, y=V2, color = PLATE))  + #label = ID
-      geom_point(size=2) +
-      stat_ellipse(level = 0.95, size = 1) +
-      geom_hline(yintercept = 0) +
-      geom_vline(xintercept = 0) +
-      theme_bw() +
-      xlab(paste0("PC1 [",paste0(round(eig[1], 2)), "%]")) +
-      ylab(paste0("PC2 [",paste0(round(eig[2], 2)), "%]")) +
-      ggtitle("PCA colored by Batch/Sequencing Plate")
-)
-```
-
-* Do any observed clusters match a source of batch effect? [how to remove batch effects](#batch)
-
-#### Visually inspect alignments {#align}  
-Another useful approach is to visually inspect alignments for SNPs or loci that are identified as "weird" in the subsequent analyses.
-In ipyrad this can be done by searching the .loci file, in dDocent...
-If the alignment looks wonky, you can feel better about removing the whole locus (?).
-<!-- `r colorize("code example .loci", "red")` -->
-<!-- `r colorize("code dDocent", "red")`   -->
-<!-- `r colorize("code stacks", "red")`     -->
-
-#### Batch effects {#batch}
-As discussed in [Setting up for success](#labwork), batch effects can arise by minor (or not so minor) differences during the library prep and sequencing stage.
-Randomizing samples across libraries and sequencing lanes can help mitigate the influence of batch effects on your downstream analyses, but sometimes this isn't possible, esp.
-when using pre-existing data.
-Even if you randomize your samples, it is a good idea to check for batch effects with a PCA and then try to mitigate them.
-
-<!-- `r colorize("PCA figure of batch effect", "red")` -->
-
-You can identify which loci are driving the batch effect by 1) identifying the top PC loadings (SNPs) and/or 2) do an Fst outlier analysis with the batches specified as populations (eg, Bayescan).
-Both of these methods are most appropriate if you've randomized your samples between batches.
-If you have not randomized samples but are certain there is a batch effect driving variation in your dataset, you can still remove loci this way but you may also be removing biologically informative loci.
-
-Nicolas Lou and Nina Therkildsen have a [great preprint](https://www.authorea.com/users/380682/articles/532568-batch-effects-in-population-genomic-studies-with-low-coverage-whole-genome-sequencing-data-causes-detection-and-mitigation) and [Github repo](https://github.com/therkildsen-lab/batch-effect) discussing causes, detection, and mitigation of batch effects in low coverage WGS, where batch effects can have particularly large effects.
-While some of their recommendations are specific to reference-based analyses, others are relevant to RAD methods too (eg, batch effects due to different sequencing platforms, DNA degradation levels, and sequencing depth).
-
-Once you've identified these batch effect SNPs, it can be useful to [visualize the alignments](#align) before removing them.
-If the entire alignment for that locus looks wonky, you should remove all SNPs from that RAD locus from your final dataset before conducting analyses.
-<!-- `r colorize("code to identify top loading snps", "red")` -->
-
-```{r top loading SNPS}
-```
-
-#### Cryptic species/contamination/clones {#cryptic}  
-One issue that is particularly common, especially for marine invertebrates, are cryptic, yet highly diverged lineages occurring in the same geographic location.
-If you are trying to do a landscape/seascape genetics study, cryptic species can dramatically throw off population genetic analyses.
-A PCA using all the samples can help identify cryptic species if samples are forming clusters that do not align with any other possible factor (eg, geography, sequencing batch).
-<!-- `r colorize("PCA figure of cryptic species", "red")`   -->
-
-```{r}
-```
-
-Note: Notice the huge variation explained along the first PC-axis.
-
-You can also check heterozygosity, and missingness!
-
-```{r}
-```
-
- Note: Two outliers are obvious.
-Large differences in heterozygosity between samples are clear, with one location showing two distinct clusters of samples.
-These samples were later determined using mtDNA and morphological analysis to be cryptic species.
-*Unpublished data from a marine species in Australian waters.*
-
-You can then estimate genetic divergence between these clusters using pairwise Fst and/or phylogenetic inference to determine how diverged the groups are.
-If they are very diverged, it may be better to assemble each cryptic species separately, especially for denovo assemblies with dDocent and Stacks, especially since often times restriction cut-sites are not always conserved; see [Wagner et al., 2013](link), [Rubin et al., 2012](link)  These two pipelines used a set of given samples to create a "reference" or "catalog" to which all other samples are mapped.
-If highly diverged samples are included in creating this reference, it can lead to [oversplitting](#cluster).
-
-Cryptic clones can also dramatically throw off population genetic analyses.
-In a PCA, clones will often appear as their own very tight cluster, similar to sample replicates.
-Below the PCA shows x and a simple calculate of pairwise-distance (i.e.
-here manhattan distance) shows x known duplicates, and x unknown clones.
-*other ways to tell they are clones*
-<!-- `r colorize("PCA figure of cryptic clones", "red")` -->
-
-#### "Bad" samples {#bad}  
-Sometimes a sample doesn't sequence well (few sequencing reads, higher error rate).
-This can be due to DNA quality, an issue during library prep, or not enough sequencing depth (average # of reads per sample).
-Generally, it will lead to a sample with fewer sequencing reads, higher missing data in a SNP dataset, and fewer shared loci with other samples.
-Identifying and then removing these samples *prior* to the final RADseq assembly analysis can help minimize mis-assembled loci, genotyping errors, and excessive filtering of acceptable loci.
-
-The process of identifying low quality individuals is usually iterative, as the way you initially filter your SNPs will influence the amount of missing data and locus sharing among samples.
-This is why we recommend minimally filtering your SNPs for sample coverage (the # of individuals a locus is called in) when initially exploring your data.
-Some ways to identify bad samples:
-
-* For every SNP dataset you generate, it is a good idea is always evaluate the missingness per sample (and report this distribution in your manuscript!).
-Identify samples with way more missingness than the rest, and observe how they look in a PCA and locus sharing plot.
-If they stick out, then try removing them from the assembly and seeing if it changes downstream analyses.
-If so, you may want to specify a missingness cutoff for including samples in the final analysis.
-**vcftools example, on the command line**  
-```{bash, eval = F}
-vcftools --vcf data/stickleback_ferchaud_2015.vcf --missing-indv --out analysis/stickleback
-# sort the file by most missing data and print the top 10  
-cat analysis/stickleback.imiss | (read h; echo "$h"; sort -k5 -r) | head -n 10
-```
-##### Plot missingness {.tabset .tabset-fade .tabset-pills}
-###### With R
-```{r}
-miss <- read.table("analysis/stickleback.imiss",header = T,sep="\t")
-hist(miss$F_MISS,breaks=50)
-```
-
-###### With bash (requires gnuplot installed)
-```{bash}
-#code from Jon Puritz
-mawk '!/IN/' analysis/stickleback.imiss | cut -f5 > totalmissing
-gnuplot << \EOF
-set terminal dumb size 120, 30
-set autoscale
-unset label
-set title "Histogram of % missing data per individual"
-set ylabel "Number of Occurrences"
-set xlabel "% of missing data"
-#set yr [0:100000]
-binwidth=0.01
-bin(x,width)=width*floor(x/width) + binwidth/2.0
-plot 'totalmissing' using (bin($1,binwidth)):(1.0) smooth freq with boxes
-pause -1
-EOF
-```
-##### {-}
-
-* Construct a locus sharing heatmap between pairs of samples, which shows the proportion of loci that are genotyped in both samples.
-Generally, samples that are more closely related will share more loci with each other due to cutsite dropout.
-Deviations from that pattern can pinpoint bad samples as well as batch effects.
-  + [ipyrad API tutorial](https://ipyrad.readthedocs.io/en/latest/API-analysis/cookbook-sharing.html) for making a locus sharing heatmap, requires ipyrad output or ipyrad's vcf2hdf5 conversion  
-  + [code to generate locus sharing/missingness heatmap from a VCF](https://github.com/atcg/clustOpt)
-
-### Test a range of key parameters {#test}
-
-#### Clustering threshold (de novo assembly) {#cluster}  
-During de novo RADseq assembly, reads are assembled into contigs so that each contig represents a single locus.
-This can occur separately for each individual (ipyrad) or jointly for all or a subset of individuals (dDocent, Stacks).
-To call SNPs, in the case of ipyrad, contigs are then clustered and aligned across individuals, while in dDocent and Stacks, reads for each individual are mapped to the "reference" contigs.
-All of these pipelines use sequence similarity thresholds to determine if fragments are homologous (from the same place in the genome) and therefore should be clustered together.
-If the sequence similarity threshold is too low, *undersplitting* can occur so that multiple loci are combined into a single contig.
-<!-- `r colorize("figure example, undersplit", "red")`   -->
-Undersplitting is common with repetitive regions and paralogs. It can inflate mean observed heterozygosity.
-
-If the threshold is too high, it can lead to *oversplitting*, where alleles of the same locus are split into two or more contigs. This can lead to reduced mean observed heterozygosity.  
-<!-- `r colorize("figure example, oversplit", "red")`   -->
-If severely off, both of these issues can bias the models used to identify SNPs.
-Undersplitting a little is better than oversplitting, as it is easier to identify lumped paralogs downstream and remove them.
-**What influences the "optimal" clustering threshold of your data?**  
-Clustering threshold is usually determined by how much genetic diversity you have in your dataset.
-If your dataset has high genetic diversity (eg., a single species with high heterozygosity, multiple highly diverged populations, multiple species), then you will need a lower clustering threshold.
-If your dataset has low genetic diversity (a single species with lower heterozygosity, a single family), then you will need a higher clustering threshold.
-Polyploid or highly repetitive (ie large) genomes may also have different requirements than diplod species with smaller genomes.
-Other factors, such as sequencing error rate, may play a role as well so it is usually a good idea to always test a few clustering thresholds with a new dataset.
-However, as long as your clustering threshold is in the ballpark it shouldn't affect downstream analyses (according to Isaac).
-
-There are a few strategies to identify a good clustering threshold for your data, all of which require assembling your data with a range of clustering thresholds.  
-1) Metrics developed by [McCartney-Melstad 2017](link), based on popgen theory and landscape genetic expectations  
-2) Utilize sample replicates (either exact replicates or individuals from the same locality) to quantify genotyping error and minimize error while maximizing the number of informative loci [Mastretta-Yanes 2015](link)
-3) Evaluating the number of haplotypes across loci, with many 3+ haplotypes indicating undersplitting [Harvey 2015](link), [Ilut 2014](link)  
-4) Make assemblies at a range of thresholds and parameters, plot the # of SNPs or contigs per assembly, and make a judgement call [example in dDocent tutorial](link).
-If most parameters give you about the same # of SNPs or contigs, you can feel confident that this parameter doesn't matter much for your data and just choose a reasonable value.
-
-#### Mapping parameters (reference-based and some de novo assemblers)   
-When a sequencing read maps to the wrong assembled contig or place in a reference genome (ie, a non-homologous locus), this can result in erroneous SNPs.
-In many cases, these SNPs can be filtered out later (link to filtering).
-However, data should be checked to make sure there aren't systematic, wide-scale mapping issues that need to be addressed.
-
-For referenced-based assembly, both dDocent and ipyrad make use of bwa(link) to map reads, while Stacks allows the user to choose their aligner. dDocent also utilizes bwa for de novo assemblies.   *
-
-* how many reads aligned for each sample? If many reads didn't aligned, this could be an issue with the reference, read trimming/filtering, or the mapping parameters.
-* What is the distribution of mapping scores?  
-* ratio of mean (across inds) mapping quality scores for reference and alternate allele (O'Leary cite). 0.25-1.75 cut off implemented in dDocent.  
-  + this info might not be coded in all VCFs
-
-(link to WGS mapping quality section)
-
-Mapping defaults:
-dDocent: bwa mem -L 20,5 -t 8 -a -M -T 10 -A1 -B 3 -O 5 -R
-L: clipping penalty for 5' and 3'
-A: matching score
-O gap open penalty
-T: mapping quality filter
-ipyrad: (defaults) -L 5,5 -B 4 -A1 -O 6 -T 30
-
-#### Read depth threshold {#seq}
-
-Mistakes that occur during sequencing can lead to a base-calling error in a read, which can then lead to an erroneous SNP.
-This type of error most commonly manifests as rare alleles (eg, only observed in one individual) or singletons (only one copy of the minor allele across all individuals).
-A conservative approach to deal with these errors is to filter out all loci with rare alleles (link filtering), but for some demographic analyses this filtering is inappropriate as they require the site frequency spectrum.
-The best ways to minimize the impact of sequencing error during bioinformatic processing is to 1) properly trim and filter your data, and 2) require higher read depth thresholds for assembly and genotyping.
-
-2) Require higher read depth thresholds for assembly and genotyping.  
-For RADseq, ideally you are generating an average of *at least* 10-20 sequence reads (10x-20x coverage) for each RAD locus in your library preparation (but this doesn't always happen...).
-For diploid species, this would be sequencing each RAD haplotype at least 5x-10x.
-The SNP calling step of RAD pipelines all require setting a minimum number of reads to determine whether a site is homozygous or heterozygous **add image example**.
-If this threshold is too low, sequencing errors are more likely to be included and the SNP caller will have a harder time accurately identifying SNPs ([but see our section on low coverage WGS methods!](link).
-If it's too high, then real polymorphic loci will be excluded.
-All of the three major pipelines allow users to implement read depth thresholds
-
-Generating multiple sequencing reads per locus is helpful when calling SNPs in order to  distinguishing between reads with sequencing errors and
-
-<!-- `r colorize("Fountain et al 2016 found error rates > 0.06 with minimum 20 read depth and Stacks v1.
-Can this be due to not optimizing other Stacks parameters?
-Hard to interpret/ know how relevant it is now, even to Stacks2", "red")` -->
-
-#### Filtering reads
-
-FASTQ files that typically come from the sequencing center include a [Phred scale quality score](link) for each base, which indicates the probability of that base call being correct.
-This information can be used prior to assembly to trim low-quality sections from either end of reads or remove reads altogether.
-Some SNP callers, including FreeBayes which is part of dDocent, also use Phred scores when calling SNPs (ipyrad and Stacks do not).
-All three of the most popular RAD pipelines include steps to filter reads due to low quality bases low quality bases or adapter sequences, but vary slightly on what kinds of read trimming are allowed.
-Always make sure to filter out adapter sequences (this is not default with stacks), otherwise default parameters are usually adequate.
-If your initial [fastqc](#fastqc) indicates that you have many low quality bases in your reads, you may lose too much of your data during filtering.
-In this case, you may want to make your read filtering a little more permissible (keeping in mind that this may increase SNP error rate).
-**Always make sure to check with your sequencing center to see if they have done anything to the raw reads before you received them!**
-<!---turn these into html tabs--->
-
-**dDocent: **
-
-* does not allow you to use reads that have already been trimmed and filtered for de novo assembly, only referenced-based.
-* Uses Trimmomatic to trim Illumina TruSeq adaters and low quality bases. Make sure to change the defaults if you used a different adapter/sequencing system.
-
-**Stacks: **
-
-* requires your reads to be all the same length, so if they were previously trimmed of low quality bases or adapter sequence this must be specified and they will be truncated to the same length.
-
-### Filter your data many ways (and embrace the missingness!)
-* Don't over filter for missingness
- + papers showing how too much filtering can skew results  
-  + Empirical: Diaz-Arce 2019, Tripp 2017, Silliman 2021, Eaton 2017, Martin-Hernanz (2019)
-  + Simulations: Huang & Knowles 2016, Leaché et al., 2015, Gautier et al., 2012, Rivera-Colón et al., 2021
-
-* some ways to account for missing data in downstream analyses
- + impute (PCA), subsample (PCA, structure, pretty much anything with SNPs), try a few different missing filters and see what impact it has on results
- + use [pcANGSD](link)
-
-## Assembly and Genotyping Pipelines  {#assembly}
+### Run an assembly pipeline {#assembly}
 
@@
 Based on the MarineOmics RAD panel, the general consensus is that all the popular pipelines can produce adequately accurate datasets (with appropriate parameter optimization and data filtering).
-![Table 3 from LaCava et al 2019. CD-HIT is used in dDocent, VSEARCH is used in ipyrad.](../RAD_code/images/lacavaTable3.png)
+![Table 3 from LaCava et al 2019. CD-HIT is used in dDocent, VSEARCH is used in ipyrad.](images/lacavaTable3.png)
 
-### iPyrad walkthrough
+#### iPyrad walkthrough
 
 Link to annotated ipyrad notebook
+
+### Evaluate potential confounding factors {#confound}
+
+Once you have processed your samples (or a subset of samples) through a genotyping pipeline, you will have a bunch of different output file options for genetic data.
+The VCF file is one of the most popular file formats, and is the most versatile for initial data exploration as many programs exist to filter and accept VCF files.
+Below are a list of potential confounding factors that may exist in your data, and how to tease them out.
+
+**Using SeqArray in R** 
+If you are following along with the R code examples, we need to 1st read in our data (skip this section if not running R code).
+
+For the code examples in this section, we primarily make use of the R package SeqArray [Zheng et al., 2017](https://doi.org/10.1093/bioinformatics/btx145), which can read and manipulate VCFs.
+If you’re familiar with R, SeqArray is simple to use.
+The package can efficiently store sequence variant calls with hundreds of thousands of individuals,  variants (e.g., SNVs, indels and structural variation calls) and annotations.
+It employs a data container called a CoreArray Genomic Data Structure (GDS).
+It’s super-fast (5X faster than PLINK v1.9; 16X faster than vcftools) and it integrates well with other R packages you might use in your analysis pipeline.
+(i.e.
+SNPRelate, SeqVarTools).
+We also like it because you can filter your data before running certain analyses without generating a separate filtered VCF file.
+
+First, we need to convert our VCF file into the GDS format. We will do this once here, then use the GDS file for subsequent code examples.
+```{r}
+filename = "OL_subset" #replace with your file name 
+filename.gds = paste0("RAD_data/", paste0(filename, ".gds"))
+filename.vcf = paste0("RAD_data/", paste0(filename, ".vcf"))
+  # 1 . Convert VCF to GDS
+SeqArray::seqVCF2GDS(vcf.fn = filename.vcf, out.fn = filename.gds, storage.option="ZIP_RA")
+gdsin = SeqArray::seqOpen(filename.gds)
+print(paste0("The number of SAMPLES in data: ", length(c(SeqArray::seqGetData(gdsin, "sample.id")))))
+print(paste0("The number of SNPs in data: ",  length(c(SeqArray::seqGetData(gdsin, "variant.id")))))
+```
+It is always helpful to have a metadata file with information for each sample, such as sampling site, sequencing library, etc.
+In our example, our matadata file (OL.popmap) is tab-delimited and has the column headers:
+ID: sample ID
+STRATA: sampling location/population
+PLATE: sequencing batch  
+Next we read in our metadata file, and make sure the samples are in the same order as your VCF file:  
+```{r}
+metafile = "RAD_data/OL.popmap"
+sample.ids = seqGetData(gdsin, "sample.id")
+sample.strata =  read.table(metafile, header = T, sep = "\t") %>%
+                  dplyr::select(ID, STRATA, PLATE)
+```
+Now, on to evaluating our data!
+
+#### "Bad" samples {#bad}
+
+Sometimes a sample doesn't sequence well (few sequencing reads, higher error rate).
+This can be due to DNA quality, an issue during library prep, or not enough sequencing depth (average # of reads per sample).
+Generally, it will lead to a sample with fewer sequencing reads, higher missing data in a SNP dataset, and fewer shared loci with other samples.
+Identifying and then removing these samples *prior* to the final RADseq assembly analysis can help minimize mis-assembled loci, genotyping errors, and excessive filtering of acceptable loci.
+
+The process of identifying low quality individuals is usually iterative, as the way you initially filter your SNPs will influence the amount of missing data and locus sharing among samples.
+This is why we recommend minimally filtering your SNPs for sample coverage (the # of individuals a locus is called in) when initially exploring your data.
+Some ways to identify bad samples:
+
+1) For every SNP dataset you generate, it is a good idea is always evaluate the missingness per sample (and report this distribution in your manuscript!).
+Identify samples with way more missingness than the rest, and observe how they look in a PCA and locus sharing plot.
+If they stick out or all cluster together in the middle, then try removing them from the assembly and seeing if it changes downstream analyses.
+If so, you may want to specify a missingness cutoff for including samples in the final analysis.
+
+**Missingness in R with SeqArray**
+```{r}
+#using previously loaded gdsin object
+print("Per variant: ")
+summary(m1 <- SeqArray::seqMissing(gdsin, per.variant=TRUE))
+print("Per sample: ")
+summary(m2 <- SeqArray::seqMissing(gdsin, per.variant=FALSE))
+samples <- SeqArray::seqGetData(gdsin, "sample.id")
+cbind(samples,m2)[order(-m2),]
+#plot histogram 
+hist(m2,breaks=50)
+
+# close the GDS file
+seqClose(gdsin)
+```
+
+Another common method of filtering and evaluating you data without ever using R is VCFtools(link). 
+**Missingness with vcftools, on the command line**  
+```{bash, eval = T}
+vcftools --vcf RAD_data/OL_subset.vcf --missing-indv --out RAD_data/OL_subset
+# sort the file by most missing data and print the top 10 samples  
+cat RAD_data/OL_subset.imiss | (read h; echo "$h"; sort -k5 -r) 
+```
+You can open the file analysis/OL_subset.imiss in any text editor an look at the missingness values by eye.
+
+To plot the missingness on the command line, you can use gnuplot:
+```{bash}
+#code from Jon Puritz
+mawk '!/IN/' RAD_data/OL_subset.imiss | cut -f5 > totalmissing
+gnuplot << \EOF
+set terminal dumb size 120, 30
+set autoscale
+unset label
+set title "Histogram of % missing data per individual"
+set ylabel "Number of Occurrences"
+set xlabel "% of missing data"
+#set yr [0:100000]
+binwidth=0.01
+bin(x,width)=width*floor(x/width) + binwidth/2.0
+plot 'totalmissing' using (bin($1,binwidth)):(1.0) smooth freq with boxes
+pause -1
+EOF
+```
+
+It looks like samples X,Y,Z have a lot of missing data relative to other samples.
+We can remove them from our dataset for now, and just keep exploring our data, but if you decide to exclude them permanently from the analysis you should eventually rerun your genotyping pipeline without those samples, esp.
+if doing a de novo assembly.
+
+2) Another great way to explore missingness with your data is to construct a heatmap of loci that are genoptyped between pairs of samples.
+Generally, samples that are more closely related will share more loci with each other due to cutsite dropout (link).
+Deviations from that pattern can pinpoint bad samples as well as batch effects.
+  + [ipyrad API tutorial](https://ipyrad.readthedocs.io/en/latest/API-analysis/cookbook-sharing.html) for making a locus sharing heatmap, requires ipyrad output or ipyrad's vcf2hdf5 conversion  
+  + [code to generate locus sharing/missingness heatmap from a VCF](https://github.com/atcg/clustOpt)
+
+#### The power of PCA
+
+One of the most powerful methods for exploring your data is a Principle Components Analysis.
+From [this genetics tutorial](https://comppopgenworkshop2019.readthedocs.io/en/latest/contents/03_pca/pca.html): "To understand how PCA works, consider a single individual and its representation by its 593,124 markers.
+Formally, each individual is a point in a 593,124-dimensional space, where each dimension can take only the three possible genotypes indicated above, or have missing data.
+To visualize this high-dimensional dataset, we would like to project it down to two dimensions.
+But as there are many ways to project the shadow of a three-dimensional object on a two dimensional plane, there are many (and even more) ways to project a 593,124-dimensional cloud of points to two dimensions.
+What PCA does is figuring out the “best” way to do this project in order to visualise the major components of variance in the data." See [here](https://michaelaalcorn.medium.com/yet-another-pca-explanation-727dff6ce26a) for a linear algebra-based explanation of PCA).
+
+When you have a VCF file with SNPs, use PCA before extensive filtering or playing with parameters to look at the data. Check which SNPs are associated with axes showing the most variation.
+
+Here we use SeqArray and SNPRelate to run a PCA in R. Other popular methods for PCA include pcangsd(link), ...., but note that ... requires some file conversions everytime you filter.
+
+*Reminder:* Missing data is a feature of RAD.
+Be aware of how different analysis tools deal with missingness, especially PCA which will fill in all missing data with some values.
+Here, the PCA from SNPRelate imputes missing genotypes as the dosage mean.
+
+##### Doing a PCA with SeqArray and SNPRelate in R.
+
+```{r}
+#open the gds file previously created
+filename.gds = "RAD_data/OL_subset.gds" #replace with your file name 
+gdsin = SeqArray::seqOpen(filename.gds)
+# exclude samples previously identified as having too much missin g data
+bad_samples = c("OR3_13_C3","CA4_12_C1","OR3_3_C7","CA4_13_C1","BC2_6_C2","BC2_9_C4")
+sample.ids = seqGetData(gdsin, "sample.id")
+keep = sample.ids[which(!sample.ids %in% bad_samples)]
+
+```
+Whether your data are denovo or reference-based, it is important to filter out linked sites before performing a PCA.
+Here's how to do it in SeqArray.
+Note, if you have already filtered your VCF file to have only 1 SNP per RAD locus, you don't need to do this:
+```{r,results='hide'}
+snpset <- SNPRelate::snpgdsLDpruning(gdsin, ld.threshold=0.2, autosome.only = F, start.pos="random", num.thread=1, remove.monosnp = T, sample.id = keep)  
+snpset.id <- unlist(unname(snpset))
+```
+
+Now we will actually run the PCA, again removing the samples with missing data, keeping unlinked SNPs, and removing SNPs with less than 5% minor allele frequency. 
+```{r}
+# PCA only on SNPs with a minor allele freq greater than 2.5%
+pca.out = SNPRelate::snpgdsPCA(autosome.only = F, gdsin, num.thread=2, remove.monosnp = T, maf = 0.05,
+                               snp.id=snpset.id,
+                               sample.id = keep) # filtering for pruned SNPs
+#close the gds file (saves memory)
+#seqClose(gdsin)
+
+eig = pca.out$eigenval[!is.na(pca.out$eigenval)]
+barplot(100*eig/sum(eig), main="PCA Eigenvalues")
+```
+
+Data Info: Below we use RAD data from an Australian marine species complex.
+```{r, eval =F}
+
+## The following code will read from a text file containing a list of the VCF files you wish to explore (one file per line) 
+## Each line must contain the full file path 
+## The code will convert your VCF to GDS
+## Perform PCA
+## Plot the results
+
+list_vcfs = read.table("analysis/Pop_Map/vcf.list")[[1]]
+lapply(list(list_vcfs), function(x) { 
+  filename = gsub(".vcf", "", gsub("data/", "", x)) ## NOTE:: you want a pattern that will remove the file path 
+  print(paste0("PCA for ... ", filename))
+  filename.gds = paste0("data/gds/", paste0(filename, ".gds"))
+  gdsfmt::showfile.gds(closeall=TRUE)
+  # 1 . Convert VCF to GDS
+  SeqArray::seqVCF2GDS(vcf.fn = x, out.fn = filename.gds, storage.option="ZIP_RA")
+  gdsin = SeqArray::seqOpen(filename.gds)
+  print(paste0("The number of SAMPLES in data: ", length(c(SeqArray::seqGetData(gdsin, "sample.id")))))
+  print(paste0("The number of SNPs in data: ",  length(c(SeqArray::seqGetData(gdsin, "variant.id")))))
+}) # end apply
+
+sample.ids = seqGetData(gdsin, "sample.id")
+sample.strata =  read.table("data/test.metadata.csv", header = T, sep = "\t") %>%
+                  dplyr::select(TARGET_ID, STRATA, LONG, LAT, PLATE)
+id.order = sapply(sample.ids, function(x,df){which(df$TARGET_ID == x)}, df=sample.strata) #incase your strata file is not in the same order as your vcf
+sample.strata.order = sample.strata[id.order,]
+
+# get pruned SNPs
+# NOT WORKING YET, issue with chromosome names
+# Try this with vcf https://github.com/zhengxwen/SNPRelate/issues/10  
+#snpset <- SNPRelate::snpgdsLDpruning(gdsin, ld.threshold=0.2, autosome.only = F, start.pos="random", num.thread=1)  
+#snpset.id <- unlist(unname(snpset))
+
+# PCA only on SNPs with a minor allele freq greater than 2.5%
+pca.out = SNPRelate::snpgdsPCA(autosome.only = F, gdsin, num.thread=2, remove.monosnp = T, maf = 0.025)
+#                               snp.id=snpset.id) # filtering for pruned SNPs
+eig = pca.out$eigenval[!is.na(pca.out$eigenval)]
+barplot(100*eig/sum(eig), main="PCA Eigenvalues")
+```
+
+First color/shape the individuals by the factor you expect to matter (eg, sampling site or region, family, ecotype).  
+```{r, eval=T}
+#PLOT PCA
+#PC1 v PC2 colored by collection location
+id.order = sapply(keep, function(x,df){which(df$ID == x)}, df=sample.strata) #incase your strata file is not in the same order as your vcf
+sample.strata.order = sample.strata[id.order,]
+print(
+  as.data.frame(pca.out$eigenvect) %>%
+      tibble::add_column(., STRATA =  sample.strata.order$STRATA) %>%
+      ggplot(., aes(x=V1, y=V2, color = STRATA)) + 
+      geom_point(size=2) +
+      stat_ellipse(level = 0.95, size = 1) +
+      geom_hline(yintercept = 0) +
+      geom_vline(xintercept = 0) +
+      theme_bw() +
+      xlab(paste0("PC1 [",paste0(round(eig[1], 2)), "%]")) +
+      ylab(paste0("PC2 [",paste0(round(eig[2], 2)), "%]")) +
+      ggtitle("PCA Colored by Collection Location")
+)
+```
+
+#### Questions to ask about your PCA: {.unlisted .unnumbered}
+
+* Is there any clustering? If so, is it different than your expectation?
+  + If different than expectation, suggests either [batch effects](#batch), [cryptic variation](#cryptic) among individuals, or systematic biological issues affecting assembly/mapping
+* Are there outlier samples driving a large amount of variation?
+  + If so, they may be [cryptic species](#cryptic), clones/sample replicates, or they [sequenced/genotyped poorly](#bad)
+  + In our example we see 2 potential outlier samples in the bottom left corner. We can check which samples they are by plotting the PCA with sample ID and/or looking at the actual PC scores.
+
+```{r, eval=T}
+#PLOT PCA
+#PC1 v PC2 colored by collection location
+print(
+  as.data.frame(pca.out$eigenvect) %>%
+      tibble::add_column(., ID =  sample.strata.order$ID) %>%
+      ggplot(., aes(x=V1, y=V2, label = ID)) +
+      geom_text(size =1.5) +
+      geom_hline(yintercept = 0) +
+      geom_vline(xintercept = 0) +
+      theme_bw() +
+      xlab(paste0("PC1 [",paste0(round(eig[1], 2)), "%]")) +
+      ylab(paste0("PC2 [",paste0(round(eig[2], 2)), "%]")) +
+      ggtitle("PCA with Labels")
+)
+```
+
+```{r}
+# which samples have PC1 > -0.7 and PC2 < -0.3?
+as.data.frame(pca.out$eigenvect) %>%
+      tibble::add_column(., ID =  sample.strata.order$ID) %>%
+  filter(V1 > 0.7 & V2 < -0.3) %>%
+  select(ID,V1,V2)
+```
+
+* Is there no clustering at all? If so, it may be due to:  
+  + bioinformatic artifacts leading to noise, [see Testing Parameters](#test)
+  + issues with how you are [filtering the data](#filter) (eg, too strict or too lax)
+  + issues with how your PCA is treating missing data
+  + or you actually have no structure in your dataset!
+  + `r colorize("figure example, no clustering", "red")`
+
+* If you used sample replicates, first see if replicates cluster very close together.
+If not, then there may be batch effects or bioinformatic artifacts leading to genotyping error that must be addressed.
+Then remove replicates for a subsequent PCA evaluation.
+
+Here is an example of a PCA with sample replicates, and then with sample replicates removed:
+
+When removing a sample replicate, you can choose one at random or pick the replicate with the least missing data(link). 
+```{r, eval=FALSE}
+# example code for removing samples from a SNPRelate PCA 
+# assumes you have already loaded a vcf file as a GDS
+samples.to.remove = c("SampleA_rep","SampleB_rep")
+sample.ids = seqGetData(gdsin, "sample.id")
+keep = sample.ids[which(!sample.ids %in% bad_samples)]
+
+# PCA only on SNPs with a minor allele freq greater than 2.5%, only keeping samples in keep
+pca.out.noreps= SNPRelate::snpgdsPCA(autosome.only = F, gdsin, num.thread=1, remove.monosnp = T, maf = 0.05,
+                               snp.id=snpset.id,
+                               sample.id = keep)
+#close gds file (saves memory)
+
+```
+
+Next, color the individuals by potential sources of batch effects (sequencing lane, library prep batch, age of tissue sample, person doing the DNA extraction...).    
+```{r, eval=T}
+# again, but colored by batch
+print(
+  as.data.frame(pca.out$eigenvect) %>%
+      tibble::add_column(., PLATE =  as.factor(sample.strata.order$PLATE)) %>%
+      ggplot(., aes(x=V1, y=V2, color = PLATE))  + #label = ID
+      geom_point(size=2) +
+      stat_ellipse(level = 0.95, size = 1) +
+      geom_hline(yintercept = 0) +
+      geom_vline(xintercept = 0) +
+      theme_bw() +
+      xlab(paste0("PC1 [",paste0(round(eig[1], 2)), "%]")) +
+      ylab(paste0("PC2 [",paste0(round(eig[2], 2)), "%]")) +
+      ggtitle("PCA colored by Batch/Sequencing Plate")
+)
+
+```
+
+* Do any observed clusters match a source of batch effect? [how to remove batch effects](#batch)
+
+#### Visually inspect alignments {#align}  
+Another useful approach is to visually inspect alignments for SNPs or loci that are identified as "weird" in the subsequent analyses.
+In ipyrad this can be done by searching the .loci file, in dDocent...
+If the alignment looks wonky, you can feel better about removing the whole locus (?).
+`r colorize("code example .loci", "red")`
+`r colorize("code dDocent", "red")`  
+`r colorize("code stacks", "red")`
+
+#### Batch effects {#batch}
+As discussed in [Setting up for success](#labwork), batch effects can arise by minor (or not so minor) differences during the library prep and sequencing stage.
+Randomizing samples across libraries and sequencing lanes can help mitigate the influence of batch effects on your downstream analyses, but sometimes this isn't possible, esp.
+when using pre-existing data.
+Even if you randomize your samples, it is a good idea to check for batch effects with a PCA and then try to mitigate them.
+
+Nicolas Lou and Nina Therkildsen have a [great preprint](https://www.authorea.com/users/380682/articles/532568-batch-effects-in-population-genomic-studies-with-low-coverage-whole-genome-sequencing-data-causes-detection-and-mitigation) and [Github repo](https://github.com/therkildsen-lab/batch-effect) discussing causes, detection, and mitigation of batch effects in low coverage WGS, where batch effects can have particularly large effects.
+While some of their recommendations are specific to reference-based analyses, others are relevant to RAD methods too (eg, batch effects due to different sequencing platforms, DNA degradation levels, and sequencing depth).
+
+`r colorize("PCA figure of batch effect", "red")`
+
+You can identify which loci are driving the batch effect by 1) identifying the top SNPs contributing to PC variation (termed "loadings") and/or 2) do an Fst outlier analysis with the batches specified as populations (eg, Bayescan).
+Both of these methods are most appropriate if you've randomized your samples between batches.
+If you have not randomized samples but are certain there is a batch effect driving variation in your dataset, you can still remove loci this way but you may also be removing biologically informative loci.
+
+Let's pretend like we have batch effects that were observed on PC 1. With SNPRelate we would extract the PC loadings for each SNP at each PC, then 
+```{r}
+
+SnpLoad <- snpgdsPCASNPLoading(pca.out, gdsin)
+#load.df <- cbind(SnpLoad$snp.id,SnpLoad$snploading)
+#colnames(load.df) <- c("Snp.ID",)
+
+```
+
+Once you've identified these batch effect SNPs, it can be useful to [visualize the alignments](#align) before removing them.
+If the entire alignment for that locus looks wonky, you should remove all SNPs from that RAD locus from your final dataset before conducting analyses.
+
+#### Cryptic species/contamination/clones {#cryptic}  
+One issue that is particularly common, especially for marine invertebrates, are cryptic, yet highly diverged lineages occurring in the same geographic location.
+If you are trying to do a landscape/seascape genetics study, cryptic species can dramatically throw off population genetic analyses.
+A PCA using all the samples can help identify cryptic species if samples are forming clusters that do not align with any other possible factor (eg, geography, sequencing batch).
+`r colorize("PCA figure of cryptic species", "red")`
+
+Note: Notice the huge variation explained along the first PC-axis.
+
+You can also check heterozygosity, and missingness!
+
+```{r}
+```
+
+ Note: Two outliers are obvious.
+Large differences in heterozygosity between samples are clear, with one location showing two distinct clusters of samples.
+These samples were later determined using mtDNA and morphological analysis to be cryptic species.
+*Unpublished data from a marine species in Australian waters.*
+
+You can then estimate genetic divergence between these clusters using pairwise Fst and/or phylogenetic inference to determine how diverged the groups are.
+If they are very diverged, it may be better to assemble each cryptic species separately, especially for denovo assemblies with dDocent and Stacks, especially since often times restriction cut-sites are not always conserved; see [Wagner et al., 2013](link), [Rubin et al., 2012](link)  These two pipelines used a set of given samples to create a "reference" or "catalog" to which all other samples are mapped.
+If highly diverged samples are included in creating this reference, it can lead to [oversplitting](#cluster).
+
+Cryptic clones can also dramatically throw off population genetic analyses.
+In a PCA, clones will often appear as their own very tight cluster, similar to sample replicates.
+Below the PCA shows x and a simple calculate of pairwise-distance (i.e.
+here manhattan distance) shows x known duplicates, and x unknown clones.
+*other ways to tell they are clones*
+`r colorize("PCA figure of cryptic clones", "red")`
+
+### Test a range of key parameters {#test}
+
+#### Clustering threshold (de novo assembly) {#cluster}  
+During de novo RADseq assembly, reads are assembled into contigs so that each contig represents a single locus.
+This can occur separately for each individual (ipyrad) or jointly for all or a subset of individuals (dDocent, Stacks).
+To call SNPs, in the case of ipyrad, contigs are then clustered and aligned across individuals, while in dDocent and Stacks, reads for each individual are mapped to the "reference" contigs.
+All of these pipelines use sequence similarity thresholds to determine if fragments are homologous (from the same place in the genome) and therefore should be clustered together.
+If the sequence similarity threshold is too low, *undersplitting* can occur so that multiple loci are combined into a single contig.
+Undersplitting is common with repetitive regions and paralogs. It can inflate mean observed heterozygosity.
+
+If the threshold is too high, it can lead to *oversplitting*, where alleles of the same locus are split into two or more contigs. This can lead to reduced mean observed heterozygosity.
+
+If severely off, both of these issues can bias the models used to identify SNPs.
+Undersplitting a little is better than oversplitting, as it is easier to identify lumped paralogs downstream and remove them.
+**What influences the "optimal" clustering threshold of your data?**  
+Clustering threshold is usually determined by how much genetic diversity you have in your dataset.
+If your dataset has high genetic diversity (eg., a single species with high heterozygosity, multiple highly diverged populations, multiple species), then you will need a lower clustering threshold.
+If your dataset has low genetic diversity (a single species with lower heterozygosity, a single family), then you will need a higher clustering threshold.
+Polyploid or highly repetitive (ie large) genomes may also have different requirements than diplod species with smaller genomes.
+Other factors, such as sequencing error rate, may play a role as well so it is usually a good idea to always test a few clustering thresholds with a new dataset.
+However, as long as your clustering threshold is in the ballpark it shouldn't affect downstream analyses (according to Isaac).
+
+There are a few strategies to identify a good clustering threshold for your data, all of which require assembling your data with a range of clustering thresholds.  
+1) Metrics developed by [McCartney-Melstad 2017](link), based on popgen theory and landscape genetic expectations  
+2) Utilize sample replicates (either exact replicates or individuals from the same locality) to quantify genotyping error and minimize error while maximizing the number of informative loci [Mastretta-Yanes 2015](link)
+3) Evaluating the number of haplotypes across loci, with many 3+ haplotypes indicating undersplitting [Harvey 2015](link), [Ilut 2014](link)  
+4) Make assemblies at a range of thresholds and parameters, plot the # of SNPs or contigs per assembly, and make a judgement call [example in dDocent tutorial](link).
+If most parameters give you about the same # of SNPs or contigs, you can feel confident that this parameter doesn't matter much for your data and just choose a reasonable value.
+
+#### Mapping parameters (reference-based and some de novo assemblers)   
+When a sequencing read maps to the wrong assembled contig or place in a reference genome (ie, a non-homologous locus), this can result in erroneous SNPs.
+In many cases, these SNPs can be filtered out later (link to filtering).
+However, data should be checked to make sure there aren't systematic, wide-scale mapping issues that need to be addressed.
+
+For referenced-based assembly, both dDocent and ipyrad make use of bwa(link) to map reads, while Stacks allows the user to choose their aligner. dDocent also utilizes bwa for de novo assemblies.
+
+Mapping defaults for popular RADseq pipelines:
+L: clipping penalty for 5' and 3'
+A: matching score
+O gap open penalty
+T: mapping quality filter
+
+dDocent: bwa mem -L 20,5 -t 8 -a -M -T 10 -A1 -B 3 -O 5 -R
+
+* can change -A and -O during the dDocent pipeline.    
+ipyrad: (defaults) -L 5,5 -B 4 -A1 -O 6 -T 30
+
+* requires changing the ipyrad source code to change parameters
+
+Our whole genome resequencing section has some good content on mapping(link). General considerations with RAD data:
+
+* how many reads aligned for each sample? If many reads didn't aligned, this could be an issue with the reference, read trimming/filtering, or the mapping parameters.
+* What is the distribution of mapping scores?  
+* ratio of mean (across inds) mapping quality scores for reference and alternate allele (O'Leary cite). 0.25-1.75 cut off implemented in dDocent.  
+  + this info might not be coded in all VCFs
+
+#### Read depth threshold {#seq}
+
+Mistakes that occur during sequencing can lead to a base-calling error in a read, which can then lead to an erroneous SNP.
+This type of error most commonly manifests as rare alleles (eg, only observed in one individual) or singletons (only one copy of the minor allele across all individuals).
+A conservative approach to deal with these errors is to filter out all loci with rare alleles (link filtering), but for some demographic analyses this filtering is inappropriate as they require the site frequency spectrum.
+The best ways to minimize the impact of sequencing error during bioinformatic processing is to 1) properly trim and filter your data, and 2) require higher read depth thresholds for assembly and genotyping.
+
+##### 1) Filtering reads
+
+FASTQ files that typically come from the sequencing center include a [Phred scale quality score](link) for each base, which indicates the probability of that base call being correct.
+This information can be used prior to assembly to trim low-quality sections from either end of reads or remove reads altogether.
+Some SNP callers, including FreeBayes which is part of dDocent, also use Phred scores when calling SNPs (ipyrad and Stacks do not).
+All three of the most popular RAD pipelines include steps to filter reads due to low quality bases low quality bases or adapter sequences, but vary slightly on what kinds of read trimming are allowed.
+Always make sure to filter out adapter sequences (this is not default with stacks), otherwise default parameters are usually adequate.
+If your initial [fastqc](#fastqc) indicates that you have many low quality bases in your reads, you may lose too much of your data during filtering.
+In this case, you may want to make your read filtering a little more permissible (keeping in mind that this may increase SNP error rate).
+**Always make sure to check with your sequencing center to see if they have done anything to the raw reads before you received them!**
+<!---turn these into html tabs--->
+
+**dDocent: **
+
+* does not allow you to use reads that have already been trimmed and filtered for de novo assembly, only referenced-based.
+* Uses Trimmomatic to trim Illumina TruSeq adaters and low quality bases. Make sure to change the defaults if you used a different adapter/sequencing system.
+
+**Stacks: **
+
+* requires your reads to be all the same length, so if they were previously trimmed of low quality bases or adapter sequence this must be specified and they will be truncated to the same length.
+
+##### 2) Require higher read depth thresholds for assembly and genotyping.  
+For RADseq, ideally you are generating an average of *at least* 10-20 sequence reads (10x-20x coverage) for each RAD locus in your library preparation (but this doesn't always happen...).
+For diploid species, this would be sequencing each RAD haplotype at least 5x-10x.
+The SNP calling step of RAD pipelines all require setting a minimum number of reads to determine whether a site is homozygous or heterozygous **add image example**.
+If this threshold is too low, sequencing errors are more likely to be included and the SNP caller will have a harder time accurately identifying SNPs ([but see our section on low coverage WGS methods!](link).
+If it's too high, then real polymorphic loci will be excluded.
+All of the three major pipelines allow users to implement read depth thresholds
+
+Generating multiple sequencing reads per locus is helpful when calling SNPs in order to  distinguishing between reads with sequencing errors and.
+
+### Filter your data many ways (and embrace the missingness!)
+* Don't over filter for missingness
+ + papers showing how too much filtering can skew results  
+  + Empirical: Diaz-Arce 2019, Tripp 2017, Silliman 2021, Eaton 2017, Martin-Hernanz (2019)
+  + Simulations: Huang & Knowles 2016, Leaché et al., 2015, Gautier et al., 2012, Rivera-Colón et al., 2021
+
+* some ways to account for missing data in downstream analyses
+ + impute (PCA), subsample (PCA, structure, pretty much anything with SNPs), try a few different missing filters and see what impact it has on results
+ + use [pcANGSD](link)
 
@@
 
-``````{r, echo=FALSE,eval=FALSE}
-library(knitr)
-my.jekyll.site <- "marineomics.github.io"
-KnitPost <- function(input, output) {
-  base.dir <- "~/Postdoc_Projects/RCN_MarineOmics/Website/marineomics.github.io/"
-  base.url <- "/"
-  fig.path <- "RAD_code/images/"
-  #opts_knit$set(base.url = base.url)
-  #fig.path <- paste0("img/", sub(".Rmd$", "", basename(input)), "/")
-  opts_knit$set(base.dir = base.dir, base.url = base.url)
-  opts_chunk$set(fig.path = fig.path) 
-  #opts_chunk$set(fig.path = "../RAD_code/images/")
-  opts_chunk$set(fig.cap = "center")
-  render_jekyll()
-  knit(input, envir = parent.frame(), output = output)
-}
-setwd("~/Postdoc_Projects/RCN_MarineOmics/Website/marineomics.github.io/RAD_code")
-KnitPost("~/Postdoc_Projects/RCN_MarineOmics/Website/marineomics.github.io/RAD_code/RADpage_text.Rmd",output = "~/Postdoc_Projects/RCN_MarineOmics/Website/marineomics.github.io/guidelines/RADpage_text.md")
-```
+# References

### 2021-11-04 Setup.Rmd
Changed format of site to Bootswatch

```{r, setup, include = FALSE, message=FALSE, echo=FALSE}
# library("knitcitations")
library("knitr")
# cite_options(max.names = 3, style = "html", hyperlink = "to.doc")
# bib <- read.bibtex("bibtexlib.bib")
opts_chunk$set(tidy = FALSE, message = FALSE, warning = FALSE, cache = TRUE) 
# use this to set knitr options: 
# http://yihui.name/knitr/options #chunk_options 
#library(printr)
```
---
title: "Required Components"
---

This website is an introduction to some of the principles of reproducible 
science in R. While we are using R and other tools for reproducible research, 
our focus here is not to teach you the tools, but how you can use them to make 
your research more reproducible. Before we get started on the bulk of the
content, we will go over how to install the software needed for this tutorial
and briefly introduce working with R.

Installing R and RStudio
-----

First, install R and RStudio:

1. Download and install the [R](http://www.r-project.org) statistical computing
   and graphing environment. This works cross-platform on Windows, OS X and
   Linux operating systems.

2. Download and install the free, open source edition of the [RStudio
   Desktop](https://www.rstudio.com/products/rstudio/download/) integrated 
   development environment (IDE), which we recommend. This is basically a point-and-click
   interface for R that includes a text editor, file browser, and some other conveniences.

Installing the required packages
----

The following packages are used in this primer:

1. [rmarkdown](http://cran.r-project.org/web/packages/rmarkdown/index.html) (creating reports)
1. [agricolae](http://cran.r-project.org/web/packages/agricolae/index.html) (agricultural disease data analysis)
1. [ggplot2](http://cran.r-project.org/web/packages/ggplot2/index.html) (graphs)
1. [poppr](http://cran.r-project.org/web/packages/poppr/index.html) (genetic data analysis)
1. [dplyr](http://cran.r-project.org/web/packages/dplyr/index.html) (data manipulation)
1. [tidyr](http://cran.r-project.org/web/packages/tidyr/index.html) (data manipulation)

To install these packages, open RStudio and copy and paste the following code into the console:

```{r, eval = FALSE}
my_packages <- c("rmarkdown","poppr", "agricolae", "dplyr", "tidyr", "ggplot2")
install.packages(my_packages, repos = "http://cran.rstudio.com", dependencies = TRUE)
```

Congratulations! You should now be all set for using R. To ensure that
everything is set up correctly, you can go through steps 1 and 2 in the
[markdown chapter](Markdown.html).

Optional software
-----------------

### PDFs with LaTeX

If you want to create PDF documents, you will need a
[$\LaTeX$](https://latex-project.org/) installation. 
For OSX and Ubuntu users, this can be a large download and you will want to ensure you have a good connection:

 - [LaTeX for OSX](https://tug.org/mactex/)
 - [LaTeX for Windows](http://miktex.org/)
 - [LaTeX for Ubuntu](https://help.ubuntu.com/community/LaTeX)

To those that know of and fear LaTeX: don't worry, you don't need to write any LaTeX code to produce PDFs from Markdown.

### Git

Git is a version control program that we will cover, but since installation
requirements can vary between operating systems, we are not requiring it for the
workshop. However, if you would like to install it, [Here's a website that
covers installation for the major operating
systems](http://happygitwithr.com/install-git.html).

A quick introduction to R using RStudio
------------------

Let's review some of the basic features and functions of R. To start R,
open the RStudio application from your programs folder or start menu. This will
initialize your R session. To exit R, simply close the RStudio application.

> Note that R is a case-sensitive language!

Let's get comfortable with R by submitting the following command on the console
(where R prompts you with a `>` in the lower left RStudio window pane) that will
tell you something about your version of R and the packages you have, which is
useful for reporting reproducible research.

```{r, eval = TRUE}
sessionInfo() # This command will tell you information about your current R session
```

> Note that any text after a '#' symbol is a comment and does not affect the code execution;
> you can just type `getwd()` after the ">".

Packages and getting help
-----------------------

One way that R shines above other languages is that R packages in CRAN are all
documented and easy to install. Help files are written in HTML and give the user
a brief overview of:

 - the purpose of a function
 - the parameters it takes
 - the output it yields
 - examples demonstrating its usage

To get help on any R function, type a question mark before the empty function. 
Here's an example of how to get help about the `gather()` function from the
*tidyr* package:

```{r}
library('tidyr') # The package with the gather() function.
?gather          # open the R documentation of the function gather()
```

If you want to run the examples, you can either copy and paste the commands to
your R console, or you can run them all with:

```{r, eval = FALSE}
example("gather", package = "tidyr")
```

Other ways of getting help:

```{r, eval = FALSE}
help(package = "tidyr")  # Get help for a package.
help("gather")           # Get help for the gather function
?gather                  # same as above
??multilocus             # Search for help that has the keyword 'multilocus' in all packages
```

Some packages include vignettes that can have different formats such as being
introductions, tutorials, or reference cards in PDF format. You can look at a
list of vignettes in all packages by typing:

```{r, eval=FALSE}
browseVignettes()                     # see vignettes from all packages
browseVignettes(package = 'poppr')    # see vignettes from a specific package.
```

and to look at a specific vignette you can type:

```{r, eval=FALSE}
vignette('mlg')
```

```{r, echo = FALSE}
#detach("package:printr", unload = TRUE)
```

### 2021-11-04 TOC.Rmd
Changed format of site to Bootswatch

---
title: Table of contents
---

- **[Panel Seminars](panels.html)**

- **Guidelines for Omics Analyses**
    - [Choosing a Population Genomics Approach](popgen_intro.html)
    - ](Population_Strata.html)
    - [Locus-based statistics and missing data](Locus_Stats.html)
    - [Genotypic evenness, richness, and diversity](Genotypic_EvenRichDiv.html)
    - [Linkage disequilibrium](Linkage_disequilibrium.html)
    - [Population structure](Pop_Structure.html)
    - [Minimum Spanning Networks](Minimum_Spanning_Networks.html)
    - [AMOVA](AMOVA.html)
    - [Discriminant analysis of principal components (DAPC)](DAPC.html)

- **Part III: Using genomic data in population genetics**
    - [Population genomics based on high throughut sequencing (HTS)](intro_vcf.html)
    - [Reading VCF data](reading_vcf.html)
    - [Quality control](qc.html)
    - [Analysis of genomic data](analysis_of_genome.html)
    - [Analysis of GBS data](gbs_analysis.html)

- **About**
    - [About the authors](Authors.html)

- **Appendices**
    - [Data sets](Data_sets.html)
    - [Function glossary](funpendix.html)
    - [Source code for this primer](https://github.com/grunwaldlab/Population_Genetics_in_R/)
    - [Introduction to R](intro_to_R.html)

### 2021-11-04 WGSFullWalkthrough.Rmd
Changed format of site to Bootswatch

---
layout: default
title: WGS fastq2vcf walkthrough
parent: Whole Genome Resequencing
grand_parent: Choosing a PopGenomics Approach
editor_options: 
  chunk_output_type: console
---

## Walkthrough Data
In this walkthrough, data files from whole genome sequences at 15x coverage from Atlantic cod will be referenced. Their naming struction are a population ID in the form of "Pop#" with the number being which population the sample came from, a "\_"\, and a 5 digit sample ID which is two digits for the year it was sampled and a three digit sample ID number. Here is an example, "Pop6_18001". We will walk through steps from demultiplexing when unique indexes are used through variant calling with Freebayes.

# **Stacks**
If your samples need to be demultiplexed, running the program Stacks is a necessary first step. Stack’s process_radtags function searches sequence reads for barcodes that you provide and demultiplexes the reads using your barcode file. This file specifies which barcode belongs to which sample and will trim off those barcodes once it demultiplexes.

### Example scenario 
Samples used as examples throughout this walkthrough were multiplexed using 3 different unique sequences. Every sample had a unique combination of an i5 primer, i7 primer and adapter barcode. Data from the sequencing facility will demultiplex to the level of unique combinations of i5 and i7 primers. Therefore for these samples, there were 32 unique i5 and i7 primer combinations and so the sequencing facility sent 64 fastq.gz files: one for the forward sequence and one for the reverse for all 32 combinations. We will take one file as an example in the following code.

### Helpful Links
```process_radtags``` manual and specific walkthrough for ```process_radtags```  
[Manual](https://catchenlab.life.illinois.edu/stacks/manual/)  
[Walkthrough](https://catchenlab.life.illinois.edu/stacks/comp/process_radtags.php)

#### Example Fastq File from Sequencing Facility

##### Read 1
```
[schaal.s@login-00 CodGenomes]$ zcat i5-2-i7-9_R1_001.fastq.gz | head -n 10

@GWNJ-1012:218:GW191226406th:1:1101:1090:1000 1:N:0:GATCAG+ATAGAGGC
ACTTGACTGTGCGTTGGCCTGCGGGCTGACTCGGTCCTGAGATGGACTGCTGTGTAGTTTGAACCATAGATTCATTATATAGAACACGGTCTCCTCTGCGCTGCTGGCCAATGGAGCCGAACGTCCGCACTGGCGGGCGGCCATCTTGCC
+
FF:,FF,FFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFF:FFFFFFFFFFFFFFFFFFFFFFF,:FFFFFFFFF:FFFFFFF:FFFFFFFFFFFFF:FFFFFFF
@GWNJ-1012:218:GW191226406th:1:1101:1687:1000 1:N:0:GATCAG+ATAGAGGC
ACTTGATCCCTCTCACTCTCCTCTCCGTCTCCTCTTTTGTCCTCGTCTCTCTCCTCTCTCCCTCTCTCCCATCTCCCTCTCTATCAAGTAGATCGGAAGAGCACACGTCTGAACTCCAGTCACGATCAGATCTCGTATGCCGTCTTCTGC
+
FFFFFF,FFFFFFFFFFFFFF:FFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFF:FFFFFFFFFFFFFFFFFF:FFFFFFFFFFFFFFFF:FFFFFFFFFFFFFFFFFFFFFFFF:FFFF:::F,F::FFFFF
@GWNJ-1012:218:GW191226406th:1:1101:6840:1000 1:N:0:GATCAG+ATAGAGGC
ACTTGAAAAAAATACATAGCGGCCATGGACAGGATGACCTCTATGACAATGATAGAAACAGAAAGGACGCGGAGACTCTTGAGTCATCAAGTAGATCGGAAGAGCACACGTCTGAACTCCAGTCACGATCAGATCTCGTATGCCGTCTTC
```	
##### Read 2
```
[schaal.s@login-00 CodGenomes]$ zcat i5-2-i7-9_R2_001.fastq.gz | head -n 10

@GWNJ-1012:218:GW191226406th:1:1101:1090:1000 2:N:0:GATCAG+ATAGAGGC
ACTTGAAGGCAAGATGGCCGCCCGCCAGTGCGGACGTTCGGCTCCATTGGCCAGCAGCGCAGAGGAGACCGTGTTCTATATAATGAATCTATGGTTCAAACTACACAGCAGTCCATCTCAGGACCGAGTCAGCCCGCAGGCCAACGCACA
+
FFFFFF,FFFF:FFFFFF:FFFFFFFFFFFFFFFFFF:FFFFFFF:,FFFFF:FF,FFFFFFFFFFFFFFF,F,:FFFF:F,FFFFF:FF,FFFF,FFFFFF,FFF,FF:FFFFF:F:FFFF,FFF:F:F:FFFFFFFFFFF,:FFFFFF
@GWNJ-1012:218:GW191226406th:1:1101:1687:1000 2:N:0:GATCAG+ATAGAGGC
ACTTGAAAGAGAGGGAGATGGGAGAGAGGGAGAGAGGAGAGAGACGAGGACAAAAGAGGAGACGGAGAGGAGAGTGAGAGGGATCAAGTAGATCGGAAGAGCGTCGTGTAGGGAAAGAGTGTGCCTCTATGTGTAGATCTCGGTGGTCGC
+
FFFFFF,FFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFF,FFFF:FFFFFFFFFFFFFFFFFFFFFFFFFFFF:FF,FF:
@GWNJ-1012:218:GW191226406th:1:1101:6840:1000 2:N:0:GATCAG+ATAGAGGC
ACTTGAAGACTCAAGAGTCTCCGCGTCCTTTCTGTTTCTATCATTGTCATAGAGGTCATCCTGTCCATGGCCGCTATGTATTTTTATCAAGTAGATCGGAAGAGCGTCGTGTAGGGAAAGAGTGTGCCTCTATGTGTAGATCTCGGTGGT
```

### Stacks - process_radtags

Barcode File: For each set of paired end files that need to be demultiplexed, a barcode file needs to be created. For example, for files i5-4-i7-11_R1_001.fastq.gz and i5-4-i7-11_R2_001.fastq.gz its barcode file is as follows:

```
ATCACG	ATCACG	Pop6_18001
CGATGT	CGATGT	Pop7_18173
TTAGGC	TTAGGC	Pop8_18148
TGGCCA	TGGCCA	Pop9_18064
ACAGTG	ACAGTG	Pop3_16228
GCCAAT	GCCAAT	Pop4_17210
CAGATC	CAGATC	Pop6_18035
ACTTGA	ACTTGA	Pop6_18038
GATCAG	GATCAG	Pop8_18130
TAGCTT	TAGCTT	Pop8_18113
GGCTAC	GGCTAC	Pop7_18190
CTTGCA	CTTGCA	Pop1_17327
```

This file contains the 12 barcodes that have the i5_4 and i7_11 index and which sample that combination belongs to. This file is tab delimited and because I had inline barcodes, meaning they were the same on both ends of the read, I have the same barcode sequence in the first and second column. Then the sample ID is on the right.

Sample naming convention is important for downstream steps. The way the example samples are named are first a “Pop” identifier for what a priori population they belonged to (see Sample Notes file for example population IDs). Then a 5-digit identifier for the specific sample from that population where the first two digits represent the year the sample was collected and the last 3 digits represent the individual sample ID from that sampling year.

We provide a simple R script to create the barcode files for each primer index combination. See file: Stacks_FileCreate.R

#### Additional Flags

```-p``` path to your genome files. If you have paired data (forward and reverse reads), put your two fastq files in the same folder and only have those two files in it. I created folders for each index combination to keep everything organized and easy to run stacks on.

```-o``` path to where you want your output files to go. For subsequent steps, you'll want all of these demultiplexed files in the same folder. Create a folder like "Stacks_Out" where you write all your files to.

```-b``` path to your barcode file

```--inline_inline ``` The barcode option flag has 6 different options and you have to use the proper flag for your barcodes or indexes need to be demultiplexed. Because my samples have inline barcodes on either end of the sequence I need to use the --inline_inline flag.

```-P``` If paired end sequencing was used you need to add the -P flag to indicate this.

```--disable_rad_check``` this flag is needed if you are doing anything other than RAD-seq. Stacks defaults to requiring cutsite sequences that are used in RAD-seq, but for WGS you won't have cut sites and need to disable the rad check.

```-r``` this flag allows Stacks to rescue any barcodes that are off by one bp

#### Code for running Stacks
```
process_radtags -P -p {Path to genome files} -o .{Path for output} -b {barcode file} --inline_inline --disable_rad_check -r

```

#### Example code for running on a Cluster 
```
#!/bin/bash
#SBATCH --job-name=stacks_i54_i711                # sets the job name
#SBATCH --mem=1Gb                                 # reserves 10 GB memory
#SBATCH --partition=lotterhos                     # requests that the job is executed in lotterhos partition 
#SBATCH --mail-user=schaal.s@northeastern.edu
#SBATCH --mail-type=FAIL
#SBATCH --time=24:00:00                           # reserves machines/cores for 24 hours.
#SBATCH --output=stacksi54_i711.%j.out                # sets the standard output to be stored in file, where %j is the job id
#SBATCH --error=stacksi54_i711.%j.err                 # sets the standard error to be stored in file

process_radtags -P -p i54_i711/GenomeFilesi54_i711/ -o ../Stacks_Out/ -b i54_i711/barcodeFilei54_i711.txt --inline_inline --disable_rad_check -r
```

### Results 
Time: 				each file took approx. 7 hrs  
Memory: 			approx. 80 MB RAM

#### Read 1 Post Stacks for example sample Pop3_17304

```
[schaal.s@login-00 Stacks_Out]$ zcat Pop3_17304.1.fq.gz | head -n 10

@218_1_1101_1090_1000/1
CTGTGCGTTGGCCTGCGGGCTGACTCGGTCCTGAGATGGACTGCTGTGTAGTTTGAACCATAGATTCATTATATAGAACACGGTCTCCTCTGCGCTGCTGGCCAATGGAGCCGAACGTCCGCACTGGCGGGCGGCCATCTTGCC
+
,FFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFF:FFFFFFFFFFFFFFFFFFFFFFF,:FFFFFFFFF:FFFFFFF:FFFFFFFFFFFFF:FFFFFFF
@218_1_1101_1687_1000/1	TCCCTCTCACTCTCCTCTCCGTCTCCTCTTTTGTCCTCGTCTCTCTCCTCTCTCCCTCTCTCCCATCTCCCTCTCTATCAAGTAGATCGGAAGAGCACACGTCTGAACTCCAGTCACGATCAGATCTCGTATGCCGTCTTCTGC
+
,FFFFFFFFFFFFFF:FFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFF:FFFFFFFFFFFFFFFFFF:FFFFFFFFFFFFFFFF:FFFFFFFFFFFFFFFFFFFFFFFF:FFFF:::F,F::FFFFF
@218_1_1101_6840_1000/1
AAAAAATACATAGCGGCCATGGACAGGATGACCTCTATGACAATGATAGAAACAGAAAGGACGCGGAGACTCTTGAGTCATCAAGTAGATCGGAAGAGCACACGTCTGAACTCCAGTCACGATCAGATCTCGTATGCCGTCTTC
```

#### Read 2 Post Stacks for example sample Pop3_17304 
```
[schaal.s@login-00 Stacks_Out]$ zcat Pop3_17304.2.fq.gz | head -n 10

@218_1_1101_1090_1000/2
AGGCAAGATGGCCGCCCGCCAGTGCGGACGTTCGGCTCCATTGGCCAGCAGCGCAGAGGAGACCGTGTTCTATATAATGAATCTATGGTTCAAACTACACAGCAGTCCATCTCAGGACCGAGTCAGCCCGCAGGCCAACGCACA
+
,FFFF:FFFFFF:FFFFFFFFFFFFFFFFFF:FFFFFFF:,FFFFF:FF,FFFFFFFFFFFFFFF,F,:FFFF:F,FFFFF:FF,FFFF,FFFFFF,FFF,FF:FFFFF:F:FFFF,FFF:F:F:FFFFFFFFFFF,:FFFFFF
@218_1_1101_1687_1000/2
AAGAGAGGGAGATGGGAGAGAGGGAGAGAGGAGAGAGACGAGGACAAAAGAGGAGACGGAGAGGAGAGTGAGAGGGATCAAGTAGATCGGAAGAGCGTCGTGTAGGGAAAGAGTGTGCCTCTATGTGTAGATCTCGGTGGTCGC
+
,FFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFF,FFFF:FFFFFFFFFFFFFFFFFFFFFFFFFFFF:FF,FF:
@218_1_1101_6840_1000/2
AGACTCAAGAGTCTCCGCGTCCTTTCTGTTTCTATCATTGTCATAGAGGTCATCCTGTCCATGGCCGCTATGTATTTTTATCAAGTAGATCGGAAGAGCGTCGTGTAGGGAAAGAGTGTGCCTCTATGTGTAGATCTCGGTGGT
```

<br>
<br>

# **Quality Trimming**
Before sequence alignment to a reference genome, you need to trim your sequence data of bases that you have low confidence in being correct. This is to ensure that you have high confidence in both your alignment and the downstream variant calls. There are a few programs that do this including Trimmomatic and Fast P, but this walkthrough will describe Fast P and settings used for this program.

### Helpful Links  
[Fast P Manual](https://github.com/OpenGene/fastp#:~:text=fastp%20supports%20global%20trimming%2C%20which,or%20%2D%2Dtrim_tail1%3D1%20option.)

#### Parameters

```--in1``` This is read 1 for a given sample that uses paired end sequencing. For paired end sequencing, you will have two input files that need to be identified by ```--in1``` and ```--in2```. If doing single end sequencing, you only need the ```--in``` flag.

```--in2``` This is read 2 for a given sample that uses paired end sequencing.

```--out1``` This is the location and name of the out file for the read 1.

```--out2``` This is the location and name of the out file for the read 2.

```-q``` this is used as a filter on the PHRED quality score that each base is given from the sequencing facility. PHRED quality scores range from 0-40 with 40 being the highest quality. Here I removed any base that had a quality score of less than or equal to 15.

```-u``` this is used as a filter on the unqualified percent limit by specifying how many bases are allowed to be unqualified (0-100).

```--trim_front1``` This flag will trim the front base of read 1 by 1. This is necessary if you have an A-overhang on your reads that needs trimming.

```--trim_front2``` This flag will trim the front base of read 2 by 1.

```--cut_front``` this moves a sliding window, based on the size you set with the ```--cut_window_size```, flag from the front 5' to tail and drops bases in the window if its mean quality is below what you set as ```--cut_front_mean_quality```

```--cut_tail``` this moves a sliding window, based on the size you set with the ```--cut_window_size```, flag from the tail 3' end to front and drops bases in the window if its mean quality is below what you set as ```--cut_front_mean_quality```

```--cut_window_size``` the window size option for ```--cut_front```,  Range: 1-1000. The window can slide from either 5′ to 3′ or from 3′ to 5′, and the average quality score within the window is evaluated. If the average quality is lower than the threshold set in ```--cut_mean_quality``` then the bases in the window will be discarded

```--cut_mean_quality``` This is the mean quality requirement option shared by ```--cut_front```, ```--cut_tail``` or ```--cut_sliding```. Range: 1-36

```-j``` outputs a json format report to the file you specify

```-h``` outputs an html format report to the file you specify

#### Code for running fastp with paired end data
```
fastp --in1 {Path to fastq file for forward reads} --in2 {Path to fastq file for reverse reads} --out1 {Path to output quality trimmed forward reads} --out2 {Path to output quality trimmed forward reads} -q {min quality score to keep} -u {percent of unqualified bases allowed} --trim_front1 {number of bases to trim from front of forward read} --trim_front2 {number of bases to trim from front of reverse read} --cut_front --cut_tail --cut_window_size {window size to assess quality} --cut_mean_quality {mean min quality to cut} -j {path to output json file} -h {path to output html file} &> {path to output .log file}

```

#### Example for running on a Cluster
```
#!/bin/bash
#SBATCH --job-name=Pop9_18101_submitFastP.txt
#SBATCH --mem=2Gb
#SBATCH --mail-user=schaal.s@northeastern.edu
#SBATCH --mail-type=FAIL
#SBATCH --partition=lotterhos
#SBATCH --time=1:00:00
#SBATCH --nodes=1
#SBATCH --tasks-per-node=1
#SBATCH --output=FastP_Out/jobSum/clustOut/Pop9_18101.%j.out
#SBATCH --error=FastP_Out/jobSum/clustOut/Pop9_18101.%j.err

fastp --in1 Stacks_Out/Pop9_18101.1.fq.gz --in2 Stacks_Out/Pop9_18101.2.fq.gz --out1 FastP_Out/Pop9_18101.R1.fq.gz --out2 FastP_Out/Pop9_18101.R2.fq.gz -q 15 -u 50 --trim_front1 1 --trim_front2 1 --cut_front --cut_tail --cut_window_size 5 --cut_mean_quality 15 -j FastP_Out/jobSum/Pop9_18101.fp.json -h FastP_Out/jobSum/Pop9_18101.fp.html &> FastP_Out/jobSum/Pop9_18101.fp.trim.log
```

#### Results
Time: 	20-30 min per sample
Memory: 1.5 - 2GB of RAM

#### Other programs
Trimmomatic
Stacks

<br>
<br>

# **Alignment**
There are multiple programs used for aligning short reads to a reference: Bowtie, Bowtie2, BWA, STAR, etc. Here I will describe the steps I took using BWA.

### Helpful Links
[bwa Github](https://github.com/lh3/bwa)
[bwa Manual](http://bio-bwa.sourceforge.net/bwa.shtml)

### First step: index your genome 
Download your genome from the [NCBI database](https://www.ncbi.nlm.nih.gov/) or [UCSC database](http://hgdownload.cse.ucsc.edu/downloads.html) and run bwa index to index the genome. This allows BWA to essentially make a catalog (i.e., index) of the locations in the genome to make alignment faster.

#### Code for running BWA index
```
bwa index {Path to .fna genome file}
```

#### Example code for running on a cluster
```
#!/bin/bash
#SBATCH --job-name=subGenomesBWA                              
#SBATCH --mem=2Gb
#SBATCH --mail-user=schaal.s@northeastern.edu
#SBATCH --mail-type=FAIL
#SBATCH --partition=lotterhos
#SBATCH --time=6:00:00
#SBATCH --nodes=1
#SBATCH --tasks-per-node=1                        
#SBATCH --output=indexGenomeBWA.%j.out                
#SBATCH --error=indexGenomeBWA.%j.err

bwa index Cod_genome/GCF_902167405.1_gadMor3.0_genomic.fna
```
#### Results
Time: 9 minutes 
Memory: 1 GB of RAM

### Second step: alignment with BWA-MEM
There are a number of parameters to change in the bwa alignment. It is worth playing with parameters and outputting summary information to evaluate how well the alignment worked.

Dr. Jon Puritz provides an open sourced [guide](https://github.com/jpuritz/Winter.School2018/blob/master/Exercises/Day1/Mapping%20Exercise.md) for testing parameters to ensure you get the best alignment results (DOUBLE CHECK WITH HIM THAT IT IS OKAY FOR US TO PUT THIS IN HERE). General advice is that a gap penalty of 5 and a mismatch penalty of 3 consistently gives better results with a lot of marine species. However, it’s worth doing a couple of test runs with 3–5 samples comparing that to the default.

#### Arguments

```-B``` mismatch penalty used to adjust alignment scores based on the number of mismatches in the aligned sequence

```-O``` gap penalty used to adjust alignment scores based on the number and length of gaps and mismatches are when there are bases that deviate from the reference that the read is aligning to. This helps alleviate sequencing errors influencing the alignment.

```-a``` output all found alignments for SE or unpaired paired end; flagged as secondary alignments

```-M``` mark shorter split hits as secondary

```-R``` RGline (reference) because you will eventually be merging files this flag is important to add in the sample ID associated with these reads in the actual alignment file. This will remain in the file then for downstream analyses.

#### Code for running BWA index
```
bwa mem -O {num} -B {num} -a -M -R {Path to .fna genome file} {Path to sample .fg.gz forward read} {Path to sample .fg.gz reverse read}  > {Path to outfile location for .sam file}

```

#### Example for running on a Cluster

```
#!/bin/bash
#SBATCH --job-name=Pop3_17304_submitBWA
#SBATCH --mem=3Gb
#SBATCH --mail-user=schaal.s@northeastern.edu
#SBATCH --mail-type=FAIL
#SBATCH --partition=lotterhos
#SBATCH --time=15:00:00
#SBATCH --nodes=1
#SBATCH --tasks-per-node=1
#SBATCH --output=BWA_Out/clustOut/Pop3_17304.%j.out
#SBATCH --error=BWA_Out/clustOut/Pop3_17304.%j.err

bwa mem -O 5 -B 3 -a -M -R BWA_genome/GCF_902167405.1_gadMor3.0_genomic.fna FastP_Out/Pop3_17304.R1.fq.gz FastP_Out/Pop3_17304.R2.fq.gz  > BWA_Out/Pop3_17304aln.sam

```
#### Results (per sample):
Time: 7-10 hours 
Memory: about 1.5 GB

### Third Step: check alignment stats using samtools view and flagstat

#### Code for running samtools 
```
samtools view -Sbt {Path to .fna genome file} {Path to .sam aligned sample file} | samtools flagstat -
```

#### Example for running on a Cluster
```
#!/bin/bash
#SBATCH --job-name=Pop7_18156_alnCheckDef
#SBATCH --mem=2Gb
#SBATCH --mail-user=schaal.s@northeastern.edu
#SBATCH --mail-type=FAIL
#SBATCH --partition=lotterhos
#SBATCH --time=4:00:00
#SBATCH --nodes=1
#SBATCH --tasks-per-node=1
#SBATCH --output=clustOut/Pop7_18156_def.%j.out
#SBATCH --error=clustOut/Pop7_18156_def.%j.err

samtools view -Sbt ../BWA_genome/GCF_902167405.1_gadMor3.0_genomic.fna ../BWA_Out/Pop7_18156alnDef.sam | samtools flagstat -
```

#### example output from flagstat (sample: Pop1_16216):
```
75779849 + 0 in total (QC-passed reads + QC-failed reads)
0 + 0 secondary
1183259 + 0 supplementary
0 + 0 duplicates
74456805 + 0 mapped (98.25% : N/A)
74596590 + 0 paired in sequencing
37298295 + 0 read1
37298295 + 0 read2
70915282 + 0 properly paired (95.07% : N/A)
73157116 + 0 with itself and mate mapped
116430 + 0 singletons (0.16% : N/A)
2026064 + 0 with mate mapped to a different chr
739870 + 0 with mate mapped to a different chr (mapQ>=5)

```
Here you want to maximize your properly paired reads and percent mapped to the genome. Try a few parameters and see how you can maximize this.

Line 1 - total: the total number of alignments that pass the quality filter, this isn’t the total number  of reads. To get the total number of reads, subtract the secondary, supplementary, and duplicate values from this number which will be R1 + R2.  
Line 2 - secondary: (0x100 bit set)
Line 3 - supplementary: Supplementary alignments are those alignments that are part of a chimeric alignment (0x800 bit set)
Line 4 - duplicates: these mark potential PCR duplicates. This value should be zero if you filtered out all the pcr duplicates using picard tools (0x400 bit set)
Line 5 - mapped: (0x4 bit not set)
Line 6 - paired in sequencing: (0x1 bit set)
Line 7 - read1: (both 0x1 and 0x40 bits set)
Line 8 - read2: (both 0x1 and 0x80 bits set)
Line 9 - properly paired: (both 0x1 and 0x2 bits set and 0x4 bit not set)
Line 10 - with itself and mate mapped: (0x1 bit set and neither 0x4 nor 0x8 bits set)
Line 11 - singletons: Singleton is a mapped read whose mate is unmapped (both 0x1 and 0x8 bits set and bit 0x4 not set)
Line 12 - with mate mapped to different chr: low quality reads mapping to different chromosomes
Line 13 - with mate mapped to different chr (mapQ>=5): higher quality reads mapping to different chromosome

<br>
<br>

# **Sort and index your alignment**
Once you have aligned your reads to the reference you need to sort and index them for downstream use. This can be done using samtools.

### Helpful links
[Samtools Manual](http://www.htslib.org/doc/samtools.html)

### Step 1: convert your sam files to bam
Bam files are much smaller than sam files. By converting your files to Bam, you willincrease downstream efficiency.

####  Code for running samtools for Sam to Bam conversion
```
samtools view -S -b {Path to .sam file} > {Path for output .bam file}
```

#### Example for running on a Cluster
```
#!/bin/bash
#SBATCH --job-name=Pop4_17226_alnSamToBam
#SBATCH --mem=2Gb
#SBATCH --mail-user=schaal.s@northeastern.edu
#SBATCH --mail-type=FAIL
#SBATCH --partition=lotterhos
#SBATCH --time=4:00:00
#SBATCH --nodes=1
#SBATCH --tasks-per-node=1
#SBATCH --output=samtools_samToBam_Out/clustOut/Pop4_17226.%j.out
#SBATCH --error=samtools_samToBam_Out/clustOut/Pop4_17226.%j.err

samtools view -S -b BWA_Out/Pop4_17226aln.sam > samtools_samToBam_Out/Pop4_17226aln.bam
```
#### Results (per sample):
Time: 35-40 minutes 
Memory: about 20MB

<br>

### Step 2: Sort your bam files
Sorting files takes a lot of memory, but not a lot of time. There is one trick about the memory part to keep in mind. You need to set memory for the cluster, but you also need to set the memory allotment allowed for samtools sort to store data in. If you make the memory in samtools too low it will create 100s to 1000s of temporary files that it will later need to knit back together. For the example shown here, the memory was increased so that samtools didn't have to create temporary files. HOWEVER, you need to set the memory to a value that is lower than what you give to the cluster. If it is exactly the same or near the amount you gave to the cluster, it will fail and say you will get an error saying the program went over the memory limits. Therefore, for this example 40GB of memory was set in the samtools code and 50 GB of memory was set for each sample on the cluster.

#### Code for running samtools for sorting bam file
```
samtools sort -m {memory allocation to samtools} {Path to .bam file} > {Path for output sorted .bam file}
```

#### Example for running on a Cluster
```
#!/bin/bash
#SBATCH --job-name=Pop4_17226_alnSortedBam
#SBATCH --mem=50Gb
#SBATCH --mail-user=schaal.s@northeastern.edu
#SBATCH --mail-type=FAIL
#SBATCH --partition=lotterhos
#SBATCH --time=24:00:00
#SBATCH --nodes=1
#SBATCH --tasks-per-node=1
#SBATCH --output=samtools_sortedBam_Out/clustOut/Pop4_17226.%j.out
#SBATCH --error=samtools_sortedBam_Out/clustOut/Pop4_17226.%j.err

samtools sort -m 40G samtools_samToBam_Out/Pop4_17226aln.bam > samtools_sortedBam_Out/Pop4_17226aln.sorted.bam
```

#### Results (per sample):
Time: about 10 min  
Memory: 40GB

<br>

### Step 3: Index your sorted bam file

#### Code for running samtools for indexing sorted bam file
```
samtools index {Path to sorted .bam file} > {Path for output sorted and indexed .bam file}
```

#### Example for running on a Cluster
```
#!/bin/bash
#SBATCH --job-name=Pop4_17226_alnIndexBam
#SBATCH --mem=50Mb
#SBATCH --mail-user=schaal.s@northeastern.edu
#SBATCH --mail-type=FAIL
#SBATCH --partition=lotterhos
#SBATCH --time=2:00:00
#SBATCH --nodes=1
#SBATCH --tasks-per-node=1
#SBATCH --output=samtools_indexedBam_Out/clustOut/Pop4_17226.%j.out
#SBATCH --error=samtools_indexedBam_Out/clustOut/Pop4_17226.%j.err

samtools index samtools_sortedBam_Out/Pop4_17226aln.sorted.bam > samtools_indexedBam_Out/Pop4_17226aln_sorted_indexed
```

#### Results (per sample):
Time: about 3-5 min  
Memory: about 15 MB

<br>
<br>

# **Checking alignment**

### First step: identifying regions to calculate coverage
It is important to check your coverage along each chromosome to determine whether you have regions of low or no coverage that could bias downstream analyses. One way to evaluate this is to first download the GFF file from NCBI for the genome of your species. As an example here is a link to the Atlantic cod genome used for the analysis examples in this walkthrough: [Atlantic cod genome](https://www.ncbi.nlm.nih.gov/genome/?term=txid8049[orgn])

Then subset the genome file for the chromosomes or genomic regions you want to evaluate coverage. Below is an example of subsetting the Atlantic cod genome for all 23 chromosomes in the R programming language: 
```
cod.gff <- read.table("src/alignment/GCF_902167405.1_gadMor3.0_genomic.gff", sep = "\t", quote = "")
scaffolds.gff <- cod.gff[cod.gff$V3 == "region",]
chrom.gff <- scaffolds.gff[1:23,]
write.table(chrom.gff, "src/alignment/GCF_902167405.1_gadMor3.0_genomic_chroms.gff", row.names = FALSE, 
            sep = "\t", col.names = FALSE, quote = FALSE)
```

### Second step: calculating coverage

#### Code for running bedtools coverage 
```
bedtools coverage -a {Path to GFF file created in previous step} -b {Path to aligned, sorted and indexed .bam file} -sorted -d > {Path for output .txt file}

```

#### Example for running on a Cluster
```
#!/bin/bash
#SBATCH --job-name=Pop1_16216_bedCov
#SBATCH --mem=50Gb
#SBATCH --mail-user=schaal.s@northeastern.edu
#SBATCH --mail-type=FAIL
#SBATCH --partition=lotterhos
#SBATCH --time=4:00:00
#SBATCH --nodes=1
#SBATCH --tasks-per-node=1
#SBATCH --output=bedtools_coverage/clustOut/Pop1_16216bedCov.%j.out
#SBATCH --error=bedtools_coverage/clustOut/Pop1_16216bedCov.%j.err
bedtools coverage -a Cod_genome_data/GCF_902167405.1_gadMor3.0_genomic_chroms.gff -b samtools_sortedBam_Out/Pop1_16216aln.sorted.bam -sorted -d > bedtools_coverage/Pop1_16216.coverageCalcDflag.txt 
```
```-a``` reference genome

```-b``` sorted bam file

```-d``` give per base coverage

```-sorted``` tells bedtools that this file is already sorted

### Third step: subsetting
Subset for just the chromosome data and the columns of interest in the output file (reduces file sizes from 111GB to 17GB this part can easily be piped so you aren't creating that large intermediate file)  
```
#!/bin/bash
#SBATCH --job-name=Pop1_16216_alnCheck
#SBATCH --mem=2Gb
#SBATCH --mail-user=schaal.s@northeastern.edu
#SBATCH --mail-type=FAIL
#SBATCH --partition=lotterhos
#SBATCH --time=4:00:00
#SBATCH --nodes=1
#SBATCH --tasks-per-node=1
#SBATCH --output=bedtools_coverage/clustOut/Pop1_16216awk.%j.out
#SBATCH --error=bedtools_coverage/clustOut/Pop1_16216awk.%j.err
awk -F"\t" '$1~/NC*/' bedtools_coverage/Pop1_16216.coverageCalcDflag.txt | awk '{print $1,$11,$12}' > bedtools_coverage/Pop1_16216.coverageCalcChr.txt
```

### Fourth step: calculate coverage average 
run the output file through the R script for calculating averages for different window sizes and plot results in ggplot   
```

### PROCESS COVERAGE DATA TO VISUALIZE ACROSS THE GENOME ###
## Sara M. Schaal

## LOAD LIBRARIES
library(ggplot2)
library(data.table)

## USER INPUTS
setwd("/scratch/schaal.s/CodGenomes") # SET YOUR WORKING DIRECTORY
samps <- c("Pop1_16216", "Pop1_17291", "Pop4_17236", "Pop5_17278", "Pop6_18017") # INPUT SAMPLES OF INTEREST
colors <- c("steelblue2", "chartreuse3", "orchid2", "firebrick1", "goldenrod2") # INPUT COLORS 
n <- 10000 # INPUT SIZE OF WINDOWS TO CALCULATE AVERAGE COVERAGE

## PROCESS DATA 
# first step through each sample
for(i in 1:length(samps)){
  df <- fread(paste0("bedtools_coverage/", samps[i], ".coverageCalcChr.txt"), 
                                                      sep =" ", quote = "", data.table = FALSE)
  colnames(df) <- c("chrom", "base", "coverage")
  chroms <- unique(df[,1])
  df.sample.data <- NULL
  # step through each chromosome and break into the increment chunks you set with n
  for(j in 1:length(chroms)){
    df.chrom <- df[df$chrom == chroms[j], ]
    extra <- nrow(df.chrom) %% n
    # this next part is taking our windows set by n and giving each window a number id
    # this grouped dataframe is made by binding the original df.chrom with grouping values for the window size you want 
    # then finding what the last grouping value would be using because it will most like not be an even increment of your n
    # for example if you divide a chromosome by your n and get 3000.3 then you can easily input the first 3000
    # grouping values in with rep (middle part of the following cbind function) then the last grouping value will be 3001 which
    # you get by rounding using ceiling in the last part of this cbind
    grouped <- cbind(df.chrom, c(rep(1:(nrow(df.chrom)/n), each = n), rep(ceiling(nrow(df.chrom)/n), extra)))
    colnames(grouped)[4] <- "grouping"
    # now bind this chromosomes data in the full dataframe
    df.sample.data <- rbind(df.sample.data, grouped)
  }
  # finally take your new dataframe and calculate the average coverage for your increments using
  # this new grouping variable and the chromosome
  df.covAve <- aggregate(coverage~grouping + chrom, data = df.sample.data, FUN = mean)

  #### PLOTTING ####
  pdf(paste0("figures/", samps[i], "MaxcoveragePlot", n, ".pdf"), height= 15, width=15)

  ## I make two plots because there will undoubtedly be some loci that have really high coverage which makes
  # it hard to see all the spread of the majority of the data. This first graph is set to the max coverage
  # found in the data frame and the second plot is setting your y limit to a more reasonable value for your
  # data. For me 100 x was good but feel free to change to what is appropriate for your data.

  print(ggplot(data = df.covAve, aes(x = grouping, y = coverage)) +
    geom_point(col = colors[i], alpha = 0.5) +
    facet_wrap(~chrom) +
    labs(y = "Coverage", x = paste0("location every ", n, " bases"), 
         title = paste0(samps[i], "Genome Coverage up to Max Coverage")) +
    ylim(0, max(df.covAve$coverage)) +
    xlim(0, max(df.covAve$grouping)) +
    theme_bw() + 
    theme(panel.border = element_blank(), panel.grid.major = element_blank(), 
          panel.grid.minor = element_blank(),
          axis.line = element_line(colour = "black"), 
          legend.position = "none"))
  dev.off()

  pdf(paste0("figures/", samps[i], "100XcoveragePlot", n, ".pdf"), height= 15, width=15)

  print(ggplot(data = df.covAve, aes(x = grouping, y = coverage)) +
          geom_point(col = colors[i], alpha = 0.5) +
          facet_wrap(~chrom) +
          labs(y = "Coverage", x = paste0("location every ", n, " bases"),
               title = paste0(samps[i], "Genome Coverage up to 100X")) +
          ylim(0, 100) +
          xlim(0, max(df.covAve$grouping)) +
          theme_bw() + 
          theme(panel.border = element_blank(), panel.grid.major = element_blank(), 
                panel.grid.minor = element_blank(),
                axis.line = element_line(colour = "black"), 
                legend.position = "none"))
  dev.off()

}
```
<!-- <img src="./Pop1_17291100XcoveragePlot10000.png" width="500">  -->

#### Results (R script for 5 samples):
Time: about 4.5 hours  
Memory: about 85 GB

<br>
<br>

# **Picard MarkDuplicates**

The tool's main output is a new SAM or BAM file, in which duplicates have been identified in the SAM flags field for each read. Duplicates are marked with the hexadecimal value of 0x400, which corresponds to a decimal value of 1024. For the example data shown in this walkthrough,  OPTICAL_DUPLICATE_PIXEL_DISTANCE=2500 needs to be set which is for data coming from a patterned flow cell. The example whole genome data comes from a NovaSeq run which uses the patterned flow cell. The default for that flag is only 100 which is appropriate for unpatterned flow cells. Make sure you know which technology your data come from and set this parameter as needed.

### Helpful Links
[Picard MarkDuplicates Manual](https://gatk.broadinstitute.org/hc/en-us/articles/360037052812-MarkDuplicates-Picard-)
[Breakdown and explanation of different duplicates from Illumina sequencing](http://core-genomics.blogspot.com/2016/05/increased-read-duplication-on-patterned.html)

#### Code for running PicardTools
```
picard MarkDuplicates I={Path to aligned, sorted and indexed .bam file} O={Path for output of duplicated marked .bam file} M={Path for metrics output file} OPTICAL_DUPLICATE_PIXEL_DISTANCE={Value specific to sequencing technology} TAGGING_POLICY={Type of duplicates to mark} &> {Path to .log file}

```

#### Example for running on a Cluster
```
#!/bin/bash
#SBATCH --nodes=1
#SBATCH --time=8:00:00
#SBATCH --job-name=picard_benchmark
#SBATCH --partition=short
#SBATCH --cpus-per-task=8
#SBATCH --output=picard_log.o
#SBATCH --error=picard_log.e

picard MarkDuplicates I=../samtools_sortedBam_Out/Pop9_18093aln.sorted.bam O=../picard_Out/Pop9_18093aln.sorted.md.bam M=md_metrics.txt OPTICAL_DUPLICATE_PIXEL_DISTANCE=2500 TAGGING_POLICY=OpticalOnly &> pop9_18093aln.sorted.md.log
```
#### Results (per sample):
Time: about 30 min  
Memory: about 2-3GB

<br>
<br>

# **Samtools filtering**

Before running the SNP caller, you need to filter the samples to ensure you are only using good quality aligned data. To do this, the bam file is filtered using samtools view.

#### Code from dDocent for quality filtering: 
```
samtools view -@32 -h -q 10 -F 0x100 -F 0x400 $1-RGmd.bam | mawk '$6 !~/[8-9].[SH]/ && $6 !~ /[1-9][0-9].[SH]/' | samtools view -@ 32 -b
```

In dDocent, all reads are filtered if mapping quality is less than 10, removes PCR duplicates (marked in previous Picard step this quality filter then removes them) with the -F 0x400 flag and removes non primary alignments with the 0x100. The mawk line of code then removes hard and soft clipped reads. Finally, the last samtools view code exports the results as a bam file.

```-@``` option allocates threads for the job

```-h``` include the header in the output

```-q``` skip alignments with MAPQ smaller than INT[0]

```-F``` Do not output alignments with any bits set in INT present in the FLAG field. INT can be specified in hex by beginning with 0x' (i.e. /^0x[0-9A-F]+/) or in octal by beginning with 0' (i.e. /^0[0-7]+/) [0]

```-b``` output in the bam format

#### Example for an altered dDocent code
Changed the mawk code to not remove the hard clipped reads because I am interested in inversion breakpoints. This would most likely filter those out.

#### Example for running on a cluster
```
#!/bin/bash
#SBATCH --job-name=samtoolsFilter
#SBATCH --mem=750Mb
#SBATCH --mail-user=schaal.s@northeastern.edu
#SBATCH --mail-type=FAIL
#SBATCH --partition=lotterhos
#SBATCH --time=0:30:00
#SBATCH --cpus-per-task=2
#SBATCH --output=../samtools_filter_Out/clustOut/samtoolsFilter.%j.out
#SBATCH --error=../samtools_filter_Out/clustOut/samtoolsFilter.%j.err

samtools view -@2 -h -q 10 -F 0x100 -F 0x400 ../picard_Out/Pop1_16216aln.sorted.md.bam | mawk '$6 !~ /[1-9][0-9].[SH]/'| samtools view -@2 -b > ../samtools_filter_Out/Pop1_16216.f.bam

```
#### Results (per sample):
Time: 15-20 min  
Memory: about 500MB

<br>
<br>

# **Merge Reads**

Before calling SNPs you need to merge all your per sample Bam files. This is why we had to add the RG ids during the mapping step. This ensures all reads have a unique identifier for which sample they belong to once merged. By merging all the reads, the SNP caller doesn’t need to keep opening and closing individual sample files, which will increase the runtime. You need to make a text file that includes all the bamfile names that you want to merge and then run the line of code below. Because this is computationally intensive, I used 32 cpus on a single compute node.

### Helpful Links
[Samtools Merge Manual](http://www.htslib.org/doc/samtools-merge.html)

#### Code for running samtools merge
```
BAMLIST={Path to .txt Bamfile list}
samtools merge {Path for output of merged .bam file} -b ${BAMLIST} -@{number of threads to use}

```

#### Example for running on a Cluster
```
#!/bin/bash
#SBATCH -p short
#SBATCH --nodes 1
#SBATCH --cpus-per-task=32
#SBATCH -t 120:00:00
#SBATCH --constraint=zen2
#SBATCH --mem=5G
#SBATCH -o clustOut/slurm.%N.%j.out
#SBATCH -e clustOut/slurm.%N.%j.err
#SBATCH --job-name="Samtools-merge_test"
#SBATCH --mail-type=END,FAIL
#SBATCH --mail-user=schaal.s@northeastern.edu

BAMLIST=./bamlist.txt
samtools merge mergedBam_n32_all.bam -b ${BAMLIST} -@32

```
#### Results (merging 296 samples):
Time: about 36 hours  
Memory: about 150MB

<br>
<br>

# **SNP calling**

There are two major variant callers that are widely used: GATK and Freebayes. Here we walk through using freebayes-parallel to optimize runtime on our cluster. SNP calling is computationally intensive so this step will take some work to optimize both the size of genome chunks you want to analyze at a time and how many cpus to run your samples on.

There are a few files that you need for this step: 1) your population list file which has a line per sample where the sample name is the first item in the line and the population ID is the second item in the line (see example header of a population list file below) 2) your .fna genome reference that you are calling snps from and 3) regions file where you tell freebayes what size chunks of the genome you want to analyze at a time. Depending on your coverage, this last file is very important to troubleshoot to maximize memory use vs. run times. In the example data, we found we were reaching about 90% efficiency on our cluster with 9-13 hour runs using 64 cores and allotting 90 GB of memory when we split the genome up into 100kb chunks. So this is how we proceeded and ran each chromosome individually by splitting them each up into 100kb chunks.

### Helpful links
[Freebayes Github Page](https://github.com/freebayes/freebayes)
[Freebayes README](https://ngseasy.readthedocs.io/en/latest/containerized/ngseasy_dockerfiles/ngseasy_freebayes/README/)

#### example header of a population list file
```
Pop4_17219	Pop4
Pop2_17008	Pop2
Pop9_18076	Pop9
Pop1_16232	Pop1
Pop7_18161	Pop7
Pop8_18147	Pop8
```
#### example header of a regions list file
```
NC_044048.1:1-100000
NC_044048.1:100001-200000
NC_044048.1:200001-300000
NC_044048.1:300001-400000
NC_044048.1:400001-500000

```

#### Code for running freebayes parallel
```
POPFILE={Path to population .txt file}
REF={Path to .fna reference genome file}
freebayes-parallel {path to regions .txt file} {num threads} -f ${REF} -b {Path to merged bam file} --populations ${POPFILE} -m {num} -q {num} -E {num} --min-repeat-entropy {num} -n {num} -F {fraction} >> {Path for .vcf out file}

```

#### Arguments

```-m```--min-mapping-quality Q. Exclude alignments from analysis if they have a mapping quality less than Q

```-q``` --min-base-quality Q. Exclude alleles from analysis if their supporting base quality is less than Q

```-E``` Allow complex alleles with contiguous embedded matches of up to this length

```-n``` --use-best-n-alleles N. Evaluate only the best N SNP alleles, ranked by sum of supporting quality scores.

```-F``` --min-alternate-fraction N. Require at least this fraction of observations supporting an alternate allele within a single individual in order to evaluate the position

#### Example for running on a Cluster
```
#!/bin/bash
#SBATCH -p long
#SBATCH --nodes 1
#SBATCH --cpus-per-task=64
#SBATCH -t 120:00:00
#SBATCH --constraint=zen2
#SBATCH --mem=90GB
#SBATCH -o noRegion.%N.%j.out
#SBATCH -e noRegion.%N.%j.err
#SBATCH --job-name="Freebayes_parallel_test"
#SBATCH --mail-type=END,FAIL
#SBATCH --mail-user=schaal.s@northeastern.edu

POPFILE=/scratch/schaal.s/CodGenomes/10_freebayes/poplist.txt
REF=/scratch/schaal.s/CodGenomes/Cod_genome_data/GCF_902167405.1_gadMor3.0_genomic.fna
freebayes-parallel regionsFiles/NC_044048.1_100kbRegions.txt 128 -f ${REF} -b ../labeled_bam_Out/mergedBam_n128_all_lot.bam --populations ${POPFILE} -m 5 -q 5 -E 3 --min-repeat-entropy 1 -n 10 -F 0.1 >> outFiles/VarCall_freebayes-par.chrom_NC_044048.1.vcf

```

#### Results (running the 23 chromosomes across nodes with 64 cores each): 
Time: 6-12 hours each (total about 36 hours)   
Memory: 60-65GB

```{r, echo=FALSE,eval=FALSE}
library(knitr)
my.jekyll.site <- "marineomics.github.io"
KnitPost <- function(input, base.url = my.jekyll.site) {
  opts_knit$set(base.url = base.url)
  fig.path <- paste0("images/", sub(".Rmd$", "", basename(input)), "/")
  opts_chunk$set(fig.path = fig.path)
  opts_chunk$set(fig.cap = "center")
  render_jekyll()
  knit(input, envir = parent.frame())
}

setwd("~/Postdoc_Projects/RCN_MarineOmics/Website/marineomics.github.io/guidelines")
KnitPost("~/Postdoc_Projects/RCN_MarineOmics/Website/marineomics.github.io/guidelines/WGSFullWalkthrough.Rmd")

```

### 2021-11-04 WGS_intro.Rmd
Changed format of site to Bootswatch

--- 2021-10-26 guidelines/WGS_intro.Rmd
@@
 ---
-layout: default
-title: Whole Genome Resequencing
-parent: Choosing a PopGenomics Approach
-grand_parent: Best Principles
-has_children: true
-#editor_options: 
-#  chunk_output_type: console
+bibliography: bibtexlib.bib
 ---
@@
 knitr::opts_chunk$set(echo = TRUE)
+library(knitcitations)
+opts_chunk$set(fig.width = 10,
+               fig.height = 5,
+               cache = FALSE)
+cite_options(citation_format = "pandoc", max.names = 3, style = "html", 
+             hyperlink = "to.doc")
 ```
 
-# Choosing a PopGenomics Approach  
-{: .no_toc }
-
-<details open markdown="block">
-  <summary>
-    Table of contents
-  </summary>
-  {: .text-delta }
-1. TOC
-{:toc}
-</details>
-
-## Whole Genome Resequencing for Population Genomics  
+# Whole Genome Resequencing for Population Genomics  
 For the purposes of this guide, we use "whole genome (re)sequencing (WGS)) to refer to  methods where a reference genome exists (whether for the focal species or a related species), and samples are sequenced and then mapped to most or all of the reference genome.
@@
 
-### Tutorials
+## Tutorials
 
@@
 
-## Considerations Before Lab Work
-
-```{r, echo=FALSE,eval=FALSE}
-library(knitr)
-my.jekyll.site <- "marineomics.github.io"
-KnitPost <- function(input, base.url = my.jekyll.site) {
-  opts_knit$set(base.url = base.url)
-  fig.path <- paste0("images/", sub(".Rmd$", "", basename(input)), "/")
-  opts_chunk$set(fig.path = fig.path)
-  opts_chunk$set(fig.cap = "center")
-  render_jekyll()
-  knit(input, envir = parent.frame())
-}
-
-setwd("~/Postdoc_Projects/RCN_MarineOmics/Website/marineomics.github.io/guidelines/")
-KnitPost("~/Postdoc_Projects/RCN_MarineOmics/Website/marineomics.github.io/guidelines/WGS_intro.Rmd")
-
-```
+# Considerations Before Lab Work

### 2021-11-04 dapc.Rmd
Changed format of site to Bootswatch

```{r,message = FALSE,echo = FALSE}
  library("knitcitations")
  library("knitr")
  cite_options(citation_format = "pandoc", max.names = 3, style = "html", hyperlink = "to.doc")
  bib <- read.bibtex("bibtexlib.bib")
  opts_chunk$set(tidy = FALSE, message = FALSE, warning = FALSE,
                 fig.width = 10, fig.height = 6, cache = TRUE)
  # use this to set knitr options:
  # http://yihui.name/knitr/options #chunk_options
```
---
title: 'Discriminant analysis of principal components (DAPC)'
subtitle: '*NJ Gr&uuml;nwald, ZN Kamvar, and SE Everhart*'
---

# Introduction
```{r}
library(ggplot2)
library(SeqArray)
library(tidyverse)
library(SNPRelate)
```

## Code chunk 
```{r}
x <- c(1,2,3,4)
y <- c(50, 99, 21, 20)
plot(x,y)
```

# References

### 2021-11-04 index.Rmd
Changed format of site to Bootswatch

--- 2021-06-04 index.Rmd
@@
----
-layout: default
-title: "Panel Seminars"
-nav_order: 1
-editor_options: 
-  chunk_output_type: console
----
-
 ```{r setup, include=FALSE}
@@
 ```
+---
+title: "Introduction"
+---
 
-# About
+# BETA TEST OF SITE!
 
-The MarineOmics working group is part of the [RCN for Evolution in Changing Seas](https://rcn-ecs.github.io/), and was formed to promote reproducible and robust genomic research in marine science.
+The MarineOmics working group is part of the [RCN for Evolution in Changing Seas](https://rcn-ecs.github.io/), and was formed to promote **reproducible** and **robust** genomic research in nonmodel species, but especially marine and aquatic studies.
 Methods in next generation sequencing and bioinformatics advance at a pace that quickly render any static set of best practices as obsolete.
-To support robust 'omics research now and moving forward, the working group will disseminate guidelines through an easily-interfaced and dynamic website that will be deployed in August 2021.
-*Keep an eye on this URL for updates!*
+To support robust 'omics research now and moving forward, the working group helps disseminate guidelines through this dynamic, open-sourced website.
 
-# Panel Seminar Series
+**Rigorous**:  following best principles to identify and mitigate sources of error in data generation and analysis
 
-To facilitate discussions about practices and pitfalls in applying 'omics to non-model systems, we are hosting Q & A panel seminars with experts on topics ranging from RADseq, population genomics with WGS, and transcriptomics.
-Seminars are every two weeks, on **Fridays from 11:00-12:15 US Central Time**.
-3-5 expert panelists will discuss each topic, with questions coming from the working group and the audience.
-If you would like to submit a question for the panelists ahead of time, please send an email to rcnecs.marineomics[at]gmail.com .
+**Reproducible**: when others can reproduce the results with accessible data, code, and documentation
 
-Everyone is welcome to join! All seminars will be recorded and available on this website.
+# Guidelines for 'omics analyses
 
-**To receive a Zoom link, please fill out the registration form [here](
-https://auburn.zoom.us/meeting/register/tZ0ode2rpz4sG9Il_TuYexWfqd6HyjdQ58Zi).**
+On this website you will find a collection of guidelines, tutorials, recommended readings, and method evaluations for various 'omics analyses that are commonly used in studies of nonmodel systems (particularly marine species).
+Instead of providing rigid "best practices", we aim to encourage data exploration and rigorous quality control through "best practices".
 
-## Seminar Schedule
+Based on the expertise of the working group, we have developed content for these topics:
 
-| Date        | Topic       | Panelists     | Video     |
-| :---        | :---        | :---          | :---      |
-| May 7         | GBS/RADseq genotyping pipelines     | [Melanie LaCava](https://melanielacava.weebly.com/), [Alicia Mastretta-Yanes](http://mastrettayanes-lab.org/), [Isaac Overcast](https://isaacovercast.github.io/), [Jon Puritz](https://www.marineevoeco.com/)   | [View video](https://www.youtube.com/watch?v=C74GBESeIq4)     |
-| May 21       |  WGS genotyping (low and high coverage)     | [Jeffrey Good](http://www.thegoodlab.org/), [Mikhail "Misha" Matz](https://matzlab.weebly.com/), [Noah Reid](http://nreid.github.io/), [Tim Sackton](https://scholar.harvard.edu/tsackton), [Nina Overgaard Therkildsen](https://www.therkildsenlab.com/) | [View Video](https://www.youtube.com/watch?v=ObJ3OT3dwSw)    |
-| June 4       |  Population genomic analyses and filtering data     | [Zach Gompert](https://gompertlab.com/), [Morgan Kelly](http://www.morgankelly.biology.lsu.edu/), [Katie Lotterhos](https://sites.google.com/site/katielotterhos/home), [Jeff Ross-Ibarra](https://rilab.ucdavis.edu/) |     |
-| June 18       |  RNAseq (experimental design through differential expression)     | [Ana Conesa](http://conesalab.org/), [Sarah Davies](http://sites.bu.edu/davieslab/), [Martin Hölzer](https://hoelzer-lab.github.io/), [Joanna Kelley](https://labs.wsu.edu/genomes/), [Harold Pimentel](https://pimentellab.com/)     |     |
+## Topics {.tabset .tabset-fade .tabset-pills}
+### Population Genomics
+
+* [Choosing a population genomics sequencing approach](WGS_intro.html)  
+* [Whole Genome Resequencing](WGS_intro.html)  
+* [Reduced Representation Sequencing](RADseq.html)
+
+### Functional Genomics
+
+* Transcriptomics with RNA-Seq (coming soon!)  
+* Differential gene expression (coming soon!)
+
+## {-}
+
+## Panel Seminars 
+To facilitate discussions about practices and pitfalls in applying 'omics to non-model systems, we hosted [Q & A panel seminars](panels.html) with experts on topics ranging from RADseq, population genomics with WGS, and transcriptomics.
+Check them out on [YouTube](https://www.youtube.com/channel/UCosG6JfIj7k0DizfsD5ADfQ)!
 
 # Reproducibility in Marine Genomics  
-The working group is also currently hosting three undergraduate research interns with funding from the [RCN-ECS](https://rcn-ecs.github.io/).
-These interns are part of a sub-committee conducting a literature survey to characterize trends in data availability, code accessibility, and methods reproducibility in papers that use genomics to study adaptation in marine systems.
-The results of this survey will be shared in a peer-reviewed article, along with guidance to improve the state of reproducibility in the field.
+The working group is also interested in evaluating and improving reproducibility in studies that use genomics to study adaptation in marine systems.
+utilizing genomics in marine species, through a peer-reviewed literature survey of data availability, code accessibility, and methods reproducibility.
+In Spring 2021, three undergraduate research interns contributed to this literature survey, with funding from the [RCN-ECS](https://rcn-ecs.github.io/).
 
@@
 
-#### [Katherine Silliman, PhD](https://ksilliman.weebly.com/), Auburn University - kes0132[at]auburn.edu
-#### [Sam Bogan](http://www.hofmannlab.com/sam-bogan.html), UC-Santa Barbara -snbogan[at]ucsb.edu
-#### [Alan Downey-Wall](https://cos.northeastern.edu/people/alan-downey-wall/), Northeastern University 
-#### Chris Mantegna, University of Washington  
-#### [Danielle Davenport](https://www.researchgate.net/profile/Danielle-Davenport-2), University of Queensland  
-#### [Hanny Rivera, PhD](https://www2.whoi.edu/staff/hrivera/), Woods Hole Oceanographic Institute  
-#### Jecy Klinkam, College of William and Mary  
-#### [Joanna Griffiths, PhD](https://joannasgriffiths.wordpress.com/), UC-Davis
-#### [Joe McGirr, PhD](https://joemcgirr.github.io/), UC-Davis 
-#### Kaitlin Macaranas, CSU-Bakersfield 
-#### Megan Guidry, University of Rhode Island
-#### Natalie Ameral, University of Rhode Island & RI Dept. of Environmental Management  
-#### Nicolas Lou, Cornell University
-#### [Sara Schaal](https://schaals.wixsite.com/smschaal), Northeastern University 
-#### [Sara Wuitchik, PhD](https://sjswuitchik.weebly.com/), Harvard University
+#### [Katherine Silliman, PhD](https://ksilliman.weebly.com/), Auburn University - kes0132[at]auburn.edu {.unlisted .unnumbered}
+#### [Sam Bogan](http://www.hofmannlab.com/sam-bogan.html), UC-Santa Barbara -snbogan[at]ucsb.edu {.unlisted .unnumbered}
+#### [Alan Downey-Wall](https://cos.northeastern.edu/people/alan-downey-wall/), Northeastern University {.unlisted .unnumbered}
+#### Chris Mantegna, University of Washington  {.unlisted .unnumbered}
+#### [Danielle Davenport](https://www.researchgate.net/profile/Danielle-Davenport-2), University of Queensland {.unlisted .unnumbered}  
+#### [Hanny Rivera, PhD](https://www2.whoi.edu/staff/hrivera/), Woods Hole Oceanographic Institute  {.unlisted .unnumbered}
+#### Jecy Klinkam, College of William and Mary  {.unlisted .unnumbered}
+#### [Joanna Griffiths, PhD](https://joannasgriffiths.wordpress.com/), UC-Davis {.unlisted .unnumbered}
+#### [Joe McGirr, PhD](https://joemcgirr.github.io/), UC-Davis {.unlisted .unnumbered}
+#### Kaitlin Macaranas, CSU-Bakersfield {.unlisted .unnumbered}
+#### Megan Guidry, University of Rhode Island {.unlisted .unnumbered}
+#### Natalie Ameral, University of Rhode Island & RI Dept. of Environmental Management {.unlisted .unnumbered}  
+#### Nicolas Lou, Cornell University {.unlisted .unnumbered}
+#### [Sara Schaal](https://schaals.wixsite.com/smschaal), Northeastern University {.unlisted .unnumbered} 
+#### [Sara Wuitchik, PhD](https://sjswuitchik.weebly.com/), Harvard University {.unlisted .unnumbered}
 
-### Faculty Advisors
-#### [Andrew Whitehead, PhD](https://whiteheadresearch.wordpress.com/), UC-Davis 
-#### [Jon Puritz, PhD](https://www.marineevoeco.com/), University of Rhode Island 
-#### [Steven Roberts, PhD](http://faculty.washington.edu/sr320/?page_id=10850), University of Washington
+### Faculty Advisors {.unlisted .unnumbered}
+#### [Andrew Whitehead, PhD](https://whiteheadresearch.wordpress.com/), UC-Davis {.unlisted .unnumbered}
+#### [Jon Puritz, PhD](https://www.marineevoeco.com/), University of Rhode Island {.unlisted .unnumbered}
+#### [Steven Roberts, PhD](http://faculty.washington.edu/sr320/?page_id=10850), University of Washington {.unlisted .unnumbered}
 
@@
 For more information about the reproducibility review and undergraduate internship, contact Sam Bogan: snbogan[at]ucsb.edu
-
-``````{r, echo=FALSE,eval=FALSE}
-library(knitr)
-my.jekyll.site <- "marineomics.github.io"
-KnitPost <- function(input, base.url = my.jekyll.site) {
-  opts_knit$set(base.url = base.url)
-  fig.path <- paste0("images/", sub(".Rmd$", "", basename(input)), "/")
-  opts_chunk$set(fig.path = fig.path)
-  opts_chunk$set(fig.cap = "center")
-  render_jekyll()
-  knit(input, envir = parent.frame())
-}
-
-setwd("~/Postdoc_Projects/RCN_MarineOmics/Website/marineomics.github.io/")
-KnitPost("~/Postdoc_Projects/RCN_MarineOmics/Website/marineomics.github.io/index.Rmd")
-
-```

### 2021-11-04 popgen_intro.Rmd
Changed format of site to Bootswatch

---
bibliography: bibtexlib.bib
title: Choosing a Population Genomics Approach 
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
library(knitcitations)
library(kableExtra)
opts_chunk$set(fig.width = 10,
               fig.height = 5,
               cache = FALSE)
cite_options(citation_format = "pandoc", max.names = 3, style = "html", 
             hyperlink = "to.doc")
```

Deciding on a sequencing approach for a new project can be challenging, and depends primarily on the questions you are interested in, but also constraints set by funding, existing genomic resources, sample size, genome characteristics, and expertise in the research group. The main categories of population genomic sequencing approaches are:

1) High coverage individual whole genome resequencing (hcWGS)  
 i) Whole-genome resequencing (WGS) is where whole genomes of multiple individuals are sequenced and mapped to an assembled reference genome in order to compare variable sites between individuals or populations. Sequence coverage refers to the average number of sequencing reads per region of the genome. With more sequencing reads at a region (higher coverage), errors that arise during sequencing have a smaller impact on accurately identifying variable sites (termed "genotyping").  
2) Low coverage individual WGS (lcWGS)  
 i) 
3) pooled individuals (Poolseq)  
4) reduced-representation sequencing 
 i) can be split into restriction-based methods (eg RADseq) and targeted capture methods (eg exome capture)

Below we list a number of great review articles and empirical comparison studies that can help when making the decision. This topic was also discussed during the MarineOmics seminars (link).

References:

* Matz 2018[@zeeberg2004mistaken]: great intro to ecological genomics methods in species with few to no prior genomic resources, accessible to those just getting started.  
* Fuentes-Pardo and Ruzzante 2017: a VERY thorough review of the different approaches to generating population genomics data, with non-model species in mind. Some of the content about relative cost and available software is a little dated, but otherwise a very valuable reference.   
* Benjelloun et al 2019: empirical comparison of high coverage WGS, low coverage WGS, random variants (eg. from RADseq), exome capture, and commercial SNP chip for mammals (genome size=2.6Gb). For neutral genomic diversity, found 5k-10k random variants were enough to accurately estimate, but comercial panales/exome capture had acertainment bias. To detect selection and accurate estimates of LD, at least 1M variants were required. For the studied species, 5x coverage was just as accurate as 12x coverage. Low coverage WGS (< 5x) should use genotype likelihood methods, however the rate of false positives for heterozygotes is higher than classic genotyping methods at higher coverage.  
* Dorant et al. 2019: Case study comparing Poolseq vs. reduced representation seq (GBS) vs Rapture (targeted capture of RAD loci) for inferring population structure with weak genetic differentiation. All three detected same weak pop structure pattern, but Poolseq gave Fst values 3-5x higher than Rapture/GBS.

It should be noted that you don't need to choose only one approach for all samples! If you are starting a research program on a species without any genomic resources, you may want to start with de novo RADseq to understand basic population structure and genetic diversity, then once a reference genome is assembled move to high/low coverage WGS. In the [MarineOmics seminar](link), Misha Matz recommended choosing one individual to use for genome assembly, doing high coverage WGS on a few other individuals that represent most of the variation in your taxa, and then WGS many other individuals at low coverage. The high coverage individuals can be useful for imputing missing genotypes in the low coverage samples.

This table briefly summarizes the pros and cons of five sequencing approaches and their appropriateness for answering specific questions. If you are interested in answering multiple questions (eg, neutral population structure and adaptive variation) then you would often want a method that can do both (eg, WGS).

```{r,echo=FALSE}
text_tbl <- data.frame(
  Goals = c("Pros","Cons","Population structure, genetic diversity","Demography (mig. rates, pop. size through time)",
            "Signatures of selection, GWAS","Phylogenetic inference","Genetic crosses/mapping panels",
            "molecular evolution (accurate low-freq alleles required)"),
  hcWGS = c("provides high quality, high density genotypes; can be used to improve development of a reference genome","Most expensive per sample, requires a reference genome assembly, large computational resources","Great, but cheaper methods work almost as well","Great if incorporating haplotype-based methods ","Great","Computationally challenging, but good at all divergence levels ","Expensive","Best"),
  lcWGS = c("Provides high density genotypes at a reduced cost","Requires a reference, can produce false (+) heterozygotes, sensitive to batch effects, may be inappropriate to use individual SNP calls for some analyses","Good, but limited to popgen analyses based on genotype likelihoods ","Good, esp. with some high coverage samples to help with imputation","Good","Good for tree shapes but not branch lengths, good for extracting organelle sequences and cost-effective primers for species and hybrid ID studies ","Great","Poor"),
 RADseq = c("Generally cheaper per sample, density of genotypes can be tuned to fit question, does not require a reference","Only 1-5% of genome covered which limits studies on adaptive variation, de novo assembly can result in paralogs without quality control, can be hard to capture the same loci between batches","Great, esp. for many samples","Good for methods based on site frequency spectrum (eg moments), but not ideal for methods using extended haplotype or phasing info","Hotly debated, but unless your genome is way too big, WGS is the way to go","good at shallow to medium divergence levels, but need to play with filtering parameters ", "great as it allows for many invididuals","poor" ),
 Poolseq = c("many individuals can be mixed for a low library prep cost, only option for some larval studies","need higher sequencing depth so it can be more expensive than RAD (but less than other options), no info on individual genotypes (duh), requires a reference, requires some replicates to account for batch effects","good for methods that only use allele frequencies, won't work for individual based methods (eg Admixture)","active area of method development, but still not common except in cases of multiple temporal samples (cite) ","good if covering majority of the genome","would work in certain situations","maybe","poor"),
 Target.Capture = c("allows consistent capture of loci between batches, good for sequencing targeted regions across many samples","expensive to start a new project and requires reference of some kind to design baits (unless using method like eecSeq), only gives info on targeted region so can result in ascertainment bias for some analyses","ascertainment bias is likely to skew results, esp. for measures of genetic diversity","likely not appropriate due to ascertainment bias","only works for targeted regions ","an improvement over RAD as there will be less missing data among samples","poor","good if only interested in targeted regions")
  )

kbl(text_tbl) %>%
 # kable_paper(full_width = F) %>%
  column_spec(1, bold = T, border_right = T) %>%
  kable_styling(bootstrap_options = "striped", font_size = 13)

```

Presently, our site covers four approaches (WGS=whole genome (re)sequencing):

For each approach, we provide guiding principles and tutorials for bioinformatic processing from raw data to genotypes. While we currently don't discuss library preparation methods in depth, we do include advice on how to organize your sequencing runs in order to improve quality control downstream.

# References

### 2021-11-05 choosing-population-genetics.Rmd
Edited WGS intro

--- 2021-11-04 popgen_intro.Rmd
@@
 
-1) High coverage individual whole genome resequencing (hcWGS)  
- i) Whole-genome resequencing (WGS) is where whole genomes of multiple individuals are sequenced and mapped to an assembled reference genome in order to compare variable sites between individuals or populations.
-Sequence coverage refers to the average number of sequencing reads per region of the genome.
-With more sequencing reads at a region (higher coverage), errors that arise during sequencing have a smaller impact on accurately identifying variable sites (termed "genotyping").
-2) Low coverage individual WGS (lcWGS)  
- i) 
-3) pooled individuals (Poolseq)  
-4) reduced-representation sequencing 
- i) can be split into restriction-based methods (eg RADseq) and targeted capture methods (eg exome capture)
+### 1) High coverage individual whole genome resequencing (hcWGS)  {.unlisted .unnumbered}
+**Whole-genome resequencing (WGS)** is where whole genomes of multiple individuals are sequenced and mapped to an assembled reference genome in order to compare variable sites between individuals or populations.
+**Coverage depth** refers to the average number of sequencing reads per region of the genome.
+With more sequencing reads at a region (higher coverage depth), errors that arise during sequencing have a smaller impact on accurately identifying variable sites (termed "genotyping") (cite).
+
+### 2) Low coverage individual WGS (lcWGS)  {.unlisted .unnumbered}
+Low coverage whole genome sequencing is where whole genomes of many individuals are sequenced, but depth of coverage so low that individual genotypes cannot be confidently assigned (cite).
+Instead of assigning specific, "hard called genotypes, lcWGS data requires probabilistic analyses that account for uncertainty about the true genotype and can incororate uncertanty accross all sequenced samples.
+Typically, sequencing depth of > 3 reads is considered "low coverage".
+
+### 3) Reduced-representation sequencing {.unlisted .unnumbered}
+As the name suggests, this approach aims to sequence a reduced portion of the genome across multiple individuals at moderate to high coverage for variant discovery and genotyping with high confidence.
+The genome fragments can be selected at random, or through targeted probe assays.
+#### RADseq/Genotype-by-sequencing  {.unlisted .unnumbered}
+This very popular class of methods involve sequencing a subset of genomic regions randomly distributed throughout the genome, often using restriction enzymes (i.e., RADseq: restriction site-associated DNA).
+#### Targeted capture methods (e.g., exome capture)  {.unlisted .unnumbered}
+
+### 4) Pooled individuals (Poolseq)  {.unlisted .unnumbered}
+
+![Figure 1 from Lou et al. 2021 comparing the distribution of sequencing reads mapped to a reference genome.](images/mec16077-fig-0001-m.jpg)
 

### 2021-11-05 index.Rmd
Edited WGS intro

@@
 ---
-
-# BETA TEST OF SITE!
 
@@
 
-On this website you will find a collection of guidelines, tutorials, recommended readings, and method evaluations for various 'omics analyses that are commonly used in studies of nonmodel systems (particularly marine species).
-Instead of providing rigid "best practices", we aim to encourage data exploration and rigorous quality control through "best practices".
+On this website you will find a growing and evolving collection of guidelines, tutorials, recommended readings, and method evaluations for various 'omics analyses that are commonly used in studies of nonmodel systems (particularly marine species).
+Instead of providing rigid "best practices", we aim to encourage data exploration and rigorous quality control through "best principles".
+The guidelines are written by members of the MarineOmics working group or the community, then shared for peer review.
+Comments on current website content or suggestions for new content can be made through the [Discussion Forum](discussion.html).
 
-Based on the expertise of the working group, we have developed content for these topics:
+Based on the expertise of the MarineOmics working group, the first topics include:
 
@@
 
-## Panel Seminars 
+### The guidelines portion of the website is still under peer review and should be considered a "preprint".
+If you are interested in contributing to the peer review process, check out the [Discussion Forum](discussion.html).
+{.unlisted .unnumbered}
+
+# Panel Seminars 
 To facilitate discussions about practices and pitfalls in applying 'omics to non-model systems, we hosted [Q & A panel seminars](panels.html) with experts on topics ranging from RADseq, population genomics with WGS, and transcriptomics.

### 2021-11-05 choosing-population-genetics.Rmd
Added new reference bib file

@@
 ---
-bibliography: bibtexlib.bib
+bibliography: RCN_site-Nov05.bib
 title: Choosing a Population Genomics Approach 
@@
 **Coverage depth** refers to the average number of sequencing reads per region of the genome.
-With more sequencing reads at a region (higher coverage depth), errors that arise during sequencing have a smaller impact on accurately identifying variable sites (termed "genotyping") (cite).
+With more sequencing reads at a region (higher coverage depth), errors that arise during sequencing have a smaller impact on accurately identifying variable sites (termed "genotyping") [@Fuentes-Pardo2017-nu].
 
@@
 
-![Figure 1 from Lou et al. 2021 comparing the distribution of sequencing reads mapped to a reference genome.](images/mec16077-fig-0001-m.jpg)
+![Figure 1 from [@Lou2021-me] comparing the distribution of sequencing reads mapped to a reference genome.](images/mec16077-fig-0001-m.jpg)
 
@@
 
-References:
+Review articles:
 
-* Matz 2018[@zeeberg2004mistaken]: great intro to ecological genomics methods in species with few to no prior genomic resources, accessible to those just getting started.  
-* Fuentes-Pardo and Ruzzante 2017: a VERY thorough review of the different approaches to generating population genomics data, with non-model species in mind.
+* [@Matz2018-vl]: great intro to ecological genomics methods in species with few to no prior genomic resources, accessible to those just getting started.  
+* [@Fuentes-Pardo2017-nu]: a VERY thorough review of the different approaches to generating population genomics data, with non-model species in mind.
 Some of the content about relative cost and available software is a little dated, but otherwise a very valuable reference.
-* Benjelloun et al 2019: empirical comparison of high coverage WGS, low coverage WGS, random variants (eg.
+* [@Lou2021-me]: Great recent primer on low coverage WGS and how it compares to RADseq and Poolseq.
+
+Empiracal comparisions:
+
+* [@Benjelloun2019-dq]: empirical comparison of high coverage WGS, low coverage WGS, random variants (eg.
 from RADseq), exome capture, and commercial SNP chip for mammals (genome size=2.6Gb).
@@
 Low coverage WGS (< 5x) should use genotype likelihood methods, however the rate of false positives for heterozygotes is higher than classic genotyping methods at higher coverage.
-* Dorant et al.
-2019: Case study comparing Poolseq vs.
+* [@Dorant2019-fb]: Case study comparing Poolseq vs.
 reduced representation seq (GBS) vs Rapture (targeted capture of RAD loci) for inferring population structure with weak genetic differentiation.

### 2021-11-05 RADseq.Rmd
Added poolseq content

@@
 title: 'RADseq'
-subtitle: '*NJ Gr&uuml;nwald, ZN Kamvar, and SE Everhart*'
+subtitle: '*Katherine Silliman, Danielle Davenport*'
 ---

### 2021-11-05 choosing-population-genetics.Rmd
Added poolseq content

@@
 
+# Population genomic sequencing approaches
 Deciding on a sequencing approach for a new project can be challenging, and depends primarily on the questions you are interested in, but also constraints set by funding, existing genomic resources, sample size, genome characteristics, and expertise in the research group.
@@
 ![Figure 1 from [@Lou2021-me] comparing the distribution of sequencing reads mapped to a reference genome.](images/mec16077-fig-0001-m.jpg)
-
-Below we list a number of great review articles and empirical comparison studies that can help when making the decision. This topic was also discussed during the MarineOmics seminars (link).
-
-Review articles:
-
-* [@Matz2018-vl]: great intro to ecological genomics methods in species with few to no prior genomic resources, accessible to those just getting started.  
-* [@Fuentes-Pardo2017-nu]: a VERY thorough review of the different approaches to generating population genomics data, with non-model species in mind.
-Some of the content about relative cost and available software is a little dated, but otherwise a very valuable reference.
-* [@Lou2021-me]: Great recent primer on low coverage WGS and how it compares to RADseq and Poolseq.
-
-Empiracal comparisions:
-
-* [@Benjelloun2019-dq]: empirical comparison of high coverage WGS, low coverage WGS, random variants (eg.
-from RADseq), exome capture, and commercial SNP chip for mammals (genome size=2.6Gb).
-For neutral genomic diversity, found 5k-10k random variants were enough to accurately estimate, but comercial panales/exome capture had acertainment bias.
-To detect selection and accurate estimates of LD, at least 1M variants were required.
-For the studied species, 5x coverage was just as accurate as 12x coverage.
-Low coverage WGS (< 5x) should use genotype likelihood methods, however the rate of false positives for heterozygotes is higher than classic genotyping methods at higher coverage.
-* [@Dorant2019-fb]: Case study comparing Poolseq vs.
-reduced representation seq (GBS) vs Rapture (targeted capture of RAD loci) for inferring population structure with weak genetic differentiation.
-All three detected same weak pop structure pattern, but Poolseq gave Fst values 3-5x higher than Rapture/GBS.
 
@@
 
-Presently, our site covers four approaches (WGS=whole genome (re)sequencing):
+# Recommended readings  
+Below we list a number of great review articles and empirical comparison studies that can help when making the decision. This topic was also discussed during the MarineOmics seminars (link).
+
+Review articles:
+
+* [@Matz2018-vl]: great intro to ecological genomics methods in species with few to no prior genomic resources, accessible to those just getting started.  
+* [@Fuentes-Pardo2017-nu]: a VERY thorough review of the different approaches to generating population genomics data, with non-model species in mind.
+Some of the content about relative cost and available software is a little dated, but otherwise a very valuable reference.
+* [@Lou2021-me]: Great recent primer on low coverage WGS and how it compares to RADseq and Poolseq.
+
+Empirical comparisons:
+
+* [@Benjelloun2019-dq]: empirical comparison of high coverage WGS, low coverage WGS, random variants (eg.
+from RADseq), exome capture, and commercial SNP chip for mammals (genome size=2.6Gb).
+For neutral genomic diversity, found 5k-10k random variants were enough to accurately estimate, but comercial panales/exome capture had acertainment bias.
+To detect selection and accurate estimates of LD, at least 1M variants were required.
+For the studied species, 5x coverage was just as accurate as 12x coverage.
+Low coverage WGS (< 5x) should use genotype likelihood methods, however the rate of false positives for heterozygotes is higher than classic genotyping methods at higher coverage.
+* [@Dorant2019-fb]: Case study comparing Poolseq vs.
+reduced representation seq (GBS) vs Rapture (targeted capture of RAD loci) for inferring population structure with weak genetic differentiation.
+All three detected same weak pop structure pattern, but Poolseq gave Fst values 3-5x higher than Rapture/GBS.
+
+# MarineOmics Guidelines
+Presently, our site covers four approaches:
+
+* [RADseq](RADseq.html)  
+* [hcWGS](WGS_intro.html)
+* [lcWGS, maintained by MarineOmics member Nicolas Lou](https://github.com/therkildsen-lab/genomic-data-analysis/blob/master/lcwgs_data_analysis.md)
+* [Poolseq](poolseq.html)
 

### 2021-11-05 WGS_intro.Rmd
Added WGS tutorials

@@
 ---
-bibliography: bibtexlib.bib
+bibliography: RCN_site-Nov05.bib
 ---
@@
 This is separate from whole-genome de novo sequencing, which aims to produce a reference genome by sequencing and assembling a complete genome for the first time.
-WGS can be divided into three main approaches, and choosing between these or RADseq(link to ) mostly depends on your questions of interest.   
-1) high coverage individual WGS (hcWGS)  
-+ description  
-2) low coverage individual WGS (lcWGS)  
-+ description
-3) pooled individuals (Poolseq)  
-RADseq is... and m
 
- also a factor of your funds and how you choose to allocate them.
-There is a financial trade-off between the number of samples you can genotype individually and the amount of sequencing coverage per sample.
-If you have 1000s of samples, it may not be financially feasible to sequence most of them to high depth.
-Large genome sizes can also make hcWGS cost-prohibited.
-This topic was discussed during the MarineOmics WGS panel seminar (link).
-An excellent and thorough review of the pros  between these approaches, esp. as they apply to nonmodel organisms, is [Fuentes-Pardo and Ruzzante 2017)(link).
+An excellent and thorough review of the pros  between these approaches, esp. as they apply to nonmodel organisms, is [@Fuentes-Pardo2017-nu].
 
@@
 
-* Fastq-to-VCF pipeline(link): In-depth explanation of automated short-read mapping and variant calling pipeline maintained by [Harvard Informatics](https://github.com/harvardinformatics/shortRead_mapping_variantCalling).
+* [Fastq-to-VCF SnakeMate pipeline](fastq2vcf_pipeline.html): In-depth explanation of automated short-read mapping and variant calling pipeline maintained by [Harvard Informatics](https://github.com/harvardinformatics/shortRead_mapping_variantCalling).
 Useful for moderate to high coverage data, where *add description of why one would use pipeline*
-*
+* [Fastq-to-VCF workflow](WGS-Full-Walkthrough.html): detailed walkthrough of processing 15x sequencing depth WGS data for cod, from raw reads to a VCF.
 
-# Considerations Before Lab Work
+# References

### 2021-11-05 fastq2vcf_pipeline.Rmd
Added WGS tutorials

---
title: "FASTQ -> VCF workflow for WGS data"
author: "Sara JS Wuitchik"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# fastq2bam
This is a more fleshed out explanation of the pipeline developed and maintained by [Harvard Informatics](https://github.com/harvardinformatics) in the [shortRead_mapping_variantCalling repo](https://github.com/harvardinformatics/shortRead_mapping_variantCalling)

### Trimming adapters

Before aligning your sequences to a reference genome, you will need to trim off the adapters. We use FastP in our pipeline as below, but there are other programs that could be suitable (e.g., Trimmomatic, TrimGalore/Cutadapt, BBDuk).

```{bash, eval = F}
fastp --in1 {r1} --in2 {r2} --out1 {r1} --out2 {r2} --detect_adapter_for_pe 2> {summary.out}
```

Where <br>
```--in``` and ```--out``` are the flags for the input and output reads <br>
```--detect_adapter_for_pe``` enables auto-detection of adapters for paired end data (because auto-detection is only enabled for single end data by default).

FastP (and other trimming software) can be used for quality control before aligning your sequences to the reference genome, but there are arguments on both sides for whether or not trimming your reads at this stage is useful (add references in here).

### Mapping

Our pipeline uses the Burrows-Wheeler transform (Li and Durbin 2010) to align reads to the reference genome, but there are many software that can be used for aligning short reads to a reference (e.g., Bowtie2, STAR). To make the alignment process more efficient, you need to index the reference genome.

```{bash, eval = F}
bwa index {reference.fa}
```

Where ```reference.fa``` is your reference genome. Once your reference has been indexed, you can do a local alignment of your FASTQs, and pipe directly into ```samtools``` to sort the alignment by coordinate and converted to the BAM (Binary ALignment/Map) format.<br>

```{bash, eval = F}
bwa mem -M -R {string} {ref} {r1} {r2} | samtools sort -o {output.bam}
```

Where <br>
```-M``` marks shorter split hits as secondary (required for Picard compatibility downstream) <br>
```-R``` defines a read group header e.g., @RG\\tID:{sample}\\tSM:{sample}\\tPL:ILLUMINA <br>
```{ref}``` is the indexed reference genome <br>
```{r1}``` and ```{r2}``` are the first and second reads you are aligning <br>
```-o``` is the output file name

Once you have a sorted and indexed alignment, you can locate and tag duplicate records, which are reads that have originated from a single fragment of DNA. This can happen during sample preparation (e.g., during PCR) or from sequencing artifacts (e.g., optical duplicates).If you have multiple runs for the same sample, you would want to merge and index the BAMs before the deduplication tagging, which you can do with ```samtools merge```.

```{bash, eval = F}
picard MarkDuplicates I={input.bam} O={output.bam} METRICS_FILE={output} REMOVE_DUPLICATES=false TAGGING_POLICY=All
```

Where <br>
```I``` is the input BAM that has been coordinate sorted (the output of ```samtools sort```) <br>
```O``` is the output BAM file <br>
```METRICS_FILE``` writes the deduplication metrics to the output file name (this is a required argument) <br>
```REMOVE_DUPLICATES``` writes the apropriate flags to the record rather than removing the reads without telling you what they are (```false``` is the default setting) <br>
```TAGGING_POLICY``` will set the duplicates to be recorded in the DT attribute when the parameter is set to ```All```

Once you have tagged the duplicated records in your BAM, you will want to generate a .bai index file

```{bash, eval = F}
picard BuildBamIndex I={input.bam}
```

Where <br>
```I``` is the input dedup'd BAM (the output of ```picard MarkDuplicates```)

### Quality control and filtering

It's important to take a look at your alignments to check for quality, potentially filter out sites or samples that don't meet adequate thresholds, etc. The pipeline outputs various statistics automatically along the way, but we have collected some of the QC commands here. <br>

We use ```samtools``` to compute the depth at each position or region, with the tabulated output containing information for the chromosome, start position, end position, number of reads aligned in that region after filtering, number of covered bases with a depth greater than or equal to 1 in that region, the proportion of coverage bases, the mean depth of coverage, mean baseQ, and mean mapQ.
```{bash, eval = F}
samtools coverage --output {output} {input.bam}
```

Where <br>
```--output``` is the file name you want to write the output to <br>
```input.bam``` is the input BAM file <br><br>

To produce a summary of the alignment metrics, including the quality of read alignments and the quality filters, we use ```Picard```
```{bash, eval = F}
picard ValidateSamFile I={input.bam} R={reference.fa} O={output}
```

Where <br>
```I``` is the input BAM file <br> 
```R``` is the reference genome <br>
```O``` is the file name you want to write the output to <br><br>

To ensure the alignments meet file format specifications:
```{bash, eval = F}
picard ValidateSamFile I={input.bam} R={reference.fa} O={output} IGNORE=VALID_TAG_NM || true
```

# bam2vcf
### Variant Calling

Two of the major variant callers you can use (at the time of writing) are Genome Analysis Toolkit (GATK) and freebayes. You can call variants with either of these software in our pipeline and the downstream steps are the same, so here we present parameters for variant calling with both GATK and freebayes. Regardless of the software used, variant calling is computationally intensive but can be parallelized for more efficient resource use by splitting the reference genome into intervals. The interval creation is automated in our pipeline but parameter guidelines will differ depending on your reference genome. <br><br>

To call germline single nucleotide polymorphisms (SNPs) and insertion/deletions (indels) via local re-assembly of haplotypes using GATK: 
```{bash, eval = F}
gatk HaplotypeCaller -R {reference.fa} -I {input.bam} -O {output.gvcf} -L {interval.list} --emit-ref-confidence GVCF --min-pruning 2 --min-dangling-branch-length 4
```

Where <br>
```-R``` is the reference genome <br>
```-I``` is the input BAM <br>
```-O``` is the output gVCF <br>
```-L``` is the interval file <br>
```--emit-ref-confidence``` yields the reference confidence scores as gVCF <br>
```--min-pruning``` sets the minimum support to not prune paths in the graph <br>
```--min-dangling-branch-length``` sets the minimum lenght of a dangling branch to attempt recovery <br><br>

A note about coverage: when calling variants where the coverage is less than 10x, we suggest setting ```--min-pruning``` and ```--min-dangling-branch-length``` each to 1. The parameters in the code above are more appropriate for data where the coverage exceeds 10x. <br>

There are some constraints when using clusters and job scheduling managers, including command lengths. When you have many samples and are trying to run GATK on all of them, the command may get too long, causing SLURM (or another job scheduler) to throw an error. To get around this, you can create DB map files (which is done automatically in our pipeline). Then we can import many gVCFs into a GenomicsDB for a given scaffold:
```{bash, eval = F}
export TILEDB_DISABLE_FILE_LOCKING=1
gatk GenomicsDBImport --genomicsdb-workspace-path {output.db} -L --sample-name-map {input.db}
```

Where <br>
the ```export``` command can remedy sluggish performance <br> 
```--genomicsdb-workspace-path``` is the working directory to point to for GenomicsDB <br>
```-L``` is the interval file <br>
```--sample-name-map``` is the path to a file that contains a sample to file map in a tab delimited format <br><br>

Now you can use the genomic databases from ```gatk GenomicsDBImport``` to create VCF files (one per interval file)
```{bash, eval = F}
gatk GenotypeGVCFs -R {reference.fa} -V {input.DB} -O {output.vcf}
```

Where <br>
```-R``` is the reference genome <br>
```-V``` is the DB map file <br>
```-O``` is the output VCF <br><br>

If you want to use freebayes the variant caller, rather than starting with ```gatk HaplotypeCaller```, you would instead use:
```{bash, eval = F}
freebayes -f {reference.fa} -r {interval} -g {maxDP} {inputs.bam} > {output.vcf}
```

Where <br>
```-f``` is the reference genome (creates a .fai index if one doesn't already exist) <br>
```-r``` limits the analysis to a specific interval (region) <br>
```-g``` skips processing of alignments overlapping positions with coverage greater than the value; we use the maxDP for this parameter <br> 
```{inputs.bam}``` are the input BAMs; you can input all BAMs at once but you need to set the output as one VCF per interval <br>
```{output.vcf}``` the output VCF for each interval e.g., {i}.vcf per interval <br><br>

Regardless of the variant caller you use, before combining all the VCFs for the intervals into one final VCF, it is computationally more efficient to filter each of the VCFs, then gather them. 
```{bash, eval = F}
gatk VariantFiltration -R {reference.fa} -V {input.vcf} --output {output.vcf} \
--filter-name "RPRS_filter" \
--filter-expression "(vc.isSNP() && (vc.hasAttribute('ReadPosRankSum') && ReadPosRankSum < -8.0)) || ((vc.isIndel() || vc.isMixed()) && (vc.hasAttribute('ReadPosRankSum') && ReadPosRankSum < -20.0)) || (vc.hasAttribute('QD') && QD < 2.0)" \
--filter-name "FS_SOR_filter" \
--filter-expression "(vc.isSNP() && ((vc.hasAttribute('FS') && FS > 60.0) || (vc.hasAttribute('SOR') &&  SOR > 3.0))) || ((vc.isIndel() || vc.isMixed()) && ((vc.hasAttribute('FS') && FS > 200.0) || (vc.hasAttribute('SOR') &&  SOR > 10.0)))" \
--filter-name "MQ_filter" \
--filter-expression "vc.isSNP() && ((vc.hasAttribute('MQ') && MQ < 40.0) || (vc.hasAttribute('MQRankSum') && MQRankSum < -12.5))" \
--filter-name "QUAL_filter" \
--filter-expression "QUAL < 30.0" \
--invalidate-previous-filters

gatk GatherVcfs {input.vcfs} -O {output.vcf}
```

Where <br>
```-R``` is the reference genome <br>
```-V``` is the input VCF (per interval)
```--output``` is the filtered VCF (per interval)
<br>
```{input.vcfs}``` is a custom script invoked by the pipeline to gather all the filtered VCFs for each interval into a single input command using the intervals list <br>
```-O``` is the final output VCF
<br>

The filter expressions more or less follow the GATK hard filtering recommendations but are split to treat SNPs and indels with different thresholds within each filter expression, as well as keeping different quality tags separate to make downstream filtering easier.

<br>

#### Quality control and statistics <br>
NB: this section will likely be better suited to be included in a QC section (and it's not fully fleshed out itself yet), but here are two of the pipeline outputs that can be used for downstream QC. <br>

```{bash, eval = F}
vcftools --gzvcf {input.vcf} --remove-filtered-all --minDP 1 --missing-indv --stdout > {output}
```
Where <br>
```--gzvcf``` is the input VCF, gzipped (use ```--vcf``` if working with an uncompressed VCF) <br>
```--remove-filtered-all``` removes all the sites with a FILTER flag other than PASS <br>
```--minDP``` includes only the genotypes greater than or equal to the parameter value (this requires the DP FORMAT tag to be specified) <br>
```--missing-indv``` generates a file reporting the missingness per indivdual <br>
```--stdout``` redirects the output to stdout so it can be redirected to a file name of your choosing (in this code, the file name would be where ```{output}``` is)
<br>
<br>
```{bash, eval = F}
bedtools intersect -a {input.intervals} -b {input.vcf} -c > {output}
```
Where <br>
```-a``` is the intervals BED produced by the pipeline automatically for use in freebayes <br>
```-b``` is the final VCF file output by ```gatk GatherVcfs``` <br>
```-c``` reports the number of hits in ```-b``` for each entry in ```-a``` <br>
```{output}``` is the file name of the output, which is a text file that reports the number of SNPs per interval

### 2021-11-05 WGS-Full-Walkthrough.Rmd
Added WGS tutorials
Added WGS image

--- 2021-11-04 WGSFullWalkthrough.Rmd
@@
 ---
-layout: default
 title: WGS fastq2vcf walkthrough
-parent: Whole Genome Resequencing
-grand_parent: Choosing a PopGenomics Approach
-editor_options: 
-  chunk_output_type: console
+
 ---
@@
 ```
+![Output](images/Pop1_17291100XcoveragePlot10000.png)
+
 <!-- <img src="./Pop1_17291100XcoveragePlot10000.png" width="500">  -->

### 2021-11-05 index.Rmd

@@
 
-* [Choosing a population genomics sequencing approach](WGS_intro.html)  
+* [Choosing a population genomics sequencing approach](choosing-population-genetics.html)  
 * [Whole Genome Resequencing](WGS_intro.html)  
 * [Reduced Representation Sequencing](RADseq.html) 
+* [Poolseq](poolseq.html)
 

### 2021-11-06 index.Rmd

--- 2021-11-05 index.Rmd
@@
 The guidelines are written by members of the MarineOmics working group or the community, then shared for peer review.
-Comments on current website content or suggestions for new content can be made through the [Discussion Forum](discussion.html).
+Comments on current website content or suggestions for new content can be made through the [Discussion Forum](https://github.com/MarineOmics/marineomics.github.io/discussions).
 
@@
 
-* [Choosing a population genomics sequencing approach](WGS_intro.html)  
-* [Whole Genome Resequencing](WGS_intro.html)  
+* [Choosing a population genomics sequencing approach](choosing-population-genetics.html)  
+* [Whole Genome Resequencing (WGS)](WGS_intro.html)  
+  * [Low coverage WGS](https://github.com/therkildsen-lab/genomic-data-analysis/blob/master/lcwgs_data_analysis.md)
 * [Reduced Representation Sequencing](RADseq.html) 
+* [Poolseq](poolseq.html)
 

### 2021-11-08 WGS_intro.Rmd

@@
 bibliography: RCN_site-Nov05.bib
+title:  Whole Genome Resequencing for Population Genomics 
 ---
@@
 
-# Whole Genome Resequencing for Population Genomics  
 For the purposes of this guide, we use "whole genome (re)sequencing (WGS)) to refer to  methods where a reference genome exists (whether for the focal species or a related species), and samples are sequenced and then mapped to most or all of the reference genome.
-This method can provide high density SNPs, structural information...for population genomic analyses.
-This is separate from whole-genome de novo sequencing, which aims to produce a reference genome by sequencing and assembling a complete genome for the first time.
+This method can provide high density genetic variants (e.g., SNPs) and structural information for population genomic analyses, while also facilitating functional insights if the reference genome is functionally annotated.
+This is separate from whole-genome *de novo* sequencing, which aims to produce a reference genome by sequencing and assembling a complete genome for the first time.
 
-An excellent and thorough review of the pros  between these approaches, esp. as they apply to nonmodel organisms, is [@Fuentes-Pardo2017-nu].
+An excellent and thorough review of the methodology and consideration for WGS approaches, especially as they apply to nonmodel organisms, is [@Fuentes-Pardo2017-nu].
 
@@
 Here we provide a couple different tutorials by our working group members, as well as links to other detailed tutorials on the web.
-The goal of the MarineOmics tutorials are to provide extensive detail on "why" certain parameters are chosen, and some guidance on how to evaluate different parameter options to fit your data.
+The goal of these MarineOmics tutorials are to provide extensive detail on "why" certain parameters are chosen, and some guidance on how to evaluate different parameter options to fit your data.
+
+### High to moderate coverage WGS
 
@@
 
+### Low coverage WGS  
+Check out [@Lou2021-me](https://onlinelibrary.wiley.com/doi/abs/10.1111/mec.16077) and the associated [Github repository](https://github.com/therkildsen-lab/genomic-data-analysis/blob/master/lcwgs_data_analysis.md) for a great primer on lcWGS analysis.
+
 # References

### 2021-11-08 poolseq.Rmd

---
title: "PoPoolation Pipeline"
subtitle: '*Joanna Griffiths*'
date: "8/12/2021"
#output:
#  html_document: default
#  pdf_document: default
#  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

I mostly followed the PoPoolation and PoPoolation2 pipelines by Robert Kofler and Christian Schlötterer (https://sourceforge.net/p/popoolation2/wiki/Main/). The website has an easy to follow tutorial for using PoPoolation2 and how to use their scripts. This pipeline extends the use of the scripts for running on a Linux cluster. I highly recommend following the PoPoolation2 tutorial and manual and using these scripts if you have any modifications from default flags.
*Note: These scripts were written for use on Louisiana State University’s high performance cluster “SuperMike-II” running Red Hat Enterprise Linux 6 (more info found here: http://www.hpc.lsu.edu/resources/hpc/system.php?system=SuperMike-II)

The following scripts were used to analyze data from an experimental evolution study on Tigriopus californicus copepods (Griffiths JS, Kawji Y, Kelly MW. 2020. An experimental test of adaptive introgression in locally adapted populations of splash pool copepods. Mol Biol and Evol. 38(4):1306-1316. https://doi.org/10.1093/molbev/msaa289). Raw sequencing data are deposited in NCBI's Short Reads Archive (https://www.ncbi.nlm.nih.gov/bioproject/PRJNA597336/). To download this dataset, go to: https://trace.ncbi.nlm.nih.gov/Traces/sra/sra.cgi?view=run_browser , then type in the accession number for the identified files you want to download (e.g., the accession no. for the file Bodega population file is SRR10760095, identified under the "Run" heading). Then click on the Data access tab and there will be a downloadable link. Additional analyses and scripts can be found at: https://github.com/JoannaGriffiths/Tigriopus_HER

***Strengths and weaknesses of Pool-seq***
Due to the nature of Pool-seq where multiple individuals are pooled to prepare a single library, this results in a loss of haplotype information. Therefore, analyses rely on allele frequency estimates rather than hard “genotype calls”. PoPoolation2 is primarily used for comparing allele frequency differences between populations or treatments in comparison to a reference genome. The pooling of many individuals can be a cost-effective tool to character genetic variation present at the population level (e.g., genome wide association studies and experimental evolution). For example, to identify small changes in allele frequencies, you would neet genotype hundreds of individuals across the entire genome, which may be too cost-prohibitive with individual barcoding. However, individual genotype information is lost, as well as information on dominance and
effect sizes. For example, individual genotypes cannot be linked to individual phenotypes. Additionally, estimations of linkage disequilibrium are limited to a single sequencing read since that is the known limit of haplotype information for a single individual. Finally, distinguishing between sequencing errors and rare, low-frequency alleles is limited. Unlike sequencing of individuals, this cannot be solved by analysing multiple reads from the same region of a single chromosome. One way to ameliorate this issue is to sequencing replicates of pools. In addition, Pool-seq analyses can be prone to inaccurate estimations of allele frequencies based on the experimental methods. Care must be taken to ensure the the experimentla design maximizes the ratio sequencing coverage to the number of individuals pooled.

***Notes on Experimental Design for Downstream Analyses:***
It is important to try to standardize the number of individuals (and thus DNA) pooled for each sample. Robert Kofler and team also recommend that the pool size should be larger than the goal coverage for sequencing, because this minimizes re-sampling the same allele from a single individual several times. Because PoPoolation is designed for comparing differences in allele frequencies among treatments/population, accuracy will depend on the pool size and sequencing coverage. For example, smaller differences in allele frequency changes can be seen if you have a large number of individuals in a single pool and more coverage. The PoPoolation2 Manual page provides the following calculation to detect significant allele frequency differences: "To detect a singificant (Fisher's exact test; p=0.01) allele frequency difference between two populations of 30%, a coverage (and pool size) of 50 will be sufficient. If however significant (Fisher's exact test; p=0.01) allele frequency differences of 10% need to be detected the coverage as well as the pool size should be about 400."
The sequencing accuracy of Pool-seq increases with larger pool sizes; if the number of sequenced individuals is approximately equal to the coverage, then the sequency accuracy is no longer superior to an individual level of sequencing (see Fig. 1a from Schlötterer et al., 2014).

Reference: Schlötterer C, Tobler R, Kofler R, Nolte V. 2014 Sequencing pools of individuals-mining genome-wide polymorphism data without big funding. Nat. Rev. Genet. 15, 749–763. (doi:10.1038/nrg3803)

***Alternative Pipelines:***
Low Coverage Pool-seq Data: 
https://github.com/petrov-lab/HAFpipe-line
Alternative pipelines have been created for low coverage (<5x) pool-seq data that can provide accurate estimates of allele frequencies. This pipeline is designed for Evolve and Re-sequence studies where the founder haplotypes are known.

Customizeable Filtering and Assessing Quality of Pool-Seq Data:
https://github.com/ToBoDev/assessPool
This pipeline filters SNPs based on adjustable criterion with suggestions for pooled data. It determines pool number and prepares proper data structure for analysis, creates a customizable run script for Popoolation2 for all pairwise comparisons, runs Popoolation2 and poolfstat, imports Popoolation2 and poolfstat output, and finally, generates population genetic statistics and plots for data visualization.

## Filter Raw Sequences
This script uses the wrapper TrimGalore to remove Adapter/Barcode sequences and removes poor quality reads. 
This example dataset uses paired end sequencing so I used the flag (--paired) in TrimGalore. Because these example files had high coverage, I was able to use conservative filters. 
For example, the default length of reads is 20 bp (so if you read is less than 20 bp then the read will be removed). Since I had high coverage and long read lengths (150 bp), I used a minimum read length of 40 bp. 
The --max_n flag specificies how many "Ns" can be in the read before the read if discarded as "poor quality". Again, depending on your sequencing depth you may want to be more or less conservative. There is no default value specified in TrimGalore, but I used a highly conservative value of 1 (ie any instances of N in the read and that read will be discarded). This is probably more conservative than you need to be, becasue the read may still contain infromative sequencign data.
Please note that these parameters are not specific to a Pool Seq dataset.
```{bash eval = FALSE}
#!/bin/bash

#PBS -q checkpt
#PBS -A hpc_kelly_19_3
#PBS -l nodes=1:ppn=16
#PBS -l walltime=72:00:00
#PBS -o /work/jgrif61/Tigs/output_files
#PBS -j oe
#PBS -M jgrif61@lsu.edu
#PBS -N trimgalore_Tigs

date

#Go to the directory where the raw sequencing files are located
cd /work/jgrif61/Tigs/raw_data/raw

trim_galore --paired --length 40 --max_n 1 Tig-1S_S6_L008_R1_001.fastq.gz Tig-1S_S6_L008_R2_001.fastq.gz -o /work/jgrif61/Tigs/raw_data/raw
trim_galore --paired --length 40 --max_n 1 Tig-1U_S3_L001_R1_001.fastq.gz Tig-1U_S3_L001_R2_001.fastq.gz -o /work/jgrif61/Tigs/raw_data/raw
trim_galore --paired --length 40 --max_n 1 Tig-2S_S7_L008_R1_001.fastq.gz  Tig-2S_S7_L008_R2_001.fastq.gz  -o /work/jgrif61/Tigs/raw_data/raw
trim_galore --paired --length 40 --max_n 1 Tig-2U_S9_L008_R1_001.fastq.gz Tig-2U_S9_L008_R2_001.fastq.gz -o /work/jgrif61/Tigs/raw_data/raw
trim_galore --paired --length 40 --max_n 1 Tig-3S_S1_L001_R1_001.fastq.gz Tig-3S_S1_L001_R2_001.fastq.gz -o /work/jgrif61/Tigs/raw_data/raw
trim_galore --paired --length 40 --max_n 1 Tig-4S_S2_L001_R1_001.fastq.gz Tig-4S_S2_L001_R2_001.fastq.gz -o /work/jgrif61/Tigs/raw_data/raw
trim_galore --paired --length 40 --max_n 1 Tig-4U_S10_L008_R1_001.fastq.gz Tig-4U_S10_L008_R2_001.fastq.gz -o /work/jgrif61/Tigs/raw_data/raw
trim_galore --paired --length 40 --max_n 1 Tig-5S_S8_L008_R1_001.fastq.gz Tig-5S_S8_L008_R2_001.fastq.gz -o /work/jgrif61/Tigs/raw_data/raw
trim_galore --paired --length 40 --max_n 1 Tig-5U_S4_L001_R1_001.fastq.gz Tig-5U_S4_L001_R2_001.fastq.gz -o /work/jgrif61/Tigs/raw_data/raw
trim_galore --paired --length 40 --max_n 1 Tig-6U_S11_L008_R1_001.fastq.gz Tig-6U_S11_L008_R2_001.fastq.gz -o /work/jgrif61/Tigs/raw_data/raw
trim_galore --paired --length 40 --max_n 1 Tig-BR_S5_L001_R1_001.fastq.gz Tig-BR_S5_L001_R2_001.fastq.gz -o /work/jgrif61/Tigs/raw_data/raw
trim_galore --paired --length 40 --max_n 1 Tig-SD_S12_L008_R1_001.fastq.gz Tig-SD_S12_L008_R2_001.fastq.gz -o /work/jgrif61/Tigs/raw_data/raw
date
exit 0
```

## Map Reads to Reference Genome 
```{bash eval = FALSE}
#!/bin/bash

#PBS -q workq
#PBS -A hpc_kelly_19_3
#PBS -l nodes=1:ppn=16
#PBS -l walltime=72:00:00
#PBS -o /work/jgrif61/Tigs/output_files
#PBS -j oe
#PBS -M jgrif61@lsu.edu
#PBS -N bowtie_tigs_BRSD

cd /work/jgrif61/Tigs/raw_data/raw

#prepare and index the reference genome so you can map reads to it
bowtie2-build reference/full_genome_mito_tigs.fasta reference/reference_index

bowtie2 -x ../reference/reference_index -1 Tig-1S_S6_L008_R1_001_val_1.fq.gz -2 Tig-1S_S6_L008_R2_001_val_2.fq.gz -S 1S_out_PE.sam

bowtie2 -x ../reference/reference_index -1 Tig-1U_S3_L001_R1_001_val_1.fq.gz -2 Tig-1U_S3_L001_R2_001_val_2.fq.gz -S 1U_out_PE.sam

bowtie2 -x ../reference/reference_index -1 Tig-2S_S7_L008_R1_001_val_1.fq.gz -2 Tig-2S_S7_L008_R2_001_val_2.fq.gz -S 2S_out_PE.sam

bowtie2 -x ../reference/reference_index -1 Tig-2U_S9_L008_R1_001_val_1.fq.gz -2 Tig-2U_S9_L008_R2_001_val_2.fq.gz -S 2U_out_PE.sam

bowtie2 -x ../reference/reference_index -1 Tig-3S_S1_L001_R1_001_val_1.fq.gz -2 Tig-3S_S1_L001_R2_001_val_2.fq.gz -S 3S_out_PE.sam

bowtie2 -x ../reference/reference_index -1 Tig-4S_S2_L001_R1_001_val_1.fq.gz -2 Tig-4S_S2_L001_R2_001_val_2.fq.gz -S 4S_out_PE.sam

bowtie2 -x ../reference/reference_index -1 Tig-4U_S10_L008_R1_001_val_1.fq.gz -2 Tig-4U_S10_L008_R2_001_val_2.fq.gz -S 4U_out_PE.sam

bowtie2 -x ../reference/reference_index -1 Tig-5S_S8_L008_R1_001_val_1.fq.gz -2 Tig-5S_S8_L008_R2_001_val_2.fq.gz -S 5S_out_PE.sam

bowtie2 -x ../reference/reference_index -1 Tig-5U_S4_L001_R1_001_val_1.fq.gz -2 Tig-5U_S4_L001_R2_001_val_2.fq.gz -S 5U_out_PE.sam

bowtie2 -x ../reference/reference_index -1 Tig-6U_S11_L008_R1_001_val_1.fq.gz -2 Tig-6U_S11_L008_R2_001_val_2.fq.gz -S 6U_out_PE.sam

bowtie2 -x ../reference/reference_index -1 Tig-BR_S5_L001_R1_001_val_1.fq.gz -2 Tig-BR_S5_L001_R2_001_val_2.fq.gz -S BR_out_PE.sam

bowtie2 -x ../reference/reference_index -1 Tig-SD_S12_L008_R1_001_val_1.fq.gz -2 Tig-SD_S12_L008_R2_001_val_2.fq.gz -S SD_out_PE.sam

bowtie2 -x ../reference/reference_index -1 Tig-allU_R1_001_val1_new.fq -2 Tig-allU_R2_001_val2_new.fq -S allU_out_PE.sam

#turn mapped sam files into bam file and then sort and remove ambiguously mapped reads
samtools view -bS 1S_out_PE.sam | samtools sort -o 1S_out.sorted
samtools view -bS 1U_out_PE.sam | samtools sort -o 1U_out.sorted
samtools view -bS 2S_out_PE.sam | samtools sort -o 2S_out.sorted
samtools view -bS 2U_out_PE.sam | samtools sort -o 2U_out.sorted
samtools view -bS 3S_out_PE.sam | samtools sort -o 3S_out.sorted
samtools view -bS 4S_out_PE.sam | samtools sort -o 4S_out.sorted
samtools view -bS 4U_out_PE.sam | samtools sort -o 4U_out.sorted
samtools view -bS 5S_out_PE.sam | samtools sort -o 5S_out.sorted
samtools view -bS 5U_out_PE.sam | samtools sort -o 5U_out.sorted
samtools view -bS 6U_out_PE.sam | samtools sort -o 6U_out.sorted
samtools view -bS BR_out_PE.sam | samtools sort -o BR_out.sorted
samtools view -bS SD_out_PE.sam | samtools sort -o SD_out.sorted
samtools view -bS allU_out_PE.sam | samtools sort -o allU_out.sorted

date 
exit
```

## Determine mean coverage across the genome
This code can be used to determine the mean coverage across the genome for each sample/library. This information can be used to inform the min and max coverage used in downstream analyses (such as the Fisher's exact test and Fst analyses). You will want to choose a min and max coverage window spanning the mean coverage.
```{bash eval = FALSE}
#!/bin/bash

#PBS -q workq
#PBS -A hpc_kelly_19_3
#PBS -l nodes=1:ppn=16
#PBS -l walltime=24:00:00
#PBS -o /work/jgrif61/Tigs/output_files
#PBS -j oe
#PBS -M jgrif61@lsu.edu
#PBS -N coverage

cd /work/jgrif61/Tigs/raw_data/raw
/home/jgrif61/samtools-1.13/samtools coverage 1S_out.sorted -o 1S.coverage
/home/jgrif61/samtools-1.13/samtools coverage 1U_out.sorted -o 1U.coverage
/home/jgrif61/samtools-1.13/samtools coverage 2S_out.sorted -o 2U.coverage
/home/jgrif61/samtools-1.13/samtools coverage 3S_out.sorted -o 3S.coverage
/home/jgrif61/samtools-1.13/samtools coverage 4S_out.sorted -o 4S.coverage
/home/jgrif61/samtools-1.13/samtools coverage 4U_out.sorted -o 4U.coverage
/home/jgrif61/samtools-1.13/samtools coverage 5S_out.sorted -o 5S.coverage
/home/jgrif61/samtools-1.13/samtools coverage 5U_out.sorted -o 5U.coverage
/home/jgrif61/samtools-1.13/samtools coverage 6U_out.sorted -o 6U.coverage
/home/jgrif61/samtools-1.13/samtools coverage BR_out.sorted -o BR.coverage
/home/jgrif61/samtools-1.13/samtools coverage SD_out.sorted -o SD.coverage
/home/jgrif61/samtools-1.13/samtools coverage allU_out.sorted -o allU.coverage

date 
exit
```

## Compiling SNPs for all samples
This script compiles all the identified SNPs for each sample into a matrix. This sync file format is required for downstream analyses such as Fst, creating Manhattan plots, or CMH tests to be used in both the PoPoolation and PoPoolation2 pipelines. You will want to make sure that the fastq-type flag is set to sanger (this is for Phred quality score matching--yes, this is correct even if you used Illumina sequencing). The min-qual flag must be above 0, the default is 20 which gives you a 99% base call accuracy rate).

What is an Mpileup file?
Pileup format is a text-based format for summarizing the base calls of aligned reads to a reference sequence. This format facilitates visual display of SNP/indel calling and alignment. (Source: Wikipedia)
```{bash eval = FALSE}
#!/bin/bash

#PBS -q checkpt
#PBS -A hpc_kelly_19_3
#PBS -l nodes=1:ppn=16
#PBS -l walltime=72:00:00
#PBS -o /work/jgrif61/Tigs/output_files
#PBS -j oe
#PBS -M jgrif61@lsu.edu
#PBS -N mpileup

cd /work/jgrif61/Tigs/raw_data/raw

samtools mpileup -f full_genome_mito_tigs.fasta 1S_out.sorted 1U_out.sorted 2S_out.sorted 2U_out.sorted 3S_out.sorted 4S_out.sorted 4U_out.sorted 5S_out.sorted 5U_out.sorted 6U_out.sorted BR_out.sorted SD_out.sorted > all_indiv.mpileup

perl /work/jgrif61/Tigs/popoolation2_1201/mpileup2sync.pl --fastq-type sanger --min-qual 20 --input all_indiv.mpileup --output all_indiv.sync

#Alternative code for generation sync file that is 78x faster
java -ea -Xmx7g -jar /work/jgrif61/Tigs/popoolation2_1201/mpileup2sync.jar --input all_indiv.mpileup --output all_indiv_java.sync --fastq-type sanger --min-qual 20 --threads 8

#Making mpileup files for each sample. Can be used for calculating nucleotide diversity for each sample.
samtools mpileup -f full_genome_mito_tigs.fasta 1S_out.sorted BR_out.sorted > 1S.mpileup
samtools mpileup -f full_genome_mito_tigs.fasta 1U_out.sorted > 1U.mpileup
samtools mpileup -f full_genome_mito_tigs.fasta 2S_out.sorted > 2S.mpileup
samtools mpileup -f full_genome_mito_tigs.fasta 1U_out.sorted > 2U.mpileup
samtools mpileup -f full_genome_mito_tigs.fasta 3S_out.sorted > 3S.mpileup
samtools mpileup -f full_genome_mito_tigs.fasta 4S_out.sorted > 4S.mpileup
samtools mpileup -f full_genome_mito_tigs.fasta 4U_out.sorted > 4U.mpileup
samtools mpileup -f full_genome_mito_tigs.fasta 5S_out.sorted > 5S.mpileup
samtools mpileup -f full_genome_mito_tigs.fasta 5U_out.sorted > 5U.mpileup
samtools mpileup -f full_genome_mito_tigs.fasta 6U_out.sorted > 6U.mpileup
samtools mpileup -f full_genome_mito_tigs.fasta BR_out.sorted > BR.mpileup
samtools mpileup -f full_genome_mito_tigs.fasta SD_out.sorted > SD.mpileup

```

## Pop Gen and Downstream Statistical Analyses
These in depth notes are not on the tutorial or manual page, but detailed descriptions for each flag can be found using the --help option when running each script. e.g. perl fst-sliding.pl --help

You will want to play around with the min-coverage, max-coverage, and min-count based on the mean coverage of your samples/populations. Minimum coverages that are too high and max coverages that are too low may result in an empty output file, which is why its a good idea to have an estimate of your mean/min/max coverage for each sample/population. The minimum count must be less than the minimum coverage. You may want to modify this flag based on the number of inidividuals pooled per sample/population.

***Important considerations:***
--suppress-noninformative: Output files will not report results for windows with no SNPs or insufficient coverage. Including this flag will make it easier for downstream manipulation of the output file, such as creating figures.

--pool-size: the number of individuals pooled per sample; If the pool sizes differ among each sample, you can provide the size for each sample individually:
    --pool-size 500 .. all samples/populations have a pool size of 500
    --pool-size 500:300:600 .. first sample/population has a pool size of 500, the second of 300 etc;

--min-coverage: the minimum coverage used for SNP identification, the coverage in ALL samples/populations has to be     higher or equal to this threshold, otherwise no SNP will be called. default=4

--max-coverage: The maximum coverage; All populations are required to have coverages lower or equal than the maximum coverage; Mandatory
The maximum coverage may be provided as one of the following:

 '500' a maximum coverage of 500 will be used for all samples/populations
 '300,400,500' a maximum coverage of 300 will be used for the first sample/population, a maximum coverage of 400 for the second sample/population and so on
 '2%' the 2% highest coverages will be ignored, this value is independently estimated for every sample/population

***NOTE ABOUT MAX COVERAGE***
When genotyping individually barcoded samples, max coverage requirements are generally not recommended. However, pool-seq analyses require a max coverage for calculating Fst, cmh, allele frequency changes, etc, because allele frequency estimates are based on the number allele count, which are prone to PCR duplicates. PCR duplicates from a single individual in the pool may skew allele frequency estimates.

--min-covered-fraction:the minimum fraction of a window being between min-coverage and max-coverage in ALL                    samples/populations; float; default=0.0   
    The tutorial provides an example script with a min-covered fraction of 1, however, you will most likely run into an       error or an empty output file. The min-covered-fraction flag must be smaller than 1: this flag defines what percent     of the window must meet the requirements (e.g. flags: min-count, min-coverage, max-coverage). It is most likely           impossible that your entire sequencing window will meet all of these requirements (unless of course you chose a         window size of 1 bp!).

--min-count: the minimum count of the minor allele, used for SNP identification. SNPs will be identified considering all samples/populations simultanously. default=2. The minimum count MUST be smaller than the min-coverage flag for this script to run.

## Calculate sliding window Fst
This script will calculate Fst pair-wise comparisons for all samples.

```{bash eval = FALSE}
#!/bin/bash

#PBS -q workq
#PBS -A hpc_Kelly_19_3
#PBS -l nodes=1:ppn=16
#PBS -l walltime=4:00:00
#PBS -o /work/jgrif61/Tigs/output_files
#PBS -j oe
#PBS -M jgrif61@lsu.edu
#PBS -N stats

date

cd /work/jgrif61/Tigs/raw_data

perl /work/jgrif61/Tigs/popoolation2_1201/fst-sliding.pl --input all_indiv.sync --output all_PE_w10000_step10000.fst --suppress-noninformative --min-count 5 --min-coverage 20 --max-coverage 200 --min-covered-fraction 0.2 --window-size 10000 --step-size 10000 --pool-size 100

date
exit

```

## Calculate Fisher's Exact Test
This script will calculate significant allele frequency differences for each SNP position between two samples. Results can be plotted in a Manhattan plot. 
Depending on the experimental method, the --window-summary-method flag can be used to determine pvalue for windows of SNPs. The default is to multiply the pvalue of all SNPs for the window. Alternatively, you can specify how pvalues within a window should be summarized. Options include multiple, geometric mean, or median.  The summary of p-values within a window is dependnt on the experimental design and how recent selection was expected to occur. For example, the default parameter might be appropriate if your treatment comparisons are geographically separated populations, where selection is expect to have occurred over hundreds of thousands of years. However, this might be inappropriate for short-term experimental evolution studies, such as the example dataset here. In this dataset, copepods were selected in the lab for 15-20 generations, therefore, we expect relatively large linkage blocks that prevent us from idenitfying the true targets of selection, and therefore I use the geometric mean option.

```{bash eval=FALSE}
#!/bin/bash

#PBS -q workq
#PBS -A hpc_Kelly_19_3
#PBS -l nodes=1:ppn=16
#PBS -l walltime=4:00:00
#PBS -o /work/jgrif61/Tigs/output_files
#PBS -j oe
#PBS -M jgrif61@lsu.edu
#PBS -N stats

date

cd /work/jgrif61/Tigs/raw_data

#You may need to run this on the command line to install twotailed perl Module before running Fisher-test.pl: 
perl -MCPAN -e 'install Text::NSP::Measures::2D::Fisher2::twotailed', see also LSU HPC for details: http://www.hpc.lsu.edu/docs/faq/installation-details.php

perl /work/jgrif61/Tigs/popoolation2_1201/fisher-test.pl --input all_indiv.sync --output all_indiv.fet --suppress-noninformative --min-count 2 --min-coverage 20 --max-coverage 200 --min-covered-fraction 0.2 --window-summary-method geometricmean

```

You may need to run the following code on the command line (not in the queues) to install twotailed perl Module before running Fisher-test.pl. See LSU HPC for details: http://www.hpc.lsu.edu/docs/faq/installation-details.php

```{bash eval=FALSE}
perl -MCPAN -e 'install Text::NSP::Measures::2D::Fisher2::twotailed'
```

## Calculate Exact Allele Frequencies in Samples
```{bash eval=FALSE}
#!/bin/bash

#PBS -q workq
#PBS -A hpc_Kelly_19_3
#PBS -l nodes=1:ppn=16
#PBS -l walltime=4:00:00
#PBS -o /work/jgrif61/Tigs/output_files
#PBS -j oe
#PBS -M jgrif61@lsu.edu
#PBS -N stats

date

cd /work/jgrif61/Tigs/raw_data

perl /work/jgrif61/Tigs/popoolation2_1201/snp-frequency-diff.pl --input all_indiv.sync --output-prefix all_indiv --min-count 6 --min-coverage 20 --max-coverage 200

date
exit

```

## Cochran-Mantel-Haenszel (CMH) test
This script will detect consistent allele frequency changes in biological replicates.
In my case, I was comparing selected samples to one control sample, but if you have replicates of treatments and controls, you can compare columns 1-2, 3-4, etc…

```{bash eval=FALSE}
#!/bin/bash

#PBS -q workq
#PBS -A hpc_Kelly_19_3
#PBS -l nodes=1:ppn=16
#PBS -l walltime=4:00:00
#PBS -o /work/jgrif61/Tigs/output_files
#PBS -j oe
#PBS -M jgrif61@lsu.edu
#PBS -N stats

date

cd /work/jgrif61/Tigs/raw_data

#perl /work/jgrif61/Tigs/popoolation2_1201/cmh-test.pl --input all_indiv.sync_for_cmh --output all_perl_PE_2.cmh --min-count 12 --min-coverage 50 --max-coverage 200 --population 1-6,2-6,3-6,4-6,5-6

date
exit
```

Due to constraints of the test, you can only compare allele frequency changes for individual SNPs and not windows. In my data, I expected high levels of linkage disequilibrium in my samples because individuals were exposed to strong selection pressure. Depending on the amount of linkage disequilibrium you expect in your populations, you may want to account for the non-independence of SNPs following the CMH test. One way to do this, would be to calculate the mean (or geometric mean) p-value for all SNPs in a window. I provide some example R code below for how to do this.
I used the program SeqMonk to map cmh results to the genome and an annotated genome file I created with probes for the desired window size (i.e. 10,000 bp). Output files from SeqMonk will tell you which 10,000 bp window a particular SNP falls into. Thus, SNPs can be grouped for each 10,000bp window on each chromosome.
```{r eval=FALSE}
library("psych")

cmh <-read.table("10000window_BR_cmh_overlap.txt", header=TRUE)
cmh <- cmh[c(1,2,3,4,6)] #Keeping only the columns I care about
cmh$ID <- paste(cmh$Chromosome,cmh$Start,sep=".") #creating an ID column so each SNP location is tied to its chromosome

cmh_p <-read.table("Tig.cmh", header=TRUE) #read in the original output from the CMH test, because SeqMonk output discards the important pvalue column.
colnames(cmh_p) <- c("CHR", "BP", "Allele", "1S", "2S", "3S", "4S", "5S", "U", "p") #rename columns based on sample names
cmh_p$ID <- paste(cmh_p$CHR,cmh_p$BP,sep=".")
cmh_p <- cmh_p[c(10,11)]

cmh_all <- merge(cmh, cmh_p, by="ID", all = TRUE)

#Now we can calculate the geometric mean for each window one chromosome at a time
library("EnvStats")
chr1<- cmh_all[cmh_all$Chromosome == "Chromosome_1",]
Chr1_ave <- aggregate(p~FeatureID, data=chr1, FUN=function(x) c(mean=geoMean(x)))
Chr1_ave$CHR <- rep(1,nrow(Chr1_ave))
Chr1_ave$CHR2 <- rep("Chromosome_1",nrow(Chr1_ave))

chr2<- cmh_all[cmh_all$Chromosome == "Chromosome_2",]
Chr2_ave <- aggregate(p~FeatureID, data=chr2, FUN=function(x) c(mean=geoMean(x)))
Chr2_ave$CHR <- rep(2,nrow(Chr2_ave))
Chr2_ave$CHR2 <- rep("Chromosome_2",nrow(Chr2_ave))

chr3<- cmh_all[cmh_all$Chromosome == "Chromosome_3",]
Chr3_ave <- aggregate(p~FeatureID, data=chr3, FUN=function(x) c(mean=geoMean(x)))
Chr3_ave$CHR <- rep(3,nrow(Chr3_ave))
Chr3_ave$CHR2 <- rep("Chromosome_3",nrow(Chr3_ave))

chr4<- cmh_all[cmh_all$Chromosome == "Chromosome_4",]
Chr4_ave <- aggregate(p~FeatureID, data=chr4, FUN=function(x) c(mean=geoMean(x)))
Chr4_ave$CHR <- rep(4,nrow(Chr4_ave))
Chr4_ave$CHR2 <- rep("Chromosome_4",nrow(Chr4_ave))

chr5<- cmh_all[cmh_all$Chromosome == "Chromosome_5",]
Chr5_ave <- aggregate(p~FeatureID, data=chr5, FUN=function(x) c(mean=geoMean(x)))
Chr5_ave$CHR <- rep(5,nrow(Chr5_ave))
Chr5_ave$CHR2 <- rep("Chromosome_5",nrow(Chr5_ave))

chr6<- cmh_all[cmh_all$Chromosome == "Chromosome_6",]
Chr6_ave <- aggregate(p~FeatureID, data=chr6, FUN=function(x) c(mean=geoMean(x)))
Chr6_ave$CHR <- rep(6,nrow(Chr6_ave))
Chr6_ave$CHR2 <- rep("Chromosome_6",nrow(Chr6_ave))

chr7<- cmh_all[cmh_all$Chromosome == "Chromosome_7",]
Chr7_ave <- aggregate(p~FeatureID, data=chr7, FUN=function(x) c(mean=geoMean(x)))
Chr7_ave$CHR <- rep(7,nrow(Chr7_ave))
Chr7_ave$CHR2 <- rep("Chromosome_7",nrow(Chr7_ave))

chr8<- cmh_all[cmh_all$Chromosome == "Chromosome_8",]
Chr8_ave <- aggregate(p~FeatureID, data=chr8, FUN=function(x) c(mean=geoMean(x)))
Chr8_ave$CHR <- rep(8,nrow(Chr8_ave))
Chr8_ave$CHR2 <- rep("Chromosome_8",nrow(Chr8_ave))

chr9<- cmh_all[cmh_all$Chromosome == "Chromosome_9",]
Chr9_ave <- aggregate(p~FeatureID, data=chr9, FUN=function(x) c(mean=geoMean(x)))
Chr9_ave$CHR <- rep(9,nrow(Chr9_ave))
Chr9_ave$CHR2 <- rep("Chromosome_9",nrow(Chr9_ave))

chr10<- cmh_all[cmh_all$Chromosome == "Chromosome_10",]
Chr10_ave <- aggregate(p~FeatureID, data=chr10, FUN=function(x) c(mean=geoMean(x)))
Chr10_ave$CHR <- rep(10,nrow(Chr10_ave))
Chr10_ave$CHR2 <- rep("Chromosome_10",nrow(Chr10_ave))

chr11<- cmh_all[cmh_all$Chromosome == "Chromosome_11",]
Chr11_ave <- aggregate(p~FeatureID, data=chr11, FUN=function(x) c(mean=geoMean(x)))
Chr11_ave$CHR <- rep(11,nrow(Chr11_ave))
Chr11_ave$CHR2 <- rep("Chromosome_11",nrow(Chr11_ave))

chr12<- cmh_all[cmh_all$Chromosome == "Chromosome_12",]
Chr12_ave <- aggregate(p~FeatureID, data=chr12, FUN=function(x) c(mean=geoMean(x)))
Chr12_ave$CHR <- rep(12,nrow(Chr12_ave))
Chr12_ave$CHR2 <- rep("Chromosome_12",nrow(Chr12_ave))

cmh_p_ave_allchr <- rbind(Chr1_ave, Chr2_ave, Chr3_ave, Chr4_ave, Chr5_ave, Chr6_ave, Chr7_ave,Chr8_ave, Chr9_ave, Chr10_ave, Chr11_ave, Chr12_ave)

library(tidyr)
cmh_p_ave_allchr2 <- separate(cmh_p_ave_allchr, FeatureID, c("BP_start", "BP_end"))
cmh_p_ave_allchr2$BP_start <- as.numeric(cmh_p_ave_allchr2$BP_start)
cmh_p_ave_allchr2$BP_ID <- paste(cmh_p_ave_allchr2$BP_start + 5000) #I wanted the window ID to start in the middle of the window
cmh_p_ave_allchr2$ID <- paste(cmh_p_ave_allchr2$CHR,cmh_p_ave_allchr2$BP_start,sep=".")

write.table(cmh_p_ave_allchr2, file = "cmh_window_geomean_p", sep = "\t")

```

## PoPoolation Pipelines for analyses
## Nucleotide Diversity
Calulate Pi (nucleotide diversity) for individual samples. Use mpileup files that were created for a single sample as the input. Below is an example for one sample.
```{bash eval=FALSE}
#!/bin/bash

#PBS -q workq
#PBS -A hpc_Kelly_19_3
#PBS -l nodes=1:ppn=16
#PBS -l walltime=4:00:00
#PBS -o /work/jgrif61/Tigs/output_files
#PBS -j oe
#PBS -M jgrif61@lsu.edu
#PBS -N stats

date

cd /work/jgrif61/Tigs/raw_data

perl /work/jgrif61/Tigs/popoolation_1.2.2/Variance-sliding.pl --measure pi --input 1S.mpileup --output 1S.pi --min-count 2 --min-coverage 10 --max-coverage 200 --window-size 10000 --step-size 10000 --pool-size 100 --fastq-type sanger

date
exit

```

### 2021-11-08 RADseq.Rmd
Cleaned up RADseq

@@
 
-* vcftools(link)
+* [vcftools](https://vcftools.github.io/man_latest.html)
 * gnuplot
@@
 
-## Considerations During Lab Work -- Setting up for Success {#labwork}
+# Considerations During Lab Work {#labwork}
 
@@
 
-## Principles for Analyzing Your Data
+# Principles for Analyzing Your Data
 
-### Steps for a robust RAD analysis 
+## Steps for a robust RAD analysis 
 This is just one approach for working through your data.
@@
 
-### First, look at the raw data! {#fastqc}  
+## First, look at the raw data! {#fastqc}  
 Always look at your data with fastqc(link) before starting an assembly.
@@
 
-### Run an assembly pipeline {#assembly}
+## Run an assembly pipeline {#assembly}
 
@@
 
-#### iPyrad walkthrough
-
-Link to annotated ipyrad notebook
-
-### Evaluate potential confounding factors {#confound}
+## Evaluate potential confounding factors {#confound}
 
@@
 
-#### "Bad" samples {#bad}
+### "Bad" samples {#bad}
 
@@
 hist(m2,breaks=50)
-
-# close the GDS file
-seqClose(gdsin)
 ```
 
-Another common method of filtering and evaluating you data without ever using R is VCFtools(link). 
+Another common method of filtering and evaluating you data without ever using R is [VCFtools](https://vcftools.github.io/man_latest.html). 
 **Missingness with vcftools, on the command line**  
@@
 
-#### The power of PCA
+### The power of PCA
 
@@
 ```{r}
-#open the gds file previously created
-filename.gds = "RAD_data/OL_subset.gds" #replace with your file name 
-gdsin = SeqArray::seqOpen(filename.gds)
+#open the gds file previously created, if not already open
+#filename.gds = "RAD_data/OL_subset.gds" #replace with your file name 
+#gdsin = SeqArray::seqOpen(filename.gds)
 # exclude samples previously identified as having too much missin g data
@@
 ```{r}
-# PCA only on SNPs with a minor allele freq greater than 2.5%
+# PCA only on SNPs with a minor allele freq greater than 5%
 pca.out = SNPRelate::snpgdsPCA(autosome.only = F, gdsin, num.thread=2, remove.monosnp = T, maf = 0.05,
@@
 
-Data Info: Below we use RAD data from an Australian marine species complex.
-```{r, eval =F}
+<!-- Data Info: Below we use RAD data from an Australian marine species complex. ---> 
+```{r, eval =F,include=F}
 
@@
 #PLOT PCA
-#PC1 v PC2 colored by collection location
+#PC1 v PC2 with Sample Labels
 print(
@@
       ggplot(., aes(x=V1, y=V2, label = ID)) +
-      geom_text(size =1.5) +
+      geom_text(size =3) +
       geom_hline(yintercept = 0) +
@@
 
-```{r}
+```{r, eval=F}
 # which samples have PC1 > -0.7 and PC2 < -0.3?
@@
   + or you actually have no structure in your dataset!
-  + `r colorize("figure example, no clustering", "red")`
+
+<!--`r colorize("figure example, no clustering", "red")`-->
 
@@
 
-Here is an example of a PCA with sample replicates, and then with sample replicates removed:
+<!--Here is an example of a PCA with sample replicates, and then with sample replicates removed: -->
 
-When removing a sample replicate, you can choose one at random or pick the replicate with the least missing data(link). 
+When removing a sample replicate, you can choose one at random or pick the replicate with the least [missing data](#bad). 
 ```{r, eval=FALSE}
@@
                                sample.id = keep)
-#close gds file (saves memory)
 
@@
 
-#### Visually inspect alignments {#align}  
+<!-- #### Visually inspect alignments {#align}  
 Another useful approach is to visually inspect alignments for SNPs or loci that are identified as "weird" in the subsequent analyses.
@@
 `r colorize("code stacks", "red")`    
+-->
 
-#### Batch effects {#batch}
+### Batch effects {#batch}
 As discussed in [Setting up for success](#labwork), batch effects can arise by minor (or not so minor) differences during the library prep and sequencing stage.
@@
 
-Let's pretend like we have batch effects that were observed on PC 1. With SNPRelate we would extract the PC loadings for each SNP at each PC, then 
+In SeqArray/SNPRelate, if you have batch effects that are identifiable by PCA, you would extract the PC loadings for each SNP atthe PC which represents the batch effect, then filter those SNPs out and create a new VCF for downstream analyses.
+```{r, eval=T}
+# assume you have a GDS object called gdsin and a PCA object called pca.out
+# extract PCA loadings
+SnpLoad <- snpgdsPCASNPLoading(pca.out, gdsin)
+names(SnpLoad)
+dim(SnpLoad$snploading)
+#plot a histogram of absolute PC loading values for PC1, with larger loading meaning more effect on PC1
+pc<-1
+hist(sort(abs(SnpLoad$snploading[pc,]),decreasing=T,index.return=T)[[1]],breaks = 30,main="PC loadings: PC1",)
+```
+Let's remove the largest effect loci over abs(0.010) and see if it changes our PCA
 ```{r}
+#extract SNP.IDs for loadings on PC 1 > abs(0.015) 
+batch.snp.ids <- SnpLoad$snp.id[which(abs(SnpLoad$snploading[1,]) >= 0.010)]
 
-SnpLoad <- snpgdsPCASNPLoading(pca.out, gdsin)
-#load.df <- cbind(SnpLoad$snp.id,SnpLoad$snploading)
-#colnames(load.df) <- c("Snp.ID",)
+#rerun PCA
+snp.id.allfilt <- setdiff(snpset.id,batch.snp.ids)
+pca.out.pcfilt = SNPRelate::snpgdsPCA(autosome.only = F, gdsin, num.thread=2, remove.monosnp = T, maf = 0.05,
+                               snp.id=snp.id.allfilt,
+                               sample.id = keep) # filtering for pruned SNPs
+print(
+  as.data.frame(pca.out.pcfilt$eigenvect) %>%
+      tibble::add_column(., STRATA =  sample.strata.order$STRATA) %>%
+      ggplot(., aes(x=V1, y=V2, color = STRATA)) + 
+      geom_point(size=2) +
+      stat_ellipse(level = 0.95, size = 1) +
+      geom_hline(yintercept = 0) +
+      geom_vline(xintercept = 0) +
+      theme_bw() +
+      xlab(paste0("PC1 [",paste0(round(eig[1], 2)), "%]")) +
+      ylab(paste0("PC2 [",paste0(round(eig[2], 2)), "%]")) +
+      ggtitle("PCA Colored by Collection Location")
+)
+
+```
+To save your filtered GDS object as a VCF: 
+```{r}
+# set a filter to exclude samples with missing data, and SNPs in linkage or with a batch effect
+seqSetFilter(gdsin, sample.id=keep,variant.id=snp.id.allfilt)
+
+# convert to vcf
+seqGDS2VCF(gdsin, "RAD_data/OL-miss-pc1-linkage-filt.vcf.gz")
+
+#close the gds file to save memory
+closefn.gds(gdsin)
 
@@
 
-Once you've identified these batch effect SNPs, it can be useful to [visualize the alignments](#align) before removing them.
-If the entire alignment for that locus looks wonky, you should remove all SNPs from that RAD locus from your final dataset before conducting analyses.
-
-#### Cryptic species/contamination/clones {#cryptic}  
+### Cryptic species/contamination/clones {#cryptic}  
 One issue that is particularly common, especially for marine invertebrates, are cryptic, yet highly diverged lineages occurring in the same geographic location.
@@
 A PCA using all the samples can help identify cryptic species if samples are forming clusters that do not align with any other possible factor (eg, geography, sequencing batch).
-`r colorize("PCA figure of cryptic species", "red")`
 
-Note: Notice the huge variation explained along the first PC-axis.
+<!--Note: Notice the huge variation explained along the first PC-axis.-->
 
-You can also check heterozygosity, and missingness!
+You can also check heterozygosity. Large differences in heterozygosity between outliers can suggest different species or cryptic hybrids.
 
-```{r}
-```
-
- Note: Two outliers are obvious.
+<!--Note: Two outliers are obvious.
 Large differences in heterozygosity between samples are clear, with one location showing two distinct clusters of samples.
 These samples were later determined using mtDNA and morphological analysis to be cryptic species.
-*Unpublished data from a marine species in Australian waters.*
+*Unpublished data from a marine species in Australian waters.*-->
 
@@
 In a PCA, clones will often appear as their own very tight cluster, similar to sample replicates.
-Below the PCA shows x and a simple calculate of pairwise-distance (i.e.
-here manhattan distance) shows x known duplicates, and x unknown clones.
-*other ways to tell they are clones*
-`r colorize("PCA figure of cryptic clones", "red")`
+A calculation of sample pairwise distance (e.g., Manhattan distance) can very duplicates/clones.
 
-### Test a range of key parameters {#test}
+<!--Below the PCA shows x and a simple calculate of pairwise-distance (i.e. here manhattan distance) shows x known duplicates, and x unknown clones. 
+`r colorize("PCA figure of cryptic clones", "red")`-->
 
-#### Clustering threshold (de novo assembly) {#cluster}  
+## Test a range of key parameters {#test}
+
+### Clustering threshold (de novo assembly) {#cluster}  
 During de novo RADseq assembly, reads are assembled into contigs so that each contig represents a single locus.
@@
 
-#### Mapping parameters (reference-based and some de novo assemblers)   
+### Mapping parameters (reference-based and some de novo assemblers)   
 When a sequencing read maps to the wrong assembled contig or place in a reference genome (ie, a non-homologous locus), this can result in erroneous SNPs.
@@
 
-#### Read depth threshold {#seq}
+### Read depth threshold {#seq}
 
@@
 
-### Filter your data many ways (and embrace the missingness!)
+## Filter your data many ways
+### (and embrace the missingness!) {.unlisted .unnumbered}
 * Don't over filter for missingness
@@
 
-## Special Cases  
+<!--## Special Cases  
 ### DARTseq  
@@
 ### Reference-based vs. de novo RADseq
-<!---brief outline on general pipeline--->
+-->
 
-## Good tutorials  
+# Helpful tutorials  
 * [Marine Genomics open course](https://baylab.github.io/MarineGenomics/)

### 2021-11-08 RADseq.Rmd
Added radseq refs

@@
 ```{r,message = FALSE,echo = FALSE}
-  library("knitcitations")
-  library("knitr")
-  cite_options(citation_format = "pandoc", max.names = 3, style = "html", hyperlink = "to.doc")
-  bib <- read.bibtex("bibtexlib.bib")
-  opts_chunk$set(tidy = FALSE, message = FALSE, warning = FALSE,
-                 fig.width = 10, fig.height = 6, cache = TRUE)
-  # use this to set knitr options:
-  # http://yihui.name/knitr/options #chunk_options
+library(knitr)
+knitr::opts_chunk$set(echo = TRUE)
+library(knitcitations)
+library(kableExtra)
+opts_chunk$set(fig.width = 10,
+               fig.height = 5,
+               cache = FALSE)
+cite_options(citation_format = "pandoc", max.names = 3, style = "html", 
+             hyperlink = "to.doc")
 ```
@@
 subtitle: '*Katherine Silliman, Danielle Davenport*'
+bibliography: RCN_site-Nov05.bib
 ---
@@
 * [vcftools](https://vcftools.github.io/man_latest.html)
-* gnuplot
+* [gnuplot](https://riptutorial.com/gnuplot/example/11275/installation-or-setup)
 
@@
 Double-digest RADseq (ddRAD; Peterson et al.,2012)  selects  markers  with  two  restriction enzymes  with  different  cut  frequencies.
-Genotype-by-Sequencing  (GBS;  Elshire  et  al.,  2011)  uses  a  frequent-cutting  restriction  enzyme  with PCR size selection.
-See this review for a comparison of some of the more popular protocols (link).
+Genotype-by-Sequencing  (GBS;  [@Elshire2011-or])  uses  a  frequent-cutting  restriction  enzyme  with PCR size selection.
+There are a number of reviews comparing different RADseq.GBS methods (e.g., [@Andrews2016-ay]).
 In this guide we use the term "RADseq" to refer to any of these protocols, including those that don't involve random shearing of data.
@@
 
-Some aspects of RADseq data that are unique compared to whole genome resequencing (more):
+<!--Some aspects of RADseq data that are unique compared to whole genome resequencing (more):-->
 
@@
 Every dataset is different [add more about philosophy over best practices].
-Much of this guide is directly inspired by this [excellent review paper by O'Leary et al.
-(2018)](link), especially the section on minimizing errors due to library prep.
+Much of this guide is directly inspired by [this excellent review paper](https://onlinelibrary.wiley.com/doi/10.1111/mec.14792) by [@OLeary2018-iy], especially the section on minimizing errors due to library prep.
 We recommend reading and cross-referencing with this paper, and citing it if you follow its suggestions.
@@
 
-![Table 1 from O'Leary et al. 2018.](images/oleary_table1.png)
+![Table 1 from [@OLeary2018-iy].](images/oleary_table1.png)
 
@@
 Just make sure to keep track of everything!
+
 * Randomize samples across library prep batches and sequencing lanes!
@@
 If you are preparing groups of individuals in different batches to be pooled later, randomize across these batches.
-Alicia MY even recommends randomizing across DNA extractions, especially if the person doing the extractions is new to molecular work.
+[Alicia Mastretta-Yanes](https://mastrettayanes-lab.org/) even recommends randomizing across DNA extractions, especially if the person doing the extractions is new to molecular work.
 This is to allow you to control for potential [batch effects](#batch) that are often observed with RAD data.
@@
 1) [Look at your raw data.](#fastqc)
-2) [Run an assembly pipeline](link assembly) through to a SNP dataset, using either all samples or a representative subset.
-Parameters can come from those used in a similar study, or default parameters.
+2) [Run an assembly pipeline](#assembly) through to a SNP dataset, using either all samples or a representative subset. Parameters can come from those used in a similar study, or default parameters.
 3) Filter your data minimally, then [evaluate for potential sources of error.](#error)
@@
 ## First, look at the raw data! {#fastqc}  
-Always look at your data with fastqc(link) before starting an assembly.
+Always look at your data with [FastQC](https://www.bioinformatics.babraham.ac.uk/projects/fastqc/) before starting an assembly.
 First, this is a good check to just make sure the sequencing worked.
-If you have demultiplexed data, you can use multiqc(link) to generate fastqc plots for all individuals and quickly identify ones that did not sequence well.
-Check out [this informative tutorial](https://datacarpentry.org/wrangling-genomics/02-quality-control/index.html) on running fastqc and interpreting the output.  
-Do you have a lot of adapter sequences?
+If you have demultiplexed data, you can use [MultiQC](https://multiqc.info/) to generate FastQC plots for all individuals and quickly identify ones that did not sequence well.
+Check out [this informative tutorial](https://datacarpentry.org/wrangling-genomics/02-quality-control/index.html) on running FastQC and interpreting the output.  
+**Questions to ask:** Do you have a lot of adapter sequences?
 Are the ends of your reads very low quality?
@@
 Sometimes you can look at your fastq data files and see that there was a problem with the sequencing such that the cut site is either offset by one or more bases, or contains many errors.
-If this is not being addressed by the default filtering steps in your assembly pipeline, you can trim off N bases from the beginning or end of R1 and R2 reads in ipyrad (`trim_reads` param), or with cutadapt before using stacks, or customize the Trimmoatic step for dDocent.
+If this is not being addressed by the default filtering steps in your assembly pipeline, you can trim off N bases from the beginning or end of R1 and R2 reads in [ipyrad](https://ipyrad.readthedocs.io/en/latest/index.html) (`trim_reads` param), or with cutadapt before using stacks, or customize the Trimmoatic step for [dDocent](https://www.ddocent.com/).
 
@@
 
-There are a number of freely available pipelines for processing RADseq data, with the most popular being dDocent, Stacks, and ipyrad.
+There are a number of freely available pipelines for processing RADseq data, with the most popular being [dDocent](https://www.ddocent.com/), [Stacks](https://catchenlab.life.illinois.edu/stacks/), and [ipyrad](https://ipyrad.readthedocs.io/en/latest/index.html).
 These pipelines vary slightly in their underlying methodologies, customization options, and additional included analyses.
 
-|  | dDocent | ipyrad | Stacks2 |
+|  | [dDocent](https://www.ddocent.com/) | [ipyrad](https://ipyrad.readthedocs.io/en/latest/index.html)| [Stacks2](https://catchenlab.life.illinois.edu/stacks/) |
 |---|---|---|---|
-| Supported datatypes | Paired-end: ddRAD, ezRAD, RAD (random shearing), data with large overlap  between forward and reverse reads Single-end: any RADseq method  de novo and reference-based  If doing de novo assembly, reads cannot be trimmed outside of dDocent |  |  |
+| Supported datatypes | Paired-end: ddRAD, ezRAD, RAD (random shearing), data with large overlap  between forward and reverse reads  
+Single-end: any RADseq method  de novo and reference-based.  
+If doing de novo assembly, reads cannot be trimmed outside of dDocent |  |  |
 | Unique aspects | Novel data reduction approach used to inform coverage cutoffs |  |  |
@@
 
-Some groups have also developed pipelines for specific flavors of lab (e.g., Matz lab and 2bRAD) or proprietary software (eg, DARTseq), which we touch on below.
-**In most cases, it is recommended that you use a pipeline developed and tested for RAD data, especially if you are making a de novo assembly.** LaCava et al, 2019 have an excellent study where they review various de novo assemblers that are used in these pipelines.
+Some groups have also developed pipelines for specific flavors of RAD (e.g., [Matz lab for 2b-RAD](https://github.com/z0on/2bRAD_denovo/blob/master/2bRAD_README.sh)) or proprietary software (eg, [DARTseq](http://georges.biomatix.org/storage/app/media/uploaded-files/dartR_Workbook.pdf)), which we touch on below.
+**In most cases, it is recommended that you use a pipeline developed and tested for RAD data, especially if you are making a de novo assembly.** [@LaCava2020-bg] have an excellent study where they review various *de novo* assemblers that are used in these pipelines.
 Assemblers that were not explicitly developed for short reads (eg, Velvet, ABySS) performed very poorly, while CD-HIT (the assembler in dDocent) performed the best.
@@
 
-For the code examples in this section, we primarily make use of the R package SeqArray [Zheng et al., 2017](https://doi.org/10.1093/bioinformatics/btx145), which can read and manipulate VCFs.
+For the code examples in this section, we primarily make use of the R package SeqArray [@Zheng2017-pl], which can read and manipulate VCFs.
 If you’re familiar with R, SeqArray is simple to use.
-The package can efficiently store sequence variant calls with hundreds of thousands of individuals,  variants (e.g., SNVs, indels and structural variation calls) and annotations.
+The package can efficiently store sequence variant calls with hundreds of thousands of individuals, variants (e.g., SNPs, indels and structural variation calls) and annotations.
 It employs a data container called a CoreArray Genomic Data Structure (GDS).
@@
 SNPRelate, SeqVarTools).
-We also like it because you can filter your data before running certain analyses without generating a separate filtered VCF file.
+We also like it because you can filter your data before running certain analyses without 1st generating a separate filtered VCF file.
 
@@
 It is always helpful to have a metadata file with information for each sample, such as sampling site, sequencing library, etc.
-In our example, our matadata file (OL.popmap) is tab-delimited and has the column headers:
+In our example, our metadata file (OL.popmap) is tab-delimited and has the column headers:
 ID: sample ID
@@
 Another common method of filtering and evaluating you data without ever using R is [VCFtools](https://vcftools.github.io/man_latest.html).
+
 **Missingness with vcftools, on the command line**  
@@
 2) Another great way to explore missingness with your data is to construct a heatmap of loci that are genoptyped between pairs of samples.
-Generally, samples that are more closely related will share more loci with each other due to cutsite dropout (link).
+Generally, samples that are more closely related will share more loci with each other due to cutsite dropout (cite).
 Deviations from that pattern can pinpoint bad samples as well as batch effects.
@@
 
-Here we use SeqArray and SNPRelate to run a PCA in R. Other popular methods for PCA include pcangsd(link), ...., but note that ... requires some file conversions everytime you filter.
+Here we use SeqArray and SNPRelate to run a PCA in R.
+
+<!--Other popular methods for PCA include pcangsd(link), ...., but note that ... requires some file conversions everytime you filter.-->
 
@@
 #gdsin = SeqArray::seqOpen(filename.gds)
-# exclude samples previously identified as having too much missin g data
+# exclude samples previously identified as having too much missing data
 bad_samples = c("OR3_13_C3","CA4_12_C1","OR3_3_C7","CA4_13_C1","BC2_6_C2","BC2_9_C4")
@@
 ### Batch effects {#batch}
-As discussed in [Setting up for success](#labwork), batch effects can arise by minor (or not so minor) differences during the library prep and sequencing stage.
+As discussed in [Considerations During Lab Work](#labwork), batch effects can arise by minor (or not so minor) differences during the library prep and sequencing stage.
 Randomizing samples across libraries and sequencing lanes can help mitigate the influence of batch effects on your downstream analyses, but sometimes this isn't possible, esp.
@@
 
-`r colorize("PCA figure of batch effect", "red")`
+<!--`r colorize("PCA figure of batch effect", "red")`-->
 
@@
 
-In SeqArray/SNPRelate, if you have batch effects that are identifiable by PCA, you would extract the PC loadings for each SNP atthe PC which represents the batch effect, then filter those SNPs out and create a new VCF for downstream analyses.
+In SeqArray/SNPRelate, if you have batch effects that are identifiable by PCA, you would extract the PC loadings for each SNP at the PC which represents the batch effect, plot a histogram to see the distribution, choose a cutoff, then filter those SNPs out and create a new VCF for downstream analyses.
+For this example, let's pretend that we see a batch effect on PC1.
 ```{r, eval=T}
@@
 ```
-Let's remove the largest effect loci over abs(0.010) and see if it changes our PCA
+Let's remove the largest effect loci over abs(0.010) and see if it changes our PCA.
 ```{r}
-#extract SNP.IDs for loadings on PC 1 > abs(0.015) 
+#extract SNP.IDs for loadings on PC 1 > abs(0.010) 
 batch.snp.ids <- SnpLoad$snp.id[which(abs(SnpLoad$snploading[1,]) >= 0.010)]
@@
 ```
-To save your filtered GDS object as a VCF: 
+
+Our PCA is now clustering better by population, so we will now export a new VCF that includes all the filtering we've done so far: 1) removed individuals with missing data, 2) thinned our SNPs for linkage, and 3) removed SNPs driving a batch effect.
+
+To save your filtered GDS object as a VCF in SeqArray: 
 ```{r}
@@
 You can then estimate genetic divergence between these clusters using pairwise Fst and/or phylogenetic inference to determine how diverged the groups are.
-If they are very diverged, it may be better to assemble each cryptic species separately, especially for denovo assemblies with dDocent and Stacks, especially since often times restriction cut-sites are not always conserved; see [Wagner et al., 2013](link), [Rubin et al., 2012](link)  These two pipelines used a set of given samples to create a "reference" or "catalog" to which all other samples are mapped.
+If they are very diverged, it may be better to assemble each cryptic species separately, especially for denovo assemblies with dDocent and Stacks, especially since often times restriction cut-sites are not always conserved; see [@Wagner2013-ff,@Rubin2012-hw].
+These two pipelines used a set of given samples to create a "reference" or "catalog" to which all other samples are mapped.
 If highly diverged samples are included in creating this reference, it can lead to [oversplitting](#cluster).
@@
 
-### Clustering threshold (de novo assembly) {#cluster}  
-During de novo RADseq assembly, reads are assembled into contigs so that each contig represents a single locus.
-This can occur separately for each individual (ipyrad) or jointly for all or a subset of individuals (dDocent, Stacks).
+There are many parameter choices to make when assembling and genotyping, and those parameters vary slightly across pipelines.
+The parameters below are general across RADseq pipelines (although with different names), and are shown to have the largest impact on genoyping error rates.
+### Clustering threshold {#cluster}
+#### (de novo assembly only)
+During *de novo* RADseq assembly, reads are assembled into contigs (contiguous sequences), so that each contig represents a single locus.
+This assembly can occur separately for each individual (ipyrad) or jointly for multiple individuals (dDocent, Stacks).
 To call SNPs, in the case of ipyrad, contigs are then clustered and aligned across individuals, while in dDocent and Stacks, reads for each individual are mapped to the "reference" contigs.
-All of these pipelines use sequence similarity thresholds to determine if fragments are homologous (from the same place in the genome) and therefore should be clustered together.
-If the sequence similarity threshold is too low, *undersplitting* can occur so that multiple loci are combined into a single contig.
-Undersplitting is common with repetitive regions and paralogs. It can inflate mean observed heterozygosity.
+All of these pipelines use sequence similarity (or **clustering threshold**) to determine if fragments are homologous (from the same place in the genome) and therefore should be clustered together.
 
-If the threshold is too high, it can lead to *oversplitting*, where alleles of the same locus are split into two or more contigs. This can lead to reduced mean observed heterozygosity.
+If the clustering threshold is too low, *undersplitting* can occur so that multiple loci are combined into a single contig.
+Undersplitting is common with repetitive regions and paralogs.
+It can inflate mean observed heterozygosity.
+If the threshold is too high, it can lead to *oversplitting*, where alleles of the same locus are split into two or more contigs.
+This can lead to reduced mean observed heterozygosity.
 
@@
 Undersplitting a little is better than oversplitting, as it is easier to identify lumped paralogs downstream and remove them.
-**What influences the "optimal" clustering threshold of your data?**  
+
+#### **What influences the "optimal" clustering threshold of your data?**  
 Clustering threshold is usually determined by how much genetic diversity you have in your dataset.
@@
 If your dataset has low genetic diversity (a single species with lower heterozygosity, a single family), then you will need a higher clustering threshold.
-Polyploid or highly repetitive (ie large) genomes may also have different requirements than diplod species with smaller genomes.
-Other factors, such as sequencing error rate, may play a role as well so it is usually a good idea to always test a few clustering thresholds with a new dataset.
-However, as long as your clustering threshold is in the ballpark it shouldn't affect downstream analyses (according to Isaac).
+Polyploid or highly repetitive (i.e., large) genomes may also have different requirements than diploid species with smaller genomes.
+Other factors, such as sequencing error rate, may play a role as well.
+**We recommend always testing a few clustering thresholds with a new dataset and/or new species**.
+However, as long as your clustering threshold is in the ballpark it shouldn't affect downstream analyses (according to [Isaac Overcast](https://isaacovercast.github.io/)).
 
 There are a few strategies to identify a good clustering threshold for your data, all of which require assembling your data with a range of clustering thresholds.  
-1) Metrics developed by [McCartney-Melstad 2017](link), based on popgen theory and landscape genetic expectations  
-2) Utilize sample replicates (either exact replicates or individuals from the same locality) to quantify genotyping error and minimize error while maximizing the number of informative loci [Mastretta-Yanes 2015](link)
-3) Evaluating the number of haplotypes across loci, with many 3+ haplotypes indicating undersplitting [Harvey 2015](link), [Ilut 2014](link)  
-4) Make assemblies at a range of thresholds and parameters, plot the # of SNPs or contigs per assembly, and make a judgement call [example in dDocent tutorial](link).
+1) Metrics developed by [@McCartney-Melstad2019-ac], based on popgen theory and landscape genetic expectations  
+2) Utilize sample replicates (either exact replicates or individuals from the same locality) to quantify genotyping error and minimize error while maximizing the number of informative loci [@Mastretta-Yanes2015-jb]
+3) Evaluating the number of haplotypes across loci, with many 3+ haplotypes indicating undersplitting [@Harvey2015-vh;@Ilut2014-ja]
+4) Make assemblies at a range of thresholds and parameters, plot the # of SNPs or contigs per assembly, and make a judgement call [example in dDocent tutorial](https://www.ddocent.com/assembly/).
 If most parameters give you about the same # of SNPs or contigs, you can feel confident that this parameter doesn't matter much for your data and just choose a reasonable value.
 
-### Mapping parameters (reference-based and some de novo assemblers)   
+### Mapping parameters 
+#### (reference-based and some de novo assemblers) {.unlisted .unnumbered}    
 When a sequencing read maps to the wrong assembled contig or place in a reference genome (ie, a non-homologous locus), this can result in erroneous SNPs.
@@
 * What is the distribution of mapping scores?  
-* ratio of mean (across inds) mapping quality scores for reference and alternate allele (O'Leary cite). 0.25-1.75 cut off implemented in dDocent.  
+* ratio of mean (across individuals) mapping quality scores for reference and alternate allele [@OLeary2018-iy]. 0.25-1.75 cut off implemented in dDocent.  
   + this info might not be coded in all VCFs
@@
 
-FASTQ files that typically come from the sequencing center include a [Phred scale quality score](link) for each base, which indicates the probability of that base call being correct.
+FASTQ files that typically come from the sequencing center include a [Phred scale quality score](https://gatk.broadinstitute.org/hc/en-us/articles/360035531872-Phred-scaled-quality-scores) for each base, which indicates the probability of that base call being correct.
 This information can be used prior to assembly to trim low-quality sections from either end of reads or remove reads altogether.
@@
 For diploid species, this would be sequencing each RAD haplotype at least 5x-10x.
-The SNP calling step of RAD pipelines all require setting a minimum number of reads to determine whether a site is homozygous or heterozygous **add image example**.
-If this threshold is too low, sequencing errors are more likely to be included and the SNP caller will have a harder time accurately identifying SNPs ([but see our section on low coverage WGS methods!](link).
+The SNP calling step of RAD pipelines all require setting a minimum number of reads to determine whether a site is homozygous or heterozygous.
+If this threshold is too low, sequencing errors are more likely to be included and the SNP caller will have a harder time accurately identifying SNPs ([in which case, you'd be better off with genotype likelihood methods](https://github.com/therkildsen-lab/genomic-data-analysis/blob/master/lcwgs_data_analysis.md).
 If it's too high, then real polymorphic loci will be excluded.
@@
 
-## Filter your data many ways
+## Filtering SNPs
 ### (and embrace the missingness!) {.unlisted .unnumbered}
+
+`r colorize("This section is still under development! Please contribute on Github or provide feedback in [Discussions](https://github.com/MarineOmics/marineomics.github.io/discussions)", "red")`
+Even with a reduced representation sequencing approach, your assembly and genotyping pipeline will generate data for thousands of SNPs across your individuals.
+These SNPs will always need to be filtered before doing downstream analyses.
+#### Main reasons to filter your data:
+
+* Remove potentially erroneous SNPs that remain after optimizing your assembly pipeline
+* Remove SNPs that are uninformative for downstream analyses 
+* Subsample to remove SNPs that are in linkage disequilibrium (required for analyses that assume independence like PCA, Structure)
+
+Unique to RADseq approaches is the missingness of RAD data.
+
 * Don't over filter for missingness
@@
  + impute (PCA), subsample (PCA, structure, pretty much anything with SNPs), try a few different missing filters and see what impact it has on results
- + use [pcANGSD](link)
+
+<!-- use [pcANGSD](link)-->
 

### 2021-12-17 choosing-population-genetics.Rmd
Choosing a popgen approach corrections

@@
 # Population genomic sequencing approaches
-Deciding on a sequencing approach for a new project can be challenging, and depends primarily on the questions you are interested in, but also constraints set by funding, existing genomic resources, sample size, genome characteristics, and expertise in the research group.
-The main categories of population genomic sequencing approaches are:
+Deciding on a sequencing approach for a new project can be challenging.
+Your choice depends primarily on the questions of interest, but also constraints set by funding, existing genomic resources, sample size, genome characteristics and expertise in the research group.
+The main categories of population genomic sequencing approaches include (1) High coverage individual whole-genome resequencing (hcWGS) and (2) Low coverage individual WGS (lcWGS), (3).
 
-### 1) High coverage individual whole genome resequencing (hcWGS)  {.unlisted .unnumbered}
+**Coverage depth** refers to the average number of sequencing reads per region of the genome (distinct from coverage breadth, see [@sims2014sequencing]).
+With more sequencing reads at a region (higher coverage), errors that arise during sequencing have a smaller impact on accurately identifying variable sites (termed “genotyping”) [@Fuentes-Pardo2017-nu].
+
+### 1) High coverage individual whole-genome resequencing (hcWGS)  {.unlisted .unnumbered}
 **Whole-genome resequencing (WGS)** is where whole genomes of multiple individuals are sequenced and mapped to an assembled reference genome in order to compare variable sites between individuals or populations.
-**Coverage depth** refers to the average number of sequencing reads per region of the genome.
-With more sequencing reads at a region (higher coverage depth), errors that arise during sequencing have a smaller impact on accurately identifying variable sites (termed "genotyping") [@Fuentes-Pardo2017-nu].
+This approach is considered the is the 'gold standard' for DNA resequencing because it includes all variant types (i.evSNVs, indels, structural variants and CNVs) [@sims2014sequencing].
 
 ### 2) Low coverage individual WGS (lcWGS)  {.unlisted .unnumbered}
-Low coverage whole genome sequencing is where whole genomes of many individuals are sequenced, but depth of coverage so low that individual genotypes cannot be confidently assigned (cite).
-Instead of assigning specific, "hard called genotypes, lcWGS data requires probabilistic analyses that account for uncertainty about the true genotype and can incororate uncertanty accross all sequenced samples.
-Typically, sequencing depth of > 3 reads is considered "low coverage".
+Low coverage whole genome sequencing is where whole genomes of many individuals are sequenced, but depth of coverage is so low (sequencing depth < 3) that individual genotypes cannot be confidently assigned [@korneliussen2014angsd].
+Instead of assigning specific, “hard called" genotypes, lcWGS data requires probabilistic analyses that account for uncertainty about the true genotype and can incorporate uncertainty across all sequenced samples.
 
@@
 The genome fragments can be selected at random, or through targeted probe assays.
+Variations of this type of sequencing approach include:
+
 #### RADseq/Genotype-by-sequencing  {.unlisted .unnumbered}
-This very popular class of methods involve sequencing a subset of genomic regions randomly distributed throughout the genome, often using restriction enzymes (i.e., RADseq: restriction site-associated DNA).
-#### Targeted capture methods (e.g., exome capture)  {.unlisted .unnumbered}
+A class of methods involving sequencing a subset of genomic regions randomly distributed throughout the genome, often using restriction enzymes (i.e. restriction site-associated DNA (RADseq)).
+
+#### Targeted capture methods (e.g., whole-exome sequencing (WES))  {.unlisted .unnumbered}
+Sequence capture methods use a set of probes designed with a-priori knowledge to focus sequencing effort on a set of hundreds to tens of thousands of specific loci.
+Several capture approaches for non-model species can utilise other sequencing strategies to design probes corresponding to genomic regions (eg.
+transcriptome sequencing, ddRAD) (see [@jones2016targeted] for a review).
 
 ### 4) Pooled individuals (Poolseq)  {.unlisted .unnumbered}
+Sequencing pools of individuals to provide information on genome wide  population  allele  frequencies (see review in [@futschik2010next]).
 
@@
 
-It should be noted that you don't need to choose only one approach for all samples!
-If you are starting a research program on a species without any genomic resources, you may want to start with de novo RADseq to understand basic population structure and genetic diversity, then once a reference genome is assembled move to high/low coverage WGS.
-In the [MarineOmics seminar](link), Misha Matz recommended choosing one individual to use for genome assembly, doing high coverage WGS on a few other individuals that represent most of the variation in your taxa, and then WGS many other individuals at low coverage.
+It should be noted that you don’t need to choose only one approach for all samples!
+If you are starting a research program on a species without any genomic resources, you may want to start with denovo RADseq to understand basic population structure and genetic diversity, then once a reference genome is assembled move to high/low coverage WGS.In the [MarineOmics seminar](https://marineomics.github.io/panels.html), Misha Matz recommended choosing one individual to use for genome assembly, doing high coverage WGS on a few other individuals that represent most of the variation in your taxa, and then WGS many other individuals at low coverage.
 The high coverage individuals can be useful for imputing missing genotypes in the low coverage samples.
 
-This table briefly summarizes the pros and cons of five sequencing approaches and their appropriateness for answering specific questions.
-If you are interested in answering multiple questions (eg, neutral population structure and adaptive variation) then you would often want a method that can do both (eg, WGS).
+The table below briefly summarizes the pros and cons of five sequencing approaches and their appropriateness for answering specific questions.
+If you are interested in answering multiple questions (i.e.
+neutral population structure and adaptive variation) then you would often want a method that can do both (i.e.
+WGS).
 
@@
 # Recommended readings  
-Below we list a number of great review articles and empirical comparison studies that can help when making the decision. This topic was also discussed during the MarineOmics seminars (link).
+Below we list a number of great review articles and empirical comparison studies that can help when making the decision.
+This topic was also discussed during the [MarineOmics seminars](https://marineomics.github.io/panels.html).
 
@@
 
-* [@Benjelloun2019-dq]: empirical comparison of high coverage WGS, low coverage WGS, random variants (eg.
+* [@Benjelloun2019-dq]:an empirical comparison of high coverage WGS, low coverage WGS, random variants (eg.
 from RADseq), exome capture, and commercial SNP chip for mammals (genome size=2.6Gb).
-For neutral genomic diversity, found 5k-10k random variants were enough to accurately estimate, but comercial panales/exome capture had acertainment bias.
+For neutral genomic diversity, found 5k-10k random variants were enough to accurately estimate, but commercial panels/exome capture had ascertainment bias.
 To detect selection and accurate estimates of LD, at least 1M variants were required.
 For the studied species, 5x coverage was just as accurate as 12x coverage.
-Low coverage WGS (< 5x) should use genotype likelihood methods, however the rate of false positives for heterozygotes is higher than classic genotyping methods at higher coverage.
-* [@Dorant2019-fb]: Case study comparing Poolseq vs.
+Low coverage WGS (< 5x) should use genotype likelihood methods, however, the rate of false positives for heterozygotes is higher than classic genotyping methods at higher coverage.
+* [@Dorant2019-fb]:  A case study comparing Poolseq vs.
 reduced representation seq (GBS) vs Rapture (targeted capture of RAD loci) for inferring population structure with weak genetic differentiation.
-All three detected same weak pop structure pattern, but Poolseq gave Fst values 3-5x higher than Rapture/GBS.
+All three detected the same weak pop structure pattern, but Poolseq gave Fst values 3-5x higher than Rapture/GBS.
 
@@
 While we currently don't discuss library preparation methods in depth, we do include advice on how to organize your sequencing runs in order to improve quality control downstream.
-
-# References

### 2021-12-17 choosing-population-genetics.Rmd
popgen approach changes

@@
 It should be noted that you don’t need to choose only one approach for all samples!
-If you are starting a research program on a species without any genomic resources, you may want to start with denovo RADseq to understand basic population structure and genetic diversity, then once a reference genome is assembled move to high/low coverage WGS.In the [MarineOmics seminar](https://marineomics.github.io/panels.html), Misha Matz recommended choosing one individual to use for genome assembly, doing high coverage WGS on a few other individuals that represent most of the variation in your taxa, and then WGS many other individuals at low coverage.
+If you are starting a research program on a species without any genomic resources, you may want to start with denovo RADseq to understand basic population structure and genetic diversity, then once a reference genome is assembled move to high/low coverage WGS.In the [MarineOmics seminar](https://marineomics.github.io/panels.html), [Misha Matz](https://matzlab.weebly.com/) recommended choosing one individual to use for genome assembly, doing high coverage WGS on a few other individuals that represent most of the variation in your taxa, and then WGS many other individuals at low coverage.
 The high coverage individuals can be useful for imputing missing genotypes in the low coverage samples.

### 2021-12-18 choosing-population-genetics.Rmd
updating the html choosing pop gen approach

@@
 
-# Population genomic sequencing approaches
+# Population Genomic Sequencing Approaches
 Deciding on a sequencing approach for a new project can be challenging.
 Your choice depends primarily on the questions of interest, but also constraints set by funding, existing genomic resources, sample size, genome characteristics and expertise in the research group.
-The main categories of population genomic sequencing approaches include (1) High coverage individual whole-genome resequencing (hcWGS) and (2) Low coverage individual WGS (lcWGS), (3).
 
+The main categories of population genomic sequencing approaches include...
+
+## 1) Whole-genome resequencing
+Two approaches for whole-genome resequencing are distinguished by their coverage depth.
 **Coverage depth** refers to the average number of sequencing reads per region of the genome (distinct from coverage breadth, see [@sims2014sequencing]).
@@
 
-### 1) High coverage individual whole-genome resequencing (hcWGS)  {.unlisted .unnumbered}
+### High coverage individual whole-genome resequencing (hcWGS)  {.unlisted .unnumbered}
 **Whole-genome resequencing (WGS)** is where whole genomes of multiple individuals are sequenced and mapped to an assembled reference genome in order to compare variable sites between individuals or populations.
-This approach is considered the is the 'gold standard' for DNA resequencing because it includes all variant types (i.evSNVs, indels, structural variants and CNVs) [@sims2014sequencing].
+High-coverage (>20x) sequencing is able to find rare variants with high confidence.
+This approach is considered the is the 'gold standard' for DNA resequencing because it includes all variant types (i.e SNVs, indels, structural variants and CNVs) [@sims2014sequencing].
 
-### 2) Low coverage individual WGS (lcWGS)  {.unlisted .unnumbered}
-Low coverage whole genome sequencing is where whole genomes of many individuals are sequenced, but depth of coverage is so low (sequencing depth < 3) that individual genotypes cannot be confidently assigned [@korneliussen2014angsd].
-Instead of assigning specific, “hard called" genotypes, lcWGS data requires probabilistic analyses that account for uncertainty about the true genotype and can incorporate uncertainty across all sequenced samples.
+### Low coverage individual WGS (lcWGS)  {.unlisted .unnumbered}
+**Low coverage whole genome sequencing (lcWGS)** is where whole genomes of many individuals are sequenced, but depth of coverage is so low (<5× per site per individual, on average [@nielsen2011genotype]) that individual genotypes cannot be confidently assigned .
+Instead of assigning specific, “hard called" genotypes, lcWGS data requires probabilistic analyses that account for uncertainty about the true genotype and can incorporate uncertainty across all sequenced samples [@korneliussen2014angsd].
+This approach is useful in cases where the research questions of interest are targeted at the population-level (i.e.
+allele frequencies, linkage disequilibrium (LD) patterns, etc) [@Lou2021-me].
 
-### 3) Reduced-representation sequencing {.unlisted .unnumbered}
+## 2) Reduced-representation sequencing {.unlisted .unnumbered}
 As the name suggests, this approach aims to sequence a reduced portion of the genome across multiple individuals at moderate to high coverage for variant discovery and genotyping with high confidence.
 The genome fragments can be selected at random, or through targeted probe assays.
-Variations of this type of sequencing approach include:
+Variations of this type of sequencing approach include...
 
-#### RADseq/Genotype-by-sequencing  {.unlisted .unnumbered}
-A class of methods involving sequencing a subset of genomic regions randomly distributed throughout the genome, often using restriction enzymes (i.e. restriction site-associated DNA (RADseq)).
+### RADseq/Genotype-by-sequencing  {.unlisted .unnumbered}
+A class of methods involving sequencing a subset of genomic regions randomly distributed throughout the genome, often using restriction enzymes (i.e. **restriction site-associated DNA (RADseq)**).
 
-#### Targeted capture methods (e.g., whole-exome sequencing (WES))  {.unlisted .unnumbered}
+### Targeted capture methods  {.unlisted .unnumbered}
 Sequence capture methods use a set of probes designed with a-priori knowledge to focus sequencing effort on a set of hundreds to tens of thousands of specific loci.
-Several capture approaches for non-model species can utilise other sequencing strategies to design probes corresponding to genomic regions (eg.
-transcriptome sequencing, ddRAD) (see [@jones2016targeted] for a review).
+Several capture approaches for non-model species can utilise other sequencing strategies to design probes corresponding to genomic regions (i.e.
+**Whole-exome sequencing (WES)**, see [@jones2016targeted] for a review, **Expressed Exome Capture Sequencing (eecSeq)**[@puritz2018expressed]).
 
-### 4) Pooled individuals (Poolseq)  {.unlisted .unnumbered}
-Sequencing pools of individuals to provide information on genome wide  population  allele  frequencies (see review in [@futschik2010next]).
+## 3) Pooled individuals (Poolseq)  {.unlisted .unnumbered}
+Sequencing pools of individuals to provide information on genome-wide population allele frequencies (see review in [@futschik2010next]).
 
-![Figure 1 from [@Lou2021-me] comparing the distribution of sequencing reads mapped to a reference genome.](images/mec16077-fig-0001-m.jpg)
+![**Figure 1** from [@Lou2021-me] comparing the distribution of sequencing reads mapped to a reference genome.](images/mec16077-fig-0001-m.jpg)
 
@@
 While we currently don't discuss library preparation methods in depth, we do include advice on how to organize your sequencing runs in order to improve quality control downstream.
+
+References:

### 2021-12-28 choosing-population-genetics.Rmd
updating some formatting issues

@@
 High-coverage (>20x) sequencing is able to find rare variants with high confidence.
-This approach is considered the is the 'gold standard' for DNA resequencing because it includes all variant types (i.e SNVs, indels, structural variants and CNVs) [@sims2014sequencing].
+This approach is considered the 'gold standard' for DNA re-sequencing because it includes all variant types (i.e SNVs, indels, structural variants and CNVs) [@sims2014sequencing].
 
 ### Low coverage individual WGS (lcWGS)  {.unlisted .unnumbered}
-**Low coverage whole genome sequencing (lcWGS)** is where whole genomes of many individuals are sequenced, but depth of coverage is so low (<5× per site per individual, on average [@nielsen2011genotype]) that individual genotypes cannot be confidently assigned .
-Instead of assigning specific, “hard called" genotypes, lcWGS data requires probabilistic analyses that account for uncertainty about the true genotype and can incorporate uncertainty across all sequenced samples [@korneliussen2014angsd].
+**Low coverage whole genome sequencing (lcWGS)** is where whole genomes of many individuals are sequenced, but depth of coverage is so low (<5× per site per individual, on average [@nielsen2011genotype]) that individual genotypes cannot be confidently assigned.
+Instead of assigning specific (“hard called") genotypes, lcWGS data requires probabilistic analyses that account for uncertainty about the true genotype and can incorporate uncertainty across all sequenced samples [@korneliussen2014angsd].
 This approach is useful in cases where the research questions of interest are targeted at the population-level (i.e.
@@
 
-References:
+# References

### 2022-01-13 poolseq.Rmd
fixing typos

@@
 I mostly followed the PoPoolation and PoPoolation2 pipelines by Robert Kofler and Christian Schlötterer (https://sourceforge.net/p/popoolation2/wiki/Main/).
-The website has an easy to follow tutorial for using PoPoolation2 and how to use their scripts.
+The website has an easy-to-follow tutorial for using PoPoolation2 and how to use their scripts.
 This pipeline extends the use of the scripts for running on a Linux cluster.
@@
 
-The following scripts were used to analyze data from an experimental evolution study on Tigriopus californicus copepods (Griffiths JS, Kawji Y, Kelly MW.
+The following scripts were used to analyze data from an experimental evolution study on *Tigriopus californicus* copepods (Griffiths JS, Kawji Y, Kelly MW.
 2020.
@@
 The pooling of many individuals can be a cost-effective tool to character genetic variation present at the population level (e.g., genome wide association studies and experimental evolution).
-For example, to identify small changes in allele frequencies, you would neet genotype hundreds of individuals across the entire genome, which may be too cost-prohibitive with individual barcoding.
+For example, to identify small changes in allele frequencies, you would need to genotype hundreds of individuals across the entire genome, which may be too cost-prohibitive with individual barcoding.
 However, individual genotype information is lost, as well as information on dominance and
@@
 Finally, distinguishing between sequencing errors and rare, low-frequency alleles is limited.
-Unlike sequencing of individuals, this cannot be solved by analysing multiple reads from the same region of a single chromosome.
+Unlike sequencing of individuals, this cannot be solved by analyzing multiple reads from the same region of a single chromosome.
 One way to ameliorate this issue is to sequencing replicates of pools.
 In addition, Pool-seq analyses can be prone to inaccurate estimations of allele frequencies based on the experimental methods.
-Care must be taken to ensure the the experimentla design maximizes the ratio sequencing coverage to the number of individuals pooled.
+Care must be taken to ensure the the experimental design maximizes the ratio sequencing coverage to the number of individuals pooled.
 
@@
 For example, smaller differences in allele frequency changes can be seen if you have a large number of individuals in a single pool and more coverage.
-The PoPoolation2 Manual page provides the following calculation to detect significant allele frequency differences: "To detect a singificant (Fisher's exact test; p=0.01) allele frequency difference between two populations of 30%, a coverage (and pool size) of 50 will be sufficient.
+The PoPoolation2 Manual page provides the following calculation to detect significant allele frequency differences: "To detect a significant (Fisher's exact test; p=0.01) allele frequency difference between two populations of 30%, a coverage (and pool size) of 50 will be sufficient.
 If however significant (Fisher's exact test; p=0.01) allele frequency differences of 10% need to be detected the coverage as well as the pool size should be about 400."
@@
 
-Customizeable Filtering and Assessing Quality of Pool-Seq Data:
+Customizable Filtering and Assessing Quality of Pool-Seq Data:
 https://github.com/ToBoDev/assessPool
@@
 Since I had high coverage and long read lengths (150 bp), I used a minimum read length of 40 bp.
-The --max_n flag specificies how many "Ns" can be in the read before the read if discarded as "poor quality".
+The --max_n flag specifies how many "Ns" can be in the read before the read if discarded as "poor quality".
 Again, depending on your sequencing depth you may want to be more or less conservative.
 There is no default value specified in TrimGalore, but I used a highly conservative value of 1 (ie any instances of N in the read and that read will be discarded).
-This is probably more conservative than you need to be, becasue the read may still contain infromative sequencign data.
+This is probably more conservative than you need to be, because the read may still contain informative sequencing data.
 Please note that these parameters are not specific to a Pool Seq dataset.
@@
 ## Pop Gen and Downstream Statistical Analyses
-These in depth notes are not on the tutorial or manual page, but detailed descriptions for each flag can be found using the --help option when running each script. e.g. perl fst-sliding.pl --help
+These in-depth notes are not on the tutorial or manual page, but detailed descriptions for each flag can be found using the --help option when running each script. e.g. perl fst-sliding.pl --help
 
@@
 The minimum count must be less than the minimum coverage.
-You may want to modify this flag based on the number of inidividuals pooled per sample/population.
+You may want to modify this flag based on the number of individuals pooled per sample/population.
 
@@
 When genotyping individually barcoded samples, max coverage requirements are generally not recommended.
-However, pool-seq analyses require a max coverage for calculating Fst, cmh, allele frequency changes, etc, because allele frequency estimates are based on the number allele count, which are prone to PCR duplicates.
+However, pool-seq analyses require a max coverage for calculating Fst, cmh, allele frequency changes, etc.
+because allele frequency estimates are based on the number allele count, which are prone to PCR duplicates.
 PCR duplicates from a single individual in the pool may skew allele frequency estimates.
@@
 --min-count: the minimum count of the minor allele, used for SNP identification.
-SNPs will be identified considering all samples/populations simultanously.
+SNPs will be identified considering all samples/populations simultaneously.
 default=2.
@@
 Options include multiple, geometric mean, or median.
-The summary of p-values within a window is dependnt on the experimental design and how recent selection was expected to occur.
-For example, the default parameter might be appropriate if your treatment comparisons are geographically separated populations, where selection is expect to have occurred over hundreds of thousands of years.
+The summary of p-values within a window is dependent on the experimental design and how recent selection was expected to occur.
+For example, the default parameter might be appropriate if your treatment comparisons are geographically separated populations, where selection is expected to have occurred over hundreds of thousands of years.
 However, this might be inappropriate for short-term experimental evolution studies, such as the example dataset here.
-In this dataset, copepods were selected in the lab for 15-20 generations, therefore, we expect relatively large linkage blocks that prevent us from idenitfying the true targets of selection, and therefore I use the geometric mean option.
+In this dataset, copepods were selected in the lab for 15-20 generations, therefore, we expect relatively large linkage blocks that prevent us from identifying the true targets of selection, and therefore I use the geometric mean option.
 
@@
 This script will detect consistent allele frequency changes in biological replicates.
-In my case, I was comparing selected samples to one control sample, but if you have replicates of treatments and controls, you can compare columns 1-2, 3-4, etc…
+In my case, I was comparing selected samples to one control sample, but if you have replicates of treatments and controls, you can compare columns 1-2, 3-4, etc.
 
@@
 ## Nucleotide Diversity
-Calulate Pi (nucleotide diversity) for individual samples. Use mpileup files that were created for a single sample as the input. Below is an example for one sample.
+Calculate Pi (nucleotide diversity) for individual samples. Use mpileup files that were created for a single sample as the input. Below is an example for one sample.
 ```{bash eval=FALSE}

### 2022-01-14 poolseq.Rmd
formatting changes through line 39

@@
 for the file Bodega population file is SRR10760095, identified under the "Run" heading).
-Then click on the Data access tab and there will be a downloadable link.
+Then click on the "Data access" tab, and there will be a downloadable link.
 Additional analyses and scripts can be found at: https://github.com/JoannaGriffiths/Tigriopus_HER
 
-***Strengths and weaknesses of Pool-seq***
+***Strengths and weaknesses of Pool-seq***\
 Due to the nature of Pool-seq where multiple individuals are pooled to prepare a single library, this results in a loss of haplotype information.
@@
 PoPoolation2 is primarily used for comparing allele frequency differences between populations or treatments in comparison to a reference genome.
-The pooling of many individuals can be a cost-effective tool to character genetic variation present at the population level (e.g., genome wide association studies and experimental evolution).
+The pooling of many individuals can be a cost-effective tool to characterize genetic variation present at the population level (e.g., genome wide association studies and experimental evolution).
 For example, to identify small changes in allele frequencies, you would need to genotype hundreds of individuals across the entire genome, which may be too cost-prohibitive with individual barcoding.
@@
 Unlike sequencing of individuals, this cannot be solved by analyzing multiple reads from the same region of a single chromosome.
-One way to ameliorate this issue is to sequencing replicates of pools.
+One way to ameliorate this issue is to sequence replicates of pools.
 In addition, Pool-seq analyses can be prone to inaccurate estimations of allele frequencies based on the experimental methods.
@@
 
-***Notes on Experimental Design for Downstream Analyses:***
+***Notes on Experimental Design for Downstream Analyses:***\
 It is important to try to standardize the number of individuals (and thus DNA) pooled for each sample.
@@
 For example, smaller differences in allele frequency changes can be seen if you have a large number of individuals in a single pool and more coverage.
-The PoPoolation2 Manual page provides the following calculation to detect significant allele frequency differences: "To detect a significant (Fisher's exact test; p=0.01) allele frequency difference between two populations of 30%, a coverage (and pool size) of 50 will be sufficient.
-If however significant (Fisher's exact test; p=0.01) allele frequency differences of 10% need to be detected the coverage as well as the pool size should be about 400."
-The sequencing accuracy of Pool-seq increases with larger pool sizes; if the number of sequenced individuals is approximately equal to the coverage, then the sequency accuracy is no longer superior to an individual level of sequencing (see Fig.
+The PoPoolation2 Manual page provides the following calculation to detect significant allele frequency differences:\"To detect a significant (Fisher's exact test; p=0.01) allele frequency difference between two populations of 30%, a coverage (and pool size) of 50 will be sufficient.
+If however significant (Fisher's exact test; p=0.01) allele frequency differences of 10% need to be detected the coverage as well as the pool size should be about 400."\
+It is critical to note, the sequencing accuracy of Pool-seq increases with larger pool sizes; if the number of sequenced individuals is approximately equal to the coverage, then the sequency accuracy is no longer superior to an individual level of sequencing (see Fig.
 1a from Schlötterer et al., 2014).
@@
 
-***Alternative Pipelines:***
-Low Coverage Pool-seq Data: 
-https://github.com/petrov-lab/HAFpipe-line
+***Alternative Pipelines***\
+Low Coverage Pool-seq Data:\ 
+https://github.com/petrov-lab/HAFpipe-line \
 Alternative pipelines have been created for low coverage (<5x) pool-seq data that can provide accurate estimates of allele frequencies.
@@
 
-Customizable Filtering and Assessing Quality of Pool-Seq Data:
-https://github.com/ToBoDev/assessPool
+Customizable Filtering and Assessing Quality of Pool-Seq Data:\
+https://github.com/ToBoDev/assessPool \
 This pipeline filters SNPs based on adjustable criterion with suggestions for pooled data.

### 2022-01-14 poolseq.Rmd
typos + formatting

@@
 ## Filter Raw Sequences
-This script uses the wrapper TrimGalore to remove Adapter/Barcode sequences and removes poor quality reads. 
+The script below uses the wrapper TrimGalore to remove Adapter/Barcode sequences and removes poor quality reads. 
 This example dataset uses paired end sequencing so I used the flag (--paired) in TrimGalore. Because these example files had high coverage, I was able to use conservative filters. 
-For example, the default length of reads is 20 bp (so if you read is less than 20 bp then the read will be removed).
+For example, the default length of reads is 20 bp (so if a read is less than 20 bp then the read will be removed).
 Since I had high coverage and long read lengths (150 bp), I used a minimum read length of 40 bp.
@@
 This sync file format is required for downstream analyses such as Fst, creating Manhattan plots, or CMH tests to be used in both the PoPoolation and PoPoolation2 pipelines.
-You will want to make sure that the fastq-type flag is set to sanger (this is for Phred quality score matching--yes, this is correct even if you used Illumina sequencing).
+You will want to make sure that the --fastq-type flag is set to "sanger" (this is for Phred quality score matching--yes, this is correct even if you used Illumina sequencing).
 The min-qual flag must be above 0, the default is 20 which gives you a 99% base call accuracy rate).
 
-What is an Mpileup file?
+What is an Mpileup file?\
 Pileup format is a text-based format for summarizing the base calls of aligned reads to a reference sequence.
@@
 You will want to play around with the min-coverage, max-coverage, and min-count based on the mean coverage of your samples/populations.
-Minimum coverages that are too high and max coverages that are too low may result in an empty output file, which is why its a good idea to have an estimate of your mean/min/max coverage for each sample/population.
+Minimum coverages that are too high and max coverages that are too low may result in an empty output file, which is why it's a good idea to have an estimate of your mean/min/max coverage for each sample/population.
 The minimum count must be less than the minimum coverage.
@@
 
-***Important considerations:***
+***Important considerations:***\
 --suppress-noninformative: Output files will not report results for windows with no SNPs or insufficient coverage.
@@
 
- '500' a maximum coverage of 500 will be used for all samples/populations
- '300,400,500' a maximum coverage of 300 will be used for the first sample/population, a maximum coverage of 400 for the second sample/population and so on
+ '500' a maximum coverage of 500 will be used for all samples/populations\
+ '300,400,500' a maximum coverage of 300 will be used for the first sample/population, a maximum coverage of 400 for the second sample/population and so on\
  '2%' the 2% highest coverages will be ignored, this value is independently estimated for every sample/population
 
-***NOTE ABOUT MAX COVERAGE***
+***NOTE ABOUT MAX COVERAGE***\
 When genotyping individually barcoded samples, max coverage requirements are generally not recommended.
@@
 This script will calculate significant allele frequency differences for each SNP position between two samples. Results can be plotted in a Manhattan plot. 
-Depending on the experimental method, the --window-summary-method flag can be used to determine pvalue for windows of SNPs.
-The default is to multiply the pvalue of all SNPs for the window.
-Alternatively, you can specify how pvalues within a window should be summarized.
+Depending on the experimental method, the --window-summary-method flag can be used to determine p-value for windows of SNPs.
+The default is to multiply the p-value of all SNPs for the window.
+Alternatively, you can specify how p-values within a window should be summarized.
 Options include multiple, geometric mean, or median.
@@
 Due to constraints of the test, you can only compare allele frequency changes for individual SNPs and not windows.
-In my data, I expected high levels of linkage disequilibrium in my samples because individuals were exposed to strong selection pressure.
+I expected high levels of linkage disequilibrium in my samples because individuals were exposed to strong selection pressure.
 Depending on the amount of linkage disequilibrium you expect in your populations, you may want to account for the non-independence of SNPs following the CMH test.
@@
 
-cmh_p <-read.table("Tig.cmh", header=TRUE) #read in the original output from the CMH test, because SeqMonk output discards the important pvalue column.
+cmh_p <-read.table("Tig.cmh", header=TRUE) #read in the original output from the CMH test, because SeqMonk output discards the important p-value column.
 colnames(cmh_p) <- c("CHR", "BP", "Allele", "1S", "2S", "3S", "4S", "5S", "U", "p") #rename columns based on sample names

### 2022-01-14 poolseq.Rmd
formatting

@@
 
-What is an Mpileup file?\
+**What is an Mpileup file?**\
 Pileup format is a text-based format for summarizing the base calls of aligned reads to a reference sequence.
@@
 ## Pop Gen and Downstream Statistical Analyses
-These in-depth notes are not on the tutorial or manual page, but detailed descriptions for each flag can be found using the --help option when running each script. e.g. perl fst-sliding.pl --help
+These in-depth notes are not on the tutorial or manual page, but detailed descriptions for each flag can be found using the `--help` option when running each script. e.g. `perl fst-sliding.pl --help`
 
@@
 ***Important considerations:***\
---suppress-noninformative: Output files will not report results for windows with no SNPs or insufficient coverage.
+`--suppress-noninformative`: Output files will not report results for windows with no SNPs or insufficient coverage.
 Including this flag will make it easier for downstream manipulation of the output file, such as creating figures.
 
---pool-size: the number of individuals pooled per sample; If the pool sizes differ among each sample, you can provide the size for each sample individually:
+`--pool-size`: the number of individuals pooled per sample; If the pool sizes differ among each sample, you can provide the size for each sample individually:
     --pool-size 500 .. all samples/populations have a pool size of 500
@@
 
---min-coverage: the minimum coverage used for SNP identification, the coverage in ALL samples/populations has to be     higher or equal to this threshold, otherwise no SNP will be called. default=4
+`--min-coverage`: the minimum coverage used for SNP identification, the coverage in ALL samples/populations has to be     higher or equal to this threshold, otherwise no SNP will be called. default=4
 
---max-coverage: The maximum coverage; All populations are required to have coverages lower or equal than the maximum coverage; Mandatory
-The maximum coverage may be provided as one of the following:
-
- '500' a maximum coverage of 500 will be used for all samples/populations\
- '300,400,500' a maximum coverage of 300 will be used for the first sample/population, a maximum coverage of 400 for the second sample/population and so on\
- '2%' the 2% highest coverages will be ignored, this value is independently estimated for every sample/population
+`--max-coverage`: The maximum coverage; All populations are required to have coverages lower or equal than the maximum coverage; Mandatory.\
+The maximum coverage may be provided as one of the following:\
+`500` a maximum coverage of 500 will be used for all samples/populations\
+`300,400,500` a maximum coverage of 300 will be used for the first sample/population, a maximum coverage of 400 for the second sample/population and so on\
+`2%` the 2% highest coverages will be ignored, this value is independently estimated for every sample/population
 
@@
 
---min-covered-fraction:the minimum fraction of a window being between min-coverage and max-coverage in ALL                    samples/populations; float; default=0.0   
+`--min-covered-fraction`:the minimum fraction of a window being between min-coverage and max-coverage in ALL                    samples/populations; float; default=0.0   
     The tutorial provides an example script with a min-covered fraction of 1, however, you will most likely run into an       error or an empty output file.
@@
 
---min-count: the minimum count of the minor allele, used for SNP identification.
+`--min-count`: the minimum count of the minor allele, used for SNP identification.
 SNPs will be identified considering all samples/populations simultaneously.
@@
 This script will calculate significant allele frequency differences for each SNP position between two samples. Results can be plotted in a Manhattan plot. 
-Depending on the experimental method, the --window-summary-method flag can be used to determine p-value for windows of SNPs.
+Depending on the experimental method, the `--window-summary-method` flag can be used to determine p-value for windows of SNPs.
 The default is to multiply the p-value of all SNPs for the window.

### 2022-01-20 WGS_intro.Rmd

@@
 library(knitcitations)
-opts_chunk$set(fig.width = 10,
+knitr::opts_chunk$set(fig.width = 10,
                fig.height = 5,
@@
 
-For the purposes of this guide, we use "whole genome (re)sequencing (WGS)) to refer to  methods where a reference genome exists (whether for the focal species or a related species), and samples are sequenced and then mapped to most or all of the reference genome.
+<br>
+
+# Introduction
+
+For the purpose of this guide, we use "whole genome (re)sequencing (WGS)) to refer to methods where a reference genome already exists (whether for the focal species or a related species), and uniquely barcoded samples are sequenced and then mapped to most or all of the reference genome.
 This method can provide high density genetic variants (e.g., SNPs) and structural information for population genomic analyses, while also facilitating functional insights if the reference genome is functionally annotated.
-This is separate from whole-genome *de novo* sequencing, which aims to produce a reference genome by sequencing and assembling a complete genome for the first time.
+This is separate from whole-genome *de novo* sequencing, which aims to produce a reference genome by sequencing and assembling a complete genome for a species for the first time.
 
-An excellent and thorough review of the methodology and consideration for WGS approaches, especially as they apply to nonmodel organisms, is [@Fuentes-Pardo2017-nu].
+An excellent and thorough review of the methodology and consideration for WGS approaches, especially as they apply to non-model organisms, is [[@Fuentes-Pardo2017-nu]](https://doi.org/10.1111/mec.14264).
+But briefly, current WGS approaches can be broadly categorized into two types: high-to-moderate-coverage WGS, and low-coverage WGS.
+Their major distinction is that with high-to-moderate-coverage WGS, each individual is sequenced at a depth with which genotype can be confidently called at most sites, whereas with low-coverage WGS, each individual is sequenced at a depth too low to call genotype with, and downstream analyses should take such genotype uncertainties into acount.
 
-## Tutorials
+However, the line between high-to-moderate-coverage WGS and low-coverage WGS is not always as clear-cut as presented above.
+For example, with moderate-coverage WGS (e.g.
+5-20x), many sites within an individual can still have low coverage due to random sampling, resulting in unreliable genotype calls and/or missing data that can become problematic in downstream analysis.
+Therefore, it could be preferrable to avoid hard-calling genotypes with moderate-coverage WGS in certain applications.
+On the other hand, in populations with high levels of linkage disequilibrium (LD), it could be possible to leverage LD to carry out genotype imputation, making genotype calling a lot more accurate with low-coverage WGS data.
+Imputation is more likely to be successful when a high-quality reference panel exists in the system ([Fuller et al.
+2020](https://www.science.org/doi/10.1126/science.aba4674), [Rubinacci et al.
+2021](https://www.nature.com/articles/s41588-020-00756-0)), but methods for imputation without such reference panels have also been developed (in which case a very large sample size would be required, [Davis et al.
+2016](https://www.nature.com/articles/ng.3594)).
+
+This said, here are the main differences between these two approaches in practice: with same sample size, high-to-moderate-coverage WGS tends to provide higher resolution data that are more versatile and less susceptible to technical artefacts (especially those caused by sequencing errors) when compared with low-coverage WGS, but it could be a lot more costly.
+Low-coverage WGS, in contrast, can be used to achieve higher sample size with a fixed budget, which can then contribute to higher-resolution population-level inferences, but it does require a different computational toolbox that takes genotype uncertainties into account, and is thus constrained by limitations of the current toolbox (see Section 6 in [[@Lou2021-me]](https://onlinelibrary.wiley.com/doi/abs/10.1111/mec.16077) for a more detailed discussion on this).
+
+<br>
+
+# Tutorials
 
 Here we provide a couple different tutorials by our working group members, as well as links to other detailed tutorials on the web.
-The goal of these MarineOmics tutorials are to provide extensive detail on "why" certain parameters are chosen, and some guidance on how to evaluate different parameter options to fit your data.
+The goal of these MarineOmics tutorials are to provide extensive details on "why" certain parameters are chosen, and some guidance on how to evaluate different parameter options to fit your data.
 
-### High to moderate coverage WGS
+<br>
 
-* [Fastq-to-VCF SnakeMate pipeline](fastq2vcf_pipeline.html): In-depth explanation of automated short-read mapping and variant calling pipeline maintained by [Harvard Informatics](https://github.com/harvardinformatics/shortRead_mapping_variantCalling).
-Useful for moderate to high coverage data, where *add description of why one would use pipeline*
+## High to moderate coverage WGS
+
+* [Fastq-to-VCF SnakeMake pipeline](fastq2vcf_pipeline.html): in-depth explanation of an automated short-read mapping and variant calling pipeline maintained by [Harvard Informatics](https://github.com/harvardinformatics/shortRead_mapping_variantCalling).
+Useful if you plan to adopt this pre-packaged automated and parallelizable pipeline, would like to understand its different components but not necessarily change it substantially.
+
 * [Fastq-to-VCF workflow](WGS-Full-Walkthrough.html): detailed walkthrough of processing 15x sequencing depth WGS data for cod, from raw reads to a VCF.
+Useful if you would like to run each component of the pipeline yourself and potentially tweak some of them for your own purpose.
+In other words, you can more easily add, skip, or change parts of this pipeline, but will lose the convenience offered by an automated pipeline.
 
-### Low coverage WGS  
-Check out [@Lou2021-me](https://onlinelibrary.wiley.com/doi/abs/10.1111/mec.16077) and the associated [Github repository](https://github.com/therkildsen-lab/genomic-data-analysis/blob/master/lcwgs_data_analysis.md) for a great primer on lcWGS analysis.
+<br>
+
+## Low coverage WGS
+
+The quality control and read alignment part of the pipeline for high-to-moderate-coverage WGS also applies for low-coverage WGS.
+Therefore, the two tutorials for high-to-moderate-coverage WGS are also useful for low-coverage WGS until the point where variants and genotypes are called.
+In addition to these, here are some resources specifically designed for low-coverage WGS.
+
+* [Low-coverage WGS tutorial](https://github.com/nt246/lcwgs-guide-tutorial): a tutorial for the processing and analysis of low-coverage WGS data (i.e.
+from raw fastq files to population genomic inference), with example datasets and hands-on exercises.
+It is associated with the paper [[@Lou2021-me]](https://onlinelibrary.wiley.com/doi/abs/10.1111/mec.16077).
+
+* [Detection and mitigation of batch effects](https://github.com/therkildsen-lab/batch-effect/blob/main/tutorial/tutorial.md): a tutorial for the detection and mitigation of batch effects with low-coverage WGS data, with example datasets and hands-on exercises.
+It is associated with the paper [(Lou and Therkildsen 2021)](https://onlinelibrary.wiley.com/doi/abs/10.1111/1755-0998.13559).
+
+* [Low-coverage WGS data analysis pipeline](https://github.com/therkildsen-lab/genomic-data-analysis/blob/master/lcwgs_data_analysis.md): a collection of scripts for the efficient and reproducible analysis of low-coverage WGS data (i.e.
+from bam to population genomic inference).
+
+* [Low-coverage WGS data processing pipeline](https://github.com/therkildsen-lab/data-processing/blob/master/lcwgs_data_processing.md): a collection of script for the efficient and reproducible processing of low-coverage WGS data (i.e.
+from raw fastq to bam).
+This pipeline should also be compatible with high-to-moderate-coverage WGS data.
+
+<br>
 

### 2022-01-20 WGS_intro.Rmd
added authors
Added bibtex refs

@@
 title:  Whole Genome Resequencing for Population Genomics 
+subtitle: '*Katherine Silliman, Nicolas Lou*'
 ---
@@
 On the other hand, in populations with high levels of linkage disequilibrium (LD), it could be possible to leverage LD to carry out genotype imputation, making genotype calling a lot more accurate with low-coverage WGS data.
-Imputation is more likely to be successful when a high-quality reference panel exists in the system ([Fuller et al.
-2020](https://www.science.org/doi/10.1126/science.aba4674), [Rubinacci et al.
-2021](https://www.nature.com/articles/s41588-020-00756-0)), but methods for imputation without such reference panels have also been developed (in which case a very large sample size would be required, [Davis et al.
-2016](https://www.nature.com/articles/ng.3594)).
+Imputation is more likely to be successful when a high-quality reference panel exists in the system ([@Fuller2020](https://www.science.org/doi/10.1126/science.aba4674), [@Rubinacci2021-dm](https://www.nature.com/articles/s41588-020-00756-0)), but methods for imputation without such reference panels have also been developed (in which case a very large sample size would be required, [@Davies2016-rq](https://www.nature.com/articles/ng.3594)).
 

### 2022-01-28 RADseq.Rmd
added radseq ref

@@
 
-If you would like to run the R code examples that are scattered throughout the guide (recommended but not required!), you will need to install some R packages:  
+If you would like to run the R code examples that are scattered throughout the guide (recommended but not required!), you will need to install some R packages. Only need to run this code once:  
 ```{r, eval=F}
@@
 ```{r}
-library(ggplot2)
-library(SeqArray)
-library(tidyverse)
-library(SNPRelate)
+library(SeqArray) # efficient storage and filtering of genomic data
+library(tidyverse) # plotting data formatting and manipulation
+library(SNPRelate) # PCA and other popgen analyses
 ```
 
-We also provide alternative coding examples that are based only on the command line, requiring these software packages:
+We also provide optional alternative coding examples that are based only on the command line, requiring these software packages:
 
@@
 "Restriction-site   Associated   DNA   sequencing"  - RADseq - combines  restriction  enzymes  with next-gen, massively parallel,  short-read  sequencing.
-RADseq involves the use of restriction enzymes which are used to shear DNA at restriction enzyme cutsite.
+RADseq involves the use of restriction enzymes, which are used to shear DNA at restriction enzyme cutsites.
 RADseq comes in different flavors.
-Double-digest RADseq (ddRAD; Peterson et al.,2012)  selects  markers  with  two  restriction enzymes  with  different  cut  frequencies.
+Double-digest RADseq (ddRAD; [@Peterson2012-tr])  selects  markers  with  two  restriction enzymes  with  different  cut  frequencies.
 Genotype-by-Sequencing  (GBS;  [@Elshire2011-or])  uses  a  frequent-cutting  restriction  enzyme  with PCR size selection.
-There are a number of reviews comparing different RADseq.GBS methods (e.g., [@Andrews2016-ay]).
+There are a number of reviews comparing different RADseq and GBS methods (e.g., [@Andrews2016-ay]).
 In this guide we use the term "RADseq" to refer to any of these protocols, including those that don't involve random shearing of data.
@@
 
-Using RADseq to generate single nucleotide polymorphisms (SNPs) involves 1) library preparation in the lab, 2) bioinformatic processing through assembly and/or mapping to a reference, then 3) filtering of SNPs and individuals for quality.
-All of these steps can (and will) introduce some error, so the goal is to minimize this error through mitigation steps at all three parts of the process.
-Every dataset is different [add more about philosophy over best practices].
-Much of this guide is directly inspired by [this excellent review paper](https://onlinelibrary.wiley.com/doi/10.1111/mec.14792) by [@OLeary2018-iy], especially the section on minimizing errors due to library prep.
+Using RADseq to generate single nucleotide polymorphisms (SNPs) involves:
+
+1. library preparation in the lab
+2. bioinformatic processing through assembly and/or mapping to a reference, then 
+3. filtering of SNPs and individuals for quality.
+
+All of these steps can (and will) introduce some error, so the goal is to minimize this error through mitigation steps at all three parts of the process. Every dataset is different!  
+<!--add more about philosophy over best practices-->
+
+Much of this guide is directly inspired by [this excellent review paper](https://onlinelibrary.wiley.com/doi/10.1111/mec.14792) [@OLeary2018-iy], especially the section on minimizing errors due to library prep.
 We recommend reading and cross-referencing with this paper, and citing it if you follow its suggestions.
@@
 * Randomize samples across library prep batches and sequencing lanes!
-For example, if you are sequencing two different groups of samples on two different sequencing lanes, make sure they are randomized in respect to sample location or whatever your groups of interest are.
+For example, if you are sequencing two different groups of samples on two different sequencing lanes, make sure they are randomized with respect to sample location or whatever your groups of interest are.
 If you are preparing groups of individuals in different batches to be pooled later, randomize across these batches.
@@
 This is to allow you to control for potential [batch effects](#batch) that are often observed with RAD data.
+
+<!--ADW: add figure showing random plate design-->
 ```{r, eval = F, include=FALSE}
@@
 * Have a core set of 2-4 technical sample replicates across all libraries.
-Ideally these are true technical replicates from library preparation to sequencing, but even just sequencing replicates can be useful for downstream quality control.
+Ideally these are true technical replicates, meaning the same tissue/DNA sample is processed multiple times all the way from library preparation to sequencing.
+Sequencing replicates, where you sequence the same sample library preparation multiple times, can also be useful for downstream quality control.
+
+<!--ADW: add figure or text to define sequencing replicates vs technical replicates-->
 
@@
 8) Repeat as needed.
+
+1) [Look at your raw data.](#fastqc)
+2) [Run an assembly pipeline](#assembly), using either all samples or a representative subset.
+3) [Evaluate potential sources of error.](#error)
+4) [Remove individuals based on initial evaluation](#bad).  
+5) [Test a range of key parameters](#test) to optimize your assembly.
+6) Run your optimized assembly on all non-removed samples.
+7) [Filter your SNPs](#filter)
+8) Repeat as needed!
 
@@
 
-## Evaluate potential confounding factors {#confound}
+## Evaluate potential sources of error {#error}
 
@@
 
-## Filtering SNPs
+## Filtering SNPs {#filter}
 ### (and embrace the missingness!) {.unlisted .unnumbered}

### 2022-01-28 poolseq.Rmd
fixing Refs

@@
 date: "8/12/2021"
+bibliography: RCN_site-Nov05.bib
 #output:
@@
 
-The following scripts were used to analyze data from an experimental evolution study on *Tigriopus californicus* copepods (Griffiths JS, Kawji Y, Kelly MW.
-2020.
-An experimental test of adaptive introgression in locally adapted populations of splash pool copepods.
-Mol Biol and Evol.
-38(4):1306-1316.
-https://doi.org/10.1093/molbev/msaa289).
+The following scripts were used to analyze data from an experimental evolution study on *Tigriopus californicus* copepods (Griffiths 2020).
 Raw sequencing data are deposited in NCBI's Short Reads Archive (https://www.ncbi.nlm.nih.gov/bioproject/PRJNA597336/).
@@
 1a from Schlötterer et al., 2014).
-
-Reference: Schlötterer C, Tobler R, Kofler R, Nolte V.
-2014 Sequencing pools of individuals-mining genome-wide polymorphism data without big funding.
-Nat.
-Rev.
-Genet.
-15, 749–763.
-(doi:10.1038/nrg3803)
 
@@
 ```
+
+## References

### 2022-01-29 poolseq.Rmd
add refs
adding figures and examples of output files

@@
 bibliography: RCN_site-Nov05.bib
-#output:
-#  html_document: default
-#  pdf_document: default
-#  word_document: default
 ---
@@
 
+Sample of a synchronized file:
+
+```{bash eval = FALSE}
+2  2302    N   0:7:0:0:0:0 0:7:0:0:0:0
+2  2303    N   0:8:0:0:0:0 0:8:0:0:0:0
+2  2304    N   0:0:9:0:0:0 0:0:9:0:0:0
+2  2305    N   1:0:9:0:0:0 0:0:9:1:0:0
+```
+
+* col1: reference contig
+* col2: position within the refernce contig
+* col3: reference character
+* col4: allele frequencies of population number 1
+* col5: allele frequencies of population number 2
+* coln: allele frequencies of population number n
+
+The allele frequencies are in the format A:T:C:G:N:del, i.e: count of bases 'A', count of bases 'T',... and deletion count in the end (character '*' in the mpileup)
+
 ## Pop Gen and Downstream Statistical Analyses
@@
 
+The following R script can be used to plot Fst results in a Manhattan plot
+```{r eval = FALSE}
+
+library("qqman")
+library("DataCombine")
+
+fst = read.table("all_PE_w10000_step10000.fst", header = T) 
+fst <- fst[c("CHR", "BP", "X1.11", "X1.12", "X2.11", "X2.12", "X3.11", "X3.12", "X4.11", "X4.12", "X5.11", "X5.12", "X6.11", "X6.12", "X7.11", "X7.12", "X8.11", "X8.12", "X9.11", "X9.12", "X10.11", "X10.12", "X11.12")] #rename column headers
+fst$ID <- paste(fst$CHR,fst$BP,sep=".")
+
+#tidying file
+Replaces <- data.frame(from=c("Chromosome_"), to=c(""))
+fst2 <- FindReplace(data = fst, Var = "CHR", replaceData = Replaces,
+                       from = "from", to = "to", exact = FALSE)
+fst2$CHR <- as.numeric(fst2$CHR)
+
+manhattan(fst2, chr="CHR", bp="BP", snp="ID", p="X1.11", suggestiveline = F, genomewideline = F, logp = F, ylim =c(0,1)) #p here refers to pvalue or in this case the Fst value, substitute with which column comparison you are interested in graphing
+
+```
+![Manhattan plot of mean Fst values for 10,000-bp windows of all the selected lines versus the pure Bodega population.](images/Fst_SvsBR_SDmap.png)
+
 ## Calculate Fisher's Exact Test
@@
 Alternatively, you can specify how p-values within a window should be summarized.
-Options include multiple, geometric mean, or median.
+Options include multiply, geometric mean, or median.
 The summary of p-values within a window is dependent on the experimental design and how recent selection was expected to occur.
@@
 
-#You may need to run this on the command line to install twotailed perl Module before running Fisher-test.pl: 
-perl -MCPAN -e 'install Text::NSP::Measures::2D::Fisher2::twotailed', see also LSU HPC for details: http://www.hpc.lsu.edu/docs/faq/installation-details.php
-
 perl /work/jgrif61/Tigs/popoolation2_1201/fisher-test.pl --input all_indiv.sync --output all_indiv.fet --suppress-noninformative --min-count 2 --min-coverage 20 --max-coverage 200 --min-covered-fraction 0.2 --window-summary-method geometricmean
@@
 ## Calculate Exact Allele Frequencies in Samples
+Exact allele frequencies can be used to determine differences in allele frequencies among populations or before and after a selection experiment, for example.
+These result could be plotted in a PCA or Manhattan plot.
 ```{bash eval=FALSE}
@@
 ```
+This script creates two output files having two different extensions:
+
+`_rc`: this file contains the major and minor alleles for every SNP in a concise format
+`_pwc`: this file contains the differences in allele frequencies for every pairwise comparision of the populations present in the synchronized file
+For details see the man pages of the script
+The allele frequency differences can be found in the `_pwc` file, a small sample:
+
+```{bash eval=FALSE}
+##chr   pos     rc      allele_count    allele_states   deletion_sum    snp_type        most_variable_allele    diff:1-2
+2      4459    N       2       C/T     0       pop     T       0.133
+2      9728    N       2       T/C     0       pop     T       0.116
+```
+
+The last column contains the obtained differences in allele frequencies for the allele provided in column 8.
+Note that in this example the last column refers to a pairwise comparison between population 1 vs 2, in case several populations are provided all pairwise comparisons will be appended in additional columns.
 
@@
 ```
+I then analyzed the results of the CMH test by plotting the geometric mean for each 10,000bp window on a Manhattan plot.
 
@@
 
+I used nucleotide diversity for creating a PCA plot to compare genetic diversity among my samples and to look for broad patterns across my data and I ran an Adonis test to determine if there were significant differences among groups (in this example, selected vs control lines)
+```{r eval=FALSE}
+
+#import each pi file and do some cleanup:
+pi_1S <- read.delim("1S.pi", header = F)
+colnames(pi_1S) <- c("Chromosome", "window", "num.snps", "frac", "pi_1S")
+pi_1S$ID <- paste(pi_1S$Chromosome, pi_1S$window, sep = '_')
+pi_1S <- pi_1S[!pi_1S$pi_1S == "na",] #remove rows with na
+pi_1S <- pi_1S[,c(6,5)]
+
+pi_1U <- read.delim("1U.pi", header = F)
+colnames(pi_1U) <- c("Chromosome", "window", "num.snps", "frac", "pi_1U")
+pi_1U$ID <- paste(pi_1U$Chromosome, pi_1U$window, sep = '_')
+pi_1U <- pi_1U[!pi_1U$pi_1U == "na",]
+pi_1U <- pi_1U[,c(6,5)]
+
+pi_2S <- read.delim("2S.pi", header = F)
+colnames(pi_2S) <- c("Chromosome", "window", "num.snps", "frac", "pi_2S")
+pi_2S$ID <- paste(pi_2S$Chromosome, pi_2S$window, sep = '_')
+pi_2S <- pi_2S[!pi_2S$pi_2S == "na",] #remove rows with na
+pi_2S <- pi_2S[,c(6,5)]
+
+pi_2U <- read.delim("2U.pi", header = F)
+colnames(pi_2U) <- c("Chromosome", "window", "num.snps", "frac", "pi_2U")
+pi_2U$ID <- paste(pi_2U$Chromosome, pi_2U$window, sep = '_')
+pi_2U <- pi_2U[!pi_2U$pi_2U == "na",]
+pi_2U <- pi_2U[,c(6,5)]
+
+pi_3S <- read.delim("3S.pi", header = F)
+colnames(pi_3S) <- c("Chromosome", "window", "num.snps", "frac", "pi_3S")
+pi_3S$ID <- paste(pi_3S$Chromosome, pi_3S$window, sep = '_')
+pi_3S <- pi_3S[!pi_3S$pi_3S == "na",] #remove rows with na
+pi_3S <- pi_3S[,c(6,5)]
+
+pi_4S <- read.delim("4S.pi", header = F)
+colnames(pi_4S) <- c("Chromosome", "window", "num.snps", "frac", "pi_4S")
+pi_4S$ID <- paste(pi_4S$Chromosome, pi_4S$window, sep = '_')
+pi_4S <- pi_4S[!pi_4S$pi_4S == "na",] #remove rows with na
+pi_4S <- pi_4S[,c(6,5)]
+
+pi_4U <- read.delim("4U.pi", header = F)
+colnames(pi_4U) <- c("Chromosome", "window", "num.snps", "frac", "pi_4U")
+pi_4U$ID <- paste(pi_4U$Chromosome, pi_4U$window, sep = '_')
+pi_4U <- pi_4U[!pi_4U$pi_4U == "na",]
+pi_4U <- pi_4U[,c(6,5)]
+
+pi_5S <- read.delim("5S.pi", header = F)
+colnames(pi_5S) <- c("Chromosome", "window", "num.snps", "frac", "pi_5S")
+pi_5S$ID <- paste(pi_5S$Chromosome, pi_5S$window, sep = '_')
+pi_5S <- pi_5S[!pi_5S$pi_5S == "na",] #remove rows with na
+pi_5S <- pi_5S[,c(6,5)]
+
+pi_5U <- read.delim("5U.pi", header = F)
+colnames(pi_5U) <- c("Chromosome", "window", "num.snps", "frac", "pi_5U")
+pi_5U$ID <- paste(pi_5U$Chromosome, pi_5U$window, sep = '_')
+pi_5U <- pi_5U[!pi_5U$pi_5U == "na",]
+pi_5U <- pi_5U[,c(6,5)]
+
+pi_6U <- read.delim("6U.pi", header = F)
+colnames(pi_6U) <- c("Chromosome", "window", "num.snps", "frac", "pi_6U")
+pi_6U$ID <- paste(pi_6U$Chromosome, pi_6U$window, sep = '_')
+pi_6U <- pi_6U[!pi_6U$pi_6U == "na",]
+pi_6U <- pi_6U[,c(6,5)]
+
+pi_SD <- read.delim("SD.pi", header = F)
+colnames(pi_SD) <- c("Chromosome", "window", "num.snps", "frac", "pi_SD")
+pi_SD$ID <- paste(pi_SD$Chromosome, pi_SD$window, sep = '_')
+pi_SD <- pi_SD[!pi_SD$pi_SD == "na",]
+pi_SD <- pi_SD[,c(6,5)]
+
+pi_BR <- read.delim("BR.pi", header = F)
+colnames(pi_BR) <- c("Chromosome", "window", "num.snps", "frac", "pi_BR")
+pi_BR$ID <- paste(pi_BR$Chromosome, pi_BR$window, sep = '_')
+pi_BR <- pi_BR[!pi_BR$pi_BR == "na",]
+pi_BR <- pi_BR[,c(6,5)]
+
+#merge all pi results into one file
+pi_all1 <- merge(pi_1S, pi_1U, by="ID")
+pi_all2 <- merge(pi_all1, pi_2S, by="ID")
+pi_all3 <- merge(pi_all2, pi_2U, by="ID")
+pi_all4 <- merge(pi_all3, pi_3S, by="ID")
+pi_all5 <- merge(pi_all4, pi_4S, by="ID")
+pi_all6 <- merge(pi_all5, pi_4U, by="ID")
+pi_all7 <- merge(pi_all6, pi_5S, by="ID")
+pi_all8 <- merge(pi_all7, pi_5U, by="ID")
+pi_all9 <- merge(pi_all8, pi_6U, by="ID")
+pi_all10 <- merge(pi_all9, pi_SD, by="ID")
+pi_all <- merge(pi_all10, pi_BR, by="ID")
+
+pi_all$pi_1S <- as.numeric(pi_all$pi_1S)
+pi_all$pi_1U <- as.numeric(pi_all$pi_1U)
+pi_all$pi_2S <- as.numeric(pi_all$pi_2S)
+pi_all$pi_2U <- as.numeric(pi_all$pi_2U)
+pi_all$pi_3S <- as.numeric(pi_all$pi_3S)
+pi_all$pi_4S <- as.numeric(pi_all$pi_4S)
+pi_all$pi_4U <- as.numeric(pi_all$pi_4U)
+pi_all$pi_5S <- as.numeric(pi_all$pi_5S)
+pi_all$pi_5U <- as.numeric(pi_all$pi_5U)
+pi_all$pi_6U <- as.numeric(pi_all$pi_6U)
+pi_all$pi_BR <- as.numeric(pi_all$pi_BR)
+pi_all$pi_SD <- as.numeric(pi_all$pi_SD)
+
+dds.pcoa=pcoa(vegdist(t((pi_all[,2:13])), method="euclidean")/1000)
+scores=dds.pcoa$vectors
+percent <- dds.pcoa$values$Eigenvalues
+percent / sum(percent) #percent for each axes; change in caption below
+
+windows()
+plot(scores[,1], scores[,2],
+     col=c("red", "blue", "red", "blue", "red", "red", "blue", "red", "blue", "blue", "black", "black"), 
+     pch = c(19,19,19,19,19,19,19,19,19,19,1,19),
+     xlab = "PC1 (24.33%)", ylab = "PC2 (21.24%)")
+
+treat_df <- data.frame(
+  ID = c("1S_pi", "1U_pi", "2S_pi", "2U_pi", "3S_pi", "4S_pi", "4U_pi", "5S_pi", "5U_pi","6U_pi"),
+  line = c("S", "U", "S", "U", "S", "S", "U", "S", "U", "U")
+)
+
+line <- treat_df$line
+
+adonis2(t(pi_all[,2:11])~line, data = treat_df, permutations = 1000000, method = "manhattan")
+
+```
+![Principle coordinate analysis of allele frequencies for selected lines (red), control lines (blue), the pure Bodega population (white), and the pure San Diego population (black).](images/pi_PCA.png)
+
 ## References
+
+Griffiths JS, Kawji Y, Kelly MW.
+2020.
+An experimental test of adaptive introgression in locally adapted populations of splash pool copepods.
+Mol Biol and Evol.
+38(4):1306-1316.
+https://doi.org/10.1093/molbev/msaa289
+
+Schlötterer C, Tobler R, Kofler R, Nolte V. 2014 Sequencing pools of individuals-mining genome-wide polymorphism data without big funding. Nat. Rev. Genet. 15, 749–763. (doi:10.1038/nrg3803)

### 2022-02-03 index.Rmd
demo

@@
 ---
-
+Test
 The MarineOmics working group is part of the [RCN for Evolution in Changing Seas](https://rcn-ecs.github.io/), and was formed to promote **reproducible** and **robust** genomic research in nonmodel species, but especially marine and aquatic studies.
-Methods in next generation sequencing and bioinformatics advance at a pace that quickly render any static set of best practices as obsolete.
-To support robust 'omics research now and moving forward, the working group helps disseminate guidelines through this dynamic, open-sourced website.
+Methods in next generation sequencing and bioinformatics advance at a pace that quickly render any static set of best practices obsolete.
+MarineOmics aims to support robust genomic research in ecology and evolution by disseminating guidelines through this dynamic, open-sourced website addressing introductory and advanced topics of population- and functional genomics.
 
@@
 On this website you will find a growing and evolving collection of guidelines, tutorials, recommended readings, and method evaluations for various 'omics analyses that are commonly used in studies of nonmodel systems (particularly marine species).
-Instead of providing rigid "best practices", we aim to encourage data exploration and rigorous quality control through "best principles".
+Instead of providing rigid "best practices", we aim to encourage data exploration and rigorous quality control through "best principles", paying particular attention to topics for which open access guidelines are scarce.
 The guidelines are written by members of the MarineOmics working group or the community, then shared for peer review.
@@
 
-## Topics {.tabset .tabset-fade .tabset-pills}
-### Population Genomics
+### Topics {.tabset .tabset-fade .tabset-pills}
+#### Population Genomics
 
@@
 
-### Functional Genomics
+#### Functional Genomics
 
@@
 
-## {-}
+### {-}
 
@@
 # Reproducibility in Marine Genomics  
-The working group is also interested in evaluating and improving reproducibility in studies that use genomics to study adaptation in marine systems.
-utilizing genomics in marine species, through a peer-reviewed literature survey of data availability, code accessibility, and methods reproducibility.
+Marineomics is also interested in evaluating and improving reproducibility in studies that use genomics to study adaptation in marine systems.
+We are conducting a meta-analysis of reproducibility across genomic studies of environmental adaptation in marine populations with the aim of identifying obstacles and opportunities for reproducing and synthesizing population genomic research.
 In Spring 2021, three undergraduate research interns contributed to this literature survey, with funding from the [RCN-ECS](https://rcn-ecs.github.io/).
@@
 #### [Alan Downey-Wall](https://cos.northeastern.edu/people/alan-downey-wall/), Northeastern University {.unlisted .unnumbered}
-#### Chris Mantegna, University of Washington  {.unlisted .unnumbered}
+#### Chris Mantegna, University of Washington (Intern)  {.unlisted .unnumbered}
 #### [Danielle Davenport](https://www.researchgate.net/profile/Danielle-Davenport-2), University of Queensland {.unlisted .unnumbered}  
 #### [Hanny Rivera, PhD](https://www2.whoi.edu/staff/hrivera/), Woods Hole Oceanographic Institute  {.unlisted .unnumbered}
-#### Jecy Klinkam, College of William and Mary  {.unlisted .unnumbered}
+#### Jecy Klinkam, College of William and Mary (Intern) {.unlisted .unnumbered}
 #### [Joanna Griffiths, PhD](https://joannasgriffiths.wordpress.com/), UC-Davis {.unlisted .unnumbered}
 #### [Joe McGirr, PhD](https://joemcgirr.github.io/), UC-Davis {.unlisted .unnumbered}
-#### Kaitlin Macaranas, CSU-Bakersfield {.unlisted .unnumbered}
+#### Kaitlin Macaranas, CSU-Bakersfield (Intern) {.unlisted .unnumbered}
 #### Megan Guidry, University of Rhode Island {.unlisted .unnumbered}
@@
 
+# Internship 
+In Spring 2021, MarineOmics hosted three virtual undergraduate research interns with funding from the [RCN-ECS](https://rcn-ecs.github.io/): Jecy Klinkam, Kaitlin Macaranas, and Chris Mantegna.
+These interns were paired with graduate student mentors within the MarineOmics group, as well as participated in training activities while contributing to the Reproducibility literature review.
+
 # Contact 
-For more information about the working group or panel seminars, contact Katherine Silliman: kes0132[at]auburn.edu  
-For more information about the reproducibility review and undergraduate internship, contact Sam Bogan: snbogan[at]ucsb.edu
+For any and all inquiries about the MarineOmics Working Group, our website and its material, or other activities, please contact us at rcnecs.marineomics@gmail.com.

### 2022-02-03 index.Rmd
undo demo

@@
 ---
-Test
+
 The MarineOmics working group is part of the [RCN for Evolution in Changing Seas](https://rcn-ecs.github.io/), and was formed to promote **reproducible** and **robust** genomic research in nonmodel species, but especially marine and aquatic studies.

### 2022-02-04 RADseq.Rmd
fixed overlap of navbar, some radseq edits

@@
 
-1) [Look at your raw data.](#fastqc)
+<!-- 1) [Look at your raw data.](#fastqc)
 2) [Run an assembly pipeline](#assembly) through to a SNP dataset, using either all samples or a representative subset. Parameters can come from those used in a similar study, or default parameters.
@@
 7) Evaluate the difference between multiple filtering schemes for your analyses of interest. *Popgen analysis guide coming soon*
-8) Repeat as needed.
+8) Repeat as needed. -->
 
@@
 7) [Filter your SNPs](#filter)
-8) Repeat as needed!
+8) [Repeat as needed!](#repeat)
 
@@
 If you have demultiplexed data, you can use [MultiQC](https://multiqc.info/) to generate FastQC plots for all individuals and quickly identify ones that did not sequence well.
+
 Check out [this informative tutorial](https://datacarpentry.org/wrangling-genomics/02-quality-control/index.html) on running FastQC and interpreting the output.
+
 **Questions to ask:** Do you have a lot of adapter sequences?
@@
 If that doesn't occur or if too many reads are being filtered so as to only recover a small number of SNPs, something might need to be tweaked with your trimming and filtering step.
+<!--ADW: elaborating what you mean by 'tweaked' or possibly providing a link to some resources for how someone might thoughtfully change the parameters for their trimming and filtering steps.-->
 
@@
 ## Filtering SNPs {#filter}
+
 ### (and embrace the missingness!) {.unlisted .unnumbered}
@@
 
+## Repeat as needed {#repeat}
 # Helpful tutorials  

### 2022-02-04 choosing-population-genetics.Rmd
fixed overlap of navbar, some radseq edits

@@
 ```{r setup, include=FALSE}
-library(knitr)
+#library(knitr)
 knitr::opts_chunk$set(echo = TRUE)

### 2022-02-04 WGS-Full-Walkthrough.Rmd

@@
 #### Example Fastq File from Sequencing Facility 
+To inspect an example data file, download files i5-4-i7-11_R1_exampleData.fastq.gz & i5-4-i7-11_R2_exampleData.fastq.gz which are truncated fastq files as examples for how data look from the sequencing facility.
 
@@
 ```
-[schaal.s@login-00 CodGenomes]$ zcat i5-2-i7-9_R1_001.fastq.gz | head -n 10
-
-@GWNJ-1012:218:GW191226406th:1:1101:1090:1000 1:N:0:GATCAG+ATAGAGGC
-ACTTGACTGTGCGTTGGCCTGCGGGCTGACTCGGTCCTGAGATGGACTGCTGTGTAGTTTGAACCATAGATTCATTATATAGAACACGGTCTCCTCTGCGCTGCTGGCCAATGGAGCCGAACGTCCGCACTGGCGGGCGGCCATCTTGCC
+[schaal.s@login-00 workingGroupFiles]$ head i5-4-i7-11_R1_exampleData.fastq.gz 
+@GWNJ-1012:218:GW191226406th:1:1101:1181:1000 1:N:0:GGCTAC+GGCTCTGA
+CGATGTTTGGAAAGACAAAAAATACATTTAGATCCAAATGACTTCTATAATGGATTTTGAAATTATATATACGAGATGCTTTAGTTCAAAACATTTGACACAAATGAGGTTCAACATAAAGAACATCGAGATCGGAAGAGCACACGTCTG
 +
-FF:,FF,FFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFF:FFFFFFFFFFFFFFFFFFFFFFF,:FFFFFFFFF:FFFFFFF:FFFFFFFFFFFFF:FFFFFFF
-@GWNJ-1012:218:GW191226406th:1:1101:1687:1000 1:N:0:GATCAG+ATAGAGGC
-ACTTGATCCCTCTCACTCTCCTCTCCGTCTCCTCTTTTGTCCTCGTCTCTCTCCTCTCTCCCTCTCTCCCATCTCCCTCTCTATCAAGTAGATCGGAAGAGCACACGTCTGAACTCCAGTCACGATCAGATCTCGTATGCCGTCTTCTGC
+FFF:FF,FFFFFFFFFFFFFFFFFFFFFF:FFFFFFFFFFFFFFFF:FFFFFFFFFF:F,FFF:FFFFFFFFFFFFFFFFFFFFF:FFFFFFFFFFFFFFFFFFFFFFFFFFFFFFF,FFFFFFFFFFFFF:FFFF,F:FFFFFFF:FFF
+@GWNJ-1012:218:GW191226406th:1:1101:1651:1000 1:N:0:GGCTAC+GGCTCTGA
+CTTGCATTGAGATTTCTGAAAGAAGAGCTCCATCCAGTGGCAGAAAGGGGACGTTTCAAAAGCACGAATGCAAGCCTTTTAATTTTTCCAAAATAAAAATATGCAAGAGATCGGAAGAGCACACGTCTGAACTCCAGTCACGGCTACATC
 +
-FFFFFF,FFFFFFFFFFFFFF:FFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFF:FFFFFFFFFFFFFFFFFF:FFFFFFFFFFFFFFFF:FFFFFFFFFFFFFFFFFFFFFFFF:FFFF:::F,F::FFFFF
-@GWNJ-1012:218:GW191226406th:1:1101:6840:1000 1:N:0:GATCAG+ATAGAGGC
-ACTTGAAAAAAATACATAGCGGCCATGGACAGGATGACCTCTATGACAATGATAGAAACAGAAAGGACGCGGAGACTCTTGAGTCATCAAGTAGATCGGAAGAGCACACGTCTGAACTCCAGTCACGATCAGATCTCGTATGCCGTCTTC
+FFFFFF,FFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFF
+@GWNJ-1012:218:GW191226406th:1:1101:2193:1000 1:N:0:GGCTAC+GGCTCTGA
+CGATGTTGCATGGAGTGGCTGGTGTCTGTGCCGACCACTCTGGTTCCAGAGGTAATGAAGACCAGCTTCCCCCCTGGCCACACAGTGGACTCAATCTTCCCTGAGATAAAGGTGAAGAGTTCCTCCTTGTTCTTAGGGTAACATCGAGAT
 ```	
@@
 ```
-[schaal.s@login-00 CodGenomes]$ zcat i5-2-i7-9_R2_001.fastq.gz | head -n 10
+[schaal.s@login-00 workingGroupFiles]$ head i5-4-i7-11_R2_exampleData.fastq.gz 
+@GWNJ-1012:218:GW191226406th:1:1101:1181:1000 2:N:0:GGCTAC+GGCTCTGA
+CGATGTGCTTTATGTTGAACCTCATTTGTGTCAAATGTTTTGAACTAAAGCATCTCGTATATATAATTTCAAAATCCATTATAGAAGTCATTTGGATCTAAATGTATTTTTTGTCTTTCCAAACATCGAGATCGGAAGAGCGTCGGGTAG
++
+FFFFFF,FF,FFFFFFFF:FFFFFFFFFFFFFFFFFFFFFFF:FFFFFFFFF:FFFF:F,F:FFF,:FFFFFFFFFF,,:,:FFFFFFFF,FFFFFFFFF:F:FFF:F,F,FFFFF:FFF:FFF,,F::FFFFF,,F:FFFFFFF,F,,F
+@GWNJ-1012:218:GW191226406th:1:1101:1651:1000 2:N:0:GGCTAC+GGCTCTGA
+CTTGCATATTTTTATTTTGGAAAAATTAAAAGGCTTGCATTCGTGCTTTTGAAACGTCCCCTTTCTGCCACTGGATGGAGCTCTTCTTTCAGAAATCTCAATGCAAGAGATCGGAAGAGCGTCGTGTAGGGAAAGAGTGTTCAGAGCCGT
++
+FFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFF:FFFFFFFFFFFFFFFFFFFFFFFF:FFFFFFFFFFFFFFFFF:FFFFFFFFFFFFFFF,
+@GWNJ-1012:218:GW191226406th:1:1101:2193:1000 2:N:0:GGCTAC+GGCTCTGA
+CGATGTTACCCTAAGAACAAGGAGGAACTCTTCACCTTTATCTCAGGGAAGATTGAGTCCACTGTGTGGCCAGGGGGGAAGCTGGTCTTCATTACCTCTGGAACCAGAGTGGTCGGCACAGACACCAGCCACTCCATGCAACATCGAGAT
+```
 
-@GWNJ-1012:218:GW191226406th:1:1101:1090:1000 2:N:0:GATCAG+ATAGAGGC
-ACTTGAAGGCAAGATGGCCGCCCGCCAGTGCGGACGTTCGGCTCCATTGGCCAGCAGCGCAGAGGAGACCGTGTTCTATATAATGAATCTATGGTTCAAACTACACAGCAGTCCATCTCAGGACCGAGTCAGCCCGCAGGCCAACGCACA
-+
-FFFFFF,FFFF:FFFFFF:FFFFFFFFFFFFFFFFFF:FFFFFFF:,FFFFF:FF,FFFFFFFFFFFFFFF,F,:FFFF:F,FFFFF:FF,FFFF,FFFFFF,FFF,FF:FFFFF:F:FFFF,FFF:F:F:FFFFFFFFFFF,:FFFFFF
-@GWNJ-1012:218:GW191226406th:1:1101:1687:1000 2:N:0:GATCAG+ATAGAGGC
-ACTTGAAAGAGAGGGAGATGGGAGAGAGGGAGAGAGGAGAGAGACGAGGACAAAAGAGGAGACGGAGAGGAGAGTGAGAGGGATCAAGTAGATCGGAAGAGCGTCGTGTAGGGAAAGAGTGTGCCTCTATGTGTAGATCTCGGTGGTCGC
-+
-FFFFFF,FFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFFF,FFFF:FFFFFFFFFFFFFFFFFFFFFFFFFFFF:FF,FF:
-@GWNJ-1012:218:GW191226406th:1:1101:6840:1000 2:N:0:GATCAG+ATAGAGGC
-ACTTGAAGACTCAAGAGTCTCCGCGTCCTTTCTGTTTCTATCATTGTCATAGAGGTCATCCTGTCCATGGCCGCTATGTATTTTTATCAAGTAGATCGGAAGAGCGTCGTGTAGGGAAAGAGTGTGCCTCTATGTGTAGATCTCGGTGGT
-```
+What these files show are XXXX.
 
@@
 Barcode File: For each set of paired end files that need to be demultiplexed, a barcode file needs to be created.
-For example, for files i5-4-i7-11_R1_001.fastq.gz and i5-4-i7-11_R2_001.fastq.gz its barcode file is as follows:
+For example, for files i5-4-i7-11_R1_exampleData.fastq.gz and i5-4-i7-11_R2_exampleData.fastq.gz its barcode file is as follows:
 
@@
 This file is tab delimited and because I had inline barcodes, meaning they were the same on both ends of the read, I have the same barcode sequence in the first and second column.
-Then the sample ID is on the right.
+Then the sample ID is in the far right column.
 
@@
 ```
-process_radtags -P -p {Path to genome files} -o .{Path for output} -b {barcode file} --inline_inline --disable_rad_check -r
+process_radtags -P -p {Path to genome files} -o {Path for output} -b {barcode file} --inline_inline --disable_rad_check -r
 
@@
 #SBATCH --job-name=stacks_i54_i711                # sets the job name
-#SBATCH --mem=1Gb                                 # reserves 10 GB memory
-#SBATCH --partition=lotterhos                     # requests that the job is executed in lotterhos partition 
-#SBATCH --mail-user=schaal.s@northeastern.edu
-#SBATCH --mail-type=FAIL
+#SBATCH --mem=1Gb                                 # reserves 1 GB memory
+#SBATCH --mail-user=schaal.s@northeastern.edu     # add your email address to get notifications
+#SBATCH --mail-type=FAIL                          # set what email notifications you want to receive here I only want emails when my script fails
 #SBATCH --time=24:00:00                           # reserves machines/cores for 24 hours.

### 2022-02-04 poolseq.Rmd
updated Poolseq references in bib and rmd file
updating references on rmd, bib, and html

@@
 
-The following scripts were used to analyze data from an experimental evolution study on *Tigriopus californicus* copepods (Griffiths 2020).
+The following scripts were used to analyze data from an experimental evolution study on *Tigriopus californicus* copepods [@Griffiths2020].
 Raw sequencing data are deposited in NCBI's Short Reads Archive (https://www.ncbi.nlm.nih.gov/bioproject/PRJNA597336/).
@@
 It is critical to note, the sequencing accuracy of Pool-seq increases with larger pool sizes; if the number of sequenced individuals is approximately equal to the coverage, then the sequency accuracy is no longer superior to an individual level of sequencing (see Fig.
-1a from Schlötterer et al., 2014).
+1a from [@Schlotterer2014]).
 
@@
 ## References
-
-Griffiths JS, Kawji Y, Kelly MW.
-2020.
-An experimental test of adaptive introgression in locally adapted populations of splash pool copepods.
-Mol Biol and Evol.
-38(4):1306-1316.
-https://doi.org/10.1093/molbev/msaa289
-
-Schlötterer C, Tobler R, Kofler R, Nolte V. 2014 Sequencing pools of individuals-mining genome-wide polymorphism data without big funding. Nat. Rev. Genet. 15, 749–763. (doi:10.1038/nrg3803)

### 2022-02-18 WGS-Full-Walkthrough.Rmd
Update WGS high cov walkthrough

@@
 ---
-title: WGS fastq2vcf walkthrough
-
+layout: page
+title: WGS Full Walkthrough
+permalink: /WGSwalkthrough/
 ---
-
-## Walkthrough Data
-In this walkthrough, data files from whole genome sequences at 15x coverage from Atlantic cod will be referenced.
-Their naming struction are a population ID in the form of "Pop#" with the number being which population the sample came from, a "\_"\, and a 5 digit sample ID which is two digits for the year it was sampled and a three digit sample ID number.
-Here is an example, "Pop6_18001".
-We will walk through steps from demultiplexing when unique indexes are used through variant calling with Freebayes.
 
@@
 #### Example Fastq File from Sequencing Facility
-To inspect an example data file, download files i5-4-i7-11_R1_exampleData.fastq.gz & i5-4-i7-11_R2_exampleData.fastq.gz which are truncated fastq files as examples for how data look from the sequencing facility.
 
@@
 
-What these files show are XXXX.
+### Stacks Walkthrough
 
-### Stacks - process_radtags
+#### process_radtags walkthrough
 
 Barcode File: For each set of paired end files that need to be demultiplexed, a barcode file needs to be created.
-For example, for files i5-4-i7-11_R1_exampleData.fastq.gz and i5-4-i7-11_R2_exampleData.fastq.gz its barcode file is as follows:
+For example, for files i5-4-i7-11_R1_001.fastq.gz and i5-4-i7-11_R2_001.fastq.gz its barcode file is as follows:
 
@@
 This file is tab delimited and because I had inline barcodes, meaning they were the same on both ends of the read, I have the same barcode sequence in the first and second column.
-Then the sample ID is in the far right column.
+Then the sample ID is on the right.
 
@@
 
-We provide a simple R script to create the barcode files for each primer index combination. See file: Stacks_FileCreate.R
+We provide a simple R script to create the barcode files for each primer index combination.
+See file: Stacks_FileCreate_OmicsWG.R.
+You can use the practice files provide to see how the code works or just input your own to make the files you need.
+This file may be modified to fit your needs and there are annotations where that would be the case.
 
@@
 
-```--inline_inline ``` The barcode option flag has 6 different options and you have to use the proper flag for your barcodes or indexes need to be demultiplexed.
-Because my samples have inline barcodes on either end of the sequence I need to use the --inline_inline flag.
+```--inline_inline ``` The barcode option flag has 6 different options and you have to use the proper flag for your barcodes or indexes that need to be demultiplexed.
+Because the example samples have inline barcodes on either end of the sequence if using the practice data you need to use the --inline_inline flag.
 
-```-P``` If paired end sequencing was used you need to add the -P flag to indicate this.
+```-P``` If paired end sequencing was used, you need to add the -P flag to indicate this.
 
@@
 
-```-r``` this flag allows Stacks to rescue any barcodes that are off by one bp
+```-r``` this flag allows Stacks to rescue any barcodes that are off by one bp.
+For example, the barcode ACTTGA might appear as CTTGA or TACTTGA.
+This should only be a small percentage of reads and will probably recover about 1% of reads.
 
@@
 #### Example code for running on a Cluster  
+You can run the practice data files using the format below while changing the path on your computer to fit your needs. 
 ```
@@
 #SBATCH --mem=1Gb                                 # reserves 1 GB memory
-#SBATCH --mail-user=schaal.s@northeastern.edu     # add your email address to get notifications
-#SBATCH --mail-type=FAIL                          # set what email notifications you want to receive here I only want emails when my script fails
+#SBATCH --mail-user=schaal.s@northeastern.edu     # set your email to get notifications about your run
+#SBATCH --mail-type=FAIL                          # if the job fails for any reason email the provided email
 #SBATCH --time=24:00:00                           # reserves machines/cores for 24 hours.
@@
 
-process_radtags -P -p i54_i711/GenomeFilesi54_i711/ -o ../Stacks_Out/ -b i54_i711/barcodeFilei54_i711.txt --inline_inline --disable_rad_check -r
+process_radtags -P -p ./genomeFilesi54_i711/ -o ./Stacks_Out/ -b barcodeFilei54_i711.txt --inline_inline --disable_rad_check -r
 ```
+This will create four new fastq files for each sample provided in the barcodeFile.
+Because the practice data is a subset of the full dataset, this will only fill files with reads that are included in the subset file.
+The process_radtags program wants to keep the reads in phase, so that the first read in the sample_XXX.1.fq file is the mate of the first read in the sample_XXX.2.fq file.
+Likewise for the second pair of reads being the second record in each of the two files and so on.
+When one read in a pair is discarded due to low quality or a missing restriction enzyme cut site, the remaining read can't simply be output to the sample_XXX.1.fq or sample_XXX.2.fq files as it would cause the remaining reads to fall out of phase.
+Instead, this read is considered a remainder read and is output into the sample_XXX.rem.1.fq file if the paired-end was discarded, or the sample_XXX.rem.2.fq file if the single-end was discarded.
 
-### Results 
+There will also be a log file that summarizes the number of reads per sample that were retained, were low quality, or that had noRADtag (only relevant for rad data).
+
+#### Results for a run with one primer combination on the full paired end data files that for each ~30GB in size
 Time: 				each file took approx. 7 hrs  
 Memory: 			approx. 80 MB RAM
+Run description: run on separate cluster nodes for 32 different primer combinations
 
@@
 #SBATCH --mail-type=FAIL
-#SBATCH --partition=lotterhos
 #SBATCH --time=1:00:00
@@
 ```
-![Output](images/Pop1_17291100XcoveragePlot10000.png)
-
-<!-- <img src="./Pop1_17291100XcoveragePlot10000.png" width="500">  -->
+<img src="./Pop1_17291100XcoveragePlot10000.png" width="500">
 
@@
 }
-
 setwd("~/Postdoc_Projects/RCN_MarineOmics/Website/marineomics.github.io/guidelines")
 KnitPost("~/Postdoc_Projects/RCN_MarineOmics/Website/marineomics.github.io/guidelines/WGSFullWalkthrough.Rmd")
-
 ```

### 2022-06-06 DGE_comparison.Rmd
Push .rmd for multifac RNAseq page

---
title: "Fitting Multifactorial Models of Differential Expression"
author: "Sam Bogan"
csl: molecular-ecology.csl
output:
  html_document:
    toc: yes
    toc_depth: 3
  word_document:
    toc: yes
    toc_depth: '3'
  pdf_document: default
  github_document: default
bibliography: DGE_Comp.bib
---

#Intro to multifactorial RNA-seq models

|           Studies of molecular responses to environmental change increasingly employ multifactorial experimental designs incorporating different developmental stages, stressors, populations, or non-linear dynamics that resolve interactions between variables that shape expression, physiology, and performance. After an experiment during which a study organism was exposed to abiotic variation and perhaps other variables, researchers interested in how gene expression correlates with these variables will often extract RNA across treatment groups, prepare cDNA libraries for sequencing, and conduct an analysis of differential expression (DE) across variables of interest after mapping and counting RNA-seq reads. DE analysis is achieved by fitting RNA-seq read count data to statistical models, often using tried-and-true DE packages. Most packages do not have the flexibility to accommodate a number of common experimental design components and it can be difficult to navigate their functionality for fitting multivariate models. When these packages fall short, determining how to fit customized multifactorial models yourself can be an additional challenge. Many great walkthroughs exist for conducting DE analysis across using single-variable experimental designs, but there is a lack of guidance on multifactorial approaches for testing for DE and modelling gene expression. Regardless of whether you have RNA-seq data on hand or you're in the process of planning your experiment, this walkthrough will help you navigate what DE packages and model fitting approaches are best suited to your experimental design.

##Targeted audience
|           This walkthrough is for both intermediate and expert practitioners of RNA-seq analyses, and most of all, those new to multifactorial RNA-seq. We have written this page with the assumption that readers have a basic familiarity with model fitting and statistical analysis, as well as next generation sequencing technologies, RNA-seq workflows, and the R statistical environment.

##Variables we will cover

|           This walkthrough can be broken down into different types of predictor variables that can be incorporated in models of gene expression. In the order we address them, these variables include:

* Non-linear fixed effects
* Continuous fixed effects
* Interactive fixed effects
* Random intercepts
* Random slopes

##Packages we will cover

* DESeq2 [@Love2014-fy] [Link](http://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html)
* edgeR [@Robinson2010-fd] [Link](https://www.bioconductor.org/packages/devel/bioc/vignettes/edgeR/inst/doc/edgeRUsersGuide.pdf)
* Voom (often called Limma-Voom) [@Law2014-ce] [Link](https://ucdavis-bioinformatics-training.github.io/2018-June-RNA-Seq-Workshop/thursday/DE.html)

#Features of popular DGE packages

|           Two important considerations when choosing a DE package to analyze RNA-seq data are the distribution family used by the package and the effect types it incorporate into a model. The DE packages we will cover assume the following read count distributions and are capable of fitting the following effect types:

| Program | Distribution | Continuous fixed eff. | Random intercepts | Random slopes | Interactive eff. | Non-linear eff. |
| :--- | :--- | :--- | :--- | :--- | :--- |
| edgeR | Negative binomial generalized linear model | ✔ | ✖ |  ✖ |  ✔ |  ✔ |  
| DESeq2 | Negative binomial generalized linear model | ✔ | ✖ | ✖ | ✔|   ✔ | 
| Voom | Mean-variance linear model | ✔ | ✔ | ✖ |  ✔|   ✔ |

##Summary of approach

|           We will walk you through (i) how to conduct essential checks of your data such as principal components analysis and observing the distribution of filtered read counts, (ii) evaluating each package's functionality for multifactorial model fitting, and (iii) how to fit each type of predictor by performing differential expression tests using real data. These data come from an experiment that exposed the Antarctic pteropod (a free-swimming planktonic snail) *Limacina helicina antarctica* to three continuous levels of ocean acidification (OA) for either 0.5 or 7 days [@Bogan2020-qs]. These RNA-seq counts are derived from reads mapped and counted via RSEM [@Li2011-mi] using a *de novo* reference transcriptome that includes transcript isoforms assembled by @Johnson2016-wq.

|           Throughout this walkthrough, we will provide examples of custom scripts for fitting linear, linear mixed models, generalized linear models to read count data using for loops in order to provide functionality where it cannot be met by most differential expression packages.

|           At the end of the walkthrough, we demonstrate how to compare model predictions by different packages for different effect types commonly used in multifactorial RNAseq analyses.

|           Let's start by loading our required packages...

```{r setup, include = FALSE}

# Set root directory
knitr::opts_knit$set( root.dir = '~/Documents/GitHub/DGE_comparison/' )

```

```{r, results = FALSE, message = FALSE, warning = FALSE}

## Unhash and run the code below if you believe you may need to install the packages loaded below

#invisible(lapply(c("DESeq2", "edgeR", "tidyverse", "ape", "vegan", "GGally",
                   #"arrayQualityMetrics", "rgl", "adegenet", "MASS",
                   #"data.table", "plyr", "lmtest", "reshape2", "Rmisc", "lmerTest"),
                 #function(p){
                   #if(! p %in% rownames(installed.packages())) {
                     #install.packages(p)
                  # }
                  # library(p, character.only=TRUE)
                 #}))

# Load packages
library(DESeq2)
library(edgeR)
library(tidyverse)
library(ape)
library(vegan)
library(GGally)
library(arrayQualityMetrics)
library(rgl)
library(adegenet)
library(MASS)
library(data.table)
library(plyr)
library(lmtest)
library(reshape2)
library(Rmisc)
library(lmerTest)

```

#Filter and visualize read counts

|           Before model fitting and testing for DE, it is important to visually inspect read counts pre- and post-filtering. In this walk through, we will plot the distributions of read counts and sample loading to multidimensional scaling axes (e.g., plotting principal components or principal coordinates analyses). Looking at these plots, we can (i) determine whether the distribution of our data match the assumptions of model families used by different packages and (ii) determine whether one or multiple variables within a dataset explain a significant degree of variation in gene expression across replicates. Depending on the distribution of your data and the variables that appear important (or variable you would like to incorporate into your models), we will then demonstrate how to choose an appropriate package to test for DE.
|           Most DE packages assume that read counts possess a negative binomial distribution. The negative binomial distribution is an extension of distributions for binary variables such as the Poisson distribution, allowing for estimations of "equidispersion" and "overdispersion", equal and greater-than-expected variation in expression attributed to biological variability . However, RNA-seq datasets can exhibit poor fits with the negative binomial distribution [@Hawinkel2020-lk]. It is well worth visualizing and testing the distributions of RNA-seq read counts before and after filtering. Keep in mind that the distribution applies not to the raw data, but the residuals of a model. Additionally, individual models are fit to the read counts of each gene. Thus, the most accurate representation of how appropriate a negative binomial distribution is for read counts would come from plotting the distribution of residuals across all genes following model fitting. Below is a visual example of negative binomial distributions that possess different variances from modelassist[dot]epixanalytics[dot]com:

```{r, echo = FALSE, out.height = "300%", fig.align = "center", }

knitr::include_graphics("images/negbinom.png")

```

|           Now let's first plot the distribution of unfiltered reads from our dataset:

```{r, message = FALSE, warning = FALSE}

# Read in matrix of RSEM expected read counts
data <- read.delim("Input_data/GE2_expected_counts_gene.matrix", sep = "\t", header = T, row.names = 1)

# Peak at data to check that it looks okay
head(data)

```

|           The table of read counts above has rows representing individual transcripts and columns representing different replicate samples across treatments.

```{r, message = FALSE, warning = FALSE}
# Name experimental samples: names correspond to pCO2 treatment (300, 600, 900) + days (12 h2 vs 7 days) + replicate ID
colnames(data) <- c( "300.7.a", "300.7.b", "300.7.c","300.12.a", "300.12.b", "300.12.c", 
                       "900.7.a", "900.7.b", "900.7.c", "900.12.a", "900.12.b", "900.12.c", 
                       "600.7.a", "600.7.b", "600.7.c", "600.12.a", "600.12.b", "600.12.c")

## Create 'targets' and 'Group dataframe, expressing experimental variables for DEG analysis
pCO2 <- as.numeric( c( 255, 255, 255, 255, 255, 255,
                       530, 530, 530, 530, 530, 530,
                       918, 918, 918, 918, 918, 918 ) )

treatment <- c( "B", "B", "B", "B", "B", "B",
                "R", "R", "R", "R", "R", "R",
                "Y", "Y", "Y", "Y", "Y", "Y" )

day <- as.numeric( c( 7, 7, 7, .5, .5, .5,
                      7, 7, 7, .5, .5, .5,
                      7, 7, 7, .5, .5, .5 ) )

targets <- data.frame( pCO2, day, treatment )
targets$grouping <- paste( targets$pCO2, targets$day, sep = "." )

# The group factor represents the combined levels of an experimental replicate across all variables
Group <- factor( paste( targets$day, targets$pCO2, sep = "_" ) )

# Data must be rounded to nearest integer in order to be fit for negative binomial distribution
data_input <- round( data )

# Peak at rounded data
head( data_input )

# Plot distribution of unfiltered read counts across all samples 
ggplot( data = data.frame( rowMeans( data_input ) ),
        aes( x = rowMeans.data_input. ) ) +
  geom_histogram( fill = "grey" ) +
  xlim( 0, 500 ) +
  theme_classic() +
  labs( title = "Distribution of unfiltered reads" ) +
  labs( y = "Density", x = "Raw read counts", title = "Read count distribution: untransformed, unnormalized, unfiltered" )

```

|           As you can see in the above plot, the raw distribution of all read counts takes on a left-skewed negative binomial distribution similar to the purple distribution in the example above. While looking at the distribution of your raw reads is useful, these are not in fact the data you will be inputting to tests of differential expression. Let's plot the distribution of filtered reads normalized by library size, expressed as log2 counts per million reads (logCPM). These are the reads we will use in our test, and after we plot their distribution, we will conduct one more, slightly more robust, test of our data's fit to the negative binomial distribution using residuals from fitted models.

```{r, message = FALSE, warning = FALSE}

# Make a DGEList object for edgeR
y <- DGEList( counts = data_input, remove.zeros = TRUE )

#Let's remove samples with less than 0.5 cpm (this is ~10 counts in the count file) in fewer then 9/12 samples
keep <- rowSums( cpm( y ) > .5 ) >= 9

table( keep )

# Set keep.lib.sizes = F and recalculate library sizes after filtering
y <- y[ keep, keep.lib.sizes = FALSE ]

y <- calcNormFactors( y )

# Calculate logCPM
df_log <- cpm( y, log = TRUE, prior.count = 2 )

# Plot distribution of filtered logCPM values
ggplot( data = data.frame( rowMeans( df_log ) ),
        aes( x = rowMeans.df_log. ) ) +
  geom_histogram( fill = "grey" ) +
  theme_classic() +
  labs( y = "Density", x = "Filtered read counts (logCPM)", 
        title = "Distribution of normalized, filtered read counts" )

```

|           Our raw data appear to follow a strong negative binomial distribution! Later, we will see whether this holds for residuals from fitted negative binomial GLMs.

##MDS plot visualizing experimental factors

|           Before analyzing our data, it is essential that we look at the multivariate relationships between our samples based on transcriptome-wide expression levels. Below is example code and output for a principal coordinates analysis (PCOA) plot that visualizes multifactorial RNA-seq replicates according to two predictor variables across major and minor latent variables or PCOA axes. These predictor variables, as discussed above, are *p*CO${_2}$ and time-in-treatment.

```{r, message = FALSE, warning = FALSE}

# Export pcoa loadings
dds.pcoa = pcoa( vegdist( t( df_log <- cpm( y, log = TRUE, prior.count = 2 ) ),
                          method = "euclidean") / 1000 )

# Create df of MDS vector loading
scores <- dds.pcoa$vectors

## Plot pcoa loadings of each sample, groouped by time point and pCO2 treatment

# Calculate % variation explained by each eigenvector
percent <- dds.pcoa$values$Eigenvalues
cumulative_percent_variance <- ( percent / sum( percent ) ) * 100

# Prepare information for pcoa plot, then plot
color <- c( "steelblue1", "tomato1", "goldenrod1")
par( mfrow = c( 1, 1 ) )
plot( scores[ , 1 ], scores[ , 2 ],
      cex=.5, cex.axis=1, cex.lab = 1.25,
      xlab = paste( "PC1, ", round( cumulative_percent_variance[ 1 ], 2 ), "%" ), 
      ylab = paste( "PC2, ", round( cumulative_percent_variance[ 2 ], 2 ), "%" ) )

# Add visual groupings to pcoa plot
ordihull( scores, as.factor(targets$treatment ), 
          border = NULL, lty = 2, lwd = .5, label = F,
          col = color, draw = "polygon", alpha = 100, cex = .5 )

ordispider( scores,as.factor( targets$grouping ),label = F ) # Vectors connecting samples in same pCO2 x time group

ordilabel( scores, cex = 0.5) # Label sample IDs

```

|           In the PCOA above, blue = low *p*$CO_{2}$, yellow = moderate *p*$CO_{2}$, and red = high *p*$CO_{2}$. From this plot, we can see that treatment and time both influence multivariate gene expression across the RNA-seq samples. By and large, samples cluster according to these two predictors in a manner consistent with what we would expect from our experimental design if *p*$CO_{2}$ and time were to affect gene expression.

```{r, message = FALSE, warning = FALSE}

logCPM.pca <- prcomp( t ( df_log ) )
logCPM.pca.proportionvariances <- ( ( logCPM.pca$sdev ^ 2 ) / ( sum( logCPM.pca$sdev ^ 2 ) ) ) * 100

## Do treatment groups fully segregate? Wrap samples by pCO2 x time, not just pCO2
# Replot using logCPM.pca
plot( logCPM.pca$x, type = "n", main = NA, xlab = paste( "PC1, ", round( logCPM.pca.proportionvariances[ 1 ], 2 ), "%" ), ylab = paste( "PC2, ", round( logCPM.pca.proportionvariances[ 2 ], 2 ), "%" ) )

points( logCPM.pca$x, col = "black", pch = 16, cex = 1 )
colors2 <- c( "steelblue1", "dodgerblue2", "tomato1", "coral", "goldenrod1", "goldenrod3" )

ordihull( logCPM.pca$x, targets$grouping, 
          border = NULL, lty = 2, lwd = .5, 
          col = colors2, draw = "polygon", 
          alpha = 75,cex = .5, label = T )

```

|           This plot is only a slight adjustment of the first PCOA. If you'd like to wrap each multifactorial group and get a better idea of potential overlap between clusters, the plot and code above can be applied to do so.

|           Now that we have conducted QC analyses and plots of the read count data, it is time to begin model fitting and testing for differential expression.

#Non-linear effects

|           Gene expression traits can exhibit non-linear performance curves across a continuous variable much like other physiological traits. Studies that incorporate more than two levels of a continuous predictor provide an opportunity to test for non-linear effects on gene expression. Similar to performance traits such as growth or metabolic rates, gene expression can be modelled across a continuous variable by fitting non-linear 'performance curves' to read count data. Rivera *et al.* 2021 recently discussed the benefits of fitting non-linear reaction norms to gene expression data in order to better model acclimation by tolerant and sensitive genotypes to an environmental stressor. In a conceptual figure of theirs, shown below, they depict non-linear variation in expression before, during, and after exposure to a stressor for four reaction norm phenotypes: (i) a tolerant conspecific that frontloads (e.g., constitutively upregulates) stress-induced transcripts, (ii) a tolerant conspecific that exhibits reduced overall expression of stress-responsive transcripts, (iii) a sensitive conspecific that upregulates inducible transcripts and cannot recover post-stress, and (iv) a resilient conspecific that mounts a transcriptional stress response followed by recovery and downregulation [@Rivera2021-hn].

```{r, echo = FALSE, out.height = "40%", fig.align = "center", }
knitr::include_graphics("images/Rivera_etal_fig.png")
```
From Rivera *et al.* 2021 - "Transcriptomic resilience and timing. (a) Gene expression reaction norms of four strategies during recovery after a stressor. We use triangles again for patterns that may confer tolerance and circles for patterns associated with stress sensitivity. While all triangle paths show a return to baseline (resilience) the pink (frontloading) and yellow (dampening) are also depicting differences in baseline and plasticity and are therefore labelled differently. (b) Adapted from the rolling ball analogy commonly used for ecological resilience and depicted in Hodgson et al. (2015). Each ball represents a gene showing a color-matched expression pattern in (a). Landscapes represent expression possibilities during a stress event. In the absence of stress, the ball will settle in a trough, representing baseline expression levels. Elasticity (rate of return to the baseline) is represented by the size of the arrow (i.e., larger arrows have faster rates of return). Pink dotted line is the expression landscape for the frontloaded ball. (c) Using Torres et al. (2016) loops through disease space as an alternative framework of an organism's path through stress response and recovery. The color gradient represents the resulting phenotype for a given path through stress and recovery space, though x-and y-axis can denote any two parameters that are correlated but with a time lag."

|           Non-linear reaction norms can be modelled across time, such as Rivera *et al.* have presented, or across multiple levels of an abiotic variable such as temperature, pH, etc. in order to fit a performance curve. Performance curves are a fundamental tool in ecological physiology, and enable more robust hypothesis testing in RNA-seq studies of environmental acclimation. For example, testing for variation in the shape of gene expression performance curves between early and long-term exposure timepoints can provide critical information about the role of gene expression in acclimation. Below we outline how to fit and test for non-linear gene expression performance curves in multifactorial RNA-seq experiments using examples in edgeR, DESeq2, and custom code.    
|           One of the simplest non-linear relationships that can be fitted to the expression of a transcript across an continuous variable is a second-order polynomial, otherwise known as a quadratic function, which can be expressed as $y_{i} = \mu + \beta_{1}x^2 + \beta_{2}x$ where $y$ is the abundance of a given transcript ($i$), $\mu$ is the intercept, and $y$ is the continuous variable. For the parabola generated by fitting a second-order polynomial, $\beta_{1}$ > 0 opens the parabola upwards while $\beta_{1}$ < 0 opens the parabola downwards. The vertex of the parabola is controlled by $\beta_{2}$ such that when $\beta_{1}$ is negative, greater $\beta_{2}$ values result in the vertex falling at higher values of $x$. 
|           Quadratic polynomials applied to phenotypic performance curves commonly possess negative $\beta_{1}$ values with positive $\beta_{2}$ values: a downard-opening parabola with a positive vertex. However, quadtratic curves fitted to gene expression data can take on a variety of postiive or negative forms similar to exponential curves, saturating curves, and parabolas. For instance, the expression of a gene may peak an intermediate level of an environmental level before crashing or it may exponentially decline across that variable. Such trends may better model changes in the transcription of a gene compared to a linear model. To get started, we will fit non-linear second order polynomials before testing for whether model predictions for a given gene are significantly improved by a non-linear linear model.

##Non-linear effects: example in edgeR

|           Let's fit a second-order polynomial for the effect of *p*CO$_{2}$ using edgeR. Using differential expression tests, we will then determine whether *p*CO$_{2}$ affected a gene's rate of change in expression and expression maximum by applying differential expression tests to $\beta_{1}$ and $\beta_{2}$ parameters. By testing for differential expression attributed to intereactions between time and $\beta_{1}$ or $\beta_{2}$, we will then test for whether these parameters were significantly different across exposure times such that 0.5 days and 7 days of acclimation to *p*CO$_{2}$ altered the rate of change in expression across *p*CO$_{2}$ ($\beta_{1}$) or the maximum of expression ($\beta_{2}$).

|          Since this is the first time we're fitting models to our data in this walkthrough, we will also output some important plots to help us explore and QC model predictions as we work toward testing for differential expression attributed to non-linear effects.

```{r, message = FALSE, warning = FALSE}

# Square pCO2 variable
pCO2_2 <- pCO2^2

# Estimate dispersion coefficients
y1 <- estimateDisp( y, robust = TRUE ) # Estimate mean dispersal

# Plot tagwise dispersal and impose w/ mean dispersal and trendline
plotBCV( y1 )

```

|           The above figure is an output of the plotBCV() function in edgeR, which visualises the biological coefficient of variation (otherwise known as dispersion) parameter fitted to individual genes (black circles), fitted across gene expression level (blue trendline), and averaged across the entire transcriptome (red horizontal line). The BCV/dispersion parameter is a necessary parameter, symbolized as $\theta$ in model specification, that must be estimated in negative binomial GLMs. $\theta$ represents the 'overdispersion' or the shape of a negative binomial distribution and is thus important for defining the distribution of read count data for a given gene during model fitting. Dispersion or $\theta$ can be input during model fitting in edgeR using the three estimates visualized in the above regression.

|           Now that we've estimated dispersion, we will use these estimates to fit negative binomial GLMs to our read count data using the glmQLFit() function in edgeR, one of the more robust model fitting functions offered by edgeR.

```{r, message = FALSE, warning = FALSE}
# Fit multifactorial design matrix
design_nl <- model.matrix( ~1 + pCO2_2 + pCO2 + pCO2_2:day + pCO2:day ) # Generate multivariate edgeR glm

# Fit quasi-likelihood, neg binom linear regression
nl_fit <- glmQLFit( y1, design_nl ) # Fit multivariate model to counts

## Test for effect of pCO2 and pCO2^2
nl_pCO2_2 <- glmQLFTest( nl_fit, coef = 2, contrast = NULL, poisson.bound = FALSE ) # Estimate significant DEGs
nl_pCO2 <- glmQLFTest( nl_fit, coef = 3, contrast = NULL, poisson.bound = FALSE ) # Estimate significant DEGs

# Make contrasts
is.de_nl_pCO2 <- decideTestsDGE( nl_pCO2, adjust.method = "fdr", p.value = 0.05 ) 
is.de_nl_pCO2_2 <- decideTestsDGE( nl_pCO2_2, adjust.method = "fdr", p.value = 0.05 )

# Summarize differential expression attributed to pCO2 and pCO2^2
summary( is.de_nl_pCO2 )
summary( is.de_nl_pCO2_2 )

```

|           We have just fit our first GLM to our read count data and have tested for differential expression across *p*CO$_{2}$. At this stage, it is important to output a few diagnostic plots. For example, edgeR has the function 'plotMD()' which, when input with a differential expression test object such as a glmQLFTest result, will produce a plot of differential expression log2 fold change values across gene expression level (log2 counts per million or logCPM). logCPM is a major component of gene function and statistical power, and is a useful variable to plot in order to make initial assessments of differential expression results.

|           Visualizations of differential expression (logFC) across baseline logCPM can be produced by the plotMD() (short for "mean-difference") function in edgeR. To do so, plotMD() is input with the results of glmQLFTest, which we used above to test for differential expression attributable to the parameters *p*CO$_{2}$ and *p*CO$_{2}$^2. Let's produce two such plots for both parameters where "pCO2_2" represents the *p*CO$_{2}$^2 parameter:

```{r, message = FALSE, warning = FALSE}
# Plot differential expression due to pCO2 and pCO2^2
plotMD( nl_pCO2 )
plotMD( nl_pCO2_2 )

```

|           In the two mean-difference plots above, large colored points represent genes exhibiting significant differential expression attributed to *p*CO$_{2}$ or *p*CO$_{2}$^2 below. As you can see, there is a lot of differential expression in this experiment resulting from both of these parameters. Part of the reason for this large number of significant differentially expressed genes (DEGs) is that we have not applied a cutoff value for logFC. For example, a logFC cutoff of 2.0 value would not designate any genes with an absolute logFC value less than 0.05 as a significant DEG. However, deciding on a logFC cutoff is very tricky! Since we used a continuous predictor for *p*CO$_{2}$ during model fitting, the values in the y-axis of this plot are slopes representing the rate of change in expression per unit *p*CO$_{2}$ ($\mu$atm). Determining what slope represents a significant change in expression requires informed reasoning and a strong body of prior data. For that reason, we are not applying a logFC cutoff in this walkthrough.

|           Next, we will use glmQLFTest() to test for differential expression attributed to interactions between time and *p*CO$_{2}$ or time and *p*CO$_{2}$^2 before producing mean-difference plots visualizing DEGs derived from these two interactions:

```{r}

## Test for interactions between time and pCO2 or pCO2^2
nl_pCO2_int <- glmQLFTest( nl_fit, coef = 4, contrast = NULL, poisson.bound = FALSE ) # Estimate significant DEGs
nl_pCO2_2_int <- glmQLFTest( nl_fit, coef = 5, contrast = NULL, poisson.bound = FALSE ) # Estimate significant DEGs

# Make contrasts
is.de_nl_pCO2_int <- decideTestsDGE( nl_pCO2_int, adjust.method = "fdr", p.value = 0.05 ) # Make contrasts
is.de_nl_pCO2_2_int <- decideTestsDGE( nl_pCO2_2_int, adjust.method = "fdr", p.value = 0.05 )

# Summarize differential expression attributed to pCO2 and pCO2^2
summary( is.de_nl_pCO2_int )
summary( is.de_nl_pCO2_2_int )

# Plot differential expression due to pCO2 and pCO2^2
plotMD( nl_pCO2_int )
plotMD( nl_pCO2_2_int )

```

|           In the mean-difference plots above, it is clear that there is remarkably less differential expression associated with interactions between time and *p*CO$_{2}$ or *p*CO$_{2}$^2 compared to the direct effects of *p*CO$_{2}$ and *p*CO$_{2}$^2. This is expected, as the statistical power necessary for detecting such interactions is lesser and there are grounds to expect that interactions would be less frequent during our experiment.

|           Taking a detour from tests for DEGs attributed to non-linear interactions, we can now output the residuals of our GLMs in order to better test for whether our data fit the assumptions of negative binomial distribution families. If the distribution of residuals from our GLMs are normal, this indicates that our data meet the assumption of the negative binomial distribution. We will use the equation for estimating Pearson residuals: $$ residual = \frac{observed - fitted} {\sqrt{fitted(dispersion*fitted)}} $$

```{r, message = FALSE, warning = FALSE}

# Output observed
y_nl <- nl_fit$counts

# Output fitted
mu_nl <- nl_fit$fitted.values

# Output dispersion or coefficient of variation
phi_nl <- nl_fit$dispersion

# Calculate denominator
v_nl <- mu_nl*(1+phi_nl*mu_nl)

# Calculate Pearson residual
resid.pearson <- (y_nl-mu_nl) / sqrt(v_nl)

# Plot distribution of Pearson residuals
ggplot(data = melt(as.data.frame(resid.pearson)), aes(x = value)) +
  geom_histogram( fill = "grey" ) +
  xlim(-2.5,5.0) +
  theme_classic() +
  labs(title = "Distribution of negative binomial GLM residuals", 
       x = "Pearson residuals", 
       y = "Density")

```

|           Our residuals appear to be normally distributed, indicating that our data fit the negative binomial distribution assumed by the GLM.

|           Now let's get back to analyzing non-linear effects on gene expression!

##Plotting non-linear effects

|           A variety of non-linear patterns may underly the significant effects we've observed above. For example, interactions between the vertex of expression and time may be attributed to variation in expression peaks across convex curves or the rate of decline in an exponentially decreasing curve among other patterns. Below we will assign transcripts exhibiting significant non-linear variation across *p*CO$_{2}$ to different groups based on the values of parameters such as $\beta_{1}$ and $\beta_{2}$. Then we will plot the expression of these grouped transcripts across *p*CO$_{2}$ and time to visualize different sources of non-linear variation.

```{r, message = FALSE, warning = FALSE}

## Bin transcripts based on (i) whether they have a significant positive or negative vertex and then (ii) whether they showed significant interactions between beta1 (vertex value) and time.

# Export diff expression data for transcripts with significant DE associated with PCO2^2 parameter
nl_pCO2_2_sig <- topTags(nl_pCO2_2, n = (11091 + 11915), adjust.method = "BH", p.value = 0.05)
nl_pCO2_2_sig_geneids <- row.names(nl_pCO2_2_sig) #Output a list of geneids associated with sig PCO2^2 effect

nl_pCO2_sig <- topTags(nl_pCO2, n = (14166 + 13376), adjust.method = "BH", p.value = 0.05)
nl_pCO2_sig_geneids <- row.names(nl_pCO2_sig) #Output a list of geneids associated with sig PCO2 effect

# Create tabulated dataframe of mean expression across each pCO2 level with metadata for transcript ID and timepoint
logCPM_df <- as.data.frame(df_log)

# Create tabularized df containing all replicates using 'melt' function in reshape2
logCPM_df$geneid <- row.names(logCPM_df)
tab_exp_df <- melt(logCPM_df,
                   id = c("geneid"))

# Add covariate information for time and pCO2
tab_exp_df$pCO2 <- substr(tab_exp_df$variable, 1, 3)
tab_exp_df$time <- as.numeric(substr(tab_exp_df$variable, 5, 5))

# Correct pCO2s to exact values
tab_exp_df$pCO2 <- as.numeric(
  ifelse(tab_exp_df$pCO2 == "300", 255,
         ifelse(tab_exp_df$pCO2 == "900", 930, 
                518)))

# Correct time to exact values
tab_exp_df$time <- as.numeric(
  ifelse(tab_exp_df$time == "1", 0.5, 7))

# Create binary variable in df_all_log for significant non-linear expression
tab_exp_df$pCO2_2_sig <- ifelse(tab_exp_df$geneid %in% nl_pCO2_2_sig_geneids, "Yes", "No")
tab_exp_df$pCO2_sig <- ifelse(tab_exp_df$geneid %in% nl_pCO2_sig_geneids, "Yes", "No")

# Create a binary variable related to up or down-regulation
up_genes <- filter(nl_pCO2_sig$table, logFC > 0)
tab_exp_df$logFC_dir <- ifelse(tab_exp_df$geneid %in% row.names(up_genes), "Up", "Down")

# Add geneid to nl_pCO2_int$coefficients
nl_pCO2_int$coefficients$geneid <- row.names(nl_pCO2_int$coefficients)

# Estimate average logCPM per gene per timepoint
tab_exp_avg <- summarySE(measurevar = "value",
                         groupvars = c("pCO2", "time", "geneid", "pCO2_sig", "pCO2_2_sig", "logFC_dir"),
                         data = tab_exp_df)

# First exploratory plot of non-linear expression grouping by exposure time and direction of differential expression
ggplot(data = filter(tab_exp_avg, pCO2_2_sig == "Yes"), 
       aes(x = pCO2, y = value)) +
  geom_path(alpha = 0.01, size = 0.25, stat = "identity", aes(group = as.factor(geneid))) +
    facet_grid(logFC_dir~time) +
  theme_classic() +
  theme(strip.background = element_blank()) +
  labs(y = "Avg logCPM", title = "Non-linear changes in GE output by edgeR")

```

|           Our plot of gene expression across tens of thousands of transcripts exhibiting significant non-linear changes across *p*CO$_{2}$ appears to include many trends that appear... well, linear. This is a pervasive issue in modeling non-linear regressions, and one potential pitfall of using outputs from packages such as edgeR or DESeq2 alone in testing for non-linear effects. The FDR-adjusted $p$-values we have used determine significance of non-linear effects essentially tell us the probability that a parameter value equal to or greater to what we have fitted could be generated given a random distribution of read counts. The $p$-value is not a representation of the strength of a non-linear effect relative to a linear effect. Numerous genes that nominally show significant non-linear effects of *p*CO$_{2}$ may be only weakly affected, and a linear effect may in fact be more probable than a non-linear one despite what our p-values tell us. Instead of asking "for what genes may there be significant, non-linear effects of *p*CO$_{2}$?", we should ask "for what genes *should* we test for significant, non-linear effects?". 
|           One of the best ways to determine whether a non-linear model is appropriate for a transcript is to determine whether it is more probable that its expression is linear or non-linear relative to a continuous predictor. We can calculate this relative probability using a likelihood ratio test (LRT). In the code chunk below, we will fit linear and 2nd order non-linear models to the expression of each gene before applying LRTs to each transcript. We will then further filter our edgeR dataset based on (i) significant differential expression attributed to non-linear effects and (ii) a significant LRT ratio supporting non-linear effects. Then, we will replot the expression levels of this re-filtered set. The code below fits gaussian linear models to log2-transformed CPM values, but can be adjusted to fit negative binomial GLMs to untransformed CPM similar to edgeR and DESeq2 by setting using the MASS package to set 'family = negative_binomial(theta = $\theta$)' where $\theta$ = the dispersion estimate or biological coefficient of variation for a given transcript.

```{r, message = FALSE, warning = FALSE}

## Using dlply, fit linear and non-linear models to each gene
# Create pCO2^2 variable in df_all_log
tab_exp_df$pCO2_2 <- tab_exp_df$pCO2^2

# Filter tabularized logCPM df for read depth of logCPM >= 0.5 in > 75% of samples

# Fit linear models - should take about 4 minutes
lms <- dlply(tab_exp_df, c("geneid"), function(df) 
lm(value ~ pCO2 + time + pCO2:time, data = df))

# Fit non-linear models - should take about 2 minutes
nlms <- dlply(tab_exp_df, c("geneid"), function(df) 
lm(value ~ pCO2 + pCO2_2 + time + pCO2:time + pCO2_2:time, data = df))

# Output nlm coefficients into dataframe
nlms_coeff <- ldply(nlms, coef)
head(nlms_coeff)

## Apply LRTs to lm's and nlm's for each transcript - should take about 2 minutes
lrts <- list() # Create list to add LRT results to

for (i in 1:length(lms)) {
 lrts[[i]] <- lrtest(lms[[i]], nlms[[i]]) # Apply LRTs with for loop
}

## Filter lrt results for transcripts with significantly higher likelihoods of nl model
lrt_dfs <- list()

# Turn list of LRT outputs into list of dataframes containing output info
for (i in 1:length(lrts)) {
 lrt_dfs[[i]] <- data.frame(lrts[i])
}

# Create singular dataframe with geneids and model outputs for chi-squared and LRT p-value
lrt_coeff_df  <- na.omit(bind_rows(lrt_dfs, .id = "column_label")) # na.omit removes first row of each df, which lacks these data

# Add geneid based on element number from original list of LRT outputs
lrt_coeff_df <- merge(lrt_coeff_df,
                      data.frame(geneid = names(nlms),
                                 column_label = as.character(seq(length(nlms)))),
                      by = "column_label")

# Apply FDR adjustment to LRT p-values before filtering for sig non-linear effects
lrt_coeff_df$FDR <- p.adjust(lrt_coeff_df$Pr..Chisq., method = "fdr")

# Filter LRT results for sig FDR coeff... produces 162 genes
lrt_filt <- filter(lrt_coeff_df, FDR < 0.05)

## Plot sig nl genes according to LRT, grouped by timepoint and direction of beta 1 coefficient
# Add beta coefficients to logCPM df
pCO2_pos <- filter(nlms_coeff, pCO2 > 0)
pCO2_2_pos <- filter(nlms_coeff, pCO2_2 > 0)

# Bin genes based on positive or negative pCO2 and pCO2^2 betas
tab_exp_avg$pCO2_binom <- ifelse(tab_exp_avg$geneid %in% pCO2_pos$geneid, "Positive", "Negative")
tab_exp_avg$pCO2_2_binom <- ifelse(tab_exp_avg$geneid %in% pCO2_2_pos$geneid, "Concave", "Convex")

# Filter for how many gene id's with significant likelihood of nl effect in LRT
LRT_filt_df <- filter(tab_exp_avg, geneid %in% lrt_filt$geneid)

# Plot
ggplot(data = LRT_filt_df, 
       aes(x = pCO2, y = value)) +
  geom_path(alpha = .25, size = 0.25, stat = "identity", aes(group = as.factor(geneid))) +
    facet_grid(pCO2_2_binom~time) +
  geom_smooth(method = "loess", se = TRUE, span = 1) +
  theme_classic() +
  theme(strip.background = element_blank()) +
  labs(y = "Avg logCPM", title = "Non-linear changes in GE output by LRTs")

# Count how many gene id's with significant likelihood of nl effect in LRT... 162 genes
nrow(as.data.frame(unique(LRT_filt_df$geneid)))

```

|           The above plot, which visualizes the expression of 162 genes identified in LRTs to exhibit significant and high likelihoods of non-linear variation in logCPM across *p*CO$_{2}$, appears distinct from our first plot of non-linear expression deemed significant by edgeR. The overall trends appear to be more convex or concave, consistent with a second-order polynomial regressions. This speaks to the strength of LRTs in identifying probable, non-linear effects or other multifactorial effects.
|           Let's see how many non-linear transcripts identified in LRTs exhibited significant non-linear effects within our edgeR models. To do this, we'll filter down the significant DEGs output by edgeR to include gene id's within our significant LRT output. Then, we'll plot the expression of this filtered set across *p*CO$_{2}$.

```{r, message = FALSE, warning = FALSE}

# Filter down df for gene id's exhibit pCO2 significant effect in edgeR and significant likelihood of nl effect in LRT 
edgeR_LRT_df <- filter(tab_exp_avg, geneid %in% lrt_filt$geneid & pCO2_sig == "Yes" |
                         geneid %in% lrt_filt$geneid & pCO2_2_sig == "Yes")

# Plot 
ggplot(data = edgeR_LRT_df, 
       aes(x = pCO2, y = value)) +
  geom_path(alpha = .25, size = 0.25, stat = "identity", aes(group = as.factor(geneid))) +
  facet_grid(pCO2_2_binom~time) +
  geom_smooth(method = "loess", se = TRUE, span = 1) +
  theme_classic() +
  theme(strip.background = element_blank()) +
  labs(y = "Avg logCPM", title = "Non-linear changes in GE output by edgeR & LRTs")

# Count how many gene id's exhibit pCO2 significant effect in edgeR and significant likelihood of nl effect in LRT... 89 genes
nrow(as.data.frame(unique(edgeR_LRT_df$geneid)))

```

|         89 genes exhibited a significant effect of *p*CO$_{2}$ or *p*CO$_{2}$^2 and a high likelihood of a non-linear as identified by an LRT. This number is substantially lower than the tens of thousands of genes identified by edgeR alone and the 162 identified via LRT. Additionally, there only appears to be partial overlap between edgeR and LRTs, speaking to the difference in the information you can extract from a significant effect identified in a generalized linear model (e.g., edgeR) vs. likelihood-based approaches such as an LRT. Because these two tests provide distinct and valuable information, the most robust estimate of we have laid out for non-linear effects of *p*CO$_{2}$ on gene expression comes from filtering for significant DEGs in edgeR *and* significant LRTs. 
|         What combination of approaches are most informative ultimately depends on your experimental design and the question you are addressing. For example, we can explore an additional question that our previous graphs did not touch on: how does gene expression vary between exposure times for genes exhibiting significant interactions between *p*CO$_{2}$^2 and time? Let's produce an exploratory plot of non-linear expression across 0.5 and 7 days of exposure for such genes identified using edgeR alone.

```{r, message = FALSE, warning = FALSE}

# Export diff expression data for transcripts with significant DE associated with interaction between PCO2^2 and time
nl_pCO2_2_int_sig <- topTags(nl_pCO2_2_int, n = (390 + 271), adjust.method = "BH",p.value = 0.05)
nl_pCO2_2_int_sig_geneids <- row.names(nl_pCO2_2_int_sig) #Output a list of geneids associated with sig PCO2^2 x time interaction

# Filter down df for gene id's exhibit pCO2 significant effect in edgeR and significant likelihood of nl effect in LRT 
edgeR_interaction_df <- filter(tab_exp_avg, geneid %in% nl_pCO2_2_int_sig_geneids )
edgeR_interaction_df$gene_id_time <- paste(edgeR_interaction_df$geneid,
                                           edgeR_interaction_df$time,
                                           sep = "_")

# Average logCPM across different groups according to pCO2^2 estimate and time
edgeR_interaction_avg <- summarySE(measurevar = "value",
                                   groupvars = c("time", "pCO2", "pCO2_2_binom"),
                                                 data = edgeR_interaction_df)

# Plot 
ggplot(data = edgeR_interaction_avg, 
       aes(x = pCO2, y = value, color = as.factor(time), group = as.factor(time))) +
  geom_path(stat = "identity") +
  geom_errorbar(aes(ymin = value - se, ymax = value + se), width = 0) +
  geom_point() +
  facet_wrap(~pCO2_2_binom) +
  theme_classic() +
  theme(strip.background = element_blank()) +
  labs(y = "logCPM", color = "Time (days)", title = "Interactions between pCO2^2 and exposure time")

```

|         After visualizing variation in expression across *p*CO$_{2}$ for genes identified in edgeR to be affected by interactions between *p*CO$_{2}$^2 and exposure time, we can see interesting changes in gene expression curves across time. For example, genes that exhibited a convex expression curve at 0.5 days of exposure generally modified their curve such that their expression became positively exponential relative to *p*CO$_{2}$. One potential explanation for this change is that the upregulation of some genes under intermediate *p*CO$_{2}$ was reduced following acclimation, while less acclimation may have occurred at the higher *p*CO$_{2}$ level after 7 days. Genes with a concave expression curves across *p*CO$_{2}$ at 0.5 days of exposure exhibited a marked decrease in expression as the experiment progressed, regardless of *p*CO$_{2}$ level. Thus, the interaction between *p*CO$_{2}$^2 and time may affect 'convex' and 'concave' genes in entirely different ways, underscorign the importance of partitioning or binning your data by parameters such as *p*CO$_{2}$^2. Exploring your data in this way is important for understanding interactive effects in RNA-seq analyses, regardless of whether or not you are fitting non-linear or linear continuous effects. Below we discuss further considerations for modeling interactions between linear continuous predictors.

#Interactive effects

|         Interactive effects shaping gene expression are common in nature and are becoming increasingly prevalent in models of gene expression derived from experimental studies. Below, we go a little bit deeper on interactive effects by outlining methods for fitting them using categorical and continuous variables in models of expression. We provide examples in edgeR, DESeq2, and Voom and point out how specification for interactive effects differs between these packages. Lastly, we compare correlations between these programs' fold change (logFC) predictions and test statistics. This final comparison of model predictions for interaction parameters is as much a tutorial on interactive effects as it is a tutorial on how to visualize differences in DE packages' results. Please use section of the code for any comparisons between DE packages you may want to make for other effect types, not just interactive effects!

##Interactive effects: edgeR

|         edgeR and Voom both take the same syntax for interactive effects, which we define below using the model.matrix() function as 'design_multi <- model.matrix( ~1 + pCO2 + pCO2:day )'. Then, we will use this multifactorial model design in both edgeR and Voom as specified below:

```{r, message = FALSE, warning = FALSE}

# Fit multifactoria design matrix that includes an interaction term for pCO2 x day
design_multi <- model.matrix( ~1 + pCO2 + pCO2:day ) #Generate multivariate edgeR glm

# Fit quasi-likelihood, neg binom linear regression
multi_fit <- glmQLFit( y1, design_multi ) # Fit multivariate model to counts

# Test for effect of pCO2
tr_pCO2 <- glmQLFTest( multi_fit, coef = 2, contrast = NULL, poisson.bound = FALSE ) # Estimate significant DEGs

is.de_tr_pCO2 <- decideTestsDGE( tr_pCO2, adjust.method = "fdr", p.value = 0.05 ) # Make contrasts

summary( is.de_tr_pCO2 )

# Test for interaction between pCO2 and time
tr_int <- glmQLFTest( multi_fit, coef = 3, poisson.bound = FALSE ) # Estimate significant DEGs

is.de_int <- decideTestsDGE( tr_int, adjust.method = "fdr", p.value = 0.05 ) # Make contrasts

summary( is.de_int )

```

##Interactive effects: limma-Voom

|         Below we will fit the same 'design_multi' model (y ~ 1 + *p*CO$_{2}$ + *p*CO$_{2}$:days) to our read counts using Voom rather than edgeR. As stated earlier, Voom fits interactions between fixed effects using the same syntax as edgeR, so the 'design_multi' object can be used by both programs.

|         What *is* different between edgeR and Voom is the manner by which Voom accounts for variation between replicates. Rather than using gene-wise or averaged estimations of biological coefficients of variation and inputting these values as $\theta$ in a negative binomial GLM, Voom models what is referred to as the mean-variance relationship and incorporates this relationship within a linear model rather than a GLM by "weighting" the accuracy of a given observation based on its level of expression and modelled variance. This mean-variance relationship can be plotted by Voom after model fitting as we show below:

```{r, message = FALSE, warning = FALSE}

# Perform Voom transformation
Voom <- voom( y, design_multi, plot = T )

```

|         When deciding between packages for testing DGE, it can be helpful to compare model assumptions regarding variance by plotting graphs like Voom's mean-variance relationship above and the output of edgeR's 'plotBCV()' function, which we provided an example of earlier and will replot below for the sake of making thsi visual comparison. Take note that the y-axes of these variance plots in Voom (square root of standard deviation) and edgeR (biological coefficient of variation) are different:

```{r, message = FALSE, warning = FALSE, include = FALSE}

# Replot edgeR BCV
plotBCV( y1 )

```

|         The mean-variance relationships modelled in edgeR and Voom are generally similar, but it appears that Voom assumes even lesser variance in expression among genes at the highest expression level. One result of this may be that greater statistical power is assumed in tests of differential expression by Voom for genes with high expression relative to low expression genes. Keep this in mind as we continue to move forward with fitting parameters for interactions between *p*CO$_{2}$ and time in Voom.

```{r, message = FALSE, warning = FALSE}
# Fit using Voom
lm_Voom_fit <- lmFit( Voom, design_multi )

# Create a contrast across continuous pCO2 variable
cont_pCO2 <- contrasts.fit( lm_Voom_fit, coef = "pCO2" )

# Create a contrast across interaction etween continuous pCO2 and time variables
cont_pCO2_day <- contrasts.fit( lm_Voom_fit, coef = "pCO2:day" )

# Perform empirical Bayes smoothing of standard errors
cont_pCO2 <- eBayes( cont_pCO2 )
cont_pCO2_day <- eBayes( cont_pCO2_day )

# Output test statistics
pCO2_results <- topTable( cont_pCO2, coef = "pCO2", adjust.method="fdr", n = Inf )
pCO2_day_results <- topTable( cont_pCO2_day, coef = "pCO2:day", adjust.method="fdr", n = Inf )

# How many DEGs are associated with pCO2
length( which( pCO2_results$adj.P.Val < 0.05  ) ) # number of DE genes =

# How many DEGs are associated with pCO2:day?
length( which( pCO2_day_results$adj.P.Val < 0.05  ) )

```

|         Voom has identified 1628 genes whose variation is affected by an interaction between *p*CO$_{2}$ and time, compared to 1748 identified by edgeR.

##Interactive effects: DESeq2

|         DESeq2 requires a slightly different syntax for specifying interactive effects compared to edgeR and Voom.

```{r, message = FALSE, warning = FALSE}

gcounts <- as.data.frame(data_input)

totalCounts <- colSums(gcounts)

### REMOVE GENES WITH LOW MEAN COUNTS ###

# Make a DGEList object for edgeR
y <- DGEList(counts = data_input, remove.zeros = TRUE)

# Let's remove samples with less then 0.5 cpm (this is ~10 counts in the count file) in fewer then 9/12 samples
keep_g <- rowSums(cpm( gcounts ) > .5) >= 9

table(keep_g)

# Set keep.lib.sizes = F and recalculate library sizes after filtering
#gcounts <- gcounts[ keep_g, keep.lib.sizes = FALSE ]

### BUILD A DATAFRAME ASSOCIATING SAMPLE NAMESWITH TREATMENT CONDITIONS ###
targets

### WALD TEST - FULL MODEL ###

dds <- DESeqDataSetFromMatrix(gcounts,
                              colData = targets,
                              design = formula( ~ 1 + pCO2 + day : pCO2))

rld <- rlog(dds)
rld.df <- assay(rld)

# Wald test for pCO2:day
dds_int <- DESeq(dds, minReplicatesForReplace = Inf)

design <- design( dds_int )

DESeq2_int_result_names <- resultsNames(dds_int)

# Count DEGs due to interaction
DESeq2_int_results <- results(dds_int, name = "pCO2.day", lfcThreshold = 0, alpha = 0.05)

summary(DESeq2_int_results)

```

|         DESeq2 has identified 4985 genes whose variation is affected by an interaction between *p*CO$_{2}$ and time, compared to 1748 identified by edgeR and 1628 by Voom. This is a ~3-fold greater number of DEGs predicted by DESeq2 compared to edgeR and Voom.

#Comparing test statistics: interactive effects

|         Differential expression packages can make dramatically different predictions for both the fold-change and probability of differential expression. If you find yourself deciding between different packages or wanting to compare how conservative different approaches are, it helps to run regressions of model predictions by different packages. Below we apply the ggpairs() function from GGAlly [@Schloerke2018-rq] to plot a correlation matrix of logFC values and negative log-transformed p-values attributed to interactive effects between time and *p*CO$_{2}$ predicted by edgeR, DESeq2, and Voom:

```{r, message = FALSE, warning = FALSE}

# Merge logFC and pval data from each program
Voom_edgeR_deseq_int_comp <- merge( 
  merge(
    data.frame(geneid = row.names(pCO2_day_results),
               Voom_logFC = pCO2_day_results$logFC,
               Voom_pval = pCO2_day_results$P.Value),
    data.frame(geneid = row.names(tr_int$table),
               edgeR_logFC = tr_int$table$logFC, #negate logFC because of syntax differences
               edgeR_pval = tr_int$table$PValue), 
    by = "geneid" ),
  data.frame(geneid = row.names(DESeq2_int_results),
             DESeq2_logFC = DESeq2_int_results$log2FoldChange,
             DESeq2_pval = DESeq2_int_results$pvalue),
  by = "geneid")

# Create neg log pvalues
Voom_edgeR_deseq_int_comp$Voom_neglogp <- -log(Voom_edgeR_deseq_int_comp$Voom_pval)
Voom_edgeR_deseq_int_comp$edgeR_neglogp <- -log(Voom_edgeR_deseq_int_comp$edgeR_pval)
Voom_edgeR_deseq_int_comp$DESeq2_neglogp <- -log(Voom_edgeR_deseq_int_comp$DESeq2_pval)

# Create function for adding 1:1 trendline on ggpairs plot and plotting geom_hex instead of geom_point
my_fn <- function(data, mapping, ...){
  p <- ggplot(data = data, mapping = mapping) + 
    geom_hex( bins = 100,
            aes(fill = stat(log(count))), alpha = 1 ) +
    scale_fill_viridis_c() +
    geom_abline(slope=1, intercept = 0, color = "red", lty = 2, size = 1, alpha = 0.5)
  p
}

# Correlation matrix of pvalues
pval_pairs <- ggpairs(data = Voom_edgeR_deseq_int_comp,
                      columns = c(8, 9, 10),
                      mapping = aes(alpha = 0.001),
                      lower = list(continuous = my_fn)) +
  labs(title = "Correlation matrix: interaction p-values")

pval_pairs +
  theme_classic(base_rect_size = 0)

```

|         The above plot is the output of ggpairs() showing correlations between negative log-transformed p-values attributed to interactive effects of time x *p*CO$_{2}$ estimated between each of the three packages, as well as the distributions of this parameter for each package, and correlation statistics. Red dashed lines represent a slope equal to 1. Yellow colors depict regions of each linear regression with more observations.

|         Voom and edgeR share the largest correlation between p-values estimated for an interactive effect of *p*CO$_{2}$ : days on differential expression (R$^2$ = 0.965). The bottom row of this ggpairs plot shows tranformed p-values from DESeq2 on the y-axis, which appear to be skewed toward lower p-values/higher -log p-values. This is consistent with the fact that we observed ~3-fold more DEGs attributed to the *p*CO$_{2}$ : days interaction when running DESeq2 compared to edgeR and Voom.

|         Next, let's create a ggpairs() plot contrasting logFC estimates of DE attributed to *p*CO$_{2}$ : days in edgeR, Voom, and DESeq2. One might think that logFCs should demonstrate stronger correlations between programs compared to p-values, but due to the manners by which read counts are normalized and transformed in different packages this is not always the case!

```{r, message = FALSE, warning = FALSE}

# Correlation matrix of logFC's
logFC_pairs <- ggpairs(data = Voom_edgeR_deseq_int_comp,
                       columns = c( 2, 4, 6 ),
                       mapping = aes( alpha = 0.001 ),
                       lower = list(continuous = my_fn)) +
  labs(title = "Correlation matrix: interaction logFC's")

logFC_pairs +
  theme_classic(base_rect_size = 0)

```

|         In this pairs plot of logFC estimates across packages, you can see that R$^2$ values are, on average, lower than those of p-value correlations between packages. Once again, the strongest correlation exists between edgeR and Voom. However, the slope of this correlation is less than 1, indicating that absolute edgeR logFC's are lower on average than absolute Voom logFCs. The correlation pair that shows a slope closest to 1.0 is edgeR vs. DESeq2. edgeR and DESeq2 normalize read counts in a similar manner, and thus a value derived from differences in read counts across continuous predictors such as logFC is likely to be more similar between these packages.

#Random Intercepts: example in Voom

We will skip edgeR and DESeq2 since they cannot fit random effects.

|         Random intercepts (often referred to as random effects) can be an essential component of estimating differential expression in experiments that incorporate non-independent observations (e.g., repeated measures and nested designs: individuals within populations or families) or randomly sampled groups. Most differential expression packages do not possess functionality for fitting random intercepts, extending GLMs to generalized linear mixed models or GLMMs, but Voom does! Additional packages that we don't cover here also possess random intercept functionality, including dream [@Hoffman2021-br] and [glmmSeq](https://cran.r-project.org/web/packages/glmmSeq/index.html).
|         Below is an example of how to fit a random intercept associated with the time variable of our dataset using Voom:

```{r, message = FALSE, warning = FALSE}

# Fit multifactoria design matrix
design_rand <- model.matrix( ~1 + pCO2 + ( 1 | day ) ) #Generate multivariate edgeR glm

# Perform Voom transformation
Voom_rand <- voom( y, design_rand, plot = T )

```

|         If you compare the mean-variance plot from our mixed model above, with our first mean-variance plot from the Voom model that did not include a random effect you will notice that the modelled relationship between variance and mean expression is considerably different between the two. While the linear model mean-variance plot took on a negative exponential shape by which earlier increases in expression level associate with a precipitous decline in variance, our mixed model's fit takes on a more logarithmic shape and the raw points appear parabolic. Here, variance is modelled as remaining somewhat constant during early increases in expression level before declining more rapidly across intermediate expression levels. Additionally, Voom predicted a maximum variance of 2.0 in our linear model while the mixed model above predicts a maximum of 1.5. This reduction in overall variance may partially be explained by incorporating a random intercept for time rather than modelling it as a continuous fixed effect. As we progress through the DE test derived this mixed model, we will anticipate that a reduction in variance will influence our results, likely by increasing the statistical power of genewise DE tests, and that modelling time as a random effect will also modify what and how many genes are identified as differentially expressed.

```{r, message = FALSE, warning = FALSE}

# Fit using Voom
lm_Voom_fit_rand <- lmFit( Voom_rand, design_rand )

# Create a contrast across continuous pCO2 variable
cont_rand_day <- contrasts.fit( lm_Voom_fit_rand, coef = "pCO2" )

# Perform empirical Bayes smoothing of standard errors
cont_rand_day <- eBayes( cont_rand_day )

# Output test statistics
rand_results <- topTable( cont_rand_day, coef = "pCO2", adjust.method = "fdr", n = Inf )

# How many DEG are associated with pCO2 after incorporating a random effect for day?
length( which( rand_results$adj.P.Val < 0.05  ) ) # number of DE genes

```

|         While the linear Voom model's DE test output 0 predicted DEG's associated with *p*CO${_2}$, the mixed Voom model predicted 3 genes. As we saw eariler when fitting linear models, the majority of DEG's in this study were derived from an interaction between time and *p*CO${_2}$. We modelled this as an interaction between two continuous fixed effects, but random effects and continuous effects can also jointly influence outome variables. Such effects can be fit in mixed models using 'random slopes', which we describe in the next section.

#Random slopes

|         Distinct from the random intercept, testing for variation in gene expression attributed to a random slope effect asks, "Within a set of groups, does the reaction norm of gene expression across a continuous predictor randomly by group?". This question is relevant to experimental designs testing for differences in the plasticity of gene expression across randomly-selected demographics such as genotypes or families. For these kinds of groups, modelling an interaction between the continuous predictor and a categorical effect representing different groups would be inappropriate, as the use of a categorical effect should be reserved for groups-of-interest specific selected in an experiment (e.g., different sexes, populations inhabiting distinct environments, etc.). To our knowledge however, there is no well-documented differential expression package that enables the fitting of random slopes.

|         Below we detail custom scripts for fitting gaussian linear mixed models using lme4 and negative binomial GLMs to read count data. For practical purposes, we will treat exposure time as a random variable even though timepoints were not randomly selected in the experiment that produced our data and apply. We will also apply linear mixed models using lmer, which is computationally intensive. The models fitted below are input with a highly filtered and reduced set of read count data. After fitting a LMM to each gene with a random slope parameter by which the effect of *p*CO${_2}$ randomly varies across time, we will compare the likelihoods of these models to those of a 'null' LMM that only possesses a random intercept associated with time. We will then count and report the number of genes that show significantly high likelihoods for random slopes.

```{r, message = FALSE, warning = FALSE}

## For practical purposes, reduce size of input data in order to run lmer in for loop without using up memory
# Re-filter data
keep_red <- rowSums(cpm(y) > 3 & cpm(y) < 10) >= 12

# Apply read filter to tab_exp_df
tab_exp_df_filt <- filter(tab_exp_df, geneid %in% 
                            row.names(filter(as.data.frame(keep_red), keep_red == TRUE)))

# Using dlply, fit linear mixed model to tabularized df of log2-transformed CPM values for each transcript
tab_exp_df_filt$time <- as.factor(tab_exp_df_filt$time)
rs_lmes <- dlply(tab_exp_df_filt, c("geneid"), function(df) 
lmer(value ~ pCO2 + (1 | time) + (pCO2 | time), data = df))

# Fit null model without random slope
null_lmes <- dlply(tab_exp_df_filt, c("geneid"), function(df) 
lmer(value ~ pCO2 + (1 | time), data = df))

# Apply LRT to each gene to test for effect of random slope
lmer_lrts <- list() # Create list to add LRT results to

for (i in 1:length(rs_lmes)) {
 lmer_lrts[[i]] <- lrtest(rs_lmes[[i]], null_lmes[[i]]) # Apply LRTs with for loop
}

## Filter lrt results for transcripts with significantly higher likelihoods than null model
lmer_lrt_dfs <- list()

# Turn list of LRT outputs into list of dataframes containing output info
for (i in 1:length(lmer_lrts)) {
 lmer_lrt_dfs[[i]] <- data.frame(lmer_lrts[i])
}

# Create singular dataframe with geneids and model outputs for chi-squared and LRT p-value
lmer_lrt_coeff_df  <- na.omit(bind_rows(lmer_lrt_dfs, .id = "column_label")) # na.omit removes first row of each df, which lacks these data

# Add geneid based on element number from original list of LRT outputs
lmer_lrt_coeff_df <- merge(lmer_lrt_coeff_df,
                      data.frame(geneid = names(rs_lmes),
                                 column_label = as.character(seq(length(rs_lmes)))),
                      by = "column_label")

# Apply FDR adjustment to LRT p-values before filtering for sig non-linear effects
lmer_lrt_coeff_df$FDR <- p.adjust(lmer_lrt_coeff_df$Pr..Chisq., method = "fdr")

# Filter LRT results for sig FDR coeff...
lmer_lrt_filt <- filter(lmer_lrt_coeff_df, FDR < 0.05)

# How many genes showed significant LRT?... none!
count(lmer_lrt_filt)

```

#Conclusion: Moving forward with multifactorial RNA-seq

|         If you've made it this far, you've probably found it to be a common theme that popular DE packages do not provide all of the flexibility necessary for fitting models to data coming from multifactorial experimental designs, even some of the more standard designs. It should be restated that our walkthrough only scratches the surface of (i) available packages that can facilitate multifactorial DE tests and (ii) opportunities that custom scripts for model fitting provide. For example, our example code for fitting linear and linear mixed models before testing for effects using likelihood ratio tests can easily be edited to fit negative binomial GLMs such as those used by edgeR and DESeq2 input with $\theta$ estimates (e.g., dispersion) from edgeR or custom estimates, and we encourage readers to try out this strategy! If you have personal code that you believe would make a valuable addition to this walkthrough, don't hesitate to post a GitHub issue or submit a pull-request to us. We will review the suggestions and add your name to the list this page's authors! We hope that this page and others on the website can serve as a living document that can keep pace with advances in multifactorial RNA-seq analyses, which lend fidelity to experimental designs that are necessary for generating reproducible and predictive results.

### 2022-06-06 DGE_comparison.Rmd
Push site rendered version of DGE_comparison.html

@@
 title: "Fitting Multifactorial Models of Differential Expression"
-author: "Sam Bogan"
-csl: molecular-ecology.csl
-output:
-  html_document:
-    toc: yes
-    toc_depth: 3
-  word_document:
-    toc: yes
-    toc_depth: '3'
-  pdf_document: default
-  github_document: default
+author: "Samuel N Bogan"
 bibliography: DGE_Comp.bib
@@
 We hope that this page and others on the website can serve as a living document that can keep pace with advances in multifactorial RNA-seq analyses, which lend fidelity to experimental designs that are necessary for generating reproducible and predictive results.
+
+#References

### 2022-06-24 DGE_comparison_v2.Rmd
Rendered site to include RNAseq in navigation, made minor edits to RADseq, changed to DEG_comparision_v2

--- 2022-06-06 DGE_comparison.Rmd
@@
+```{r,message = FALSE,echo = FALSE}
+library(knitr)
+knitr::opts_chunk$set(echo = TRUE)
+opts_chunk$set(fig.width = 10,
+               fig.height = 5,
+               cache = FALSE)
+```
 ---
@@
 
-#Intro to multifactorial RNA-seq models
+# Intro to multifactorial RNA-seq models
 
@@
 
-##Targeted audience
+## Targeted audience  
 |           This walkthrough is for both intermediate and expert practitioners of RNA-seq analyses, and most of all, those new to multifactorial RNA-seq.
@@
 
-##Variables we will cover
+## Variables we will cover
 
@@
 
-##Packages we will cover
+## Packages we will cover
 
@@
 
-#Features of popular DGE packages
+# Features of popular DGE packages
 
@@
 
-##Summary of approach
+## Summary of approach
 
@@
 ```{r setup, include = FALSE}
-
+# should remove from Rmd
 # Set root directory
-knitr::opts_knit$set( root.dir = '~/Documents/GitHub/DGE_comparison/' )
+#knitr::opts_knit$set( root.dir = '~/Documents/GitHub/DGE_comparison/' )
 
@@
 
-#invisible(lapply(c("DESeq2", "edgeR", "tidyverse", "ape", "vegan", "GGally",
-                   #"arrayQualityMetrics", "rgl", "adegenet", "MASS",
-                   #"data.table", "plyr", "lmtest", "reshape2", "Rmisc", "lmerTest"),
+#invisible(lapply(c( "tidyverse", "ape", "vegan", "GGally",
+                   #, "rgl", "adegenet", "MASS",
+                   #"data.table", "plyr", "lmtest", "reshape2", "Rmisc", "lmerTest","statmod"),
                  #function(p){
@@
                  #}))
+
+#if (!require("BiocManager", quietly = TRUE))
+    #install.packages("BiocManager")
+
+#BiocManager::install(c("DESeq2","edgeR","arrayQualityMetrics"))
 
@@
 
-#Filter and visualize read counts
+# Filter and visualize read counts
 
@@
 # Read in matrix of RSEM expected read counts
-data <- read.delim("Input_data/GE2_expected_counts_gene.matrix", sep = "\t", header = T, row.names = 1)
+data <- read.delim("RNA_data/GE2_expected_counts_gene.matrix", sep = "\t", header = T, row.names = 1)
 
@@
 
-##MDS plot visualizing experimental factors
+## MDS plot visualizing experimental factors
 
@@
 
-#Non-linear effects
+# Non-linear effects
 
@@
 
-##Non-linear effects: example in edgeR
+## Non-linear effects: example in edgeR
 
@@
 
-##Plotting non-linear effects
+## Plotting non-linear effects
 
@@
 
-#Interactive effects
+# Interactive effects
 
@@
 
-##Interactive effects: edgeR
+## Interactive effects: edgeR
 
@@
 
-##Interactive effects: limma-Voom
+## Interactive effects: limma-Voom
 
@@
 
-##Interactive effects: DESeq2
+## Interactive effects: DESeq2
 
@@
 
-#Comparing test statistics: interactive effects
+# Comparing test statistics: interactive effects
 
@@
 
-#Random Intercepts: example in Voom
+# Random Intercepts: example in Voom
 
@@
 
-#Random slopes
+# Random slopes
 
@@
 
-#Conclusion: Moving forward with multifactorial RNA-seq
+# Conclusion: Moving forward with multifactorial RNA-seq
 
@@
 
-#References
+# References

### 2022-06-24 RADseq.Rmd
Rendered site to include RNAseq in navigation, made minor edits to RADseq, changed to DEG_comparision_v2

@@
 ```
+
+# Setup for running code
 
@@
 |---|---|---|---|
-| Supported datatypes | Paired-end: ddRAD, ezRAD, RAD (random shearing), data with large overlap  between forward and reverse reads  
-Single-end: any RADseq method  de novo and reference-based.  
-If doing de novo assembly, reads cannot be trimmed outside of dDocent |  |  |
-| Unique aspects | Novel data reduction approach used to inform coverage cutoffs |  |  |
+| Supported datatypes | Paired-end: ddRAD, ezRAD, RAD (random shearing), data with large overlap between forward and reverse reads; Single-end: any RADseq method  de novo and reference-based.
+If doing de novo assembly, reads cannot be trimmed outside of dDocent | Wide range of RAD/GBS methods with paired/single end data, see [ipyrad documentation](https://ipyrad.readthedocs.io/en/master/4-data.html) | Paired-end and single-end data for single and double digest RAD and DART; GBS with single-end sequencing.
+NOT suitable for paired-end GBS.
+|
+| Unique aspects | Novel data reduction approach used to inform coverage cutoffs | Python API for popgene and phylogenetic analyses written specifically for RAD data |  |
 | Documentation | Very good (esp.
@@
 with higher levels of indel polymorphism), but can be addressed with downstream filtering |
-| Open source/development | Open source | Open source | no |
-| Filtering options | Minimal default filtering as implemented in VCFtools. Ideal for those who want full freedom in filtering their SNPs. |  |  |
-| Output options | Only produces a VCF, fasta of the de novo assembly, and individual BAM files  from mapping to the reference  |  |  |
-| Popgen analyses | No additional popgen analyses included  |  |  |
+| Open source/development | Open source | Open source, primarily combines existing software | No |
+| Filtering options | Minimal default filtering as implemented in VCFtools.
+Ideal for those who want full freedom in filtering their SNPs.
+| Wide range of filtering options  | Moderate range of filtering options |
+| Output options | Only produces a VCF, fasta of the de novo assembly, and individual BAM files  from mapping to the reference  | Lots of output formats (see [documentation](https://ipyrad.readthedocs.io/en/master/output_formats.html)) | Lots of Stacks-specific output and log files, as well as various inputs for popgen programs (see [documentation](https://catchenlab.life.illinois.edu/stacks/comp/populations.php)) |
+| Popgen analyses | No additional popgen analyses included  | Unique, flexible Python API implementing range of popgen and phylogenetic methods (see [documentation](https://ipyrad.readthedocs.io/en/master/API-analysis/index.html))   | Popgen summary statistics (F-statistics, pi), including kernel-smoothing along reference genome.
+Underlying equations and assumptions for these are not very clear.
+|
 |  |  |  |  |
@@
 To plot the missingness on the command line, you can use gnuplot:
-```{bash}
+```{bash, eval = T}
 #code from Jon Puritz
@@
 The parameters below are general across RADseq pipelines (although with different names), and are shown to have the largest impact on genoyping error rates.
+
 ### Clustering threshold {#cluster}  
@@
 `r colorize("This section is still under development! Please contribute on Github or provide feedback in [Discussions](https://github.com/MarineOmics/marineomics.github.io/discussions)", "red")`
+
 Even with a reduced representation sequencing approach, your assembly and genotyping pipeline will generate data for thousands of SNPs across your individuals.
 These SNPs will always need to be filtered before doing downstream analyses.
+
 #### Main reasons to filter your data:

### 2022-06-24 choosing-population-genetics.Rmd
Rendered site to include RNAseq in navigation, made minor edits to RADseq, changed to DEG_comparision_v2

--- 2021-12-28 choosing-population-genetics.Rmd
@@
 library(kableExtra)
-opts_chunk$set(fig.width = 10,
+knitr::opts_chunk$set(fig.width = 10,
                fig.height = 5,

### 2022-06-24 index.Rmd
Push index.rmd and knit with multifac RNAseq page
Push knitted index.html after formatting DGE_comparison.html
Rendered site to include RNAseq in navigation, made minor edits to RADseq, changed to DEG_comparision_v2

@@
 
-* Transcriptomics with RNA-Seq (coming soon!)  
-* Differential gene expression (coming soon!)
+* [Multifactorial RNAseq analyses](DGE_comparison_v2.html)  
+* Align and count RNAseq reads (coming soon!)
+    * De novo, reference-based, and Tag-seq methods
 
@@
 
-#### [Katherine Silliman, PhD](https://ksilliman.weebly.com/), Auburn University - kes0132[at]auburn.edu {.unlisted .unnumbered}
+#### [Katherine Silliman, PhD](https://ksilliman.weebly.com/), South Carolina Dept. of Natural Resources - sillimank[at]dnr.sc.gov {.unlisted .unnumbered}
 #### [Sam Bogan](http://www.hofmannlab.com/sam-bogan.html), UC-Santa Barbara -snbogan[at]ucsb.edu {.unlisted .unnumbered}

### 2022-07-08 index.Rmd
Updated intro page

@@
 
-The MarineOmics working group is part of the [RCN for Evolution in Changing Seas](https://rcn-ecs.github.io/), and was formed to promote **reproducible** and **robust** genomic research in nonmodel species, but especially marine and aquatic studies.
-Methods in next generation sequencing and bioinformatics advance at a pace that quickly render any static set of best practices obsolete.
-MarineOmics aims to support robust genomic research in ecology and evolution by disseminating guidelines through this dynamic, open-sourced website addressing introductory and advanced topics of population- and functional genomics.
-
-**Rigorous**:  following best principles to identify and mitigate sources of error in data generation and analysis
-
-**Reproducible**: when others can reproduce the results with accessible data, code, and documentation
-
 # Guidelines for 'omics analyses
@@
 Instead of providing rigid "best practices", we aim to encourage data exploration and rigorous quality control through "best principles", paying particular attention to topics for which open access guidelines are scarce.
-The guidelines are written by members of the MarineOmics working group or the community, then shared for peer review.
+The guidelines are written by members of the [MarineOmics working group](#marineomics) or the community, then shared for peer review.
 Comments on current website content or suggestions for new content can be made through the [Discussion Forum](https://github.com/MarineOmics/marineomics.github.io/discussions).
@@
 
-### The guidelines portion of the website is still under peer review and should be considered a "preprint".
+### The guidelines portion of the website should be considered a "preprint".
 If you are interested in contributing to the peer review process, check out the [Discussion Forum](discussion.html).
@@
 
+# What is the MarineOmics working group? {#marineomics}
+
+The original MarineOmics working group is part of the [RCN for Evolution in Changing Seas](https://rcn-ecs.github.io/), and was formed to promote **reproducible** and **robust** genomic research in nonmodel species, but especially marine and aquatic studies.
+Methods in next generation sequencing and bioinformatics advance at a pace that quickly render any static set of best practices obsolete.
+MarineOmics aims to support robust genomic research in ecology and evolution by disseminating guidelines through this dynamic, open-sourced website addressing introductory and advanced topics of population- and functional genomics.
+*We are always looking for folks to join and write content for the website, if interested please [contact us](#contact)!*
+
+**Rigorous**:  following best principles to identify and mitigate sources of error in data generation and analysis
+
+**Reproducible**: when others can reproduce the results with accessible data, code, and documentation
+
 # Reproducibility in Marine Genomics  
-Marineomics is also interested in evaluating and improving reproducibility in studies that use genomics to study adaptation in marine systems.
+MarineOmics is also interested in evaluating and improving reproducibility in studies that use genomics to study adaptation in marine systems.
 We are conducting a meta-analysis of reproducibility across genomic studies of environmental adaptation in marine populations with the aim of identifying obstacles and opportunities for reproducing and synthesizing population genomic research.
@@
 
-# Working Group Members
+# Original Working Group Members
 
@@
 
-# Contact 
-For any and all inquiries about the MarineOmics Working Group, our website and its material, or other activities, please contact us at rcnecs.marineomics@gmail.com.
+# Contact {#contact}
+For questions about the website and its materials, post to the [Discussion Forum](https://github.com/MarineOmics/marineomics.github.io/discussions)!
+For inquiries about the MarineOmics Working Group activities, please contact us at rcnecs.marineomics@gmail.com.

### 2022-07-19 DGE_comparison_v2.Rmd
Your Message

@@
-```{r,message = FALSE,echo = FALSE}
-library(knitr)
-knitr::opts_chunk$set(echo = TRUE)
-opts_chunk$set(fig.width = 10,
-               fig.height = 5,
-               cache = FALSE)
-```
 ---
 title: "Fitting Multifactorial Models of Differential Expression"
-author: "Samuel N Bogan"
+subtitle: '*Samuel N. Bogan*'
 bibliography: DGE_Comp.bib
@@
 
-|           Studies of molecular responses to environmental change increasingly employ multifactorial experimental designs incorporating different developmental stages, stressors, populations, or non-linear dynamics that resolve interactions between variables that shape expression, physiology, and performance.
+Studies of molecular responses to environmental change increasingly employ multifactorial experimental designs incorporating different developmental stages, stressors, populations, or non-linear dynamics that resolve interactions between variables that shape expression, physiology, and performance.
 After an experiment during which a study organism was exposed to abiotic variation and perhaps other variables, researchers interested in how gene expression correlates with these variables will often extract RNA across treatment groups, prepare cDNA libraries for sequencing, and conduct an analysis of differential expression (DE) across variables of interest after mapping and counting RNA-seq reads.
@@
 
+<br>
+
 ## Targeted audience  
-|           This walkthrough is for both intermediate and expert practitioners of RNA-seq analyses, and most of all, those new to multifactorial RNA-seq.
+This walkthrough is for both intermediate and expert practitioners of RNA-seq analyses, and most of all, those new to multifactorial RNA-seq.
 We have written this page with the assumption that readers have a basic familiarity with model fitting and statistical analysis, as well as next generation sequencing technologies, RNA-seq workflows, and the R statistical environment.
 
+<br>
+
 ## Variables we will cover
 
-|           This walkthrough can be broken down into different types of predictor variables that can be incorporated in models of gene expression.
-In the order we address them, these variables include:
+This walkthrough can be broken down into different types of predictor variables that can be incorporated in models of gene expression. In the order we address them, these variables include:
 
@@
 
+<br>
+
 ## Packages we will cover
@@
 
+<br>
+
 # Features of popular DGE packages
 
-|           Two important considerations when choosing a DE package to analyze RNA-seq data are the distribution family used by the package and the effect types it incorporate into a model.
+Two important considerations when choosing a DE package to analyze RNA-seq data are the distribution family used by the package and the effect types it incorporate into a model.
 The DE packages we will cover assume the following read count distributions and are capable of fitting the following effect types:
@@
 | Program | Distribution | Continuous fixed eff. | Random intercepts | Random slopes | Interactive eff. | Non-linear eff. |
-| :--- | :--- | :--- | :--- | :--- | :--- |
+| :--- | :--- | :--- | :--- | :--- | :--- | :--- |
 | edgeR | Negative binomial generalized linear model | ✔ | ✖ |  ✖ |  ✔ |  ✔ |  
@@
 
+<br>
+
 ## Summary of approach
 
-|           We will walk you through (i) how to conduct essential checks of your data such as principal components analysis and observing the distribution of filtered read counts, (ii) evaluating each package's functionality for multifactorial model fitting, and (iii) how to fit each type of predictor by performing differential expression tests using real data.
+We will walk you through (i) how to conduct essential checks of your data such as principal components analysis and observing the distribution of filtered read counts, (ii) evaluating each package's functionality for multifactorial model fitting, and (iii) how to fit each type of predictor by performing differential expression tests using real data.
 These data come from an experiment that exposed the Antarctic pteropod (a free-swimming planktonic snail) *Limacina helicina antarctica* to three continuous levels of ocean acidification (OA) for either 0.5 or 7 days [@Bogan2020-qs].
@@
 
-|           Throughout this walkthrough, we will provide examples of custom scripts for fitting linear, linear mixed models, generalized linear models to read count data using for loops in order to provide functionality where it cannot be met by most differential expression packages.
-
-|           At the end of the walkthrough, we demonstrate how to compare model predictions by different packages for different effect types commonly used in multifactorial RNAseq analyses.
-
-|           Let's start by loading our required packages...
+Throughout this walkthrough, we will provide examples of custom scripts for fitting linear, linear mixed models, generalized linear models to read count data using for loops in order to provide functionality where it cannot be met by most differential expression packages.
+
+At the end of the walkthrough, we demonstrate how to compare model predictions by different packages for different effect types commonly used in multifactorial RNAseq analyses.
+
+Let's start by loading our required packages...
+
+<br>
 
@@
 # Set root directory
-#knitr::opts_knit$set( root.dir = '~/Documents/GitHub/DGE_comparison/' )
+#knitr::opts_knit$set(root.dir = '~/Documents/GitHub/DGE_comparison/')
 
@@
 
-|           Before model fitting and testing for DE, it is important to visually inspect read counts pre- and post-filtering.
+Before model fitting and testing for DE, it is important to visually inspect read counts pre- and post-filtering.
 In this walk through, we will plot the distributions of read counts and sample loading to multidimensional scaling axes (e.g., plotting principal components or principal coordinates analyses).
@@
 Depending on the distribution of your data and the variables that appear important (or variable you would like to incorporate into your models), we will then demonstrate how to choose an appropriate package to test for DE.
-|           Most DE packages assume that read counts possess a negative binomial distribution.
+
+Most DE packages assume that read counts possess a negative binomial distribution.
 The negative binomial distribution is an extension of distributions for binary variables such as the Poisson distribution, allowing for estimations of "equidispersion" and "overdispersion", equal and greater-than-expected variation in expression attributed to biological variability .
@@
 Thus, the most accurate representation of how appropriate a negative binomial distribution is for read counts would come from plotting the distribution of residuals across all genes following model fitting.
-Below is a visual example of negative binomial distributions that possess different variances from modelassist[dot]epixanalytics[dot]com:
-
-```{r, echo = FALSE, out.height = "300%", fig.align = "center", }
-
-knitr::include_graphics("images/negbinom.png")
-
-```
-
-|           Now let's first plot the distribution of unfiltered reads from our dataset:
+Below is a visual example of negative binomial distributions that possess different means and variances:
+
+```{r, echo = FALSE, message = FALSE, warning = FALSE}
+
+## Simulate two negative binomial distributions of different variances
+# First, great function for doing so
+generate_distribution <- function(n, mus, sizes){
+  tibble(n = n, 
+       t_mu = mus, 
+       t_size = sizes)%>% 
+  mutate(rt = pmap(list(n, t_mu, t_size),
+                     ~rnbinom(n = ..1, mu = ..2, size = ..3))) %>% 
+  unnest(rt) 
+}
+
+set.seed(123)
+
+# Second, simulate distributions with chosen parameters
+rt_df <- 
+  generate_distribution(n = 2000,
+                        mus = 25,
+                        sizes = .5)
+
+rt_df <- rbind(rt_df,
+      generate_distribution(n = 2000,
+                        mus = 75,
+                        sizes = 15))
+
+# Plot distributions with histogram
+ggplot(data = rt_df, aes(x = rt, fill = as.factor(t_size), group = as.factor(t_size))) +
+  geom_histogram(data = filter(rt_df, t_mu == 25), alpha = 0.75) +
+  geom_histogram(data = filter(rt_df, t_mu == 75), alpha = 0.75) +
+  theme_classic() +
+  labs(title = "Simulated negative binomial distribution", fill = "Mean", y = "Frequency", x = "Mean")
+
+```
+
+Now let's first plot the distribution of unfiltered reads from our dataset:
+
+<br>
 
@@
 
-|           The table of read counts above has rows representing individual transcripts and columns representing different replicate samples across treatments.
+The table of read counts above has rows representing individual transcripts and columns representing different replicate samples across treatments.
+
+<br>
 
@@
 # Name experimental samples: names correspond to pCO2 treatment (300, 600, 900) + days (12 h2 vs 7 days) + replicate ID
-colnames(data) <- c( "300.7.a", "300.7.b", "300.7.c","300.12.a", "300.12.b", "300.12.c", 
+colnames(data) <- c("300.7.a", "300.7.b", "300.7.c","300.12.a", "300.12.b", "300.12.c", 
                     "900.7.a", "900.7.b", "900.7.c", "900.12.a", "900.12.b", "900.12.c", 
@@
 ## Create 'targets' and 'Group dataframe, expressing experimental variables for DEG analysis
-pCO2 <- as.numeric( c( 255, 255, 255, 255, 255, 255,
+pCO2 <- as.numeric(c( 255, 255, 255, 255, 255, 255,
                       530, 530, 530, 530, 530, 530,
-                       918, 918, 918, 918, 918, 918 ) )
-
-treatment <- c( "B", "B", "B", "B", "B", "B",
+                      918, 918, 918, 918, 918, 918))
+
+treatment <- c("B","B", "B", "B", "B", "B",
                "R", "R", "R", "R", "R", "R",
-                "Y", "Y", "Y", "Y", "Y", "Y" )
-
-day <- as.numeric( c( 7, 7, 7, .5, .5, .5,
+               "Y", "Y", "Y", "Y", "Y", "Y")
+
+day <- as.numeric(c(7, 7, 7, .5, .5, .5,
                     7, 7, 7, .5, .5, .5,
-                      7, 7, 7, .5, .5, .5 ) )
-
-targets <- data.frame( pCO2, day, treatment )
-targets$grouping <- paste( targets$pCO2, targets$day, sep = "." )
+                    7, 7, 7, .5, .5, .5))
+
+targets <- data.frame(pCO2, day, treatment)
+targets$grouping <- paste(targets$pCO2, targets$day, sep = ".")
 
 # The group factor represents the combined levels of an experimental replicate across all variables
-Group <- factor( paste( targets$day, targets$pCO2, sep = "_" ) )
+Group <- factor(paste(targets$day, targets$pCO2, sep = "_"))
 
 # Data must be rounded to nearest integer in order to be fit for negative binomial distribution
-data_input <- round( data )
+data_input <- round(data)
 
 # Peak at rounded data
-head( data_input )
+head(data_input)
 
 # Plot distribution of unfiltered read counts across all samples 
-ggplot( data = data.frame( rowMeans( data_input ) ),
-        aes( x = rowMeans.data_input. ) ) +
-  geom_histogram( fill = "grey" ) +
-  xlim( 0, 500 ) +
+ggplot(data = data.frame(rowMeans(data_input)),
+       aes(x = rowMeans.data_input.)) +
+  geom_histogram(fill = "grey") +
+  xlim(0, 500) +
   theme_classic() +
-  labs( title = "Distribution of unfiltered reads" ) +
-  labs( y = "Density", x = "Raw read counts", title = "Read count distribution: untransformed, unnormalized, unfiltered" )
-
-```
-
-|           As you can see in the above plot, the raw distribution of all read counts takes on a left-skewed negative binomial distribution similar to the purple distribution in the example above.
+  labs(title = "Distribution of unfiltered reads") +
+  labs(y = "Density", x = "Raw read counts",
+  title = "Read count distribution: untransformed, unnormalized, unfiltered")
+
+```
+
+As you can see in the above plot, the raw distribution of all read counts takes on a left-skewed negative binomial distribution similar to the purple distribution in the example above.
 While looking at the distribution of your raw reads is useful, these are not in fact the data you will be inputting to tests of differential expression.
@@
 
+<br>
+
 ```{r, message = FALSE, warning = FALSE}
@@
 # Make a DGEList object for edgeR
-y <- DGEList( counts = data_input, remove.zeros = TRUE )
+y <- DGEList(counts = data_input, remove.zeros = TRUE)
 
 #Let's remove samples with less than 0.5 cpm (this is ~10 counts in the count file) in fewer then 9/12 samples
-keep <- rowSums( cpm( y ) > .5 ) >= 9
-
-table( keep )
+keep <- rowSums(cpm(y) > .5) >= 9
+
+table(keep)
 
 # Set keep.lib.sizes = F and recalculate library sizes after filtering
-y <- y[ keep, keep.lib.sizes = FALSE ]
-
-y <- calcNormFactors( y )
+y <- y[keep, keep.lib.sizes = FALSE]
+
+y <- calcNormFactors(y)
 
 # Calculate logCPM
-df_log <- cpm( y, log = TRUE, prior.count = 2 )
+df_log <- cpm(y, log = TRUE, prior.count = 2)
 
 # Plot distribution of filtered logCPM values
-ggplot( data = data.frame( rowMeans( df_log ) ),
-        aes( x = rowMeans.df_log. ) ) +
-  geom_histogram( fill = "grey" ) +
+ggplot(data = data.frame(rowMeans(df_log)), 
+       aes(x = rowMeans.df_log.) ) +
+  geom_histogram(fill = "grey") +
   theme_classic() +
-  labs( y = "Density", x = "Filtered read counts (logCPM)", 
-        title = "Distribution of normalized, filtered read counts" )
-
-```
-
-|           Our raw data appear to follow a strong negative binomial distribution! Later, we will see whether this holds for residuals from fitted negative binomial GLMs.
+  labs(y = "Density", x = "Filtered read counts (logCPM)",
+       title = "Distribution of normalized, filtered read counts")
+
+```
+
+Our raw data appear to follow a strong negative binomial distribution! Later, we will see whether this holds for residuals from fitted negative binomial GLMs.
+
+<br>
 
@@
 
-|           Before analyzing our data, it is essential that we look at the multivariate relationships between our samples based on transcriptome-wide expression levels.
+Before analyzing our data, it is essential that we look at the multivariate relationships between our samples based on transcriptome-wide expression levels.
 Below is example code and output for a principal coordinates analysis (PCOA) plot that visualizes multifactorial RNA-seq replicates according to two predictor variables across major and minor latent variables or PCOA axes.
@@
 # Export pcoa loadings
-dds.pcoa = pcoa( vegdist( t( df_log <- cpm( y, log = TRUE, prior.count = 2 ) ),
-                          method = "euclidean") / 1000 )
+dds.pcoa = pcoa(vegdist(t(df_log <- cpm(y, log = TRUE, prior.count = 2)),
+                          method = "euclidean") / 1000)
 
@@
 percent <- dds.pcoa$values$Eigenvalues
-cumulative_percent_variance <- ( percent / sum( percent ) ) * 100
+cumulative_percent_variance <- (percent / sum( percent)) * 100
 
 # Prepare information for pcoa plot, then plot
-color <- c( "steelblue1", "tomato1", "goldenrod1")
-par( mfrow = c( 1, 1 ) )
-plot( scores[ , 1 ], scores[ , 2 ],
-      cex=.5, cex.axis=1, cex.lab = 1.25,
-      xlab = paste( "PC1, ", round( cumulative_percent_variance[ 1 ], 2 ), "%" ), 
-      ylab = paste( "PC2, ", round( cumulative_percent_variance[ 2 ], 2 ), "%" ) )
+color <- c("steelblue1", "tomato1", "goldenrod1")
+par(mfrow = c(1, 1))
+plot(
+  scores[, 1],
+  scores[, 2],
+  cex = .5,
+  cex.axis = 1,
+  cex.lab = 1.25,
+  xlab = paste("PC1, ", round(cumulative_percent_variance[1], 2), "%"),
+  ylab = paste("PC2, ", round(cumulative_percent_variance[2], 2), "%")
+  )
 
 # Add visual groupings to pcoa plot
-ordihull( scores, as.factor(targets$treatment ), 
-          border = NULL, lty = 2, lwd = .5, label = F,
-          col = color, draw = "polygon", alpha = 100, cex = .5 )
-
-ordispider( scores,as.factor( targets$grouping ),label = F ) # Vectors connecting samples in same pCO2 x time group
-
-ordilabel( scores, cex = 0.5) # Label sample IDs
-
-```
-
-|           In the PCOA above, blue = low *p*$CO_{2}$, yellow = moderate *p*$CO_{2}$, and red = high *p*$CO_{2}$.
+ordihull(
+  scores,
+  as.factor(targets$treatment),
+  border = NULL,
+  lty = 2,
+  lwd = .5,
+  label = F,
+  col = color,
+  draw = "polygon",
+  alpha = 100,
+  cex = .5
+  )
+
+ordispider(scores, as.factor(targets$grouping), label = F) # Vectors connecting samples in same pCO2 x time group
+
+ordilabel(scores, cex = 0.5) # Label sample IDs
+
+```
+
+In the PCOA above, blue = low *p*$CO_{2}$, yellow = moderate *p*$CO_{2}$, and red = high *p*$CO_{2}$.
 From this plot, we can see that treatment and time both influence multivariate gene expression across the RNA-seq samples.
@@
 
-```{r, message = FALSE, warning = FALSE}
-
-logCPM.pca <- prcomp( t ( df_log ) )
-logCPM.pca.proportionvariances <- ( ( logCPM.pca$sdev ^ 2 ) / ( sum( logCPM.pca$sdev ^ 2 ) ) ) * 100
+<br>
+
+```{r, message = FALSE, warning = FALSE}
+
+logCPM.pca <- prcomp(t (df_log))
+logCPM.pca.proportionvariances <-
+((logCPM.pca$sdev ^ 2) / (sum(logCPM.pca$sdev ^ 2))) * 100
 
@@
 # Replot using logCPM.pca
-plot( logCPM.pca$x, type = "n", main = NA, xlab = paste( "PC1, ", round( logCPM.pca.proportionvariances[ 1 ], 2 ), "%" ), ylab = paste( "PC2, ", round( logCPM.pca.proportionvariances[ 2 ], 2 ), "%" ) )
-
-points( logCPM.pca$x, col = "black", pch = 16, cex = 1 )
-colors2 <- c( "steelblue1", "dodgerblue2", "tomato1", "coral", "goldenrod1", "goldenrod3" )
-
-ordihull( logCPM.pca$x, targets$grouping, 
-          border = NULL, lty = 2, lwd = .5, 
-          col = colors2, draw = "polygon", 
-          alpha = 75,cex = .5, label = T )
-
-```
-
-|           This plot is only a slight adjustment of the first PCOA.
+plot(
+  logCPM.pca$x,
+  type = "n",
+  main = NA,
+  xlab = paste("PC1, ", round(logCPM.pca.proportionvariances[1], 2), "%"),
+  ylab = paste("PC2, ", round(logCPM.pca.proportionvariances[2], 2), "%")
+  )
+
+points(logCPM.pca$x,
+       col = "black",
+       pch = 16,
+       cex = 1)
+       colors2 <-
+       c("steelblue1",
+         "dodgerblue2",
+         "tomato1",
+         "coral",
+         "goldenrod1",
+         "goldenrod3")
+
+       ordihull(
+       logCPM.pca$x,
+       targets$grouping,
+       border = NULL,
+       lty = 2,
+       lwd = .5,
+       col = colors2,
+       draw = "polygon",
+       alpha = 75,
+       cex = .5,
+       label = T
+       )
+
+```
+
+This plot is only a slight adjustment of the first PCOA.
 If you'd like to wrap each multifactorial group and get a better idea of potential overlap between clusters, the plot and code above can be applied to do so.
 
-|           Now that we have conducted QC analyses and plots of the read count data, it is time to begin model fitting and testing for differential expression.
+Now that we have conducted QC analyses and plots of the read count data, it is time to begin model fitting and testing for differential expression.
+
+<br>
 
@@
 
-|           Gene expression traits can exhibit non-linear performance curves across a continuous variable much like other physiological traits.
+Gene expression traits can exhibit non-linear performance curves across a continuous variable much like other physiological traits.
 Studies that incorporate more than two levels of a continuous predictor provide an opportunity to test for non-linear effects on gene expression.
@@
 
+<br>
+
 ```{r, echo = FALSE, out.height = "40%", fig.align = "center", }
+
 knitr::include_graphics("images/Rivera_etal_fig.png")
+
 ```
@@
 
-|           Non-linear reaction norms can be modelled across time, such as Rivera *et al.* have presented, or across multiple levels of an abiotic variable such as temperature, pH, etc.
+Non-linear reaction norms can be modelled across time, such as Rivera *et al.* have presented, or across multiple levels of an abiotic variable such as temperature, pH, etc.
 in order to fit a performance curve.
@@
 Below we outline how to fit and test for non-linear gene expression performance curves in multifactorial RNA-seq experiments using examples in edgeR, DESeq2, and custom code.
-|           One of the simplest non-linear relationships that can be fitted to the expression of a transcript across an continuous variable is a second-order polynomial, otherwise known as a quadratic function, which can be expressed as $y_{i} = \mu + \beta_{1}x^2 + \beta_{2}x$ where $y$ is the abundance of a given transcript ($i$), $\mu$ is the intercept, and $y$ is the continuous variable.
+
+One of the simplest non-linear relationships that can be fitted to the expression of a transcript across an continuous variable is a second-order polynomial, otherwise known as a quadratic function, which can be expressed as $y_{i} = \mu + \beta_{1}x^2 + \beta_{2}x$ where $y$ is the abundance of a given transcript ($i$), $\mu$ is the intercept, and $y$ is the continuous variable.
 For the parabola generated by fitting a second-order polynomial, $\beta_{1}$ > 0 opens the parabola upwards while $\beta_{1}$ < 0 opens the parabola downwards.
 The vertex of the parabola is controlled by $\beta_{2}$ such that when $\beta_{1}$ is negative, greater $\beta_{2}$ values result in the vertex falling at higher values of $x$.
-|           Quadratic polynomials applied to phenotypic performance curves commonly possess negative $\beta_{1}$ values with positive $\beta_{2}$ values: a downard-opening parabola with a positive vertex.
+
+Quadratic polynomials applied to phenotypic performance curves commonly possess negative $\beta_{1}$ values with positive $\beta_{2}$ values: a downard-opening parabola with a positive vertex.
 However, quadtratic curves fitted to gene expression data can take on a variety of postiive or negative forms similar to exponential curves, saturating curves, and parabolas.
@@
 
+<br>
+
 ## Non-linear effects: example in edgeR
 
-|           Let's fit a second-order polynomial for the effect of *p*CO$_{2}$ using edgeR.
+Let's fit a second-order polynomial for the effect of *p*CO$_{2}$ using edgeR.
 Using differential expression tests, we will then determine whether *p*CO$_{2}$ affected a gene's rate of change in expression and expression maximum by applying differential expression tests to $\beta_{1}$ and $\beta_{2}$ parameters.
@@
 
-|          Since this is the first time we're fitting models to our data in this walkthrough, we will also output some important plots to help us explore and QC model predictions as we work toward testing for differential expression attributed to non-linear effects.
+Since this is the first time we're fitting models to our data in this walkthrough, we will also output some important plots to help us explore and QC model predictions as we work toward testing for differential expression attributed to non-linear effects.
+
+<br>
 
@@
 # Square pCO2 variable
-pCO2_2 <- pCO2^2
+pCO2_2 <- pCO2 ^ 2
 
 # Estimate dispersion coefficients
-y1 <- estimateDisp( y, robust = TRUE ) # Estimate mean dispersal
+y1 <- estimateDisp(y, robust = TRUE) # Estimate mean dispersal
 
 # Plot tagwise dispersal and impose w/ mean dispersal and trendline
-plotBCV( y1 )
-
-```
-
-|           The above figure is an output of the plotBCV() function in edgeR, which visualises the biological coefficient of variation (otherwise known as dispersion) parameter fitted to individual genes (black circles), fitted across gene expression level (blue trendline), and averaged across the entire transcriptome (red horizontal line).
+plotBCV(y1)
+
+```
+
+The above figure is an output of the plotBCV() function in edgeR, which visualises the biological coefficient of variation (otherwise known as dispersion) parameter fitted to individual genes (black circles), fitted across gene expression level (blue trendline), and averaged across the entire transcriptome (red horizontal line).
 The BCV/dispersion parameter is a necessary parameter, symbolized as $\theta$ in model specification, that must be estimated in negative binomial GLMs.
@@
 
-|           Now that we've estimated dispersion, we will use these estimates to fit negative binomial GLMs to our read count data using the glmQLFit() function in edgeR, one of the more robust model fitting functions offered by edgeR.
+Now that we've estimated dispersion, we will use these estimates to fit negative binomial GLMs to our read count data using the glmQLFit() function in edgeR, one of the more robust model fitting functions offered by edgeR.
+
+<br>
 
@@
 # Fit multifactorial design matrix
-design_nl <- model.matrix( ~1 + pCO2_2 + pCO2 + pCO2_2:day + pCO2:day ) # Generate multivariate edgeR glm
+design_nl <-
+    model.matrix(~ 1 + pCO2_2 + pCO2 + pCO2_2:day + pCO2:day) # Generate multivariate edgeR glm
 
 # Fit quasi-likelihood, neg binom linear regression
-nl_fit <- glmQLFit( y1, design_nl ) # Fit multivariate model to counts
+nl_fit <-
+    glmQLFit(y1, design_nl) # Fit multivariate model to counts
 
 ## Test for effect of pCO2 and pCO2^2
-nl_pCO2_2 <- glmQLFTest( nl_fit, coef = 2, contrast = NULL, poisson.bound = FALSE ) # Estimate significant DEGs
-nl_pCO2 <- glmQLFTest( nl_fit, coef = 3, contrast = NULL, poisson.bound = FALSE ) # Estimate significant DEGs
+nl_pCO2_2 <-
+    glmQLFTest(nl_fit,
+    coef = 2,
+    contrast = NULL,
+    poisson.bound = FALSE) # Estimate significant DEGs
+nl_pCO2 <-
+    glmQLFTest(nl_fit,
+    coef = 3,
+    contrast = NULL,
+    poisson.bound = FALSE) # Estimate significant DEGs
 
 # Make contrasts
-is.de_nl_pCO2 <- decideTestsDGE( nl_pCO2, adjust.method = "fdr", p.value = 0.05 ) 
-is.de_nl_pCO2_2 <- decideTestsDGE( nl_pCO2_2, adjust.method = "fdr", p.value = 0.05 )
+is.de_nl_pCO2 <-
+  decideTestsDGE(nl_pCO2, adjust.method = "fdr", p.value = 0.05)
+is.de_nl_pCO2_2 <-
+  decideTestsDGE(nl_pCO2_2, adjust.method = "fdr", p.value = 0.05)
 
 # Summarize differential expression attributed to pCO2 and pCO2^2
-summary( is.de_nl_pCO2 )
-summary( is.de_nl_pCO2_2 )
-
-```
-
-|           We have just fit our first GLM to our read count data and have tested for differential expression across *p*CO$_{2}$.
+summary(is.de_nl_pCO2)
+summary(is.de_nl_pCO2_2)
+
+```
+
+We have just fit our first GLM to our read count data and have tested for differential expression across *p*CO$_{2}$.
 At this stage, it is important to output a few diagnostic plots.
@@
 
-|           Visualizations of differential expression (logFC) across baseline logCPM can be produced by the plotMD() (short for "mean-difference") function in edgeR.
+Visualizations of differential expression (logFC) across baseline logCPM can be produced by the plotMD() (short for "mean-difference") function in edgeR.
 To do so, plotMD() is input with the results of glmQLFTest, which we used above to test for differential expression attributable to the parameters *p*CO$_{2}$ and *p*CO$_{2}$^2.
@@
 
+<br>
+
 ```{r, message = FALSE, warning = FALSE}
 # Plot differential expression due to pCO2 and pCO2^2
-plotMD( nl_pCO2 )
-plotMD( nl_pCO2_2 )
-
-```
-
-|           In the two mean-difference plots above, large colored points represent genes exhibiting significant differential expression attributed to *p*CO$_{2}$ or *p*CO$_{2}$^2 below.
+plotMD(nl_pCO2)
+plotMD(nl_pCO2_2)
+
+```
+
+In the two mean-difference plots above, large colored points represent genes exhibiting significant differential expression attributed to *p*CO$_{2}$ or *p*CO$_{2}$^2 below.
 As you can see, there is a lot of differential expression in this experiment resulting from both of these parameters.
@@
 
-|           Next, we will use glmQLFTest() to test for differential expression attributed to interactions between time and *p*CO$_{2}$ or time and *p*CO$_{2}$^2 before producing mean-difference plots visualizing DEGs derived from these two interactions:
+Next, we will use glmQLFTest() to test for differential expression attributed to interactions between time and *p*CO$_{2}$ or time and *p*CO$_{2}$^2 before producing mean-difference plots visualizing DEGs derived from these two interactions:
+
+<br>
 
@@
 ## Test for interactions between time and pCO2 or pCO2^2
-nl_pCO2_int <- glmQLFTest( nl_fit, coef = 4, contrast = NULL, poisson.bound = FALSE ) # Estimate significant DEGs
-nl_pCO2_2_int <- glmQLFTest( nl_fit, coef = 5, contrast = NULL, poisson.bound = FALSE ) # Estimate significant DEGs
+nl_pCO2_int <-
+  glmQLFTest(nl_fit,
+  coef = 4,
+  contrast = NULL,
+  poisson.bound = FALSE) # Estimate significant DEGs
+nl_pCO2_2_int <-
+  glmQLFTest(nl_fit,
+  coef = 5,
+  contrast = NULL,
+  poisson.bound = FALSE) # Estimate significant DEGs
 
 # Make contrasts
-is.de_nl_pCO2_int <- decideTestsDGE( nl_pCO2_int, adjust.method = "fdr", p.value = 0.05 ) # Make contrasts
-is.de_nl_pCO2_2_int <- decideTestsDGE( nl_pCO2_2_int, adjust.method = "fdr", p.value = 0.05 )
+is.de_nl_pCO2_int <-
+  decideTestsDGE(nl_pCO2_int, adjust.method = "fdr", p.value = 0.05) # Make contrasts
+is.de_nl_pCO2_2_int <-
+  decideTestsDGE(nl_pCO2_2_int, adjust.method = "fdr", p.value = 0.05)
 
 # Summarize differential expression attributed to pCO2 and pCO2^2
-summary( is.de_nl_pCO2_int )
-summary( is.de_nl_pCO2_2_int )
+summary(is.de_nl_pCO2_int)
+summary(is.de_nl_pCO2_2_int)
 
 # Plot differential expression due to pCO2 and pCO2^2
-plotMD( nl_pCO2_int )
-plotMD( nl_pCO2_2_int )
-
-```
-
-|           In the mean-difference plots above, it is clear that there is remarkably less differential expression associated with interactions between time and *p*CO$_{2}$ or *p*CO$_{2}$^2 compared to the direct effects of *p*CO$_{2}$ and *p*CO$_{2}$^2.
+plotMD(nl_pCO2_int)
+plotMD(nl_pCO2_2_int)
+
+```
+
+In the mean-difference plots above, it is clear that there is remarkably less differential expression associated with interactions between time and *p*CO$_{2}$ or *p*CO$_{2}$^2 compared to the direct effects of *p*CO$_{2}$ and *p*CO$_{2}$^2.
 This is expected, as the statistical power necessary for detecting such interactions is lesser and there are grounds to expect that interactions would be less frequent during our experiment.
 
-|           Taking a detour from tests for DEGs attributed to non-linear interactions, we can now output the residuals of our GLMs in order to better test for whether our data fit the assumptions of negative binomial distribution families.
+Taking a detour from tests for DEGs attributed to non-linear interactions, we can now output the residuals of our GLMs in order to better test for whether our data fit the assumptions of negative binomial distribution families.
 If the distribution of residuals from our GLMs are normal, this indicates that our data meet the assumption of the negative binomial distribution.
@@
 ggplot(data = melt(as.data.frame(resid.pearson)), aes(x = value)) +
-  geom_histogram( fill = "grey" ) +
-  xlim(-2.5,5.0) +
+  geom_histogram(fill = "grey") +
+  xlim(-2.5, 5.0) +
   theme_classic() +
@@
 
-|           Our residuals appear to be normally distributed, indicating that our data fit the negative binomial distribution assumed by the GLM.
-
-|           Now let's get back to analyzing non-linear effects on gene expression!
+Our residuals appear to be normally distributed, indicating that our data fit the negative binomial distribution assumed by the GLM.
+
+Now let's get back to analyzing non-linear effects on gene expression!
+
+<br>
 
@@
 
-|           A variety of non-linear patterns may underly the significant effects we've observed above.
+A variety of non-linear patterns may underly the significant effects we've observed above.
 For example, interactions between the vertex of expression and time may be attributed to variation in expression peaks across convex curves or the rate of decline in an exponentially decreasing curve among other patterns.
@@
 # Export diff expression data for transcripts with significant DE associated with PCO2^2 parameter
-nl_pCO2_2_sig <- topTags(nl_pCO2_2, n = (11091 + 11915), adjust.method = "BH", p.value = 0.05)
+nl_pCO2_2_sig <-
+  topTags(
+  nl_pCO2_2,
+  n = (11091 + 11915),
+  adjust.method = "BH",
+  p.value = 0.05
+  )
 nl_pCO2_2_sig_geneids <- row.names(nl_pCO2_2_sig) #Output a list of geneids associated with sig PCO2^2 effect
 
-nl_pCO2_sig <- topTags(nl_pCO2, n = (14166 + 13376), adjust.method = "BH", p.value = 0.05)
+nl_pCO2_sig <-
+  topTags(
+  nl_pCO2,
+  n = (14166 + 13376),
+  adjust.method = "BH",
+  p.value = 0.05
+  )
 nl_pCO2_sig_geneids <- row.names(nl_pCO2_sig) #Output a list of geneids associated with sig PCO2 effect
@@
 # Correct pCO2s to exact values
-tab_exp_df$pCO2 <- as.numeric(
-  ifelse(tab_exp_df$pCO2 == "300", 255,
+tab_exp_df$pCO2 <- as.numeric(ifelse(
+  tab_exp_df$pCO2 == "300",
+  255,
   ifelse(tab_exp_df$pCO2 == "900", 930,
-                518)))
+  518)
+  ))
 
 # Correct time to exact values
-tab_exp_df$time <- as.numeric(
-  ifelse(tab_exp_df$time == "1", 0.5, 7))
+tab_exp_df$time <- as.numeric(ifelse(tab_exp_df$time == "1", 0.5, 7))
 
 # Create binary variable in df_all_log for significant non-linear expression
-tab_exp_df$pCO2_2_sig <- ifelse(tab_exp_df$geneid %in% nl_pCO2_2_sig_geneids, "Yes", "No")
-tab_exp_df$pCO2_sig <- ifelse(tab_exp_df$geneid %in% nl_pCO2_sig_geneids, "Yes", "No")
+tab_exp_df$pCO2_2_sig <-
+  ifelse(tab_exp_df$geneid %in% nl_pCO2_2_sig_geneids, "Yes", "No")
+tab_exp_df$pCO2_sig <-
+  ifelse(tab_exp_df$geneid %in% nl_pCO2_sig_geneids, "Yes", "No")
 
@@
 up_genes <- filter(nl_pCO2_sig$table, logFC > 0)
-tab_exp_df$logFC_dir <- ifelse(tab_exp_df$geneid %in% row.names(up_genes), "Up", "Down")
+tab_exp_df$logFC_dir <-
+  ifelse(tab_exp_df$geneid %in% row.names(up_genes), "Up", "Down")
 
@@
 # Estimate average logCPM per gene per timepoint
-tab_exp_avg <- summarySE(measurevar = "value",
-                         groupvars = c("pCO2", "time", "geneid", "pCO2_sig", "pCO2_2_sig", "logFC_dir"),
-                         data = tab_exp_df)
+tab_exp_avg <- summarySE(
+  measurevar = "value",
+  groupvars = c("pCO2",     "time",       "geneid", 
+                "pCO2_sig", "pCO2_2_sig", "logFC_dir"),
+  data = tab_exp_df
+  )
 
@@
        aes(x = pCO2, y = value)) +
-  geom_path(alpha = 0.01, size = 0.25, stat = "identity", aes(group = as.factor(geneid))) +
-    facet_grid(logFC_dir~time) +
+  geom_path(
+    alpha = 0.01,
+    size = 0.25,
+    stat = "identity",
+    aes(group = as.factor(geneid))) +
+  facet_grid(logFC_dir ~ time) +
   theme_classic() +
@@
 
-|           Our plot of gene expression across tens of thousands of transcripts exhibiting significant non-linear changes across *p*CO$_{2}$ appears to include many trends that appear...
+Our plot of gene expression across tens of thousands of transcripts exhibiting significant non-linear changes across *p*CO$_{2}$ appears to include many trends that appear...
 well, linear.
@@
 Instead of asking "for what genes may there be significant, non-linear effects of *p*CO$_{2}$?", we should ask "for what genes *should* we test for significant, non-linear effects?".
-|           One of the best ways to determine whether a non-linear model is appropriate for a transcript is to determine whether it is more probable that its expression is linear or non-linear relative to a continuous predictor.
+
+One of the best ways to determine whether a non-linear model is appropriate for a transcript is to determine whether it is more probable that its expression is linear or non-linear relative to a continuous predictor.
 We can calculate this relative probability using a likelihood ratio test (LRT).
@@
 tab_exp_df$pCO2_2 <- tab_exp_df$pCO2^2
-
-# Filter tabularized logCPM df for read depth of logCPM >= 0.5 in > 75% of samples
 
@@
                       data.frame(geneid = names(nlms),
-                                 column_label = as.character(seq(length(nlms)))),
+                      column_label = as.character(seq(length(
+                      nlms
+                      )))),
                       by = "column_label")
@@
        aes(x = pCO2, y = value)) +
-  geom_path(alpha = .25, size = 0.25, stat = "identity", aes(group = as.factor(geneid))) +
-    facet_grid(pCO2_2_binom~time) +
+  geom_path(
+    alpha = .25,
+    size = 0.25,
+    stat = "identity",
+    aes(group = as.factor(geneid))
+    ) +
+  facet_grid(pCO2_2_binom ~ time) +
   geom_smooth(method = "loess", se = TRUE, span = 1) +
@@
 
-|           The above plot, which visualizes the expression of 162 genes identified in LRTs to exhibit significant and high likelihoods of non-linear variation in logCPM across *p*CO$_{2}$, appears distinct from our first plot of non-linear expression deemed significant by edgeR.
+The above plot, which visualizes the expression of 162 genes identified in LRTs to exhibit significant and high likelihoods of non-linear variation in logCPM across *p*CO$_{2}$, appears distinct from our first plot of non-linear expression deemed significant by edgeR.
 The overall trends appear to be more convex or concave, consistent with a second-order polynomial regressions.
 This speaks to the strength of LRTs in identifying probable, non-linear effects or other multifactorial effects.
-|           Let's see how many non-linear transcripts identified in LRTs exhibited significant non-linear effects within our edgeR models.
+
+Let's see how many non-linear transcripts identified in LRTs exhibited significant non-linear effects within our edgeR models.
 To do this, we'll filter down the significant DEGs output by edgeR to include gene id's within our significant LRT output.
 Then, we'll plot the expression of this filtered set across *p*CO$_{2}$.
+
+<br>
 
@@
        aes(x = pCO2, y = value)) +
-  geom_path(alpha = .25, size = 0.25, stat = "identity", aes(group = as.factor(geneid))) +
+  geom_path(
+    alpha = .25,
+    size = 0.25,
+    stat = "identity",
+    aes(group = as.factor(geneid))
+    ) +
   facet_grid(pCO2_2_binom~time) +
@@
 
-|         89 genes exhibited a significant effect of *p*CO$_{2}$ or *p*CO$_{2}$^2 and a high likelihood of a non-linear as identified by an LRT.
+89 genes exhibited a significant effect of *p*CO$_{2}$ or *p*CO$_{2}$^2 and a high likelihood of a non-linear as identified by an LRT.
 This number is substantially lower than the tens of thousands of genes identified by edgeR alone and the 162 identified via LRT.
@@
 Because these two tests provide distinct and valuable information, the most robust estimate of we have laid out for non-linear effects of *p*CO$_{2}$ on gene expression comes from filtering for significant DEGs in edgeR *and* significant LRTs.
-|         What combination of approaches are most informative ultimately depends on your experimental design and the question you are addressing.
+
+What combination of approaches are most informative ultimately depends on your experimental design and the question you are addressing.
 For example, we can explore an additional question that our previous graphs did not touch on: how does gene expression vary between exposure times for genes exhibiting significant interactions between *p*CO$_{2}$^2 and time?
 Let's produce an exploratory plot of non-linear expression across 0.5 and 7 days of exposure for such genes identified using edgeR alone.
+
+<br>
 
@@
 ggplot(data = edgeR_interaction_avg, 
-       aes(x = pCO2, y = value, color = as.factor(time), group = as.factor(time))) +
+       aes(
+         x = pCO2,
+         y = value,
+         color = as.factor(time),
+         group = as.factor(time)
+         )) +
          geom_path(stat = "identity") +
@@
 
-|         After visualizing variation in expression across *p*CO$_{2}$ for genes identified in edgeR to be affected by interactions between *p*CO$_{2}$^2 and exposure time, we can see interesting changes in gene expression curves across time.
+After visualizing variation in expression across *p*CO$_{2}$ for genes identified in edgeR to be affected by interactions between *p*CO$_{2}$^2 and exposure time, we can see interesting changes in gene expression curves across time.
 For example, genes that exhibited a convex expression curve at 0.5 days of exposure generally modified their curve such that their expression became positively exponential relative to *p*CO$_{2}$.
@@
 
+<br>
+
 # Interactive effects
 
-|         Interactive effects shaping gene expression are common in nature and are becoming increasingly prevalent in models of gene expression derived from experimental studies.
+Interactive effects shaping gene expression are common in nature and are becoming increasingly prevalent in models of gene expression derived from experimental studies.
 Below, we go a little bit deeper on interactive effects by outlining methods for fitting them using categorical and continuous variables in models of expression.
@@
 
+<br>
+
 ## Interactive effects: edgeR
 
-|         edgeR and Voom both take the same syntax for interactive effects, which we define below using the model.matrix() function as 'design_multi <- model.matrix( ~1 + pCO2 + pCO2:day )'.
+edgeR and Voom both take the same syntax for interactive effects, which we define below using the model.matrix() function as 'design_multi <- model.matrix( ~1 + pCO2 + pCO2:day )'.
 Then, we will use this multifactorial model design in both edgeR and Voom as specified below:
 
+<br>
+
 ```{r, message = FALSE, warning = FALSE}
@@
 # Fit multifactoria design matrix that includes an interaction term for pCO2 x day
-design_multi <- model.matrix( ~1 + pCO2 + pCO2:day ) #Generate multivariate edgeR glm
+design_multi <-
+  model.matrix(~ 1 + pCO2 + pCO2:day) #Generate multivariate edgeR glm
 
 # Fit quasi-likelihood, neg binom linear regression
-multi_fit <- glmQLFit( y1, design_multi ) # Fit multivariate model to counts
+multi_fit <-
+  glmQLFit(y1, design_multi) # Fit multivariate model to counts
 
 # Test for effect of pCO2
-tr_pCO2 <- glmQLFTest( multi_fit, coef = 2, contrast = NULL, poisson.bound = FALSE ) # Estimate significant DEGs
-
-is.de_tr_pCO2 <- decideTestsDGE( tr_pCO2, adjust.method = "fdr", p.value = 0.05 ) # Make contrasts
-
-summary( is.de_tr_pCO2 )
+tr_pCO2 <-
+  glmQLFTest(
+  multi_fit,
+  coef = 2,
+  contrast = NULL,
+  poisson.bound = FALSE
+  ) # Estimate significant DEGs
+
+is.de_tr_pCO2 <-
+  decideTestsDGE(tr_pCO2, adjust.method = "fdr", p.value = 0.05) # Make contrasts
+
+summary(is.de_tr_pCO2)
 
 # Test for interaction between pCO2 and time
-tr_int <- glmQLFTest( multi_fit, coef = 3, poisson.bound = FALSE ) # Estimate significant DEGs
-
-is.de_int <- decideTestsDGE( tr_int, adjust.method = "fdr", p.value = 0.05 ) # Make contrasts
-
-summary( is.de_int )
+tr_int <-
+  glmQLFTest(multi_fit, coef = 3, poisson.bound = FALSE) # Estimate significant DEGs
+
+is.de_int <-
+  decideTestsDGE(tr_int, adjust.method = "fdr", p.value = 0.05) # Make contrasts
+
+summary(is.de_int)
 
@@
 
-|         Below we will fit the same 'design_multi' model (y ~ 1 + *p*CO$_{2}$ + *p*CO$_{2}$:days) to our read counts using Voom rather than edgeR.
+Below we will fit the same 'design_multi' model (y ~ 1 + *p*CO$_{2}$ + *p*CO$_{2}$:days) to our read counts using Voom rather than edgeR.
 As stated earlier, Voom fits interactions between fixed effects using the same syntax as edgeR, so the 'design_multi' object can be used by both programs.
 
-|         What *is* different between edgeR and Voom is the manner by which Voom accounts for variation between replicates.
+What *is* different between edgeR and Voom is the manner by which Voom accounts for variation between replicates.
 Rather than using gene-wise or averaged estimations of biological coefficients of variation and inputting these values as $\theta$ in a negative binomial GLM, Voom models what is referred to as the mean-variance relationship and incorporates this relationship within a linear model rather than a GLM by "weighting" the accuracy of a given observation based on its level of expression and modelled variance.
@@
 # Perform Voom transformation
-Voom <- voom( y, design_multi, plot = T )
-
-```
-
-|         When deciding between packages for testing DGE, it can be helpful to compare model assumptions regarding variance by plotting graphs like Voom's mean-variance relationship above and the output of edgeR's 'plotBCV()' function, which we provided an example of earlier and will replot below for the sake of making thsi visual comparison.
+Voom <- voom(y, design_multi, plot = T)
+
+```
+
+When deciding between packages for testing DGE, it can be helpful to compare model assumptions regarding variance by plotting graphs like Voom's mean-variance relationship above and the output of edgeR's 'plotBCV()' function, which we provided an example of earlier and will replot below for the sake of making thsi visual comparison.
 Take note that the y-axes of these variance plots in Voom (square root of standard deviation) and edgeR (biological coefficient of variation) are different:
 
+<br>
+
 ```{r, message = FALSE, warning = FALSE, include = FALSE}
@@
 # Replot edgeR BCV
-plotBCV( y1 )
-
-```
-
-|         The mean-variance relationships modelled in edgeR and Voom are generally similar, but it appears that Voom assumes even lesser variance in expression among genes at the highest expression level.
+plotBCV(y1)
+
+```
+
+The mean-variance relationships modelled in edgeR and Voom are generally similar, but it appears that Voom assumes even lesser variance in expression among genes at the highest expression level.
 One result of this may be that greater statistical power is assumed in tests of differential expression by Voom for genes with high expression relative to low expression genes.
@@
 
+<br>
+
 ```{r, message = FALSE, warning = FALSE}
 # Fit using Voom
-lm_Voom_fit <- lmFit( Voom, design_multi )
+lm_Voom_fit <- lmFit(Voom, design_multi)
 
 # Create a contrast across continuous pCO2 variable
-cont_pCO2 <- contrasts.fit( lm_Voom_fit, coef = "pCO2" )
+cont_pCO2 <- contrasts.fit(lm_Voom_fit, coef = "pCO2")
 
 # Create a contrast across interaction etween continuous pCO2 and time variables
-cont_pCO2_day <- contrasts.fit( lm_Voom_fit, coef = "pCO2:day" )
+cont_pCO2_day <- contrasts.fit(lm_Voom_fit, coef = "pCO2:day")
 
 # Perform empirical Bayes smoothing of standard errors
-cont_pCO2 <- eBayes( cont_pCO2 )
-cont_pCO2_day <- eBayes( cont_pCO2_day )
+cont_pCO2 <- eBayes(cont_pCO2)
+cont_pCO2_day <- eBayes(cont_pCO2_day)
 
 # Output test statistics
-pCO2_results <- topTable( cont_pCO2, coef = "pCO2", adjust.method="fdr", n = Inf )
-pCO2_day_results <- topTable( cont_pCO2_day, coef = "pCO2:day", adjust.method="fdr", n = Inf )
+pCO2_results <-
+  topTable(cont_pCO2,
+  coef = "pCO2",
+  adjust.method = "fdr",
+  n = Inf)
+pCO2_day_results <-
+  topTable(cont_pCO2_day,
+  coef = "pCO2:day",
+  adjust.method = "fdr",
+  n = Inf)
 
 # How many DEGs are associated with pCO2
-length( which( pCO2_results$adj.P.Val < 0.05  ) ) # number of DE genes =
+length(which(pCO2_results$adj.P.Val < 0.05))
 
 # How many DEGs are associated with pCO2:day?
-length( which( pCO2_day_results$adj.P.Val < 0.05  ) )
-
-```
-
-|         Voom has identified 1628 genes whose variation is affected by an interaction between *p*CO$_{2}$ and time, compared to 1748 identified by edgeR.
+length(which(pCO2_day_results$adj.P.Val < 0.05))
+
+```
+
+Voom has identified 1628 genes whose variation is affected by an interaction between *p*CO$_{2}$ and time, compared to 1748 identified by edgeR.
 
@@
 
-|         DESeq2 requires a slightly different syntax for specifying interactive effects compared to edgeR and Voom.
+DESeq2 requires a slightly different syntax for specifying interactive effects compared to edgeR and Voom.
+
+<br>
 
@@
 # Let's remove samples with less then 0.5 cpm (this is ~10 counts in the count file) in fewer then 9/12 samples
-keep_g <- rowSums(cpm( gcounts ) > .5) >= 9
+keep_g <- rowSums(cpm(gcounts) > .5) >= 9
 
@@
 
-design <- design( dds_int )
+design <- design(dds_int)
 
@@
 
-|         DESeq2 has identified 4985 genes whose variation is affected by an interaction between *p*CO$_{2}$ and time, compared to 1748 identified by edgeR and 1628 by Voom.
+DESeq2 has identified 4985 genes whose variation is affected by an interaction between *p*CO$_{2}$ and time, compared to 1748 identified by edgeR and 1628 by Voom.
 This is a ~3-fold greater number of DEGs predicted by DESeq2 compared to edgeR and Voom.
 
+<br>
+
 # Comparing test statistics: interactive effects
 
-|         Differential expression packages can make dramatically different predictions for both the fold-change and probability of differential expression.
+Differential expression packages can make dramatically different predictions for both the fold-change and probability of differential expression.
 If you find yourself deciding between different packages or wanting to compare how conservative different approaches are, it helps to run regressions of model predictions by different packages.
 Below we apply the ggpairs() function from GGAlly [@Schloerke2018-rq] to plot a correlation matrix of logFC values and negative log-transformed p-values attributed to interactive effects between time and *p*CO$_{2}$ predicted by edgeR, DESeq2, and Voom:
+
+<br>
 
@@
     data.frame(geneid = row.names(tr_int$table),
-               edgeR_logFC = tr_int$table$logFC, #negate logFC because of syntax differences
+               edgeR_logFC = tr_int$table$logFC,
                edgeR_pval = tr_int$table$PValue), 
@@
     scale_fill_viridis_c() +
-    geom_abline(slope=1, intercept = 0, color = "red", lty = 2, size = 1, alpha = 0.5)
+    geom_abline(
+    slope = 1,
+    intercept = 0,
+    color = "red",
+    lty = 2,
+    size = 1,
+    alpha = 0.5
+    )
   p
@@
 
-|         The above plot is the output of ggpairs() showing correlations between negative log-transformed p-values attributed to interactive effects of time x *p*CO$_{2}$ estimated between each of the three packages, as well as the distributions of this parameter for each package, and correlation statistics.
+The above plot is the output of ggpairs() showing correlations between negative log-transformed p-values attributed to interactive effects of time x *p*CO$_{2}$ estimated between each of the three packages, as well as the distributions of this parameter for each package, and correlation statistics.
 Red dashed lines represent a slope equal to 1.
@@
 
-|         Voom and edgeR share the largest correlation between p-values estimated for an interactive effect of *p*CO$_{2}$ : days on differential expression (R$^2$ = 0.965).
+Voom and edgeR share the largest correlation between p-values estimated for an interactive effect of *p*CO$_{2}$ : days on differential expression (R$^2$ = 0.965).
 The bottom row of this ggpairs plot shows tranformed p-values from DESeq2 on the y-axis, which appear to be skewed toward lower p-values/higher -log p-values.
@@
 
-|         Next, let's create a ggpairs() plot contrasting logFC estimates of DE attributed to *p*CO$_{2}$ : days in edgeR, Voom, and DESeq2.
+Next, let's create a ggpairs() plot contrasting logFC estimates of DE attributed to *p*CO$_{2}$ : days in edgeR, Voom, and DESeq2.
 One might think that logFCs should demonstrate stronger correlations between programs compared to p-values, but due to the manners by which read counts are normalized and transformed in different packages this is not always the case!
+
+<br>
 
@@
 
-|         In this pairs plot of logFC estimates across packages, you can see that R$^2$ values are, on average, lower than those of p-value correlations between packages.
+In this pairs plot of logFC estimates across packages, you can see that R$^2$ values are, on average, lower than those of p-value correlations between packages.
 Once again, the strongest correlation exists between edgeR and Voom.
@@
 
-|         Random intercepts (often referred to as random effects) can be an essential component of estimating differential expression in experiments that incorporate non-independent observations (e.g., repeated measures and nested designs: individuals within populations or families) or randomly sampled groups.
+Random intercepts (often referred to as random effects) can be an essential component of estimating differential expression in experiments that incorporate non-independent observations (e.g., repeated measures and nested designs: individuals within populations or families) or randomly sampled groups.
 Most differential expression packages do not possess functionality for fitting random intercepts, extending GLMs to generalized linear mixed models or GLMMs, but Voom does!
 Additional packages that we don't cover here also possess random intercept functionality, including dream [@Hoffman2021-br] and [glmmSeq](https://cran.r-project.org/web/packages/glmmSeq/index.html).
-|         Below is an example of how to fit a random intercept associated with the time variable of our dataset using Voom:
+
+Below is an example of how to fit a random intercept associated with the time variable of our dataset using Voom:
+
+<br>
 
@@
 # Fit multifactoria design matrix
-design_rand <- model.matrix( ~1 + pCO2 + ( 1 | day ) ) #Generate multivariate edgeR glm
+design_rand <- model.matrix(~ 1 + pCO2 + (1 | day)) #Generate multivariate edgeR glm
 
 # Perform Voom transformation
-Voom_rand <- voom( y, design_rand, plot = T )
-
-```
-
-|         If you compare the mean-variance plot from our mixed model above, with our first mean-variance plot from the Voom model that did not include a random effect you will notice that the modelled relationship between variance and mean expression is considerably different between the two.
+Voom_rand <- voom(y, design_rand, plot = T)
+
+```
+
+If you compare the mean-variance plot from our mixed model above, with our first mean-variance plot from the Voom model that did not include a random effect you will notice that the modelled relationship between variance and mean expression is considerably different between the two.
 While the linear model mean-variance plot took on a negative exponential shape by which earlier increases in expression level associate with a precipitous decline in variance, our mixed model's fit takes on a more logarithmic shape and the raw points appear parabolic.
@@
 # Fit using Voom
-lm_Voom_fit_rand <- lmFit( Voom_rand, design_rand )
+lm_Voom_fit_rand <- lmFit(Voom_rand, design_rand)
 
 # Create a contrast across continuous pCO2 variable
-cont_rand_day <- contrasts.fit( lm_Voom_fit_rand, coef = "pCO2" )
+cont_rand_day <- contrasts.fit(lm_Voom_fit_rand, coef = "pCO2")
 
 # Perform empirical Bayes smoothing of standard errors
-cont_rand_day <- eBayes( cont_rand_day )
+cont_rand_day <- eBayes(cont_rand_day)
 
 # Output test statistics
-rand_results <- topTable( cont_rand_day, coef = "pCO2", adjust.method = "fdr", n = Inf )
+rand_results <- topTable(cont_rand_day,
+                         coef = "pCO2",
+                         adjust.method = "fdr",
+                         n = Inf)
 
 # How many DEG are associated with pCO2 after incorporating a random effect for day?
-length( which( rand_results$adj.P.Val < 0.05  ) ) # number of DE genes
-
-```
-
-|         While the linear Voom model's DE test output 0 predicted DEG's associated with *p*CO${_2}$, the mixed Voom model predicted 3 genes.
+length(which(rand_results$adj.P.Val < 0.05)) # number of DE genes
+
+```
+
+While the linear Voom model's DE test output 0 predicted DEG's associated with *p*CO${_2}$, the mixed Voom model predicted 3 genes.
 As we saw eariler when fitting linear models, the majority of DEG's in this study were derived from an interaction between time and *p*CO${_2}$.
@@
 
+<br>
+
 # Random slopes
 
-|         Distinct from the random intercept, testing for variation in gene expression attributed to a random slope effect asks, "Within a set of groups, does the reaction norm of gene expression across a continuous predictor randomly by group?".
+Distinct from the random intercept, testing for variation in gene expression attributed to a random slope effect asks, "Within a set of groups, does the reaction norm of gene expression across a continuous predictor randomly by group?".
 This question is relevant to experimental designs testing for differences in the plasticity of gene expression across randomly-selected demographics such as genotypes or families.
@@
 
-|         Below we detail custom scripts for fitting gaussian linear mixed models using lme4 and negative binomial GLMs to read count data.
+Below we detail custom scripts for fitting gaussian linear mixed models using lme4 and negative binomial GLMs to read count data.
 For practical purposes, we will treat exposure time as a random variable even though timepoints were not randomly selected in the experiment that produced our data and apply.
@@
 
-|         If you've made it this far, you've probably found it to be a common theme that popular DE packages do not provide all of the flexibility necessary for fitting models to data coming from multifactorial experimental designs, even some of the more standard designs.
+If you've made it this far, you probably found it to be a common theme that popular DE packages do not provide all of the flexibility necessary for fitting models to data coming from multifactorial experimental designs, even some of the more standard designs.
 It should be restated that our walkthrough only scratches the surface of (i) available packages that can facilitate multifactorial DE tests and (ii) opportunities that custom scripts for model fitting provide.
 For example, our example code for fitting linear and linear mixed models before testing for effects using likelihood ratio tests can easily be edited to fit negative binomial GLMs such as those used by edgeR and DESeq2 input with $\theta$ estimates (e.g., dispersion) from edgeR or custom estimates, and we encourage readers to try out this strategy!
+One useful resource for doing so is [@Rocke2015-uo], a preprint describing transformations of RNA-seq data that enable model fitting using the 'MASS' function glm.nb().
 If you have personal code that you believe would make a valuable addition to this walkthrough, don't hesitate to post a GitHub issue or submit a pull-request to us.
@@
 
+<br>
+
 # References

### 2022-07-25 DGE_comparison_v2.Rmd
Push changes to fastq to VCF and multifac RNAseq

@@
 # Plot distributions with histogram
-ggplot(data = rt_df, aes(x = rt, fill = as.factor(t_size), group = as.factor(t_size))) +
+ggplot(data = rt_df, aes(x = rt, fill = as.factor(t_mu), group = as.factor(t_mu))) +
   geom_histogram(data = filter(rt_df, t_mu == 25), alpha = 0.75) +
@@
   theme_classic() +
-  labs(title = "Simulated negative binomial distribution", fill = "Mean", y = "Frequency", x = "Mean")
+  labs(title = "Simulated negative binomial distributions", fill = "Mean", y = "Frequency", x = "Value")
 

### 2022-07-25 fastq2vcf_pipeline.Rmd
Push changes to fastq to VCF and multifac RNAseq

@@
 ---
-title: "FASTQ -> VCF workflow for WGS data"
+title: 'snpArcher: FASTQ -> VCF workflow for WGS data'
 author: "Sara JS Wuitchik"
@@
 # fastq2bam
-This is a more fleshed out explanation of the pipeline developed and maintained by [Harvard Informatics](https://github.com/harvardinformatics) in the [shortRead_mapping_variantCalling repo](https://github.com/harvardinformatics/shortRead_mapping_variantCalling)
 
@@
 
-FastP (and other trimming software) can be used for quality control before aligning your sequences to the reference genome, but there are arguments on both sides for whether or not trimming your reads at this stage is useful (add references in here).
+FastP (and other trimming software) can be used for quality control before aligning your sequences to the reference genome, but there are arguments on both sides for whether or not trimming your reads at this stage is useful.
 
@@
 ```{bash, eval = F}
-bwa mem -M -R {string} {ref} {r1} {r2} | samtools sort -o {output.bam}
+bwa mem -M -R {string} {ref} {r1} {r2} | samtools sort -u - > {output.bam}
 ```
@@
 ```{r1}``` and ```{r2}``` are the first and second reads you are aligning <br>
-```-o``` is the output file name
+```-u``` is for uncompressed output
 
 Once you have a sorted and indexed alignment, you can locate and tag duplicate records, which are reads that have originated from a single fragment of DNA.
-This can happen during sample preparation (e.g., during PCR) or from sequencing artifacts (e.g., optical duplicates).If you have multiple runs for the same sample, you would want to merge and index the BAMs before the deduplication tagging, which you can do with ```samtools merge```.
+This can happen during sample preparation (e.g., during PCR) or from sequencing artifacts (e.g., optical duplicates).If you have multiple runs for the same sample, you would want to merge and index the BAMs before the deduplication tagging, which you can do with ```samtools merge``` (integrated in snpArcher under ```merge_bams``` rule).
 
 ```{bash, eval = F}
-picard MarkDuplicates I={input.bam} O={output.bam} METRICS_FILE={output} REMOVE_DUPLICATES=false TAGGING_POLICY=All
+sabamba markup {input} {output}
 ```
-
-Where <br>
-```I``` is the input BAM that has been coordinate sorted (the output of ```samtools sort```) <br>
-```O``` is the output BAM file <br>
-```METRICS_FILE``` writes the deduplication metrics to the output file name (this is a required argument) <br>
-```REMOVE_DUPLICATES``` writes the apropriate flags to the record rather than removing the reads without telling you what they are (```false``` is the default setting) <br>
-```TAGGING_POLICY``` will set the duplicates to be recorded in the DT attribute when the parameter is set to ```All```
-
-Once you have tagged the duplicated records in your BAM, you will want to generate a .bai index file
-
-```{bash, eval = F}
-picard BuildBamIndex I={input.bam}
-```
-
-Where <br>
-```I``` is the input dedup'd BAM (the output of ```picard MarkDuplicates```)
 
@@
 
-To produce a summary of the alignment metrics, including the quality of read alignments and the quality filters, we use ```Picard```
+And to produce a summary of the alignment metrics:
 ```{bash, eval = F}
-picard ValidateSamFile I={input.bam} R={reference.fa} O={output}
+samtools flagstat -O tsv {input.bam} > {output}
 ```
@@
 Where <br>
-```I``` is the input BAM file <br> 
-```R``` is the reference genome <br>
-```O``` is the file name you want to write the output to <br><br>
-
-To ensure the alignments meet file format specifications:
-```{bash, eval = F}
-picard ValidateSamFile I={input.bam} R={reference.fa} O={output} IGNORE=VALID_TAG_NM || true
-```
+```O``` sets the output format (in this case, as a TSV) <br> 
+```input.bam``` is the input BAM file <br>
+```output``` is the file name you want to write the output to <br><br>
 
@@
 Regardless of the software used, variant calling is computationally intensive but can be parallelized for more efficient resource use by splitting the reference genome into intervals.
-The interval creation is automated in our pipeline but parameter guidelines will differ depending on your reference genome.
+The interval creation is automated in our pipeline using ```Picard``` but parameter guidelines will differ depending on your reference genome.
 <br><br>
@@
 ```{bash, eval = F}
-gatk HaplotypeCaller -R {reference.fa} -I {input.bam} -O {output.gvcf} -L {interval.list} --emit-ref-confidence GVCF --min-pruning 2 --min-dangling-branch-length 4
+gatk HaplotypeCaller -R {reference.fa} -I {input.bam} -O {output.gvcf} -L {interval.list} --emit-ref-confidence GVCF --min-pruning {params} --min-dangling-branch-length {params}
 ```
@@
 ```-O``` is the output gVCF <br>
-```-L``` is the interval file <br>
+```-L``` is the interval file list <br>
 ```--emit-ref-confidence``` yields the reference confidence scores as gVCF <br>
-```--min-pruning``` sets the minimum support to not prune paths in the graph <br>
-```--min-dangling-branch-length``` sets the minimum lenght of a dangling branch to attempt recovery <br><br>
-
-A note about coverage: when calling variants where the coverage is less than 10x, we suggest setting ```--min-pruning``` and ```--min-dangling-branch-length``` each to 1.
-The parameters in the code above are more appropriate for data where the coverage exceeds 10x.
-<br>
+```--min-pruning``` sets the minimum support to not prune paths in the graph (low coverage option: 1 / high coverage option: 2) <br>
+```--min-dangling-branch-length``` sets the minimum length of a dangling branch to attempt recovery (low coverage option: 1 / high coverage option: 4) <br><br>
 
@@
 export TILEDB_DISABLE_FILE_LOCKING=1
-gatk GenomicsDBImport --genomicsdb-workspace-path {output.db} -L --sample-name-map {input.db}
+gatk GenomicsDBImport --genomicsdb-shared-posixfs-optimizations true --batch-size 25 --genomicsdb-workspace-path {output.db} -L {input.l} --sample-name-map {input.db}
 ```
@@
 the ```export``` command can remedy sluggish performance <br> 
+```--genomicsdb-shared-posixfs-optimizations``` allows for optimization when using shared Posix Filesystems (like Lustre) <br>
+```--batch-size``` is used in conjunction with the above argument to control the memory consumption <br>
 ```--genomicsdb-workspace-path``` is the working directory to point to for GenomicsDB <br>
-```-L``` is the interval file <br>
+```-L``` is the interval file list <br>
 ```--sample-name-map``` is the path to a file that contains a sample to file map in a tab delimited format <br><br>
@@
 ```{bash, eval = F}
-gatk GenotypeGVCFs -R {reference.fa} -V {input.DB} -O {output.vcf}
+gatk GenotypeGVCFs -R {reference.fa} --heterozygosity {params} --genomicsdb-shared-posixfs-optimizations true -V {input.DB} -O {output.vcf}
 ```
@@
 ```-R``` is the reference genome <br>
+```--heterozygosity``` value used to compute prior likelihoods for any locus - we use 0.005 as a default, but this can be changed in the config <br>
 ```-V``` is the DB map file <br>
@@
 
-If you want to use freebayes the variant caller, rather than starting with ```gatk HaplotypeCaller```, you would instead use:
-```{bash, eval = F}
-freebayes -f {reference.fa} -r {interval} -g {maxDP} {inputs.bam} > {output.vcf}
-```
-
-Where <br>
-```-f``` is the reference genome (creates a .fai index if one doesn't already exist) <br>
-```-r``` limits the analysis to a specific interval (region) <br>
-```-g``` skips processing of alignments overlapping positions with coverage greater than the value; we use the maxDP for this parameter <br> 
-```{inputs.bam}``` are the input BAMs; you can input all BAMs at once but you need to set the output as one VCF per interval <br>
-```{output.vcf}``` the output VCF for each interval e.g., {i}.vcf per interval <br><br>
-
-Regardless of the variant caller you use, before combining all the VCFs for the intervals into one final VCF, it is computationally more efficient to filter each of the VCFs, then gather them. 
+Then, before combining all the VCFs for the intervals into one final VCF, it is computationally more efficient to filter each of the VCFs, then gather them together. 
 ```{bash, eval = F}
@@
 #### Quality control and statistics <br>
-NB: this section will likely be better suited to be included in a QC section (and it's not fully fleshed out itself yet), but here are two of the pipeline outputs that can be used for downstream QC.
+There are a number of statistics generated by snpArcher at various points of the workflow, as well as a QC workflow that is implemented after the final VCF is produced.
+This QC workflow produces an interactive dashboard for the user, but we have separated some of the QC out by the type/purpose, rather than separating each rule.
+<br><br>
+
+This ```bcftools query``` pulls the designated fields in a tab-separated format that can be easily read into R (or other software) for interrogation.  
+```{bash, eval =F}
+bcftools query -f '%CHROM\t%POS\t%ID\t%INFO/AF\t%QUAL\t%INFO/ReadPosRankSum\t%INFO/FS\t%INFO/SOR\t%INFO/MQ\t%INFO/MQRankSum\n' {input.vcf} > {output}
+```
 <br>
 
+This set of commands will output statistics based on the VCF
 ```{bash, eval = F}
-vcftools --gzvcf {input.vcf} --remove-filtered-all --minDP 1 --missing-indv --stdout > {output}
+vcftools --gzvcf {input.vcf} --FILTER-summary --out {output.prefix}
+vcftools --gzvcf {input.vcf} --depth --out {output.prefix}
+vcftools --gzvcf {input.vcf} --missing-indv --out {output.prefix}
 ```
@@
 ```--gzvcf``` is the input VCF, gzipped (use ```--vcf``` if working with an uncompressed VCF) <br>
-```--remove-filtered-all``` removes all the sites with a FILTER flag other than PASS <br>
-```--minDP``` includes only the genotypes greater than or equal to the parameter value (this requires the DP FORMAT tag to be specified) <br>
+```--FILTER-summary```
+```--depth```
 ```--missing-indv``` generates a file reporting the missingness per indivdual <br>
-```--stdout``` redirects the output to stdout so it can be redirected to a file name of your choosing (in this code, the file name would be where ```{output}``` is)
+```--out`` assigns a prefix to each output e.g., the filter summary output would be titled "prefix.FILTER.summary"
+
 <br>
 <br>
+Create a PCA from PLINK, using biallelic variants only
 ```{bash, eval = F}
-bedtools intersect -a {input.intervals} -b {input.vcf} -c > {output}
+# First, remove filtered sites and keep only the biallelic SNPs
+bcftools view -v snps -m2 -M2 -f .,PASS -e 'AF==1 | AF==0 | ALT="*" | TYPE~"indel" | ref="N"' {input.vcf} -O z -o {output.filtered}
+bcftools index {output.filtered}
+
+# Then, if there are less than 150,000 SNPs, just take all of them
+ALLSITES=`bcftools query -f '%CHROM\t%POS\n' {output.filtered} | wc -l`
+SITES=`echo $(( ${{ALLSITES}} / 100000 ))`
+if [[ $SITES -gt 1 ]]
+then
+    bcftools +prune -w $SITES -n 1 -N rand -O z -o {output.pruned} {output.filtered}
+else
+    bcftools view -O z -o {output.pruned} {output.filtered}
+fi
+
+# Use PLINK 2 for the King relatedness matrix ... 
+plink2 --vcf {input.vcf} --pca 2 --out {params.prefix} --allow-extra-chr --autosome-num 95 --make-bed --make-king square --const-fid --bad-freqs
+
+# and PLINK 1.9 for the distance matrix 
+plink --vcf {input.vcf} --out {params.prefix} --allow-extra-chr --autosome-num 95 --distance square --const-fid
 ```
-Where <br>
-```-a``` is the intervals BED produced by the pipeline automatically for use in freebayes <br>
-```-b``` is the final VCF file output by ```gatk GatherVcfs``` <br>
-```-c``` reports the number of hits in ```-b``` for each entry in ```-a``` <br>
-```{output}``` is the file name of the output, which is a text file that reports the number of SNPs per interval
+<br>
+<br>
+
+snpArcher also runs ADMIXTURE at k=2 and k=3: 
+```{bash, eval = F}
+# First, make a BIM without any characters in the chromosome names
+mv {input.bim} {input.bim}.orig
+paste <(cut -f 1 {input.bim}.orig | sed 's/[^0-9]//g') <(cut -f 2,3,4,5,6 {input.bim}.orig) >  {input.bim}
+
+# Then run ADMIXTURE
+admixture {input.bed} 2
+admixture {input.bed} 3
+```

### 2022-08-17 RADseq.Rmd
Fixed references in RADseq, updated fastq2vcf to most recent version from SW

@@
 RADseq comes in different flavors.
-Double-digest RADseq (ddRAD; [@Peterson2012-tr])  selects  markers  with  two  restriction enzymes  with  different  cut  frequencies.
-Genotype-by-Sequencing  (GBS;  [@Elshire2011-or])  uses  a  frequent-cutting  restriction  enzyme  with PCR size selection.
-There are a number of reviews comparing different RADseq and GBS methods (e.g., [@Andrews2016-ay]).
+Double-digest RADseq (ddRAD; @Peterson2012-tr)  selects  markers  with  two  restriction enzymes  with  different  cut  frequencies.
+Genotype-by-Sequencing  (GBS;  @Elshire2011-or)  uses  a  frequent-cutting  restriction  enzyme  with PCR size selection.
+There are a number of reviews comparing different RADseq and GBS methods (e.g., @Andrews2016-ay).
 In this guide we use the term "RADseq" to refer to any of these protocols, including those that don't involve random shearing of data.
@@
 |
-| Unique aspects | Novel data reduction approach used to inform coverage cutoffs | Python API for popgene and phylogenetic analyses written specifically for RAD data |  |
+| Unique aspects | Novel data reduction approach used to inform coverage cutoffs | Python API for popgen and phylogenetic analyses written specifically for RAD data | `populations` module calculates sliding window and site-specific popgen metrics |
 | Documentation | Very good (esp.
 tutorials), active community support on Google Groups | Excellent (esp.
-installation, parameter explanations, and tutorials), active community support on Gitter | Very good, active community support on Google Groups |
-| Speed/Accuracy | fastest and most accurate | close to dDocent accuracy, more over-splitting is possible  if parameters are not tuned | produces some untrue genome fragments (esp.
+installation, parameter explanations, and tutorials), active community support on Gitter | Very good, in depth tutorials as published manuscripts, active community support on Google Groups |
+| Speed/Accuracy (based on discussion in [panel seminar](https://www.youtube.com/watch?v=C74GBESeIq4)) | fastest and most accurate | close to dDocent accuracy, more over-splitting is possible  if parameters are not tuned | produces some untrue genome fragments (esp.
 with higher levels of indel polymorphism), but can be addressed with downstream filtering |
-| Open source/development | Open source | Open source, primarily combines existing software | No |
+| Open source/development | Open source, primarily combines existing software | Open source | No |
 | Filtering options | Minimal default filtering as implemented in VCFtools.
 Ideal for those who want full freedom in filtering their SNPs.
-| Wide range of filtering options  | Moderate range of filtering options |
+| Wide range of filtering options, some hard-coded filtering to deal with paralogs  | Moderate range of filtering options |
 | Output options | Only produces a VCF, fasta of the de novo assembly, and individual BAM files  from mapping to the reference  | Lots of output formats (see [documentation](https://ipyrad.readthedocs.io/en/master/output_formats.html)) | Lots of Stacks-specific output and log files, as well as various inputs for popgen programs (see [documentation](https://catchenlab.life.illinois.edu/stacks/comp/populations.php)) |
@@
 
-Some groups have also developed pipelines for specific flavors of RAD (e.g., [Matz lab for 2b-RAD](https://github.com/z0on/2bRAD_denovo/blob/master/2bRAD_README.sh)) or proprietary software (eg, [DARTseq](http://georges.biomatix.org/storage/app/media/uploaded-files/dartR_Workbook.pdf)), which we touch on below.
+Some groups have also developed pipelines for specific flavors of RAD (e.g., [Matz lab for 2b-RAD](https://github.com/z0on/2bRAD_denovo/blob/master/2bRAD_README.sh)) or proprietary software (eg, [DARTseq](http://georges.biomatix.org/storage/app/media/uploaded-files/dartR_Workbook.pdf)).
 **In most cases, it is recommended that you use a pipeline developed and tested for RAD data, especially if you are making a de novo assembly.** [@LaCava2020-bg] have an excellent study where they review various *de novo* assemblers that are used in these pipelines.
@@
 It should be noted that all of these RAD pipelines have optimized the specific parameters of these de novo assemblers to work with RAD data of various flavors, which may not be fully reflected in the LaCava study.
+
 Based on the MarineOmics RAD panel, the general consensus is that all the popular pipelines can produce adequately accurate datasets (with appropriate parameter optimization and data filtering).     
@@
 
-It looks like samples X,Y,Z have a lot of missing data relative to other samples.
+It looks like samples OR3_13_C3,CA4_12_C1, OR3_3_C7, CA4_13_C1, BC2_6_C2, BC2_9_C4 have a lot of missing data relative to other samples.
 We can remove them from our dataset for now, and just keep exploring our data, but if you decide to exclude them permanently from the analysis you should eventually rerun your genotyping pipeline without those samples, esp.
@@
 2) Another great way to explore missingness with your data is to construct a heatmap of loci that are genoptyped between pairs of samples.
-Generally, samples that are more closely related will share more loci with each other due to cutsite dropout (cite).
+Generally, samples that are more closely related will share more loci with each other due to cutsite dropout.
 Deviations from that pattern can pinpoint bad samples as well as batch effects.
@@
 But as there are many ways to project the shadow of a three-dimensional object on a two dimensional plane, there are many (and even more) ways to project a 593,124-dimensional cloud of points to two dimensions.
-What PCA does is figuring out the “best” way to do this project in order to visualise the major components of variance in the data." See [here](https://michaelaalcorn.medium.com/yet-another-pca-explanation-727dff6ce26a) for a linear algebra-based explanation of PCA).
+What PCA does is figuring out the “best” way to do this project in order to visualise the major components of variance in the data." See [here](https://michaelaalcorn.medium.com/yet-another-pca-explanation-727dff6ce26a) for a linear algebra-based explanation of PCA.
 
@@
 
-Here we use SeqArray and SNPRelate to run a PCA in R.
+Here we use _SeqArray_ and _SNPRelate_ to run a PCA in R.
 
@@
 Be aware of how different analysis tools deal with missingness, especially PCA which will fill in all missing data with some values.
-Here, the PCA from SNPRelate imputes missing genotypes as the dosage mean.
+Here, the PCA from _SNPRelate_ imputes missing genotypes as the mean genotype across samples.
 
@@
 
-* Do any observed clusters match a source of batch effect? [how to remove batch effects](#batch)
+* Do any observed clusters match a source of batch effect? [How to remove batch effects](#batch)
 
@@
 
-Nicolas Lou and Nina Therkildsen have a [great preprint](https://www.authorea.com/users/380682/articles/532568-batch-effects-in-population-genomic-studies-with-low-coverage-whole-genome-sequencing-data-causes-detection-and-mitigation) and [Github repo](https://github.com/therkildsen-lab/batch-effect) discussing causes, detection, and mitigation of batch effects in low coverage WGS, where batch effects can have particularly large effects.
+Nicolas Lou and Nina Therkildsen have a [great article](https://onlinelibrary.wiley.com/doi/abs/10.1111/1755-0998.13559) and [Github repo](https://github.com/therkildsen-lab/batch-effect) discussing causes, detection, and mitigation of batch effects in low coverage WGS, where batch effects can have particularly large effects [@@Lou2021-ew].
 While some of their recommendations are specific to reference-based analyses, others are relevant to RAD methods too (eg, batch effects due to different sequencing platforms, DNA degradation levels, and sequencing depth).
@@
 
-You can also check heterozygosity. Large differences in heterozygosity between outliers can suggest different species or cryptic hybrids.
-
-<!--Note: Two outliers are obvious.
+You can also check heterozygosity.
+Large differences in heterozygosity between outliers can suggest different species or cryptic hybrids.
+Two outliers are obvious in the previous PCA.
 Large differences in heterozygosity between samples are clear, with one location showing two distinct clusters of samples.
 These samples were later determined using mtDNA and morphological analysis to be cryptic species.
-*Unpublished data from a marine species in Australian waters.*-->
+*Unpublished data from a marine species in Australian waters.*
+
+<!--Note: -->
 
 You can then estimate genetic divergence between these clusters using pairwise Fst and/or phylogenetic inference to determine how diverged the groups are.
-If they are very diverged, it may be better to assemble each cryptic species separately, especially for denovo assemblies with dDocent and Stacks, especially since often times restriction cut-sites are not always conserved; see [@Wagner2013-ff,@Rubin2012-hw].
+If they are very diverged, it may be better to assemble each cryptic species separately, especially for denovo assemblies with dDocent and Stacks, especially since often times restriction cut-sites are not always conserved; see [@Wagner2013-ff; @Rubin2012-hw].
 These two pipelines used a set of given samples to create a "reference" or "catalog" to which all other samples are mapped.
@@
 Mapping defaults for popular RADseq pipelines:  
-L: clipping penalty for 5' and 3'
-A: matching score
-O gap open penalty
-T: mapping quality filter
+`L`: clipping penalty for 5' and 3'  
+`A`: matching score  
+`O`: gap open penalty  
+`T`: mapping quality filter
 
-dDocent: bwa mem -L 20,5 -t 8 -a -M -T 10 -A1 -B 3 -O 5 -R
+dDocent: `bwa mem -L 20,5 -t 8 -a -M -T 10 -A1 -B 3 -O 5 -R`
 
-* can change -A and -O during the dDocent pipeline.    
-ipyrad: (defaults) -L 5,5 -B 4 -A1 -O 6 -T 30
+* can change -A and -O interactively during the dDocent pipeline.
+
+ipyrad: (defaults) `-L 5,5 -B 4 -A1 -O 6 -T 30`
 
@@
 
-Our whole genome resequencing section has some good content on mapping(link). General considerations with RAD data:
+Our [whole genome resequencing tutorials](https://marineomics.github.io/WGS_intro.html)] has some good content on mapping. General considerations with RAD data:
 
@@
 If it's too high, then real polymorphic loci will be excluded.
-All of the three major pipelines allow users to implement read depth thresholds
-
-Generating multiple sequencing reads per locus is helpful when calling SNPs in order to  distinguishing between reads with sequencing errors and.
+The *too high* threshold for your dataset can usually best be determined by using a few different read depth thresholds and comparing how the results change downstream analyses.
 
@@
 
-`r colorize("This section is still under development! Please contribute on Github or provide feedback in [Discussions](https://github.com/MarineOmics/marineomics.github.io/discussions)", "red")`
+`r colorize("Tutorials for this section are still under development!
+Please contribute on Github or provide feedback in [Discussions](https://github.com/MarineOmics/marineomics.github.io/discussions)", "red")`
 
@@
 * Don't over filter for missingness
- + papers showing how too much filtering can skew results  
-  + Empirical: Diaz-Arce 2019, Tripp 2017, Silliman 2021, Eaton 2017, Martin-Hernanz (2019)
-  + Simulations: Huang & Knowles 2016, Leaché et al., 2015, Gautier et al., 2012, Rivera-Colón et al., 2021
+    + Papers showing how too much filtering can skew results:
+       + Empirical:  [@Eaton2017-gx; @Tripp2017-cp; @Martin-Hernanz2019-bu; @Silliman2021-oo; @Diaz-Arce2019-nh]  
+       + Simulation investigating effects and sources of missing data: [@Huang2016-na; @Leache2015-eg; @Gautier2013-vj; @Rivera-Colon2021-en]
+* Some ways to account for missing data in downstream analyses:
+   + impute mising data (PCA), subsample (PCA, structure, pretty much anything with SNPs), try a few different missing filters and see what impact it has on results
 
-* some ways to account for missing data in downstream analyses
- + impute (PCA), subsample (PCA, structure, pretty much anything with SNPs), try a few different missing filters and see what impact it has on results
+Jon Puritz has a [great tutorial](http://www.ddocent.com/filtering/) on filtering SNPs from RADseq data assembled with dDocent.
+While it requires certain VCF fields to be present that aren't generated by ipyrad or Stacks, it contains great concepts on how to filter.
 
@@
 ## Repeat as needed {#repeat}
+
 # Helpful tutorials  

### 2022-08-17 WGS_intro.Rmd
Added a ref
Fixed references in RADseq, updated fastq2vcf to most recent version from SW

@@
 
-* [Fastq-to-VCF SnakeMake pipeline](fastq2vcf_pipeline.html): in-depth explanation of an automated short-read mapping and variant calling pipeline maintained by [Harvard Informatics](https://github.com/harvardinformatics/shortRead_mapping_variantCalling).
-Useful if you plan to adopt this pre-packaged automated and parallelizable pipeline, would like to understand its different components but not necessarily change it substantially.
+* [Fastq-to-VCF SnakeMake pipeline](fastq2vcf_june2022_update.html): in-depth explanation of an automated short-read mapping and variant calling pipeline maintained by [Harvard Informatics](https://github.com/harvardinformatics/shortRead_mapping_variantCalling).
+Useful if you plan to adopt this pre-packaged automated and parallelizable pipeline, and would like to understand its different components, but not necessarily change it substantially.
 
@@
 * [Detection and mitigation of batch effects](https://github.com/therkildsen-lab/batch-effect/blob/main/tutorial/tutorial.md): a tutorial for the detection and mitigation of batch effects with low-coverage WGS data, with example datasets and hands-on exercises.
-It is associated with the paper [(Lou and Therkildsen 2021)](https://onlinelibrary.wiley.com/doi/abs/10.1111/1755-0998.13559).
+It is associated with the paper [@Lou2021-ew](https://onlinelibrary.wiley.com/doi/abs/10.1111/1755-0998.13559).
 

### 2022-09-08 index.Rmd
push Jason's name to front page

@@
 #### [Joanna Griffiths, PhD](https://joannasgriffiths.wordpress.com/), UC-Davis {.unlisted .unnumbered}
+#### Jason Johns, UC-Santa Barbara (Website Manager) {.unlisted .unnumbered}
 #### [Joe McGirr, PhD](https://joemcgirr.github.io/), UC-Davis {.unlisted .unnumbered}

### 2022-10-26 choosing-population-genetics.Rmd
listing all 3 types of sequencing stretegies for choosing a population genetic sequencing approach

@@
 
-## 2) Reduced-representation sequencing {.unlisted .unnumbered}
+## 2) Reduced-representation sequencing
 As the name suggests, this approach aims to sequence a reduced portion of the genome across multiple individuals at moderate to high coverage for variant discovery and genotyping with high confidence.
@@
 
-## 3) Pooled individuals (Poolseq)  {.unlisted .unnumbered}
+## 3) Pooled individuals (Poolseq)
 Sequencing pools of individuals to provide information on genome-wide population allele frequencies (see review in [@futschik2010next]).

### 2022-10-26 choosing-population-genetics.Rmd
standardizing capitalization and spacing; adding a citation to poolseq-demograpjy table

@@
 text_tbl <- data.frame(
-  Goals = c("Pros","Cons","Population structure, genetic diversity","Demography (mig. rates, pop. size through time)",
-            "Signatures of selection, GWAS","Phylogenetic inference","Genetic crosses/mapping panels",
+  Goals = c("pros","cons","population structure, genetic diversity","demography (mig. rates, pop. size through time)",
+            "signatures of selection, GWAS","phylogenetic inference","genetic crosses/mapping panels",
             "molecular evolution (accurate low-freq alleles required)"),
-  hcWGS = c("provides high quality, high density genotypes; can be used to improve development of a reference genome","Most expensive per sample, requires a reference genome assembly, large computational resources","Great, but cheaper methods work almost as well","Great if incorporating haplotype-based methods ","Great","Computationally challenging, but good at all divergence levels ","Expensive","Best"),
-  lcWGS = c("Provides high density genotypes at a reduced cost","Requires a reference, can produce false (+) heterozygotes, sensitive to batch effects, may be inappropriate to use individual SNP calls for some analyses","Good, but limited to popgen analyses based on genotype likelihoods ","Good, esp.
-with some high coverage samples to help with imputation","Good","Good for tree shapes but not branch lengths, good for extracting organelle sequences and cost-effective primers for species and hybrid ID studies ","Great","Poor"),
- RADseq = c("Generally cheaper per sample, density of genotypes can be tuned to fit question, does not require a reference","Only 1-5% of genome covered which limits studies on adaptive variation, de novo assembly can result in paralogs without quality control, can be hard to capture the same loci between batches","Great, esp.
-for many samples","Good for methods based on site frequency spectrum (eg moments), but not ideal for methods using extended haplotype or phasing info","Hotly debated, but unless your genome is way too big, WGS is the way to go","good at shallow to medium divergence levels, but need to play with filtering parameters ", "great as it allows for many invididuals","poor" ),
- Poolseq = c("many individuals can be mixed for a low library prep cost, only option for some larval studies","need higher sequencing depth so it can be more expensive than RAD (but less than other options), no info on individual genotypes (duh), requires a reference, requires some replicates to account for batch effects","good for methods that only use allele frequencies, won't work for individual based methods (eg Admixture)","active area of method development, but still not common except in cases of multiple temporal samples (cite) ","good if covering majority of the genome","would work in certain situations","maybe","poor"),
+  hcWGS = c("provides high quality, high density genotypes; can be used to improve development of a reference genome","most expensive per sample, requires a reference genome assembly, large computational resources","great, but cheaper methods work almost as well","great if incorporating haplotype-based methods ","great","computationally challenging, but good at all divergence levels ","expensive","best"),
+  lcWGS = c("provides high density genotypes at a reduced cost","requires a reference, can produce false (+) heterozygotes, sensitive to batch effects, may be inappropriate to use individual SNP calls for some analyses","good, but limited to popgen analyses based on genotype likelihoods ","good, esp.
+with some high coverage samples to help with imputation","good","good for tree shapes but not branch lengths, good for extracting organelle sequences and cost-effective primers for species and hybrid ID studies ","great","poor"),
+ RADseq = c("generally cheaper per sample, density of genotypes can be tuned to fit question, does not require a reference","only 1-5% of genome covered which limits studies on adaptive variation, de novo assembly can result in paralogs without quality control, can be hard to capture the same loci between batches","great, esp.
+for many samples","good for methods based on site frequency spectrum (eg moments), but not ideal for methods using extended haplotype or phasing info","hotly debated, but unless your genome is way too big, WGS is the way to go","good at shallow to medium divergence levels, but need to play with filtering parameters ", "great as it allows for many invididuals","poor" ),
+ Poolseq = c("many individuals can be mixed for a low library prep cost, only option for some larval studies","need higher sequencing depth so it can be more expensive than RAD (but less than other options), no info on individual genotypes (duh), requires a reference, requires some replicates to account for batch effects","good for methods that only use allele frequencies, won't work for individual based methods (eg Admixture)","active area of method development, but still not common except in cases of multiple temporal samples [@nielsen2020bc] ","good if covering majority of the genome","would work in certain situations","maybe","poor"),
  Target.Capture = c("allows consistent capture of loci between batches, good for sequencing targeted regions across many samples","expensive to start a new project and requires reference of some kind to design baits (unless using method like eecSeq), only gives info on targeted region so can result in ascertainment bias for some analyses","ascertainment bias is likely to skew results, esp.
@@
 Some of the content about relative cost and available software is a little dated, but otherwise a very valuable reference.
-* [@Lou2021-me]: Great recent primer on low coverage WGS and how it compares to RADseq and Poolseq.
+* [@Lou2021-me]: great recent primer on low coverage WGS and how it compares to RADseq and Poolseq.
 
@@
 
-* [@Benjelloun2019-dq]:an empirical comparison of high coverage WGS, low coverage WGS, random variants (eg.
+* [@Benjelloun2019-dq]: an empirical comparison of high coverage WGS, low coverage WGS, random variants (eg.
 from RADseq), exome capture, and commercial SNP chip for mammals (genome size=2.6Gb).
@@
 Low coverage WGS (< 5x) should use genotype likelihood methods, however, the rate of false positives for heterozygotes is higher than classic genotyping methods at higher coverage.
-* [@Dorant2019-fb]:  A case study comparing Poolseq vs.
+* [@Dorant2019-fb]: a case study comparing Poolseq vs.
 reduced representation seq (GBS) vs Rapture (targeted capture of RAD loci) for inferring population structure with weak genetic differentiation.

### 2022-10-26 data_processing_test.Rmd
testing an attempt to consolidate all materials to this website rather than linking to another git repo, starting with the first tutorial in lcWGS

---
bibliography: RCN_site-Nov05.bib
title: "Tutorial 1: Data processing - from .fastq to .bam"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitcitations)
knitr::opts_chunk$set(fig.width = 10,
               fig.height = 5,
               cache = FALSE)
cite_options(citation_format = "pandoc", max.names = 3, style = "html", 
             hyperlink = "to.doc")
```

<br>
<br>

In this tutorial, you will learn how to go from raw sequencing files in `fastq` format through alignment files in `bam` format that we can use for downstream analysis. Along the way, you will perform multiple quality control (QC) procedures, and will map the short sequences to a snippet of a reference genome.

<br>

# Case study for practicals

Throughout this course, you will be working with data from the Atlantic silverside, *Menidia menidia*, a small estuarine fish.

```{r, echo=FALSE,out.width="100%", out.height="40%",fig.show='hold',fig.align='center'}

knitr::include_graphics(c("/Users/Jason/github/marineomics.github.io/images_test/IMG_8658_SnyderCredit.jpg"))

```

The Atlantic silverside is distributed along the east coast of North America and shows a remarkable degree of local adaptation in growth rates and a suite of other traits (Conover et al. 2005). You will be exploring patterns of genomic variation across the species range, which spans one of the steepest thermal gradient in the world.

### Today's data
Today, we will work with subsets of two different fastq files from each of three Atlantic silversides used in recent studies of fisheries-induced evolution (Therkildsen et al. 2019) and local adaptation (Wilder et al. 2020). The samples we'll use today originate from the MAQU and PANY populations shown in this map:

<br>

```{r, echo=FALSE,out.width="49%", out.height="20%"}
knitr::include_graphics("/Users/Jason/github/marineomics.github.io/images_test/Silverside_Sample_Map.png") 
```

<br>

The libraries were prepared as described in Therkildsen and Palumbi (2017) and sequenced with 125bp paired-end reads on an Illumina HiSeq instrument. There are two different fastq files for each individual because our libraries were sequenced in two different sequencing runs, to even out sequence coverage among individuals (as discussed in lecture).

We will map these raw sequence files to a snippet of the sparkling new Atlantic silverside genome (Tigano et al. nearly submitted!). To minimize computational time, we are just working with a small 2 Mb section of chromosome 24 for all the exercises in this course.

<br>

<span style='font-size:75%'> **References:**

<span style='font-size:75%'> Conover, D. O., Arnott, S. A., Walsh, M. R., & Munch, S. B. (2005). Darwinian fishery science: lessons from the Atlantic silverside (Menidia menidia). Canadian Journal of Fisheries and Aquatic Sciences, 62(4), 730–737. http://doi.org/10.1139/f05-069

<span style='font-size:75%'> Therkildsen, N. O., & Palumbi, S. R. (2017). Practical low-coverage genomewide sequencing of hundreds of individually barcoded samples for population and evolutionary genomics in nonmodel species. Molecular Ecology Resources, 17(2), 194–208. http://doi.org/10.1111/1755-0998.12593

<span style='font-size:75%'> Therkildsen, N. O., Wilder, A. P., Conover, D. O., Munch, S. B., Baumann, H., & Palumbi, S. R. (2019). Contrasting genomic shifts underlie parallel phenotypic evolution in response to fishing. Science, 365(6452), 487–490. http://doi.org/10.1126/science.aaw7271

<span style='font-size:75%'> Wilder, A. P., Palumbi, S. R., Conover, D. O., and Therkildsen, N. O. 2020. Footprints of local adaptation span hundreds of linked genes in the Atlantic silverside genome. Evolution Letters 9:177. https://doi.org/10.1002/evl3.189

</span>

<br>
<br>

# Initial preparation

## 1. Make sure you're up to speed on basic shell scripting

We'll be working almost exclusively through the command line, so if you have not used shell scripting before or are getting rusty on it, it may be helpful to have a look at a tutorial like [this one](https://linuxconfig.org/bash-scripting-tutorial-for-beginners) or a cheat sheet like [this one](https://bioinformaticsworkbook.org/Appendix/Unix/UnixCheatSheet.html#gsc.tab=0) before proceeding to the next step.

<br>

## 2. Clone this GitHub repository to your Linux server

Let's first get set up and retrieve a copy the data we will be working on, which is contained in this GitHub repository.

```{bash, eval = FALSE}

cd ~ ## Change this to the directory where you would like to store this GitHub repo

git clone https://github.com/nt246/lcwgs-guide-tutorial.git

# Go into your new tutorial1_data_processing directory and examine its contents
cd lcwgs-guide-tutorial/tutorial1_data_processing

```

<br>

Your own copy of the `tutorial1_data_processing` directory will be referred to as `BASEDIR` in many of the scripts below. Have a look inside; `tutorial1_data_processing` contains the following subdirectories:

  * `raw_fastq` has the raw fastq files we'll be working on today

  * `sample_lists` is for storing sample tables, sample lists, and other small text files

  * `reference` currently contains the reference genome file and a list of adapter sequences

  * `scripts` is for storing scripts

> Hint: We move between directories using the `cd` command and can view the content of directories with the `ls` command in the Unix shell.

<br>

In addition, you will create the following directories within `tutorial1_data_processing`

```{bash eval=FALSE}

BASEDIR=~/lcwgs-guide-tutorial/tutorial1_data_processing ## Change this to where tutorial1_data_processing is located on your server

mkdir ${BASEDIR}/adapter_clipped ## For storing your adapter clipped fastq files

mkdir ${BASEDIR}/bam ## For storing your bam (alignment) files

mkdir ${BASEDIR}/fastqc ## For storing your FastQC output

```

<br>

## 3. Orient yourself to the formatting of our fastq table and fastq list

When we get data files back from the sequencing center, the files often have obscure names, so we need a data table that let's us link those file names to our sample IDs and other information. Often, part of the fastq file name will be identical among all samples in a run and part of it will reflect some kind of unique sample identifier (either a name you supplied or a name given by the sequencing center). As an example, look in the `raw_fastq` folder and notice how all the files end in either `_1.fastq.gz` or `_2.fastq.gz` (these are the forward and reverse sequences) while the first part of the file names differs (the unique sample identifier). We call the sample identifier the `prefix`.

Our pipeline is set up to link up these `prefix` names with sample details based on a fastq table set up as the example you can find in `sample_lists/fastq_table.tsv`.

For our scripts below to work, the sample table has to be a **tab deliminated** table with the following six columns, strictly in this order:

  * `prefix` the prefix of raw fastq file names

  * `lane_number` lane number; each sequencing lane or batch should be assigned a unique identifier. This is important so that if you sequence a library across multiple different sequencing lanes, you can keep track of which lane/batch a particular set of reads came from (important for accounting for sequencing error patterns or batch effects).

  * `seq_id` sequence ID; this variable is only relevant when different libraries were prepared out of the same sample and were run in the same lane (e.g. if you wanted to include a replicate). In this case, seq_id should be used to distinguish these separate libraries. If you only have a single library prepared from each of your samples (even if you sequence that same library across multiple lanes), you can just put 1 for all samples in this column.

  * `sample_id` sample ID; a unique identifier for each individual sequenced

  * `population` population name; the population or other relevant grouping variable that the individual belongs to

  * `data_type` data type; there are only two allowed entries here: `pe` (for paired-end data) or `se` (for single end data). We need this in the table because for some of our processing steps, the commands are slightly different for paired-end and single-end data.

It is important to make sure that the combination of sample_id, seq_id, and lane_number is unique for each fastq file.

<br>

We'll also use a second file that we call a fastq list. This is simply a list of prefixes for the samples we want to analyze. Our sample table can contain data for all individuals in our study, but at any given time, we may only want to perform an operation on a subset of them. Like today, in the interest of time, we only want to run 6 sets of fastq files through each processing step.

Have a look at the list we'll be using in `sample_lists/fastq_list.txt` and note that it's just a list of fastq name prefixes, each on a separate line and there should be no header in this file.

<br>

### Activity
Compare the `fastq_list.txt` to the `fastq_table.txt`. Which populations do the samples we'll be analyzing today originate from?

<br>

With our small fastq table and fastq list here, we can easily look this up manually. But if we have hundreds of samples, that becomes more cumbersome. Let's automate the sample lookup with our first `for loop`.

<br>

## 4. Make sure you're familiar with `for loops` and how to assign and call variables in bash

In low-coverage whole genome sequencing datasets, we'll typically have data from hundreds of individuals, so we need an efficient way to process all of these files without having to write a separate line of code for each file. `for loops` are a powerful way to achieve this, and we will be using them in every step of our pipeline, so let's first take a moment to make sure we understand the syntax.

A ‘for loop’ is a bash programming language statement which allows code to be repeatedly executed. If you've never worked with `for loops` in bash before, it might be helpful to look over a tutorial, e.g. [this one from Software Carpentry](https://swcarpentry.github.io/shell-novice/05-loop/index.html).

Here is a key extract inspired from that tutorial:

The basic syntax of a `for loop` is as follows

```{bash, eval = FALSE}

for thing in list_of_things; do    # ; is equivalent to an end-of-line. We can alternatively put "do" on its own line

    operation_using $thing    # Indentation within the loop is not required, but aids legibility

done

```

When the shell sees the keyword `for`, it knows to repeat a command (or group of commands) once for each item in a list. Each time the loop runs (called an iteration), an item in the list is assigned in sequence to the variable, and the commands inside the loop are executed, before moving on to the next item in the list. Inside the loop, we call the variable’s value by putting `$` in front of it. The `$` tells the shell interpreter to treat the variable as a variable name and substitute its value in its place, rather than treat it as text or an external command.

<br>

## 5. Practice using bash `for loops` to iterate over target samples
First, let's just look up the sample IDs. For each prefix in our `fastq_list.txt`, we will use `grep` to extract the relevant line from the fastq table, use `cut` to extract the column with sample ID, and then `echo` to print the sample ID

```{bash, eval = FALSE}

BASEDIR=~/lcwgs-guide-tutorial/tutorial1_data_processing ## Change this to where tutorial1_data_processing is located on your server

SAMPLELIST=$BASEDIR/sample_lists/fastq_list.txt # Path to a list of prefixes of the raw fastq files. It should be a subset of the the 1st column of the fastq table.

SAMPLETABLE=$BASEDIR/sample_lists/fastq_table.tsv # Path to a fastq table where the 1st column is the prefix of the raw fastq files. The 4th column is the sample ID.

for SAMPLEFILE in `cat $SAMPLELIST`; do   # Loop through each of the prefixes listed in our fastq list

	# For each prefix, extract the associated sample ID (column 4) from the table
	SAMPLE_ID=`grep -P "${SAMPLEFILE}\t" $SAMPLETABLE | cut -f 4`

	echo $SAMPLEFILE refers to sample $SAMPLE_ID

done

```

<br>

### Exercise 
Now change the `for loop` so it also outputs which population each fastq file has data for.

<br>

#### See a solution
<details>
  <summary>Click here to expand</summary>

```{bash, eval = FALSE}

for SAMPLEFILE in `cat $SAMPLELIST`; do   # Loop through each of the prefixes listed in our sample list

	# For each prefix, extract the associated sample ID (column 4) and population ID (column 5) from the table
	SAMPLE_ID=`grep -P "${SAMPLEFILE}\t" $SAMPLETABLE | cut -f 4` 
	POPULATION=`grep -P "${SAMPLEFILE}\t" $SAMPLETABLE | cut -f 5`

	echo $SAMPLEFILE refers to sample $SAMPLE_ID from $POPULATION

done

```

</details>

<br>

## 6. Define paths to the project directory and programs

We need to make sure the server knows where to find the programs we'll be running and our input and output directories. This will always need to be specified every time we run our scripts in a new login session.

<br>

### Set the project directory as a variable named `BASEDIR`

```{bash, eval = FALSE}

BASEDIR=~/lcwgs-guide-tutorial/tutorial1_data_processing ## Change this to where tutorial1_data_processing is located on your server

```

<br>

### Specify the paths to required programs as variables

You will need to install the programs below on your server if they are not already installed, and define their paths as follows. The paths to these program on Cornell's BioHPC servers are provided below.

```{bash, eval = FALSE}

FASTQC=/programs/bin/fastqc/fastqc
TRIMMOMATIC=/programs/trimmomatic/trimmomatic-0.39.jar
PICARD=/programs/picard-tools-2.19.2/picard.jar
SAMTOOLS=/programs/bin/samtools/samtools
BOWTIEBUILD=/programs/bin/bowtie2/bowtie2-build
BOWTIE=/programs/bin/bowtie2/bowtie2
BAMUTIL=/programs/bamUtil/bam

```

The following programs are used for indel realignment and are optional.

```{bash, eval = FALSE}

GATK=/programs/GenomeAnalysisTK-3.7/GenomeAnalysisTK.jar ## Note that GATK-3.7 is needed for indel-realignment, which is no longer suppported in new versions
JAVA=/usr/local/jdk1.8.0_121/bin/java ## An older version of java is required for GATK-3.7 to run on Cornell's BioHPC server

```

<br>
If you will be running these programs on a different system, you will have to specify the paths to the different programs on that system (or add them to your $PATH).

<br>
<br>

# Data processing pipeline

Now let's get started processing the data!

<br>

## Examine the raw fastq files

### fastq file structure

A FASTQ file normally contains four lines per sequence.

  * Line 1 contains the sequence identifier, with information on the sequencing run and the cluster. The exact content of this line varies depending on how fastq files are generated from the sequencer.
  * Line 2 is the raw sequence. 
  * Line 3 often consists of a single `+` symbol. 
  * Line 4 encodes the quality of each base in the sequence in Line 2 (i.e. the probability of sequencing error in log scale). For most current sequencers, these base qualities are encoded in the [Phred33 format](https://drive5.com/usearch/manual/quality_score.html), but always check to make sure how your quality scores are encoded.

Now read the code below, guess what it does, and run it on your own. Does it do what you expect it to do? Inspect the output and try to identify the group of four lines for each read.

<br>

```{bash, eval = FALSE}

SAMPLELIST=$BASEDIR/sample_lists/fastq_list.txt # Path to the sample list.
RAWFASTQSUFFIX1=_1.fastq.gz # Suffix to raw fastq files. Use forward reads with paired-end data.

for SAMPLE in `cat $SAMPLELIST`; do

  echo $SAMPLE
  zcat $BASEDIR'/raw_fastq/'$SAMPLE$RAWFASTQSUFFIX1 | head -n 8
  echo ' '

done

```

<br>

### Evaluate the overall data quality

With a new batch of data, it is always to good idea to start out by getting an overview of the data quality and look for any signs of quality issues. The [FastQC](https://www.bioinformatics.babraham.ac.uk/projects/fastqc/) program provides a useful set of diagnostics, so we'll run it on each on our fastq files to check their quality. In the interest of time, we will only look at three of our fastqs. We'll loop over each of these and call FastQC on the forward file only (the program only takes the path to the input and the path to where you want the output as parameters). When working with your own data, always make sure to look both at the forward and reverse files because sometimes issues can arise in only one of the read directions.

<br>

```{bash, eval = FALSE}

SAMPLELIST=$BASEDIR/sample_lists/fastq_list.txt # Path to the sample list.
RAWFASTQSUFFIX1=_1.fastq.gz # Suffix to raw fastq files. We'll only look at the forward reads here

for SAMPLE in `cat $SAMPLELIST | head -n 3`; do  # The head -n 3 is taking just the first three elements of our fastq list to loop over

  $FASTQC $BASEDIR'/raw_fastq/'$SAMPLE$RAWFASTQSUFFIX1 -o $BASEDIR'/fastqc/'

done

```

<br>

If the program ran, you should now see the output (in html format and a zip file with various files) in your `fastqc` directory. To view the .html, use `scp` or FileZilla to transfer the html output files to your local machine and open them in a web browser, or use something like R-Studio server to directory view it.

<br>

#### Question: 
Do you notice anything different about the fastQC reports from the three different fastq files?

<br>

The libraries for these samples were prepared in different batches. Below are representative Bioanalyzer traces for each of the batches. Which sample do you think came from which batch?

<br>

```{r, echo=FALSE,out.width="49%", out.height="20%",fig.show='hold',fig.align='center'}

knitr::include_graphics(c("/Users/Jason/github/marineomics.github.io/images_test/Bioanalyzer_Batch2.png"))

```

<br>
<br>

## Adapter clipping

When the insert length of a library fragment is shorter than the read length, the sequencer will read into the adapter sequence (as shown below). This means that the end of the read will not be from our actual sample, but will be adapter sequence, which may lead to lower alignment performance and even biases in the result if not removed.

![](https://www.ecseq.com/support/ngs/img/fragmentsize.png)

<br>

We saw in our FastQC report that we have substantial adapter content in some of our libraries, so will will need to clip that off. Here, we use [Trimmomatic](http://www.usadellab.org/cms/?page=trimmomatic) to clip the adapter sequence off the ends of reads where they appear. This step requires us to input the known adapter sequences that we used when preparing the libraries. In this exercise, the libraries were prepared using Illumina's Nextera adapters (sequences listed in `reference/NexteraPE_NT.fa`), and we will read those into the `ADAPTERS` variable.

Look over the code below. The first block of text specifies which files we are using as input. Then we start looping over our samples. Within the loop, the first step is to extract the relevant sample data from our sample table and assign those as temporary variables. Then we have two `if statements` to call Trimmomatic with slightly different parameters for paired-end and single-end data. Trimmomatic has lots of different filtering modules that can be useful in different contexts. Here we only clip sequence that match to our adapter sequence and remove reads that end up being <40bp after clipping.

<br>

```{bash, eval = FALSE}

SAMPLELIST=$BASEDIR/sample_lists/fastq_list.txt # Path to a list of prefixes of the raw fastq files. It should be a subset of the the 1st column of the sample table.
SAMPLETABLE=$BASEDIR/sample_lists/fastq_table.tsv # Path to a sample table where the 1st column is the prefix of the raw fastq files. The 4th column is the sample ID, the 2nd column is the lane number, and the 3rd column is sequence ID. The combination of these three columns have to be unique. The 6th column should be data type, which is either pe or se. 
RAWFASTQDIR=$BASEDIR/raw_fastq/ # Path to raw fastq files. 
RAWFASTQSUFFIX1=_1.fastq.gz # Suffix to raw fastq files. Use forward reads with paired-end data.
RAWFASTQSUFFIX2=_2.fastq.gz # Suffix to raw fastq files. Use reverse reads with paired-end data. 
ADAPTERS=$BASEDIR/reference/NexteraPE_NT.fa # Path to a list of adapter/index sequences.

## Loop over each sample
for SAMPLEFILE in `cat $SAMPLELIST`; do

	## Extract relevant values from a table of sample, sequencing, and lane ID (here in columns 4, 3, 2, respectively) for each sequenced library
	SAMPLE_ID=`grep -P "${SAMPLEFILE}\t" $SAMPLETABLE | cut -f 4`
	POP_ID=`grep -P "${SAMPLEFILE}\t" $SAMPLETABLE | cut -f 5`
	SEQ_ID=`grep -P "${SAMPLEFILE}\t" $SAMPLETABLE | cut -f 3`
	LANE_ID=`grep -P "${SAMPLEFILE}\t" $SAMPLETABLE | cut -f 2`
	SAMPLE_UNIQ_ID=$SAMPLE_ID'_'$POP_ID'_'$SEQ_ID'_'$LANE_ID  # When a sample has been sequenced in multiple lanes, we need to be able to identify the files from each run uniquely

	## Extract data type from the sample table
	DATATYPE=`grep -P "${SAMPLEFILE}\t" $SAMPLETABLE | cut -f 6`

	## The input and output path and file prefix
	RAWFASTQ_ID=$RAWFASTQDIR$SAMPLEFILE
	SAMPLEADAPT=$BASEDIR'/adapter_clipped/'$SAMPLE_UNIQ_ID

	## Adapter clip the reads with Trimmomatic
	# The options for ILLUMINACLIP are: ILLUMINACLIP:<fastaWithAdaptersEtc>:<seed mismatches>:<palindrome clip threshold>:<simple clip threshold>:<minAdapterLength>:<keepBothReads>
	# The MINLENGTH drops the read if it is below the specified length in bp
	# For definitions of these options, see http://www.usadellab.org/cms/uploads/supplementary/Trimmomatic/TrimmomaticManual_V0.32.pdf

	if [ $DATATYPE = pe ]; then
		java -jar $TRIMMOMATIC PE -threads 1 -phred33 $RAWFASTQ_ID$RAWFASTQSUFFIX1 $RAWFASTQ_ID$RAWFASTQSUFFIX2 $SAMPLEADAPT'_adapter_clipped_f_paired.fastq.gz' $SAMPLEADAPT'_adapter_clipped_f_unpaired.fastq.gz' $SAMPLEADAPT'_adapter_clipped_r_paired.fastq.gz' $SAMPLEADAPT'_adapter_clipped_r_unpaired.fastq.gz' 'ILLUMINACLIP:'$ADAPTERS':2:30:10:1:true MINLENGTH:40'

	elif [ $DATATYPE = se ]; then
		java -jar $TRIMMOMATIC SE -threads 1 -phred33 $RAWFASTQ_ID$RAWFASTQSUFFIX1 $SAMPLEADAPT'_adapter_clipped_se.fastq.gz' 'ILLUMINACLIP:'$ADAPTERS':2:30:10 MINLENGTH:40'
	fi

done

```

<br>

Have a look at the output printed to the screen as we're iterating over the samples. Note the first time it says

`TrimmomaticPE: Started with arguments:`

Following this, you will see that actual variable names that were added to the command in each iteration of our loop (e.g. what `$RAWFASTQ_ID$RAWFASTQSUFFIX1` expanded to (what value was assigned to this variable)).

Also examine the section that says
`ILLUMINACLIP: Using 2 prefix pairs, 8 forward/reverse sequences, 0 forward only sequences, 0 reverse only sequences
Input Read Pairs:`

This will show how many reads were removed from our filtering.

<br>

#### Question
The first three samples are the samples for which we examined the FastQC outputs. Do you expect different amount of sequence getting removed from each of these? Is that what you see? Why/Why not?

<br>

<details>
  <summary>Discuss first, then you can click here for a hint</summary>

<br>

The output from Trimmomatic only shows how many full reads get removed, not how much the reads within the file get truncated. In the library with lots of adapter, many of the reads will now be shorter, but as long as they're still longer than our threshold of 40bp, they will not get removed. If we wanted to know how much sequence we lost from each fastq, counting the number of bases lost is more informative than the number of sequences. It is also always a good idea to check the adapter clipped fastq files with FastQC to make sure that you did in fact get rid of the adapter sequence. We don't have time to do this in class, but if you're interested, you can use the fastqc loop above and just change it to run on the files in your `adapter_clipped` folder.

</details>

<br>
<br>

## OPTIONAL: Quality trimming
As we saw in our FastQC output, the base call quality score tends to drop off towards the ends of the reads. As we'll learn more about tomorrow, probabilistic analysis frameworks, like `ANGSD` and others based on genotype likelihoods, can take the base call quality into account and that way give less weight to a base call that is less certain.

However, as a conservative measure, we may want to just trim off the rest of the read if the quality score drops too low over multiple bases. We can do this with the `SLIDINGWINDOW` module in [Trimmomatic](http://www.usadellab.org/cms/uploads/supplementary/Trimmomatic/TrimmomaticManual_V0.32.pdf). We won't have time to do that in this practical, but if you're interested, you can modify the Trimmomatic code above to also trim off low-quality bases.

<br>
<br>

## Build reference index files

There are lots of different programs developed for mapping short reads to a reference sequence. We will use the program [bowtie2](http://bowtie-bio.sourceforge.net/bowtie2/index.shtml). This program requires a set of reference index files to be able to perform the sequence alignment. So we will start by indexing our reference.

<br>

```{bash, eval = FALSE}

REFERENCE=$BASEDIR/reference/mme_physalia_testdata_chr24.fa   # This is a fasta file with the reference genome sequence we will map to 
REFBASENAME="${REFERENCE%.*}"
$SAMTOOLS faidx $REFERENCE

java -jar $PICARD CreateSequenceDictionary R=$REFERENCE O=$REFBASENAME'.dict'

$BOWTIEBUILD $REFERENCE $REFBASENAME

```

<br>
<br>

## Map to the reference, sort, and quality filter

In this step, we align the short reads within each fastq file to the reference genome using `bowtie2`. The resulting alignment file, in `sam` format, will be converted to a binary format `bam` for more efficient storage. Each mapped read will have a mapping quality, which indicates how confident that mapper is that a read is mapped in the correct position. The [bowtie2 manual](http://bowtie-bio.sourceforge.net/bowtie2/manual.shtml) defines it as "a non-negative integer Q = -10 log10 p, where p is an estimate of the probability that the alignment does not correspond to the read's true point of origin." Accordingly, a mapping quality (or MAPQ) of 10 or less indicates that there is at least a 1 in 10 chance that the read truly originated elsewhere, and a MAPQ of 20 indicates at least a 1 in 100 chance.

Here, to only retain reads for which we are reasonably certain have been mapped in the correct place, we will filter out reads with a mapping quality lower than 20, and after that sort the filtered alignment file for easier computation in the next step.

Look over the code and make sure you understand what it's doing, then copy and run it.

<br>

```{bash, eval = FALSE}

SAMPLELIST=$BASEDIR/sample_lists/fastq_list.txt # Path to a list of prefixes of the raw fastq files. It should be a subset of the the 1st column of the sample table.
SAMPLETABLE=$BASEDIR/sample_lists/fastq_table.tsv # Path to a sample table where the 1st column is the prefix of the raw fastq files. The 4th column is the sample ID, the 2nd column is the lane number, and the 3rd column is sequence ID. The combination of these three columns have to be unique. The 6th column should be data type, which is either pe or se. 
FASTQDIR=$BASEDIR/adapter_clipped/ # Path to the directory where fastq file are stored. 
FASTQSUFFIX1=_adapter_clipped_f_paired.fastq.gz # Suffix to fastq files. Use forward reads with paired-end data. 
FASTQSUFFIX2=_adapter_clipped_r_paired.fastq.gz # Suffix to fastq files. Use reverse reads with paired-end data. 
MAPPINGPRESET=very-sensitive # The pre-set option to use for mapping in bowtie2 (very-sensitive for end-to-end (global) mapping [typically used when we have a full genome reference], very-sensitive-local for partial read mapping that allows soft-clipping [typically used when mapping genomic reads to a transcriptome]
REFERENCE=$BASEDIR/reference/mme_physalia_testdata_chr24.fa # Path to reference fasta file and file name
REFNAME=mme_physalia_testdata_chr24 # Reference name to add to output files, e.g. gadMor2

## Loop over each sample
for SAMPLEFILE in `cat $SAMPLELIST`; do

	## Extract relevant values from a table of sample, sequencing, and lane ID (here in columns 4, 3, 2, respectively) for each sequenced library
	SAMPLE_ID=`grep -P "${SAMPLEFILE}\t" $SAMPLETABLE | cut -f 4`
	POP_ID=`grep -P "${SAMPLEFILE}\t" $SAMPLETABLE | cut -f 5`
	SEQ_ID=`grep -P "${SAMPLEFILE}\t" $SAMPLETABLE | cut -f 3`
	LANE_ID=`grep -P "${SAMPLEFILE}\t" $SAMPLETABLE | cut -f 2`
	SAMPLE_UNIQ_ID=$SAMPLE_ID'_'$POP_ID'_'$SEQ_ID'_'$LANE_ID  # When a sample has been sequenced in multiple lanes, we need to be able to identify the files from each run uniquely

	## Extract data type from the sample table
	DATATYPE=`grep -P "${SAMPLEFILE}\t" $SAMPLETABLE | cut -f 6`

	## The input and output path and file prefix
	SAMPLETOMAP=$FASTQDIR$SAMPLE_UNIQ_ID
	SAMPLEBAM=$BASEDIR'/bam/'$SAMPLE_UNIQ_ID

	## Define platform unit (PU), which is the lane number
	PU=`grep -P "${SAMPLEFILE}\t" $SAMPLETABLE | cut -f 2`

	## Define reference base name
	REFBASENAME="${REFERENCE%.*}"

	## Map reads to the reference 
	echo $SAMPLE_UNIQ_ID

	# Map the paired-end reads
	if [ $DATATYPE = pe ]; then 
	# We ignore the reads that get orphaned during adapter clipping because that is typically a very small proportion of reads. If a large proportion of reads get orphaned (loose their mate so they become single-end), these can be mapped in a separate step and the resulting bam files merged with the paired-end mapped reads.
	$BOWTIE -q --phred33 --$MAPPINGPRESET -p 1 -I 0 -X 1500 --fr --rg-id $SAMPLE_UNIQ_ID --rg SM:$SAMPLE_ID --rg LB:$SAMPLE_ID --rg PU:$PU --rg PL:ILLUMINA -x $REFBASENAME -1 $SAMPLETOMAP$FASTQSUFFIX1 -2 $SAMPLETOMAP$FASTQSUFFIX2 -S $SAMPLEBAM'_'$DATATYPE'_bt2_'$REFNAME'.sam'

	# Map the single-end reads
	elif [ $DATATYPE = se ]; then
	$BOWTIE -q --phred33 --$MAPPINGPRESET -p 1 --rg-id $SAMPLE_UNIQ_ID --rg SM:$SAMPLE_ID --rg LB:$SAMPLE_ID --rg PU:$PU --rg PL:ILLUMINA -x $REFBASENAME -U $SAMPLETOMAP$FASTQSUFFIX1 -S $SAMPLEBAM'_'$DATATYPE'_bt2_'$REFNAME'.sam'

	fi

	## Convert to bam file for storage (including all the mapped reads)
	$SAMTOOLS view -bS -F 4 $SAMPLEBAM'_'$DATATYPE'_bt2_'$REFNAME'.sam' > $SAMPLEBAM'_'$DATATYPE'_bt2_'$REFNAME'.bam'
	rm -f $SAMPLEBAM'_'$DATATYPE'_bt2_'$REFNAME'.sam'

	## Filter the mapped reads (to onky retain reads with high mapping quality)
	# Filter bam files to remove poorly mapped reads (non-unique mappings and mappings with a quality score < 20) -- do we want the quality score filter??
	$SAMTOOLS view -h -q 20 $SAMPLEBAM'_'$DATATYPE'_bt2_'$REFNAME'.bam' | $SAMTOOLS view -buS - | $SAMTOOLS sort -o $SAMPLEBAM'_'$DATATYPE'_bt2_'$REFNAME'_minq20_sorted.bam'

done

```

<br>
<br>

Have a look at the output printed to the screen. How many of our reads are mapping to the reference? Does it vary between samples? Is there anything that seems weird to you about the stats reported?

<br>
<br>

## Examine the bam files

SAM stands for Sequence Alignment/Map format. BAM is the binary format for sam files (which takes up much less space). It is a TAB-delimited text format consisting of a header section, which is optional, and an alignment section. If present, the header must be prior to the alignments. Header lines start with ‘@’, while alignment lines do not. Each alignment line has 11 mandatory fields for essential alignment information such as mapping position, and variable number of optional fields for flexible or aligner specific information. BAM is the binary version of the SAM format.

See the full documentation of the sam file format [here](https://samtools.github.io/hts-specs/SAMv1.pdf) or a quick overview of the column descriptors [here](https://en.wikipedia.org/wiki/SAM_(file_format))

Note that we have converted our sam files to bam files. That's useful for saving disk space, but because bam files are binary, they are not human readable. However, we can use the `view` utility in the program [samtools](http://www.htslib.org/doc/samtools.html) to convert the content back to human readable output to we can examine our alignments.

As an example, let's look at the output for `985_PANY`. The following command can be used to inspect the first eight lines the sorted bam file for this sample.

```{bash, eval = FALSE}

$SAMTOOLS view $BASEDIR/bam/985_PANY_1_lane1_pe_bt2_mme_physalia_testdata_chr24_minq20_sorted.bam | head -n 8

```

<br>

Take a few minutes to look at the output and look at the [column descriptors](https://en.wikipedia.org/wiki/SAM_(file_format)) to understand its content.

<br>

**OPTIONAL exercise:** Write a loop to print the first three alignments of all the sorted bam files that you generated in the last step. You can use the general template code below as a starting point.

```{bash, eval = FALSE}

$SAMTOOLS view $SAMPLEBAM'_'$DATATYPE'_bt2_'$REFNAME'_minq20_sorted.bam' | head -n 3

```

<br>
<br>

## Merge samples that were sequenced in multiple batches

We have now mapped the two separate sets of fastq files for each sample (the separate sets generated in independent sequencing runs), so we also have two bam files for each sample. If the two sequencing runs were performed with aliquots of the same library, we will need to merge the bam files before we remove duplicate sequences because the two sequencing lanes would have been sequencing the same set of molecules (so each may contain duplicate fragments also sequenced in the other). For most downstream analysis, it's also a lot more convenient to only have a single bam file per individual.

<br>

We will merge the two bam files for each individual with [samtools merge](http://www.htslib.org/doc/samtools-merge.html) with the following parameters

```{bash, eval = FALSE}

$SAMTOOLS merge output.bam input1.bam input2.bam   # We replace the output.bam with the name we want to give the output merged bam and the two input names with the names of the bam files we want to merge.

```

We could write a for loop to get the separate fastq files for each individual merged. But in this case, we will instead run a shell script that has a line that calls `samtools merge` for each individual. You can find the script `merge_bams.sh` in the `scripts` folder. Have a look at it.

With just three samples, we could quickly copy and edit these three lines to construct this shell script manually. But if we had hundreds of samples, that approach would become error-prone so we want to use a script to automate it. In the interest of time, let's move forward with the merging script we have already generated, but we provide R-code below that generates this script from the your fastq-level sample table and list, so you don't need to do this manually if you're working with your own samples.

Since we'll be working on bam files rather than fastq files downstream from this point, we need to list bam IDs rather than fastq IDs in the sample lists that specify which samples we want to loop over in a particular pipeline step. The R-code below generates that as well, but for today, we'll just use the list that we've already added to `$BASEDIR/sample_lists`.

<br>

#### The following R code is just provided for your reference. You DON'T need to run it - the output is already generated in your `scripts` and `sample_lists` folders

<details>
  <summary>Click here to view the R code</summary>

<br>

```{r eval=F}

library(tidyverse)

basedir <- "~/tutorial1_data_processing/"
refname <- "mme_physalia_testdata_chr24"

fastq_list <- read_lines(paste0(basedir, "sample_lists/fastq_list.txt"))
fastq_table <- read_tsv(paste0(basedir, "sample_lists/fastq_table.tsv"))

#Subset the table to the samples we're currently interested in analyzing
target_samples <- filter(fastq_table, prefix %in% fastq_list)

## Find all duplicated samples
duplicated_samples <- (target_samples$sample_id)[duplicated(target_samples$sample_id)] %>% unique()

# Write cd
write_lines(c("BASEDIR=$1", "cd $BASEDIR'/bam'\n"), paste0(basedir, "scripts/merge_bam.sh"))

## Loop through all duplicated samples 
for (i in 1:length(duplicated_samples)){
  dup_sample_dat <- filter(fastq_table, sample_id==duplicated_samples[i])

  ## Extract the bam file names from the unmerged sample table
  input <- dup_sample_dat %>%
    mutate(unmerged_bam = paste(sample_id, population, seq_id, lane_number, data_type, "bt2", refname, "minq20_sorted.bam", sep = "_")) %>% 
    # We are reconstructing the $SAMPLE_SEQ_ID that is the first part of the separate bam file names
    .$unmerged_bam

  output <- paste(dup_sample_dat[1,"sample_id"], dup_sample_dat[1, "population"], "merged_bt2", refname, "minq20_sorted.bam", sep = "_")

  ## Paste together the command line
  #merging_script[i+1] <- paste("samtools merge", as.character(output), input[1], input[2], sep = " ")
  write_lines(paste("samtools merge", as.character(output), input[1], input[2], sep = " "), paste0(basedir, "scripts/merge_bam.sh"), append = TRUE)

  # Also write a target bam list that we'll use for downstream looping over merged bam files
  if (i == 1){
  write_lines(paste(dup_sample_dat[1,"sample_id"], dup_sample_dat[1, "population"], "merged", sep = "_"), paste0(basedir, "sample_lists/merged_bam_list.txt"))
  }

  if (i > 1){
  write_lines(paste(dup_sample_dat[1,"sample_id"], dup_sample_dat[1, "population"], "merged", sep = "_"), paste0(basedir, "sample_lists/merged_bam_list.txt"), append = TRUE)
  }

}

```

</details>

<br>
<br>

### Run the merging script

To execute the merging, run the bash script with the following command:

```{bash, eval = FALSE}

bash $BASEDIR/scripts/merge_bam.sh $BASEDIR

```

<br>

Check that your merged bam files get generated. As a sanity check, we can compare the number of lines in our merged and set of unmerged bam files for each sample with [samtools view](http://www.htslib.org/doc/samtools-view.html) and the command `samtools view in.bam | wc -l`

Run on our server, for the merged file for sample 985, the command would like like

```{bash, eval = FALSE}

$SAMTOOLS view $BASEDIR/bam/985_PANY_merged_bt2_mme_physalia_testdata_chr24_minq20_sorted.bam | wc -l

```

Check the line count in the bam files for the two individual fastqs and see if the numbers add up.

```{bash, eval = FALSE}

$SAMTOOLS view $BASEDIR/bam/985_PANY_1_lane2_pe_bt2_mme_physalia_testdata_chr24_minq20_sorted.bam | wc -l
$SAMTOOLS view $BASEDIR/bam/985_PANY_1_lane1_pe_bt2_mme_physalia_testdata_chr24_minq20_sorted.bam | wc -l

```

<br>

**OPTIONAL exercise**: You could write a for loop that will extract the line count for each file.

<br>

### Discussion question
This merging procedure required some extra effort. Why didn't we just merge the fastq files for each individual before mapping?

<br>

##### Answer

<details>
  <summary>Click here</summary>

<br>

Because there may be particular issues associated each sequencing lane (in particular the base call quality score calibration may vary), so for some downstream analyses, we need to keep track of what data were sequenced in what lane. By mapping the fastq files separately, we were able to add read platform unit `PU` information to the read group tag in the bam file while mapping with `bowtie2` (see the script). This way we can continue to keep track of which read came from which lane, and could even filter our bam file based on this later on, if we ended up wanting to compare data from different lanes for troubleshooting or other reasons.

</details>

<br>
<br>

## Deduplicate and clip overlapping read pairs

Here, we remove the PCR duplicates and trim the overlapping part of each read pair in pair-end data. It is important to wait to deduplicate until after merging, because PCR duplicates for the same sample may exist in different lanes. We use the [Picard Tools MarkDuplicates](https://gatk.broadinstitute.org/hc/en-us/articles/360037052812-MarkDuplicates-Picard-).

We also want to clip overlapping reads. We will use the [BamUtil clipOverlap](https://genome.sph.umich.edu/wiki/BamUtil:_clipOverlap)

```{r, echo=FALSE,out.width="49%", out.height="20%"}
knitr::include_graphics("/Users/Jason/github/marineomics.github.io/images_test/clip_overlap.png") 
```

<br>

```{bash, eval = FALSE}

BAMLIST=$BASEDIR/sample_lists/merged_bam_list.txt # Path to a list of merged bam files.
REFNAME=mme_physalia_testdata_chr24 # Reference name to add to output files

## Loop over each sample
for SAMPLEBAM in `cat $BAMLIST`; do

	## Remove duplicates and print dupstat file
	# We used to be able to just specify picard.jar on the CBSU server, but now we need to specify the path and version
	java -Xmx60g -jar $PICARD MarkDuplicates I=$BASEDIR'/bam/'$SAMPLEBAM'_bt2_'$REFNAME'_minq20_sorted.bam' O=$BASEDIR'/bam/'$SAMPLEBAM'_bt2_'$REFNAME'_minq20_sorted_dedup.bam' M=$BASEDIR'/bam/'$SAMPLEBAM'_bt2_'$REFNAME'_minq20_sorted_dupstat.txt' VALIDATION_STRINGENCY=SILENT REMOVE_DUPLICATES=true

	## Clip overlapping paired end reads (only necessary for paired-end data, so if you're only running se samples, you can comment this step out)
	$BAMUTIL clipOverlap --in $BASEDIR'/bam/'$SAMPLEBAM'_bt2_'$REFNAME'_minq20_sorted_dedup.bam' --out $BASEDIR'/bam/'$SAMPLEBAM'_bt2_'$REFNAME'_minq20_sorted_dedup_overlapclipped.bam' --stats

done
```

<br>

MarkDuplicates has very verbose output - take a look at it to make sure the programs ran and didn't throw an error.

Next, look at the output from the clipOverlap. Was there a substantial difference in how much sequence got clipped in the three samples, and does that make sense in light of their FastQC reports?

Use `head` to look at the top lines of the dupstat report for each sample (in `$BASEDIR/bam/XXX_merged_bt2_mme_physalia_testdata_chr24_minq20_sorted_dupstat.txt` [replace the XXX with the sample [prefix]]). Can you spot where the duplication rate is reported. Does it vary between our samples?

<br>
<br>

## Indel realignment (optional)

Unlike other variant detector programs (like the [GATK Haplotype Caller](https://gatk.broadinstitute.org/hc/en-us/articles/360037225632-HaplotypeCaller) or [Freebayes](https://github.com/ekg/freebayes)), [ANGSD](http://www.popgen.dk/angsd/index.php/ANGSD) does not realign reads during its analysis. Because it can be difficult to distinguish indels from SNPs at the end of reads if each alignment is considered separately, indels may interfere with genotype likelihood estimation. We therefore recommend running your bam files through a program that realigns reads around indels prior to running `ANGSD`. The [GATK IndelRealigner](https://github.com/broadinstitute/gatk-docs/blob/master/gatk3-tutorials/(howto)_Perform_local_realignment_around_indels.md) takes all the aligned sequences from all samples in to account to validate the indels discovered from the mapping process and then realigns each read locally. This is not a mandatory step and tends to be very time-consuming if you have a large dataset, but the code is provided here if you want to give it a try (it takes ~3 minutes). We will not use these output for the rest of this course though.

<details>
  <summary>Click here to see the GATK IndelRealigner code</summary>

```{bash, eval = FALSE}
BAMLIST=$BASEDIR/sample_lists/bam_list_dedup_overlapclipped.list # Path to a list of merged, deduplicated, and overlap clipped bam files. Full paths should be included. This file has to have a suffix of ".list"
REFERENCE=$BASEDIR/reference/mme_physalia_testdata_chr24.fa # Path to reference fasta file and file name
REFNAME=mme_physalia_testdata_chr24 # Reference name to add to output files

for SAMPLEBAM in `cat $BASEDIR/sample_lists/merged_bam_list.txt`; do
echo $BASEDIR'/bam/'$SAMPLEBAM'_bt2_mme_physalia_testdata_chr24_minq20_sorted_dedup_overlapclipped.bam' >> $BAMLIST
done

## Loop over each sample
cd $BASEDIR/bam/
for SAMPLEBAM in `cat $BAMLIST`; do
	$SAMTOOLS index $SAMPLEBAM
done

## Realign around in-dels
# This is done across all samples at once

## Create list of potential in-dels
$JAVA -Xmx40g -jar $GATK \
-T RealignerTargetCreator \
-R $REFERENCE \
-I $BAMLIST \
-o $BASEDIR'/bam/all_samples_for_indel_realigner.intervals' \
-drf BadMate

## Run the indel realigner tool
$JAVA -Xmx40g -jar $GATK \
-T IndelRealigner \
-R $REFERENCE \
-I $BAMLIST \
-targetIntervals $BASEDIR'/bam/all_samples_for_indel_realigner.intervals' \
--consensusDeterminationModel USE_READS  \
--nWayOut _realigned.bam
```

</details>

<br>
<br>

## Estimate read depth in your bam files

After all the filtering steps, we want to know what final depth of coverage we have for each sample for downstream analysis. Here, we will use [samtools depth](http://www.htslib.org/doc/samtools-depth.html) to first compute the read depth at each bp position in the genome. Then we will pull the output file to our local machines and compute depth summary stats in R. This is just one way to summarize the depth. There are other programs available to provide summaries of the depth in a bam file, including [samtools coverage](http://www.htslib.org/doc/samtools-coverage.html), [Mosdepth](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6030888/) and [Indexcov](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5737511/).

<br>

First, run `samtools depth` to get depth per sample per position.

```{bash eval=F}

BAMLIST=$BASEDIR/sample_lists/merged_bam_list.txt # Path to a list of unique sample prefixes for merged bam files.	
REFNAME=mme_physalia_testdata_chr24 # Reference name to add to output files

for SAMPLEBAM in `cat $BAMLIST`; do	
	## Count per position depth per sample
	$SAMTOOLS depth -aa $BASEDIR'/bam/'$SAMPLEBAM'_bt2_'$REFNAME'_minq20_sorted_dedup_overlapclipped.bam' | cut -f 3 | gzip > $BASEDIR'/bam/'$SAMPLEBAM'_bt2_'$REFNAME'_minq20_sorted_dedup_overlapclipped.bam.depth.gz'

done

```

<br>

Then, we'll process and visualize the data in R. We recommend that you use R-Studio server for this, but running R on command line also works.

```{r eval = FALSE}
## install.packages("tidyverse") ## Install tidyverse if you don't have it installed yet
library(tidyverse)

basedir <- "~/lcwgs-guide-tutorial/tutorial1_data_processing" # Make sure to edit this to match your $BASEDIR
bam_list <- read_lines(paste0(basedir, "/sample_lists/merged_bam_list.txt"))

for (i in 1:length(bam_list)){
  bamfile = bam_list[i]
  # Compute depth stats
  depth <- read_tsv(paste0(basedir, "/bam/", bamfile, "_bt2_mme_physalia_testdata_chr24_minq20_sorted_dedup_overlapclipped.bam.depth.gz"), col_names = F)$X1
  mean_depth <- mean(depth)
  sd_depth <- sd(depth)
  mean_depth_nonzero <- mean(depth[depth > 0])
  mean_depth_within2sd <- mean(depth[depth < mean_depth + 2 * sd_depth])
  median <- median(depth)
  presence <- as.logical(depth)
  proportion_of_reference_covered <- mean(presence)
  output_temp <- tibble(bamfile, mean_depth, sd_depth, mean_depth_nonzero, mean_depth_within2sd, median, proportion_of_reference_covered)

  # Bind stats into dataframe and store sample-specific per base depth and presence data
  if (i==1){
    output <- output_temp
    total_depth <- depth
    total_presence <- presence
  } else {
    output <- bind_rows(output, output_temp)
    total_depth <- total_depth + depth
    total_presence <- total_presence + presence
  }
}

output %>%
  mutate(across(where(is.numeric), round, 3))

# Plot the depth distribution (this may take a few minutes to run)
tibble(total_depth = total_depth, position = 1:length(total_depth))  %>%
  ggplot(aes(x = position, y = total_depth)) +
  geom_point(size = 0.1)

# Total depth per site across all individuals 
total_depth_summary <- count(tibble(total_depth = total_depth), total_depth)
total_presence_summary <- count(tibble(total_presence = total_presence), total_presence)
total_depth_summary %>%
  ggplot(aes(x = total_depth, y = n)) +
  geom_point()
total_depth_summary %>%
  ggplot(aes(x = total_depth, y = n)) +
  geom_point() +
  coord_cartesian(xlim=c(NA, 20))
total_presence_summary %>%
  ggplot(aes(x = total_presence, y = n)) +
  geom_col()

```

<br>
You can see that there is a lot of variation among the different depth statistics. Which of them do you think are most relevant to report? Why?

<br>

If you're interested, you can go back and compare the depth distribution to in the raw mapped files.

<br>
<br>

## END OF TUTORIAL 1

### 2022-10-27 fastq2vcf_june2022_update.Rmd
Fixed references in RADseq, updated fastq2vcf to most recent version from SW
changing back to original .Rmd files, will make sure it's ok before I go ahead with doing this

--- 2022-07-25 fastq2vcf_pipeline.Rmd
@@
 ---
-title: 'snpArcher: FASTQ -> VCF workflow for WGS data'
+title: "snpArcher: FASTQ -> VCF workflow for WGS data"
 author: "Sara JS Wuitchik"
@@
 knitr::opts_chunk$set(echo = TRUE)
+library(knitcitations)
 ```

### 2022-10-27 contributors.Rmd
adding contributor page

--- 2022-09-08 index.Rmd
@@
+---
+title: "Contributors"
+---
+
 ```{r setup, include=FALSE}
@@
 ```
----
-title: "Introduction"
----
-
-# Guidelines for 'omics analyses
-
-On this website you will find a growing and evolving collection of guidelines, tutorials, recommended readings, and method evaluations for various 'omics analyses that are commonly used in studies of nonmodel systems (particularly marine species).
-Instead of providing rigid "best practices", we aim to encourage data exploration and rigorous quality control through "best principles", paying particular attention to topics for which open access guidelines are scarce.
-The guidelines are written by members of the [MarineOmics working group](#marineomics) or the community, then shared for peer review.
-Comments on current website content or suggestions for new content can be made through the [Discussion Forum](https://github.com/MarineOmics/marineomics.github.io/discussions).
-
-Based on the expertise of the MarineOmics working group, the first topics include:
-
-### Topics {.tabset .tabset-fade .tabset-pills}
-#### Population Genomics
-
-* [Choosing a population genomics sequencing approach](choosing-population-genetics.html)  
-* [Whole Genome Resequencing (WGS)](WGS_intro.html)  
-  * [Low coverage WGS](https://github.com/therkildsen-lab/genomic-data-analysis/blob/master/lcwgs_data_analysis.md)
-* [Reduced Representation Sequencing](RADseq.html) 
-* [Poolseq](poolseq.html)
-
-#### Functional Genomics
-
-* [Multifactorial RNAseq analyses](DGE_comparison_v2.html)  
-* Align and count RNAseq reads (coming soon!)
-    * De novo, reference-based, and Tag-seq methods
-
-### {-}
-
-### The guidelines portion of the website should be considered a "preprint".
-If you are interested in contributing to the peer review process, check out the [Discussion Forum](discussion.html).
-{.unlisted .unnumbered}
-
-# Panel Seminars 
-To facilitate discussions about practices and pitfalls in applying 'omics to non-model systems, we hosted [Q & A panel seminars](panels.html) with experts on topics ranging from RADseq, population genomics with WGS, and transcriptomics.
-Check them out on [YouTube](https://www.youtube.com/channel/UCosG6JfIj7k0DizfsD5ADfQ)!
-
-# What is the MarineOmics working group? {#marineomics}
-
-The original MarineOmics working group is part of the [RCN for Evolution in Changing Seas](https://rcn-ecs.github.io/), and was formed to promote **reproducible** and **robust** genomic research in nonmodel species, but especially marine and aquatic studies.
-Methods in next generation sequencing and bioinformatics advance at a pace that quickly render any static set of best practices obsolete.
-MarineOmics aims to support robust genomic research in ecology and evolution by disseminating guidelines through this dynamic, open-sourced website addressing introductory and advanced topics of population- and functional genomics.
-*We are always looking for folks to join and write content for the website, if interested please [contact us](#contact)!*
-
-**Rigorous**:  following best principles to identify and mitigate sources of error in data generation and analysis
-
-**Reproducible**: when others can reproduce the results with accessible data, code, and documentation
-
-# Reproducibility in Marine Genomics  
-MarineOmics is also interested in evaluating and improving reproducibility in studies that use genomics to study adaptation in marine systems.
-We are conducting a meta-analysis of reproducibility across genomic studies of environmental adaptation in marine populations with the aim of identifying obstacles and opportunities for reproducing and synthesizing population genomic research.
-In Spring 2021, three undergraduate research interns contributed to this literature survey, with funding from the [RCN-ECS](https://rcn-ecs.github.io/).
 
@@
 #### [Steven Roberts, PhD](http://faculty.washington.edu/sr320/?page_id=10850), University of Washington {.unlisted .unnumbered}
-
-# Internship 
-In Spring 2021, MarineOmics hosted three virtual undergraduate research interns with funding from the [RCN-ECS](https://rcn-ecs.github.io/): Jecy Klinkam, Kaitlin Macaranas, and Chris Mantegna.
-These interns were paired with graduate student mentors within the MarineOmics group, as well as participated in training activities while contributing to the Reproducibility literature review.
-
-# Contact {#contact}
-For questions about the website and its materials, post to the [Discussion Forum](https://github.com/MarineOmics/marineomics.github.io/discussions)!
-For inquiries about the MarineOmics Working Group activities, please contact us at rcnecs.marineomics@gmail.com.

### 2022-10-27 index.Rmd
adding 'Genome-Phenome' place holder tab
adding contributor page

--- 2022-07-08 index.Rmd
@@
 
+#### Genome-Phenome
+
+* coming soon!
+
 ### {-}
@@
 
-# Original Working Group Members
-
-#### [Katherine Silliman, PhD](https://ksilliman.weebly.com/), South Carolina Dept. of Natural Resources - sillimank[at]dnr.sc.gov {.unlisted .unnumbered}
-#### [Sam Bogan](http://www.hofmannlab.com/sam-bogan.html), UC-Santa Barbara -snbogan[at]ucsb.edu {.unlisted .unnumbered}
-#### [Alan Downey-Wall](https://cos.northeastern.edu/people/alan-downey-wall/), Northeastern University {.unlisted .unnumbered}
-#### Chris Mantegna, University of Washington (Intern)  {.unlisted .unnumbered}
-#### [Danielle Davenport](https://www.researchgate.net/profile/Danielle-Davenport-2), University of Queensland {.unlisted .unnumbered}  
-#### [Hanny Rivera, PhD](https://www2.whoi.edu/staff/hrivera/), Woods Hole Oceanographic Institute  {.unlisted .unnumbered}
-#### Jecy Klinkam, College of William and Mary (Intern) {.unlisted .unnumbered}
-#### [Joanna Griffiths, PhD](https://joannasgriffiths.wordpress.com/), UC-Davis {.unlisted .unnumbered}
-#### [Joe McGirr, PhD](https://joemcgirr.github.io/), UC-Davis {.unlisted .unnumbered}
-#### Kaitlin Macaranas, CSU-Bakersfield (Intern) {.unlisted .unnumbered}
-#### Megan Guidry, University of Rhode Island {.unlisted .unnumbered}
-#### Natalie Ameral, University of Rhode Island & RI Dept. of Environmental Management {.unlisted .unnumbered}  
-#### Nicolas Lou, Cornell University {.unlisted .unnumbered}
-#### [Sara Schaal](https://schaals.wixsite.com/smschaal), Northeastern University {.unlisted .unnumbered} 
-#### [Sara Wuitchik, PhD](https://sjswuitchik.weebly.com/), Harvard University {.unlisted .unnumbered}
-
-### Faculty Advisors {.unlisted .unnumbered}
-#### [Andrew Whitehead, PhD](https://whiteheadresearch.wordpress.com/), UC-Davis {.unlisted .unnumbered}
-#### [Jon Puritz, PhD](https://www.marineevoeco.com/), University of Rhode Island {.unlisted .unnumbered}
-#### [Steven Roberts, PhD](http://faculty.washington.edu/sr320/?page_id=10850), University of Washington {.unlisted .unnumbered}
-
 # Internship 

### 2022-11-01 submissions_instructions.Rmd
new submissions instructions template (currently just the 'choosing a pop gen approach' and the RADseq tutorial)

--- 2022-08-17 RADseq.Rmd
@@
 ```
+
 ---
-title: 'RADseq'
-subtitle: '*Katherine Silliman, Danielle Davenport*'
+title: 'Instructions for submitting new pages and tutorials to the MarineOmics webpage'
 bibliography: RCN_site-Nov05.bib
 ---
+
+*Choosing a Population Genomics Approach code*
+
+# Population Genomic Sequencing Approaches
+Deciding on a sequencing approach for a new project can be challenging.
+Your choice depends primarily on the questions of interest, but also constraints set by funding, existing genomic resources, sample size, genome characteristics and expertise in the research group.
+
+The main categories of population genomic sequencing approaches include...
+
+## 1) Whole-genome resequencing
+Two approaches for whole-genome resequencing are distinguished by their coverage depth.
+**Coverage depth** refers to the average number of sequencing reads per region of the genome (distinct from coverage breadth, see [@sims2014sequencing]).
+With more sequencing reads at a region (higher coverage), errors that arise during sequencing have a smaller impact on accurately identifying variable sites (termed “genotyping”) [@Fuentes-Pardo2017-nu].
+
+### High coverage individual whole-genome resequencing (hcWGS)  {.unlisted .unnumbered}
+**Whole-genome resequencing (WGS)** is where whole genomes of multiple individuals are sequenced and mapped to an assembled reference genome in order to compare variable sites between individuals or populations.
+High-coverage (>20x) sequencing is able to find rare variants with high confidence.
+This approach is considered the 'gold standard' for DNA re-sequencing because it includes all variant types (i.e SNVs, indels, structural variants and CNVs) [@sims2014sequencing].
+
+### Low coverage individual WGS (lcWGS)  {.unlisted .unnumbered}
+**Low coverage whole genome sequencing (lcWGS)** is where whole genomes of many individuals are sequenced, but depth of coverage is so low (<5× per site per individual, on average [@nielsen2011genotype]) that individual genotypes cannot be confidently assigned.
+Instead of assigning specific (“hard called") genotypes, lcWGS data requires probabilistic analyses that account for uncertainty about the true genotype and can incorporate uncertainty across all sequenced samples [@korneliussen2014angsd].
+This approach is useful in cases where the research questions of interest are targeted at the population-level (i.e.
+allele frequencies, linkage disequilibrium (LD) patterns, etc) [@Lou2021-me].
+
+## 2) Reduced-representation sequencing
+As the name suggests, this approach aims to sequence a reduced portion of the genome across multiple individuals at moderate to high coverage for variant discovery and genotyping with high confidence.
+The genome fragments can be selected at random, or through targeted probe assays.
+Variations of this type of sequencing approach include...
+
+### RADseq/Genotype-by-sequencing  {.unlisted .unnumbered}
+A class of methods involving sequencing a subset of genomic regions randomly distributed throughout the genome, often using restriction enzymes (i.e. **restriction site-associated DNA (RADseq)**).
+
+### Targeted capture methods  {.unlisted .unnumbered}
+Sequence capture methods use a set of probes designed with a-priori knowledge to focus sequencing effort on a set of hundreds to tens of thousands of specific loci.
+Several capture approaches for non-model species can utilise other sequencing strategies to design probes corresponding to genomic regions (i.e.
+**Whole-exome sequencing (WES)**, see [@jones2016targeted] for a review, **Expressed Exome Capture Sequencing (eecSeq)**[@puritz2018expressed]).
+
+## 3) Pooled individuals (Poolseq)
+Sequencing pools of individuals to provide information on genome-wide population allele frequencies (see review in [@futschik2010next]).
+
+![**Figure 1** from [@Lou2021-me] comparing the distribution of sequencing reads mapped to a reference genome.](images/mec16077-fig-0001-m.jpg)
+
+It should be noted that you don’t need to choose only one approach for all samples!
+If you are starting a research program on a species without any genomic resources, you may want to start with denovo RADseq to understand basic population structure and genetic diversity, then once a reference genome is assembled move to high/low coverage WGS.In the [MarineOmics seminar](https://marineomics.github.io/panels.html), [Misha Matz](https://matzlab.weebly.com/) recommended choosing one individual to use for genome assembly, doing high coverage WGS on a few other individuals that represent most of the variation in your taxa, and then WGS many other individuals at low coverage.
+The high coverage individuals can be useful for imputing missing genotypes in the low coverage samples.
+
+The table below briefly summarizes the pros and cons of five sequencing approaches and their appropriateness for answering specific questions.
+If you are interested in answering multiple questions (i.e.
+neutral population structure and adaptive variation) then you would often want a method that can do both (i.e.
+WGS).
+
+```{r,echo=FALSE}
+text_tbl <- data.frame(
+  Goals = c("pros","cons","population structure, genetic diversity","demography (mig. rates, pop. size through time)",
+            "signatures of selection, GWAS","phylogenetic inference","genetic crosses/mapping panels",
+            "molecular evolution (accurate low-freq alleles required)"),
+  hcWGS = c("provides high quality, high density genotypes; can be used to improve development of a reference genome","most expensive per sample, requires a reference genome assembly, large computational resources","great, but cheaper methods work almost as well","great if incorporating haplotype-based methods ","great","computationally challenging, but good at all divergence levels ","expensive","best"),
+  lcWGS = c("provides high density genotypes at a reduced cost","requires a reference, can produce false (+) heterozygotes, sensitive to batch effects, may be inappropriate to use individual SNP calls for some analyses","good, but limited to popgen analyses based on genotype likelihoods ","good, esp.
+with some high coverage samples to help with imputation","good","good for tree shapes but not branch lengths, good for extracting organelle sequences and cost-effective primers for species and hybrid ID studies ","great","poor"),
+ RADseq = c("generally cheaper per sample, density of genotypes can be tuned to fit question, does not require a reference","only 1-5% of genome covered which limits studies on adaptive variation, de novo assembly can result in paralogs without quality control, can be hard to capture the same loci between batches","great, esp.
+for many samples","good for methods based on site frequency spectrum (eg moments), but not ideal for methods using extended haplotype or phasing info","hotly debated, but unless your genome is way too big, WGS is the way to go","good at shallow to medium divergence levels, but need to play with filtering parameters ", "great as it allows for many invididuals","poor" ),
+ Poolseq = c("many individuals can be mixed for a low library prep cost, only option for some larval studies","need higher sequencing depth so it can be more expensive than RAD (but less than other options), no info on individual genotypes (duh), requires a reference, requires some replicates to account for batch effects","good for methods that only use allele frequencies, won't work for individual based methods (eg Admixture)","active area of method development, but still not common except in cases of multiple temporal samples [@nielsen2020bc] ","good if covering majority of the genome","would work in certain situations","maybe","poor"),
+ Target.Capture = c("allows consistent capture of loci between batches, good for sequencing targeted regions across many samples","expensive to start a new project and requires reference of some kind to design baits (unless using method like eecSeq), only gives info on targeted region so can result in ascertainment bias for some analyses","ascertainment bias is likely to skew results, esp.
+for measures of genetic diversity","likely not appropriate due to ascertainment bias","only works for targeted regions ","an improvement over RAD as there will be less missing data among samples","poor","good if only interested in targeted regions")
+  )
+
+kbl(text_tbl) %>%
+ # kable_paper(full_width = F) %>%
+  column_spec(1, bold = T, border_right = T) %>%
+  kable_styling(bootstrap_options = "striped", font_size = 13)
+
+```
+
+# Recommended readings  
+Below we list a number of great review articles and empirical comparison studies that can help when making the decision.
+This topic was also discussed during the [MarineOmics seminars](https://marineomics.github.io/panels.html).
+
+Review articles:
+
+* [@Matz2018-vl]: great intro to ecological genomics methods in species with few to no prior genomic resources, accessible to those just getting started.  
+* [@Fuentes-Pardo2017-nu]: a VERY thorough review of the different approaches to generating population genomics data, with non-model species in mind.
+Some of the content about relative cost and available software is a little dated, but otherwise a very valuable reference.
+* [@Lou2021-me]: great recent primer on low coverage WGS and how it compares to RADseq and Poolseq.
+
+Empirical comparisons:
+
+* [@Benjelloun2019-dq]: an empirical comparison of high coverage WGS, low coverage WGS, random variants (eg.
+from RADseq), exome capture, and commercial SNP chip for mammals (genome size=2.6Gb).
+For neutral genomic diversity, found 5k-10k random variants were enough to accurately estimate, but commercial panels/exome capture had ascertainment bias.
+To detect selection and accurate estimates of LD, at least 1M variants were required.
+For the studied species, 5x coverage was just as accurate as 12x coverage.
+Low coverage WGS (< 5x) should use genotype likelihood methods, however, the rate of false positives for heterozygotes is higher than classic genotyping methods at higher coverage.
+* [@Dorant2019-fb]: a case study comparing Poolseq vs.
+reduced representation seq (GBS) vs Rapture (targeted capture of RAD loci) for inferring population structure with weak genetic differentiation.
+All three detected the same weak pop structure pattern, but Poolseq gave Fst values 3-5x higher than Rapture/GBS.
+
+# MarineOmics Guidelines
+Presently, our site covers four approaches:
+
+* [RADseq](RADseq.html)  
+* [hcWGS](WGS_intro.html)
+* [lcWGS, maintained by MarineOmics member Nicolas Lou](https://github.com/therkildsen-lab/genomic-data-analysis/blob/master/lcwgs_data_analysis.md)
+* [Poolseq](poolseq.html)
+
+For each approach, we provide guiding principles and tutorials for bioinformatic processing from raw data to genotypes.
+While we currently don't discuss library preparation methods in depth, we do include advice on how to organize your sequencing runs in order to improve quality control downstream.
+
+# References
+
+## *RADseq tutorial code*
 
@@
 To plot the missingness on the command line, you can use gnuplot:
-```{bash, eval = T}
-#code from Jon Puritz
-mawk '!/IN/' RAD_data/OL_subset.imiss | cut -f5 > totalmissing
-gnuplot << \EOF
-set terminal dumb size 120, 30
-set autoscale
-unset label
-set title "Histogram of % missing data per individual"
-set ylabel "Number of Occurrences"
-set xlabel "% of missing data"
-#set yr [0:100000]
-binwidth=0.01
-bin(x,width)=width*floor(x/width) + binwidth/2.0
-plot 'totalmissing' using (bin($1,binwidth)):(1.0) smooth freq with boxes
-pause -1
-EOF
-```
+<!-- ```{bash, eval = T} -->
+<!-- #code from Jon Puritz -->
+<!-- mawk '!/IN/' RAD_data/OL_subset.imiss | cut -f5 > totalmissing -->
+<!-- gnuplot << \EOF -->
+<!-- set terminal dumb size 120, 30 -->
+<!-- set autoscale -->
+<!-- unset label -->
+<!-- set title "Histogram of % missing data per individual" -->
+<!-- set ylabel "Number of Occurrences" -->
+<!-- set xlabel "% of missing data" -->
+<!-- #set yr [0:100000] -->
+<!-- binwidth=0.01 -->
+<!-- bin(x,width)=width*floor(x/width) + binwidth/2.0 -->
+<!-- plot 'totalmissing' using (bin($1,binwidth)):(1.0) smooth freq with boxes -->
+<!-- pause -1 -->
+<!-- EOF -->
+<!-- ``` -->
 

### 2022-11-04 DGE_comparison_v2.Rmd
added cache = TRUE to some of the tutorials that take a long time to render. doing this to try to make it easier & quicker to re-render the website and update the nav bar

@@
 #knitr::opts_knit$set(root.dir = '~/Documents/GitHub/DGE_comparison/')
+knitr::opts_chunk$set(echo = TRUE)
 

### 2022-11-04 WGS_intro.Rmd
testing an attempt to consolidate all materials to this website rather than linking to another git repo, starting with the first tutorial in lcWGS
added cache = TRUE to some of the tutorials that take a long time to render. doing this to try to make it easier & quicker to re-render the website and update the nav bar

@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
 library(knitcitations)

### 2022-11-06 submissions_instructions.Rmd
setting the outline of the sumbissions_instructions page and developing the first part of my material

--- 2022-06-24 RADseq.Rmd
@@
-```{r,message = FALSE,echo = FALSE}
+---
+title: 'Instructions for developing new pages to the MarineOmics webpage'
+bibliography: RCN_site-Nov05.bib
+---
+
+```{r setup, message = FALSE, echo = FALSE}
 library(knitr)
-knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
 library(knitcitations)
@@
 ```
----
-title: 'RADseq'
-subtitle: '*Katherine Silliman, Danielle Davenport*'
-bibliography: RCN_site-Nov05.bib
----
 
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+## Overview
 
-colorize <- function(x, color) {
-  if (knitr::is_latex_output()) {
-    sprintf("\\textcolor{%s}{%s}", color, x)
-  } else if (knitr::is_html_output()) {
-    sprintf("<span style='color: %s;'>%s</span>", color,
-      x)
-  } else x
-}
-#set bash code chunks to use bash_profile
-knitr::opts_chunk$set(engine.opts = list(bash = "-l"))
+This is a guide for creating a new page for the website.
+Of course, the website managers will work with anyone who develops new content to answer questions and publish the site, but this can be used as a resource as you develop your page.
+In this guide are general instructions as well as examples of how to build specific features.
+If there is anything else you would like to see on this guide please contact Jason Johns (jasonjohns[at]ucsb.edu).
+Of course, do not hesitate to reach out with any other questions or comments as well.
 
-```
+\*\* Before publishing any content, please contact Jason Johns (jasonjohns[at]ucsb.edu) so we can have working group contributors review the page.
 
-# Setup for running code
+This website is published through Github Pages directly from the [MarineOmics github repository](https://github.com/MarineOmics/marineomics.github.io).
+In short, each page of the website is rendered as an .html from an RMarkdown (.Rmd) document.
+The two most important files for publishing the website are the .site_yml ("the yaml file") and the index.Rmd/index.html.
 
-If you would like to run the R code examples that are scattered throughout the guide (recommended but not required!), you will need to install some R packages. Only need to run this code once:  
-```{r, eval=F}
-install.packages("tidyverse")
-if (!requireNamespace("BiocManager", quietly = TRUE))
-    install.packages("BiocManager")
+**.site_yml**: the 'under the hood' backbone of the website, where things like the visual appearance of the website the navigation bar are built
 
-BiocManager::install("SeqArray")
-BiocManager::install("SNPRelate")
-```
+**index.html**: rendered from index.Rmd - the home page from which all other pages branch
 
-Now load those packages, if using:  
-```{r}
-library(SeqArray) # efficient storage and filtering of genomic data
-library(tidyverse) # plotting data formatting and manipulation
-library(SNPRelate) # PCA and other popgen analyses
-```
+\* The website manager will help modify the index.html and .site_yml files for your page when the time comes, so you don't need to know exactly how they work, but it's good to know what they are at a basic level.
 
-We also provide optional alternative coding examples that are based only on the command line, requiring these software packages:
+# R Markdown basics
+The rendered .Rmd and .html files for each page on this site can be seen in the [MarineOmics github repository](https://github.com/MarineOmics/marineomics.github.io).
+If you don't have much experience with RStudio, it's quite intuitive with a shallow learning curve and many existing online resources for using it.
 
-* [vcftools](https://vcftools.github.io/man_latest.html)
-* [gnuplot](https://riptutorial.com/gnuplot/example/11275/installation-or-setup)
+*Note: it's a good idea to look at existing .Rmd scripts to see examples of how a particular feature can be accomplished, such as citing references, linking a website, building a table, incorporating figures, etc.
+We will explicitly go over some of those feautures below, but there are many other bells and whistles that can be explored through other resources online.*
 
-# Reduced Representation Sequencing (RADseq/GBS)
+## Initial Page Setup
+- caching
+- installing & loading packages
+- etc.
 
-"Restriction-site   Associated   DNA   sequencing"  - RADseq - combines  restriction  enzymes  with next-gen, massively parallel,  short-read  sequencing.
-RADseq involves the use of restriction enzymes, which are used to shear DNA at restriction enzyme cutsites.
-RADseq comes in different flavors.
-Double-digest RADseq (ddRAD; [@Peterson2012-tr])  selects  markers  with  two  restriction enzymes  with  different  cut  frequencies.
-Genotype-by-Sequencing  (GBS;  [@Elshire2011-or])  uses  a  frequent-cutting  restriction  enzyme  with PCR size selection.
-There are a number of reviews comparing different RADseq and GBS methods (e.g., [@Andrews2016-ay]).
-In this guide we use the term "RADseq" to refer to any of these protocols, including those that don't involve random shearing of data.
-When recommendations are specific to a certain type of RAD/GBS, we will explicitly say so.
+## Formatting
+- text vs. code chunks, how to include code on the page or not, include output or not, etc.
+- navigation bar on the left, font size, bold, italics, etc
+- in-text citations, hyperlinks
 
-<!--Some aspects of RADseq data that are unique compared to whole genome resequencing (more):-->
+## Making a table - Katherine/Sam
 
-Using RADseq to generate single nucleotide polymorphisms (SNPs) involves:
+## Including a figure - Katherine/Sam
 
-1. library preparation in the lab
-2. bioinformatic processing through assembly and/or mapping to a reference, then 
-3. filtering of SNPs and individuals for quality.
+## Including an image - Katherine/Sam
 
-All of these steps can (and will) introduce some error, so the goal is to minimize this error through mitigation steps at all three parts of the process. Every dataset is different!  
-<!--add more about philosophy over best practices-->
+## Running code in another language such as bash - Katherine/Sam
 
-Much of this guide is directly inspired by [this excellent review paper](https://onlinelibrary.wiley.com/doi/10.1111/mec.14792) [@OLeary2018-iy], especially the section on minimizing errors due to library prep.
-We recommend reading and cross-referencing with this paper, and citing it if you follow its suggestions.
-Table 1 from this paper summarizes the various potential issues that can arise from RAD datasets and some mitigation steps.
-The goal of this guide is to expand on the O'Leary paper and provide some example code to help implement quality control and mitigation steps.
-
-![Table 1 from [@OLeary2018-iy].](images/oleary_table1.png)
-
-# Considerations During Lab Work {#labwork}
-
-There are steps you can take before you even sequence RAD libraries that can help minimize issues downstream.
-Here, we use "library" to refer to a set of RADseq fragments from a group of individuals that are barcoded and sequenced together on a single lane or group of lanes.
-While specific RAD library prep methods have their own nuances for minimizing error, there are some steps you can take that are common across methods.
-
-* If this RAD/GBS method has not been done in your species or in your molecular lab setup before, spend some time optimizing the protocol using a representative subset of individuals.
-Then try to keep everything about the library prep as consistent as possible across samples (eg, DNA extraction kit, PCR cycles, sequencing platform).
-This isn't always possible, especially if you need to optimize the protocol for certain tricky samples.
-Just make sure to keep track of everything!
-
-* Randomize samples across library prep batches and sequencing lanes!
-For example, if you are sequencing two different groups of samples on two different sequencing lanes, make sure they are randomized with respect to sample location or whatever your groups of interest are.
-If you are preparing groups of individuals in different batches to be pooled later, randomize across these batches.
-[Alicia Mastretta-Yanes](https://mastrettayanes-lab.org/) even recommends randomizing across DNA extractions, especially if the person doing the extractions is new to molecular work.
-This is to allow you to control for potential [batch effects](#batch) that are often observed with RAD data.
-
-<!--ADW: add figure showing random plate design-->
-```{r, eval = F, include=FALSE}
-# This code reads in a comma-delimited STRATA/meta-data file, randomises samples, and returns a file with wells and plates and duplicates which may be used to assist with plating libraries
-# ** TODO** will need to consider where to duplicate samples - need to consider STARTA, DOC, ect..?
-# maybe have user write in a list of samples to replicate
-
-wells = 96 # how many wells can you use on your plate?
-data.in = read.table("./data/example.metadata.csv", sep = ",", header = T) # each sample should be in its own ROW
-l = length(data.in[,1])
-plates = rep(1:length(ceiling(101/ wells)), wells)
-data.out = cbind(
-   data.in[sample(1:l, l, replace = F),], # randomise samples
-   PLATEID = plates[c(1:l)],  # add plate info
-   WELLID = rep(1:wells, length(unique(plates)))[1:l] # add well info
-) # add duplicates **TODO**
-
-# print example
- print(head(data.out))
-
-#save to file
-
-```
-* Keep track of all potential batch effect sources in a Metadata file (eg, storage conditions of tissue/DNA, date/method of DNA extraction, date/method library prep batch, sequencing lane).   
-* Have a core set of 2-4 technical sample replicates across all libraries.
-Ideally these are true technical replicates, meaning the same tissue/DNA sample is processed multiple times all the way from library preparation to sequencing.
-Sequencing replicates, where you sequence the same sample library preparation multiple times, can also be useful for downstream quality control.
-
-<!--ADW: add figure or text to define sequencing replicates vs technical replicates-->
-
-# Principles for Analyzing Your Data
-
-## Steps for a robust RAD analysis 
-This is just one approach for working through your data.
-Some people will prefer to run just a subset of samples through a pipeline at first and evaluate parameters, then run all the samples through.
-Either way, be prepared to make MULTIPLE assemblies and go through this process iteratively, especially if this is a new-to-you study system.
-
-<!-- 1) [Look at your raw data.](#fastqc)
-2) [Run an assembly pipeline](#assembly) through to a SNP dataset, using either all samples or a representative subset. Parameters can come from those used in a similar study, or default parameters.
-3) Filter your data minimally, then [evaluate for potential sources of error.](#error)
-4) Subset or remove individuals based on initial evaluation.  
-5) Using a representative subset of samples, [test key parameters](#test) to optimize your assembly.
-6) Run your optimized assembly on all non-removed samples.
-7) Evaluate the difference between multiple filtering schemes for your analyses of interest. *Popgen analysis guide coming soon*
-8) Repeat as needed. -->
-
-1) [Look at your raw data.](#fastqc)
-2) [Run an assembly pipeline](#assembly), using either all samples or a representative subset.
-3) [Evaluate potential sources of error.](#error)
-4) [Remove individuals based on initial evaluation](#bad).  
-5) [Test a range of key parameters](#test) to optimize your assembly.
-6) Run your optimized assembly on all non-removed samples.
-7) [Filter your SNPs](#filter)
-8) [Repeat as needed!](#repeat)
-
-## First, look at the raw data! {#fastqc}  
-Always look at your data with [FastQC](https://www.bioinformatics.babraham.ac.uk/projects/fastqc/) before starting an assembly.
-First, this is a good check to just make sure the sequencing worked.
-If you have demultiplexed data, you can use [MultiQC](https://multiqc.info/) to generate FastQC plots for all individuals and quickly identify ones that did not sequence well.
-
-Check out [this informative tutorial](https://datacarpentry.org/wrangling-genomics/02-quality-control/index.html) on running FastQC and interpreting the output.
-
-**Questions to ask:** Do you have a lot of adapter sequences?
-Are the ends of your reads very low quality?
-If so, you should expect a fair amount of trimming and read filtering to occur prior to assembly.
-If that doesn't occur or if too many reads are being filtered so as to only recover a small number of SNPs, something might need to be tweaked with your trimming and filtering step.
-<!--ADW: elaborating what you mean by 'tweaked' or possibly providing a link to some resources for how someone might thoughtfully change the parameters for their trimming and filtering steps.-->
-
-You should also look at the top few reads in the Terminal.  
-```
-## zcat: unZip and conCATenate the file to the screen
-## head -n 20: Just take the first 8 lines of input
-
-$ zcat raw-fastqs/BC2_10_C5.fastq.gz | head -n 20
-@BC2_10_C5.1 1 length=96
-CAGCGTGTAATAGTCACCGGCGGCTCCCTCTGGAGAATAGCACAAGTGATCATTTTGCTCATCTTCCGTCCACTGGTGATTGTGGACCAGCCTCAC
-+BC2_10_C5.1 1 length=96
-<GGGGGA<GGGGIIIGIGGIGGGIIIIIGGGGGGGGGGGGGGGGGIIIIIIIIGGIIIGGGGIGIIGIGGGIIIIIAGGGGIIIGIIGGGGGAAGG
-@BC2_10_C5.2 2 length=96
-CTGCTACATGCAGTGTTCTGTATTACTTTTATTGTACGTTGATATGAATGAATGAGTGTTTTGTATACTTAGAGTACAAGTTTGTCAGTCATATCG
-+BC2_10_C5.2 2 length=96
-GIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIGGIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIG
-@BC2_10_C5.3 3 length=96
-CAGCACATGTTCCTGTGTAGAAAGCTTGTTAGTAGAATAAATAACACATGGCTGGTCAAACACAACACATGAAGAAACAACTTTCTGAACAGTTTT
-+BC2_10_C5.3 3 length=96
-GIIGIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIIGIIIIIIIIIIIIIIIIIIIIIIIIGGGGIIGIIIGGGII
-@BC2_10_C5.4 4 length=96
-CAGCGATTCGGCCCAAATTTGCACCACATCAGGCCCTTGACAGGGCGCTTCGATGGTGCAAATTTGGTGCGATTCGCTGCGCACCTAGCATATATG
-+BC2_10_C5.4 4 length=96
-AGGGIGIIIGIGGGIIGIGGGIIIGGGGAGGGGGAGGIGIGGIGGGGAGGGGGGGGGGGGGGGIIIII.G<GGIGGGIIIGGGGGIGGIIIGAGGI
-@BC2_10_C5.5 5 length=96
-CAGCAGTTTGGTGGAGTTCTGCAACCTTCCATTTCCAAAGAATTACCCAGGAGCTCTTCCCAGTGAATTTCTTCGGCACTTTTCATTGACCTTTTA
-+BC2_10_C5.5 5 length=96
-GGAGAAA.<AGAA.G.GGAGA<.GGA.<GAAGGAAGGIA<...<GA<..<G.<.<<.<AAAGGGG..<GGGGG<G.A.<GGGII.AG..<.GGGGG
-
-```
-
-If the sequencing center gave you one fastq file with all your samples, you should expect to see a barcode sequence, followed by the cutsite at the start of the read.
-If the Data are already demultiplexed (as the example is), you should only see the cutsite overhang (in this case, CWGC).
-Sometimes you can look at your fastq data files and see that there was a problem with the sequencing such that the cut site is either offset by one or more bases, or contains many errors.
-If this is not being addressed by the default filtering steps in your assembly pipeline, you can trim off N bases from the beginning or end of R1 and R2 reads in [ipyrad](https://ipyrad.readthedocs.io/en/latest/index.html) (`trim_reads` param), or with cutadapt before using stacks, or customize the Trimmoatic step for [dDocent](https://www.ddocent.com/).
-
-## Run an assembly pipeline {#assembly}
-
-There are a number of freely available pipelines for processing RADseq data, with the most popular being [dDocent](https://www.ddocent.com/), [Stacks](https://catchenlab.life.illinois.edu/stacks/), and [ipyrad](https://ipyrad.readthedocs.io/en/latest/index.html).
-These pipelines vary slightly in their underlying methodologies, customization options, and additional included analyses.
-
-|  | [dDocent](https://www.ddocent.com/) | [ipyrad](https://ipyrad.readthedocs.io/en/latest/index.html)| [Stacks2](https://catchenlab.life.illinois.edu/stacks/) |
-|---|---|---|---|
-| Supported datatypes | Paired-end: ddRAD, ezRAD, RAD (random shearing), data with large overlap between forward and reverse reads; Single-end: any RADseq method  de novo and reference-based.
-If doing de novo assembly, reads cannot be trimmed outside of dDocent | Wide range of RAD/GBS methods with paired/single end data, see [ipyrad documentation](https://ipyrad.readthedocs.io/en/master/4-data.html) | Paired-end and single-end data for single and double digest RAD and DART; GBS with single-end sequencing.
-NOT suitable for paired-end GBS.
-|
-| Unique aspects | Novel data reduction approach used to inform coverage cutoffs | Python API for popgene and phylogenetic analyses written specifically for RAD data |  |
-| Documentation | Very good (esp.
-tutorials), active community support on Google Groups | Excellent (esp.
-installation, parameter explanations, and tutorials), active community support on Gitter | Very good, active community support on Google Groups |
-| Speed/Accuracy | fastest and most accurate | close to dDocent accuracy, more over-splitting is possible  if parameters are not tuned | produces some untrue genome fragments (esp.
-with higher levels of indel polymorphism), but can be addressed with downstream filtering |
-| Open source/development | Open source | Open source, primarily combines existing software | No |
-| Filtering options | Minimal default filtering as implemented in VCFtools.
-Ideal for those who want full freedom in filtering their SNPs.
-| Wide range of filtering options  | Moderate range of filtering options |
-| Output options | Only produces a VCF, fasta of the de novo assembly, and individual BAM files  from mapping to the reference  | Lots of output formats (see [documentation](https://ipyrad.readthedocs.io/en/master/output_formats.html)) | Lots of Stacks-specific output and log files, as well as various inputs for popgen programs (see [documentation](https://catchenlab.life.illinois.edu/stacks/comp/populations.php)) |
-| Popgen analyses | No additional popgen analyses included  | Unique, flexible Python API implementing range of popgen and phylogenetic methods (see [documentation](https://ipyrad.readthedocs.io/en/master/API-analysis/index.html))   | Popgen summary statistics (F-statistics, pi), including kernel-smoothing along reference genome.
-Underlying equations and assumptions for these are not very clear.
-|
-|  |  |  |  |
-
-Some groups have also developed pipelines for specific flavors of RAD (e.g., [Matz lab for 2b-RAD](https://github.com/z0on/2bRAD_denovo/blob/master/2bRAD_README.sh)) or proprietary software (eg, [DARTseq](http://georges.biomatix.org/storage/app/media/uploaded-files/dartR_Workbook.pdf)), which we touch on below.
-**In most cases, it is recommended that you use a pipeline developed and tested for RAD data, especially if you are making a de novo assembly.** [@LaCava2020-bg] have an excellent study where they review various *de novo* assemblers that are used in these pipelines.
-Assemblers that were not explicitly developed for short reads (eg, Velvet, ABySS) performed very poorly, while CD-HIT (the assembler in dDocent) performed the best.
-Stacks/Stacks2 and vsearch (the assembler in ipyrad) both performed worse when analyzing simulated data with indels.
-It should be noted that all of these RAD pipelines have optimized the specific parameters of these de novo assemblers to work with RAD data of various flavors, which may not be fully reflected in the LaCava study.
-Based on the MarineOmics RAD panel, the general consensus is that all the popular pipelines can produce adequately accurate datasets (with appropriate parameter optimization and data filtering).
-![Table 3 from LaCava et al 2019. CD-HIT is used in dDocent, VSEARCH is used in ipyrad.](images/lacavaTable3.png)
-
-## Evaluate potential sources of error {#error}
-
-Once you have processed your samples (or a subset of samples) through a genotyping pipeline, you will have a bunch of different output file options for genetic data.
-The VCF file is one of the most popular file formats, and is the most versatile for initial data exploration as many programs exist to filter and accept VCF files.
-Below are a list of potential confounding factors that may exist in your data, and how to tease them out.
-
-**Using SeqArray in R** 
-If you are following along with the R code examples, we need to 1st read in our data (skip this section if not running R code).
-
-For the code examples in this section, we primarily make use of the R package SeqArray [@Zheng2017-pl], which can read and manipulate VCFs.
-If you’re familiar with R, SeqArray is simple to use.
-The package can efficiently store sequence variant calls with hundreds of thousands of individuals, variants (e.g., SNPs, indels and structural variation calls) and annotations.
-It employs a data container called a CoreArray Genomic Data Structure (GDS).
-It’s super-fast (5X faster than PLINK v1.9; 16X faster than vcftools) and it integrates well with other R packages you might use in your analysis pipeline.
-(i.e.
-SNPRelate, SeqVarTools).
-We also like it because you can filter your data before running certain analyses without 1st generating a separate filtered VCF file.
-
-First, we need to convert our VCF file into the GDS format. We will do this once here, then use the GDS file for subsequent code examples.
-```{r}
-filename = "OL_subset" #replace with your file name 
-filename.gds = paste0("RAD_data/", paste0(filename, ".gds"))
-filename.vcf = paste0("RAD_data/", paste0(filename, ".vcf"))
-  # 1 . Convert VCF to GDS
-SeqArray::seqVCF2GDS(vcf.fn = filename.vcf, out.fn = filename.gds, storage.option="ZIP_RA")
-gdsin = SeqArray::seqOpen(filename.gds)
-print(paste0("The number of SAMPLES in data: ", length(c(SeqArray::seqGetData(gdsin, "sample.id")))))
-print(paste0("The number of SNPs in data: ",  length(c(SeqArray::seqGetData(gdsin, "variant.id")))))
-```
-It is always helpful to have a metadata file with information for each sample, such as sampling site, sequencing library, etc.
-In our example, our metadata file (OL.popmap) is tab-delimited and has the column headers:
-ID: sample ID
-STRATA: sampling location/population
-PLATE: sequencing batch  
-Next we read in our metadata file, and make sure the samples are in the same order as your VCF file:  
-```{r}
-metafile = "RAD_data/OL.popmap"
-sample.ids = seqGetData(gdsin, "sample.id")
-sample.strata =  read.table(metafile, header = T, sep = "\t") %>%
-                  dplyr::select(ID, STRATA, PLATE)
-```
-Now, on to evaluating our data!
-
-### "Bad" samples {#bad}
-
-Sometimes a sample doesn't sequence well (few sequencing reads, higher error rate).
-This can be due to DNA quality, an issue during library prep, or not enough sequencing depth (average # of reads per sample).
-Generally, it will lead to a sample with fewer sequencing reads, higher missing data in a SNP dataset, and fewer shared loci with other samples.
-Identifying and then removing these samples *prior* to the final RADseq assembly analysis can help minimize mis-assembled loci, genotyping errors, and excessive filtering of acceptable loci.
-
-The process of identifying low quality individuals is usually iterative, as the way you initially filter your SNPs will influence the amount of missing data and locus sharing among samples.
-This is why we recommend minimally filtering your SNPs for sample coverage (the # of individuals a locus is called in) when initially exploring your data.
-Some ways to identify bad samples:
-
-1) For every SNP dataset you generate, it is a good idea is always evaluate the missingness per sample (and report this distribution in your manuscript!).
-Identify samples with way more missingness than the rest, and observe how they look in a PCA and locus sharing plot.
-If they stick out or all cluster together in the middle, then try removing them from the assembly and seeing if it changes downstream analyses.
-If so, you may want to specify a missingness cutoff for including samples in the final analysis.
-
-**Missingness in R with SeqArray**
-```{r}
-#using previously loaded gdsin object
-print("Per variant: ")
-summary(m1 <- SeqArray::seqMissing(gdsin, per.variant=TRUE))
-print("Per sample: ")
-summary(m2 <- SeqArray::seqMissing(gdsin, per.variant=FALSE))
-samples <- SeqArray::seqGetData(gdsin, "sample.id")
-cbind(samples,m2)[order(-m2),]
-#plot histogram 
-hist(m2,breaks=50)
-```
-
-Another common method of filtering and evaluating you data without ever using R is [VCFtools](https://vcftools.github.io/man_latest.html).
-
-**Missingness with vcftools, on the command line**  
-```{bash, eval = T}
-vcftools --vcf RAD_data/OL_subset.vcf --missing-indv --out RAD_data/OL_subset
-# sort the file by most missing data and print the top 10 samples  
-cat RAD_data/OL_subset.imiss | (read h; echo "$h"; sort -k5 -r) 
-```
-You can open the file analysis/OL_subset.imiss in any text editor an look at the missingness values by eye.
-
-To plot the missingness on the command line, you can use gnuplot:
-```{bash, eval = T}
-#code from Jon Puritz
-mawk '!/IN/' RAD_data/OL_subset.imiss | cut -f5 > totalmissing
-gnuplot << \EOF
-set terminal dumb size 120, 30
-set autoscale
-unset label
-set title "Histogram of % missing data per individual"
-set ylabel "Number of Occurrences"
-set xlabel "% of missing data"
-#set yr [0:100000]
-binwidth=0.01
-bin(x,width)=width*floor(x/width) + binwidth/2.0
-plot 'totalmissing' using (bin($1,binwidth)):(1.0) smooth freq with boxes
-pause -1
-EOF
-```
-
-It looks like samples X,Y,Z have a lot of missing data relative to other samples.
-We can remove them from our dataset for now, and just keep exploring our data, but if you decide to exclude them permanently from the analysis you should eventually rerun your genotyping pipeline without those samples, esp.
-if doing a de novo assembly.
-
-2) Another great way to explore missingness with your data is to construct a heatmap of loci that are genoptyped between pairs of samples.
-Generally, samples that are more closely related will share more loci with each other due to cutsite dropout (cite).
-Deviations from that pattern can pinpoint bad samples as well as batch effects.
-  + [ipyrad API tutorial](https://ipyrad.readthedocs.io/en/latest/API-analysis/cookbook-sharing.html) for making a locus sharing heatmap, requires ipyrad output or ipyrad's vcf2hdf5 conversion  
-  + [code to generate locus sharing/missingness heatmap from a VCF](https://github.com/atcg/clustOpt)
-
-### The power of PCA
-
-One of the most powerful methods for exploring your data is a Principle Components Analysis.
-From [this genetics tutorial](https://comppopgenworkshop2019.readthedocs.io/en/latest/contents/03_pca/pca.html): "To understand how PCA works, consider a single individual and its representation by its 593,124 markers.
-Formally, each individual is a point in a 593,124-dimensional space, where each dimension can take only the three possible genotypes indicated above, or have missing data.
-To visualize this high-dimensional dataset, we would like to project it down to two dimensions.
-But as there are many ways to project the shadow of a three-dimensional object on a two dimensional plane, there are many (and even more) ways to project a 593,124-dimensional cloud of points to two dimensions.
-What PCA does is figuring out the “best” way to do this project in order to visualise the major components of variance in the data." See [here](https://michaelaalcorn.medium.com/yet-another-pca-explanation-727dff6ce26a) for a linear algebra-based explanation of PCA).
-
-When you have a VCF file with SNPs, use PCA before extensive filtering or playing with parameters to look at the data. Check which SNPs are associated with axes showing the most variation.
-
-Here we use SeqArray and SNPRelate to run a PCA in R.
-
-<!--Other popular methods for PCA include pcangsd(link), ...., but note that ... requires some file conversions everytime you filter.-->
-
-*Reminder:* Missing data is a feature of RAD.
-Be aware of how different analysis tools deal with missingness, especially PCA which will fill in all missing data with some values.
-Here, the PCA from SNPRelate imputes missing genotypes as the dosage mean.
-
-##### Doing a PCA with SeqArray and SNPRelate in R.
-
-```{r}
-#open the gds file previously created, if not already open
-#filename.gds = "RAD_data/OL_subset.gds" #replace with your file name 
-#gdsin = SeqArray::seqOpen(filename.gds)
-# exclude samples previously identified as having too much missing data
-bad_samples = c("OR3_13_C3","CA4_12_C1","OR3_3_C7","CA4_13_C1","BC2_6_C2","BC2_9_C4")
-sample.ids = seqGetData(gdsin, "sample.id")
-keep = sample.ids[which(!sample.ids %in% bad_samples)]
-
-```
-Whether your data are denovo or reference-based, it is important to filter out linked sites before performing a PCA.
-Here's how to do it in SeqArray.
-Note, if you have already filtered your VCF file to have only 1 SNP per RAD locus, you don't need to do this:
-```{r,results='hide'}
-snpset <- SNPRelate::snpgdsLDpruning(gdsin, ld.threshold=0.2, autosome.only = F, start.pos="random", num.thread=1, remove.monosnp = T, sample.id = keep)  
-snpset.id <- unlist(unname(snpset))
-```
-
-Now we will actually run the PCA, again removing the samples with missing data, keeping unlinked SNPs, and removing SNPs with less than 5% minor allele frequency. 
-```{r}
-# PCA only on SNPs with a minor allele freq greater than 5%
-pca.out = SNPRelate::snpgdsPCA(autosome.only = F, gdsin, num.thread=2, remove.monosnp = T, maf = 0.05,
-                               snp.id=snpset.id,
-                               sample.id = keep) # filtering for pruned SNPs
-#close the gds file (saves memory)
-#seqClose(gdsin)
-
-eig = pca.out$eigenval[!is.na(pca.out$eigenval)]
-barplot(100*eig/sum(eig), main="PCA Eigenvalues")
-```
-
-<!-- Data Info: Below we use RAD data from an Australian marine species complex. ---> 
-```{r, eval =F,include=F}
-
-## The following code will read from a text file containing a list of the VCF files you wish to explore (one file per line) 
-## Each line must contain the full file path 
-## The code will convert your VCF to GDS
-## Perform PCA
-## Plot the results
-
-list_vcfs = read.table("analysis/Pop_Map/vcf.list")[[1]]
-lapply(list(list_vcfs), function(x) { 
-  filename = gsub(".vcf", "", gsub("data/", "", x)) ## NOTE:: you want a pattern that will remove the file path 
-  print(paste0("PCA for ... ", filename))
-  filename.gds = paste0("data/gds/", paste0(filename, ".gds"))
-  gdsfmt::showfile.gds(closeall=TRUE)
-  # 1 . Convert VCF to GDS
-  SeqArray::seqVCF2GDS(vcf.fn = x, out.fn = filename.gds, storage.option="ZIP_RA")
-  gdsin = SeqArray::seqOpen(filename.gds)
-  print(paste0("The number of SAMPLES in data: ", length(c(SeqArray::seqGetData(gdsin, "sample.id")))))
-  print(paste0("The number of SNPs in data: ",  length(c(SeqArray::seqGetData(gdsin, "variant.id")))))
-}) # end apply
-
-sample.ids = seqGetData(gdsin, "sample.id")
-sample.strata =  read.table("data/test.metadata.csv", header = T, sep = "\t") %>%
-                  dplyr::select(TARGET_ID, STRATA, LONG, LAT, PLATE)
-id.order = sapply(sample.ids, function(x,df){which(df$TARGET_ID == x)}, df=sample.strata) #incase your strata file is not in the same order as your vcf
-sample.strata.order = sample.strata[id.order,]
-
-# get pruned SNPs
-# NOT WORKING YET, issue with chromosome names
-# Try this with vcf https://github.com/zhengxwen/SNPRelate/issues/10  
-#snpset <- SNPRelate::snpgdsLDpruning(gdsin, ld.threshold=0.2, autosome.only = F, start.pos="random", num.thread=1)  
-#snpset.id <- unlist(unname(snpset))
-
-# PCA only on SNPs with a minor allele freq greater than 2.5%
-pca.out = SNPRelate::snpgdsPCA(autosome.only = F, gdsin, num.thread=2, remove.monosnp = T, maf = 0.025)
-#                               snp.id=snpset.id) # filtering for pruned SNPs
-eig = pca.out$eigenval[!is.na(pca.out$eigenval)]
-barplot(100*eig/sum(eig), main="PCA Eigenvalues")
-```
-
-First color/shape the individuals by the factor you expect to matter (eg, sampling site or region, family, ecotype).  
-```{r, eval=T}
-#PLOT PCA
-#PC1 v PC2 colored by collection location
-id.order = sapply(keep, function(x,df){which(df$ID == x)}, df=sample.strata) #incase your strata file is not in the same order as your vcf
-sample.strata.order = sample.strata[id.order,]
-print(
-  as.data.frame(pca.out$eigenvect) %>%
-      tibble::add_column(., STRATA =  sample.strata.order$STRATA) %>%
-      ggplot(., aes(x=V1, y=V2, color = STRATA)) + 
-      geom_point(size=2) +
-      stat_ellipse(level = 0.95, size = 1) +
-      geom_hline(yintercept = 0) +
-      geom_vline(xintercept = 0) +
-      theme_bw() +
-      xlab(paste0("PC1 [",paste0(round(eig[1], 2)), "%]")) +
-      ylab(paste0("PC2 [",paste0(round(eig[2], 2)), "%]")) +
-      ggtitle("PCA Colored by Collection Location")
-)
-```
-
-#### Questions to ask about your PCA: {.unlisted .unnumbered}
-
-* Is there any clustering? If so, is it different than your expectation?
-  + If different than expectation, suggests either [batch effects](#batch), [cryptic variation](#cryptic) among individuals, or systematic biological issues affecting assembly/mapping
-* Are there outlier samples driving a large amount of variation?
-  + If so, they may be [cryptic species](#cryptic), clones/sample replicates, or they [sequenced/genotyped poorly](#bad)
-  + In our example we see 2 potential outlier samples in the bottom left corner. We can check which samples they are by plotting the PCA with sample ID and/or looking at the actual PC scores.
-
-```{r, eval=T}
-#PLOT PCA
-#PC1 v PC2 with Sample Labels
-print(
-  as.data.frame(pca.out$eigenvect) %>%
-      tibble::add_column(., ID =  sample.strata.order$ID) %>%
-      ggplot(., aes(x=V1, y=V2, label = ID)) +
-      geom_text(size =3) +
-      geom_hline(yintercept = 0) +
-      geom_vline(xintercept = 0) +
-      theme_bw() +
-      xlab(paste0("PC1 [",paste0(round(eig[1], 2)), "%]")) +
-      ylab(paste0("PC2 [",paste0(round(eig[2], 2)), "%]")) +
-      ggtitle("PCA with Labels")
-)
-```
-
-```{r, eval=F}
-# which samples have PC1 > -0.7 and PC2 < -0.3?
-as.data.frame(pca.out$eigenvect) %>%
-      tibble::add_column(., ID =  sample.strata.order$ID) %>%
-  filter(V1 > 0.7 & V2 < -0.3) %>%
-  select(ID,V1,V2)
-```
-
-* Is there no clustering at all? If so, it may be due to:  
-  + bioinformatic artifacts leading to noise, [see Testing Parameters](#test)
-  + issues with how you are [filtering the data](#filter) (eg, too strict or too lax)
-  + issues with how your PCA is treating missing data
-  + or you actually have no structure in your dataset!
-
-<!--`r colorize("figure example, no clustering", "red")`-->
-
-* If you used sample replicates, first see if replicates cluster very close together.
-If not, then there may be batch effects or bioinformatic artifacts leading to genotyping error that must be addressed.
-Then remove replicates for a subsequent PCA evaluation.
-
-<!--Here is an example of a PCA with sample replicates, and then with sample replicates removed: -->
-
-When removing a sample replicate, you can choose one at random or pick the replicate with the least [missing data](#bad). 
-```{r, eval=FALSE}
-# example code for removing samples from a SNPRelate PCA 
-# assumes you have already loaded a vcf file as a GDS
-samples.to.remove = c("SampleA_rep","SampleB_rep")
-sample.ids = seqGetData(gdsin, "sample.id")
-keep = sample.ids[which(!sample.ids %in% bad_samples)]
-
-# PCA only on SNPs with a minor allele freq greater than 2.5%, only keeping samples in keep
-pca.out.noreps= SNPRelate::snpgdsPCA(autosome.only = F, gdsin, num.thread=1, remove.monosnp = T, maf = 0.05,
-                               snp.id=snpset.id,
-                               sample.id = keep)
-
-```
-
-Next, color the individuals by potential sources of batch effects (sequencing lane, library prep batch, age of tissue sample, person doing the DNA extraction...).    
-```{r, eval=T}
-# again, but colored by batch
-print(
-  as.data.frame(pca.out$eigenvect) %>%
-      tibble::add_column(., PLATE =  as.factor(sample.strata.order$PLATE)) %>%
-      ggplot(., aes(x=V1, y=V2, color = PLATE))  + #label = ID
-      geom_point(size=2) +
-      stat_ellipse(level = 0.95, size = 1) +
-      geom_hline(yintercept = 0) +
-      geom_vline(xintercept = 0) +
-      theme_bw() +
-      xlab(paste0("PC1 [",paste0(round(eig[1], 2)), "%]")) +
-      ylab(paste0("PC2 [",paste0(round(eig[2], 2)), "%]")) +
-      ggtitle("PCA colored by Batch/Sequencing Plate")
-)
-
-```
-
-* Do any observed clusters match a source of batch effect? [how to remove batch effects](#batch)
-
-<!-- #### Visually inspect alignments {#align}  
-Another useful approach is to visually inspect alignments for SNPs or loci that are identified as "weird" in the subsequent analyses.
-In ipyrad this can be done by searching the .loci file, in dDocent...
-If the alignment looks wonky, you can feel better about removing the whole locus (?).
-`r colorize("code example .loci", "red")`
-`r colorize("code dDocent", "red")`  
-`r colorize("code stacks", "red")`    
--->
-
-### Batch effects {#batch}
-As discussed in [Considerations During Lab Work](#labwork), batch effects can arise by minor (or not so minor) differences during the library prep and sequencing stage.
-Randomizing samples across libraries and sequencing lanes can help mitigate the influence of batch effects on your downstream analyses, but sometimes this isn't possible, esp.
-when using pre-existing data.
-Even if you randomize your samples, it is a good idea to check for batch effects with a PCA and then try to mitigate them.
-
-Nicolas Lou and Nina Therkildsen have a [great preprint](https://www.authorea.com/users/380682/articles/532568-batch-effects-in-population-genomic-studies-with-low-coverage-whole-genome-sequencing-data-causes-detection-and-mitigation) and [Github repo](https://github.com/therkildsen-lab/batch-effect) discussing causes, detection, and mitigation of batch effects in low coverage WGS, where batch effects can have particularly large effects.
-While some of their recommendations are specific to reference-based analyses, others are relevant to RAD methods too (eg, batch effects due to different sequencing platforms, DNA degradation levels, and sequencing depth).
-
-<!--`r colorize("PCA figure of batch effect", "red")`-->
-
-You can identify which loci are driving the batch effect by 1) identifying the top SNPs contributing to PC variation (termed "loadings") and/or 2) do an Fst outlier analysis with the batches specified as populations (eg, Bayescan).
-Both of these methods are most appropriate if you've randomized your samples between batches.
-If you have not randomized samples but are certain there is a batch effect driving variation in your dataset, you can still remove loci this way but you may also be removing biologically informative loci.
-
-In SeqArray/SNPRelate, if you have batch effects that are identifiable by PCA, you would extract the PC loadings for each SNP at the PC which represents the batch effect, plot a histogram to see the distribution, choose a cutoff, then filter those SNPs out and create a new VCF for downstream analyses.
-For this example, let's pretend that we see a batch effect on PC1.
-```{r, eval=T}
-# assume you have a GDS object called gdsin and a PCA object called pca.out
-# extract PCA loadings
-SnpLoad <- snpgdsPCASNPLoading(pca.out, gdsin)
-names(SnpLoad)
-dim(SnpLoad$snploading)
-#plot a histogram of absolute PC loading values for PC1, with larger loading meaning more effect on PC1
-pc<-1
-hist(sort(abs(SnpLoad$snploading[pc,]),decreasing=T,index.return=T)[[1]],breaks = 30,main="PC loadings: PC1",)
-```
-Let's remove the largest effect loci over abs(0.010) and see if it changes our PCA.
-```{r}
-#extract SNP.IDs for loadings on PC 1 > abs(0.010) 
-batch.snp.ids <- SnpLoad$snp.id[which(abs(SnpLoad$snploading[1,]) >= 0.010)]
-
-#rerun PCA
-snp.id.allfilt <- setdiff(snpset.id,batch.snp.ids)
-pca.out.pcfilt = SNPRelate::snpgdsPCA(autosome.only = F, gdsin, num.thread=2, remove.monosnp = T, maf = 0.05,
-                               snp.id=snp.id.allfilt,
-                               sample.id = keep) # filtering for pruned SNPs
-print(
-  as.data.frame(pca.out.pcfilt$eigenvect) %>%
-      tibble::add_column(., STRATA =  sample.strata.order$STRATA) %>%
-      ggplot(., aes(x=V1, y=V2, color = STRATA)) + 
-      geom_point(size=2) +
-      stat_ellipse(level = 0.95, size = 1) +
-      geom_hline(yintercept = 0) +
-      geom_vline(xintercept = 0) +
-      theme_bw() +
-      xlab(paste0("PC1 [",paste0(round(eig[1], 2)), "%]")) +
-      ylab(paste0("PC2 [",paste0(round(eig[2], 2)), "%]")) +
-      ggtitle("PCA Colored by Collection Location")
-)
-
-```
-
-Our PCA is now clustering better by population, so we will now export a new VCF that includes all the filtering we've done so far: 1) removed individuals with missing data, 2) thinned our SNPs for linkage, and 3) removed SNPs driving a batch effect.
-
-To save your filtered GDS object as a VCF in SeqArray: 
-```{r}
-# set a filter to exclude samples with missing data, and SNPs in linkage or with a batch effect
-seqSetFilter(gdsin, sample.id=keep,variant.id=snp.id.allfilt)
-
-# convert to vcf
-seqGDS2VCF(gdsin, "RAD_data/OL-miss-pc1-linkage-filt.vcf.gz")
-
-#close the gds file to save memory
-closefn.gds(gdsin)
-
-```
-
-### Cryptic species/contamination/clones {#cryptic}  
-One issue that is particularly common, especially for marine invertebrates, are cryptic, yet highly diverged lineages occurring in the same geographic location.
-If you are trying to do a landscape/seascape genetics study, cryptic species can dramatically throw off population genetic analyses.
-A PCA using all the samples can help identify cryptic species if samples are forming clusters that do not align with any other possible factor (eg, geography, sequencing batch).
-
-<!--Note: Notice the huge variation explained along the first PC-axis.-->
-
-You can also check heterozygosity. Large differences in heterozygosity between outliers can suggest different species or cryptic hybrids.
-
-<!--Note: Two outliers are obvious.
-Large differences in heterozygosity between samples are clear, with one location showing two distinct clusters of samples.
-These samples were later determined using mtDNA and morphological analysis to be cryptic species.
-*Unpublished data from a marine species in Australian waters.*-->
-
-You can then estimate genetic divergence between these clusters using pairwise Fst and/or phylogenetic inference to determine how diverged the groups are.
-If they are very diverged, it may be better to assemble each cryptic species separately, especially for denovo assemblies with dDocent and Stacks, especially since often times restriction cut-sites are not always conserved; see [@Wagner2013-ff,@Rubin2012-hw].
-These two pipelines used a set of given samples to create a "reference" or "catalog" to which all other samples are mapped.
-If highly diverged samples are included in creating this reference, it can lead to [oversplitting](#cluster).
-
-Cryptic clones can also dramatically throw off population genetic analyses.
-In a PCA, clones will often appear as their own very tight cluster, similar to sample replicates.
-A calculation of sample pairwise distance (e.g., Manhattan distance) can very duplicates/clones.
-
-<!--Below the PCA shows x and a simple calculate of pairwise-distance (i.e. here manhattan distance) shows x known duplicates, and x unknown clones. 
-`r colorize("PCA figure of cryptic clones", "red")`-->
-
-## Test a range of key parameters {#test}
-
-There are many parameter choices to make when assembling and genotyping, and those parameters vary slightly across pipelines.
-The parameters below are general across RADseq pipelines (although with different names), and are shown to have the largest impact on genoyping error rates.
-
-### Clustering threshold {#cluster}  
-#### (de novo assembly only)  
-During *de novo* RADseq assembly, reads are assembled into contigs (contiguous sequences), so that each contig represents a single locus.
-This assembly can occur separately for each individual (ipyrad) or jointly for multiple individuals (dDocent, Stacks).
-To call SNPs, in the case of ipyrad, contigs are then clustered and aligned across individuals, while in dDocent and Stacks, reads for each individual are mapped to the "reference" contigs.
-All of these pipelines use sequence similarity (or **clustering threshold**) to determine if fragments are homologous (from the same place in the genome) and therefore should be clustered together.
-
-If the clustering threshold is too low, *undersplitting* can occur so that multiple loci are combined into a single contig.
-Undersplitting is common with repetitive regions and paralogs.
-It can inflate mean observed heterozygosity.
-If the threshold is too high, it can lead to *oversplitting*, where alleles of the same locus are split into two or more contigs.
-This can lead to reduced mean observed heterozygosity.
-
-If severely off, both of these issues can bias the models used to identify SNPs.
-Undersplitting a little is better than oversplitting, as it is easier to identify lumped paralogs downstream and remove them.
-
-#### **What influences the "optimal" clustering threshold of your data?**  
-Clustering threshold is usually determined by how much genetic diversity you have in your dataset.
-If your dataset has high genetic diversity (eg., a single species with high heterozygosity, multiple highly diverged populations, multiple species), then you will need a lower clustering threshold.
-If your dataset has low genetic diversity (a single species with lower heterozygosity, a single family), then you will need a higher clustering threshold.
-Polyploid or highly repetitive (i.e., large) genomes may also have different requirements than diploid species with smaller genomes.
-Other factors, such as sequencing error rate, may play a role as well.
-**We recommend always testing a few clustering thresholds with a new dataset and/or new species**.
-However, as long as your clustering threshold is in the ballpark it shouldn't affect downstream analyses (according to [Isaac Overcast](https://isaacovercast.github.io/)).
-
-There are a few strategies to identify a good clustering threshold for your data, all of which require assembling your data with a range of clustering thresholds.  
-1) Metrics developed by [@McCartney-Melstad2019-ac], based on popgen theory and landscape genetic expectations  
-2) Utilize sample replicates (either exact replicates or individuals from the same locality) to quantify genotyping error and minimize error while maximizing the number of informative loci [@Mastretta-Yanes2015-jb]
-3) Evaluating the number of haplotypes across loci, with many 3+ haplotypes indicating undersplitting [@Harvey2015-vh;@Ilut2014-ja]
-4) Make assemblies at a range of thresholds and parameters, plot the # of SNPs or contigs per assembly, and make a judgement call [example in dDocent tutorial](https://www.ddocent.com/assembly/).
-If most parameters give you about the same # of SNPs or contigs, you can feel confident that this parameter doesn't matter much for your data and just choose a reasonable value.
-
-### Mapping parameters 
-#### (reference-based and some de novo assemblers) {.unlisted .unnumbered}    
-When a sequencing read maps to the wrong assembled contig or place in a reference genome (ie, a non-homologous locus), this can result in erroneous SNPs.
-In many cases, these SNPs can be filtered out later (link to filtering).
-However, data should be checked to make sure there aren't systematic, wide-scale mapping issues that need to be addressed.
-
-For referenced-based assembly, both dDocent and ipyrad make use of bwa(link) to map reads, while Stacks allows the user to choose their aligner. dDocent also utilizes bwa for de novo assemblies.
-
-Mapping defaults for popular RADseq pipelines:
-L: clipping penalty for 5' and 3'
-A: matching score
-O gap open penalty
-T: mapping quality filter
-
-dDocent: bwa mem -L 20,5 -t 8 -a -M -T 10 -A1 -B 3 -O 5 -R
-
-* can change -A and -O during the dDocent pipeline.    
-ipyrad: (defaults) -L 5,5 -B 4 -A1 -O 6 -T 30
-
-* requires changing the ipyrad source code to change parameters
-
-Our whole genome resequencing section has some good content on mapping(link). General considerations with RAD data:
-
-* how many reads aligned for each sample? If many reads didn't aligned, this could be an issue with the reference, read trimming/filtering, or the mapping parameters.
-* What is the distribution of mapping scores?  
-* ratio of mean (across individuals) mapping quality scores for reference and alternate allele [@OLeary2018-iy]. 0.25-1.75 cut off implemented in dDocent.  
-  + this info might not be coded in all VCFs
-
-### Read depth threshold {#seq}
-
-Mistakes that occur during sequencing can lead to a base-calling error in a read, which can then lead to an erroneous SNP.
-This type of error most commonly manifests as rare alleles (eg, only observed in one individual) or singletons (only one copy of the minor allele across all individuals).
-A conservative approach to deal with these errors is to filter out all loci with rare alleles (link filtering), but for some demographic analyses this filtering is inappropriate as they require the site frequency spectrum.
-The best ways to minimize the impact of sequencing error during bioinformatic processing is to 1) properly trim and filter your data, and 2) require higher read depth thresholds for assembly and genotyping.
-
-##### 1) Filtering reads
-
-FASTQ files that typically come from the sequencing center include a [Phred scale quality score](https://gatk.broadinstitute.org/hc/en-us/articles/360035531872-Phred-scaled-quality-scores) for each base, which indicates the probability of that base call being correct.
-This information can be used prior to assembly to trim low-quality sections from either end of reads or remove reads altogether.
-Some SNP callers, including FreeBayes which is part of dDocent, also use Phred scores when calling SNPs (ipyrad and Stacks do not).
-All three of the most popular RAD pipelines include steps to filter reads due to low quality bases low quality bases or adapter sequences, but vary slightly on what kinds of read trimming are allowed.
-Always make sure to filter out adapter sequences (this is not default with stacks), otherwise default parameters are usually adequate.
-If your initial [fastqc](#fastqc) indicates that you have many low quality bases in your reads, you may lose too much of your data during filtering.
-In this case, you may want to make your read filtering a little more permissible (keeping in mind that this may increase SNP error rate).
-**Always make sure to check with your sequencing center to see if they have done anything to the raw reads before you received them!**
-<!---turn these into html tabs--->
-
-**dDocent: **
-
-* does not allow you to use reads that have already been trimmed and filtered for de novo assembly, only referenced-based.
-* Uses Trimmomatic to trim Illumina TruSeq adaters and low quality bases. Make sure to change the defaults if you used a different adapter/sequencing system.
-
-**Stacks: **
-
-* requires your reads to be all the same length, so if they were previously trimmed of low quality bases or adapter sequence this must be specified and they will be truncated to the same length.
-
-##### 2) Require higher read depth thresholds for assembly and genotyping.  
-For RADseq, ideally you are generating an average of *at least* 10-20 sequence reads (10x-20x coverage) for each RAD locus in your library preparation (but this doesn't always happen...).
-For diploid species, this would be sequencing each RAD haplotype at least 5x-10x.
-The SNP calling step of RAD pipelines all require setting a minimum number of reads to determine whether a site is homozygous or heterozygous.
-If this threshold is too low, sequencing errors are more likely to be included and the SNP caller will have a harder time accurately identifying SNPs ([in which case, you'd be better off with genotype likelihood methods](https://github.com/therkildsen-lab/genomic-data-analysis/blob/master/lcwgs_data_analysis.md).
-If it's too high, then real polymorphic loci will be excluded.
-All of the three major pipelines allow users to implement read depth thresholds
-
-Generating multiple sequencing reads per locus is helpful when calling SNPs in order to  distinguishing between reads with sequencing errors and.
-
-## Filtering SNPs {#filter}
-
-### (and embrace the missingness!) {.unlisted .unnumbered}
-
-`r colorize("This section is still under development! Please contribute on Github or provide feedback in [Discussions](https://github.com/MarineOmics/marineomics.github.io/discussions)", "red")`
-
-Even with a reduced representation sequencing approach, your assembly and genotyping pipeline will generate data for thousands of SNPs across your individuals.
-These SNPs will always need to be filtered before doing downstream analyses.
-
-#### Main reasons to filter your data:
-
-* Remove potentially erroneous SNPs that remain after optimizing your assembly pipeline
-* Remove SNPs that are uninformative for downstream analyses 
-* Subsample to remove SNPs that are in linkage disequilibrium (required for analyses that assume independence like PCA, Structure)
-
-Unique to RADseq approaches is the missingness of RAD data.
-
-* Don't over filter for missingness
- + papers showing how too much filtering can skew results  
-  + Empirical: Diaz-Arce 2019, Tripp 2017, Silliman 2021, Eaton 2017, Martin-Hernanz (2019)
-  + Simulations: Huang & Knowles 2016, Leaché et al., 2015, Gautier et al., 2012, Rivera-Colón et al., 2021
-
-* some ways to account for missing data in downstream analyses
- + impute (PCA), subsample (PCA, structure, pretty much anything with SNPs), try a few different missing filters and see what impact it has on results
-
-<!-- use [pcANGSD](link)-->
-
-<!--## Special Cases  
-### DARTseq  
-### 2bRAD
-
-### Reference-based vs. de novo RADseq
--->
-
-## Repeat as needed {#repeat}
-# Helpful tutorials  
-* [Marine Genomics open course](https://baylab.github.io/MarineGenomics/)
-* [UNIX tutorial](http://evomics.org/learning/unix-tutorial/)
-* [what is a VCF file and how can I filter it?](http://evomics.org/learning/population-and-speciation-genomics/2020-population-and-speciation-genomics/first-steps-in-genomic-data-analysis/)
-
-# References
+## References

### 2022-11-16 submissions_instructions.Rmd
more headway on the submissions_instructions page

@@
 ---
-title: 'Instructions for developing new pages to the MarineOmics webpage'
+title: 'Instructions for developing new pages for the MarineOmics webpage'
 bibliography: RCN_site-Nov05.bib
@@
 
-```{r setup, message = FALSE, echo = FALSE}
+```{r setup, echo = FALSE, message = FALSE}
 library(knitr)
-knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
+knitr::opts_chunk$set(cache = TRUE)
 library(knitcitations)
@@
 This is a guide for creating a new page for the website.
-Of course, the website managers will work with anyone who develops new content to answer questions and publish the site, but this can be used as a resource as you develop your page.
+Of course, the website managers will work with anyone who develops new content to help develop and publish your page, but this can be used as a resource as you develop your content.
 In this guide are general instructions as well as examples of how to build specific features.
@@
 
-\*\* Before publishing any content, please contact Jason Johns (jasonjohns[at]ucsb.edu) so we can have working group contributors review the page.
+\* Even if you comfortable with publishing websites in Github/RMarkdown, please contact Jason Johns (jasonjohns[at]ucsb.edu) before publishing any new pages so we can ensure smooth incorporation into the website.
 
@@
 In short, each page of the website is rendered as an .html from an RMarkdown (.Rmd) document.
-The two most important files for publishing the website are the .site_yml ("the yaml file") and the index.Rmd/index.html.
+The two most important files for publishing the general website are the .site_yml ("the yaml file") and the index.Rmd/index.html.
 
-**.site_yml**: the 'under the hood' backbone of the website, where things like the visual appearance of the website the navigation bar are built
+**.site_yml**: the 'under the hood' backbone of the website, where things like the visual appearance of the website and the navigation bar are built
 
@@
 
+<hr />
+
 # R Markdown basics
@@
 If you don't have much experience with RStudio, it's quite intuitive with a shallow learning curve and many existing online resources for using it.
+It is possible to use .Rmd files in base R, but we recommend using RStudio.
 
-*Note: it's a good idea to look at existing .Rmd scripts to see examples of how a particular feature can be accomplished, such as citing references, linking a website, building a table, incorporating figures, etc.
-We will explicitly go over some of those feautures below, but there are many other bells and whistles that can be explored through other resources online.*
+*Note: it's a good idea to look at existing .Rmd scripts (including the one that codes for this page) to see examples of how a particular feature can be accomplished, such as citing references, linking a website, building a table, incorporating figures, etc.
+We will explicitly go over some of those features below, but there are many other bells and whistles that can be explored through online resources.*
 
-## Initial Page Setup
-- caching
-- installing & loading packages
-- etc.
+Here are some general RMarkdown tutorials that we like to reference:
+<br>
+- [Chapter 10 from RMarkdown the Definitive Guide: Xi, Allaire, & Grolemund 2022](https://bookdown.org/yihui/rmarkdown/rmarkdown-site.html)
+<br>
+- [R Markdown Websites: Garrett Grolemund](https://garrettgman.github.io/rmarkdown/rmarkdown_websites.html)
 
-## Formatting
-- text vs. code chunks, how to include code on the page or not, include output or not, etc.
-- navigation bar on the left, font size, bold, italics, etc
-- in-text citations, hyperlinks
+<br/>
 
-## Making a table - Katherine/Sam
+## RMarkdown formatting
+One of the strengths of .Rmd documents is their versatility.
+They can hold everything from code (not only in R but in other languages), formatted text, figure outputs, and more.
+Basically, code must be enclosed within 'code chunks' and anything outside of code chunks will be read as text.
+So to include non-code text, we don't need any #, unless of course we're in a code chunk.
 
-## Including a figure - Katherine/Sam
+### **code chunks**
+Below is a basic code chunk which can be inserted by clicking the green +C button on the RStudio toolbar or typing the keyboard shortcut:<br>
+- macOS: command + option + I <br>
+- Windows: control + alt + I
 
-## Including an image - Katherine/Sam
+````
+```{r, echo = TRUE, include = TRUE} `r ''`
 
-## Running code in another language such as bash - Katherine/Sam
+```
+````
+Whether code chunks and/or their output are included in the knitted .html can be modified with the logical commands "echo" and "include".<br>
+- echo: include the code chunk itself or not<br>
+- include: include the output of the code chunk or not
 
-## References
+\* Other options exist for code chunk output which can be explored online and in other .Rmd scripts in this repository.
+
+<br>
+
+#### *global code chunk options*
+This is the chunk where we establish global settings for the entire document.
+This is done with the command from the knitr package "opts_chunk$set" as shown in the example below.
+This command is typically included in the first chunk, where we also load our packages and adjust the settings for citations.
+When a new .Rmd is made, this command is automatically included in the first chunk, which is named "setup" by default.
+The name of a code chunk can be specified after the r in ```{r ____}.
+
+````
+```{r setup} `r ''`
+knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, cache = TRUE)
+
+library(knitr)
+library(knitcitations)
+library(kableExtra)
+
+knitr::opts_chunk$set(fig.width = 10, fig.height = 5) #setting the dimensions for figure outputs in a different line of code for cleanliness
+
+cite_options(citation_format = "pandoc", max.names = 3, style = "html", 
+             hyperlink = "to.doc") #setting format for in-text citations for the references section at the bottom of the page
+```
+````
+\* Note the "cache = " option.
+When set to "true" R will cache all of the output created from knitting the RMarkdown document.
+This is especially useful for pages with heavy computation that take a long time to knit.
+<br>
+
+### **formatting text**
+Options abound for formatting text in .Rmd documents using [Pandoc Markdown](https://pandoc.org/MANUAL.html), and many of these can be found with a quick search, but we will go over a few basics here.
+
+As with writing code, any text symbol that codes for something, such as a \# or a \* must be escaped with a backslash (\\) if you just want to use the symbol itself.
+
+**\#**: *Outside of a code chunk*, pound signs dictate the size and organizational level of the text.
+A single \# will make large text and a main section in the navigation window.
+Two \#\# will make slightly smaller text and a sub-section in the navigation window.
+<br>
+
+For example, the "RMarkdown formatting" text and category was created with "\# RMarkdown formatting" while the "code chunks" subcategory was created with "\#\# code chunks."
+
+**\***: Asterisks are used to bold and italicize when surrounding texts. Use \*\***two asterisks to bold**\*\* and \**one asterisk*\* to italicize.
+
+**\<br>**: adds a line break
+
+**\<hr />**: adds a horizontal line across the page
+
+### **installing & loading packages**
+If you will be using any R packages for your information page or tutorial, it's a good idea to highlight them toward the beginning of your page, including code for installing them.
+Here's an example from the RADseq tutorial:
+
+If you would like to run the R code examples that are scattered throughout the guide (recommended but not required!), you will need to install some R packages. Only need to run this code once:  
+```{r, eval=F}
+install.packages("tidyverse")
+if (!requireNamespace("BiocManager", quietly = TRUE))
+    install.packages("BiocManager")
+#the BiocManager package facilitates the installation of packages from the Bioconductor project
+BiocManager::install("SeqArray")
+BiocManager::install("SNPRelate")
+```
+
+### **caching**
+If your page includes any code that takes a long time to run, please set "cache = TRUE" in the knitr::opts_chunk$set command in your global code chunk.
+
+### **making a table - Katherine/Sam**
+
+### **including a figure - Katherine/Sam**
+
+### **including an image - Katherine/Sam**
+
+### **running code in another language such as bash - Katherine/Sam**
+
+<hr />
+
+### **References**

### 2022-12-02 submissions_instructions.Rmd
Push Sam's formatting details for tables, citations, equations, and image insertions

@@
 ---
-title: 'Instructions for developing new pages for the MarineOmics webpage'
+title: "Instructions for developing new pages for the MarineOmics webpage"
 bibliography: RCN_site-Nov05.bib
@@
 
+### Example of table syntax
+
+| Term A | Term B | Term C | Term D | Term E | Term F | Term G |
+| :--- | :--- | :--- | :--- | :--- | :--- | :--- |
+| xxx | xxx | xxx | xxx | xxx | xxx | xxx |  
+| yyy | yyy | yyy | yyy | yyy | yyy | yyy | 
+| zzz | zzz | zzz | zzz | zzz | zzz | zzz |
+
+<br>
+
+### Example of how to insert an image
+
+```{r, echo = FALSE, out.height = "40%", fig.align = "center", }
+
+knitr::include_graphics("images/Rivera_etal_fig.png")
+
+```
+
+### Example of how to write and print an equation
+
+For illustrative purposes, here is a simple and famous equation
+
+$$
+E = mc^2
+$$
+
+<br>
+
+As an example of a more complex equation, here is an equation for the t-statistic
+
+$$
+t = \frac{\overline{x_{1}}-\overline{x_{2}}}
+{\sqrt{(s^2(
+\frac{1}{n_{1}} + \frac{1}{n_{2}}
+))}}
+$$
+
+### Example of how to cite a reference from your .bib in-text
+
+RADseq can produce insightful analyses even in the absencee of data, if missing values are handled properly [@Tripp2017-cp].
+
+<br>
+
 ### **code chunks**
@@
 knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, cache = TRUE) 
-
 library(knitr)
@@
 library(kableExtra)
-
 knitr::opts_chunk$set(fig.width = 10, fig.height = 5) #setting the dimensions for figure outputs in a different line of code for cleanliness
-
 cite_options(citation_format = "pandoc", max.names = 3, style = "html", 
@@
 <br>
-
 ### **formatting text**
 Options abound for formatting text in .Rmd documents using [Pandoc Markdown](https://pandoc.org/MANUAL.html), and many of these can be found with a quick search, but we will go over a few basics here.
-
 As with writing code, any text symbol that codes for something, such as a \# or a \* must be escaped with a backslash (\\) if you just want to use the symbol itself.
-
 **\#**: *Outside of a code chunk*, pound signs dictate the size and organizational level of the text.
@@
 <br>
-
 For example, the "RMarkdown formatting" text and category was created with "\# RMarkdown formatting" while the "code chunks" subcategory was created with "\#\# code chunks."
-
 **\***: Asterisks are used to bold and italicize when surrounding texts. Use \*\***two asterisks to bold**\*\* and \**one asterisk*\* to italicize.
-
 **\<br>**: adds a line break
-
 **\<hr />**: adds a horizontal line across the page
-
 ### **installing & loading packages**
@@
 Here's an example from the RADseq tutorial:
-
 If you would like to run the R code examples that are scattered throughout the guide (recommended but not required!), you will need to install some R packages. Only need to run this code once:  
@@
 ```
-
 ### **caching**

### 2023-01-16 submissions_instructions.Rmd
formatting test
some edits from Joanna, some rearranging and reformatting

@@
 
-<br/>
+<br>
 
-## RMarkdown formatting
+## RMarkdown formatting basics
 One of the strengths of .Rmd documents is their versatility.
 They can hold everything from code (not only in R but in other languages), formatted text, figure outputs, and more.
-Basically, code must be enclosed within 'code chunks' and anything outside of code chunks will be read as text.
-So to include non-code text, we don't need any #, unless of course we're in a code chunk.
-
-### Example of table syntax
-
-| Term A | Term B | Term C | Term D | Term E | Term F | Term G |
-| :--- | :--- | :--- | :--- | :--- | :--- | :--- |
-| xxx | xxx | xxx | xxx | xxx | xxx | xxx |  
-| yyy | yyy | yyy | yyy | yyy | yyy | yyy | 
-| zzz | zzz | zzz | zzz | zzz | zzz | zzz |
-
-<br>
-
-### Example of how to insert an image
-
-```{r, echo = FALSE, out.height = "40%", fig.align = "center", }
-
-knitr::include_graphics("images/Rivera_etal_fig.png")
-
-```
-
-### Example of how to write and print an equation
-
-For illustrative purposes, here is a simple and famous equation
-
-$$
-E = mc^2
-$$
-
-<br>
-
-As an example of a more complex equation, here is an equation for the t-statistic
-
-$$
-t = \frac{\overline{x_{1}}-\overline{x_{2}}}
-{\sqrt{(s^2(
-\frac{1}{n_{1}} + \frac{1}{n_{2}}
-))}}
-$$
-
-### Example of how to cite a reference from your .bib in-text
-
-RADseq can produce insightful analyses even in the absencee of data, if missing values are handled properly [@Tripp2017-cp].
-
-<br>
+Basically, code must be enclosed within 'code chunks' and anything outside of code chunks will be read as text, similar to a basic text file.
 
 ### **code chunks**
-Below is a basic code chunk which can be inserted by clicking the green +C button on the RStudio toolbar or typing the keyboard shortcut:<br>
+Below is a basic code chunk which can be inserted by clicking the green +C button on the RStudio toolbar or typing the keyboard shortcut: <br>
+
 - macOS: command + option + I <br>
@@
 ````
-Whether code chunks and/or their output are included in the knitted .html can be modified with the logical commands "echo" and "include".<br>
-- echo: include the code chunk itself or not<br>
+Whether code chunks and/or their output are included in the knitted .html can be modified with the logical commands "echo" and "include". <br>
+
+- echo: include the code chunk itself or not <br>
 - include: include the output of the code chunk or not
@@
 \* Other options exist for code chunk output which can be explored online and in other .Rmd scripts in this repository.
-
 <br>
@@
 <br>
+
 ### **formatting text**
-Options abound for formatting text in .Rmd documents using [Pandoc Markdown](https://pandoc.org/MANUAL.html), and many of these can be found with a quick search, but we will go over a few basics here.
-As with writing code, any text symbol that codes for something, such as a \# or a \* must be escaped with a backslash (\\) if you just want to use the symbol itself.
-**\#**: *Outside of a code chunk*, pound signs dictate the size and organizational level of the text.
+Options abound for formatting text in .Rmd documents using [Pandoc Markdown](https://pandoc.org/MANUAL.html), and many of these can be found with a quick search, but we will go over a few basics here.<br>
+
+- As with writing code, any text symbol that codes for something, such as a \# or a \* must be escaped with a backslash (\\) if you just want to use the symbol itself. <br>
+- **\#**: *Outside of a code chunk*, pound signs dictate the size and organizational level of the text.
 A single \# will make large text and a main section in the navigation window.
@@
 <br>
-For example, the "RMarkdown formatting" text and category was created with "\# RMarkdown formatting" while the "code chunks" subcategory was created with "\#\# code chunks."
-**\***: Asterisks are used to bold and italicize when surrounding texts. Use \*\***two asterisks to bold**\*\* and \**one asterisk*\* to italicize.
-**\<br>**: adds a line break
-**\<hr />**: adds a horizontal line across the page
+  + For example, the "RMarkdown formatting" text and category was created with "\# RMarkdown formatting" while the "code chunks" subcategory was created with "\#\# code chunks."<br>
+- **\***: Asterisks are used to bold and italicize when surrounding texts. Use \*\***two asterisks to bold**\*\* and \**one asterisk*\* to italicize.<br>
+- **\<br>**: adds a line break <br>
+- **\<hr />**: adds a horizontal line across the page <br>
+
 ### **installing & loading packages**
 If you will be using any R packages for your information page or tutorial, it's a good idea to highlight them toward the beginning of your page, including code for installing them.
-Here's an example from the RADseq tutorial:
-If you would like to run the R code examples that are scattered throughout the guide (recommended but not required!), you will need to install some R packages. Only need to run this code once:  
+Here's an example from the RADseq tutorial:<br>
+<br>
+*If you would like to run the R code examples that are scattered throughout the guide (recommended but not required!), you will need to install some R packages. Only need to run this code once:*  
 ```{r, eval=F}
@@
 
-### **making a table - Katherine/Sam**
+<hr />
 
-### **including a figure - Katherine/Sam**
+## Examples of how to include visual feautures such as tables, images, equations, etc.
 
-### **including an image - Katherine/Sam**
+### **insert a table**
+
+| Term A | Term B | Term C | Term D | Term E | Term F | Term G |
+| :--- | :--- | :--- | :--- | :--- | :--- | :--- |
+| xxx | xxx | xxx | xxx | xxx | xxx | xxx |  
+| yyy | yyy | yyy | yyy | yyy | yyy | yyy | 
+| zzz | zzz | zzz | zzz | zzz | zzz | zzz |
+
+<br>
+
+### **insert an image**
+
+```{r, echo = FALSE, out.height = "40%", fig.align = "center", }
+
+knitr::include_graphics("images/Rivera_etal_fig.png")
+
+```
+
+### **write and print an equation**
+
+For illustrative purposes, here is a simple and famous equation
+
+$$
+E = mc^2
+$$
+
+<br>
+
+As an example of a more complex equation, here is an equation for the t-statistic
+
+$$
+t = \frac{\overline{x_{1}}-\overline{x_{2}}}
+{\sqrt{(s^2(
+\frac{1}{n_{1}} + \frac{1}{n_{2}}
+))}}
+$$
+
+### **in-text citations from your .bib**
+
+RADseq can produce insightful analyses even in the absencee of data, if missing values are handled properly [@Tripp2017-cp].
+
+<br>
 
@@
 
-### **References**
+## **References**

### 2023-01-17 submissions_instructions.Rmd
added section on in text citations, and some re-formatting. Also added code output to Sam's sections

@@
 
-```{r setup, echo = FALSE, message = FALSE}
+```{r #setup, echo = FALSE, message = FALSE}
 library(knitr)
-knitr::opts_chunk$set(cache = TRUE)
+knitr::opts_chunk$set(
+	fig.height = 5,
+	fig.width = 10,
+	cache = FALSE
+)
 library(knitcitations)
@@
 
-## RMarkdown formatting basics
+## Formatting basics
 One of the strengths of .Rmd documents is their versatility.
@@
 \* Other options exist for code chunk output which can be explored online and in other .Rmd scripts in this repository.
+
 <br>
@@
 
+<a id="cite_options"></a>
 ````
@@
 This is especially useful for pages with heavy computation that take a long time to knit.
+
+<br>
+
+#### *including code from another language*
+We can also run code such as bash in an R script by modifying the code chunk header as below: <br>
+
+````
+```{bash, eval = FALSE} `r ''`
+
+```
+````
 <br>
@@
 - **\<br>**: adds a line break <br>
-- **\<hr />**: adds a horizontal line across the page <br>
+- **\<hr />**: adds a horizontal line across the page
+
+<br>
+
+### **in-text citations from your .bib**
+In-text citations can be made throughout the document that will automatically add a reference to your reference section via the following steps: <br>
+
+- add your citations to the RCN_site-Nov05.bib document via your reference manager
+- add the line 'bibliography: RCN_site-Nov05.bib' to your header at the top of the .Rmd document, which looks like this:
+````
+---
+title: "Instructions for developing new pages for the MarineOmics webpage"
+bibliography: RCN_site-Nov05.bib
+---
+````
+- include the 'cite-options' chunk in your 'setup' code chunk. See the initial ['r setup' chunk](#cite_options) above for the syntax of the 'cite_options' command.
+
+You can add Bibtex references to the file RCN_site-Nov05.bib, and then refer to them in your RMarkdown by adding bibliography: RCN_site-Nov05.bib to the header and citing them in the RMarkdown document.
+Check out existing pages for examples.
+
+RADseq can produce insightful analyses even in the absence of data, if missing values are handled properly [@Tripp2017-cp].
+
+<br>
 
@@
 ```
+
+<br>
+
 ### **caching**
@@
 ### **insert a table**
+*text in .Rmd **(not in a code chunk)***:
+````
+| Term A | Term B | Term C | Term D | Term E | Term F | Term G |
+| :--- | :--- | :--- | :--- | :--- | :--- | :--- |
+| xxx | xxx | xxx | xxx | xxx | xxx | xxx |  
+| yyy | yyy | yyy | yyy | yyy | yyy | yyy | 
+| zzz | zzz | zzz | zzz | zzz | zzz | zzz | 
+````
+
+<br>
+
+*knitted output*:
 
@@
 ### **insert an image**
+*code chunk in .Rmd*:
 
-```{r, echo = FALSE, out.height = "40%", fig.align = "center", }
+````
+```{r, echo = TRUE, include = TRUE, out.height = "40%", fig.align = "center"} `r ''`
+
+knitr::include_graphics("images/Rivera_etal_fig.png")
+
+```
+````
+
+<br>
+
+*knitted output*:
+```{r, echo = TRUE, include = TRUE, out.height = "40%", fig.align = "center", }
 
@@
 
+<br>
+
 ### **write and print an equation**
 
-For illustrative purposes, here is a simple and famous equation
-
+For illustrative purposes, here is a simple and famous equation. <br>
+*syntax in .Rmd*:
+````
 $$
@@
 $$
+````
 
@@
 
-As an example of a more complex equation, here is an equation for the t-statistic
+*knitted output*:
+$$
+E = mc^2
+$$
+<br>
 
+As an example of a more complex equation, here is an equation for the t-statistic: <br>
+*syntax in .Rmd*:
+````
+$$
+t = \frac{\overline{x_{1}}-\overline{x_{2}}}
+{\sqrt{(s^2(
+\frac{1}{n_{1}} + \frac{1}{n_{2}}
+))}}
+$$ 
+````
+
+*knitted output*:
 $$
@@
 
-### **in-text citations from your .bib**
-
-RADseq can produce insightful analyses even in the absencee of data, if missing values are handled properly [@Tripp2017-cp].
-
 <br>
-
-### **running code in another language such as bash - Katherine/Sam**
 
@@
 
-## **References**
+## References

### 2023-02-03 rda_tutorial/RDAtraitPredictionTutorial.Rmd
initial push of katie l's rda tutorial files

---
title: "RDA trait prediction tutorial"
author: "KE Lotterhos"
date: "2023-01-20"
output:
  pdf_document: default
  html_document: default
---

This tutorial accompanies the paper "The paradox of adaptive trait clines with non-clinal patterns in the underlying genes"

If the following packages are not installed, be sure to install them first:
```
install.packages("vegan")
install.packages("lfmm")

if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("LEA")
```

Load the libraries and set your working directory:

```{r setup}

libraries_needed <- c("vegan", "LEA", "lfmm")

for (i in 1:length(libraries_needed)){
  library(libraries_needed[i],character.only = TRUE) #laptop
}

knitr::opts_knit$set(root.dir = "~/Documents/GitHub/MVP-NonClinalAF/tutorial") 
#change code to your working directory

```

## Load the data

* A matrix of genotypes in 012 format (counts of reference allele)
  * number of rows = number of individuals
  * number of columns = number of SNPs
* A table with information about sampled individuals (each individual in a row)
* A table with information about SNPs (each SNP in a row)

This data was simulated in SLiM and is associated with the complex multivariate simulation presented in Figure 6 in the paper. Briefly, a non-Wright-Fisher model was simulated on a landscape with 6 environmental variables that reflect different aspects of thermal stress and precipitation in British Columbia. The simulation included 6 environmental traits, each of which adapted to a different environmental variable.

The `ind` table includes the xy location for each individual, the 6 exact trait values (note that these won't exactly equal the trait value calculated from the genotype matrix because of MAF filtering), and the 6 environmental values at their xy location.

The `muts` table includes the linkage group `LG`, the position of the mutation on the genetic map `pos_pyslim`, a unique ID `mutname`, the allele frequency based on the 1000 sampled individuals `a_freq_subset`, and whether or not it had effects on one or more phenotypes `causal`.

Note that you will have to change the working directory to where the data is stored on your computer.

```{r load data}

G <- read.table(unz("Genotypes.txt.zip", "Genotypes.txt"))
dim(G) # 1000 individuals and 26371 loci

ind <- read.table("Individuals.txt", header=TRUE)
dim(ind) #corresponds to rows in G
head(ind)

muts <-  read.table("SNPs.txt", header=TRUE)
dim(muts) #corresponds to columns in G
head(muts)

rownames(G) <- as.character(paste0("i_",ind$ind_index))
colnames(G) <- as.character(muts$mutname)
#G <- as.matrix(G)
head(G[,1:10])
```

## RDA trait prediction funciton

This function predicts an environmental trait through the back-transformation of the RDA "site score" of an individual to a chosen environmental variable (Equation 1 in the manuscript). It makes the prediction for all the individuals that were used to run the RDA.

```{r function}
rda_trait_pred <- function(rdaobj, env_row, K){
  #rdaobj is RDA object
  #envi row is the row of the environment in the biplot output
  #K is the number of RDA axes
  scores <- scores(rdaobj, choices=1:K)
  ind.sc <- scores$sites
  pred <- matrix(NA, nrow=nrow(ind.sc), ncol=K)
  for (k in 1:K){
    pred[,k] <- ind.sc[,k]*eigenvals(rdaobj)[k]*summary(rdaobj)$biplot[env_row,k]
  }
  trait_pred <- scale(rowSums(pred))
 return(trait_pred) 
}
```

## Example of an RDA-predicted environmental trait value

1. First, run the RDA:

Scale the environmental variables to have a mean of 0 and standard deviation of 1
```{r example}

ind$env1_mat <- scale(ind$env1_mat)
ind$env2_MTWetQ <- scale(ind$env2_MTWetQ)
ind$env3_MTDQ <- scale(ind$env3_MTDQ)
ind$env4_PDM <- scale(ind$env4_PDM)
ind$env5_PwarmQ <- scale(ind$env5_PwarmQ)
ind$env6_PWM <- scale(ind$env6_PWM)

# Run the RDA
rdaout <- rda(G ~ ind$env1_mat +
                ind$env2_MTWetQ +
                ind$env3_MTDQ + 
                ind$env4_PDM +
                ind$env5_PwarmQ +
                ind$env6_PWM
              )

```

2. Next, check the biplot output and decide how many RDA axes to use in the prediction.

```{r}
# Check the biplot output
rdaout$CCA$biplot

# Decide how many RDA axes to use in calculation
  a<- screeplot(rdaout)
  str(a)
  a$y # save this it's the eigenvalues
  prop_var <- round(a$y[1:6]/sum(a$y),3)
  cumsum(prop_var)
  plot(cumsum(prop_var), xlab="Number of RDA axes", 
       ylab="Cumulative percent of variation explained", ylim=c(0,1))
```

3. In this case, the first 3 RDA axes explain 80% of the variance. Note that choosing too many axes may result in overfitting. Here is an example of a trait prediction for MAT using the first 3 RDA axes:
```{r}
# Make the trait prediction for MAT (1st row in biplot output)
K = 3 # use 3 RDA axes to make the trait prediciton

MATtraitPredict <- rda_trait_pred(rdaout, 1, K)

# Since this is a simulation, we can compare the prediction to the true value
# Similarly, an empirical study could compare an empirically measured trait value 
# to the RDA-predicted trait value to test how well landscape genomic data 
# can predict functional traits
plot(scale(ind$phenotype1_mat), MATtraitPredict, xlab="Evolved trait value in simulations",
     ylab="RDA trait prediction")
abline(0,1)

#Correlation between the prediction and the true value:
cor(ind$phenotype1_mat, MATtraitPredict) 
```

### Compare to other functions in RDA

Note that the `predict` function and it's variations in the R package `vegan` do not make the same kind of predictions as `rda_trait_pred`. Here are the types of outputs produced by the function `predict` and it's variations:

```{r}

# This option in the `predict` function outputs the scores for each locus in RDA space
loci_scores_predict <- predict(rdaout, type="sp", newdata=G, scaling=2)
str(loci_scores_predict)

# This option in the `predict` function outputs the fitted values from the multiple regression
# performed on each locus within each individual 
fitted_values_predict <- predict(rdaout, newdata=G, type="response")
str(fitted_values_predict)

  # As a side note, it outputs the same thing as the `fitted` funciton
  fitted_values_predict2 <- fitted(rdaout)
  str(fitted_values_predict2)

# This option  in the `predict` function outputs the individual scores in RDA space 
# based on a linear combination of the predictor variables
X <- data.frame(ind$env1_mat ,
                ind$env2_MTWetQ ,
                ind$env3_MTDQ ,
                ind$env4_PDM ,
                ind$env5_PwarmQ ,
                ind$env6_PWM)
ind_scores_predict <- predict(rdaout, type="lc", new=X, scal=2)
str(ind_scores_predict)
```

The `predict` function and it's variations make predictions in RDA space, and therefore do not output the same kind of predictions as `RDA_trait_predict` and Equation 1 in the paper.

### Understanding how the RDA is built on multiple regressions

Prior to ordination in the RDA, each locus is used in a multiple regression model with the environmental variables to produce fitted values for that locus across individuals.

_SNP Genotype_ ~ _Env1_ + _Env2_ + _Env3_ etc.

For example for the first SNP in the data:

```{r}
 # multiple regression of 1st locus

mod <- lm(G[,1] ~ ind$env1_mat + ind$env2_MTWetQ + ind$env3_MTDQ +  ind$env4_PDM +
                ind$env5_PwarmQ +   ind$env6_PWM)
 coef(summary(mod))
```

Although multiple regression is a linear combination of multiple variables, it is able to model complex multivariate responses that appear to be non-monotonic in any one dimension. For example, let's look at a the relationship between explanatory variable temperature and the response variable genotype, across decreasing and increasing values of the other explanatory variables:

```{r}
otherenv <- c(seq(1,0,length.out=100), seq(0,1,length.out=101))
newdata=data.frame(ind.env1_mat = seq(-1,1, by=0.01),
                   ind.env2_MTWetQ = otherenv,
                   ind.env3_MTDQ = otherenv,
                   ind.env4_PDM = otherenv,
                   ind.env5_PwarmQ =otherenv,
                   ind.env6_PWM = otherenv)
pred <- t(newdata)*(coef(summary(mod))[2:7,1]) + coef(summary(mod))[1,1]
plot(seq(-1,1, by=0.01), colSums(pred), xlab="Temperature", ylab="Genotype prediction")
```

Thus, there is flexibility with the RDA to capture the way environmental variables may influence the patterns at one locus in a different way than at another locus, which may not correlate with the relationship between the environment and population structure.

It may be interesting for some studies to understand how each locus is shaped by the environment - in other words, what are the slopes associated with the environmental variables in the multiple regression model for each locus?

Unfortunately there is not a way to output these slopes in the R package `vegan`, but we can reproduce the first step of the RDA to get the regression coefficients:
(vegan source code at https://github.com/cran/vegan/blob/master/R/simpleRDA2.R)

```{r}

X <- data.frame(ind$env1_mat ,
                ind$env2_MTWetQ ,
                ind$env3_MTDQ ,
                ind$env4_PDM ,
                ind$env5_PwarmQ ,
                ind$env6_PWM)

# Perform qr decomposition to do the regression for all SNPs at the same time 
 Q <- qr(X, tol=1e-6)
# str(Q) run this line if you want to understand the structure of Q

# Get the matrix of regression coefficients
 Qr.coeff <- qr.coef(Q, G)
 # This matrix has each SNP in a column and the regression coefficients 
 # for that SNP corresponds to each environmental variable. 
 # This is the step that is not performed in the `vegan` package - 
 # the package skips directly to predicting the fitted values, 
 # on which the ordination is performed.

# Here is an example of regression coefficients for the first 10 SNPs:
 head(Qr.coeff[,1:10])

# Note that the regression coefficients for the first SNP from this 
# approach is exactly the same as from our model above:
 Qr.coeff[,1]
 coef(summary(mod))[,1]
```

We can visualize the regression coefficients with a heatmap. In this case, we know the causal loci in the simulations, so we will just visualize those loci.

This visualization illustrates how there are unique ways in which environments are combined in the model to predict the pattern at each SNP.
```{r}
# look at the range of slopes
summary(as.numeric(Qr.coeff[,which(muts$causal)]))

brks <- seq(-0.7, 0.7, by=0.05) #set the color scale

heatmap(t(Qr.coeff[,which(muts$causal)]),
        scale="none", col = cm.colors(length(brks)-1), breaks=brks)

```

Here is the information about the session when the tutorial was built:
```{r}
sessionInfo()
```

### 2023-02-07 choosing-population-genetics.Rmd
caching choosing pop gen info page, update index and yaml files for rda tutorial, fixed few typos rda tutorial

@@
 #library(knitr)
-knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
 library(knitcitations)
@@
                fig.height = 5,
-               cache = FALSE)
+               cache = TRUE)
 cite_options(citation_format = "pandoc", max.names = 3, style = "html",

### 2023-02-07 index.Rmd
caching choosing pop gen info page, update index and yaml files for rda tutorial, fixed few typos rda tutorial

@@
 * [Poolseq](poolseq.html)
+* [Redundancy Analysis (RDA) Trait Prediction](rda_tutorial/RDAtraitPredictionTutorial.html)
 

### 2023-02-07 rda_tutorial/RDAtraitPredictionTutorial.Rmd
caching choosing pop gen info page, update index and yaml files for rda tutorial, fixed few typos rda tutorial

@@
 date: "2023-01-20"
-output:
-  pdf_document: default
-  html_document: default
 ---
@@
 
-knitr::opts_knit$set(root.dir = "~/Documents/GitHub/MVP-NonClinalAF/tutorial") 
 #change code to your working directory
+#knitr::opts_knit$set(root.dir = "~/Documents/GitHub/MVP-NonClinalAF/tutorial") 
+knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, cache = TRUE)
 
@@
 
-## RDA trait prediction funciton
+## RDA trait prediction function
 
@@
   #rdaobj is RDA object
-  #envi row is the row of the environment in the biplot output
+  #env_row is the row of the environment in the biplot output
   #K is the number of RDA axes
@@
 
-Scale the environmental variables to have a mean of 0 and standard deviation of 1
+Scale the environmental variables to have a mean of 0 and standard deviation of 1.
 ```{r example}
@@
 # Make the trait prediction for MAT (1st row in biplot output)
-K = 3 # use 3 RDA axes to make the trait prediciton
+K = 3 # use 3 RDA axes to make the trait prediction
 
@@
 Note that the `predict` function and it's variations in the R package `vegan` do not make the same kind of predictions as `rda_trait_pred`.
-Here are the types of outputs produced by the function `predict` and it's variations:
+Here are the types of outputs produced by the function `predict` and its variations:
 
@@
 
-  # As a side note, it outputs the same thing as the `fitted` funciton
+  # As a side note, it outputs the same thing as the `fitted` function
   fitted_values_predict2 <- fitted(rdaout)
@@
 
-The `predict` function and it's variations make predictions in RDA space, and therefore do not output the same kind of predictions as `RDA_trait_predict` and Equation 1 in the paper.
+The `predict` function and its variations make predictions in RDA space, and therefore do not output the same kind of predictions as `rda_trait_predict` and Equation 1 in the paper.
 
@@
 
+dev.new()
 heatmap(t(Qr.coeff[,which(muts$causal)]),

### 2023-02-08 rda_tutorial/RDAtraitPredictionTutorial.Rmd

@@
 author: "KE Lotterhos"
-date: "2023-01-20"
+date: "2023-02-08"
+output: pdf_document
 ---
 
-This tutorial accompanies the paper "The paradox of adaptive trait clines with non-clinal patterns in the underlying genes"
+This tutorial accompanies the paper "The paradox of adaptive trait clines with non-clinal patterns in the underlying genes" published in PNAS [add link].
+
+# Abstract
+
+Multivariate climate change presents an urgent need to understand how species adapt to complex environments.
+Population genetic theory predicts that loci under selection will form monotonic allele frequency clines with their selective environment, which has led to the wide use of genotype-environment associations (GEAs).
+However, the accuracy of GEA methods to identify adapted loci is limited, as shown in the main paper.
+
+This tutorial shows how to apply a novel extension of multivariate ordination, which accurately predicted individual multivariate traits from genotype and environmental data on simulated data regardless of whether inference from GEAs was accurate.
+
+# Install packages
 
@@
 
-Load the libraries and set your working directory:
+## Load the libraries:
 
-```{r setup}
+```{r setup, eval=TRUE, echo=FALSE}
 
-libraries_needed <- c("vegan", "LEA", "lfmm")
+libraries_needed <- c("vegan", "LEA", "lfmm", "gplots")
 
@@
 
-#change code to your working directory
-#knitr::opts_knit$set(root.dir = "~/Documents/GitHub/MVP-NonClinalAF/tutorial") 
-knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, cache = TRUE)
-
+knitr::opts_chunk$set(message = FALSE, warning = FALSE, cache = TRUE) 
 ```
 
-## Load the data
+Don't forget to set your working directory!
+
+# Background on the simulation
+
+This data was simulated in SLiM and is associated with the complex multivariate simulation presented in Figure 6 in the paper.
+Briefly, a non-Wright-Fisher model was simulated on a landscape with 6 environmental variables that reflect different aspects of thermal stress and precipitation in British Columbia.
+The simulation included 6 environmental traits, each of which adapted to a different environmental variable.
+
+The six environmental variables are based on real data from western Canada and are shown below, clockwise from upper right: Clockwise from upper left: precipitation of driest month, precipitation of warmest quarter, mean annual temperature, precipitation of wettest month, mean temperature of wettest quarter, mean temperature of driest quarter.
+Each small square is a simulated individual, with its color representing its trait value in that environment.
+
+![](multivar.png)
+
+# Load the data
 
@@
 * A table with information about SNPs (each SNP in a row)
-
-This data was simulated in SLiM and is associated with the complex multivariate simulation presented in Figure 6 in the paper.
-Briefly, a non-Wright-Fisher model was simulated on a landscape with 6 environmental variables that reflect different aspects of thermal stress and precipitation in British Columbia.
-The simulation included 6 environmental traits, each of which adapted to a different environmental variable.
 
@@
 
-## RDA trait prediction function
+# RDA trait prediction function
 
@@
 
-### Understanding how the RDA is built on multiple regressions
+# Understanding how the RDA is built on multiple regressions
 
@@
 This visualization illustrates how there are unique ways in which environments are combined in the model to predict the pattern at each SNP.
-```{r}
-# look at the range of slopes
+
+```{r, fig.height=7, fig.width=8}
+dim(Qr.coeff)
+colnames(Qr.coeff) <- muts$mutname
+
+# look at the range of coefficients in the multiple regression model
 summary(as.numeric(Qr.coeff[,which(muts$causal)]))
@@
 
-dev.new()
-heatmap(t(Qr.coeff[,which(muts$causal)]),
-        scale="none", col = cm.colors(length(brks)-1), breaks=brks)
+heatmap.2(t(Qr.coeff[,which(muts$causal)]),
+        scale="none", 
+        col = cm.colors(length(brks)-1), 
+        breaks=brks,
+        dendrogram = "column",
+        Rowv=FALSE, #set this to "TRUE" if you would like to see which groups
+        trace="none",
+        key.title = "Coefficient in multiple\nregression model",
+        ylab="SNPs",
+        cexCol=1)
+```
 
+In the above heatmap, each row is a SNP.
+The SNPs are named according to their linkage group (1 through 10) and cumulative position in the genome (e.g.
+9-448632 is on the 9th linkage group).
+Each linkage group was 50,000 bases long, so the cumulative position ranges from 1 to 500,000 over the 10 linkage groups.
+Each column in the heatmap is an environment, which is an explanatory variable in the model.
+
+The color of the heatmap cells for a SNP shows the coefficients in the multiple regression model for each corresponding environment.
+In other words, the colors show how environments are combined in a multiple regression model to predict the patterns at that SNP on the landscape.
+
+If we want to visualize clusters of SNPs that have similar coefficients in the multiple regression model, we can allow for clustering in the heatmap:
+
+```{r, fig.height=7, fig.width=8}
+heatmap.2(t(Qr.coeff[,which(muts$causal)]),
+        scale="none", 
+        col = cm.colors(length(brks)-1), 
+        breaks=brks,
+        dendrogram = "both",
+        Rowv=TRUE, 
+        trace="none",
+        key.title = "Coefficient in multiple\nregression model",
+        ylab="SNPs",
+        cexCol=1)
 ```

### 2023-02-08 index.Rmd
changes to yml and index for rda tutorial

--- 2022-10-27 index.Rmd
@@
 * [Poolseq](poolseq.html)
+* [Redundancy Analysis (RDA) Trait Prediction](RDAtraitPredictionTutorial.html)
 

### 2023-02-08 submissions_instructions.Rmd
Push Sam's edits to submission instruction .Rmd

@@
 title: "Instructions for developing new pages for the MarineOmics webpage"
-bibliography: RCN_site-Nov05.bib
+bibliography: "RCN_site-Nov05.bib"
 ---
@@
 
-This is a guide for creating a new page for the website.
+This is a guide for creating a new page for MarineOmics.github.io.
 Of course, the website managers will work with anyone who develops new content to help develop and publish your page, but this can be used as a resource as you develop your content.
@@
 Of course, do not hesitate to reach out with any other questions or comments as well.
-
-\* Even if you comfortable with publishing websites in Github/RMarkdown, please contact Jason Johns (jasonjohns[at]ucsb.edu) before publishing any new pages so we can ensure smooth incorporation into the website.
+Please also reach out to him before publishing any new pages.
 
@@
 In short, each page of the website is rendered as an .html from an RMarkdown (.Rmd) document.
-The two most important files for publishing the general website are the .site_yml ("the yaml file") and the index.Rmd/index.html.
+The two most important files for publishing the general website are the _site.yml and the index.Rmd/index.html.
 
-**.site_yml**: the 'under the hood' backbone of the website, where things like the visual appearance of the website and the navigation bar are built
+**_site.yml**: the 'under the hood' backbone of the website, where things like the visual appearance of the website and the navigation bar are built
 
@@
 ### **code chunks**
-Below is a basic code chunk which can be inserted by clicking the green +C button on the RStudio toolbar or typing the keyboard shortcut: <br>
+Below is a basic R code chunk which can be inserted by clicking the green +C button on the RStudio toolbar or typing the keyboard shortcut: <br>
 
@@
 ```{r, echo = TRUE, include = TRUE} `r ''`
+
+#Here's a simple linear model in R as an example
+my_lm <- lm(y ~ x, data = df)
 
@@
 
+# Here's a simple line of bash as an example
+mkdir ~/Path/To/My/NewDir
+
 ```
@@
 You can add Bibtex references to the file RCN_site-Nov05.bib, and then refer to them in your RMarkdown by adding bibliography: RCN_site-Nov05.bib to the header and citing them in the RMarkdown document.
-Check out existing pages for examples.
+Let's say you have a paper you want to cite by LastName et al.
+2023 and the paper's ID in your .bib file is 'LastName2023-xy'.
+
+````
+We can reference this paper in-text by adding [@LastName2023-xy] to the end of a sentence. This will produce a reference such as the one below for Tripp et al. 2017.
+````
 

### 2023-02-08 index.Rmd
changed the index file to not yet link to the rda analysis until it's totally ready

--- 2022-10-27 index.Rmd
@@
 * [Poolseq](poolseq.html)
+* Redundancy Analysis (RDA) Trait Prediction (coming soon!)
 

### 2023-02-08 submissions_instructions.Rmd
incoroprated Sam's edits to submissions instructions

@@
 2023 and the paper's ID in your .bib file is 'LastName2023-xy'.
+We can reference this paper in-text by adding [\@LastName2023-xy] to the end of a sentence.
+This will produce a reference such as the one below for Tripp et al.
+2017.
 
+*syntax in .Rmd **(not in a code chunk)***:
 ````
-We can reference this paper in-text by adding [@LastName2023-xy] to the end of a sentence. This will produce a reference such as the one below for Tripp et al. 2017.
+RADseq can produce insightful analyses even in the absence of data, if missing values are handled properly [@Tripp2017-cp].
 ````
 
+*knitted output*: <br>
 RADseq can produce insightful analyses even in the absence of data, if missing values are handled properly [@Tripp2017-cp].

### 2023-02-10 DGE_comparison_v2.Rmd
re-rendered website (minus some cached files too large for now) with katie's rda tutorial page

--- 2022-07-25 DGE_comparison_v2.Rmd
@@
 #knitr::opts_knit$set(root.dir = '~/Documents/GitHub/DGE_comparison/')
+knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
 
@@
 #invisible(lapply(c( "tidyverse", "ape", "vegan", "GGally",
-                   #, "rgl", "adegenet", "MASS",
+                   #"rgl", "adegenet", "MASS",
                    #"data.table", "plyr", "lmtest", "reshape2", "Rmisc", "lmerTest","statmod"),
@@
                      #install.packages(p)
-                  # }
-                  # library(p, character.only=TRUE)
+                   #}
+                   #library(p, character.only=TRUE)
                  #}))

### 2023-02-10 Funct_Enrich_files/Funct_Enrich.Rmd
re-rendered website (minus some cached files too large for now) with katie's rda tutorial page

---
title: 'Functional Enrichment Tests: Intro and methods comparison'
subtitle: '*Thea Gessler*, *Samuel N. Bogan*, and *Steven Roberts*'
bibliography: Funct_Enrich.bib
---

# Getting to Enrichment Analysis

Finding your subset of genes of interest

* Differential expression
* Clustering
* WGCNA

<br>

# Introduction and Considerations

* The Gene Ontology (Thea)
* Annotation and filtration decisions (Sam) - include key terms
* Overrepresentation vs Enrichment (Thea)
* Explanation of statistics (Sam) - include key terms

<br>

# Tools and Approaches

* Ontologizer (command line) (Thea)
* Panther (online GUI) - endorsed by The Gene Ontology (Thea)
* R...Mann-Whitney U Test and Fisher’s Exact Test (Sam)

<br>

# What to Do with the Output

* Visualizing results
* Interpreting results - gather consensus advice 
* R...Mann-Whitney U Test and Fisher’s Exact Test (Sam)

<br>

# Formatting Examples

<br>

## Example of table syntax

| Term A | Term B | Term C | Term D | Term E | Term F | Term G |
| :--- | :--- | :--- | :--- | :--- | :--- | :--- |
| xxx | xxx | xxx | xxx | xxx | xxx | xxx |  
| yyy | yyy | yyy | yyy | yyy | yyy | yyy | 
| zzz | zzz | zzz | zzz | zzz | zzz | zzz |

<br>

## Example of how to insert an image

```{r, echo = FALSE, out.height = "40%", fig.align = "center", }

knitr::include_graphics("images/Rivera_etal_fig.png")

```

## Example of how to write and print an equation

For illustrative purposes, here is a simple and famous equation

$$
E = mc^2
$$

<br>

As an example of a more complex equation, here is an equation for the t-statistic

$$
t = \frac{\overline{x_{1}}-\overline{x_{2}}}
{\sqrt{(s^2(
\frac{1}{n_{1}} + \frac{1}{n_{2}}
))}}
$$

## Example of how to cite a reference from your .bib in-text

Functional enrichment tests applied to mRNA analyses often impose significant biases [@Bleazard2015-vn].

```{r setup, include = FALSE}

# Set root directory... in this draft, ensure that you have the same path
knitr::opts_knit$set(root.dir = '~/Documents/GitHub/marineomics.github.io/')

```

<br>

# Let's get started

Below is an example of what should be the first, visible code chunk

```{r, results = FALSE, message = FALSE, warning = FALSE}

## Unhash and run the code below if you believe you may need to install the packages loaded below
# Edit code below to list packages that are used in this draft... at present these are just examples

if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install(c("DESeq2","edgeR","arrayQualityMetrics"))

# Load packages
library(tidyverse)

```

# References

Below are the markdown-formatted references for our document derived from the

### 2023-02-10 RDAtraitPredictionTutorial.Rmd
re-rendered website (minus some cached files too large for now) with katie's rda tutorial page

--- 2023-02-08 rda_tutorial/RDAtraitPredictionTutorial.Rmd
@@
 date: "2023-02-08"
-output: pdf_document
+output:
+  html_document:
+    df_print: paged
 ---
@@
 install.packages("lfmm")
+install.packages("gplots")
 
@@
 The six environmental variables are based on real data from western Canada and are shown below, clockwise from upper right: Clockwise from upper left: precipitation of driest month, precipitation of warmest quarter, mean annual temperature, precipitation of wettest month, mean temperature of wettest quarter, mean temperature of driest quarter.
-Each small square is a simulated individual, with its color representing its trait value in that environment.
+Background colors correspond to the optimum trait value on each landscape, and each small square is a simulated individual, with its color representing its trait value in that environment.
 
@@
 
-G <- read.table(unz("Genotypes.txt.zip", "Genotypes.txt"))
+G <- read.table(unz("rda_tutorial/Genotypes.txt.zip", "Genotypes.txt"))
 dim(G) # 1000 individuals and 26371 loci
 
-ind <- read.table("Individuals.txt", header=TRUE)
+ind <- read.table("rda_tutorial/Individuals.txt", header=TRUE)
 dim(ind) #corresponds to rows in G
@@
 
-muts <-  read.table("SNPs.txt", header=TRUE)
+muts <-  read.table("rda_tutorial/SNPs.txt", header=TRUE)
 dim(muts) #corresponds to columns in G

### 2023-02-13 submissions_instructions.Rmd
KS edits to submission instructions

@@
 bibliography: "RCN_site-Nov05.bib"
+editor_options: 
+  markdown: 
+    wrap: 72
 ---
@@
 
-This is a guide for creating a new page for MarineOmics.github.io.
-Of course, the website managers will work with anyone who develops new content to help develop and publish your page, but this can be used as a resource as you develop your content.
-In this guide are general instructions as well as examples of how to build specific features.
-If there is anything else you would like to see on this guide please contact Jason Johns (jasonjohns[at]ucsb.edu).
-Of course, do not hesitate to reach out with any other questions or comments as well.
-Please also reach out to him before publishing any new pages.
+If you are interested in submitting new content to the
+MarineOmics.github.io website, this is the guide for you! While the
+website managers will work with you to help prepare and publish your new
+page, this guide can be a useful resource as you develop your content.
+Below we provide general instructions as well as examples of how to
+build specific features (eg., tables, code chunks). Prior to publishing
+new pages, please contact the current Website Manager Jason Johns
+(jasonjohns[at]ucsb.edu). Do not hesitate to reach out with any other
+questions, comments, or features you'd like added to this guide.
 
-This website is published through Github Pages directly from the [MarineOmics github repository](https://github.com/MarineOmics/marineomics.github.io).
-In short, each page of the website is rendered as an .html from an RMarkdown (.Rmd) document.
-The two most important files for publishing the general website are the _site.yml and the index.Rmd/index.html.
+The MarineOmics website is published through Github Pages directly from
+the [MarineOmics github
+repository](https://github.com/MarineOmics/marineomics.github.io). In
+short, each page of the website is rendered as an .html from an
+RMarkdown (.Rmd) document. The two most important files for publishing
+the general website are the \_site.yml and the index.Rmd/index.html.
 
-**_site.yml**: the 'under the hood' backbone of the website, where things like the visual appearance of the website and the navigation bar are built
+\*\*\_site.yml\*\*: the 'under the hood' backbone of the website, where
+things like the visual appearance of the website and the navigation bar
+are specified
 
-**index.html**: rendered from index.Rmd - the home page from which all other pages branch
+**index.html**: rendered from index.Rmd - the home page from which all
+other pages branch
 
-\* The website manager will help modify the index.html and .site_yml files for your page when the time comes, so you don't need to know exactly how they work, but it's good to know what they are at a basic level.
+\* The website manager will modify the index.html and .site_yml files
+for your page when you are ready to publish it, so you don't need to
+know exactly how they work, but it's good to know what they are at a
+basic level.
 
@@
 # R Markdown basics
-The rendered .Rmd and .html files for each page on this site can be seen in the [MarineOmics github repository](https://github.com/MarineOmics/marineomics.github.io).
-If you don't have much experience with RStudio, it's quite intuitive with a shallow learning curve and many existing online resources for using it.
-It is possible to use .Rmd files in base R, but we recommend using RStudio.
 
-*Note: it's a good idea to look at existing .Rmd scripts (including the one that codes for this page) to see examples of how a particular feature can be accomplished, such as citing references, linking a website, building a table, incorporating figures, etc.
-We will explicitly go over some of those features below, but there are many other bells and whistles that can be explored through online resources.*
+The rendered .Rmd and .html files for each page on this site can be seen
+in the [MarineOmics github
+repository](https://github.com/MarineOmics/marineomics.github.io). It is
+possible to use .Rmd files in base R, but we recommend using RStudio. If
+you don't have much experience with RStudio, it's quite intuitive with a
+shallow learning curve and many existing online resources for using it.
+It is possible to use .Rmd files in base R, but we recommend using
+RStudio.
+
+*Note: it's a good idea to look at existing .Rmd scripts (including [the
+one that codes for this
+page](https://github.com/MarineOmics/marineomics.github.io/blob/main/submissions_instructions.Rmd))
+to see examples of how a particular feature can be accomplished, such as
+citing references, linking a website, building a table, incorporating
+figures, etc. We will explicitly go over some of those features below,
+but there are many other bells and whistles that can be explored through
+online resources.*
 
 Here are some general RMarkdown tutorials that we like to reference:
-<br>
-- [Chapter 10 from RMarkdown the Definitive Guide: Xi, Allaire, & Grolemund 2022](https://bookdown.org/yihui/rmarkdown/rmarkdown-site.html)
-<br>
-- [R Markdown Websites: Garrett Grolemund](https://garrettgman.github.io/rmarkdown/rmarkdown_websites.html)
+<br> - [Chapter 10 from RMarkdown the Definitive Guide: Xi, Allaire, &
+Grolemund
+2022](https://bookdown.org/yihui/rmarkdown/rmarkdown-site.html) <br> -
+[R Markdown Websites: Garrett
+Grolemund](https://garrettgman.github.io/rmarkdown/rmarkdown_websites.html)
 
@@
 ## Formatting basics
-One of the strengths of .Rmd documents is their versatility.
-They can hold everything from code (not only in R but in other languages), formatted text, figure outputs, and more.
-Basically, code must be enclosed within 'code chunks' and anything outside of code chunks will be read as text, similar to a basic text file.
+
+One of the strengths of .Rmd documents is their versatility. They can
+hold everything from code (not only in R but in other languages),
+formatted text, figure outputs, and more. Basically, code must be
+enclosed within 'code chunks' and anything outside of code chunks will
+be read as text, similar to a basic text file.
 
 ### **code chunks**
-Below is a basic R code chunk which can be inserted by clicking the green +C button on the RStudio toolbar or typing the keyboard shortcut: <br>
+
+Below is a basic R code chunk which can be inserted by clicking the
+green +C button on the RStudio toolbar or typing the keyboard shortcut:
+<br>
 
 -   macOS: command + option + I <br>
-- Windows: control + alt + I
+-   Windows/Linux: control + alt + I
 
-````
+<!-- -->
+
     ```{r, echo = TRUE, include = TRUE} `r ''`
@@
     ```
-````
-Whether code chunks and/or their output are included in the knitted .html can be modified with the logical commands "echo" and "include". <br>
+
+Whether code chunks and/or their output are included in the knitted
+.html can be modified with the logical commands "echo" and "include".
+<br>
 
@@
 
-\* Other options exist for code chunk output which can be explored online and in other .Rmd scripts in this repository.
+\* Other options exist for code chunk output which can be explored
+online and in other .Rmd scripts in this repository.
 
@@
 #### *global code chunk options*
-This is the chunk where we establish global settings for the entire document.
-This is done with the command from the knitr package "opts_chunk$set" as shown in the example below.
-This command is typically included in the first chunk, where we also load our packages and adjust the settings for citations.
-When a new .Rmd is made, this command is automatically included in the first chunk, which is named "setup" by default.
-The name of a code chunk can be specified after the r in ```{r ____}.
+
+This is the chunk where we establish global settings for the entire
+document. This is done with the command from the knitr package
+"opts_chunk\$set" as shown in the example below. This command is
+typically included in the first code chunk, where we also load our page
+formatting packages and adjust the settings for citations. When a new
+.Rmd is made, this command is automatically included in the first chunk,
+which is named "setup" by default. The name of a code chunk can be
+specified after the r in \`\`\`{r \_\_\_\_}.
 
 <a id="cite_options"></a>
-````
+
     ```{r setup} `r ''`
     knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, cache = TRUE) 
+    # load page formatting packages
     library(knitr)
@@
     ```
-````
-\* Note the "cache = " option.
-When set to "true" R will cache all of the output created from knitting the RMarkdown document.
-This is especially useful for pages with heavy computation that take a long time to knit.
+
+\* Note the "cache =" option. When set to "true" R will cache all of the
+output created from knitting the RMarkdown document. This is especially
+useful for pages with heavy computation that take a long time to knit.
 
@@
 #### *including code from another language*
-We can also run code such as bash in an R script by modifying the code chunk header as below: <br>
 
-````
-```{bash, eval = FALSE} `r ''`
+In addition to R code, we can also run code such as bash shell script in
+the R markdown file by modifying the code chunk header as below: <br>
+
+    ```{bash, eval = T} `r ''`
 
     # Here's a simple line of bash as an example
-mkdir ~/Path/To/My/NewDir
+    echo "Hi world"
 
     ```
-````
+
 <br>
@@
 ### **formatting text**
-Options abound for formatting text in .Rmd documents using [Pandoc Markdown](https://pandoc.org/MANUAL.html), and many of these can be found with a quick search, but we will go over a few basics here.<br>
 
-- As with writing code, any text symbol that codes for something, such as a \# or a \* must be escaped with a backslash (\\) if you just want to use the symbol itself. <br>
-- **\#**: *Outside of a code chunk*, pound signs dictate the size and organizational level of the text.
-A single \# will make large text and a main section in the navigation window.
-Two \#\# will make slightly smaller text and a sub-section in the navigation window.
+Options abound for formatting text in .Rmd documents using [Pandoc
+Markdown](https://pandoc.org/MANUAL.html), and many of these can be
+found with a quick search, but we will go over a few basics here.<br>
+
+-   As with writing code, any text symbol that codes for something, such
+    as a \# or a \* must be escaped with a backslash (\\) if you just
+    want to use the symbol itself. <br>
+-   **\#**: *Outside of a code chunk*, pound signs dictate the size and
+    organizational level of the text. A single \# will make large text
+    and a main section in the navigation window. Two \## will make
+    slightly smaller text and a sub-section in the navigation window.
     <br>
-  + For example, the "RMarkdown formatting" text and category was created with "\# RMarkdown formatting" while the "code chunks" subcategory was created with "\#\# code chunks."<br>
-- **\***: Asterisks are used to bold and italicize when surrounding texts. Use \*\***two asterisks to bold**\*\* and \**one asterisk*\* to italicize.<br>
-- **\<br>**: adds a line break <br>
-- **\<hr />**: adds a horizontal line across the page
+    -   For example, the "RMarkdown formatting" text and category was
+        created with "\# RMarkdown formatting" while the "code chunks"
+        subcategory was created with "\## code chunks."<br>
+-   **\***: Asterisks are used to bold and italicize when surrounding
+    texts. Use \*\***two asterisks to bold**\*\* and \**one asterisk*\*
+    to italicize.<br>
+-   **\<br\>**: adds a line break <br>
+-   **\<hr /\>**: adds a horizontal line across the page
 
@@
 ### **in-text citations from your .bib**
-In-text citations can be made throughout the document that will automatically add a reference to your reference section via the following steps: <br>
 
-- add your citations to the RCN_site-Nov05.bib document via your reference manager
-- add the line 'bibliography: RCN_site-Nov05.bib' to your header at the top of the .Rmd document, which looks like this:
-````
+In-text citations can be made throughout the document that will
+automatically add a reference to your reference section via the
+following steps: <br>
+
+-   add your citations to the RCN_site-Nov05.bib document by exporting
+    BibTeX-formatted citations from your reference manager
+-   add the line 'bibliography: RCN_site-Nov05.bib' to your header at
+    the top of the .Rmd document, which looks like this:
+
+<!-- -->
+
     ---
@@
     ---
-````
-- include the 'cite-options' chunk in your 'setup' code chunk. See the initial ['r setup' chunk](#cite_options) above for the syntax of the 'cite_options' command.
 
-You can add Bibtex references to the file RCN_site-Nov05.bib, and then refer to them in your RMarkdown by adding bibliography: RCN_site-Nov05.bib to the header and citing them in the RMarkdown document.
-Let's say you have a paper you want to cite by LastName et al.
-2023 and the paper's ID in your .bib file is 'LastName2023-xy'.
-We can reference this paper in-text by adding [\@LastName2023-xy] to the end of a sentence.
-This will produce a reference such as the one below for Tripp et al.
-2017.
+-   include the 'cite-options' chunk in your 'setup' code chunk. See the
+    initial ['r setup' chunk](#cite_options) above for the syntax of the
+    'cite_options' command.
 
-*syntax in .Rmd **(not in a code chunk)***:
-````
+After following these step, you can cite them in the RMarkdown document.
+Let's say you have a paper you want to cite by LastName et al. 2023 and
+the paper's ID in your .bib file is 'LastName2023-xy'. We can reference
+this paper in-text by adding [\@LastName2023-xy] to the end of a
+sentence. This will produce a reference such as the one below for Tripp
+et al. 2017.
+
+*syntax in .Rmd text **(not in a code chunk)***:
+
     RADseq can produce insightful analyses even in the absence of data, if missing values are handled properly [@Tripp2017-cp].
-````
 
-*knitted output*: <br>
-RADseq can produce insightful analyses even in the absence of data, if missing values are handled properly [@Tripp2017-cp].
+*knitted output*: <br> RADseq can produce insightful analyses even in
+the absence of data, if missing values are handled properly
+[@Tripp2017-cp].
 
@@
 ### **installing & loading packages**
-If you will be using any R packages for your information page or tutorial, it's a good idea to highlight them toward the beginning of your page, including code for installing them.
-Here's an example from the RADseq tutorial:<br>
-<br>
-*If you would like to run the R code examples that are scattered throughout the guide (recommended but not required!), you will need to install some R packages. Only need to run this code once:*  
+
+If you will be using any R packages or external software for your
+content, it's a good idea to highlight them toward the beginning of your
+page, including code for installing them. Here's an example from the
+RADseq tutorial: <br> <br> *If you would like to run the R code examples
+that are scattered throughout the guide (recommended but not required!),
+you will need to install some R packages. Only need to run this code
+once:*
+
 ```{r, eval=F}
@@
 
-<br>
+<br> *These external software are referenced in the tutorial:* <br>
+
+-   [vcftools](https://vcftools.github.io/man_latest.html)
+-   [gnuplot](https://riptutorial.com/gnuplot/example/11275/installation-or-setup)
 
 ### **caching**
-If your page includes any code that takes a long time to run, please set "cache = TRUE" in the knitr::opts_chunk$set command in your global code chunk.
+
+If your page includes any code that takes a long time to run, please set
+"cache = TRUE" in the knitr::opts_chunk\$set command in your global code
+chunk.
 
@@
 ### **insert a table**
+
 *text in .Rmd **(not in a code chunk)***:
-````
-| Term A | Term B | Term C | Term D | Term E | Term F | Term G |
-| :--- | :--- | :--- | :--- | :--- | :--- | :--- |
-| xxx | xxx | xxx | xxx | xxx | xxx | xxx |  
-| yyy | yyy | yyy | yyy | yyy | yyy | yyy | 
-| zzz | zzz | zzz | zzz | zzz | zzz | zzz | 
-````
-
-<br>
-
-*knitted output*:
 
@@
 
+*knitted output*:
+
+| Term A | Term B | Term C | Term D | Term E | Term F | Term G |
+|:-------|:-------|:-------|:-------|:-------|:-------|:-------|
+| xxx    | xxx    | xxx    | xxx    | xxx    | xxx    | xxx    |
+| yyy    | yyy    | yyy    | yyy    | yyy    | yyy    | yyy    |
+| zzz    | zzz    | zzz    | zzz    | zzz    | zzz    | zzz    |
+
+<br>
+
+You can generate a Markdown-formatted table using a free online editor,
+such as [this
+one](https://tabletomarkdown.com/generate-markdown-table/). The
+[**kableExtra**](https://bookdown.org/yihui/rmarkdown-cookbook/kableextra.html)
+package has a lot of functionality for fancier tables.
+
+<br>
+
 ### **insert an image**
+
+To insert an image, you need the image file to also be uploaded to the
+Github repositority. We store the images hosted on the website in the
+`images/` folder. For any image, please make sure you have permission to
+share the image and that it is appropriately cited.
+
 *code chunk in .Rmd*:
 
-````
-```{r, echo = TRUE, include = TRUE, out.height = "40%", fig.align = "center"} `r ''`
+    ```{r, echo = TRUE, include = TRUE, out.height = "40%", fig.align = "center", fig.cap="Figure from Rivera et al. 2021"} `r ''`
 
@@
     ```
-````
 
@@
 *knitted output*:
-```{r, echo = TRUE, include = TRUE, out.height = "40%", fig.align = "center", }
+
+```{r, echo = TRUE, include = TRUE, out.height = "40%", fig.align = "center",fig.cap="Figure from Rivera et al. 2021" }
 
@@
 
+Equations follow LaTeX style formatting.
+
 For illustrative purposes, here is a simple and famous equation. <br>
 *syntax in .Rmd*:
-````
+
     $$
@@
     $$
-````
 
@@
 
-*knitted output*:
-$$
+*knitted output*: $$
 E = mc^2
-$$
-<br>
+$$ <br>
 
-As an example of a more complex equation, here is an equation for the t-statistic: <br>
-*syntax in .Rmd*:
-````
+As an example of a more complex equation, here is an equation for the
+t-statistic: <br> *syntax in .Rmd*:
+
     $$
@@
     $$
-````
 
-*knitted output*:
-$$
+*knitted output*: $$
 t = \frac{\overline{x_{1}}-\overline{x_{2}}}
@@
 
+After you have finished editing your R markdown file, you can "knit" it
+to an HTML file by selecting the Knit button at the top of RStudio. You
+can review this HTML file to get an idea of what the page will look like
+on the website (minus the navigation bar and other site-specific
+formatting). You will need to upload both the R markdown file and the
+HTML file to Github prior to you page being added to the website.
+
 <hr />

### 2023-02-13 contributions.Rmd
publishing submissions_instructions page to website and pushing some other cache files

--- 2022-10-27 contributors.Rmd
@@
 ---
-title: "Contributors"
+title: "Contributions"
 ---
@@
 ```
+# How to contribute new material to the website
+If you are interested in submitting new content to the
+MarineOmics.github.io website, we have a [guide](submissions_instructions.html) for you!
+While the website managers will work with you to help prepare and publish your new
+page, this guide can be a useful resource as you develop your content.
+The guide provides general instructions as well as examples of how to build specific features (eg., tables, code chunks).
+Prior to developing new pages, please contact the current Website Manager Jason Johns (jasonjohns[at]ucsb.edu).
+Do not hesitate to reach out with any other questions, comments, or features you'd like added to the guide.
 
-# Original Working Group Members
+# Original working group members
 

### 2023-02-13 index.Rmd
publishing submissions_instructions page to website and pushing some other cache files

--- 2023-02-08 index.Rmd
@@
 
+# Making contributions to the MarineOmics website
+We are excited to incorporate new material from contributors of various backgrounds, expertise, and experience.
+To date these tutorials, guidelines,  and discussions have been authored and led by graduate students and postdocs with contributions from scientists at all levels, including undergraduates, technicians, and faculty members.
+A diverse range of perspectives and backgrounds will make these tutorials more inclusive, informative, and useful for everyone in the marine-omics community.
+If you are interested in developing an information page or tutorial for the website,  we ask that you review our [page with instructions for submitting content](submissions_instructions.html) and contact our Website Manager Jason Johns (jasonjohns[at]ucsb.edu).
+
 # Internship 

### 2023-02-13 index.Rmd
minor change to index file to reference new section on contributions

@@
 MarineOmics aims to support robust genomic research in ecology and evolution by disseminating guidelines through this dynamic, open-sourced website addressing introductory and advanced topics of population- and functional genomics.
-*We are always looking for folks to join and write content for the website, if interested please [contact us](#contact)!*
+*We are always looking for folks to join and write content for the website, if interested please [contact us](#contact) and see the [below section](#contributions)!*
 
@@
 
-# Making contributions to the MarineOmics website
+# Making contributions to the MarineOmics website {#contributions}
 We are excited to incorporate new material from contributors of various backgrounds, expertise, and experience.

### 2023-02-13 index.Rmd
couple minor wording changes to index & contribution pages

@@
 
-# Making contributions to the MarineOmics website {#contributions}
+# Making contributions to the MarineOmics.github.io website {#contributions}
 We are excited to incorporate new material from contributors of various backgrounds, expertise, and experience.

### 2023-02-16 DNA-methylation.Rmd
started DNA methylation analysis tutorial

---
title: 'DNA Methylation Analysis'
subtitle: '*Steven Roberts*, and *Sam White* '
bibliography: xxxxx.bib
---

# Introduction
The purpose of this tutorial is to provide an example of taking Bisulfite treated DNA sequence data through identification of methylation at the loci and gene level. The principle behind creating bisulfite treated DNA libraries is that when cytosines are not methylated they will be converted to uracil, which ultimately result in thymine. When aligning to a reference genome the percent methylation at a given CpG loci is determined by examine the ratio to thymines and cytosines. For example if there

# Sequence Quality

### 2023-02-16 DNA-methylation.Rmd
methylation - quality

@@
 subtitle: '*Steven Roberts*, and *Sam White* '
-bibliography: xxxxx.bib
+bibliography: DNA-Methylation.bib
 ---
@@
 When aligning to a reference genome the percent methylation at a given CpG loci is determined by examine the ratio to thymines and cytosines.
-For example if there
+For example if there is a given CpG loci with 30% of the alignment with the cytosine on the reference being and thymine, one would consider this cytosine loci (CpG) to be 70% methylated.
 
 # Sequence Quality
+An example of what raw data might look like from a marine invertebrate is shown.
+_Note that DNA methylation patters are diverse across taxa with most marine invertebrates demonstrating a mosiac pattern and global methylation at CpG loci is approximately 15%._
+
+Following trimming one would expect to see something similar to this. This particular data set was trimmed with the following parameters:
+
+![bp-chart](images/meth-bp-chart.png)

### 2023-02-16 DNA-methylation.Rmd
added trim code

@@
 title: 'DNA Methylation Analysis'
-subtitle: '*Steven Roberts*, and *Sam White* '
+subtitle: '*Steven Roberts* and *Sam White* '
 bibliography: DNA-Methylation.bib
@@
 
+```
+/gscratch/srlab/programs/TrimGalore-0.4.5/trim_galore \
+--output_dir /gscratch/scrubbed/strigg/analyses/20200311/WGBS_MBD \
+--paired \
+--fastqc_args \
+--outdir /gscratch/scrubbed/strigg/analyses/20200311/WGBS_MBD/FASTQC \
+--threads 28 \
+--illumina \
+--clip_R1 10 \
+--clip_R2 10 \
+--three_prime_clip_R1 10 \
+--three_prime_clip_R2 10 \
+--path_to_cutadapt /gscratch/srlab/programs/miniconda3/bin/cutadapt \
+/gscratch/scrubbed/sr320/froger-raw/00_fastq/Meth17_R1_001.fastq.gz \
+```
+
 ![bp-chart](images/meth-bp-chart.png)
+
+# Read Alignment

### 2023-02-16 DNA-methylation.Rmd
taking through bismark

@@
 # Read Alignment
+The software used as part of this tutorial is **Bismark Bisulfite Mapper**.
+As with any software it is best to be famililar with the [User Manual](https://rawgit.com/FelixKrueger/Bismark/master/Docs/Bismark_User_Guide.html).
+Thus content below will provided with the assumption that the reader has read the manual and is meant to serve as guidance based on experience working with marine invertebrates.
+
+The first step in the process in preparing the genome. Example code:
+
+```
+# Directories and programs
+bismark_dir="/programs/Bismark-0.21.0"
+bowtie2_dir="/programs/bowtie2-2.3.4.1-linux-x86_64/"
+genome_folder="/where/the/fastafile/lives/"
+
+${bismark_dir}/bismark_genome_preparation \
+--verbose \
+--parallel 28 \
+--path_to_aligner ${bowtie2_dir} \
+${genome_folder}
+
+```
+
+You should expect to prepared genome with new directories indluding
+
+```
+./Bisulfite_Genome
+./Bisulfite_Genome/GA_conversion
+./Bisulfite_Genome/CT_conversion
+```
+
+For doing the alignment of the trim reads to the genome the following code structured is used
+
+```
+find ${reads_dir}*_R1_001_val_1.fq.gz \
+| xargs basename -s _R1_001_val_1.fq.gz | xargs -I{} ${bismark_dir}/bismark \
+--path_to_bowtie ${bowtie2_dir} \
+-genome ${genome_folder} \
+-p 4 \
+-score_min L,0,-0.6 \
+--non_directional \
+-1 ${reads_dir}{}_R1_001_val_1.fq.gz \
+-2 ${reads_dir}{}_R2_001_val_2.fq.gz \
+-o Mcap_tg
+```
+
+This will create bam files (sequence alignment files)
+
+Files are then deduplicated if they are whole genome bisulfite treated samples.
+This command will deduplicate the Bismark alignment BAM file and remove all reads but one which align to the the very same position and in the same orientation.
+This step is recommended for whole-genome bisulfite samples, but should not be used for reduced representation libraries such as RRBS, amplicon or target enrichment libraries.
+
+```
+find *.bam | \
+xargs basename -s .bam | \
+xargs -I{} ${bismark_dir}/deduplicate_bismark \
+--bam \
+--paired \
+{}.bam
+```
+
+Methylation levels are then extracted using the `bismark_methylation_extractor`. For example:
+
+```
+${bismark_dir}/bismark_methylation_extractor \
+--bedGraph \
+--counts \
+--comprehensive \
+--merge_non_CpG \
+--multicore 28 \
+--buffer_size 75% \
+*deduplicated.bam
+```
+
+This will create deduplicated.bismark.cov.gz, uncompressed is in this format.
+
+```
+NC_035784.1 141 141 37.5    3   5
+NC_035784.1 142 142 100 2   0
+NC_035784.1 155 155 70  7   3
+NC_035784.1 156 156 100 2   0
+NC_035784.1 291 291 0   0   2
+NC_035784.1 292 292 0   0   3
+NC_035784.1 313 313 0   0   1
+NC_035784.1 314 314 66.6666666666667    2   1
+NC_035784.1 470 470 66.6666666666667    4   2
+NC_035784.1 611 611 0   0   4
+```
+
+```
+<chromosome> <start position> <end position> <methylation percentage> <count methylated> <count unmethylated>
+```
+
+A genome-wide cytosine report ins generated, with this code as an example:
+
+```
+find *deduplicated.bismark.cov.gz \
+| xargs basename -s _trimmed_bismark_bt2.deduplicated.bismark.cov.gz \
+| xargs -I{} ${bismark_dir}/coverage2cytosine \
+--genome_folder ${genome_folder} \
+-o {} \
+--merge_CpG \
+--zero_based \
+{}_trimmed_bismark_bt2.deduplicated.bismark.cov.gz
+```
+
+Generating  a file `.CpG_report.merged_CpG_evidence.cov`
+
+```
+NC_035785.1 217 219 100.000000  17  0
+NC_035785.1 523 525 87.500000   7   1
+NC_035785.1 556 558 50.000000   5   5
+NC_035785.1 727 729 100.000000  16  0
+NC_035785.1 1330    1332    0.000000    0   2
+NC_035785.1 1403    1405    0.000000    0   2
+NC_035785.1 1494    1496    66.666667   2   1
+NC_035785.1 1747    1749    100.000000  8   0
+NC_035785.1 2024    2026    100.000000  24  0
+NC_035785.1 2054    2056    93.333333   14  1
+```
+
+From here it could be useful to do some simple file reorganizaion to obtain bedgraph or tab files for down stream analysis.
+
+```
+
+#creating bedgraphs post merge
+
+for f in *merged_CpG_evidence.cov
+do
+  STEM=$(basename "${f}" .CpG_report.merged_CpG_evidence.cov)
+  cat "${f}" | awk -F $'\t' 'BEGIN {OFS = FS} {if ($5+$6 >= 10) {print $1, $2, $3, $4}}' \
+  > "${STEM}"_10x.bedgraph
+done
+
+for f in *merged_CpG_evidence.cov
+do
+  STEM=$(basename "${f}" .CpG_report.merged_CpG_evidence.cov)
+  cat "${f}" | awk -F $'\t' 'BEGIN {OFS = FS} {if ($5+$6 >= 5) {print $1, $2, $3, $4}}' \
+  > "${STEM}"_5x.bedgraph
+done
+
+#creating tab files with raw count for glms
+
+for f in *merged_CpG_evidence.cov
+do
+  STEM=$(basename "${f}" .CpG_report.merged_CpG_evidence.cov)
+  cat "${f}" | awk -F $'\t' 'BEGIN {OFS = FS} {if ($5+$6 >= 10) {print $1, $2, $3, $4, $5, $6}}' \
+  > "${STEM}"_10x.tab
+done
+
+for f in *merged_CpG_evidence.cov
+do
+  STEM=$(basename "${f}" .CpG_report.merged_CpG_evidence.cov)
+  cat "${f}" | awk -F $'\t' 'BEGIN {OFS = FS} {if ($5+$6 >= 5) {print $1, $2, $3, $4, $5, $6}}' \
+  > "${STEM}"_5x.tab
+done
+```

### 2023-02-16 DNA-methylation.Rmd
meth ds file conversion

@@
 ```
+
+It is often useful to sort files for downstream analysis (eg methylkit, IGV)
+
+```
+# Sort files for methylkit and IGV
+
+find *.bam | \
+xargs basename -s .bam | \
+xargs -I{} ${samtools} \
+sort --threads 28 {}.bam \
+-o {}.sorted.bam
+
+# Index sorted files for IGV
+
+find *.sorted.bam | \
+xargs basename -s .sorted.bam | \
+xargs -I{} ${samtools} \
+index -@ 28 {}.sorted.bam
+```

### 2023-02-16 DNA-methylation.Rmd
meth subheadings
meth bedgraph disclaimer

@@
 ---
-title: 'DNA Methylation Analysis'
+title: 'DNA Methylation Analyses'
 subtitle: '*Steven Roberts* and *Sam White* '
@@
 
+# Methylation Quantification
 Methylation levels are then extracted using the `bismark_methylation_extractor`. For example:
@@
 This will create deduplicated.bismark.cov.gz, uncompressed is in this format.
+And while a bedGraph file is generated, in this workflow a bedGraph will be generated later that uses information from merged strands (as opposed to this output).
 
@@
 
+Column organization
 ```
@@
 ```
+
+#Generating files for downstream analysis
 

### 2023-02-16 DNA-methylation.Rmd
meth tightening code

@@
 Column organization
+
 ```
@@
 ```
-
 #creating bedgraphs post merge
-
 for f in *merged_CpG_evidence.cov
@@
 #creating tab files with raw count for glms
-
 for f in *merged_CpG_evidence.cov
@@
 # Sort files for methylkit and IGV
-
 find *.bam | \
@@
 # Index sorted files for IGV
-
 find *.sorted.bam | \

### 2023-02-16 DNA-methylation.Rmd
meth - formatting

@@
 
-This will create deduplicated.bismark.cov.gz, uncompressed is in this format.
+This will create a file with the suffix `deduplicated.bismark.cov.gz` (example shown below in uncompressed format).
 And while a bedGraph file is generated, in this workflow a bedGraph will be generated later that uses information from merged strands (as opposed to this output).
@@
 
-Column organization
+column organization of the file
 
@@
 
-A genome-wide cytosine report ins generated, with this code as an example:
+A genome-wide cytosine report is generated, with this code as an example:
 
@@
 
-#Generating files for downstream analysis
+# Generating files for downstream analysis
 
-From here it could be useful to do some simple file reorganizaion to obtain bedgraph or tab files for down stream analysis.
+From here it could be useful to do some simple file reorganizaion to obtain bedgraph or tab files for downstream analysis (eg bedtools, GLMs).
 
@@
 
-It is often useful to sort files for downstream analysis (eg methylkit, IGV)
+It is sometimes useful to sort BAM files for downstream analysis (eg methylkit, IGV)
 

### 2023-02-17 DNA-methylation.Rmd
open authors
started adding references

@@
+```{r,message = FALSE,echo = FALSE}
+library(knitr)
+knitr::opts_chunk$set(echo = TRUE)
+library(knitcitations)
+library(kableExtra)
+opts_chunk$set(fig.width = 10,
+               fig.height = 5,
+               cache = FALSE)
+cite_options(citation_format = "pandoc", max.names = 3, style = "html", 
+             hyperlink = "to.doc")
+```
+
 ---
 title: 'DNA Methylation Analyses'
-subtitle: '*Steven Roberts* and *Sam White* '
+subtitle: '*Steven Roberts*, *Sam White*, and ............ '
 bibliography: DNA-Methylation.bib
@@
 For example if there is a given CpG loci with 30% of the alignment with the cytosine on the reference being and thymine, one would consider this cytosine loci (CpG) to be 70% methylated.
+
+For a good review of epigenetic mechanisms including DNA methylation in marine organisms see @doi:10.1146/annurev-marine-010318-095114.
+@https://doi.org/10.1111/1755-0998.13542 offers a comparison of three methods for quantifying DNA methylation at single base-pair resolution using whole genome bisulfite sequencing (WGBS), reduced representation bisulfite sequencing (RRBS), and methyl-CpG binding domain bisulfite sequencing (MBDBS).
+The tutorial below is based on WGBS, though the general workflow would be consistent.
 
@@
 
-# Generating files for downstream analysis
+# File Conversions
 
@@
 ```
+
+# References

### 2023-02-17 DNA-methylation.Rmd
fmt

@@
 # Introduction
+
 The purpose of this tutorial is to provide an example of taking Bisulfite treated DNA sequence data through identification of methylation at the loci and gene level.
@@
 # Sequence Quality
+
 An example of what raw data might look like from a marine invertebrate is shown.
-_Note that DNA methylation patters are diverse across taxa with most marine invertebrates demonstrating a mosiac pattern and global methylation at CpG loci is approximately 15%._
+*Note that DNA methylation patters are diverse across taxa with most marine invertebrates demonstrating a mosiac pattern and global methylation at CpG loci is approximately 15%.*
 
@@
 
-```
     /gscratch/srlab/programs/TrimGalore-0.4.5/trim_galore \
@@
     /gscratch/scrubbed/sr320/froger-raw/00_fastq/Meth17_R1_001.fastq.gz \
-```
 
@@
 # Read Alignment
+
 The software used as part of this tutorial is **Bismark Bisulfite Mapper**.
@@
 
-```
     # Directories and programs
@@
 
-```
-
 You should expect to prepared genome with new directories indluding
 
-```
     ./Bisulfite_Genome
@@
     ./Bisulfite_Genome/CT_conversion
-```
 
@@
 
-```
     find ${reads_dir}*_R1_001_val_1.fq.gz \
@@
     -o Mcap_tg
-```
 
@@
 
-```
     find *.bam | \
@@
     {}.bam
-```
 
 # Methylation Quantification
+
 Methylation levels are then extracted using the `bismark_methylation_extractor`. For example:
 
-```
     ${bismark_dir}/bismark_methylation_extractor \
@@
     *deduplicated.bam
-```
 
@@
 
-```
     NC_035784.1 141 141 37.5    3   5
@@
     NC_035784.1 611 611 0   0   4
-```
 
@@
 
-```
     <chromosome> <start position> <end position> <methylation percentage> <count methylated> <count unmethylated>
-```
 
@@
 
-```
     find *deduplicated.bismark.cov.gz \
@@
     {}_trimmed_bismark_bt2.deduplicated.bismark.cov.gz
-```
 
@@
 
-```
     NC_035785.1 217 219 100.000000  17  0
@@
     NC_035785.1 2054    2056    93.333333   14  1
-```
 
@@
 
-```
     #creating bedgraphs post merge
@@
     done
-```
 
@@
 
-```
     # Sort files for methylkit and IGV
@@
     index -@ 28 {}.sorted.bam
-```
 

### 2023-02-22 contributions.Rmd
testing whether we can keep each page's files in one folder to organize the repo

@@
 ```
-# How to contribute new material to the website
+# How to contribute new material to the MarineOmics.github.io website
 If you are interested in submitting new content to the
-MarineOmics.github.io website, we have a [guide](submissions_instructions.html) for you!
+MarineOmics.github.io website, we have a [guide](submissions_instructions/submissions_instructions.html) for you!
 While the website managers will work with you to help prepare and publish your new

### 2023-02-22 index.Rmd
testing whether we can keep each page's files in one folder to organize the repo

@@
 A diverse range of perspectives and backgrounds will make these tutorials more inclusive, informative, and useful for everyone in the marine-omics community.
-If you are interested in developing an information page or tutorial for the website,  we ask that you review our [page with instructions for submitting content](submissions_instructions.html) and contact our Website Manager Jason Johns (jasonjohns[at]ucsb.edu).
+If you are interested in developing an information page or tutorial for the website,  we ask that you review our [page with instructions for submitting content](submissions_instructions/submissions_instructions.html) and contact our Website Manager Jason Johns (jasonjohns[at]ucsb.edu).
 

### 2023-02-22 submissions_instructions.Rmd
publishing submissions_instructions page to website and pushing some other cache files
testing whether we can keep each page's files in one folder to organize the repo

@@
 	fig.width = 10,
-	cache = FALSE
+	cache = TRUE
 )
@@
                fig.height = 5,
-               cache = FALSE)
+               cache = TRUE)
 cite_options(citation_format = "pandoc", max.names = 3, style = "html", 
@@
 Below we provide general instructions as well as examples of how to
-build specific features (eg., tables, code chunks). Prior to publishing
+build specific features (eg., tables, code chunks). Prior to developing
 new pages, please contact the current Website Manager Jason Johns
@@
 
-    knitr::include_graphics("images/Rivera_etal_fig.png")
+    knitr::include_graphics("submissions_instructions/Rivera_etal_fig.png")
 
@@
 
-knitr::include_graphics("images/Rivera_etal_fig.png")
+knitr::include_graphics("submissions_instructions/Rivera_etal_fig.png")
 

### 2023-02-22 contributions.Rmd
same as previous but missed index and contributions pages

--- 2023-02-13 contributions.Rmd
@@
 ```
-# How to contribute new material to the website
+# How to contribute new material to the MarineOmics.github.io website
 If you are interested in submitting new content to the

### 2023-02-27 DNA-methylation.Rmd
meth - reduced image size
Update DNA-methylation.Rmd
clarified descriptions, added links to bismark manual where appropriate

@@
 
-The purpose of this tutorial is to provide an example of taking Bisulfite treated DNA sequence data through identification of methylation at the loci and gene level.
-The principle behind creating bisulfite treated DNA libraries is that when cytosines are not methylated they will be converted to uracil, which ultimately result in thymine.
-When aligning to a reference genome the percent methylation at a given CpG loci is determined by examine the ratio to thymines and cytosines.
-For example if there is a given CpG loci with 30% of the alignment with the cytosine on the reference being and thymine, one would consider this cytosine loci (CpG) to be 70% methylated.
+The purpose of this tutorial is to provide an example of identifying loci- and gene-level methylation in bisulfite-treated DNA sequence data.
+The principle behind creating bisulfite-treated DNA libraries is that when cytosines are not methylated, then they will be converted to uracil, which ultimately result in thymine nucleotides in the sequence data.
+When aligning to a reference genome the percent methylation at a given CpG loci is determined by examining the ratio of thymines to cytosines.
+For example if there is a given cytosine adjacent to a guanine (ie CpG locus) with 30% of the reads containing thymine, one would consider this cytosine loci (CpG) to be 70% methylated.
 
@@
 
-![bp-chart](images/meth-bp-chart.png)
+```{r meth-bp, echo = FALSE, out.width = "70%", fig.align = "center"}
+knitr::include_graphics("images/meth-bp-chart.png")
+```
+
+For WGBS and MBDBS data, it is general practice to hard-trim 10 bp from the beginning and end of each read.
+[RRBS data often needs different trimming specifications](https://rawgit.com/FelixKrueger/Bismark/master/Docs/Bismark_User_Guide.html#ix-notes-about-different-library-types-and-commercial-kits).
 
@@
 
-You should expect to prepared genome with new directories indluding
+You should expect to prepared genome with new directories including
 
@@
 
-For doing the alignment of the trim reads to the genome the following code structured is used
+For aligning the trimmed reads to the genome the following code structure is used
 
@@
 Files are then deduplicated if they are whole genome bisulfite treated samples.
-This command will deduplicate the Bismark alignment BAM file and remove all reads but one which align to the the very same position and in the same orientation.
-This step is recommended for whole-genome bisulfite samples, but should not be used for reduced representation libraries such as RRBS, amplicon or target enrichment libraries.
+This command will deduplicate the Bismark alignment BAM file and remove all reads that align to the the very same position and in the same orientation except for one.
+This step is recommended for whole-genome bisulfite samples, but **should not** be used for reduced representation libraries such as RRBS, amplicon, or target enrichment libraries.
+The [`bismark` manual](https://rawgit.com/FelixKrueger/Bismark/master/Docs/Bismark_User_Guide.html#ix-notes-about-different-library-types-and-commercial-kits) provides additional information.
 
@@
 This will create a file with the suffix `deduplicated.bismark.cov.gz` (example shown below in uncompressed format).
-And while a bedGraph file is generated, in this workflow a bedGraph will be generated later that uses information from merged strands (as opposed to this output).
+The resultant bedGraph file contains information from *unmerged* strands, meaning reads aligning to forward and reverse strands in the same position are not combined.
+Later in this workflow, a bedGraph with *merged* information will be created.
 
@@
 
-A genome-wide cytosine report is generated, with this code as an example:
+A genome-wide cytosine report is generated (including merged bedGraph file), with this code as an example:
 
@@
 
-Generating a file `.CpG_report.merged_CpG_evidence.cov`
+Generating a file `.CpG_report.merged_CpG_evidence.cov` (merged bedGraph file)
 
@@
 
-From here it could be useful to do some simple file reorganizaion to obtain bedgraph or tab files for downstream analysis (eg bedtools, GLMs).
+From here it could be useful to do some simple file reorganizaion to obtain bedGraph or tab files for downstream analysis (eg bedtools, GLMs).
 
-    #creating bedgraphs post merge
+Creating bedGraphs post-merge at a specific coverage threshold (ex. 10x or 5x):
+
     for f in *merged_CpG_evidence.cov
@@
 
-    #creating tab files with raw count for glms
+Creating tab files with raw count data for GLMs at 10x and 5x coverage:
+
     for f in *merged_CpG_evidence.cov

### 2023-03-20 ADMIN_01_submissions_instructions.Rmd
implementing new organizational structure to repo

--- 2023-02-22 submissions_instructions.Rmd
@@
 title: "Instructions for developing new pages for the MarineOmics webpage"
-bibliography: "RCN_site-Nov05.bib"
+bibliography: "ADMIN_01_submissions_instructions_files/common-bib_01.bib"
 editor_options: 
@@
 
--   add your citations to the RCN_site-Nov05.bib document by exporting
+-   add your citations to the common-bib_01.bib document by exporting
     BibTeX-formatted citations from your reference manager
--   add the line 'bibliography: RCN_site-Nov05.bib' to your header at
+-   add the line 'bibliography: common-bib_01.bib' to your header at
     the top of the .Rmd document, which looks like this:
@@
     title: "Instructions for developing new pages for the MarineOmics webpage"
-    bibliography: RCN_site-Nov05.bib
+    bibliography: common-bib_01.bib
     ---
@@
 
-    knitr::include_graphics("submissions_instructions/Rivera_etal_fig.png")
+    knitr::include_graphics("ADMIN_01_submissions_instructions_files/Rivera_etal_fig.png")
 
@@
 
-knitr::include_graphics("submissions_instructions/Rivera_etal_fig.png")
+knitr::include_graphics("ADMIN_01_submissions_instructions_files/Rivera_etal_fig.png")
 

### 2023-03-20 ADMIN_02_contributions.Rmd
implementing new organizational structure to repo

--- 2023-02-13 contributions.Rmd
@@
 ```
-# How to contribute new material to the website
+# How to contribute new material to the MarineOmics.github.io website
 If you are interested in submitting new content to the
-MarineOmics.github.io website, we have a [guide](submissions_instructions.html) for you!
+MarineOmics.github.io website, we have a [guide](ADMIN_01_submissions_instructions.html) for you!
 While the website managers will work with you to help prepare and publish your new

### 2023-03-20 FUN_01_DGE_comparison_v2.Rmd
implementing new organizational structure to repo

--- 2023-02-10 DGE_comparison_v2.Rmd
@@
 subtitle: '*Samuel N. Bogan*'
-bibliography: DGE_Comp.bib
+bibliography: FUN_01_DGE_comparison_v2_files/DGE_Comp.bib
 ---
@@
 # Read in matrix of RSEM expected read counts
-data <- read.delim("RNA_data/GE2_expected_counts_gene.matrix", sep = "\t", header = T, row.names = 1)
+data <- read.delim("FUN_01_DGE_comparison_v2_files/RNA_data/GE2_expected_counts_gene.matrix", sep = "\t", header = T, row.names = 1)
 
@@
 
-knitr::include_graphics("images/Rivera_etal_fig.png")
+knitr::include_graphics("FUN_01_DGE_comparison_v2_files/Rivera_etal_fig.png")
 

### 2023-03-20 POP_01_choosing_population_genetics.Rmd
implementing new organizational structure to repo

--- 2023-02-07 choosing-population-genetics.Rmd
@@
 ---
-bibliography: RCN_site-Nov05.bib
+bibliography: POP_01_choosing_population_genetics_files/common-bib_01.bib
 title: Choosing a Population Genomics Approach
@@
 
-![**Figure 1** from [@Lou2021-me] comparing the distribution of sequencing reads mapped to a reference genome.](images/mec16077-fig-0001-m.jpg)
+![**Figure 1** from [@Lou2021-me] comparing the distribution of sequencing reads mapped to a reference genome.](POP_01_choosing_population_genetics_files/mec16077-fig-0001-m.jpg)
 
@@
 Below we list a number of great review articles and empirical comparison studies that can help when making the decision.
-This topic was also discussed during the [MarineOmics seminars](https://marineomics.github.io/panels.html).
+This topic was also discussed during the [MarineOmics seminars](https://marineomics.github.io/ADMIN_03_panels.html).
 
@@
 
-* [RADseq](RADseq.html)  
-* [hcWGS](WGS_intro.html)
+* [RADseq](POP_02_RADseq.html)  
+* [hcWGS](POP_04_WGS_intro.html)
 * [lcWGS, maintained by MarineOmics member Nicolas Lou](https://github.com/therkildsen-lab/genomic-data-analysis/blob/master/lcwgs_data_analysis.md)
-* [Poolseq](poolseq.html)
+* [Poolseq](POP_03_poolseq.html)
 

### 2023-03-20 POP_04_WGS_intro.Rmd
implementing new organizational structure to repo

--- 2022-11-04 WGS_intro.Rmd
@@
 ---
-bibliography: RCN_site-Nov05.bib
+bibliography: POP_04_WGS_intro_files/common-bib_01.bib
 title:  Whole Genome Resequencing for Population Genomics 
@@
 
-* [Fastq-to-VCF SnakeMake pipeline](fastq2vcf_june2022_update.html): in-depth explanation of an automated short-read mapping and variant calling pipeline maintained by [Harvard Informatics](https://github.com/harvardinformatics/shortRead_mapping_variantCalling).
+* [Fastq-to-VCF SnakeMake pipeline](POP_06_fastq2vcf_june2022_update/POP_06_fastq2vcf_june2022_update.html): in-depth explanation of an automated short-read mapping and variant calling pipeline maintained by [Harvard Informatics](https://github.com/harvardinformatics/shortRead_mapping_variantCalling).
 Useful if you plan to adopt this pre-packaged automated and parallelizable pipeline, and would like to understand its different components, but not necessarily change it substantially.
 
-* [Fastq-to-VCF workflow](WGS-Full-Walkthrough.html): detailed walkthrough of processing 15x sequencing depth WGS data for cod, from raw reads to a VCF.
+* [Fastq-to-VCF workflow](POP_07_WGS_Full_Walkthrough/POP_07_WGS_Full_Walkthrough.html): detailed walkthrough of processing 15x sequencing depth WGS data for cod, from raw reads to a VCF.
 Useful if you would like to run each component of the pipeline yourself and potentially tweak some of them for your own purpose.

### 2023-03-20 POP_05_RDAtraitPredictionTutorial.Rmd
implementing new organizational structure to repo

--- 2023-02-10 RDAtraitPredictionTutorial.Rmd
@@
 
-```{r setup, eval=TRUE, echo=FALSE}
+```{r setup, eval=TRUE, echo=TRUE, message = FALSE, include = TRUE}
 
@@
 
-G <- read.table(unz("rda_tutorial/Genotypes.txt.zip", "Genotypes.txt"))
+G <- read.table(unz("POP_05_RDAtraitPredictionTutorial_files/Genotypes.txt.zip", "Genotypes.txt"))
 dim(G) # 1000 individuals and 26371 loci
 
-ind <- read.table("rda_tutorial/Individuals.txt", header=TRUE)
+ind <- read.table("POP_05_RDAtraitPredictionTutorial_files/Individuals.txt", header=TRUE)
 dim(ind) #corresponds to rows in G
@@
 
-muts <-  read.table("rda_tutorial/SNPs.txt", header=TRUE)
+muts <-  read.table("POP_05_RDAtraitPredictionTutorial_files/SNPs.txt", header=TRUE)
 dim(muts) #corresponds to columns in G

### 2023-03-20 POP_05_RDAtraitPredictionTutorial_files/RDAtraitPredictionTutorial.tex
implementing new organizational structure to repo

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={RDA trait prediction tutorial},
  pdfauthor={KE Lotterhos},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{RDA trait prediction tutorial}
\author{KE Lotterhos}
\date{2023-02-08}

\begin{document}
\maketitle

This tutorial accompanies the paper ``The paradox of adaptive trait
clines with non-clinal patterns in the underlying genes'' published in
PNAS {[}add link{]}.

\hypertarget{abstract}{%
\section{Abstract}\label{abstract}}

Multivariate climate change presents an urgent need to understand how
species adapt to complex environments. Population genetic theory
predicts that loci under selection will form monotonic allele frequency
clines with their selective environment, which has led to the wide use
of genotype-environment associations (GEAs). However, the accuracy of
GEA methods to identify adapted loci is limited, as shown in the main
paper.

This tutorial shows how to apply a novel extension of multivariate
ordination, which accurately predicted individual multivariate traits
from genotype and environmental data on simulated data regardless of
whether inference from GEAs was accurate.

\hypertarget{install-packages}{%
\section{Install packages}\label{install-packages}}

If the following packages are not installed, be sure to install them
first:

\begin{verbatim}
install.packages("vegan")
install.packages("lfmm")
install.packages("gplots")

if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("LEA")
\end{verbatim}

\hypertarget{load-the-libraries}{%
\subsection{Load the libraries:}\label{load-the-libraries}}

\begin{verbatim}
## Loading required package: permute
\end{verbatim}

\begin{verbatim}
## Loading required package: lattice
\end{verbatim}

\begin{verbatim}
## This is vegan 2.6-4
\end{verbatim}

\begin{verbatim}
## 
## Attaching package: 'LEA'
\end{verbatim}

\begin{verbatim}
## The following object is masked from 'package:lattice':
## 
##     barchart
\end{verbatim}

\begin{verbatim}
## 
## Attaching package: 'gplots'
\end{verbatim}

\begin{verbatim}
## The following object is masked from 'package:stats':
## 
##     lowess
\end{verbatim}

Don't forget to set your working directory!

\hypertarget{background-on-the-simulation}{%
\section{Background on the
simulation}\label{background-on-the-simulation}}

This data was simulated in SLiM and is associated with the complex
multivariate simulation presented in Figure 6 in the paper. Briefly, a
non-Wright-Fisher model was simulated on a landscape with 6
environmental variables that reflect different aspects of thermal stress
and precipitation in British Columbia. The simulation included 6
environmental traits, each of which adapted to a different environmental
variable.

The six environmental variables are based on real data from western
Canada and are shown below, clockwise from upper right: Clockwise from
upper left: precipitation of driest month, precipitation of warmest
quarter, mean annual temperature, precipitation of wettest month, mean
temperature of wettest quarter, mean temperature of driest quarter. Each
small square is a simulated individual, with its color representing its
trait value in that environment.

\includegraphics{multivar.png}

\hypertarget{load-the-data}{%
\section{Load the data}\label{load-the-data}}

\begin{itemize}
\tightlist
\item
  A matrix of genotypes in 012 format (counts of reference allele)

  \begin{itemize}
  \tightlist
  \item
    number of rows = number of individuals
  \item
    number of columns = number of SNPs
  \end{itemize}
\item
  A table with information about sampled individuals (each individual in
  a row)
\item
  A table with information about SNPs (each SNP in a row)
\end{itemize}

The \texttt{ind} table includes the xy location for each individual, the
6 exact trait values (note that these won't exactly equal the trait
value calculated from the genotype matrix because of MAF filtering), and
the 6 environmental values at their xy location.

The \texttt{muts} table includes the linkage group \texttt{LG}, the
position of the mutation on the genetic map \texttt{pos\_pyslim}, a
unique ID \texttt{mutname}, the allele frequency based on the 1000
sampled individuals \texttt{a\_freq\_subset}, and whether or not it had
effects on one or more phenotypes \texttt{causal}.

Note that you will have to change the working directory to where the
data is stored on your computer.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{G }\OtherTok{\textless{}{-}} \FunctionTok{read.table}\NormalTok{(}\FunctionTok{unz}\NormalTok{(}\StringTok{"Genotypes.txt.zip"}\NormalTok{, }\StringTok{"Genotypes.txt"}\NormalTok{))}
\FunctionTok{dim}\NormalTok{(G) }\CommentTok{\# 1000 individuals and 26371 loci}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1]  1000 26371
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{ind }\OtherTok{\textless{}{-}} \FunctionTok{read.table}\NormalTok{(}\StringTok{"Individuals.txt"}\NormalTok{, }\AttributeTok{header=}\ConstantTok{TRUE}\NormalTok{)}
\FunctionTok{dim}\NormalTok{(ind) }\CommentTok{\#corresponds to rows in G}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] 1000   15
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{head}\NormalTok{(ind)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   ind_index         x        y phenotype1_mat phenotype2_MTWetQ phenotype3_MTDQ
## 1        33 0.4061840 0.233272      0.7379670        -0.1914210        0.142992
## 2        34 0.4256970 0.837158     -0.2740810        -0.4283320       -0.012880
## 3        44 0.6716730 0.581744     -0.6763260         0.0866287       -0.470671
## 4        45 0.0174659 0.329922     -0.2832210        -0.4039420       -0.333718
## 5        46 0.0697436 0.121221      0.0576163        -0.3032010        0.731600
## 6        64 0.1433610 0.233237      0.2656280        -0.2318730        0.214004
##   phenotype4_PDM phenotype5_PwarmQ phenotype6_PWM  env1_mat env2_MTWetQ
## 1      -0.131894         -0.452497     -0.5416510  0.762432  -0.1622380
## 2       0.221058          0.229296      0.0665969 -0.339330  -0.4076670
## 3       0.218813          0.260098     -0.1034650 -0.567733   0.0909278
## 4      -0.200554         -0.345915     -0.4100010 -0.245860  -0.2007590
## 5       0.195679         -0.367110      0.1962960 -0.121500  -0.3063790
## 6      -0.239776         -0.517086     -0.2582660  0.103057  -0.2863170
##    env3_MTDQ  env4_PDM env5_PwarmQ   env6_PWM
## 1  0.2612810 -0.332078   -0.471502 -0.4778210
## 2 -0.0582962  0.288030    0.198810  0.0176800
## 3 -0.4143530  0.234899    0.230255 -0.0535109
## 4 -0.2674550 -0.265260   -0.389828 -0.4589790
## 5  0.7028580  0.211275   -0.352307  0.2127510
## 6  0.1657060 -0.190813   -0.476650 -0.2927740
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{muts }\OtherTok{\textless{}{-}}  \FunctionTok{read.table}\NormalTok{(}\StringTok{"SNPs.txt"}\NormalTok{, }\AttributeTok{header=}\ConstantTok{TRUE}\NormalTok{)}
\FunctionTok{dim}\NormalTok{(muts) }\CommentTok{\#corresponds to columns in G}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] 26371     5
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{head}\NormalTok{(muts)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   LG pos_pyslim mutname a_freq_subset causal
## 1  1          8     1-8        0.0175  FALSE
## 2  1         27    1-27        0.0120  FALSE
## 3  1         34    1-34        0.0370  FALSE
## 4  1         81    1-81        0.0330  FALSE
## 5  1         97    1-97        0.0170  FALSE
## 6  1        153   1-153        0.0225  FALSE
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{rownames}\NormalTok{(G) }\OtherTok{\textless{}{-}} \FunctionTok{as.character}\NormalTok{(}\FunctionTok{paste0}\NormalTok{(}\StringTok{"i\_"}\NormalTok{,ind}\SpecialCharTok{$}\NormalTok{ind\_index))}
\FunctionTok{colnames}\NormalTok{(G) }\OtherTok{\textless{}{-}} \FunctionTok{as.character}\NormalTok{(muts}\SpecialCharTok{$}\NormalTok{mutname)}
\CommentTok{\#G \textless{}{-} as.matrix(G)}
\FunctionTok{head}\NormalTok{(G[,}\DecValTok{1}\SpecialCharTok{:}\DecValTok{10}\NormalTok{])}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##      1-8 1-27 1-34 1-81 1-97 1-153 1-154 1-206 1-287 1-304
## i_33   0    0    0    0    0     0     0     0     0     0
## i_34   0    0    0    0    0     0     0     0     0     0
## i_44   0    0    0    0    0     0     0     0     0     0
## i_45   0    0    0    0    0     0     0     0     0     0
## i_46   0    0    0    0    0     0     0     0     1     0
## i_64   2    0    0    0    0     0     0     0     0     0
\end{verbatim}

\hypertarget{rda-trait-prediction-function}{%
\section{RDA trait prediction
function}\label{rda-trait-prediction-function}}

This function predicts an environmental trait through the
back-transformation of the RDA ``site score'' of an individual to a
chosen environmental variable (Equation 1 in the manuscript). It makes
the prediction for all the individuals that were used to run the RDA.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{rda\_trait\_pred }\OtherTok{\textless{}{-}} \ControlFlowTok{function}\NormalTok{(rdaobj, env\_row, K)\{}
  \CommentTok{\#rdaobj is RDA object}
  \CommentTok{\#env\_row is the row of the environment in the biplot output}
  \CommentTok{\#K is the number of RDA axes}
\NormalTok{  scores }\OtherTok{\textless{}{-}} \FunctionTok{scores}\NormalTok{(rdaobj, }\AttributeTok{choices=}\DecValTok{1}\SpecialCharTok{:}\NormalTok{K)}
\NormalTok{  ind.sc }\OtherTok{\textless{}{-}}\NormalTok{ scores}\SpecialCharTok{$}\NormalTok{sites}
\NormalTok{  pred }\OtherTok{\textless{}{-}} \FunctionTok{matrix}\NormalTok{(}\ConstantTok{NA}\NormalTok{, }\AttributeTok{nrow=}\FunctionTok{nrow}\NormalTok{(ind.sc), }\AttributeTok{ncol=}\NormalTok{K)}
  \ControlFlowTok{for}\NormalTok{ (k }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\NormalTok{K)\{}
\NormalTok{    pred[,k] }\OtherTok{\textless{}{-}}\NormalTok{ ind.sc[,k]}\SpecialCharTok{*}\FunctionTok{eigenvals}\NormalTok{(rdaobj)[k]}\SpecialCharTok{*}\FunctionTok{summary}\NormalTok{(rdaobj)}\SpecialCharTok{$}\NormalTok{biplot[env\_row,k]}
\NormalTok{  \}}
\NormalTok{  trait\_pred }\OtherTok{\textless{}{-}} \FunctionTok{scale}\NormalTok{(}\FunctionTok{rowSums}\NormalTok{(pred))}
 \FunctionTok{return}\NormalTok{(trait\_pred) }
\NormalTok{\}}
\end{Highlighting}
\end{Shaded}

\hypertarget{example-of-an-rda-predicted-environmental-trait-value}{%
\subsection{Example of an RDA-predicted environmental trait
value}\label{example-of-an-rda-predicted-environmental-trait-value}}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  First, run the RDA:
\end{enumerate}

Scale the environmental variables to have a mean of 0 and standard
deviation of 1.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{ind}\SpecialCharTok{$}\NormalTok{env1\_mat }\OtherTok{\textless{}{-}} \FunctionTok{scale}\NormalTok{(ind}\SpecialCharTok{$}\NormalTok{env1\_mat)}
\NormalTok{ind}\SpecialCharTok{$}\NormalTok{env2\_MTWetQ }\OtherTok{\textless{}{-}} \FunctionTok{scale}\NormalTok{(ind}\SpecialCharTok{$}\NormalTok{env2\_MTWetQ)}
\NormalTok{ind}\SpecialCharTok{$}\NormalTok{env3\_MTDQ }\OtherTok{\textless{}{-}} \FunctionTok{scale}\NormalTok{(ind}\SpecialCharTok{$}\NormalTok{env3\_MTDQ)}
\NormalTok{ind}\SpecialCharTok{$}\NormalTok{env4\_PDM }\OtherTok{\textless{}{-}} \FunctionTok{scale}\NormalTok{(ind}\SpecialCharTok{$}\NormalTok{env4\_PDM)}
\NormalTok{ind}\SpecialCharTok{$}\NormalTok{env5\_PwarmQ }\OtherTok{\textless{}{-}} \FunctionTok{scale}\NormalTok{(ind}\SpecialCharTok{$}\NormalTok{env5\_PwarmQ)}
\NormalTok{ind}\SpecialCharTok{$}\NormalTok{env6\_PWM }\OtherTok{\textless{}{-}} \FunctionTok{scale}\NormalTok{(ind}\SpecialCharTok{$}\NormalTok{env6\_PWM)}

\CommentTok{\# Run the RDA}
\NormalTok{rdaout }\OtherTok{\textless{}{-}} \FunctionTok{rda}\NormalTok{(G }\SpecialCharTok{\textasciitilde{}}\NormalTok{ ind}\SpecialCharTok{$}\NormalTok{env1\_mat }\SpecialCharTok{+}
\NormalTok{                ind}\SpecialCharTok{$}\NormalTok{env2\_MTWetQ }\SpecialCharTok{+}
\NormalTok{                ind}\SpecialCharTok{$}\NormalTok{env3\_MTDQ }\SpecialCharTok{+} 
\NormalTok{                ind}\SpecialCharTok{$}\NormalTok{env4\_PDM }\SpecialCharTok{+}
\NormalTok{                ind}\SpecialCharTok{$}\NormalTok{env5\_PwarmQ }\SpecialCharTok{+}
\NormalTok{                ind}\SpecialCharTok{$}\NormalTok{env6\_PWM}
\NormalTok{              )}
\end{Highlighting}
\end{Shaded}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{1}
\tightlist
\item
  Next, check the biplot output and decide how many RDA axes to use in
  the prediction.
\end{enumerate}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Check the biplot output}
\NormalTok{rdaout}\SpecialCharTok{$}\NormalTok{CCA}\SpecialCharTok{$}\NormalTok{biplot}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##                       RDA1        RDA2       RDA3        RDA4        RDA5
## ind$env1_mat    -0.5004451  0.01863416 -0.5476147  0.54418964 -0.37298079
## ind$env2_MTWetQ  0.4523477  0.19158505 -0.8296969 -0.11392353  0.03132509
## ind$env3_MTDQ   -0.7128197 -0.22757653  0.1200761  0.48309488  0.33816827
## ind$env4_PDM    -0.3843437  0.04904761  0.8297919 -0.24744529  0.09707124
## ind$env5_PwarmQ  0.6524152  0.56221376  0.4505148  0.02427266 -0.04630498
## ind$env6_PWM     0.3209795 -0.05176393  0.7346369  0.19495732 -0.08780176
##                       RDA6
## ind$env1_mat    -0.1186113
## ind$env2_MTWetQ -0.2373183
## ind$env3_MTDQ   -0.2791780
## ind$env4_PDM    -0.3011107
## ind$env5_PwarmQ -0.2292885
## ind$env6_PWM    -0.5557731
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Decide how many RDA axes to use in calculation}
\NormalTok{  a}\OtherTok{\textless{}{-}} \FunctionTok{screeplot}\NormalTok{(rdaout)}
\end{Highlighting}
\end{Shaded}

\includegraphics{RDAtraitPredictionTutorial_files/figure-latex/unnamed-chunk-1-1.pdf}

\begin{Shaded}
\begin{Highlighting}[]
  \FunctionTok{str}\NormalTok{(a)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## List of 4
##  $ x   : num [1:6] 0.7 1.9 3.1 4.3 5.5 6.7
##  $ y   : num [1:6] 189 103.1 66.1 37.8 30.9 ...
##  $ xlab: NULL
##  $ ylab: NULL
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{  a}\SpecialCharTok{$}\NormalTok{y }\CommentTok{\# save this it\textquotesingle{}s the eigenvalues}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] 189.02481 103.10740  66.10749  37.79033  30.86247  23.24983
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{  prop\_var }\OtherTok{\textless{}{-}} \FunctionTok{round}\NormalTok{(a}\SpecialCharTok{$}\NormalTok{y[}\DecValTok{1}\SpecialCharTok{:}\DecValTok{6}\NormalTok{]}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(a}\SpecialCharTok{$}\NormalTok{y),}\DecValTok{3}\NormalTok{)}
  \FunctionTok{cumsum}\NormalTok{(prop\_var)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] 0.420 0.649 0.796 0.880 0.949 1.001
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
  \FunctionTok{plot}\NormalTok{(}\FunctionTok{cumsum}\NormalTok{(prop\_var), }\AttributeTok{xlab=}\StringTok{"Number of RDA axes"}\NormalTok{, }
       \AttributeTok{ylab=}\StringTok{"Cumulative percent of variation explained"}\NormalTok{, }\AttributeTok{ylim=}\FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\includegraphics{RDAtraitPredictionTutorial_files/figure-latex/unnamed-chunk-1-2.pdf}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{2}
\tightlist
\item
  In this case, the first 3 RDA axes explain 80\% of the variance. Note
  that choosing too many axes may result in overfitting. Here is an
  example of a trait prediction for MAT using the first 3 RDA axes:
\end{enumerate}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Make the trait prediction for MAT (1st row in biplot output)}
\NormalTok{K }\OtherTok{=} \DecValTok{3} \CommentTok{\# use 3 RDA axes to make the trait prediction}

\NormalTok{MATtraitPredict }\OtherTok{\textless{}{-}} \FunctionTok{rda\_trait\_pred}\NormalTok{(rdaout, }\DecValTok{1}\NormalTok{, K)}

\CommentTok{\# Since this is a simulation, we can compare the prediction to the true value}
\CommentTok{\# Similarly, an empirical study could compare an empirically measured trait value }
\CommentTok{\# to the RDA{-}predicted trait value to test how well landscape genomic data }
\CommentTok{\# can predict functional traits}
\FunctionTok{plot}\NormalTok{(}\FunctionTok{scale}\NormalTok{(ind}\SpecialCharTok{$}\NormalTok{phenotype1\_mat), MATtraitPredict, }\AttributeTok{xlab=}\StringTok{"Evolved trait value in simulations"}\NormalTok{,}
     \AttributeTok{ylab=}\StringTok{"RDA trait prediction"}\NormalTok{)}
\FunctionTok{abline}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{RDAtraitPredictionTutorial_files/figure-latex/unnamed-chunk-2-1.pdf}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#Correlation between the prediction and the true value:}
\FunctionTok{cor}\NormalTok{(ind}\SpecialCharTok{$}\NormalTok{phenotype1\_mat, MATtraitPredict) }
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##           [,1]
## [1,] 0.6461756
\end{verbatim}

\hypertarget{compare-to-other-functions-in-rda}{%
\subsubsection{Compare to other functions in
RDA}\label{compare-to-other-functions-in-rda}}

Note that the \texttt{predict} function and it's variations in the R
package \texttt{vegan} do not make the same kind of predictions as
\texttt{rda\_trait\_pred}. Here are the types of outputs produced by the
function \texttt{predict} and its variations:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# This option in the \textasciigrave{}predict\textasciigrave{} function outputs the scores for each locus in RDA space}
\NormalTok{loci\_scores\_predict }\OtherTok{\textless{}{-}} \FunctionTok{predict}\NormalTok{(rdaout, }\AttributeTok{type=}\StringTok{"sp"}\NormalTok{, }\AttributeTok{newdata=}\NormalTok{G, }\AttributeTok{scaling=}\DecValTok{2}\NormalTok{)}
\FunctionTok{str}\NormalTok{(loci\_scores\_predict)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  num [1:26371, 1:6] -0.00957 -0.01685 -0.00659 0.08155 0.03396 ...
##  - attr(*, "dimnames")=List of 2
##   ..$ : chr [1:26371] "1-8" "1-27" "1-34" "1-81" ...
##   ..$ : chr [1:6] "RDA1" "RDA2" "RDA3" "RDA4" ...
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# This option in the \textasciigrave{}predict\textasciigrave{} function outputs the fitted values from the multiple regression}
\CommentTok{\# performed on each locus within each individual }
\NormalTok{fitted\_values\_predict }\OtherTok{\textless{}{-}} \FunctionTok{predict}\NormalTok{(rdaout, }\AttributeTok{newdata=}\NormalTok{G, }\AttributeTok{type=}\StringTok{"response"}\NormalTok{)}
\FunctionTok{str}\NormalTok{(fitted\_values\_predict)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  num [1:1000, 1:26371] 0.1031 0.0345 -0.0436 0.1821 0.1753 ...
##  - attr(*, "dimnames")=List of 2
##   ..$ : chr [1:1000] "i_33" "i_34" "i_44" "i_45" ...
##   ..$ : chr [1:26371] "1-8" "1-27" "1-34" "1-81" ...
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
  \CommentTok{\# As a side note, it outputs the same thing as the \textasciigrave{}fitted\textasciigrave{} function}
\NormalTok{  fitted\_values\_predict2 }\OtherTok{\textless{}{-}} \FunctionTok{fitted}\NormalTok{(rdaout)}
  \FunctionTok{str}\NormalTok{(fitted\_values\_predict2)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  num [1:1000, 1:26371] 0.1031 0.0345 -0.0436 0.1821 0.1753 ...
##  - attr(*, "METHOD")= chr "PCA"
##  - attr(*, "dimnames")=List of 2
##   ..$ : chr [1:1000] "i_33" "i_34" "i_44" "i_45" ...
##   ..$ : chr [1:26371] "1-8" "1-27" "1-34" "1-81" ...
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# This option  in the \textasciigrave{}predict\textasciigrave{} function outputs the individual scores in RDA space }
\CommentTok{\# based on a linear combination of the predictor variables}
\NormalTok{X }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(ind}\SpecialCharTok{$}\NormalTok{env1\_mat ,}
\NormalTok{                ind}\SpecialCharTok{$}\NormalTok{env2\_MTWetQ ,}
\NormalTok{                ind}\SpecialCharTok{$}\NormalTok{env3\_MTDQ ,}
\NormalTok{                ind}\SpecialCharTok{$}\NormalTok{env4\_PDM ,}
\NormalTok{                ind}\SpecialCharTok{$}\NormalTok{env5\_PwarmQ ,}
\NormalTok{                ind}\SpecialCharTok{$}\NormalTok{env6\_PWM)}
\NormalTok{ind\_scores\_predict }\OtherTok{\textless{}{-}} \FunctionTok{predict}\NormalTok{(rdaout, }\AttributeTok{type=}\StringTok{"lc"}\NormalTok{, }\AttributeTok{new=}\NormalTok{X, }\AttributeTok{scal=}\DecValTok{2}\NormalTok{)}
\FunctionTok{str}\NormalTok{(ind\_scores\_predict)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  num [1:1000, 1:6] -2.034 -0.19 0.543 0.253 -0.508 ...
##  - attr(*, "dimnames")=List of 2
##   ..$ : chr [1:1000] "1" "2" "3" "4" ...
##   ..$ : chr [1:6] "RDA1" "RDA2" "RDA3" "RDA4" ...
\end{verbatim}

The \texttt{predict} function and its variations make predictions in RDA
space, and therefore do not output the same kind of predictions as
\texttt{rda\_trait\_predict} and Equation 1 in the paper.

\hypertarget{understanding-how-the-rda-is-built-on-multiple-regressions}{%
\section{Understanding how the RDA is built on multiple
regressions}\label{understanding-how-the-rda-is-built-on-multiple-regressions}}

Prior to ordination in the RDA, each locus is used in a multiple
regression model with the environmental variables to produce fitted
values for that locus across individuals.

\emph{SNP Genotype} \textasciitilde{} \emph{Env1} + \emph{Env2} +
\emph{Env3} etc.

For example for the first SNP in the data:

\begin{Shaded}
\begin{Highlighting}[]
 \CommentTok{\# multiple regression of 1st locus}

\NormalTok{mod }\OtherTok{\textless{}{-}} \FunctionTok{lm}\NormalTok{(G[,}\DecValTok{1}\NormalTok{] }\SpecialCharTok{\textasciitilde{}}\NormalTok{ ind}\SpecialCharTok{$}\NormalTok{env1\_mat }\SpecialCharTok{+}\NormalTok{ ind}\SpecialCharTok{$}\NormalTok{env2\_MTWetQ }\SpecialCharTok{+}\NormalTok{ ind}\SpecialCharTok{$}\NormalTok{env3\_MTDQ }\SpecialCharTok{+}\NormalTok{  ind}\SpecialCharTok{$}\NormalTok{env4\_PDM }\SpecialCharTok{+}
\NormalTok{                ind}\SpecialCharTok{$}\NormalTok{env5\_PwarmQ }\SpecialCharTok{+}\NormalTok{   ind}\SpecialCharTok{$}\NormalTok{env6\_PWM)}
 \FunctionTok{coef}\NormalTok{(}\FunctionTok{summary}\NormalTok{(mod))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##                     Estimate  Std. Error    t value     Pr(>|t|)
## (Intercept)      0.035000000 0.007404071  4.7271290 2.605782e-06
## ind$env1_mat    -0.035729076 0.011338501 -3.1511287 1.675100e-03
## ind$env2_MTWetQ -0.062538545 0.012165353 -5.1407094 3.295576e-07
## ind$env3_MTDQ    0.007471565 0.013224098  0.5649962 5.722040e-01
## ind$env4_PDM    -0.088670803 0.014835808 -5.9768099 3.169042e-09
## ind$env5_PwarmQ -0.069588746 0.014273932 -4.8752331 1.264746e-06
## ind$env6_PWM     0.047203246 0.013582893  3.4751982 5.325440e-04
\end{verbatim}

Although multiple regression is a linear combination of multiple
variables, it is able to model complex multivariate responses that
appear to be non-monotonic in any one dimension. For example, let's look
at a the relationship between explanatory variable temperature and the
response variable genotype, across decreasing and increasing values of
the other explanatory variables:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{otherenv }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\FunctionTok{seq}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{0}\NormalTok{,}\AttributeTok{length.out=}\DecValTok{100}\NormalTok{), }\FunctionTok{seq}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{,}\AttributeTok{length.out=}\DecValTok{101}\NormalTok{))}
\NormalTok{newdata}\OtherTok{=}\FunctionTok{data.frame}\NormalTok{(}\AttributeTok{ind.env1\_mat =} \FunctionTok{seq}\NormalTok{(}\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{,}\DecValTok{1}\NormalTok{, }\AttributeTok{by=}\FloatTok{0.01}\NormalTok{),}
                   \AttributeTok{ind.env2\_MTWetQ =}\NormalTok{ otherenv,}
                   \AttributeTok{ind.env3\_MTDQ =}\NormalTok{ otherenv,}
                   \AttributeTok{ind.env4\_PDM =}\NormalTok{ otherenv,}
                   \AttributeTok{ind.env5\_PwarmQ =}\NormalTok{otherenv,}
                   \AttributeTok{ind.env6\_PWM =}\NormalTok{ otherenv)}
\NormalTok{pred }\OtherTok{\textless{}{-}} \FunctionTok{t}\NormalTok{(newdata)}\SpecialCharTok{*}\NormalTok{(}\FunctionTok{coef}\NormalTok{(}\FunctionTok{summary}\NormalTok{(mod))[}\DecValTok{2}\SpecialCharTok{:}\DecValTok{7}\NormalTok{,}\DecValTok{1}\NormalTok{]) }\SpecialCharTok{+} \FunctionTok{coef}\NormalTok{(}\FunctionTok{summary}\NormalTok{(mod))[}\DecValTok{1}\NormalTok{,}\DecValTok{1}\NormalTok{]}
\FunctionTok{plot}\NormalTok{(}\FunctionTok{seq}\NormalTok{(}\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{,}\DecValTok{1}\NormalTok{, }\AttributeTok{by=}\FloatTok{0.01}\NormalTok{), }\FunctionTok{colSums}\NormalTok{(pred), }\AttributeTok{xlab=}\StringTok{"Temperature"}\NormalTok{, }\AttributeTok{ylab=}\StringTok{"Genotype prediction"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{RDAtraitPredictionTutorial_files/figure-latex/unnamed-chunk-5-1.pdf}

Thus, there is flexibility with the RDA to capture the way environmental
variables may influence the patterns at one locus in a different way
than at another locus, which may not correlate with the relationship
between the environment and population structure.

It may be interesting for some studies to understand how each locus is
shaped by the environment - in other words, what are the slopes
associated with the environmental variables in the multiple regression
model for each locus?

Unfortunately there is not a way to output these slopes in the R package
\texttt{vegan}, but we can reproduce the first step of the RDA to get
the regression coefficients: (vegan source code at
\url{https://github.com/cran/vegan/blob/master/R/simpleRDA2.R})

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{X }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(ind}\SpecialCharTok{$}\NormalTok{env1\_mat ,}
\NormalTok{                ind}\SpecialCharTok{$}\NormalTok{env2\_MTWetQ ,}
\NormalTok{                ind}\SpecialCharTok{$}\NormalTok{env3\_MTDQ ,}
\NormalTok{                ind}\SpecialCharTok{$}\NormalTok{env4\_PDM ,}
\NormalTok{                ind}\SpecialCharTok{$}\NormalTok{env5\_PwarmQ ,}
\NormalTok{                ind}\SpecialCharTok{$}\NormalTok{env6\_PWM)}

\CommentTok{\# Perform qr decomposition to do the regression for all SNPs at the same time }
\NormalTok{ Q }\OtherTok{\textless{}{-}} \FunctionTok{qr}\NormalTok{(X, }\AttributeTok{tol=}\FloatTok{1e{-}6}\NormalTok{)}
\CommentTok{\# str(Q) run this line if you want to understand the structure of Q}

\CommentTok{\# Get the matrix of regression coefficients}
\NormalTok{ Qr.coeff }\OtherTok{\textless{}{-}} \FunctionTok{qr.coef}\NormalTok{(Q, G)}
 \CommentTok{\# This matrix has each SNP in a column and the regression coefficients }
 \CommentTok{\# for that SNP corresponds to each environmental variable. }
 \CommentTok{\# This is the step that is not performed in the \textasciigrave{}vegan\textasciigrave{} package {-} }
 \CommentTok{\# the package skips directly to predicting the fitted values, }
 \CommentTok{\# on which the ordination is performed.}

\CommentTok{\# Here is an example of regression coefficients for the first 10 SNPs:}
 \FunctionTok{head}\NormalTok{(Qr.coeff[,}\DecValTok{1}\SpecialCharTok{:}\DecValTok{10}\NormalTok{])}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##                         [,1]         [,2]         [,3]          [,4]
## ind.env1_mat    -0.035729076  0.007247705 -0.001047802 -6.104895e-02
## ind.env2_MTWetQ -0.062538545  0.039624398  0.101004048  9.680467e-05
## ind.env3_MTDQ    0.007471565 -0.005698865  0.039132724  7.995858e-02
## ind.env4_PDM    -0.088670803  0.017639011  0.027586476 -1.237919e-01
## ind.env5_PwarmQ -0.069588746 -0.043489428  0.005967464  1.262363e-01
## ind.env6_PWM     0.047203246 -0.012944110 -0.061687810 -6.657131e-03
##                         [,5]        [,6]        [,7]         [,8]        [,9]
## ind.env1_mat    -0.063099759  0.02064869  0.02539415  0.002292975 -0.07305243
## ind.env2_MTWetQ  0.023736921 -0.03019428  0.01360083  0.112457789 -0.02354021
## ind.env3_MTDQ   -0.019240464 -0.02199593 -0.06875513  0.043179895  0.10418350
## ind.env4_PDM    -0.003946225 -0.11767281  0.05887330  0.032681938 -0.09642371
## ind.env5_PwarmQ -0.032384913  0.05976386 -0.07535789  0.008011003 -0.03393128
## ind.env6_PWM     0.018107020  0.06710304  0.06788231 -0.068827598  0.02928042
##                       [,10]
## ind.env1_mat     0.03694418
## ind.env2_MTWetQ -0.04964375
## ind.env3_MTDQ   -0.05660635
## ind.env4_PDM     0.01149236
## ind.env5_PwarmQ -0.01743073
## ind.env6_PWM    -0.01118472
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Note that the regression coefficients for the first SNP from this }
\CommentTok{\# approach is exactly the same as from our model above:}
\NormalTok{ Qr.coeff[,}\DecValTok{1}\NormalTok{]}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##    ind.env1_mat ind.env2_MTWetQ   ind.env3_MTDQ    ind.env4_PDM ind.env5_PwarmQ 
##    -0.035729076    -0.062538545     0.007471565    -0.088670803    -0.069588746 
##    ind.env6_PWM 
##     0.047203246
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
 \FunctionTok{coef}\NormalTok{(}\FunctionTok{summary}\NormalTok{(mod))[,}\DecValTok{1}\NormalTok{]}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##     (Intercept)    ind$env1_mat ind$env2_MTWetQ   ind$env3_MTDQ    ind$env4_PDM 
##     0.035000000    -0.035729076    -0.062538545     0.007471565    -0.088670803 
## ind$env5_PwarmQ    ind$env6_PWM 
##    -0.069588746     0.047203246
\end{verbatim}

We can visualize the regression coefficients with a heatmap. In this
case, we know the causal loci in the simulations, so we will just
visualize those loci.

This visualization illustrates how there are unique ways in which
environments are combined in the model to predict the pattern at each
SNP.

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{dim}\NormalTok{(Qr.coeff)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1]     6 26371
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{colnames}\NormalTok{(Qr.coeff) }\OtherTok{\textless{}{-}}\NormalTok{ muts}\SpecialCharTok{$}\NormalTok{mutname}

\CommentTok{\# look at the range of coefficients in the multiple regression model}
\FunctionTok{summary}\NormalTok{(}\FunctionTok{as.numeric}\NormalTok{(Qr.coeff[,}\FunctionTok{which}\NormalTok{(muts}\SpecialCharTok{$}\NormalTok{causal)]))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##      Min.   1st Qu.    Median      Mean   3rd Qu.      Max. 
## -0.657086 -0.054464  0.001058  0.001041  0.056986  0.447236
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{brks }\OtherTok{\textless{}{-}} \FunctionTok{seq}\NormalTok{(}\SpecialCharTok{{-}}\FloatTok{0.7}\NormalTok{, }\FloatTok{0.7}\NormalTok{, }\AttributeTok{by=}\FloatTok{0.05}\NormalTok{) }\CommentTok{\#set the color scale}

\FunctionTok{heatmap.2}\NormalTok{(}\FunctionTok{t}\NormalTok{(Qr.coeff[,}\FunctionTok{which}\NormalTok{(muts}\SpecialCharTok{$}\NormalTok{causal)]),}
        \AttributeTok{scale=}\StringTok{"none"}\NormalTok{, }
        \AttributeTok{col =} \FunctionTok{cm.colors}\NormalTok{(}\FunctionTok{length}\NormalTok{(brks)}\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{), }
        \AttributeTok{breaks=}\NormalTok{brks,}
        \AttributeTok{dendrogram =} \StringTok{"column"}\NormalTok{,}
        \AttributeTok{Rowv=}\ConstantTok{FALSE}\NormalTok{, }\CommentTok{\#set this to "TRUE" if you would like to see which groups}
        \AttributeTok{trace=}\StringTok{"none"}\NormalTok{,}
        \AttributeTok{key.title =} \StringTok{"Coefficient in multiple}\SpecialCharTok{\textbackslash{}n}\StringTok{regression model"}\NormalTok{,}
        \AttributeTok{ylab=}\StringTok{"SNPs"}\NormalTok{,}
        \AttributeTok{cexCol=}\DecValTok{1}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{RDAtraitPredictionTutorial_files/figure-latex/unnamed-chunk-7-1.pdf}

In the above heatmap, each row is a SNP. The SNPs are named according to
their linkage group (1 through 10) and cumulative position in the genome
(e.g.~9-448632 is on the 9th linkage group). Each linkage group was
50,000 bases long, so the cumulative position ranges from 1 to 500,000
over the 10 linkage groups. Each column in the heatmap is an
environment, which is an explanatory variable in the model.

The color of the heatmap cells for a SNP shows the coefficients in the
multiple regression model for each corresponding environment. In other
words, the colors show how environments are combined in a multiple
regression model to predict the patterns at that SNP on the landscape.

If we want to visualize clusters of SNPs that have similar coefficients
in the multiple regression model, we can allow for clustering in the
heatmap:

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{heatmap.2}\NormalTok{(}\FunctionTok{t}\NormalTok{(Qr.coeff[,}\FunctionTok{which}\NormalTok{(muts}\SpecialCharTok{$}\NormalTok{causal)]),}
        \AttributeTok{scale=}\StringTok{"none"}\NormalTok{, }
        \AttributeTok{col =} \FunctionTok{cm.colors}\NormalTok{(}\FunctionTok{length}\NormalTok{(brks)}\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{), }
        \AttributeTok{breaks=}\NormalTok{brks,}
        \AttributeTok{dendrogram =} \StringTok{"both"}\NormalTok{,}
        \AttributeTok{Rowv=}\ConstantTok{TRUE}\NormalTok{, }
        \AttributeTok{trace=}\StringTok{"none"}\NormalTok{,}
        \AttributeTok{key.title =} \StringTok{"Coefficient in multiple}\SpecialCharTok{\textbackslash{}n}\StringTok{regression model"}\NormalTok{,}
        \AttributeTok{ylab=}\StringTok{"SNPs"}\NormalTok{,}
        \AttributeTok{cexCol=}\DecValTok{1}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{RDAtraitPredictionTutorial_files/figure-latex/unnamed-chunk-8-1.pdf}

Here is the information about the session when the tutorial was built:

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{sessionInfo}\NormalTok{()}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## R version 4.2.2 (2022-10-31)
## Platform: aarch64-apple-darwin20 (64-bit)
## Running under: macOS Ventura 13.1
## 
## Matrix products: default
## BLAS:   /Library/Frameworks/R.framework/Versions/4.2-arm64/Resources/lib/libRblas.0.dylib
## LAPACK: /Library/Frameworks/R.framework/Versions/4.2-arm64/Resources/lib/libRlapack.dylib
## 
## locale:
## [1] en_US.UTF-8/en_US.UTF-8/en_US.UTF-8/C/en_US.UTF-8/en_US.UTF-8
## 
## attached base packages:
## [1] stats     graphics  grDevices utils     datasets  methods   base     
## 
## other attached packages:
## [1] gplots_3.1.3    lfmm_1.1        LEA_3.10.2      vegan_2.6-4    
## [5] lattice_0.20-45 permute_0.9-7  
## 
## loaded via a namespace (and not attached):
##  [1] Rcpp_1.0.10        rstudioapi_0.14    knitr_1.42         cluster_2.1.4     
##  [5] splines_4.2.2      MASS_7.3-58.2      rlang_1.0.6        fastmap_1.1.0     
##  [9] foreach_1.5.2      highr_0.10         caTools_1.18.2     tools_4.2.2       
## [13] parallel_4.2.2     grid_4.2.2         nlme_3.1-162       mgcv_1.8-41       
## [17] xfun_0.37          KernSmooth_2.23-20 cli_3.6.0          gtools_3.9.4      
## [21] htmltools_0.5.4    iterators_1.0.14   yaml_2.3.7         digest_0.6.31     
## [25] Matrix_1.5-3       bitops_1.0-7       codetools_0.2-19   evaluate_0.20     
## [29] rmarkdown_2.20     compiler_4.2.2
\end{verbatim}

\end{document}

### 2023-03-20 index.Rmd
implementing new organizational structure to repo

--- 2023-02-13 index.Rmd
@@
 
-* [Choosing a population genomics sequencing approach](choosing-population-genetics.html)  
-* [Whole Genome Resequencing (WGS)](WGS_intro.html)  
+* [Choosing a population genomics sequencing approach](POP_01_choosing_population_genetics.html)  
+* [Whole Genome Resequencing (WGS)](POP_04_WGS_intro.html)  
   * [Low coverage WGS](https://github.com/therkildsen-lab/genomic-data-analysis/blob/master/lcwgs_data_analysis.md)
-* [Reduced Representation Sequencing](RADseq.html) 
-* [Poolseq](poolseq.html)
-* [Redundancy Analysis (RDA) Trait Prediction](RDAtraitPredictionTutorial.html)
+* [Reduced Representation Sequencing](POP_02_RADseq.html) 
+* [Poolseq](POP_03_poolseq.html)
+* [Redundancy Analysis (RDA) Trait Prediction](POP_05_RDAtraitPredictionTutorial.html)
 
@@
 
-* [Multifactorial RNAseq analyses](DGE_comparison_v2.html)  
+* [Multifactorial RNAseq analyses](FUN_01_DGE_comparison_v2.html)  
 * Align and count RNAseq reads (coming soon!)
@@
 # Panel Seminars 
-To facilitate discussions about practices and pitfalls in applying 'omics to non-model systems, we hosted [Q & A panel seminars](panels.html) with experts on topics ranging from RADseq, population genomics with WGS, and transcriptomics.
+To facilitate discussions about practices and pitfalls in applying 'omics to non-model systems, we hosted [Q & A panel seminars](ADMIN_03_panels.html) with experts on topics ranging from RADseq, population genomics with WGS, and transcriptomics.
 Check them out on [YouTube](https://www.youtube.com/channel/UCosG6JfIj7k0DizfsD5ADfQ)!
@@
 A diverse range of perspectives and backgrounds will make these tutorials more inclusive, informative, and useful for everyone in the marine-omics community.
-If you are interested in developing an information page or tutorial for the website,  we ask that you review our [page with instructions for submitting content](submissions_instructions.html) and contact our Website Manager Jason Johns (jasonjohns[at]ucsb.edu).
+If you are interested in developing an information page or tutorial for the website,  we ask that you review our [page with instructions for submitting content](ADMIN_01_submissions_instructions.html) and contact our Website Manager Jason Johns (jasonjohns[at]ucsb.edu).
 

### 2023-03-21 ADMIN_03_panels.Rmd

---
title: Panel Seminars
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

To facilitate discussions about practices and pitfalls in applying 'omics to non-model systems, we hosted Q & A panel seminars with experts on topics ranging from RADseq, population genomics with WGS, and transcriptomics in May-June 2021. 3-5 expert panelists discussed each topic, with questions coming from the working group and the audience. Videos are available on YouTube!

## Seminar Schedule {.unlisted .unnumbered}

| Date        | Topic       | Panelists     | Video     |
| :---        | :---        | :---          | :---      |
| May 7         | GBS/RADseq genotyping pipelines     | [Melanie LaCava](https://melanielacava.weebly.com/), [Alicia Mastretta-Yanes](http://mastrettayanes-lab.org/), [Isaac Overcast](https://isaacovercast.github.io/), [Jon Puritz](https://www.marineevoeco.com/)   | [View video](https://www.youtube.com/watch?v=C74GBESeIq4)     |
| May 21       |  WGS genotyping (low and high coverage)     | [Jeffrey Good](http://www.thegoodlab.org/), [Mikhail "Misha" Matz](https://matzlab.weebly.com/), [Noah Reid](http://nreid.github.io/), [Tim Sackton](https://scholar.harvard.edu/tsackton), [Nina Overgaard Therkildsen](https://www.therkildsenlab.com/) | [View Video](https://www.youtube.com/watch?v=ObJ3OT3dwSw)    |
| June 4       |  Population genomic analyses and filtering data     | [Zach Gompert](https://gompertlab.com/), [Morgan Kelly](http://www.morgankelly.biology.lsu.edu/), [Katie Lotterhos](https://sites.google.com/site/katielotterhos/home), [Jeff Ross-Ibarra](https://rilab.ucdavis.edu/) |  [View Video](https://youtu.be/dZyuqVEGPMo)   |
| June 18       |  RNAseq (experimental design through differential expression)     | [Ana Conesa](http://conesalab.org/), [Sarah Davies](http://sites.bu.edu/davieslab/), [Martin Hölzer](https://hoelzer-lab.github.io/), [Joanna Kelley](https://labs.wsu.edu/genomes/), [Harold Pimentel](https://pimentellab.com/)     |  [View Video](https://youtu.be/-gQ59m2cck0)   |

### 2023-03-21 POP_02_RADseq.Rmd

--- 2022-08-17 RADseq.Rmd
@@
 subtitle: '*Katherine Silliman, Danielle Davenport*'
-bibliography: RCN_site-Nov05.bib
+bibliography: POP_02_RADseq_files/common-bib_01.bib
 ---
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
 
@@
 
-![Table 1 from [@OLeary2018-iy].](images/oleary_table1.png)
+![Table 1 from [@OLeary2018-iy].](POP_02_RADseq_files/oleary_table1.png)
 
@@
 Based on the MarineOmics RAD panel, the general consensus is that all the popular pipelines can produce adequately accurate datasets (with appropriate parameter optimization and data filtering).     
-![Table 3 from LaCava et al 2019. CD-HIT is used in dDocent, VSEARCH is used in ipyrad.](images/lacavaTable3.png)
+![Table 3 from LaCava et al 2019. CD-HIT is used in dDocent, VSEARCH is used in ipyrad.](POP_02_RADseq_files/lacavaTable3.png)
 
@@
 ```{r}
-filename = "OL_subset" #replace with your file name 
-filename.gds = paste0("RAD_data/", paste0(filename, ".gds"))
-filename.vcf = paste0("RAD_data/", paste0(filename, ".vcf"))
+filename = "OL_subset" #replace with y our file name 
+filename.gds = paste0("POP_02_RADseq_files/", paste0(filename, ".gds"))
+filename.vcf = paste0("POP_02_RADseq_files/", paste0(filename, ".vcf"))
   # 1 . Convert VCF to GDS
@@
 Next we read in our metadata file, and make sure the samples are in the same order as your VCF file:  
-```{r}
-metafile = "RAD_data/OL.popmap"
+```{r, eval = T}
+metafile = "POP_02_RADseq_files/OL.popmap"
 sample.ids = seqGetData(gdsin, "sample.id")
@@
 **Missingness in R with SeqArray**
-```{r}
+```{r, eval = T}
 #using previously loaded gdsin object
@@
 ```{bash, eval = T}
-vcftools --vcf RAD_data/OL_subset.vcf --missing-indv --out RAD_data/OL_subset
+vcftools --vcf POP_02_RADseq_files/OL_subset.vcf --missing-indv --out POP_02_RADseq_files/OL_subset
 # sort the file by most missing data and print the top 10 samples  
-cat RAD_data/OL_subset.imiss | (read h; echo "$h"; sort -k5 -r) 
+cat POP_02_RADseq_files/OL_subset.imiss | (read h; echo "$h"; sort -k5 -r) 
 ```
@@
 #code from Jon Puritz
-mawk '!/IN/' RAD_data/OL_subset.imiss | cut -f5 > totalmissing
+mawk '!/IN/' POP_02_RADseq_files/OL_subset.imiss | cut -f5 > totalmissing
 gnuplot << \EOF
@@
 #open the gds file previously created, if not already open
-#filename.gds = "RAD_data/OL_subset.gds" #replace with your file name 
+#filename.gds = "POP_02_RADseq_files/OL_subset.gds" #replace with your file name 
 #gdsin = SeqArray::seqOpen(filename.gds)
@@
 # convert to vcf
-seqGDS2VCF(gdsin, "RAD_data/OL-miss-pc1-linkage-filt.vcf.gz")
+seqGDS2VCF(gdsin, "POP_02_RADseq_files/OL-miss-pc1-linkage-filt.vcf.gz")
 

### 2023-03-21 POP_03_poolseq.Rmd

--- 2022-02-04 poolseq.Rmd
@@
 date: "8/12/2021"
-bibliography: RCN_site-Nov05.bib
+bibliography: "POP_03_poolseq_files/common-bib_01.bib"
 ---
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
 ```
@@
 ```
-![Manhattan plot of mean Fst values for 10,000-bp windows of all the selected lines versus the pure Bodega population.](images/Fst_SvsBR_SDmap.png)
+![Manhattan plot of mean Fst values for 10,000-bp windows of all the selected lines versus the pure Bodega population.](POP_03_poolseq_files/Fst_SvsBR_SDmap.png)
 
@@
 ```
-![Principle coordinate analysis of allele frequencies for selected lines (red), control lines (blue), the pure Bodega population (white), and the pure San Diego population (black).](images/pi_PCA.png)
+![Principle coordinate analysis of allele frequencies for selected lines (red), control lines (blue), the pure Bodega population (white), and the pure San Diego population (black).](POP_03_poolseq_files/pi_PCA.png)
 

### 2023-03-21 DRAFT_DNA_methylation.Rmd
added YV revisions

--- 2023-02-27 DNA-methylation.Rmd
@@
 title: 'DNA Methylation Analyses'
-subtitle: '*Steven Roberts*, *Sam White*, and ............ '
-bibliography: DNA-Methylation.bib
+subtitle: '*Steven Roberts*, *Sam White*, and *Yaamini Venkataraman*'
+bibliography: DRAFT_DNA_methylation/DNA-Methylation.bib
 ---
@@
 ```{r meth-bp, echo = FALSE, out.width = "70%", fig.align = "center"}
-knitr::include_graphics("images/meth-bp-chart.png")
+knitr::include_graphics("DRAFT_DNA_methylation/meth-bp-chart.png")
 ```

### 2023-03-21 POP_04_WGS_intro.Rmd
removed unused images and fixed a couple file paths

@@
 
-* [Fastq-to-VCF SnakeMake pipeline](POP_06_fastq2vcf_june2022_update/POP_06_fastq2vcf_june2022_update.html): in-depth explanation of an automated short-read mapping and variant calling pipeline maintained by [Harvard Informatics](https://github.com/harvardinformatics/shortRead_mapping_variantCalling).
+* [Fastq-to-VCF SnakeMake pipeline](POP_06_fastq2vcf_june2022_update.html): in-depth explanation of an automated short-read mapping and variant calling pipeline maintained by [Harvard Informatics](https://github.com/harvardinformatics/shortRead_mapping_variantCalling).
 Useful if you plan to adopt this pre-packaged automated and parallelizable pipeline, and would like to understand its different components, but not necessarily change it substantially.
 
-* [Fastq-to-VCF workflow](POP_07_WGS_Full_Walkthrough/POP_07_WGS_Full_Walkthrough.html): detailed walkthrough of processing 15x sequencing depth WGS data for cod, from raw reads to a VCF.
+* [Fastq-to-VCF workflow](POP_07_WGS_Full_Walkthrough.html): detailed walkthrough of processing 15x sequencing depth WGS data for cod, from raw reads to a VCF.
 Useful if you would like to run each component of the pipeline yourself and potentially tweak some of them for your own purpose.

### 2023-03-21 index02.Rmd
beta testing landing page layout

--- 2023-03-20 index.Rmd
@@
 ---
-title: "Introduction"
+title: "Welcome"
 ---
-
-# Guidelines for 'omics analyses
 
@@
 
-Based on the expertise of the MarineOmics working group, the first topics include:
-
-### Topics {.tabset .tabset-fade .tabset-pills}
-#### Population Genomics
+# Population Genomics
 
@@
 
-#### Functional Genomics
+# Functional Genomics
 
@@
 
-#### Genome-Phenome
-
+# Genome-Phenome
 * coming soon!
-
-### {-}
-
-### The guidelines portion of the website should be considered a "preprint".
-If you are interested in contributing to the peer review process, check out the [Discussion Forum](discussion.html).
-{.unlisted .unnumbered}
 

### 2023-03-21 index02.Rmd
index beta

@@
 
-# What is the MarineOmics working group? {#marineomics}
+# How to contribute!
+We are excited to incorporate new material from contributors of various backgrounds, expertise, and experience.
+To date these tutorials, guidelines,  and discussions have been authored and led by graduate students and postdocs with contributions from scientists at all levels, including undergraduates, technicians, and faculty members.
+A diverse range of perspectives and backgrounds will make these tutorials more inclusive, informative, and useful for everyone in the marine-omics community.
+If you are interested in developing an information page or tutorial for the website,  we ask that you review our [page with instructions for submitting content](ADMIN_01_submissions_instructions.html) and contact our Website Manager Jason Johns (jasonjohns[at]ucsb.edu).
 
+# Background
 The original MarineOmics working group is part of the [RCN for Evolution in Changing Seas](https://rcn-ecs.github.io/), and was formed to promote **reproducible** and **robust** genomic research in nonmodel species, but especially marine and aquatic studies.
@@
 
-# Reproducibility in Marine Genomics  
 MarineOmics is also interested in evaluating and improving reproducibility in studies that use genomics to study adaptation in marine systems.
@@
 
-# Making contributions to the MarineOmics.github.io website {#contributions}
-We are excited to incorporate new material from contributors of various backgrounds, expertise, and experience.
-To date these tutorials, guidelines,  and discussions have been authored and led by graduate students and postdocs with contributions from scientists at all levels, including undergraduates, technicians, and faculty members.
-A diverse range of perspectives and backgrounds will make these tutorials more inclusive, informative, and useful for everyone in the marine-omics community.
-If you are interested in developing an information page or tutorial for the website,  we ask that you review our [page with instructions for submitting content](ADMIN_01_submissions_instructions.html) and contact our Website Manager Jason Johns (jasonjohns[at]ucsb.edu).
-
-# Internship 
 In Spring 2021, MarineOmics hosted three virtual undergraduate research interns with funding from the [RCN-ECS](https://rcn-ecs.github.io/): Jecy Klinkam, Kaitlin Macaranas, and Chris Mantegna.
@@
 
-# Contact {#contact}
-For questions about the website and its materials, post to the [Discussion Forum](https://github.com/MarineOmics/marineomics.github.io/discussions)!
-For inquiries about the MarineOmics Working Group activities, please contact us at rcnecs.marineomics@gmail.com.
+Original working group members
+
+#### [Katherine Silliman, PhD](https://ksilliman.weebly.com/), South Carolina Dept. of Natural Resources - sillimank[at]dnr.sc.gov {.unlisted .unnumbered}
+#### [Sam Bogan](http://www.hofmannlab.com/sam-bogan.html), UC-Santa Barbara -snbogan[at]ucsb.edu {.unlisted .unnumbered}
+#### [Alan Downey-Wall](https://cos.northeastern.edu/people/alan-downey-wall/), Northeastern University {.unlisted .unnumbered}
+#### Chris Mantegna, University of Washington (Intern)  {.unlisted .unnumbered}
+#### [Danielle Davenport](https://www.researchgate.net/profile/Danielle-Davenport-2), University of Queensland {.unlisted .unnumbered}  
+#### [Hanny Rivera, PhD](https://www2.whoi.edu/staff/hrivera/), Woods Hole Oceanographic Institute  {.unlisted .unnumbered}
+#### Jecy Klinkam, College of William and Mary (Intern) {.unlisted .unnumbered}
+#### [Joanna Griffiths, PhD](https://joannasgriffiths.wordpress.com/), UC-Davis {.unlisted .unnumbered}
+#### Jason Johns, UC-Santa Barbara (Website Manager) {.unlisted .unnumbered}
+#### [Joe McGirr, PhD](https://joemcgirr.github.io/), UC-Davis {.unlisted .unnumbered}
+#### Kaitlin Macaranas, CSU-Bakersfield (Intern) {.unlisted .unnumbered}
+#### Megan Guidry, University of Rhode Island {.unlisted .unnumbered}
+#### Natalie Ameral, University of Rhode Island & RI Dept. of Environmental Management {.unlisted .unnumbered}  
+#### Nicolas Lou, Cornell University {.unlisted .unnumbered}
+#### [Sara Schaal](https://schaals.wixsite.com/smschaal), Northeastern University {.unlisted .unnumbered} 
+#### [Sara Wuitchik, PhD](https://sjswuitchik.weebly.com/), Harvard University {.unlisted .unnumbered}
+
+### Faculty Advisors {.unlisted .unnumbered}
+#### [Andrew Whitehead, PhD](https://whiteheadresearch.wordpress.com/), UC-Davis {.unlisted .unnumbered}
+#### [Jon Puritz, PhD](https://www.marineevoeco.com/), University of Rhode Island {.unlisted .unnumbered}
+#### [Steven Roberts, PhD](http://faculty.washington.edu/sr320/?page_id=10850), University of Washington {.unlisted .unnumbered}

### 2023-03-22 index02.Rmd
menu mod
added panel seminars to index02

@@
 # Panel Seminars
-To facilitate discussions about practices and pitfalls in applying 'omics to non-model systems, we hosted [Q & A panel seminars](ADMIN_03_panels.html) with experts on topics ranging from RADseq, population genomics with WGS, and transcriptomics.
-Check them out on [YouTube](https://www.youtube.com/channel/UCosG6JfIj7k0DizfsD5ADfQ)!
 
-# How to contribute!
+To facilitate discussions about practices and pitfalls in applying 'omics to non-model systems, we hosted Q & A panel seminars with experts on topics ranging from RADseq, population genomics with WGS, and transcriptomics in May-June 2021.
+3-5 expert panelists discussed each topic, with questions coming from the working group and the audience.
+Videos are available on YouTube!
+
+| Date        | Topic       | Panelists     | Video     |
+| :---        | :---        | :---          | :---      |
+| May 7         | GBS/RADseq genotyping pipelines     | [Melanie LaCava](https://melanielacava.weebly.com/), [Alicia Mastretta-Yanes](http://mastrettayanes-lab.org/), [Isaac Overcast](https://isaacovercast.github.io/), [Jon Puritz](https://www.marineevoeco.com/)   | [View video](https://www.youtube.com/watch?v=C74GBESeIq4)     |
+| May 21       |  WGS genotyping (low and high coverage)     | [Jeffrey Good](http://www.thegoodlab.org/), [Mikhail "Misha" Matz](https://matzlab.weebly.com/), [Noah Reid](http://nreid.github.io/), [Tim Sackton](https://scholar.harvard.edu/tsackton), [Nina Overgaard Therkildsen](https://www.therkildsenlab.com/) | [View Video](https://www.youtube.com/watch?v=ObJ3OT3dwSw)    |
+| June 4       |  Population genomic analyses and filtering data     | [Zach Gompert](https://gompertlab.com/), [Morgan Kelly](http://www.morgankelly.biology.lsu.edu/), [Katie Lotterhos](https://sites.google.com/site/katielotterhos/home), [Jeff Ross-Ibarra](https://rilab.ucdavis.edu/) |  [View Video](https://youtu.be/dZyuqVEGPMo)   |
+| June 18       |  RNAseq (experimental design through differential expression)     | [Ana Conesa](http://conesalab.org/), [Sarah Davies](http://sites.bu.edu/davieslab/), [Martin Hölzer](https://hoelzer-lab.github.io/), [Joanna Kelley](https://labs.wsu.edu/genomes/), [Harold Pimentel](https://pimentellab.com/)     |  [View Video](https://youtu.be/-gQ59m2cck0)   |
+
+# How to Contribute
 We are excited to incorporate new material from contributors of various backgrounds, expertise, and experience.

### 2023-03-22 DRAFT_Funct_Enrich.Rmd
Add GOMWU directory and code

--- 2023-02-10 Funct_Enrich_files/Funct_Enrich.Rmd
@@
 subtitle: '*Thea Gessler*, *Samuel N. Bogan*, and *Steven Roberts*'
-bibliography: Funct_Enrich.bib
+bibliography: DRAFT_Funct_Enrich/Funct_Enrich.bib
 ---
@@
 
-knitr::include_graphics("images/Rivera_etal_fig.png")
+knitr::include_graphics("DRAFT_Funct_Enrich/Rivera_etal_fig.png")
 
@@
 
-```{r setup, include = FALSE}
+    ```{r setup, include = FALSE} `r ''`
 
@@
 
+# Run GOMWU and Fisher's exact tests on data of interest (see https://github.com/z0on/GO_MWU)
+
+```{r, include = TRUE}
+
+# GO enrichments using Mann Whitney U tests
+setwd("GO_MWU-master/")
+
+#Now we multiply them together into a new column
+
+## Ok we also want to bring in our annotations
+#I would suggest putting a GO_MWU folder into your working directory for these steps. There is some stuff to download such as the obo files 
+#See https://github.com/z0on/GO_MWU
+
+# GO_MWU uses continuous measure of significance (such as fold-change or -log(p-value) ) to identify GO categories that are significantly enriches with either up- or down-regulated genes.
+The advantage - no need to impose arbitrary significance cutoff.
+# If the measure is binary (0 or 1) the script will perform a typical "GO enrichment" analysis based Fisher's exact test: it will show GO categories over-represented among the genes that have 1 as their measure.
+# On the plot, different fonts are used to indicate significance and color indicates enrichment with either up (red) or down (blue) regulated genes. No colors are shown for binary measure analysis.
+# The tree on the plot is hierarchical clustering of GO categories based on shared genes. Categories with no branch length between them are subsets of each other.
+# The fraction next to GO category name indicates the fracton of "good" genes in it; "good" genes being the ones exceeding the arbitrary absValue cutoff (option in gomwuPlot).
+For Fisher's based test, specify absValue=0.5.
+This value does not affect statistics and is used for plotting only.
+# Stretch the plot manually to match tree to text
+
+# Mikhail V. Matz, UT Austin, February 2015; matz@utexas.edu
+
+################################################################
+# First, press command-D on mac or ctrl-shift-H in Rstudio and navigate to the directory containing scripts and input files. Then edit, mark and execute the following bits of code, one after another.
+
+# Edit these to match your data file names: 
+input = "GOMWU_input.csv" # two columns of comma-separated values: gene id, continuous measure of significance.
+To perform standard GO enrichment analysis based on Fisher's exact test, use binary measure (0 or 1, i.e., either sgnificant or not).
+goAnnotations = "goAnnot_spu.tab" # two-column, tab-delimited, one line per gene, multiple GO terms separated by semicolon.
+If you have multiple lines per gene, use nrify_GOtable.pl prior to running this script.
+goDatabase = "go.obo" # download from http://www.geneontology.org/GO.downloads.ontology.shtml
+goDivision = "MF" # either MF, or BP, or CC
+source("gomwu.functions.R")
+
+# Calculating stats.
+It might take ~3 min for MF and BP.
+Do not rerun it if you just want to replot the data with different cutoffs, go straight to gomwuPlot.
+If you change any of the numeric values below, delete the files that were generated in previos runs first.
+gomwuStats(input, goDatabase, goAnnotations, goDivision,
+	perlPath = "perl", # replace with full path to perl executable if it is not in your system's PATH already
+	largest = 0.1, # a GO category will not be considered if it contains more than this fraction of the total number of genes
+	smallest = 25,   # a GO category should contain at least this many genes to be considered
+	clusterCutHeight = 0.25, # threshold for merging similar (gene-sharing) terms. See README for details.
+	Alternative = "g") # # threshold for merging similar (gene-sharing) terms. See README for details.
+           #	Alternative="g" # by default the MWU test is two-tailed; specify "g" or "l" of you want to test for "greater" or "less" instead. 
+           #	Module=TRUE,Alternative="g" # un-remark this if you are analyzing a SIGNED WGCNA module (values: 0 for not in module genes, kME for in-module genes).
+In the call to gomwuPlot below, specify absValue=0.001 (count number of "good genes" that fall into the module)
+           #	Module=TRUE # un-remark this if you are analyzing an UNSIGNED WGCNA module
+# do not continue if the printout shows that no GO terms pass 10% FDR.
+
+quartz.save(file = "quartz.png", type = "png")
+results=gomwuPlot(input, goAnnotations, goDivision,
+#	absValue=-log(0.05,10),  # genes with the measure value exceeding this will be counted as "good genes".
+Specify absValue=0.001 if you are doing Fisher's exact test for standard GO enrichment or analyzing a WGCNA module (all non-zero genes = "good genes").
+	absValue= .2,
+	level1=0.01, # FDR threshold for plotting. Specify level1=1 to plot all GO categories containing genes exceeding the absValue.
+	level2=0.005, # FDR cutoff to print in regular (not italic) font.
+	level3=0.001, # FDR cutoff to print in large bold font.
+	txtsize=1.2,    # decrease to fit more on one page, or increase (after rescaling the plot so the tree fits the text) for better "word cloud" effect
+	treeHeight=0.5 # height of the hierarchical clustering tree
+#	colors=c("dodgerblue2","firebrick1","skyblue","lightcoral") # these are default colors, un-remar and change if needed
+)
+# manually rescale the plot so the tree matches the text 
+# if there are too many categories displayed, try make it more stringent with level1=0.05,level2=0.01,level3=0.001.
+
+# text representation of results, with actual adjusted p-values
+results
+
+# Export list of GO terms enriched among genes in lowest quartiles of TSS acess and expression
+write.csv(results[1], 
+          "~/Documents/GitHub/marineomics.github.io/DRAFT_Funct_Enrich/MF_output.csv", 
+          row.names = TRUE)
+
+# Next, manually repeat this code chunk using BP and CC terms and change exported name of file
+
+```
+
 # References

### 2023-03-22 DRAFT_Funct_Enrich/GO_MWU-master/GO_MWU.R

@@ -1,0 +1,98 @@
+
+# GO_MWU uses continuous measure of significance (such as fold-change or -log(p-value) ) to identify GO categories that are significantly enriches with either up- or down-regulated genes. The advantage - no need to impose arbitrary significance cutoff.
+
+# If the measure is binary (0 or 1) the script will perform a typical "GO enrichment" analysis based Fisher's exact test: it will show GO categories over-represented among the genes that have 1 as their measure. 
+
+# On the plot, different fonts are used to indicate significance and color indicates enrichment with either up (red) or down (blue) regulated genes. No colors are shown for binary measure analysis.
+
+# The tree on the plot is hierarchical clustering of GO categories based on shared genes. Categories with no branch length between them are subsets of each other.
+
+# The fraction next to GO category name indicates the fracton of "good" genes in it; "good" genes being the ones exceeding the arbitrary absValue cutoff (option in gomwuPlot). For Fisher's based test, specify absValue=0.5. This value does not affect statistics and is used for plotting only.
+
+# Stretch the plot manually to match tree to text
+
+# Mikhail V. Matz, UT Austin, February 2015; matz@utexas.edu
+
+################################################################
+# First, press command-D on mac or ctrl-shift-H in Rstudio and navigate to the directory containing scripts and input files. Then edit, mark and execute the following bits of code, one after another.
+
+
+# Edit these to match your data file names: 
+input="heats.csv" # two columns of comma-separated values: gene id, continuous measure of significance. To perform standard GO enrichment analysis based on Fisher's exact test, use binary measure (0 or 1, i.e., either sgnificant or not).
+goAnnotations="amil_defog_iso2go.tab" # two-column, tab-delimited, one line per gene, multiple GO terms separated by semicolon. If you have multiple lines per gene, use nrify_GOtable.pl prior to running this script.
+goDatabase="go.obo" # download from http://www.geneontology.org/GO.downloads.ontology.shtml
+goDivision="MF" # either MF, or BP, or CC
+source("gomwu.functions.R")
+
+
+# ------------- Calculating stats
[…]

### 2023-03-22 DRAFT_Funct_Enrich/GO_MWU-master/GO_deltaRanks_correlation.R

@@ -1,0 +1,32 @@
+#setwd("~/Dropbox/Documents/dixon_2017_RT-GBM/reciprocal_transplant_methylationV2/clean_july12/scripts")
+
+data1=read.table("MWU_MF_heat_lpv.csv",header=T)
+data2=read.table("MWU_MF_surv_lpv.csv",header=T)
+
+goods=intersect(data1$term,data2$term)
+#goods=unique(as.character(c(data1$term[data1$p.adj<=0.1],data2$term[data2$p.adj<=0.1])))
+length(goods)
+
+data1=data1[data1$term %in% goods,]
+data2=data2[data2$term %in% goods,]
+
+# all overlapping GO terms
+ress=merge(data1,data2,by="term")
+plot(delta.rank.x~delta.rank.y,ress,xlab="GBM", ylab="GE",mgp=c(2.3,1,0))
+abline(v=0,lty=3)
+abline(h=0,lty=3)
+
+# GO terms highly signifcant in any of the two datasets
+sigs=(ress$p.adj.x<=0.01 | ress$p.adj.y<=0.01)
+sum(sigs) # 71
+plot(delta.rank.x~delta.rank.y,ress[sigs,],xlab="GBM", ylab="GE",mgp=c(2.3,1,0))
+abline(v=0,lty=3)
+abline(h=0,lty=3)
+
+# GO terms signifcant in both datasets
+sigs=(ress$p.adj.x<=0.1 & ress$p.adj.y<=0.1)
+sum(sigs) # 20
+plot(delta.rank.x~delta.rank.y,ress[sigs,],xlab="GBM", ylab="GE",mgp=c(2.3,1,0))
+abline(v=0,lty=3)
+abline(h=0,lty=3)
+

### 2023-03-22 DRAFT_Funct_Enrich.Rmd
added background on FET

@@
 ---
+
+# Background
+
+Functional gene enrichment analysis is a powerful tool that can aid in the interpretation of biological data by identifying biological processes and pathways that are significantly overrepresented in a gene list of interest.
+However, to effectively use this tool, it is essential to have a proper annotation of genes.
+
+Proper annotation of genes is crucial because it ensures that the genes are correctly identified and annotated with their appropriate biological functions.
+Incorrect annotation of genes can lead to inaccurate interpretations of experimental results and hinder progress in understanding the biological systems of non-model species.
+
+Therefore, the quality of gene annotation is critical for accurate functional gene enrichment analysis.
+This involves careful curation of gene annotations based on experimental evidence from various sources, including transcriptomic and proteomic data, literature mining, and functional assays.
+This curation process can be particularly challenging for non-model species with limited genomic and transcriptomic resources.
+More information on this can be found at
+
+- Example 1
+- Example 2
+
+**In some instances high quality annotation might not be possible but it is critical that before you undergo enrichment analysis you have a full understanding of how your data set was annotated, and any associated limitations**
 

### 2023-03-23 DRAFT_04_best_principles.Rmd
new draft of best practices page

---
title: "Best Principles in Genomics Research"
bibliography: "DRAFT_04_best_principles_files/common-bib_01.bib"
_disableToc: true
editor_options: 
  markdown: 
    wrap: 72
---

```{r #setup, echo = FALSE, message = FALSE}
library(knitr)
knitr::opts_chunk$set(
	fig.height = 5,
	fig.width = 10,
	cache = TRUE
)
library(knitcitations)
library(kableExtra)
opts_chunk$set(fig.width = 10,
               fig.height = 5,
               cache = TRUE)
cite_options(citation_format = "pandoc", max.names = 3, style = "html", 
             hyperlink = "to.doc")
```

The backbone underlying the content on the website comes from a general set of “best principles” that should be applied in genomics studies, irrespective of the specific sequencing method used. Instead of providing rigid “best practices", the website promotes “best principles”, defined as a guiding set of values and goals that can be tailored to one’s hypotheses and guided by one's data, as opposed to a specific set of instructions. Motivated by rigor and reproducibility, "best principles" are designed to encourage data exploration and critical thinking during analysis and evaluation instead of ticking off boxes on a protocol or step-list. These principles are outlined below.

<br>

## Rigor

### 1. Understand the characteristics of your chosen sequencing approach. Take these characteristics into account when designing a study and during data analysis. {.unlisted .unnumbered}

Goals of study should be chosen before choosing the best sequencing approach, which will inform the total number of samples and sequencing coverage needed: e.g., PoolSeq requires larger sample sizes and deeper coverage given the lack of individual genotyping [@Guirao-Rico2021-sw].

<br>

### 2. Plot your data early and often. Get to know it in both its raw and processed forms. {.unlisted .unnumbered}

Deepen the interpretation of results and flag sources of error throughout a workflow by plotting data such as (i) read-quality metrics pre- and post-filtering, (ii) sequence coverage across a reference and across samples, (iii) principal component analysis of replicates pre- and post-filtering, and (iv) results and predictions of statistical tests.

<br>

### 3. All models and pipelines introduce some type and magnitude of error. Compare models’ nuances to find the best approach given your data. {.unlisted .unnumbered}

This issue is particularly acute in non-model species. Some quantitative approaches towards     evaluating methods include (i) comparing the performance of different methods or parameter choices using simulated data [@Lotterhos2022-ff], (ii) measuring their predictive strengths using model selection statistics [[@Hooten2015-ts], [@Johnson2004-sy]], and (iii) observed-predicted plots from model outputs. A basic understanding of the sensitivity of inference in different analyses will be helpful for determining how robust the results are to nuanced decisions, especially for non-model organisms or unique experimental designs.

<br>

## Reproducibility

### 4. Wherever your sequencing data go, their associated metadata go with them. {.unlisted .unnumbered}

Any and all metadata that can be reported should accompany sequence data in databases such as NCBI or SRA. Data on Dryad or GitHub should crosslink to NCBI/SRA.

### 5. Take detailed records on all analysis decisions you make, including for preliminary analyses and errors that occurred, so you remember what you did and can reproduce your own work. {.unlisted .unnumbered}

Use text-annotated code notebooks for bioinformatic analyses (e.g., Rmarkdown, Jupyter).

### 6. Provide a reproducible text-annotated code notebook for all final analyses, containing computing environment information (i.e. software versions), so that these methods could be reproduced by someone else. {.unlisted .unnumbered}

Provide these notebooks (in Rmarkdown or Jupyter) in a publically accessible format on services such as GitHub, GitLab, Dryad, Figshare, or Zenodo.

<br>

## References

### 2023-03-23 index02.Rmd
adding text for best principles page on index, added best principles to header, and all appropriate links, re-rendered site

@@
 On this website you will find a growing and evolving collection of guidelines, tutorials, recommended readings, and method evaluations for various 'omics analyses that are commonly used in studies of nonmodel systems (particularly marine species).
-Instead of providing rigid "best practices", we aim to encourage data exploration and rigorous quality control through "best principles", paying particular attention to topics for which open access guidelines are scarce.
+Instead of providing rigid "best practices", we aim to encourage data exploration and rigorous quality control through ["best principles"](#best_principles), paying particular attention to topics for which open access guidelines are scarce.
 The guidelines are written by members of the [MarineOmics working group](#marineomics) or the community, then shared for peer review.
 Comments on current website content or suggestions for new content can be made through the [Discussion Forum](https://github.com/MarineOmics/marineomics.github.io/discussions).
+
+# Best Principles <a id="best_principles"></a>
+
+The backbone underlying the content on the website comes from a general set of “best principles” that should be applied in genomics studies, irrespective of the specific sequencing method used.
+Instead of providing rigid “best practices", the website promotes “best principles”, defined as a guiding set of values and goals that can be tailored to one’s hypotheses and guided by one's data, as opposed to a specific set of instructions.
+Motivated by rigor and reproducibility, "best principles" are designed to encourage data exploration and critical thinking during analysis and evaluation instead of ticking off boxes on a protocol or step-list.
+Please visit our [best principles page](DRAFT_04_best_principles.html) for a detailed outline.
 

### 2023-03-24 ADMIN_04_best_principles.Rmd
dialing in navigation to contributions & best practices with new index landing page

--- 2023-03-23 DRAFT_04_best_principles.Rmd
@@
 title: "Best Principles in Genomics Research"
-bibliography: "DRAFT_04_best_principles_files/common-bib_01.bib"
+bibliography: "ADMIN_04_best_principles_files/common-bib_01.bib"
 _disableToc: true

### 2023-03-24 index.Rmd
dialing in navigation to contributions & best practices with new index landing page

@@
 ---
-title: "Introduction"
+title: "Welcome"
 ---
 
-# Guidelines for 'omics analyses
-
 On this website you will find a growing and evolving collection of guidelines, tutorials, recommended readings, and method evaluations for various 'omics analyses that are commonly used in studies of nonmodel systems (particularly marine species).
-Instead of providing rigid "best practices", we aim to encourage data exploration and rigorous quality control through "best principles", paying particular attention to topics for which open access guidelines are scarce.
-The guidelines are written by members of the [MarineOmics working group](#marineomics) or the community, then shared for peer review.
+Instead of providing rigid "best practices", we aim to encourage data exploration and rigorous quality control through ["best principles"](#best_principles), paying particular attention to topics for which open access guidelines are scarce.
+The guidelines are written by members of the [MarineOmics working group](#marineomics) and/or the community, then shared for community input.
 Comments on current website content or suggestions for new content can be made through the [Discussion Forum](https://github.com/MarineOmics/marineomics.github.io/discussions).
 
-Based on the expertise of the MarineOmics working group, the first topics include:
+# Best Principles <a id="best_principles"></a>
 
-### Topics {.tabset .tabset-fade .tabset-pills}
-#### Population Genomics
+The backbone underlying the content on the website comes from a general set of “best principles” that should be applied in genomics studies, irrespective of the specific sequencing method used.
+Instead of providing rigid “best practices", the website promotes “best principles”, defined as a guiding set of values and goals that can be tailored to one’s hypotheses and guided by one's data, as opposed to a specific set of instructions.
+Motivated by rigor and reproducibility, "best principles" are designed to encourage data exploration and critical thinking during analysis and evaluation instead of ticking off boxes on a protocol or step-list.
+Please visit our [best principles page](ADMIN_04_best_principles.html) for a detailed outline.
+
+# Population Genomics
 
@@
 
-#### Genome-Phenome
-
+# Genome-Phenome
 * coming soon!
 
-### {-}
+# Panel Seminars
 
-### The guidelines portion of the website should be considered a "preprint".
-If you are interested in contributing to the peer review process, check out the [Discussion Forum](discussion.html).
-{.unlisted .unnumbered}
+To facilitate discussions about practices and pitfalls in applying 'omics to non-model systems, we hosted Q & A panel seminars with experts on topics ranging from RADseq, population genomics with WGS, and transcriptomics in May-June 2021.
+3-5 expert panelists discussed each topic, with questions coming from the working group and the audience.
+Videos are available on YouTube!
 
-# Panel Seminars 
-To facilitate discussions about practices and pitfalls in applying 'omics to non-model systems, we hosted [Q & A panel seminars](ADMIN_03_panels.html) with experts on topics ranging from RADseq, population genomics with WGS, and transcriptomics.
-Check them out on [YouTube](https://www.youtube.com/channel/UCosG6JfIj7k0DizfsD5ADfQ)!
+| Date        | Topic       | Panelists     | Video     |
+| :---        | :---        | :---          | :---      |
+| May 7         | GBS/RADseq genotyping pipelines     | [Melanie LaCava](https://melanielacava.weebly.com/), [Alicia Mastretta-Yanes](http://mastrettayanes-lab.org/), [Isaac Overcast](https://isaacovercast.github.io/), [Jon Puritz](https://www.marineevoeco.com/)   | [View video](https://www.youtube.com/watch?v=C74GBESeIq4)     |
+| May 21       |  WGS genotyping (low and high coverage)     | [Jeffrey Good](http://www.thegoodlab.org/), [Mikhail "Misha" Matz](https://matzlab.weebly.com/), [Noah Reid](http://nreid.github.io/), [Tim Sackton](https://scholar.harvard.edu/tsackton), [Nina Overgaard Therkildsen](https://www.therkildsenlab.com/) | [View Video](https://www.youtube.com/watch?v=ObJ3OT3dwSw)    |
+| June 4       |  Population genomic analyses and filtering data     | [Zach Gompert](https://gompertlab.com/), [Morgan Kelly](http://www.morgankelly.biology.lsu.edu/), [Katie Lotterhos](https://sites.google.com/site/katielotterhos/home), [Jeff Ross-Ibarra](https://rilab.ucdavis.edu/) |  [View Video](https://youtu.be/dZyuqVEGPMo)   |
+| June 18       |  RNAseq (experimental design through differential expression)     | [Ana Conesa](http://conesalab.org/), [Sarah Davies](http://sites.bu.edu/davieslab/), [Martin Hölzer](https://hoelzer-lab.github.io/), [Joanna Kelley](https://labs.wsu.edu/genomes/), [Harold Pimentel](https://pimentellab.com/)     |  [View Video](https://youtu.be/-gQ59m2cck0)   |
 
-# What is the MarineOmics working group? {#marineomics}
+# How to Contribute
+We are excited to incorporate new material from contributors of various backgrounds, expertise, and experience.
+To date these tutorials, guidelines,  and discussions have been authored and led by graduate students and postdocs with contributions from scientists at all levels, including undergraduates and faculty members.
+If you are interested in developing an information page or tutorial for the website, we ask that you review our [page with instructions for submitting content](ADMIN_01_submissions_instructions.html) and contact our Website Manager Jason Johns (jasonjohns[at]ucsb.edu).
 
+# Background
 The original MarineOmics working group is part of the [RCN for Evolution in Changing Seas](https://rcn-ecs.github.io/), and was formed to promote **reproducible** and **robust** genomic research in nonmodel species, but especially marine and aquatic studies.
@@
 MarineOmics aims to support robust genomic research in ecology and evolution by disseminating guidelines through this dynamic, open-sourced website addressing introductory and advanced topics of population- and functional genomics.
-*We are always looking for folks to join and write content for the website, if interested please [contact us](#contact) and see the [below section](#contributions)!*
 

### 2023-03-24 index_old.Rmd
dialing in navigation to contributions & best practices with new index landing page

--- 2023-03-21 index02.Rmd
@@
 ---
-title: "Welcome"
+title: "Introduction"
 ---
 
+# Guidelines for 'omics analyses
+
 On this website you will find a growing and evolving collection of guidelines, tutorials, recommended readings, and method evaluations for various 'omics analyses that are commonly used in studies of nonmodel systems (particularly marine species).
-Instead of providing rigid "best practices", we aim to encourage data exploration and rigorous quality control through "best principles", paying particular attention to topics for which open access guidelines are scarce.
+Instead of providing rigid "best practices", we aim to encourage data exploration and rigorous quality control through ["best principles"](#best_principles), paying particular attention to topics for which open access guidelines are scarce.
 The guidelines are written by members of the [MarineOmics working group](#marineomics) or the community, then shared for peer review.
@@
 
-# Population Genomics
+# Best Principles <a id="best_principles"></a>
+
+The backbone underlying the content on the website comes from a general set of “best principles” that should be applied in genomics studies, irrespective of the specific sequencing method used.
+Instead of providing rigid “best practices", the website promotes “best principles”, defined as a guiding set of values and goals that can be tailored to one’s hypotheses and guided by one's data, as opposed to a specific set of instructions.
+Motivated by rigor and reproducibility, "best principles" are designed to encourage data exploration and critical thinking during analysis and evaluation instead of ticking off boxes on a protocol or step-list.
+Please visit our [best principles page](DRAFT_04_best_principles.html) for a detailed outline.
+
+Based on the expertise of the MarineOmics working group, the first topics include:
+
+### Topics {.tabset .tabset-fade .tabset-pills}
+#### Population Genomics
 
@@
 
-# Functional Genomics
+#### Functional Genomics
 
@@
 
-# Genome-Phenome
+#### Genome-Phenome
+
 * coming soon!
+
+### {-}
+
+### The guidelines portion of the website should be considered a "preprint".
+If you are interested in contributing to the peer review process, check out the [Discussion Forum](https://github.com/MarineOmics/marineomics.github.io/discussions).
+{.unlisted .unnumbered}
 
@@
 
-# How to contribute!
-We are excited to incorporate new material from contributors of various backgrounds, expertise, and experience.
-To date these tutorials, guidelines,  and discussions have been authored and led by graduate students and postdocs with contributions from scientists at all levels, including undergraduates, technicians, and faculty members.
-A diverse range of perspectives and backgrounds will make these tutorials more inclusive, informative, and useful for everyone in the marine-omics community.
-If you are interested in developing an information page or tutorial for the website,  we ask that you review our [page with instructions for submitting content](ADMIN_01_submissions_instructions.html) and contact our Website Manager Jason Johns (jasonjohns[at]ucsb.edu).
+# What is the MarineOmics working group? {#marineomics}
 
-# Background
 The original MarineOmics working group is part of the [RCN for Evolution in Changing Seas](https://rcn-ecs.github.io/), and was formed to promote **reproducible** and **robust** genomic research in nonmodel species, but especially marine and aquatic studies.
@@
 
+# Reproducibility in Marine Genomics  
 MarineOmics is also interested in evaluating and improving reproducibility in studies that use genomics to study adaptation in marine systems.
@@
 
+# Making contributions to the MarineOmics.github.io website {#contributions}
+We are excited to incorporate new material from contributors of various backgrounds, expertise, and experience.
+To date these tutorials, guidelines,  and discussions have been authored and led by graduate students and postdocs with contributions from scientists at all levels, including undergraduates, technicians, and faculty members.
+A diverse range of perspectives and backgrounds will make these tutorials more inclusive, informative, and useful for everyone in the marine-omics community.
+If you are interested in developing an information page or tutorial for the website,  we ask that you review our [page with instructions for submitting content](ADMIN_01_submissions_instructions.html) and contact our Website Manager Jason Johns (jasonjohns[at]ucsb.edu).
+
+# Internship 
 In Spring 2021, MarineOmics hosted three virtual undergraduate research interns with funding from the [RCN-ECS](https://rcn-ecs.github.io/): Jecy Klinkam, Kaitlin Macaranas, and Chris Mantegna.
@@
 
-Original working group members
-
-#### [Katherine Silliman, PhD](https://ksilliman.weebly.com/), South Carolina Dept. of Natural Resources - sillimank[at]dnr.sc.gov {.unlisted .unnumbered}
-#### [Sam Bogan](http://www.hofmannlab.com/sam-bogan.html), UC-Santa Barbara -snbogan[at]ucsb.edu {.unlisted .unnumbered}
-#### [Alan Downey-Wall](https://cos.northeastern.edu/people/alan-downey-wall/), Northeastern University {.unlisted .unnumbered}
-#### Chris Mantegna, University of Washington (Intern)  {.unlisted .unnumbered}
-#### [Danielle Davenport](https://www.researchgate.net/profile/Danielle-Davenport-2), University of Queensland {.unlisted .unnumbered}  
-#### [Hanny Rivera, PhD](https://www2.whoi.edu/staff/hrivera/), Woods Hole Oceanographic Institute  {.unlisted .unnumbered}
-#### Jecy Klinkam, College of William and Mary (Intern) {.unlisted .unnumbered}
-#### [Joanna Griffiths, PhD](https://joannasgriffiths.wordpress.com/), UC-Davis {.unlisted .unnumbered}
-#### Jason Johns, UC-Santa Barbara (Website Manager) {.unlisted .unnumbered}
-#### [Joe McGirr, PhD](https://joemcgirr.github.io/), UC-Davis {.unlisted .unnumbered}
-#### Kaitlin Macaranas, CSU-Bakersfield (Intern) {.unlisted .unnumbered}
-#### Megan Guidry, University of Rhode Island {.unlisted .unnumbered}
-#### Natalie Ameral, University of Rhode Island & RI Dept. of Environmental Management {.unlisted .unnumbered}  
-#### Nicolas Lou, Cornell University {.unlisted .unnumbered}
-#### [Sara Schaal](https://schaals.wixsite.com/smschaal), Northeastern University {.unlisted .unnumbered} 
-#### [Sara Wuitchik, PhD](https://sjswuitchik.weebly.com/), Harvard University {.unlisted .unnumbered}
-
-### Faculty Advisors {.unlisted .unnumbered}
-#### [Andrew Whitehead, PhD](https://whiteheadresearch.wordpress.com/), UC-Davis {.unlisted .unnumbered}
-#### [Jon Puritz, PhD](https://www.marineevoeco.com/), University of Rhode Island {.unlisted .unnumbered}
-#### [Steven Roberts, PhD](http://faculty.washington.edu/sr320/?page_id=10850), University of Washington {.unlisted .unnumbered}
+# Contact {#contact}
+For questions about the website and its materials, post to the [Discussion Forum](https://github.com/MarineOmics/marineomics.github.io/discussions)!
+For inquiries about the MarineOmics Working Group activities, please contact us at rcnecs.marineomics@gmail.com.

### 2023-05-10 POP_05_RDAtraitPredictionTutorial.Rmd
removed caching from Katie's RDA page so we won't have to use Large File Storage (LFS) so people don;t have to get LFS configured on their devices
added link  to Katie's PNAS paper on her RDA tutorial page

@@
 
-This tutorial accompanies the paper "The paradox of adaptive trait clines with non-clinal patterns in the underlying genes" published in PNAS [add link].
+This tutorial accompanies the paper ["The paradox of adaptive trait clines with non-clinal patterns in the underlying genes"](https://doi.org/10.1073/pnas.2220313120) published in PNAS.
 
@@
 
-knitr::opts_chunk$set(message = FALSE, warning = FALSE, cache = TRUE) 
+knitr::opts_chunk$set(message = FALSE, warning = FALSE, cache = FALSE) 
 ```
@@
 
-![](multivar.png)
+![](POP_05_RDAtraitPredictionTutorial_files/multivar.png)
 

### 2023-05-21 DRAFT_QTL_tutorial.Rmd
draft of QTL info page and tutorial; still needs images and code added

---
title: "QTL information page and tutorial"
author: "Jason Johns"
bibliography: "ADMIN_01_submissions_instructions_files/common-bib_01.bib"
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, echo = FALSE, message = FALSE}
library(knitr)
knitr::opts_chunk$set(
	fig.height = 5,
	fig.width = 10,
	cache = TRUE
)
library(knitcitations)
library(kableExtra)
opts_chunk$set(fig.width = 10,
               fig.height = 5,
               cache = TRUE)
cite_options(citation_format = "pandoc", max.names = 3, style = "html", 
             hyperlink = "to.doc")
```

The purpose of this page is to give the reader a broad sense of how Quantitative Trait Locus (QTL) analysis can be used to connect phenotype to genotype. It is designed for those who have little to no experience with QTL analysis, although it should be a useful resource for anyone seeking the fundamentals of the method. I first give an overview of some different approaches to the method, then provide a basic tutorial style example using real data. This resource is meant to give more of a broad strokes overview of the methodology rather than rigorous details. As with any MarineOmics page, we welcome feedback and contributions that will strengthen and diversify its content. We invite any and all commentary on the [Discussion Forum](https://github.com/MarineOmics/marineomics.github.io/discussions)!

*I owe most of this code and most of what I know about QTL analysis to Dr. Evangeline Ballerini.

For more technical details, there are many valuable resources, referenced throughout this document. In particular, Karl Broman and Śaunak Sen developed a widely used R package called R/qtl, and have written thorough tutorials freely available on their website. While these tutorials are truly helpful for understanding the basic mechanics of QTL analysis, reading their book entitled QTL mapping with R/qtl is crucial for a thorough understanding.

Other resources for explaining basics of QTL: <br>
-  [Nature Education](https://www.nature.com/scitable/topicpage/quantitative-trait-locus-qtl-analysis-53904/): Miles & Wayne, 2008 <br>
-  JEB Kearsey 1998: The principles of QTL analysis (a minimal mathematics approach) - make sure to read this and vet it <br>
-  Darvasi 1998: Experimental strategies for the genetic dissection of complex traits in animals
discusses intercross vs. backcross strategies

## Principles of QTL analysis
QTL analysis uses linkage to identify genomic regions that harbor gene(s) involved with a trait of interest. It identifies region(s) of the genome where individuals with a particular phenotype consistently have the same genotype, and vice versa. The idea is to eventually narrow the region to a specific gene or genes responsible for a trait. Even if candidate genes are not identified, it can be useful for describing the genetic architecture of a trait or for ruling out a priori candidate genes.

## Experimental design
### considerations for experimental setup
1. Is/are the trait(s) likely to be highly complex & polygenic? <br>
    * large effect QTL can be identified with a relatively small sample size
    * an increase in sample size will narrow the QTL region, decreasing the number of candidate genes to consider
    * the greater the sample size, the more power you will have to identify small effect QTL

2. Is there likely to be a significant influence of environment, life stage, etc.?
    * the more environmental influence there is on a trait, the more it will have to be controlled and/or accounted for
    * environmental variables, life stage, etc. can all be accounted for as covariates in a QTL model

3. Evolutionary implications of the cross
    * QTL analysis identifies the genomic region(s) responsible for the phenotypic difference between the parents
    * if you are interested in the evolution of the trait in the context of selection and speciation, it is better to use populations or species that are more closely related to each other
      * for example: 
        * if we cross two species that are distantly related and have different phenotypes, there are more likely to be regions of the genome associated with the trait that were not responsible for or a result of speciation
        * more closely related taxa will identify regions of the genome associated with the trait that were involved in differentiation

4. single vs. multiple joint family design
    * In some cases (like mine), you may be able to grow your organisms in the lab or greenhouse and make controlled crosses. In these cases, using a single family design is feasible and preferred. 
    * More commonly, (especially in the marine world), controlled crosses are not possible, necessitating the use of multiple joint-family QTL analysis.
<br>
<br>

    The two approaches are briefly outline and compared below: <br>
    *single-family cross*
      * cross one individual with the trait, one without -> F(filial)1 hybrid
      * ideally parents are as inbred as possible, but not crucial that they are
      * cross the F1 to:
        * itself (intercross) - *here’s to hermaphroditism and monoecy
        * to parent with trait of interest (backcross)
        * to sibling with same phenotype <br>

        * for Jason: look at advantages vs disadvantages in Broman
      * most basic approach: use 2nd generation hybrids (F2) for analysis
        * same as what Mendel did to determine modes of inheritance, but now we sequence and annotate genomes to find the actual genes responsible

      * advantages of single-family
        * higher power to detect rare QTL (QTL segregating in small % of families)
        * easier to genotype offspring with only two parents: more likely that parents will have different alleles

    *multiple joint-family*
      * can use: 
        * multiple offspring from the same outcrossed parents
        * multiple offspring from multiple parents
        * uses already existing families

      * advantages of joint-family
        * can use already existing families rather than waiting on crosses
        * do not need to be able to do controlled crosses in captivity
        * more power to detect QTL consistent across multiple families
        * can investigate QTL across multiple genetic backgrounds
        * better estimation of QTL effects
        * more alleles mean more information about function of different alleles in real populations
        * more recombination means better resolution

      * Resources for linkage maps & QTL with joint multiple family based setup:
        * Nousias et al. 2022 Nature SR: Linkage mapping, comparative genome analysis, etc.
        * Manousaki et al. 2016 - Linkage mapping with RADseq & Pandora (Sparidae fish)
        * Wu & Ma 2002 - Math behind creating linkage map with outcrossed parents & siblings
        * Ogut et al. 2015 - Joint-multiple family linkage analysis predicts within-family variation better than single-family analysis of the maize nested association mapping population

make a table comparing single and joint-family QTL

*for nice comparison paper between single and multiple see Ogut et al. 2015*

### limitations of QTL analysis
  * often huge genomic regions fall under the peak with thousands genes under them
  * limited diversity in genetic background to pull out gene interactions
  * time consuming

### data
#### *genotype: most common types of data used today*
  * reduced representation: genotype by sequencing (GBS), RADseq]
  * WGS (link ‘Choosing a Popgen Approach’)<br>
  * As far as linkage map construction goes, relatively little sequence is actually needed. Because recombination only happens so often, F2s will have a consistent genotype (PaPa, PaPb, PbPb) across a large window. Thus, we only need enough sequence data to call genotypes within recombination windows. This can be done with:
    * low coverage (a.k.a. skim) sequencing
    * aggregate data within recombination windows to call genotype over a large block

#### *phenotype*
  * can use: 
    * truly quantitative and continuous
    * categorical data, even binary
  * keep covariates in mind
    * collect as much phenotypic data as is reasonable, keep track of environmental variables

<hr />

## Tutorial
I used whole genome skim sequencing due to the ease of a multiplex library prep, making costs and wet lab time comparable to reduced representation methods

### general info on calling variants
If you have a reference genome available, you can of course just call variants against it. If not, you’ll do a de novo assembly from your sequence data (link tutorials, packages for this within and outside MarineOmics). You may also use another program for calling SNPs without a reference or enough sequence for de novo assembly.

### calling variants for this tutorial
We have a reference genome for *Aquilegia*, so we’re going to align our sequence to it and call variants from it

This tutorial will start with .bam files, a result of aligning the raw sequence data (.fastq) to the *Aquilegia* reference genome. Many tutorials exist for aligning low coverage raw sequence data, including [one on the MarineOmics page](https://github.com/nt246/lcwgs-guide-tutorial) from Nicolas Lu.

This whole idea of steps 1-3 below is to determine the genotypes of the F2 population at enough useful sites to properly represent their genotypes across recombination bins. By useful, I mean only sites that were polymorphic between the parents, and within these, only sites where the F1 inherited different alleles from each parent (heterozygous).

### Let's start by determining the useful sites for genotyping the F2s
####  Generate a genotype depth sequence file for the F1
  * Of course, this can be done by sequencing the F1 to enough coverage to confidently genotype it at many SNPs, however there is a much easier and less costly way:
    * In this case, the F2 population only inherited alleles from the F1, so if we merge all of the F2 sequence data, we're left with a genotype level sequence file of the F1!

#### Call and filter SNPs
  * call SNPs with whatever program you like: we use samtools here, there are many others
  * filter vcfs
    * min & max depth: ___
    * min quality: ___

#### Identify useful SNPs 
  * As mentioned previously, useful SNPs will be heterozygous in the F1
  * If we have enough sequence data from the parents, we can determine which allele came from which parent
  * In this case, I only had sequence data for one of the parents (Pa), which narrows the list of useful sites:
    * So not only can I only use sites where the F1 is heterozygous, I also have to use sites that with a particular genotype in Pa.
    * Sites where the F1 and Pa are both heterozygous for the same allele are not useful, because it is still unclear which allele came from which parent:
    * insert image to illustrate this
    * So the only sites that are useful are: 
      * F1: is heterozygous 
      * Pa: 1. homozygous, or.. <br>
            2. heterozygous with one allele that matches the F1 and one that doesn't. 
      * As the latter case is relatively rare, and there were plenty of sites that satisfy case 1, I decided to use sites that were heterozygous in the F1 and homozygous in Pa
      *show code for this

### Now we can actually genotype the F2s
 * determine useful window sizes depending on recombination rate
    * we used from 10kb to 1Mb
    * make a .csv or .txt defining these windows
  * calculate frequencies of parental alleles in these windows
  * we used windows of allele frequencies of the parent we had sequence for (Pa) to determine F2 genotypes
    * frequency of Pa allele
      * 0 - 0.15: homozygous Pb
      * \>0.15 - 0.35: unsure
      * \>0.35 - 0.65: heterozygous
      * \>0.65 - 0.85: unsure
      * \>0.85 - 1: homozygous Pa

### Make a genetic map
  * used r/qtl for this, along with some manual brain power
    * in my case there was discordance between genetic and physical maps, which was likely due to: 
      * true structural differences between reference genome and species used in the cross
      * mis-assembly of the original genome
        * was done with Sanger sequencing over 15 years ago
    * so I re-arranged markers manually by intuition
      * I took my genetic map and colored it by genotype so I could visualize genotypes that didn't make sense, such as impossible patterns of recombination 
        * in some cases, there was a recombination event more than expected by chance
        * in addition, some of these alleged recombination events went from PAPA to PBPB between two markers, which would only be possible if there were two recombination events between those two markers --> highly unlikely
    * Thus, I manually re-ordered markers and filled in ‘unsure’ genotypes based on intuition

### QTL analysis
  * merge phenotype data with genotype data
    * transposed phenotype data onto genetic map
  * read into to r/qtl
  * scanone
  * scantwo
  * multipleqtl
  * stepwiseqtl
  * 1.5 LODs
  * effect plots

### interpretation of analysis
  * remember, you’re often not honing in on a specific gene with QTL analysis alone, so keep your goal in mind and be reasonable with how much you can actually infer based on your data
  * are candidate genes or homologs of candidate genes under the peak?
  * are previously identified candidate genes not under the peak?
  * narrowing the list of genes under the peak
    * is there any existing expression data for your tissue of interest?
    * RNA-seq, qPCR if you have a good candidate

## Terminology glossary
  * marker
  * genetic map
  * linkage mapping
  * association mapping
  * interval mapping
  * linkage group
  * recombination rate (maybe ditch)
  * recombination fraction
  * multiple QTL regression analysis
  * significance threshold
  * missingness
  * Quantitative Trait Locus
  * LOD score
  * LOD interval (1.5, 1.8)
  * intercross
  * backcross
  * single family based QTL mapping
  * joint family based QTL mapping
  * genetic interaction
  * epistasis

graveyard
fine mapping:
recombinant inbred lines (RIL) see Takuno et al. 2012
near isogenic lines (NIL)

### 2023-06-01 DRAFT_QTL_tutorial.Rmd
added some code and some text to the QTL tutorial

@@
 ## Tutorial
-I used whole genome skim sequencing due to the ease of a multiplex library prep, making costs and wet lab time comparable to reduced representation methods
+This tutorial will walk you through the essentials of the particular analysis that I did, but my hope is that it will be useful even for those who take different approaches.
+It is not written with reproducible data, but I have included all of the necessary code and outputs.
+
+I used whole genome skim sequencing due to the ease of a multiplex library prep, making costs and wet lab time comparable to reduced representation methods.
 
@@
 You may also use another program for calling SNPs without a reference or enough sequence for de novo assembly.
+I have included all script files for this analysis here.
+For the sake of the tutorial, code is split into chunks with its output.
+For a brush up on command line code, Nicolas Lu has a great tutorial in his WGS page (if this is true, link it).
 
@@
 
-This whole idea of steps 1-3 below is to determine the genotypes of the F2 population at enough useful sites to properly represent their genotypes across recombination bins.
+The whole idea of steps 1-3 below is to determine the genotypes of the F2 population at enough useful sites to properly represent their genotypes across recombination bins.
 By useful, I mean only sites that were polymorphic between the parents, and within these, only sites where the F1 inherited different alleles from each parent (heterozygous).
 
-### Let's start by determining the useful sites for genotyping the F2s
+### We're going to start by defining all of our filepaths, variable names, and tools
+```{bash, eval = F}
+# make variables for commonly used directories
+workingdir="/Users/jason/qtl.tutorial"
+mkdir $workingdir
+
+bamdir="$workingdir/f2.alignments"
+mkdir $bamdir
+
+infodir="$workingdir/info"
+mkdir $infodir
+
+genodir="$workingdir/genotype"
+idallelesdir="$genodir/id.alleles"
+mkdir $genodir
+mkdir $idallelesdir
+
+# file with the Aquilegia reference genome
+reference="/Volumes/HD/genomes.reference/Aquilegia/V3/sequences/Aquilegia_coerulea.main_genome.scaffolds.fasta"
+f1bam="$infodir/f2.merged.bam"
+
+# on our lab computer we call samtools from it's location
+samtools="/Volumes/HD/tools/samtools-1.3/samtools"
+
+```
+
+### To begin our genotyping, let's determine the useful sites for genotyping the F2s
 ####  Generate a genotype depth sequence file for the F1
@@
     * In this case, the F2 population only inherited alleles from the F1, so if we merge all of the F2 sequence data, we're left with a genotype level sequence file of the F1!
+
+```{bash, eval = F}
+# we'll start in our working directory
+cd $workingdir
+
+# make a list file with all of the file paths of our .bams for samtools to merge
+ls -d -1 $PWD/*.* > $infodir/f2.paths.list
+
+# merge bam files
+samtools merge $infodir/f2.merged.bam -b $infodir/f2.paths.list
+
+# index bam files
+samtools index $infodir/f2.merged.bam
+```
+
+We can view the merged, indexed .bam file of the F2s in IGV (link).
+Because I got ~1x coverage on ~350 individuals, this merged .bam file is essentially a 350x coverage sequence file of the F1...plenty to call genotypes :).
 
@@
     * min quality: ___
+
+```{bash, eval = F}
+# move into the id.alleles directory
+cd $idallelesdir
+
+# run samtool mpileup to identify variant sites and create a .vcf file with those sites in our 'F1'
+samtools mpileup -q 30 -B -u -v -f $reference $f1bam | bcftools call -vm -f GQ > var.f1.vcf
+
+```
 

### 2023-06-02 ADMIN_01_submissions_instructions.Rmd
fixed typo on submissions instructions and made a bunch of progress on the QTL tutorial

@@
 To insert an image, you need the image file to also be uploaded to the
-Github repositority. We store the images hosted on the website in the
-`images/` folder. For any image, please make sure you have permission to
+Github repository. We store the images hosted on the website in each page's folder and call them accordingly. For any image, please make sure you have permission to
 share the image and that it is appropriately cited.

### 2023-06-14 DRAFT_Funct_Enrich.Rmd
adding text for best principles page on index, added best principles to header, and all appropriate links, re-rendered site
changed document back to eval = TRUE, had to make false so I could render the site from rstudio
dialing in navigation to contributions & best practices with new index landing page
added text describing overrepresentation vs enrichment analysis

@@
 * Explanation of statistics (Sam) - include key terms
+
+## Overrepresentation vs Enrichment
+
+Often, everything gets referred to as enrichment analysis, but this is not the case.
+The two primary analyses to be run that examine functional patterns are statistical over-representation analysis and statistial enrichment analysis.
+There are differences in how data are input, what kind of data are used, and what statistical tests are employed by each analysis.
+
+Over-represenation analysis is used when you do not have expression values associated with your data.
+In this case you will provide a list of a subset of genes of interest as well as a reference list.
+This reference list will contain all the genes in your study.
+For example, in a transcriptomic analysis the reference list would include all expressed genes in the experiment while the gene of interest list might include genes that are over expressed under a specific condition (e.g., genes over expressed in females compared to all genes expressed in males and females).
+These two lists are compared to one another and a Fisher's Exact Test is used to calculate whether any functional categories (based on how genes are annotated to particular GO terms) are over-represented in your list of genes of interest, relative to the representation of all functional categories present in your full data set.
+
+Enrichment analysis is used when you do have expression values associated with your data.
+In this case, you will provide the full list of genes in your experiment and their associated expression values (e.g., this might be a library from an RNA-seq experiment).
+Any expression metric can be used.
+This flexiblity is because enrichment analysis uses a Mann-Whitney U Test to calculate enrichment, which is based on rankings - thus values need to be interpretable for rank-based relationships, and an appropriate normalization metric should be applied to your data.
+
+See later sections for further discussion of Fisher's Exact Test and Mann-Whitney U Test.
 
@@
 
-```{r, include = TRUE}
+```{r, include = TRUE, eval=F}
 

### 2023-06-14 DRAFT_Funct_Enrich.Rmd
Added subsection headers and text on ontologizer.

@@
 
+## The Gene Ontology
+
+## Annotation and filtration decisions
+
 ## Overrepresentation vs Enrichment
@@
 
+## Explanation of Statistics
+
 <br>
@@
 * R...Mann-Whitney U Test and Fisher’s Exact Test (Sam)
+
+## Ontologizer
+
+[Ontologizer](http://ontologizer.de/) offers a command-line option for functional analysis using the Gene Ontology and is one of several tools endorsed by the GO.
+Functionally, it represents a tool for overrepresentation analysis as it makes use of the Fisher's Exact Test.
+The most challenging part of running ontologizer is preparing your input files and ensuring they are annotated properly.
+
+To run Ontologizer on the command line the Ontologizer.jar file will need to be downloaded from [here](http://ontologizer.de/commandline/).
+Additionally, Ontologizer requires java to run, so ensure you have this installed or call the relevant module in your hpc system.
+It needs to be supplied with the following:
+
+* The association file (-a) relating genes to GO terms (ends in .gaf for gene association file).
+A number of options are available through the annotations link on [this page](http://geneontology.org/docs/download-go-annotations/).
+* The file (-g) containing the GO terminology and structure [go-basic.obo](http://geneontology.org/docs/download-ontology/).
+* A population reference file (-p), which you will generate.
+* Files representing your gene subsets of interest (-s).
+Multiple subsets can be provided at one time to Ontologizer and run in parallel by specifying a path to the directory containing only these files.
+
+There are a number of other options that can be specified.
+Some worth consider are your calculation method (-c), whether to ignore genes with no associations in your calculations (-i), and the name of the directory that you want your output written to (-o).
+
+A sample script run on an hpc cluster may look like the following:
+
+```
+java -jar Ontologizer.jar -a goa_uniprot_all.gaf -g go-basic.obo -p pop.CPISwissProtIDs.txt -s Subset/CPIDEA/StudySet -c "MGSA" -d 0.05 -i -o CPIOntologizerSubsetOutput
+```
+
+## PANTHER
+
+## Using R
 

### 2023-06-15 DRAFT_Funct_Enrich.Rmd
Added bibliographic reference on for meta analysis of enrichment studies. Added text on guidelines for rigorous reporting of enrichment analyses.

@@
 - Example 2
+
+While functional gene enrichment analysis poses many challenges to non-model systems, it is critical the analyses conducted be as rigorous and replicable as possible.
+A recent survey of published enrichment results indicated that issues in the performance and reporting of enrichment analyses is not uncommon [@RN335].
+Basic standards advised by the authors include:
+
+* Selection of appropriate gene background set.
+This set should not necessarily be the full set of genes in the genome.
+Rather it should be the set of genes expressed in the experiment.
+This is especially an issue with overrepresentation analysis and showed a consider impact on results when done incorrectly.
+* Multiple hypothesis correction. A common metric used with enrichment results is a false discovery rate (FDR) correction. 
+* Adequate methodological reporting including software and parameters used, versions of software and annotation files, all gene lists analyzed including subset and background sets, types of statistical tests and corrections performed, and any code used to conduct the analyses.
 

### 2023-06-15 DRAFT_Funct_Enrich.Rmd
Added text on PANTHER

@@
 
-Proper annotation of genes is crucial because it ensures that the genes are correctly identified and annotated with their appropriate biological functions.
+Proper annotation of genes is crucial because it ensures the genes are correctly identified and annotated with their appropriate biological functions.
 Incorrect annotation of genes can lead to inaccurate interpretations of experimental results and hinder progress in understanding the biological systems of non-model species.
@@
 
-Over-represenation analysis is used when you do not have expression values associated with your data.
+Over-representation analysis is used when you do not have expression values associated with your data.
 In this case you will provide a list of a subset of genes of interest as well as a reference list.
@@
 ## PANTHER
+
+PANTHER is a GUI option that is endorsed by The Gene Ontology and integrated into their website. PANTHER stands for Protein ANalysis THrough Evolutionary Relationships.
+
+### Preparing data for input into Panther.
+
+Here we will go over how to map your transcripts to PANTHER IDs for input into the PANTHER analysis pipeline.
+PANTHER can accept a number of ID types, but this is method may be useful if you are working in a nonmodel system with limited annotation data available.
+
+Download the PANTHER HMM scoring tool from [here](http://pantherdb.org/downloads/index.jsp)
+
+```
+wget http://data.pantherdb.org/ftp/hmm_scoring/current_release
+```
+
+Unpack the tar file.
+
+```
+tar -xvzf PANTHER17.0_hmmscoring.tgz
+```
+
+Using the provided script located at pantherScore2.2/pantherScore2.2.pl, submit amino acid sequences to be searched against the PANTHER HMM library and assign appropriate IDs.
+Thus you may need to translate your transcript sequences using a tool such as TransDecoder.
+Searching your sequences against the PANTHER HMM library can be done on the command line.
+You will need to load perl and hmmer software packages.
+Test data are provided to check that everything installed properly.
+Once you are ready to go, you can run some version of the following.
+See command options for alternative variables to consider.
+
+```
+pantherScore2.2.pl -l ../target/famlib/rel/PANTHER17.0_altVersion/hmmscoring/PANTHER17.0/ -D B -V -i translated_coding_sequences.faa -o hits.txt -n
+```
+
+    -l PANTER (l)ibrary with HMMs when using hmmsearch, binary PANTHER HMM model when using hmmscan
+    -D display type for results
+        options: I (interproscan), B (best hit), A (all hits)
+    -o (o)utput file (redirect STDOUT)
+    -i (i)nput fasta file to score
+    -V (V)ery verbose (debug info to STDERR)
+
+The results will look something like this:
+
+```
+CpiID   PantherID       GeneName        HmmEvalue       HmmBitscore     AlignRange
+lcl|NC_024225.1_prot_XP_023959495.1_6146        PTHR46343:SF3   SUSHI REPEAT-CONTAINING PROTEIN SRPX2   1.2e-266        887.6   9-462
+lcl|NW_007359912.1_prot_XP_008164062.1_46642    PTHR10048:SF33  PHOSPHATIDYLINOSITOL 4,5-BISPHOSPHATE 3-KINASE CATALYTIC SUBUNIT BETA ISOFORM   0       2111.3  1-1040
+lcl|NW_007281451.1_prot_XP_008170789.1_21777    PTHR23119:SF48  LEUCINE-RICH REPEAT-CONTAINING PROTEIN 7        0       2508.4  7-1351
+lcl|NW_007281531.1_prot_XP_005308912.1_25857    PTHR12157:SF18  REGULATING SYNAPTIC MEMBRANE EXOCYTOSIS PROTEIN 1       0       2637.1  1-1502
+lcl|NW_007359900.1_prot_XP_005287749.1_44438    PTHR13040:SF2   AUTOPHAGY PROTEIN 5     3.2e-105        355.0   3-272
+lcl|NW_007359900.1_prot_XP_005287749.1_44438    PTHR13040       AUTOPHAGY PROTEIN 5     3.2e-105        355.0   3-272
+lcl|NW_007282136.1_prot_XP_005313942.1_33204    PTHR26450:SF179 OLFACTORY RECEPTOR      2.6e-146        489.7   1-311
+...
+```
+
+Now that you have the PANTHER annotations, the next step is to do any final parsing of your data for input into PANTHER.
+For overrepresentation analysis you will need a reference file representing your population of genes and a subset file representing your set of target genes of interest.
+They can both take the format for Generic Mapping which follows the form shown above.
+If performing enrichment analysis, some extra data formatting will be needed.
+In this case, input should contain three columns: a gene ID (in a form useful to you), the PANTHER ID (resulting from the search against the HMM library described above), and an expression metric for each gene.
+
+### Running PANTHER Overrepresentation
+
+Go to http://pantherdb.org. This will bring up the start page which will look something like this:
+
+image: ![](DRAFT_Funct_Enrich/PantherStartPage.png)
+
+Step 1 is to upload your file of interest.
+If you are doing an overrepresentation analysis, this would be a your subset list of genes.
+Your list type will need to be specified.
+If you are working with a non-model system with limited genomic resources, chances are you will want to use the PANTHER Generic Mapping option (see elsewhere for instructions on preparing this data).
+See both the Supported IDs and File format links on this page for descriptions of acceptable IDs.
+
+Step 2 depends on your selection in step one.
+Here is where you would specify your organism (and consequent reference file) if it is avialable within the PANTHER database.
+If your organism is not avialable, you will have selected PANTHER Generic Mapping and will provide your own reference set (for overrepresenation analysis only) at a later stage and this step will be skipped.
+
+Step 3 selects the type of analysis you would like to run. Here we will discuss the statistical overrperesentation test.
+
+Here is what things will look like once your data are uploaded:
+
+image: ![](DRAFT_Funct_Enrich/PantherStartPageWithData.png)
+
+Click Submit.
+
+You will now be brought to the second data entry page.
+Here is where you will provide your reference list.
+If you are working with a species present in the PANTHER DATABASE you can specify that here.
+Alternatively you can provide your own.
+In this example we will provide a PANTHER Generic Mapping File, since we also provided this type of file for our subset list.
+
+First, select the list type - in this case we select PANTHER Generic Mapping File.
+Then choose file; you will be able to upload a file from your computer.
+Select upload list and you will be brought the the analysis page.
+
+Prior to selecting upload list, the page should look something like this:
+
+image: ![](DRAFT_Funct_Enrich/PantherReferencePage.png)
+
+Once you have uploaded your reference file you will be brought to a new page.
+You should see your subset of interest file after Analyzed List and your population reference file after Reference List.
+You can select which annotation data set you want to query:
+
+* PANTHER Pathways
+* PANTHER GO-Slim Molecular Function
+* PANTHER GO-Slim Biological Process
+* PANTHER GO-Slim Cellular Component
+* PANTHER Protein Class
+
+Select your test type (either Fisher's Exact Test or Binomial) and test correction type (False Discovery Rate, Bonferroni, or None).
+
+Here is an example of what the page will look like once it is ready to go:
+
+image: ![](DRAFT_Funct_Enrich/PantherAnalysisPage.png)
+
+Launch Analysis.
+
+Results will load below. The default is to only show significant results. There are several options for downloading results. Simply select the type you wish to download and save it to your computer.
+
+Here is an example of the results page when there are significant hits:
+
+image: ![](DRAFT_Funct_Enrich/PantherResults.png)
+
+### Running PANTHER Enrichment
+
+Go to http://pantherdb.org. This will bring up the start page which will look something like this:
+
+image: ![](DRAFT_Funct_Enrich/PantherStartPage.png)
+
+Step 1 is to upload your file of interest.
+If you are running enrichment analysis, then you would upload your full set of genes with expression values.
+Your list type will need to be specified.
+If you are working with a non-model system with limited genomic resources, chances are you will want to use the PANTHER Generic Mapping option (see elsewhere for instructions on preparing this data).
+See both the Supported IDs and File format links on this page for descriptions of acceptable IDs.
+
+Step 2 depends on your selection in step one.
+Here is where you would specify your organism if it is avialable within the PANTHER database.
+If your organism is not avialable, you will have selected PANTHER Generic Mapping and this step will be skipped.
+
+Step 3 selects the type of analysis you would like to run. Here we will discuss the statistical enrichment test.
+
+Here is what things will look like once your data are uploaded:
+
+image: ![](DRAFT_Funct_Enrich/PantherStartPageWithDataEnrichment.png)
+
+Click Submit.
+
+This will bring up the results page for the analysis.
+You can select a different type of annotation data set to analyze if you wish as well as change the multiple hypothesis correction type.
+The data sets available are the same as for running the overrepresentation test.
+A false discovery rate correction is applied automatically and any significant results are displayed below, but Bonferroni and none are options as well.
+There are several options for downloading results.
+Simply select the type you wish to download and save it to your computer.
+
+Here is an example of the results page when there are significant hits:
+
+image: ![](DRAFT_Funct_Enrich/PantherResultsEnrichment.png)
 

### 2023-06-15 DRAFT_Funct_Enrich.Rmd
Add sam's intro to enrichment stats

@@
 * R...Mann-Whitney U Test and Fisher’s Exact Test (Sam)
+
+# A primer on the statistics of functional enrichment
+
+Two popular methods of testing for GO term enrichment are the Fishers exact test and Mann Whitney U tests.
+Each test can produce results that are sensitive to the kinds of data they are input with such as differential expression or WGCNA coefficients [cite].
+Neither test necessarily needs to be used alongside known networks of genes and use GO terms, enabling their use in non-model systems.
+These two methods are summarized and derived below:
+
+<br>
+
+##Fisher’s Exact Test
+
+The Fisher’s exact test evaluates the relationship between two categorical variables: whether the ontology of a gene is represented in a test set or a background set.
+In the case of differential expression (DE), the test set is al DE genes and the background set is all genes included in DE analysis after filtering, not the entire transcriptome.
+The test asks whether the number of genes for a given GO term represented in a test set relative to the background is greater than what is expected due to chance alone (p).
+
+$$
+p = \frac{{\binom{{a+b}}{{a}} \cdot \binom{{c+d}}{{c}}}}{{\binom{{n}}{{a+c}}}}
+$$
+
+Where
+
+$a =$ $n$ genes of GO term $i$ in both the test and background set
+$b =$ $n$ genes of GO $i$ in the background but not test
+$c =$ $n$ genes of GO $i$ absent from the test but present in the background
+$d =$ $n$ genes of GO $i$ absent in both the category and the background
+$n =$ $n$ genes in the background set
+
+A Fisher's exact test can be executed in base R using the fishers.text() function
+
+<br>
+
+##Mann Whitney U Test
+
+The critical difference between Mann Whitney U and Fisher’s exact tests and are that Mann Whitney U’s permit the analysis of continuous rather than categorical variables.
+These continuous values can be log fold changes of differential expression or -log p-values associated with DE or inclusion in a WGCNA.
+Because it doesn’t rely on binary values denoting the existence of an effect, but rather the strength of an effect, the Mann Whitney U test is a more quantitative approach toward functional enrichment.
+The test asks whether the distribution of ranks for this continuous value in a GO term is significantly skewed from the distribution of ranks across all other GO terms.
+Like the background set employed in a Fisher’s exact test, “all other GO terms” refers to those terms included in genes used for analysis following filtering.
+
+The skewedness of a GO term’s rank distribution relative to the background is represented by the statistic $U$. The p-value output by the Mann Whitney U test is calculated with the following equation:
+
+$$
+p = \frac{{U}}{{a \cdot b}}
+$$
+
+Where
+
+$a =$ $n$ genes in GO term $i$
+$b =$ $n$ genes in the background set
+
+A Mann Whitney U test can be executed for GO enrichment analysis using code from [@Wright2015-dc]
+
+<br>
+
+A number of different parameters can be used to filter results or modify the execution of both tests, which we detail below.
 

### 2023-06-15 DRAFT_Funct_Enrich.Rmd
Add html knit of draft w/ stats intro

@@
 
-##Fisher’s Exact Test
+## Fisher’s Exact Test
 
@@
 
-$a =$ $n$ genes of GO term $i$ in both the test and background set
-$b =$ $n$ genes of GO $i$ in the background but not test
-$c =$ $n$ genes of GO $i$ absent from the test but present in the background
-$d =$ $n$ genes of GO $i$ absent in both the category and the background
-$n =$ $n$ genes in the background set
+* $a =$ $n$ genes of GO term $i$ in both the test and background set
+* $b =$ $n$ genes of GO $i$ in the background but not test
+* $c =$ $n$ genes of GO $i$ absent from the test but present in the background
+* $d =$ $n$ genes of GO $i$ absent in both the category and the background
+* $n =$ $n$ genes in the background set
 
@@
 
-##Mann Whitney U Test
+## Mann Whitney U Test
 
@@
 
-$a =$ $n$ genes in GO term $i$
-$b =$ $n$ genes in the background set
+* $a =$ $n$ genes in GO term $i$
+* $b =$ $n$ genes in the background set
 

### 2023-06-21 DRAFT_Funct_Enrich.Rmd
knit changes
started addition of annot to FE

@@
 ---
-
-# Background
 
@@
 This curation process can be particularly challenging for non-model species with limited genomic and transcriptomic resources.
-More information on this can be found at
 
-- Example 1
-- Example 2
+# Annotation
+
+The following is a stepwise example or annotation of a gene set using Swiss-Prot such that Gene Ontology terms can be associated with each gene.
+
+```{r engine='bash', eval=FALSE}
+cd ../blastdb
+curl -O https://ftp.uniprot.org/pub/databases/uniprot/current_release/knowledgebase/complete/uniprot_sprot.fasta.gz
+mv uniprot_sprot.fasta.gz uniprot_sprot_r2022_02.fasta.gz
+gunzip -k uniprot_sprot_r2022_02.fasta.gz
+```
+
+```{r engine='bash', eval=FALSE}
+/home/shared/ncbi-blast-2.11.0+/bin/makeblastdb \
+-in ../blastdb/uniprot_sprot_r2022_02.fasta \
+-dbtype prot \
+-out ../blastdb/uniprot_sprot_r2022_02
+```
+
+# Overview
 
@@
 
-* Selection of appropriate gene background set.
+-   Selection of appropriate gene background set.
 This set should not necessarily be the full set of genes in the genome.
@@
 This is especially an issue with overrepresentation analysis and showed a consider impact on results when done incorrectly.
-* Multiple hypothesis correction. A common metric used with enrichment results is a false discovery rate (FDR) correction. 
-* Adequate methodological reporting including software and parameters used, versions of software and annotation files, all gene lists analyzed including subset and background sets, types of statistical tests and corrections performed, and any code used to conduct the analyses.
+-   Multiple hypothesis correction. A common metric used with enrichment results is a false discovery rate (FDR) correction.
+-   Adequate methodological reporting including software and parameters used, versions of software and annotation files, all gene lists analyzed including subset and background sets, types of statistical tests and corrections performed, and any code used to conduct the analyses.
 
@@
 
-* Differential expression
-* Clustering
-* WGCNA
+-   Differential expression
+-   Clustering
+-   WGCNA
 
@@
 
-* The Gene Ontology (Thea)
-* Annotation and filtration decisions (Sam) - include key terms
-* Overrepresentation vs Enrichment (Thea)
-* Explanation of statistics (Sam) - include key terms
+-   The Gene Ontology (Thea)
+-   Annotation and filtration decisions (Sam) - include key terms
+-   Overrepresentation vs Enrichment (Thea)
+-   Explanation of statistics (Sam) - include key terms
 
@@
 
-* Ontologizer (command line) (Thea)
-* Panther (online GUI) - endorsed by The Gene Ontology (Thea)
-* R...Mann-Whitney U Test and Fisher’s Exact Test (Sam)
+-   Ontologizer (command line) (Thea)
+-   Panther (online GUI) - endorsed by The Gene Ontology (Thea)
+-   R...Mann-Whitney U Test and Fisher's Exact Test (Sam)
 
@@
 
-## Fisher’s Exact Test
+## Fisher's Exact Test
 
-The Fisher’s exact test evaluates the relationship between two categorical variables: whether the ontology of a gene is represented in a test set or a background set.
+The Fisher's exact test evaluates the relationship between two categorical variables: whether the ontology of a gene is represented in a test set or a background set.
 In the case of differential expression (DE), the test set is al DE genes and the background set is all genes included in DE analysis after filtering, not the entire transcriptome.
@@
 
-* $a =$ $n$ genes of GO term $i$ in both the test and background set
-* $b =$ $n$ genes of GO $i$ in the background but not test
-* $c =$ $n$ genes of GO $i$ absent from the test but present in the background
-* $d =$ $n$ genes of GO $i$ absent in both the category and the background
-* $n =$ $n$ genes in the background set
+-   $a =$ $n$ genes of GO term $i$ in both the test and background set
+-   $b =$ $n$ genes of GO $i$ in the background but not test
+-   $c =$ $n$ genes of GO $i$ absent from the test but present in the background
+-   $d =$ $n$ genes of GO $i$ absent in both the category and the background
+-   $n =$ $n$ genes in the background set
 
@@
 
-The critical difference between Mann Whitney U and Fisher’s exact tests and are that Mann Whitney U’s permit the analysis of continuous rather than categorical variables.
+The critical difference between Mann Whitney U and Fisher's exact tests and are that Mann Whitney U's permit the analysis of continuous rather than categorical variables.
 These continuous values can be log fold changes of differential expression or -log p-values associated with DE or inclusion in a WGCNA.
-Because it doesn’t rely on binary values denoting the existence of an effect, but rather the strength of an effect, the Mann Whitney U test is a more quantitative approach toward functional enrichment.
+Because it doesn't rely on binary values denoting the existence of an effect, but rather the strength of an effect, the Mann Whitney U test is a more quantitative approach toward functional enrichment.
 The test asks whether the distribution of ranks for this continuous value in a GO term is significantly skewed from the distribution of ranks across all other GO terms.
-Like the background set employed in a Fisher’s exact test, “all other GO terms” refers to those terms included in genes used for analysis following filtering.
+Like the background set employed in a Fisher's exact test, "all other GO terms" refers to those terms included in genes used for analysis following filtering.
 
-The skewedness of a GO term’s rank distribution relative to the background is represented by the statistic $U$. The p-value output by the Mann Whitney U test is calculated with the following equation:
+The skewedness of a GO term's rank distribution relative to the background is represented by the statistic $U$. The p-value output by the Mann Whitney U test is calculated with the following equation:
 
@@
 
-* $a =$ $n$ genes in GO term $i$
-* $b =$ $n$ genes in the background set
+-   $a =$ $n$ genes in GO term $i$
+-   $b =$ $n$ genes in the background set
 
@@
 
-* The association file (-a) relating genes to GO terms (ends in .gaf for gene association file).
+-   The association file (-a) relating genes to GO terms (ends in .gaf for gene association file).
 A number of options are available through the annotations link on [this page](http://geneontology.org/docs/download-go-annotations/).
-* The file (-g) containing the GO terminology and structure [go-basic.obo](http://geneontology.org/docs/download-ontology/).
-* A population reference file (-p), which you will generate.
-* Files representing your gene subsets of interest (-s).
+-   The file (-g) containing the GO terminology and structure [go-basic.obo](http://geneontology.org/docs/download-ontology/).
+-   A population reference file (-p), which you will generate.
+-   Files representing your gene subsets of interest (-s).
 Multiple subsets can be provided at one time to Ontologizer and run in parallel by specifying a path to the directory containing only these files.
@@
 Using the provided script located at pantherScore2.2/pantherScore2.2.pl, submit amino acid sequences to be searched against the PANTHER HMM library and assign appropriate IDs.
-Thus you may need to translate your transcript sequences using a tool such as TransDecoder.
+Thus you may need to translate your transcript sequences using a tool such as [TransDecoder](https://github.com/TransDecoder/TransDecoder/wiki).
 Searching your sequences against the PANTHER HMM library can be done on the command line.
@@
 
+```         
 -l PANTER (l)ibrary with HMMs when using hmmsearch, binary PANTHER HMM model when using hmmscan
@@
 -V (V)ery verbose (debug info to STDERR)
+```
 
@@
 
-Go to http://pantherdb.org. This will bring up the start page which will look something like this:
+Go to <http://pantherdb.org>. This will bring up the start page which will look something like this:
 
@@
 
-* PANTHER Pathways
-* PANTHER GO-Slim Molecular Function
-* PANTHER GO-Slim Biological Process
-* PANTHER GO-Slim Cellular Component
-* PANTHER Protein Class
+-   PANTHER Pathways
+-   PANTHER GO-Slim Molecular Function
+-   PANTHER GO-Slim Biological Process
+-   PANTHER GO-Slim Cellular Component
+-   PANTHER Protein Class
 
@@
 
-* Visualizing results
-* Interpreting results - gather consensus advice 
-* R...Mann-Whitney U Test and Fisher’s Exact Test (Sam)
+-   Visualizing results
+-   Interpreting results - gather consensus advice
+-   R...Mann-Whitney U Test and Fisher's Exact Test (Sam)
 
@@
 | Term A | Term B | Term C | Term D | Term E | Term F | Term G |
-| :--- | :--- | :--- | :--- | :--- | :--- | :--- |
+|:-------|:-------|:-------|:-------|:-------|:-------|:-------|
 | xxx    | xxx    | xxx    | xxx    | xxx    | xxx    | xxx    |
@@
 
+````         
 ```{r setup, include = FALSE} `r ''`
@@
 ```
+````
 
@@
 
-# Run GOMWU and Fisher's exact tests on data of interest (see https://github.com/z0on/GO_MWU)
+# Run GOMWU and Fisher's exact tests on data of interest (see <https://github.com/z0on/GO_MWU>)
 

### 2023-06-21 DRAFT_Funct_Enrich.Rmd
cv annotation on FE

@@
 
-# Annotation
+# Navigating Annotation
 
-The following is a stepwise example or annotation of a gene set using Swiss-Prot such that Gene Ontology terms can be associated with each gene.
+The following is a stepwise example or annotation of a gene set using UniProt::Swiss-Prot (reviewed) such that Gene Ontology terms can be associated with each gene.
+
+In this following chunk where the fasta file is downloaded the [release](https://www.uniprot.org/help/release-statistics) is noted and the file name is modified accordingly.
 
@@
 
+A protein blast database is then made.
+
 ```{r engine='bash', eval=FALSE}
@@
 -out ../blastdb/uniprot_sprot_r2022_02
+```
+
+In a majority of cases you will want to annotate a gene set to get gene ontology information.
+If you are creating your own genome or transcriptome it should be rather straightforward to know what file to annotate.
+If using a widely studied system where there are publically available resources, it is advisable to use those as this is the best way to facilitate integration of data sets.
+In this case study we will be considering the Eastern oyster, (_Crassostrea virginica_)
+
+```{r engine='bash', eval=FALSE}
+cd ../data
+
+curl -O https://ftp.ncbi.nlm.nih.gov/genomes/all/GCF/002/022/765/GCF_002022765.2_C_virginica-3.0/GCF_002022765.2_C_virginica-3.0_protein.faa.gz
+
+gunzip -k GCF_002022765.2_C_virginica-3.0_protein.faa.gz
 ```

### 2023-06-22 DRAFT_Funct_Enrich.Rmd
more Cg annot on FE

@@
 If using a widely studied system where there are publically available resources, it is advisable to use those as this is the best way to facilitate integration of data sets.
-In this case study we will be considering the Eastern oyster, (_Crassostrea virginica_)
+In this case study we will be considering the Eastern oyster, (_Crassostrea virginica_) for which there is data at [NCBI](https://www.ncbi.nlm.nih.gov/datasets/genome/GCF_002022765.2/) and [Ensembl Metazoa](http://metazoa.ensembl.org/Crassostrea_virginica_gca002022765v4/Info/Index).
+At NCBI there is both a GenBank and RefSeq assembly available.
+
+In order to know which of the numerous fasta files should annotated with gene ontology information one should think downstream (or look to files already generated) to the identifiers in genesets that would be subject to functional enrichment tests.
+
+The resulting fpkm count matrix for our case study is from an experiment where male and female oysters where exposed to low pH (and control) conditions. 
+The count matrix is accessible [here (csv)](https://github.com/epigeneticstoocean/2018_L18-adult-methylation/blob/main/analyses/gene_fpkm.csv).
+Hisat2/Stringtie was used to generate the count matrix with `GCF_002022765.2_C_virginica-3.0_genomic.gff` formatting thus responsible for gene naming.
+Specifically the naming format is as follows `gene-LOC111099033,gene-LOC111099034,gene-LOC111099035`.
+
+No single file seems obvious or simple with respect to the primary fasta header labelled with LOCXXXXXXXXXX. Have decided to use
+
+```
+>lcl|NC_035780.1_ncrna_XR_002636969.1_1 [gene=LOC111116054] [db_xref=GeneID:111116054] [product=uncharacterized LOC111116054] [ncRNA_class=lncRNA] [transcript_id=XR_002636969.1] [location=join(13578..13603,14237..14290,14557..14594)] [gbkey=ncRNA]
+TGATATTGTTGTGTGCAGAACGTGGGTCATGCTGTCTTGTTGAAGAATACCCTCAATCCAAGGCTAGATTAATCATCTAG
+CCATGATATTGATGATAGCAAGATTTGTGTTGATCCTG
+```
 

### 2023-06-22 DRAFT_Funct_Enrich.Rmd
sp blast on FE

@@
 ```{r engine='bash', eval=FALSE}
-cd ../blastdb
+cd DRAFT_Funct_Enrich/annot
+
 curl -O https://ftp.uniprot.org/pub/databases/uniprot/current_release/knowledgebase/complete/uniprot_sprot.fasta.gz
-mv uniprot_sprot.fasta.gz uniprot_sprot_r2022_02.fasta.gz
-gunzip -k uniprot_sprot_r2022_02.fasta.gz
+
+mv uniprot_sprot.fasta.gz uniprot_sprot_r2023_02.fasta.gz
+gunzip -k uniprot_sprot_r2023_02.fasta.gz
 ```
@@
 ```{r engine='bash', eval=FALSE}
+
 /home/shared/ncbi-blast-2.11.0+/bin/makeblastdb \
--in ../blastdb/uniprot_sprot_r2022_02.fasta \
+-in DRAFT_Funct_Enrich/annot/uniprot_sprot_r2023_02.fasta \
 -dbtype prot \
--out ../blastdb/uniprot_sprot_r2022_02
+-out DRAFT_Funct_Enrich/annot/uniprot_sprot_r2023_02
 ```
@@
 
-No single file seems obvious or simple with respect to the primary fasta header labelled with LOCXXXXXXXXXX. Have decided to use
+The following fasta was selected for annotation: `GCF_002022765.2_C_virginica-3.0_translated_cds.faa.gz`
 
-```
->lcl|NC_035780.1_ncrna_XR_002636969.1_1 [gene=LOC111116054] [db_xref=GeneID:111116054] [product=uncharacterized LOC111116054] [ncRNA_class=lncRNA] [transcript_id=XR_002636969.1] [location=join(13578..13603,14237..14290,14557..14594)] [gbkey=ncRNA]
-TGATATTGTTGTGTGCAGAACGTGGGTCATGCTGTCTTGTTGAAGAATACCCTCAATCCAAGGCTAGATTAATCATCTAG
-CCATGATATTGATGATAGCAAGATTTGTGTTGATCCTG
+```{r engine='bash', eval=FALSE}
+cd DRAFT_Funct_Enrich/annot
+
+curl -O https://ftp.ncbi.nlm.nih.gov/genomes/all/GCF/002/022/765/GCF_002022765.2_C_virginica-3.0/GCF_002022765.2_C_virginica-3.0_translated_cds.faa.gz
+
+gunzip -k GCF_002022765.2_C_virginica-3.0_translated_cds.faa.gz
 ```
@@
 ```{r engine='bash', eval=FALSE}
-cd ../data
 
-curl -O https://ftp.ncbi.nlm.nih.gov/genomes/all/GCF/002/022/765/GCF_002022765.2_C_virginica-3.0/GCF_002022765.2_C_virginica-3.0_protein.faa.gz
+head DRAFT_Funct_Enrich/annot/GCF_002022765.2_C_virginica-3.0_translated_cds.faa
 
-gunzip -k GCF_002022765.2_C_virginica-3.0_protein.faa.gz
+```
+
+```{bash}
+/home/shared/ncbi-blast-2.11.0+/bin/blastp \
+-query DRAFT_Funct_Enrich/annot/GCF_002022765.2_C_virginica-3.0_translated_cds.faa \
+-db DRAFT_Funct_Enrich/annot/uniprot_sprot_r2023_02 \
+-out DRAFT_Funct_Enrich/annot/Cvir_transcds-uniprot_blastp.tab \
+-evalue 1E-20 \
+-num_threads 40 \
+-max_target_seqs 1 \
+-outfmt 6
 ```

### 2023-06-22 DRAFT_Funct_Enrich.Rmd
annot gff fE

@@
 ```{r engine='bash', eval=FALSE}
-
 /home/shared/ncbi-blast-2.11.0+/bin/makeblastdb \
@@
 ```{r engine='bash', eval=FALSE}
-
-head DRAFT_Funct_Enrich/annot/GCF_002022765.2_C_virginica-3.0_translated_cds.faa
+head -2 DRAFT_Funct_Enrich/annot/GCF_002022765.2_C_virginica-3.0_translated_cds.faa
 
@@
 
-```{bash}
+```
+>lcl|NC_035780.1_prot_XP_022327646.1_1 [gene=LOC111126949] [db_xref=GeneID:111126949] [protein=UNC5C-like protein] [protein_id=XP_022327646.1] [location=join(30535..31557,31736..31887,31977..32565,32959..33204)] [gbkey=CDS]
+MTEVCYIWASSSTTVVICGIFFIVWRCFISIKKRASPLHGSSQQVCQTCQIEGHDFGEFQLSCRRQNTNVGYDLQGRRSD
+
+```
+```{r engine='bash', eval=FALSE}
 /home/shared/ncbi-blast-2.11.0+/bin/blastp \
@@
 ```
+
+Here is what the output file looks like, and at this point we want to get the UniProt Accession number for each gene
+
+```{r engine='bash', eval=TRUE}
+head -2 DRAFT_Funct_Enrich/annot/Cvir_transcds-uniprot_blastp.tab \
+```
+
+```{r engine='bash', eval=FALSE}
+cd DRAFT_Funct_Enrich/annot
+
+curl -O https://ftp.ncbi.nlm.nih.gov/genomes/all/GCF/002/022/765/GCF_002022765.2_C_virginica-3.0/GCF_002022765.2_C_virginica-3.0_genomic.gff.gz
+
+gunzip -k GCF_002022765.2_C_virginica-3.0_genomic.gff.gz
+```
+
+You will notice the LOC iD is not present, but we will revisit that later with some
+
+!tr '|' '\t' < $blastout \
+> _blast-sep.tab
+
+!awk -v OFS='\t' '{print $4, $1, $14}' < _blast-sep.tab | sort \
+> _blast-sort.tab
 

### 2023-07-06 DRAFT_Funct_Enrich.Rmd
SPID to go

@@
 ---
+
+```{r}
+library(tidyverse)
+```
 
@@
 If using a widely studied system where there are publically available resources, it is advisable to use those as this is the best way to facilitate integration of data sets.
-In this case study we will be considering the Eastern oyster, (_Crassostrea virginica_) for which there is data at [NCBI](https://www.ncbi.nlm.nih.gov/datasets/genome/GCF_002022765.2/) and [Ensembl Metazoa](http://metazoa.ensembl.org/Crassostrea_virginica_gca002022765v4/Info/Index).
+In this case study we will be considering the Eastern oyster, (*Crassostrea virginica*) for which there is data at [NCBI](https://www.ncbi.nlm.nih.gov/datasets/genome/GCF_002022765.2/) and [Ensembl Metazoa](http://metazoa.ensembl.org/Crassostrea_virginica_gca002022765v4/Info/Index).
 At NCBI there is both a GenBank and RefSeq assembly available.
@@
 
+echo "number of sequences"
+grep -c  ">" DRAFT_Funct_Enrich/annot/GCF_002022765.2_C_virginica-3.0_translated_cds.faa
+
 ```
@@
 MTEVCYIWASSSTTVVICGIFFIVWRCFISIKKRASPLHGSSQQVCQTCQIEGHDFGEFQLSCRRQNTNVGYDLQGRRSD
+```
 
-```
 ```{r engine='bash', eval=FALSE}
@@
 ```{r engine='bash', eval=TRUE}
-head -2 DRAFT_Funct_Enrich/annot/Cvir_transcds-uniprot_blastp.tab \
+head -2 DRAFT_Funct_Enrich/annot/Cvir_transcds-uniprot_blastp.tab
+```
+
+```{r}
+blast <- read.csv("DRAFT_Funct_Enrich/annot/Cvir_transcds-uniprot_blastp.tab", sep = '\t', header = FALSE)
+```
+
+```{r engine='bash', eval=FALSE}
+cd DRAFT_Funct_Enrich/annot
+
+curl -O https://ftp.ncbi.nlm.nih.gov/genomes/all/GCF/002/022/765/GCF_002022765.2_C_virginica-3.0/GCF_002022765.2_C_virginica-3.0_genomic.fna.gz
+
+gunzip -k GCF_002022765.2_C_virginica-3.0_genomic.fna.gz
 ```
@@
 
+```{r}
+if (!requireNamespace("BiocManager", quietly = TRUE))
+    install.packages("BiocManager")
+
+BiocManager::install("rtracklayer")
+
+# Load the library
+library(rtracklayer)
+
+# Define the file path to your GFF file
+file_path <- "DRAFT_Funct_Enrich/annot/GCF_002022765.2_C_virginica-3.0_genomic.gff"
+
+# Import the GFF file
+gff_data <- import(file_path)
+
+# Convert GRanges object to a DataFrame for easier manipulation
+gff_data_frame <- as.data.frame(gff_data)
+```
+
+```{r}
+cds <- gff_data_frame %>%
+  filter(gbkey == "CDS", source == "Gnomon")
+```
+
+Convert fasta to tab
+
+```{bash}
+perl -e '$count=0; $len=0; while(<>) {s/\r?\n//; s/\t/ /g; if (s/^>//) { if ($.
+!= 1) {print "\n"} s/ |$/\t/; $count++; $_ .= "\t";} else {s/ //g; $len += length($_)} print $_;} print "\n"; warn "\nConverted $count FASTA records in $.
+lines to tabular format\nTotal sequence length: $len\n\n";' \
+DRAFT_Funct_Enrich/annot/GCF_002022765.2_C_virginica-3.0_translated_cds.faa > DRAFT_Funct_Enrich/annot/GCF_002022765.2_C_virginica-3.0_translated_cds.tab
+```
+
+```{bash}
+head -1 DRAFT_Funct_Enrich/annot/GCF_002022765.2_C_virginica-3.0_translated_cds.tab
+```
+
+```{r}
+cdsftab <- read.csv("DRAFT_Funct_Enrich/annot/GCF_002022765.2_C_virginica-3.0_translated_cds.tab", sep = '\t', header = FALSE, row.names=NULL)
+```
+
+join blast with cdsftab
+
+```{r}
+g.spid <- left_join(blast, cdsftab, by = "V1") %>%
+  mutate(gene = str_extract(V2.y, "(?<=\\[gene=)\\w+")) %>%
+  select(gene, V11, V2.x) %>%
+  mutate(SPID = str_extract(V2.x, "(?<=\\|)[^\\|]*(?=\\|)")) %>%
+  distinct(gene, SPID, .keep_all = TRUE)
+```
+
+```{r}
+left_join(blast, cdsftab, by = "V1") %>%
+  mutate(gene = str_extract(V2.y, "(?<=\\[gene=)\\w+")) %>%
+  select(gene, V11, V2.x) %>%
+  mutate(SPID = str_extract(V2.x, "(?<=\\|)[^\\|]*(?=\\|)")) %>%
+  distinct(gene, SPID, .keep_all = TRUE) %>%
+  select(SPID) %>%
+  write.table(file = "DRAFT_Funct_Enrich/annot/SPID.txt", sep = "\t", row.names = FALSE, quote = FALSE
+ ) 
+```
+
+```{python}
+python DRAFT_Funct_Enrich/annot/uniprot-retrieval.py DRAFT_Funct_Enrich/annot/SPID.txt
+```
+
 You will notice the LOC iD is not present, but we will revisit that later with some
 
-!tr '|' '\t' < $blastout \
-> _blast-sep.tab
+!tr '\|' '\t' \< \$blastout\
+\> \_blast-sep.tab
 
-!awk -v OFS='\t' '{print $4, $1, $14}' < _blast-sep.tab | sort \
-> _blast-sort.tab
+!awk -v OFS='\t''{print \$4, \$1, \$14}' \< \_blast-sep.tab \| sort\
+\> \_blast-sort.tab
 

### 2023-07-06 DRAFT_Funct_Enrich/annot/uniprot-retrieval.py

@@ -1,0 +1,56 @@
+import re
+import zlib
+import gzip
+import requests
+from requests.adapters import HTTPAdapter, Retry
+import sys
+
+
+re_next_link = re.compile(r'<(.+)>; rel="next"')
+retries = Retry(total=5, backoff_factor=0.25, status_forcelist=[500, 502, 503, 504])
+session = requests.Session()
+session.mount("https://", HTTPAdapter(max_retries=retries))
+
+
+def get_next_link(headers):
+    if "Link" in headers:
+        match = re_next_link.match(headers["Link"])
+        if match:
+            return match.group(1)
+
+
+def get_batch(batch_url):
+    while batch_url:
+        response = session.get(batch_url)
+        response.raise_for_status()
+        total = response.headers["x-total-results"]
+        yield response, total
+        batch_url = get_next_link(response.headers)
+
+
+def main(accession_file):
+    with open(accession_file, 'r') as f:
+        accessions = f.read().splitlines()
+
+    accession_query = '%29%20OR%20%28accession%3A'.join(accessions)
+    url = f"https://rest.uniprot.org/uniprotkb/search?compressed=true&fields=accession%2Creviewed%2Cid%2Cprotein_name%2Cgene_names%2Corganism_name%2Clength%2Cgo_p%2Cgo_c%2Cgo_f%2Cgo%2Cgo_id&format=tsv&query=%28%28accession%3A{accession_query}%29%29&size=500"
+
+    progress = 0
+    lines = []
+    for batch, total in get_batch(url):
+        # Decompress each batch as we want to extract the header
+        decompressed = zlib.decompress(batch.content, 16 + zlib.MAX_WBITS)
+        batch_lines = [line for line in decompressed.decode("utf-8").split("\n") if line]
+        if not progress:
+            # First line so print TSV header
+            lines = [batch_lines[0]]
+        lines += batch_lines[1:]
+        progress = len(lines) - 1
+        print(f"{progress} / {total}")
+
+    # Save lines to a gzip file
+    with gzip.open("uniprot-retrieval.tsv.gz", "wt", encoding="utf-8") as f:
+        f.write('\n'.join(lines))
+
+if __name__ == '__main__':
+    main(sys.argv[1])

### 2023-07-10 DRAFT_Funct_Enrich.Rmd
cleaning up annotation derails

@@
 
-```{r}
-library(tidyverse)
+```{r setup, include=FALSE}
+#library(tidyverse)
 ```
@@
 
-```{r}
+```{r eval=TRUE}}
 blast <- read.csv("DRAFT_Funct_Enrich/annot/Cvir_transcds-uniprot_blastp.tab", sep = '\t', header = FALSE)
-```
-
-```{r engine='bash', eval=FALSE}
-cd DRAFT_Funct_Enrich/annot
-
-curl -O https://ftp.ncbi.nlm.nih.gov/genomes/all/GCF/002/022/765/GCF_002022765.2_C_virginica-3.0/GCF_002022765.2_C_virginica-3.0_genomic.fna.gz
-
-gunzip -k GCF_002022765.2_C_virginica-3.0_genomic.fna.gz
-```
-
-```{r engine='bash', eval=FALSE}
-cd DRAFT_Funct_Enrich/annot
-
-curl -O https://ftp.ncbi.nlm.nih.gov/genomes/all/GCF/002/022/765/GCF_002022765.2_C_virginica-3.0/GCF_002022765.2_C_virginica-3.0_genomic.gff.gz
-
-gunzip -k GCF_002022765.2_C_virginica-3.0_genomic.gff.gz
-```
-
-```{r}
-if (!requireNamespace("BiocManager", quietly = TRUE))
-    install.packages("BiocManager")
-
-BiocManager::install("rtracklayer")
-
-# Load the library
-library(rtracklayer)
-
-# Define the file path to your GFF file
-file_path <- "DRAFT_Funct_Enrich/annot/GCF_002022765.2_C_virginica-3.0_genomic.gff"
-
-# Import the GFF file
-gff_data <- import(file_path)
-
-# Convert GRanges object to a DataFrame for easier manipulation
-gff_data_frame <- as.data.frame(gff_data)
-```
-
-```{r}
-cds <- gff_data_frame %>%
-  filter(gbkey == "CDS", source == "Gnomon")
 ```

### 2023-07-10 DRAFT_Funct_Enrich.Rmd
cleaning up annotation

@@
 ```{r setup, include=FALSE}
-#library(tidyverse)
+library(tidyverse)
 ```
@@
 
+This protein fasta is used as query for blast of uniprot_sprot database.
+
 ```{r engine='bash', eval=FALSE}
@@
 
-```{bash}
+```{r engine='bash', eval=TRUE}
 perl -e '$count=0; $len=0; while(<>) {s/\r?\n//; s/\t/ /g; if (s/^>//) { if ($.
@@
 
-```{bash}
+```{r engine='bash', eval=TRUE}
 head -1 DRAFT_Funct_Enrich/annot/GCF_002022765.2_C_virginica-3.0_translated_cds.tab
@@
 
-join blast with cdsftab
+Now we can take the two data frames: A) blast output of taking protein fasta and comparing to uniprot_swiss-prot and B) a tabular version of same fasta file that has ID numbers of importance.
+Note this importance was determined based on what we want to use down stream.
 
-```{r}
+```{r, eval=TRUE}
 g.spid <- left_join(blast, cdsftab, by = "V1") %>%

### 2023-07-10 DRAFT_Funct_Enrich.Rmd
have GO terms

@@
 
-```{r eval=TRUE}}
+```{r eval=TRUE}
 blast <- read.csv("DRAFT_Funct_Enrich/annot/Cvir_transcds-uniprot_blastp.tab", sep = '\t', header = FALSE)
@@
 
-```{r}
+Let's break it down step by step:
+
+1.
+**`g.spid <- left_join(blast, cdsftab, by = "V1")`** - This line is using the **`left_join()`** function from **`dplyr`** to merge the **`blast`** and **`cdsftab`** datasets by the column "V1".
+A left join retains all the rows in the **`blast`** data frame and appends the matching rows in the **`cdsftab`** data frame.
+If there is no match, the result is **`NA`**.
+The result of this operation is assigned to the **`g.spid`** object.
+
+2.
+**`mutate(gene = str_extract(V2.y, "(?<=\\[gene=)\\w+"))`** - This line is using the **`mutate()`** function from **`dplyr`** to add a new column called "gene" to the data frame.
+The new column is created by extracting substrings from the "V2.y" column based on the given regular expression pattern **`"(?<=\\[gene=)\\w+"`**.
+This regular expression matches and extracts any word (sequence of word characters, i.e., alphanumeric and underscore) that comes after "[gene=".
+
+3.  **`select(gene, V11, V2.x)`** - This line is using the **`select()`** function from **`dplyr`** to keep only the specified columns ("gene", "V11", and "V2.x") in the data frame.
+
+4.
+**`mutate(SPID = str_extract(V2.x, "(?<=\\|)[^\\|]*(?=\\|)"))`** - Again, the **`mutate()`** function is used to add another new column named "SPID".
+This column is created by extracting substrings from the "V2.x" column.
+The regular expression **`"(?<=\\|)[^\\|]*(?=\\|)"`** is designed to extract any character(s) that is/are surrounded by "\|" (pipe symbol).
+This is a common format for delimited strings.
+
+5.
+**`distinct(gene, SPID, .keep_all = TRUE)`** - This line is using the **`distinct()`** function from **`dplyr`** to remove duplicate rows based on the "gene" and "SPID" columns.
+The **`.keep_all = TRUE`** argument means that all other columns are also kept in the result, not just the "gene" and "SPID" columns.
+
+The resulting **`g.spid`** data frame should have unique rows with respect to the "gene" and "SPID" columns, and it should contain these two new columns, "gene" and "SPID", extracted from the original data based on specific string patterns.
+
+Now lets just write out SPIDs.
+
+```{r, eval=TRUE}
 left_join(blast, cdsftab, by = "V1") %>%
@@
 
-```{python}
-python DRAFT_Funct_Enrich/annot/uniprot-retrieval.py DRAFT_Funct_Enrich/annot/SPID.txt
+With a list of matching Swiss-Prot IDs, (technically UniProt Accession number) we can go back to https://www.uniprot.org and grab corresponding GO terms.
+This can be done via a web or using Python API.
+
+**Using Web**
+
+Using ID Mapping
+
+![id](http://gannet.fish.washington.edu/seashell/snaps/Monosnap_RetrieveID_mapping__UniProt_2023-07-10_10-03-33.png)
+
+![finished](http://gannet.fish.washington.edu/seashell/snaps/Monosnap_Tool_results__UniProt_2023-07-10_10-05-07.png)
+
+Now will customize columns to get GO IDs.
+
+![custcol](http://gannet.fish.washington.edu/seashell/snaps/Monosnap_RetrieveID_mapping_results__UniProtKB__UniProt_2023-07-10_10-07-05.png)
+
+```{r engine='bash', eval=TRUE}
+head -2 DRAFT_Funct_Enrich/annot/uniprotGO.tab
 ```
 
-You will notice the LOC iD is not present, but we will revisit that later with some
+Finally we can join table to get "LOCIDs" the notation for our DEGs, with GO terms.
 
-!tr '\|' '\t' \< \$blastout\
-\> \_blast-sep.tab
+```{r, eval=TRUE}
+go <- read.csv("DRAFT_Funct_Enrich/annot/uniprotGO.tab", sep = '\t', header = TRUE, row.names=NULL)
+```
 
-!awk -v OFS='\t''{print \$4, \$1, \$14}' \< \_blast-sep.tab \| sort\
-\> \_blast-sort.tab
+```{r, eval=TRUE}
+left_join(g.spid, go, by = c("SPID" = "Entry")) %>%
+  select(gene,Gene.Ontology.IDs) %>%
+  write.table(file = "DRAFT_Funct_Enrich/annot/geneGO.txt", sep = "\t", row.names = FALSE, quote = FALSE
+  )
+```
+
+```{r engine='bash', eval=TRUE}
+head DRAFT_Funct_Enrich/annot/geneGO.txt
+```
+
+**Using API** (DOES NOTE WORK)
+
+```{python}
+#DRAFT_Funct_Enrich/annot/uniprot-retrieval.py DRAFT_Funct_Enrich/annot/SPID.txt
+```
 

### 2023-07-14 DRAFT_Funct_Enrich.Rmd
rearranged some text, changed header name, added description in getting to enrichment analysis section

@@
 This curation process can be particularly challenging for non-model species with limited genomic and transcriptomic resources.
+
+**In some instances high quality annotation might not be possible but it is critical that before you undergo enrichment analysis you have a full understanding of how your data set was annotated, and any associated limitations**
 
@@
 
-# Overview
+# Considerations for a Rigourous Analysis
 
@@
 
-**In some instances high quality annotation might not be possible but it is critical that before you undergo enrichment analysis you have a full understanding of how your data set was annotated, and any associated limitations**
-
 # Getting to Enrichment Analysis
 
-Finding your subset of genes of interest
+Once the challenge of annotating data has been met, the next step is determining the focus of your functional analysis.
+In some cases, you will use the complete dataset, but in other cases you may be interested in a subset of interest.
+There are numerous approaches that can be used to achieve this, many with excellent tutorials available online.
+These include:
 
--   Differential expression
+-   Differential Expression Analysis
 -   Clustering

### 2023-07-14 DRAFT_Funct_Enrich.Rmd
added reticulate package for knitting. knit.
formatting updates

@@
 library(tidyverse)
+library(reticulate)
 ```
@@
 **In some instances high quality annotation might not be possible but it is critical that before you undergo enrichment analysis you have a full understanding of how your data set was annotated, and any associated limitations**
+
+<br>
 
@@
 
-# Considerations for a Rigourous Analysis
+<br>
+
+# Introduction and Considerations
+
+## Considerations for a Rigourous Analysis
 
@@
 
-# Getting to Enrichment Analysis
+<br>
+
+## Getting to Enrichment Analysis: Annotation and Filtration Decisions
 
@@
 There are numerous approaches that can be used to achieve this, many with excellent tutorials available online.
-These include:
+Some commonly used approaches include:
 
@@
 
-# Introduction and Considerations
-
--   The Gene Ontology (Thea)
--   Annotation and filtration decisions (Sam) - include key terms
--   Overrepresentation vs Enrichment (Thea)
--   Explanation of statistics (Sam) - include key terms
-
 ## The Gene Ontology
 
-## Annotation and filtration decisions
+<br>
 
-## Overrepresentation vs Enrichment
+## Overrepresentation vs Enrichment Approaches
 
@@
 
-## Explanation of Statistics
-
 <br>
 
-# Tools and Approaches
-
--   Ontologizer (command line) (Thea)
--   Panther (online GUI) - endorsed by The Gene Ontology (Thea)
--   R...Mann-Whitney U Test and Fisher's Exact Test (Sam)
-
-# A primer on the statistics of functional enrichment
+## A Primer on the Statistics of Functional Enrichment
 
@@
 
-## Fisher's Exact Test
+### Fisher's Exact Test
 
@@
 
-## Mann Whitney U Test
+### Mann Whitney U Test
 
@@
 A number of different parameters can be used to filter results or modify the execution of both tests, which we detail below.
+
+<br>
+
+# Tools and Approaches
+
+-   Ontologizer (command line) (Thea)
+-   Panther (online GUI) - endorsed by The Gene Ontology (Thea)
+-   R...Mann-Whitney U Test and Fisher's Exact Test (Sam)
 

### 2023-07-25 DRAFT_Funct_Enrich.Rmd
clean up api code

@@
 library(reticulate)
+
 ```
@@
 
-**Using API** (DOES NOTE WORK)
+**Using API**
+
+```{r, include=FALSE}
+knitr::opts_chunk$set(engine.path = '/usr/bin/python3')
+
+```
 
 ```{python}
-#DRAFT_Funct_Enrich/annot/uniprot-retrieval.py DRAFT_Funct_Enrich/annot/SPID.txt
+uniprot-retrieval.py SPID.txt
+```
+
+```{r, engine.path = '/usr/bin/python3'}
+uniprot-retrieval.py
 ```

### 2023-08-27 ADMIN_02_contributions.Rmd
dialing in navigation to contributions & best practices with new index landing page
updated contributions page

@@
+---
+hide:
+  - navigation
+---
 ---
@@
 
-# Original working group members
-
-#### [Katherine Silliman, PhD](https://ksilliman.weebly.com/), South Carolina Dept. of Natural Resources - sillimank[at]dnr.sc.gov {.unlisted .unnumbered}
+# Active Contributors 
+#### [Katherine Silliman, PhD](https://ksilliman.weebly.com/), NOAA Atlantic Oceanographic and Meteorological Lab - katherine.silliman[at]noaa.gov (Working Group Co-leader) {.unlisted .unnumbered}
 #### [Sam Bogan](http://www.hofmannlab.com/sam-bogan.html), UC-Santa Barbara -snbogan[at]ucsb.edu {.unlisted .unnumbered}
-#### [Alan Downey-Wall](https://cos.northeastern.edu/people/alan-downey-wall/), Northeastern University {.unlisted .unnumbered}
-#### Chris Mantegna, University of Washington (Intern)  {.unlisted .unnumbered}
-#### [Danielle Davenport](https://www.researchgate.net/profile/Danielle-Davenport-2), University of Queensland {.unlisted .unnumbered}  
-#### [Hanny Rivera, PhD](https://www2.whoi.edu/staff/hrivera/), Woods Hole Oceanographic Institute  {.unlisted .unnumbered}
-#### Jecy Klinkam, College of William and Mary (Intern) {.unlisted .unnumbered}
+#### Tait Algayer, Washington State University {.unlisted .unnumbered}
+#### Len Aguila, Texas A&M University Galveston {.unlisted .unnumbered}
+#### [Danielle Davenport, PhD](https://www.researchgate.net/profile/Danielle-Davenport-2), Fisheries and Oceans Canada {.unlisted .unnumbered}  
+#### Stephen Gaughran, PhD, Princeton University {.unlisted .unnumbered}
+#### Thea Gessler, Iowa State University {.unlisted .unnumbered}
 #### [Joanna Griffiths, PhD](https://joannasgriffiths.wordpress.com/), UC-Davis {.unlisted .unnumbered}
 #### Jason Johns, UC-Santa Barbara (Website Manager) {.unlisted .unnumbered}
-#### [Joe McGirr, PhD](https://joemcgirr.github.io/), UC-Davis {.unlisted .unnumbered}
-#### Kaitlin Macaranas, CSU-Bakersfield (Intern) {.unlisted .unnumbered}
-#### Megan Guidry, University of Rhode Island {.unlisted .unnumbered}
-#### Natalie Ameral, University of Rhode Island & RI Dept. of Environmental Management {.unlisted .unnumbered}  
-#### Nicolas Lou, Cornell University {.unlisted .unnumbered}
-#### [Sara Schaal](https://schaals.wixsite.com/smschaal), Northeastern University {.unlisted .unnumbered} 
-#### [Sara Wuitchik, PhD](https://sjswuitchik.weebly.com/), Harvard University {.unlisted .unnumbered}
+#### Sarit Truskey, Northeastern University {.unlisted .unnumbered}
+
+<br>
 
 ### Faculty Advisors {.unlisted .unnumbered}
-#### [Andrew Whitehead, PhD](https://whiteheadresearch.wordpress.com/), UC-Davis {.unlisted .unnumbered}
+#### [Katie Lotterhos, PhD](https://whiteheadresearch.wordpress.com/), Northeastern University {.unlisted .unnumbered}
 #### [Jon Puritz, PhD](https://www.marineevoeco.com/), University of Rhode Island {.unlisted .unnumbered}
 #### [Steven Roberts, PhD](http://faculty.washington.edu/sr320/?page_id=10850), University of Washington {.unlisted .unnumbered}
+
+<br>
+
+# Former Contributors 
+#### Natalie Ameral, University of Rhode Island & RI Dept. of Environmental Management {.unlisted .unnumbered}
+#### [Alan Downey-Wall](https://cos.northeastern.edu/people/alan-downey-wall/), Northeastern University {.unlisted .unnumbered}
+#### Megan Guidry, University of Rhode Island {.unlisted .unnumbered}
+#### Jecy Klinkam, College of William and Mary (Intern) {.unlisted .unnumbered}
+#### Nicolas Lou, Cornell University {.unlisted .unnumbered}
+#### Kaitlin Macaranas, CSU-Bakersfield (Intern) {.unlisted .unnumbered}
+#### Chris Mantegna, University of Washington (Intern)  {.unlisted .unnumbered}
+#### [Joe McGirr, PhD](https://joemcgirr.github.io/), Thermo Fisher Scientific {.unlisted .unnumbered}
+#### [Hanny Rivera, PhD](https://www2.whoi.edu/staff/hrivera/), Woods Hole Oceanographic Institute  {.unlisted .unnumbered}
+#### [Sara Schaal](https://schaals.wixsite.com/smschaal), NOAA Alaska Fisheries Science Center {.unlisted .unnumbered} 
+#### [Andrew Whitehead, PhD](https://whiteheadresearch.wordpress.com/), UC-Davis {.unlisted .unnumbered}
+#### [Sara Wuitchik, PhD](https://sjswuitchik.weebly.com/), Mount Royal University {.unlisted .unnumbered}

### 2023-08-27 ADMIN_01_submissions_instructions.Rmd
added policy for creatign a page and what to do when conflictsarise

@@
 build specific features (eg., tables, code chunks). Prior to developing
-new pages, please contact the current Website Manager Jason Johns
+new pages, please first contact the current Website Manager Jason Johns
 (jasonjohns[at]ucsb.edu). Do not hesitate to reach out with any other
@@
 
+### MarineOmics publishing process
+#### Content development and review process
+1. Initiation: The potential contributor contacts the website manager via email or the discussion forum with an idea.
+<br>
+2. Development: Meet with the website manager, Katherine Silliman and/or Sam Bogan (Working Group Co-leaders) <br>
+    + All active members will be notified and invited to be involved in the review process. <br>
+    + The contributor presents their idea for new material. <br>
+    + There will be regular meetings (frequency decided by those involved) with the website manager, Katherine Silliman and/or Sam Bogan, and any other participating members for progress updates and support.
+<br>
+3. Review: Full drafts of content will be reviewed by the website manager, Katherine Silliman and/or Sam Bogan, and at least one other active MarineOmics member assigned to the project.
+
+*See the [bottom of the page](#conflicts) for our policy on disagreements or conflicts during the development process
+
 <hr />
@@
 one that codes for this
-page](https://github.com/MarineOmics/marineomics.github.io/blob/main/submissions_instructions.Rmd))
+page](https://github.com/MarineOmics/marineomics.github.io/blob/main/ADMIN_01_submissions_instructions.Rmd))
 to see examples of how a particular feature can be accomplished, such as
@@
 
+<a id="conflicts"></a>
+
+#### Content/wording conflicts
+At any point in the process, disagreements on what content to include may arise.
+These include but are not limited to: recommendations for experimental design, techniques, and phrasing.
+All disagreements will be approached with the same method(s).
+<br>
+<br>
+Disagreements will be discussed with all of those involved at the time that the disagreement arises.
+The first attempt at resolving the disagreement will be via discussion.
+If a resolution is not agreed upon by all parties through discussion, it will be put to a majority vote including all of those actively involved at the time, not including the content author.
+In case of a tie, a MarineOmics senior scientist will be assigned to break the tie.
+
 ## References

### 2023-08-27 ADMIN_01_submissions_instructions.Rmd
minor edits to protocol for developing content

@@
 #### Content development and review process
-1. Initiation: The potential contributor contacts the website manager via email or the discussion forum with an idea.
+1. Initiation: The potential contributor contacts the website manager via email  (jasonjohns[at]ucsb.edu) or the discussion forum with an idea.
 <br>
-2. Development: Meet with the website manager, Katherine Silliman and/or Sam Bogan (Working Group Co-leaders) <br>
-    + All active members will be notified and invited to be involved in the review process. <br>
-    + The contributor presents their idea for new material. <br>
+2. Development: Meet with the website manager, Katherine Silliman and/or Sam Bogan (Working Group Co-leaders). <br>
+    + All active members will be invited to be involved in the review process. <br>
+    + The contributor presents their idea for new material to any involved working group members. <br>
     + There will be regular meetings (frequency decided by those involved) with the website manager, Katherine Silliman and/or Sam Bogan, and any other participating members for progress updates and support.

### 2023-08-27 index.Rmd
minor edits to protocol for developing content

@@
 To date these tutorials, guidelines,  and discussions have been authored and led by graduate students and postdocs with contributions from scientists at all levels, including undergraduates and faculty members.
-If you are interested in developing an information page or tutorial for the website, we ask that you review our [page with instructions for submitting content](ADMIN_01_submissions_instructions.html) and contact our Website Manager Jason Johns (jasonjohns[at]ucsb.edu).
+If you are interested in developing an information page or tutorial for the website, we ask that you review our [page with instructions for developing content](ADMIN_01_submissions_instructions.html) and contact our Website Manager Jason Johns (jasonjohns[at]ucsb.edu).
 

### 2023-08-28 ADMIN_02_contributions.Rmd
couple minor edits to contributors list

@@
 #### [Katherine Silliman, PhD](https://ksilliman.weebly.com/), NOAA Atlantic Oceanographic and Meteorological Lab - katherine.silliman[at]noaa.gov (Working Group Co-leader) {.unlisted .unnumbered}
-#### [Sam Bogan](http://www.hofmannlab.com/sam-bogan.html), UC-Santa Barbara -snbogan[at]ucsb.edu {.unlisted .unnumbered}
+#### [Sam Bogan](http://www.hofmannlab.com/sam-bogan.html), UC-Santa Barbara - snbogan[at]ucsb.edu (Working Group Co-leader) {.unlisted .unnumbered}
 #### Tait Algayer, Washington State University {.unlisted .unnumbered}
@@
 #### [Joanna Griffiths, PhD](https://joannasgriffiths.wordpress.com/), UC-Davis {.unlisted .unnumbered}
-#### Jason Johns, UC-Santa Barbara (Website Manager) {.unlisted .unnumbered}
+#### Jason Johns, UC-Santa Barbara - jasonjohns[at]ucsb.edu (Website Manager) {.unlisted .unnumbered}
 #### Sarit Truskey, Northeastern University {.unlisted .unnumbered}

### 2023-09-19 DRAFT_Funct_Enrich.Rmd
rm python path
updated python code for using annotation API

@@
 
-```{r, include=FALSE}
-knitr::opts_chunk$set(engine.path = '/usr/bin/python3')
+Instead of using the web interface, one can also use the API.
 
-```
-
-```{python}
-uniprot-retrieval.py SPID.txt
-```
-
-```{r, engine.path = '/usr/bin/python3'}
-uniprot-retrieval.py
+```{bash}
+python3 DRAFT_Funct_Enrich/annot/uniprot-retrieval.py DRAFT_Funct_Enrich/annot/SPID.txt
 ```

### 2023-09-19 DRAFT_Funct_Enrich/annot/uniprot-retrieval.py

@@ -1,10 +1,9 @@
-import re
 import zlib
 import gzip
 import requests
 from requests.adapters import HTTPAdapter, Retry
 import sys
-
+import shutil
 
 re_next_link = re.compile(r'<(.+)>; rel="next"')
 retries = Retry(total=5, backoff_factor=0.25, status_forcelist=[500, 502, 503, 504])
@@ -28,29 +27,57 @@ def get_batch(batch_url):
         batch_url = get_next_link(response.headers)
 
 
-def main(accession_file):
+def process_accessions(accessions):
+    accession_batches = [accessions[i:i+500] for i in range(0, len(accessions), 500)]
+    all_lines = []
+
+    for accession_batch in accession_batches:
+        accession_query = '%29%20OR%20%28accession%3A'.join(accession_batch)
+        url = f"https://rest.uniprot.org/uniprotkb/search?compressed=true&fields=accession%2Creviewed%2Cid%2Cprotein_name%2Cgene_names%2Corganism_name%2Clength%2Cgo_p%2Cgo_c%2Cgo%2Cgo_f%2Cgo_id&format=tsv&query=%28%28accession%3A{accession_query}%29%29&size=500"
+
+        progress = 0
+        lines = []
+        for batch, total in get_batch(url):
+            # Decompress each batch as we want to extract the header
+            decompressed = zlib.decompress(batch.content, 16 + zlib.MAX_WBITS)
+            batch_lines = [line for line in decompressed.decode("utf-8").split("\n") if line]
+            if not progress:
+                # First line so print TSV header
+                lines = [batch_lines[0]]
+            lines += batch_lines[1:]
+            progress = len(lines) - 1
+            print(f"{progress} / {total}")
+
+        all_lines.extend(lines)
+
+    return all_lines
+
+
+if __name__ == '__main__':
+    if len(sys.argv) < 2:
+        print("Usage: python uniprot-retrieval.py <input_file>")
+        sys.exit(1)
+
+    accession_file = sys.argv[1]
     with open(accession_file, 'r') as f:
         accessions = f.read().splitlines()
 
-    accession_query = '%29%20OR%20%28accession%3A'.join(accessions)
[…]

### 2023-10-05 index.Rmd
changed index and yaml per previous commit

@@
 
-* [Multifactorial RNAseq analyses](FUN_01_DGE_comparison_v2.html)  
+* [Multifactorial RNAseq analyses](DGE_comparison_v2.html)  
 * Align and count RNAseq reads (coming soon!)

### 2023-10-05 DRAFT_Funct_Enrich.Rmd
Push to do list for SB DE analysis and GO annot QC

@@
 
-## Using R
+## Comparing Fisher's Exact vs. Mann Whitney U tests
 
-<br>
+```{r}
 
-# What to Do with the Output
+## To-do list for Sam once he has data
 
--   Visualizing results
--   Interpreting results - gather consensus advice
--   R...Mann-Whitney U Test and Fisher's Exact Test (Sam)
+# 1. Conduct differential expression analysis w/ oyster data modeling continuous vs. categorical effects of pH (small effect; continuous) and sex (large effect; categorical) on gene expression
 
-<br>
+# 2. Using example GO annotations, perform FE and MWU tests
 
-# Formatting Examples
+# 3. Compare results and discuss differences... why was one effect type influenced more than the other? How does this inform what kind of test you should use?
 
-<br>
-
-## Example of table syntax
-
-| Term A | Term B | Term C | Term D | Term E | Term F | Term G |
-|:-------|:-------|:-------|:-------|:-------|:-------|:-------|
-| xxx    | xxx    | xxx    | xxx    | xxx    | xxx    | xxx    |
-| yyy    | yyy    | yyy    | yyy    | yyy    | yyy    | yyy    |
-| zzz    | zzz    | zzz    | zzz    | zzz    | zzz    | zzz    |
-
-<br>
-
-## Example of how to insert an image
-
-```{r, echo = FALSE, out.height = "40%", fig.align = "center", }
-
-knitr::include_graphics("DRAFT_Funct_Enrich/Rivera_etal_fig.png")
-
-```
-
-## Example of how to write and print an equation
-
-For illustrative purposes, here is a simple and famous equation
-
-$$
-E = mc^2
-$$
-
-<br>
-
-As an example of a more complex equation, here is an equation for the t-statistic
-
-$$
-t = \frac{\overline{x_{1}}-\overline{x_{2}}}
-{\sqrt{(s^2(
-\frac{1}{n_{1}} + \frac{1}{n_{2}}
-))}}
-$$
-
-## Example of how to cite a reference from your .bib in-text
-
-Functional enrichment tests applied to mRNA analyses often impose significant biases [@Bleazard2015-vn].
-
-````         
-```{r setup, include = FALSE} `r ''`
-
-# Set root directory... in this draft, ensure that you have the same path
-knitr::opts_knit$set(root.dir = '~/Documents/GitHub/marineomics.github.io/')
-
-```
-````
-
-<br>
-
-# Let's get started
-
-Below is an example of what should be the first, visible code chunk
-
-```{r, results = FALSE, message = FALSE, warning = FALSE}
-
-## Unhash and run the code below if you believe you may need to install the packages loaded below
-# Edit code below to list packages that are used in this draft... at present these are just examples
-
-if (!require("BiocManager", quietly = TRUE))
-    install.packages("BiocManager")
-
-BiocManager::install(c("DESeq2","edgeR","arrayQualityMetrics"))
-
-# Load packages
-library(tidyverse)
+# 4. Querying how annotation quality is influencing results
 

### 2023-10-09 POP_02_RADseq.Rmd
fixing html so google scholar connects to rad page, similar to what I did for DGE the other day

@@
+---
+title: 'RADseq'
+subtitle: '*Katherine Silliman, Danielle Davenport*'
+bibliography: POP_02_RADseq_files/common-bib_01.bib
+output: 
+  html_document:
+    pandoc_args: [ 
+      "--output=RADseq.html" 
+    ]
+---
+
 ```{r,message = FALSE,echo = FALSE}
@@
              hyperlink = "to.doc")
+
 ```
----
-title: 'RADseq'
-subtitle: '*Katherine Silliman, Danielle Davenport*'
-bibliography: POP_02_RADseq_files/common-bib_01.bib
----
 

### 2023-10-09 index.Rmd
fixing html so google scholar connects to rad page, similar to what I did for DGE the other day

@@
   * [Low coverage WGS](https://github.com/therkildsen-lab/genomic-data-analysis/blob/master/lcwgs_data_analysis.md)
-* [Reduced Representation Sequencing](POP_02_RADseq.html) 
+* [Reduced Representation Sequencing](RADseq.html) 
 * [Poolseq](POP_03_poolseq.html)

### 2023-10-12 index.Rmd
Push front page w citation instructions

@@
 
+# How to Cite{#how_to_cite}
+
+If referencing marineomics.io, please use the following citation:
+
+Bogan, S.
+N., Johns, J., Griffiths, J.
+S., Davenport, D., Smith, S.
+J., Schaal, S.
+M., Downey-Wall, A., Lou, R.
+N., Lotterhos, K., Guidry, M.
+E., Rivera, H.
+E., McGirr, J.
+A., Puritz, J.
+B., Roberts, S.
+B., and Silliman, K.
+(2023).
+A dynamic web resource for robust and reproducible genomics in nonmodel species: marineomics.io.
+Methods in Ecology and Evolution, 00, 1–8.
+https://doi.org/10.1111/2041-210X.14219
+<br>
+
+If referencing a specific page on marineomics.io please reference the above citation in addition to a reference to the individual page as follows:
+<br>
+LastName1, Initials1, LastName2, Initials2, etc. (Publication year) Webpage title. URL.
+
+For the page "Choosing a Population Genomics Approach" the citation would be written as
+
+Silliman, K. and Lou, R. N. (2022). Choosing a Population Genomics Approach. MarineOmics. https://marineomics.github.io/POP_01_choosing_population_genetics.html
+
 # Contact {#contact}

### 2023-10-12 index.Rmd
Push edits to citation instructions

@@
 A dynamic web resource for robust and reproducible genomics in nonmodel species: marineomics.io.
-Methods in Ecology and Evolution, 00, 1–8.
+*Methods in Ecology and Evolution*, 00, 1–8.
 https://doi.org/10.1111/2041-210X.14219
@@
 
-If referencing a specific page on marineomics.io please reference the above citation in addition to a reference to the individual page as follows:
+If citing a specific page, please use the above citation and a reference to the individual page as follows:
 <br>
-LastName1, Initials1, LastName2, Initials2, etc. (Publication year) Webpage title. URL.
+
+LastName1, Initials1, LastName2, Initials2, etc. (Publication year). Webpage title. URL
 

### 2023-10-12 ADMIN_01_submissions_instructions.Rmd
added initial publication year to pages, added link to how to cite section on main page, added bit about making sure to add date and link to how to cite on the submissions instructions page

@@
 ```
+## {#howtocite}
+
+Initial publication year: 2022
+<br>
+[How to cite](https://marineomics.github.io/#How_to_Cite)
 
@@
 *See the [bottom of the page](#conflicts) for our policy on disagreements or conflicts during the development process
+
+#### Please include the initial publication date (year at minimum) at the top of your page so it can be properly cited.
+We also ask that you include the below text at the [top of your page](#howtocite), which will make a link to a section on the main website about how to cite.
+
+      [How to cite](https://marineomics.github.io/#How_to_Cite)
 

### 2023-10-12 DRAFT_DNA_methylation.Rmd
added initial publication year to pages, added link to how to cite section on main page, added bit about making sure to add date and link to how to cite on the submissions instructions page

@@
 ---
+Initial publication year: TBD <br>
+[How to cite](https://marineomics.github.io/#How_to_Cite)
 

### 2023-10-12 DRAFT_Funct_Enrich.Rmd
re-renderd website to fix all links to dge url
added initial publication year to pages, added link to how to cite section on main page, added bit about making sure to add date and link to how to cite on the submissions instructions page

@@
 ---
+
+Initial publication year: TBD <br>
+[How to cite](https://marineomics.github.io/#How_to_Cite)
 
@@
 
-```{bash}
+```{bash, eval = F}
 python3 DRAFT_Funct_Enrich/annot/uniprot-retrieval.py DRAFT_Funct_Enrich/annot/SPID.txt

### 2023-10-12 DRAFT_QTL_tutorial.Rmd
added initial publication year to pages, added link to how to cite section on main page, added bit about making sure to add date and link to how to cite on the submissions instructions page

@@
 ---
+
+Initial publication year: 2023 <br>
+[How to cite](https://marineomics.github.io/#How_to_Cite)
 
@@
 ```{bash, eval = F}
-# make variables for commonly used directories
+# make variables for commonly used directories. as always, set your working directory wherever works for you
 workingdir="/Users/jason/qtl.tutorial"
@@
 
-# file with the Aquilegia reference genome
+# file with the Aquilegia reference genome, and bams for the f1 and the parent I have sequence for ('Pa')
 reference="/Volumes/HD/genomes.reference/Aquilegia/V3/sequences/Aquilegia_coerulea.main_genome.scaffolds.fasta"
-f1bam="$infodir/f2.merged.bam"
+f1bam="$bamdir/f2.merged.bam"
+pabam="/Volumes/HD/projects/genomes.species/V3/alignments/jonesii.sorted.bam"
 
@@
 # make a list file with all of the file paths of our .bams for samtools to merge
-ls -d -1 $PWD/*.* > $infodir/f2.paths.list
+ls -d -1 $bamdir/*.* > $bamdir/f2.paths.list
 
 # merge bam files
-samtools merge $infodir/f2.merged.bam -b $infodir/f2.paths.list
+samtools merge $bamdir/f2.merged.bam -b $bamdir/f2.paths.list
 
 # index bam files
-samtools index $infodir/f2.merged.bam
+samtools index $bamdir/f2.merged.bam
 ```
@@
 We can view the merged, indexed .bam file of the F2s in IGV (link).
-Because I got ~1x coverage on ~350 individuals, this merged .bam file is essentially a 350x coverage sequence file of the F1...plenty to call genotypes :).
+Because I got ~1x coverage on ~350 individuals, this merged .bam file is (on average) a 350x coverage sequence file of the F1...plenty to call a genotype at every base pair.
+As you can see from the below screenshot of this 53kb region, the coverage was quite variable.
+This is typical of the library prep method I used (formerly iGenomx, now Twist 96-plex kit), and is just fine for this application.
+
+```{r, echo = FALSE, include = TRUE, out.height = "100%", fig.align = "center", fig.cap="screen shot of the merged .bam file in IGV"}
+
+knitr::include_graphics("DRAFT_GxP_01_qtl_tutorial_files/images/merged.f2.igv.png")
+
+```
 
 #### Call and filter SNPs
+  * We can only use sites that are heterozygous in the F1, as those are sites where it inherited different alleles from either parent, and these are the sites that may be useful for genotyping our F2s across recombination bins.
+If this doesn't make sense yet it hopefully will after reading more.
+
+  ```{r, echo = FALSE, include = TRUE, out.width = "250px", fig.align = "center", fig.cap="no way to tell which allle came from which parent"}
+
+  knitr::include_graphics("DRAFT_GxP_01_qtl_tutorial_files/images/genotype_homf1.png")
+
+  ```
+
   * call SNPs with whatever program you like: we use samtools here, there are many others
-  * filter vcfs
-    * min & max depth: ___
-    * min quality: ___
-
     ```{bash, eval = F}
@@
 
-# run samtool mpileup to identify variant sites and create a .vcf file with those sites in our 'F1'
+    # run samtool mpileup to identify variant sites in the F1 and create a .vcf file with those sites
     samtools mpileup -q 30 -B -u -v -f $reference $f1bam | bcftools call -vm -f GQ > var.f1.vcf
+
+    # we can look at the first 10 lines (SNPs) after the 'contig', 'INFO', and 'FORMAT' lines
 
@@
 
+  * filter vcfs, I used bcftools, then output a text file with our useful SNP positions
+    * here's all of the code, which I'll break down piece by piece below:
+    ```{bash, eval = F}
+        cd $idallelesdir
+        bcftools filter -g3 var.F1.vcf | awk ' ($8 !~ /INDEL/) {print $0} ' | bcftools query -i 'QUAL>30 && DP>20 && DP<700 && GT="0/1" && GQ>100' -f '%CHROM\t%POS\n' > f1.het.pos
+    ```
+      * use bcftools to filter my .vcf, and remove SNPs within 3 base pairs of an INDEL
+      ```{bash, eval = F}
+        bcftools filter -g3 var.F1.vcf
+      ```
+      * remove INDELs: INDELs can be hard to genotype, and we'll have plenty of SNPs, so we keep it to just those.
+      ```{bash, eval = F}
+        awk ' ($8 !~ /INDEL/) {print $0} '
+      ```
+      * filter for depth & quality and define output format: 
+        * QUAL>30 (Phred score): 1/1000 chance that a base was called incorrectly
+        * DP>20 && DP<700: site must have between 20 and 700x coverage
+        * GT="0/1": heterozygous at that SNP
+        * GQ>100: we're really, really sure that the genotype call is correct
+        * -f '%CHROM\t%POS\n': defines the format of the output text file, outputting the chromosome, then a tab, then the number position of the SNP on that chromosome
+
+        ** Explanation of why I chose these cutoffs: We want to get some idea of our coverage, which we can do by looking at IGV above.
+We want to include sites that had enough coverage to confidently call a genotype (I used 20x) and not so high that many of those reads are a result of PCR duplicates.
+My data was filtered for PCR duplicates using Picard MarkDuplicates, but some likely still remain.
+I set my top end of acceptable coverage at 700x, as this includes sites that got a lot of 'extra' coverage, but will avoid many sites with PCR duplicates.
+      ```{bash, eval = F}
+        bcftools query -i 'QUAL>30 && DP>20 && DP<700 && GT="0/1" && GQ>100' -f '%CHROM\t%POS\n'
+      ```
+
 #### Identify useful SNPs 
-  * As mentioned previously, useful SNPs will be heterozygous in the F1
-  * If we have enough sequence data from the parents, we can determine which allele came from which parent
+  * As mentioned previously, useful SNPs will be heterozygous in the F1. We identified these sites above.
+  * If we have enough sequence data from the parents, we can determine which allele came from which parent.
   * In this case, I only had sequence data for one of the parents (Pa), which narrows the list of useful sites:
     * So not only can I only use sites where the F1 is heterozygous, I also have to use sites that with a particular genotype in Pa.
-    * Sites where the F1 and Pa are both heterozygous for the same allele are not useful, because it is still unclear which allele came from which parent:
-    * insert image to illustrate this
+    * Sites where Pa is heterozygous are not useful, because either allele could have come from Pa:
+    ```{r, echo = FALSE, include = TRUE, out.width = "250px", fig.align = "center", fig.cap="still unclear which allele came from which parent"}
+
+    knitr::include_graphics("DRAFT_GxP_01_qtl_tutorial_files/images/genotype_hetjon.png")
+
+    ```
     * So the only sites that are useful are: 
@@
       * As the latter case is relatively rare, and there were plenty of sites that satisfy case 1, I decided to use sites that were heterozygous in the F1 and homozygous in Pa
-      *show code for this
+    ```{r, echo = FALSE, include = TRUE, out.width = "250px", fig.align = "center", fig.cap="now we can determine that the T came from Pa and the G from Pb"}
+
+    knitr::include_graphics("DRAFT_GxP_01_qtl_tutorial_files/images/genotype_homjon.png")
+
+    ```
+    * We have our list of high quality, heterozygus sites from the F1, so now we need to determine their genotype in Pa and output that vcf file
+      * The below code uses [samtools mpileup](http://www.htslib.org/doc/samtools-mpileup.html) to take particular sites from a .bam file (f1.het.pos) and make a .vcf file from it with [bcftools call](https://samtools.github.io/bcftools/bcftools.html#call).
+You can look up the individuals commands on the linked manual pages.
+        ```{bash, eval = F}
+        samtools mpileup -l f1.het.pos -q 30 -u -B -f $reference $pabam | bcftools call -m -f GQ > pa.genotypes.vcf
+        ```
+    * Now we grab sites homozygous in Pa
+      * SNPs only
+      * high quality
+      * 20x < coverage < 100x (coverage on this parent was ~80x)
+      * note that all of our parameters are 'removing'excluding' sites with those specs due to the '-e' in bcftools query (e.g.
+DP<20 = remove sites with less than 20x coverage).
+You could do this either way.
+        ```{bash, eval = F}
+        bcftools filter -g3 pa.genotypes.vcf | awk ' ($8 !~ /INDEL/) {print $0} ' | bcftools query -e 'QUAL<30 || DP<20 || DP>100 || GQ<100 || GT="0/1" || GT="1/2"'      -f'%CHROM\t%POS\n' > snp.list
+        ```
+    * snp.list is our final list of usable SNPs, of which there were ~1,000,000 in this case
+    * Now we'll call genotypes for both the F1 and Pa at these sites using a combination of [samtools mpileup](http://www.htslib.org/doc/samtools-mpileup.html), [bcftools call](https://samtools.github.io/bcftools/bcftools.html#call), and [bcftools query](https://samtools.github.io/bcftools/bcftools.html#query).
+This is a simple text file with 4 columns:
+      * chromosome, position, reference allele, and the alternate allele:
+      ```{bash, eval = F}
+      samtools mpileup -l snp.list -q 30 -u -B -f $reference $f1 $pabam | bcftools call -vm -f GQ | awk ' ($8 !~ /INDEL/) {print $0} ' | bcftools query -f        '%CHROM\:%POS\t%REF\t%ALT[\t%GT]\n' > f1.pa.genotypes
+      ```
+    * Finally, we'll print a file with the sites, the Pa allele, and the Pb allele
+    ```{bash, eval = F}
+    awk ' { if ( $5 ~ /0\/0/ ) print $1 "\t" $2 "\t" $3; else print $1 "\t" $3 "\t" $2; } ' f1.pa.genotypes > pos.pa.pb.txt    
+    ```
 

### 2023-10-12 FUN_01_DGE_comparison_v2.Rmd
changing name of DGE html to match what's on Google, keeping same .Rmd name
added initial publication year to pages, added link to how to cite section on main page, added bit about making sure to add date and link to how to cite on the submissions instructions page

@@
 bibliography: FUN_01_DGE_comparison_v2_files/DGE_Comp.bib
+output: 
+  html_document:
+    pandoc_args: [ 
+      "--output=DGE_comparison_v2.html" 
+    ]
 ---
+
+Initial publication year: 2022 <br>
+[How to cite](https://marineomics.github.io/#How_to_Cite)
 

### 2023-10-12 POP_01_choosing_population_genetics.Rmd
added initial publication year to pages, added link to how to cite section on main page, added bit about making sure to add date and link to how to cite on the submissions instructions page

@@
 ```
+Initial publication year: 2022 <br>
+[How to cite](https://marineomics.github.io/#How_to_Cite)
 

### 2023-10-12 POP_02_RADseq.Rmd
added initial publication year to pages, added link to how to cite section on main page, added bit about making sure to add date and link to how to cite on the submissions instructions page

@@
 ---
+Initial publication year: 2022 <br>
+[How to cite](https://marineomics.github.io/#How_to_Cite)
 

### 2023-10-12 POP_07_WGS_Full_Walkthrough.Rmd
implementing new organizational structure to repo
removed unused images and fixed a couple file paths
added initial publication year to pages, added link to how to cite section on main page, added bit about making sure to add date and link to how to cite on the submissions instructions page

--- 2022-02-18 WGS-Full-Walkthrough.Rmd
@@
 ---
-layout: page
-title: WGS Full Walkthrough
-permalink: /WGSwalkthrough/
+title: WGS fastq2vcf walkthrough
 ---
+
+Initial publication year: 2022 <br>
+[How to cite](https://marineomics.github.io/#How_to_Cite)
 
@@
 ```
-<img src="./Pop1_17291100XcoveragePlot10000.png" width="500">
+<img src="POP_07_WGS_Full_Walkthrough_files/Pop1_17291100XcoveragePlot10000.png" width="500">
 

### 2023-11-07 DRAFT_PCA.Rmd
first pass at PCA page. need to fix links and such, but this should be the bulk of the material with most of the formatting in place

---
title: "PCA"
authors: "DL Davenport, KE Lotterhos"
date: "2023-11-07"
output:
  html_document:
    df_print: paged
---

```{r setup, eval=TRUE, echo=TRUE, message = FALSE, include = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, cache = TRUE) 
```

[How to cite](https://marineomics.github.io/#How_to_Cite)

Principal component analysis (PCA) is a multivariate method often used for summarising and visualising genetic data. In MarineOmics, PCA is commonly used to explore [population structure](/lnk to population structure section) and to highlight data issues such as batch effects, outliers, or variation to be explored (i.e. structural variation).

This page aims to provide a helpful resource for applying PCA in the field of population genomics. We provide a brief outline of the PCA method and then detail some ‘ins-and-outs’ as it relates to population genomics. Importantly, there are many existing resources for navigating PCA from both the primary literature and online tutorials, several of which we reference throughout this page. These useful resources can be found in the [References](link to section) section at the bottom of this page!

On this page we outline and describe many aspects of PCA:

* What is PCA and what are its common uses in population genetics?
* Considerations before sampling:
    + How one chooses to sample from a landscape
    + Sample sizes
* Considerations while processing data (aka 'weird stuff that happens in PCA'): 
    + Batch effects
    + LD and inversion effects
    + Missing data and imputation
    + Hybridization and inbreeding effects
    + Horseshoe and wave patterns
* PCA to inform filtering for other downstream analyses **keep?**
* Comparison to other common approaches to investigating population genetic structure
    + PCA  vs ADMIXTURE/STRUCTURE
* PCA in outlier analysis (i.e. genotype-environment analysis) **keep?**

<hr />

## What is PCA and what are its common uses in population genetics?

Primarily, PCA is about dimension-reduction. PCA captures the linear relationships (constrained, orthogonal axes) that best explain the correlated structure across the data.

In a population genomic context, the variables used for PCA are a matrix (n x m) of n samples (i.e. individuals, populations) and m genetic loci (usually microsatellite or SNP markers; genotypes, likelihoods or allele frequencies). In the case of an n x m (individual x SNP) PCA, each SNP is a variable, so that in a dataset with 500,000 SNPs there will be 500,000 variables (or dimensions) in the data.

Because we cannot visualise 500,000 dimensions, PCA helps us distill these down to increase the interpretability of data while preserving the maximum amount of information (aka variance) . It does this by creating new, uncorrelated variables that maximise variance called “principal components” (PCs, see  for the OG description). Like other multivariate methods, PCA involves a series of geometric operations and associated computations, usually achieved by matrix factorization methods (for details see REF).  In PCA, the first few PCs will account for a substantial proportion of the variation in the original variables, and can consequently be used to provide a convenient lower-dimensional summary of these variables.

The output of PCA are the eigenvalues and eigenvectors (eigenvalues = principal components (PCs), eigenvectors = principal axes) (see here for video explanation) which can be represented as graphical outputs that summarise the information of a large number of variables by a smaller number of dimensions (PCs). When applied to very large genetic datasets, the eigenvectors and eigenvalues are usually determined by a method called singular value decomposition (SVD) or some variation of it, a matrix factorisation method recommended when the number of variables is very large  (i.e. often more variables than observed entities, n << m, see this youtube playlist for details). Other methods use the covariance (or correlation) matrix. The data are usually centered (so it becomes zero mean gaussian, center = TRUE ), and scaled ( scale = TRUE ), where scaling is recommended when the variance is not homogeneous across variables (REF Jolliffe, I. (2005). Principal component analysis. Wiley Online Library).

The eigenvalues represent the amount of variance explained per component and are typically represented as a barplot, in descending order (Figure 1) (also called variable loadings). The percentage of explained variance by each eigenvector will vary between datasets since it is highly dependent on the size of the data and the correlation structure between variables. The eigenvalues are typically used as an empirical way of choosing the number of PCs to represent the data. Usually, the “elbow” of this barplot shows the number of the least amount of variables (PCs) explaining the most amount of variation.

```{r, echo = FALSE, include = TRUE, out.width = "60%", fig.align = "center", fig.cap="Figure 1. 'Screeplot' of PCA eigenvalues, where there is one eigenvalue for every eigenvector. From Spies et al. 2022 (Supp. Fig. 2)."}

knitr::include_graphics("DRAFT_PCA/Fig.1_Spies_2022_supfig2.png")

```

The eigenvectors are represented as a standard scatter plot, where points are samples/individuals represented on the new system of axes (PCs). Dimension reduction to eigenvectrors (PCs) means that the 1st PC is the linear combination of the original variables that explains the most/greatest amount of variation, while the 2nd PC accounts for the greatest amount of the remaining variation being orthogonal (uncorrelated) to the 1st PC, the 3rd PC accounts for the greatest amount of the remaining variation being orthogonal (uncorrelated) to the 1st and 2nd PC so on and so on. In population genomics, typically the 1st PC will capture broad scale population structure (because broadly distributed populations are usually the most differentiated) while subsequent axes capture more regional population structure (usually less differentiated at smaller geographic scales, see Figure 2A). However, other cool examples using PCA find the greatest amount of variation is between individuals, particularly in the case of inversions, structural variation or regions associated with a specific phenotype or adaption (see Figure 2B).

```{r, echo = FALSE, include = TRUE, out.width = "60%", fig.align = "center", fig.cap="Figure 2. PCA plots from two studies (A) from Figure 2 in x, showing population structure results for 19 Ostrea lurida populations Plots of individual admixture determined using the program STRUCTURE (top) and PCA plots for PCs 1–5. PC1 is plotted against the latitude of the sampling site, then PC2 versus PC3 and PC4 versus PC5 and (B) from Figure 2 in x, showing PCA scores of 734 individuals of Atlantic Halibut (Hippoglossus hippoglossus) from samling sites on the North West Atlantic on PC1 and PC3 axes from PCA on 86 043 SNPs, highlighting population structure (PC3) and a putative inversion (PC1). In both plots, colors refer to the phylogeographic regions of each sample."}

knitr::include_graphics("DRAFT_PCA/Fig.2_PCAs.png")

```

To get a better understanding of PCA, we recommend reading these useful reviews and studies of PCA:

* Jolliffe, I. (2005). Principal component analysis. Wiley Online Library.
https://doi.org/10.1098/rsta.2015.0202
* Pearson K (1901) On lines and planes of closest fit to systems of points in space. Philos Mag
2:559–572
* Hotelling H (1936) Relations between two sets of variates. Biometrika 28:321–377
* // add good reviews here: task Katherine Silliman and others
* Fentaw Abegaz et al.,2019. Principals about principal components in statistical genetics, Briefings in Bioinformatics. DOI: https://doi.org/10.1093/bib/bby08
* Josse J., Husson F. (2012) Handling missing values in exploratory multivariate data analysis methods. J. Soc. Française Stat., 153, 79–99.

<hr />

## Considerations before sampling
### How one chooses to sample from a landscape
How samples are selected from a metapopulation also affects the visual pattern in a PCA.

This study by [Gompert & Buerkle, 2021](https://onlinelibrary.wiley.com/doi/full/10.1111/eva.12380) simulated a metapopulation along a 1-D stepping stone model with 50 patches (patch - a location in space), where dispersal was allowed only between adjacent patches, leading to isolation by distance (https://onlinelibrary.wiley.com/doi/full/10.1111/eva.12380).

The authors sampled the patches from the landscape in different ways and then performed a PCA.

Figure 4 from their paper shows how different patterns in a PCA can arise from the same metapopulation, depending on how that population was sampled:

A) 50 individuals were sampled from each patch. Dark red and blue indicate patches on opposite ends, with lighter colors used for more central patches. This graph reflects the weak population structure in the simulation.

B) 5 individuals were sampled from each patch. Dark red and blue indicate patches on opposite ends, with lighter colors used for more central patches.

C) 50 individuals were included from patches 1–4, 24–27 and 47–50. Dark red and dark blue are used to denote peripheral patches and gray is used to denote central patches. This could be incorrectly interpreted as evidence of an isolated hybrid lineage or even hybrid species.

D) 50 individuals were sampled from patches 20–31. Dark red and dark blue are used to denote peripheral patches in the sample and gray is used to denote central patches. In this case, sampling only the central patches also exaggerated the level of population structure. This horseshoe pattern in a PCA is a consequence of distance metrics that saturate (https://journals.asm.org/doi/10.1128/mSystems.00166-16). This saturation property arises in the case of  isolation by distance. With increasing distance between sampled patches, there is a loss of information about dissimilarity among patches (i.e. dissimilarity saturates with distance), and PCA cannot discriminate between samples that do not share any common features.

```{r, echo = FALSE, include = TRUE, out.width = "70%", fig.align = "center", fig.cap="Figure 3. Figure 4 from https://onlinelibrary.wiley.com/doi/full/10.1111/eva.12380"}

knitr::include_graphics("DRAFT_PCA/Fig.3_Gompert_Buerkle_fig4.png")

```

### Sample sizes
Uneven sampling between groups, the number of samples and/or the number of markers/SNPs used in PCA can change the PCA projection space, and thus interpretation of results, relative to the true demographic history of the sampled groups.

In the paper “A Genealogical Interpretation of Principal Components Analysis,” McVean demonstrates the relationship between fundamental demographic parameters and the projection of samples onto the primary axes. This paper highlights how different demographic processes can lead to the same projections, and that projections can be influenced by uneven sampling. The study reviews nuances in how PCA is conducted. For example, if one chooses to normalize the rows to have equal variance, it will tend to up-weight the influence of rare variants.

**Number of SNPs**
The number of SNPs sampled has a large impact on the resolution of populations in PC space. In PCA, there is a critical signal-to-noise ratio below which the true structure of the signal cannot be recovered - in other words for genotype data, as genetic distance (i.e. FST) among populations decreases, the number of SNPs needed to separate the signal from the noise increases. The paper makes useful recommendations on how many SNPs would be needed to resolve population structure in PCA space for a given FST for a two population model (see  McVean //add ref):

FST > 1/sqrt(number_of_SNPs * number_of_samples_per_group) 		Eq1
Rearranging, if we know FST and the number of samples per group, we can calculate how many SNPs we would need to sequence for our PCA to be above the signal-to-noise ratio; or if we know FST and the number of SNPs, we can calculate how many samples per group we would need to sequence for our PCA to be above the signal-to-noise ratio:

Number_of_samples_per_group <- 1/(number_of_SNPs * FST^2)		Eq2

Number_of_SNPs <- 1/(Number_of_samples_per_group * FST^2)		Eq3

For instance, if FST = 0.01 and there is 100 samples per group, then you would need 1/(100 * 0.01^2) = 100 SNPs . If FST = 0.001 and there is 20 samples per group (not out of the question for a marine species), then you would need 1/(20 * 0.001^2) = 50,000 SNPs. Anecdotally, we have observed with SNP data for a high gene flow marine species (low FST) that >50,000 SNPs was needed to resolve structure in PC space between adjacent populations.

```{r, echo = FALSE, include = TRUE, out.width = "60%", fig.align = "center", fig.cap="Figure 4. Figure 2 from MacVean - Simulated a population (Fig 2A), where the expected locations for the two populations on the first PC are defined by the time-since-divergence (the lines in Fig 2C). The signal-to-noise threshold for a sample size of 100 and FST=0.01 is 100 SNPs (see Eq3 above). Fig 2C shows that around or below this threshold of 100 SNPs, the PCA is affected by random clustering. The author notes that the separation of samples (Fig 2C) with 10 SNPs does not correlate with population divergence, but reflects random clustering arising from the small numbers of SNPs."}

knitr::include_graphics("DRAFT_PCA/Fig.4_McVean_fig2.png")

```

**Unequal sample size per population**

MacVean (2009) used simulations of two populations of equal size to show how the number of individuals sampled per group influenced PC projections (Figure 2a, MacVean, 2009). Figure 2B (MacVean, 2009) shows how groups with larger sample sizes will lie closer to the origin of the first PC axis, while Figure 2c shows the effect of using fewer SNPs on the inferring the true population configuration, where lines indicate the expectation of population scores on the PC-axis.

The author extended this analysis to a 9-population stepping stone lattice. Figure 3 from the paper shows how differences in the number of samples per population can warp the projection space of PCA, even when migration rates and effective population size in each deme are the same. Note that FST among demes would be similar in all panels because it is based on an allele frequency in each deme (although small differences would occur due to sampling error when fewer individuals within a deme are sampled).

```{r, echo = FALSE, include = TRUE, out.width = "50%", fig.align = "center", fig.cap="Figure 5. Figure 3 from MacVean (2009). PCA projection of samples taken from a set of nine populations arranged in a lattice, each of which exchanges migrants at rate M per Ne generations with each adjoining neighbor, leads to a recovery of the migration-space if samples are of equal size (A), or a distortion of migration-space if populations are not equally represented (B,C). The left-hand panel for each pair shows analytical results, where the area of each point represents the relative sample size and lines show migration routes, while the right-hand panel shows how the projection space no longer reflects the migration matrix when sample size is uneven between projected groups."}

knitr::include_graphics("DRAFT_PCA/Fig.5_McVean_fig3.png")

```

<hr />

## Considerations while processing data (aka 'weird stuff that happens in PCA')
Author: KE Lotterhos

We often use PCA because we are interested in understanding population structure using a dataset of genetic markers. However, because PCA is a linear transformation of variables, it can be strongly influenced by varying degrees of non-independence among the samples and variables (i.e. SNPs) that go into the analysis, which in some cases may obscure the structure that we wish to detect. Simply put, samples in a population genetic dataset are never independent of each other due to shared evolutionary history among populations (e.g., population structure), variation in associations among nucleotides within a genome due to linkage disequilibrium, or even variation in data quality due to sequencing (i.e. batch effects)... which means weird things can happen!

In this respect, PCA is a useful tool to apply to your data, both prior to broader analysis, and as an analysis tool. Below we show how PCA can be applied including how it can be used to check for issues in your data during the filtering/cleaning stage (Batch effects, missing data, linkage disequilibrium, hybridization), how it can be used to highlight informative loci/samples for further exploration (i.e. structural variation, linkage-disequilibrium, sex-linked markers, hybrids), and how it it is commonly used to visualise population structure in a set of unlinked genetic markers.

### Batch effects
Batch effects are caused by technical differences among batches (i.e. groups) of samples in a dataset and may reflect differences in DNA quality, library preparation method, depth of sequencing, sequencing platform, read type (single vs paired-end), and/or read length (https://onlinelibrary.wiley.com/doi/full/10.1111/1755-0998.13559). These technical differences can result in differences in missing data, genotype error rates, allele frequency estimates, or SNP coverage among batches, causing different batches to appear as unique clusters in PC space. Batch effects can systematically bias genetic diversity estimates, population structure inference and selection scans (https://onlinelibrary.wiley.com/doi/full/10.1111/1755-0998.13559). Batch effects can be removed with careful read trimming and filtering (see https://onlinelibrary.wiley.com/doi/full/10.1111/1755-0998.13559 and references therein).

As an example, let us consider simulated results from Lou & Therkildsen (2022; https://onlinelibrary.wiley.com/doi/full/10.1111/1755-0998.13559) depicting batch effects related to differences in sequencing depth in a low-coverage whole genome dataset.  
Simulations consisted of nine populations on a 3x3 grid connected to neighbors in a low migration scenario (see Supporting Information at https://onlinelibrary.wiley.com/doi/full/10.1111/1755-0998.13559 for details). In this case, the PCA based on “true genotypes” should look similar to the figure below in which each of the populations are able to be distinguished as separate clusters in PC space.

```{r, echo = FALSE, include = TRUE, out.width = "30%", fig.align = "center", fig.cap="Figure 6. Example figure above modified from Figure 4 in https://onlinelibrary.wiley.com/doi/abs/10.1111/mec.16077. This figure depicts patterns of spatial population structure inferred through principal component analysis (PCA) with lcWGS data. At top is a schematic for a scenario with nine populations (each coloured node corresponds to a simulated population) on a 3x3 grid connected to neighbors via gene flow, and below is the PCA based on the true genotypes under low migration conditions."}

knitr::include_graphics("DRAFT_PCA/Fig.6_Lou_Thilk_fig4.png")

```

Under these conditions, two batches of sequencing data generation - one reflecting a sequencing depth of 0.125x and the other of 4x - were then simulated for different numbers of individuals sampled from each population. The performance of three different PCA/PCoA approaches to inferring patterns of spatial population structure under these conditions is compared in the figure below.

```{r, echo = FALSE, include = TRUE, out.width = "100%", fig.align = "center", fig.cap="Figure 7. Differences in the sensitivity of three PCA programs to batch effects caused by different sequencing depths and number of samples. The number at the top of each panel is the number of individuals sampled from each simulated population. Color represents sequencing depth: yellow points mark individuals that were simulated sequenced at 0.125×; grey points for 4×. Figure 6 from https://onlinelibrary.wiley.com/doi/full/10.1111/1755-0998.13559."}

knitr::include_graphics("DRAFT_PCA/Fig.7_Lou_Thilk_fig6.png")

```

Within each panel of the above plot, the only thing that differs between the two batches of simulated data is their sequencing depth. In this example, we see false patterns of clustering at lower sample sizes (n=5,10) when using the first two PCA approaches (PCAngsd and ANGSD with the -doCov 1 option) in which samples tended to group together according to sequencing depth along one of the PC axes (i.e. a batch effect associated with differences in sequencing depth). As the sample size per population increased from left to right, false patterns of clustering by read depth become less apparent and suggest that larger sample sizes can help to mitigate batch effects caused by differences in coverage.

### LD and Inversion effects
Linkage disequilibrium (LD) is the non-random association between alleles and can arise from physical proximity on a chromosome, demography, or selection. LD is often measured as a correlation in allele frequencies across a set of individuals and varies across a genomic map depending on recombination rates, demographic history, and selection. For example, if two SNPs are physically proximate, the allele at one SNP can predict the allele at the other SNP in the same individual because recombination is rare among sites. But LD can also evolve among SNPs that are not physically proximate, for example if an allele at each SNP is under selection. LD is a source of non-independence in genomic data. Variation in linkage disequilibrium across the map of the genome can warp principal components, as initial PC axes are biased to explain genetic variation in LD as opposed to genetic variation among populations (see Figure 2B above).

The following Figure 3 from Lotterhos 2019 (https://academic.oup.com/g3journal/article/9/6/1851/6028019) illustrates what happens in a PCA on landscape genomic data when there is recombination variation. She simulated a 2-D continuous landscape with local adaptation to an environmental cline, with a genome that consisted of linkage groups with quantitative trait loci that contributed to adaptation, a large neutral inversion (not involved in adaptation), and a region of low recombination (also not involved in adaptation). Even though the SNPs in the inversion made up less than 5% of the total number of SNPs in the PCA, and the inversion was not related to population structure, the first PC axis separated samples by inversion genotype (Figure A below - the second PC axes separated samples based on their haplotype in the region of low recombination).

When SNPs were thinned for LD and then a PCA was run on the data, the PCA showed a pattern of isolation-by-environment, which more accurately captured the population structure in the simulation (Figure B below - https://academic.oup.com/g3journal/article/9/6/1851/6028019).

```{r, echo = FALSE, include = TRUE, out.width = "50%", fig.align = "center", fig.cap="Figure 8. Figure 3 from https://academic.oup.com/g3journal/article/9/6/1851/6028019."}

knitr::include_graphics("DRAFT_PCA/Fig.8_Lotterhos2019_fig3.png")

```

Similar patterns have been observed for sex linked-markers. For example, https://onlinelibrary.wiley.com/doi/10.1111/mec.14217  found genetic clustering in a PCA by sex instead of by population structure, which was driven by as few as 12 sex-linked markers in the data.  Removing the sex-linked markers led to nonsignificant genetic structure in one species and a more accurate estimation of FST in the second species.

### Missing Data in PCA and Imputation

Missing data in population genomic analysis is common.  Sometimes it is random, or it may be a feature of the data type (i.e. low coverage NGS, microsatellite null alleles, mutation at RAD cut sites). In general, if missing data reflects a true population signal (e.g. null alleles in one population but not others) then it requires careful consideration during analysis.

However, PCA and other multivariate analyses typically do not allow missing data in the input, and many commonly used PCA methods cannot handle missing data which can bias results. It is therefore relevant to know how the implementations of PCA used often in MarineOmics and other conservation genomic studies account, or not,  for missing data.

Implementations of PCA in genomics typically (a) avoid missing values (i.e pcadapt, plink) usually by first determining the covariance-matrix or genomic relationship matrix (GRM)) between each pair of individuals using the variables (loci) that are available for both samples or by (b) impute missing values. Imputation is the process of replacing a missing value with a numerical value via inference. Imputation of missing data can be done by sophisticated methods such as in genomic association studies, linked haplotypes, and strong genotype-phenotype correlations (see Yi & Latch, 2021). However, in population genomic studies for species of conservation concern which may have few genomic resources, simple imputations such as mean imputation (replace missing with column mean) are commonly encountered.

Imputation via the mean genotype in the metapopulation.
In this case, missing values for all individuals at a locus are replaced with the mean for that locus (also called a mean-imputed matrix). Importantly, when the mean-imputed matrix is centered and scaled, the original missing become non-informative in PCA (missing values are placed at the origin). This can be problematic if you have a lot of missing data in your dataset, especially if that missing data is biased (not random). Yi and Latch (2021) demonstrated this effect using different simulated populations (no migration. 5% migration, isolation-by-distance-cline) with 1%, 10% and 20% missing data, either randomly distributed among individuals, or biased among some individuals and some populations. They conducted PCA using the adegenet::glPca() implementation of PCA with default mean imputation and standardization, retaining all PCA axes. They found that individuals biased with missing data would be dragged away from their real population clusters to the origin of PCA plots, making them indistinguishable from known admixed individuals, potentially leading to misinterpreted population structure (Figure 11).

```{r, echo = FALSE, include = TRUE, out.width = "80%", fig.align = "center", fig.cap="Figure 9. Figure 2(b) from Yi and Latch (2021) shows PCA on the individual-biased missing data (1%, 10% or 20%) introduced to a simulated population experiencing an isolation-by-distance cline, where one population (represented as triangles) is an admixture of the remaining two (circle and squares). Individual colors represent their amounts of missing data with relatively higher missingness shown in lighter blue in each plot (legend, notice changing scale from left to right). As the % of missing data increases, the samples with the most missing data become centered at the PCA origin."}

knitr::include_graphics("DRAFT_PCA/Fig.9_Yi_Latch_fig2.png")

```

Imputation via the mean genotype within a site, sampling location, or population. 
In this case, missing values for each individual sample at each locus is replaced with the mean genotype of other individuals from the same sampling location or population. The effect of this imputation in a PCA is that the mapping for that imputed sample will move closer to the center of the population cluster, and will make the PCA appear to be more clustered (Figure 12). This is because this method of imputation would make samples with missing data look more similar to the samples used to generate imputation values. This method is not recommended (see Yi & Latch 2021), since within-group mean imputation depends largely on the a priori population designation and can easily bring artificial biases.

```{r, echo = FALSE, include = TRUE, out.width = "80%", fig.align = "center", fig.cap="Figure 10. Figure 4 from Yi and Latch (2021) shows PCA on the cline model with missing data condensed in either the (a) admixed population and (b) one end population. The admixed population (pop2) has high migration with both end populations (pop1 and pop3) while no migration occurs between the end populations."}

knitr::include_graphics("DRAFT_PCA/Fig.10_Yi_Latch_fig4.png")

```

**Effects of different approaches to imputation on PCA.**

The choice of imputation can have drastic effects on inference if there is a lot of missing data. When there is population structure and missing genotypes are imputed with the mean genotype across all populations, the genetic differentiation among populations may be artificially reduced and the metapopulation would appear to be panmictic.  When there is no population structure and missing genotypes are imputed with the mean genotype within each sampling location, the genetic differentiation among populations may be artificially inflated and it could appear that their genetic structure when there is in fact none. As such, it is really important to understand the type of imputation a PCA method might be implementing  and its effect  on population clustering and final inferences.

Since most genomic data has missingness, it may be preferable to implement a method of PCA which explicitly accepts it (i.e pcadapt, PCAngsd).

### Hybridization and inbreeding effects
Genetic differences between two different parental species typically constitute the dominant axis of genetic variation in a PCA, with hybrids mapping in PC space between the parental species (https://onlinelibrary.wiley.com/doi/full/10.1111/eva.12380 ). However, PC axes can also pick up on differences in homozygosity caused by many generations of inbreeding. These PCA patterns can be informative during early data exploration for identifying cryptic species or hybrids that you may want to remove prior to analyzing for population structure.

Below is Figure 5 from https://onlinelibrary.wiley.com/doi/full/10.1111/eva.12380, with two parental species shown in dark red or dark blue, the F1 hybrids in yellow circles, and the F1-parental backcrosses in the lighter colors. Subsequent crosses between hybrids are shown in yellow with different symbols (F2 squares, F5 upward triangles, F20 downward triangles). F1...FN are genetically intermediate on PC1, and across all hybrids, PC1 mirrors the admixture proportion. In general, PC2 is associated with genetic variation among Fn generations.

```{r, echo = FALSE, include = TRUE, out.width = "50%", fig.align = "center", fig.cap="Figure 11. Figure 5 from https://onlinelibrary.wiley.com/doi/full/10.1111/eva.12380."}

knitr::include_graphics("DRAFT_PCA/Fig.11_Gompert_Buerkle_fig5.png")

```

### Horseshoe and Wave patterns
Horseshoe patterns arise in population genetic data that has isolation by distance structure, by which genetic differentiation among locations increases with the distance between them. The result when this type of data is put into a PCA is that the pattern looks like a horseshoe (Figure 1B from Morton et al below). In Figure 1B, Sample 0 is most distantly related from Sample 20, but they have the same loading on the PC1 axis. Horseshoe patterns also arise in microbial ecology, where they were originally misinterpreted as having an ecological explanation before they were proven to be a statistical artifact (https://journals.asm.org/doi/10.1128/mSystems.00166-16). The pattern arises in microbial data or genetic data as a consequence of distance metrics that saturate (Figure 1A and C below), typically when distance metrics as used in PCA cannot discriminate between samples that do not share any common features. Figure 1A below shows an isolation by distance or “band” pattern, in which neighboring samples have mostly similar alleles, but the similarity in allelic composition declines with the distance between samples. For example, “Sample 0” has the reference alleles at SNPs 1-10, but does not share the same alleles at any SNPs with samples 10-20. As a result, the euclidean distance (on which a PCA is based) saturates with increasing distance between samples (Figure 1C below).

```{r, echo = FALSE, include = TRUE, out.width = "70%", fig.align = "center", fig.cap="Figure 12. Figure 1 from https://journals.asm.org/doi/10.1128/mSystems.00166-16.  (a) A band table for haploids where SNPs are in rows, samples are in columns, and the color is the allele (0 or 1). Blocks that are colored black have a value of 1 while blocks that are colored white have a value of 0. (b) The first 2 components from a PCA of the band table, yielding the typical horseshoe shape. (c) The Euclidean distance (e.g., genetic relatedness) from point 0 to all of the other points. (d) An illustration of the distance saturation property."}

knitr::include_graphics("DRAFT_PCA/Fig.12_Morton_fig1.png")

```

The saturation property of PCA can also create “wave” patterns in a PCA of genetic data, which were initially misinterpreted to be signatures of specific migration events before they were shown to be statistical artifacts (Novembre & Stephens). Figure 2 from Novembre & Stephens below shows the pattern arising from simulating a 1-D stepping stone model, which is analogous to marine samples that are collected along a coastline. The horseshoe pattern arises in the first 2PC axes (Figure 2C), but this example also shows the wave patterns that arise on subsequent PC axes.

```{r, echo = FALSE, include = TRUE, out.width = "70%", fig.align = "center", fig.cap="Figure 13. Figure 2 from Novembre & Stephens. (a) Schematic of the one-dimensional habitat, with circles marking sampling locations and shades of blue marking order along the line. (b) One-dimensional PC maps (that is, plots of each PC element against the geographic position of the corresponding sample location). (c) Biplots of PC1 versus PC2, PC2 versus PC3, and PC3 versus PC4. Colors correspond to those in a. In many datasets without spatially referenced samples, the colors and the lines connecting neighboring points would not be observed; here they are shown to aid interpretation."}

knitr::include_graphics("DRAFT_PCA/Fig.13_Novembre_Stephens_fig2.png")

```

These highly structured patterns are mathematical artifacts that arise generally when PCA is applied to spatial data in which covariance (similarity) between locations tends to decay with geographic distance. These artifacts happen because the distance metrics on which PCA is based do not capture all of the information about genetic dissimilarity along a gradient - in other words, distance metrics cannot discriminate between samples that do not share any common features. Understanding the statistical reasons for the horseshoe and wave effects is useful for correct interpretation of these patterns.

<hr />

## Relationship between PCA and Admixture

In cases of simple two- or three-way admixture, where close relatives of the source populations can be identified and sampled, estimation of admixture proportions can be achieved from projecting samples onto the PCs identified from the source populations. Figure 5 from McVean 2009 below shows an example for human populations from the HapMap3 project. Samples having African ancestry in Southwest USA (ASW) and two groups of samples representing the “source” populations for ASW: (1) Yoruba in Ibadan, Nigeria (YRI) and (2) Utah USA with Northern and Western European ancestry (CEU). The ASW group is admixed between the YRI and CEU groups. The YRI and CEU groups are not necessarily the source populations for the admixed group, but they are closely related to the true source. It is important when using PCA to infer admixture that the PCA is conducted on each chromosome on the two source groups, and all samples are subsequently projected onto the PCA.

In MacVean (2009) Figure 5A below, PCA is carried out only on the haplotypes from CEU and YRI and all samples are subsequently projected onto the first PC identified from this analysis. Figure 5A shows how the loadings of individuals map onto PC1 each chromosome (in rows), and note the uniformity of the loadings for the source populations. However for the admixed group ASW, note how there is considerable variation at the level of individual chromosomes, with some chromosomes within an individual appearing essentially European (when a green point maps at the blue end) and others essentially African (when a green point maps at the orange end). 
Figure 5B below shows how the genome-wide admixture proportions can be inferred directly from the location of admixed samples on the first PC between the two source populations.

```{r, echo = FALSE, include = TRUE, out.width = "50%", fig.align = "center", fig.cap="Figure 14. Figure 5 from https://journals.plos.org/plosgenetics/article?id=10.1371/journal.pgen.1000686. See text above for explanation."}

knitr::include_graphics("DRAFT_PCA/Fig.14_McVean_fig5.png")

```

Table 1. Common PCA methods implemented in population genomics studies, a description of how they work, and how they treat missing data.  Abbreviations include Singular Value Decomposition (SVD), Single Nucleotide Polymorphism (SNP), genomic relationship matrix (GRM), minor allele frequency (MAF)

| Method or Function, command call | Treatment of Missingness | Ref |
|:---------------------------------|:-------------------------|:----|
|pcadapt|Missing data are coded as ‘9’ in the input. Applies a covariance approach to missing data, outlined in (Dray & Josse, 2015); this approach deals with missing data by only computing the covariance matrix (also called the genomic relationship matrix (GRM)) between each pair of individuals using the variables (loci) that are available for both individuals (at least until version < 3.x). The updated pcadapt (v.4.x)  is very fast, using a truncated SVD algorithm and accounts for missing data using a function which computes the product and cross-product of the (scaled) genotype matrix with a given vector (see Privé et al., 2020 supplementary materials). If you have A LOT of missing data, this method can affect the finding of ‘outlier loci’ (significance of loadings).|(Privé et al., 2020; Luu, et al.,  2016)|
|NIPALS (Nonlinear Iterative Partial Least Squares)|Missing values are initially set to the row averages, and SVD of the SNP matrix is used to create orthogonal PCs. The PCs which correspond to the largest eigenvalues are then used to reconstruct the missing SNP genotypes in the SNP matrix (REF). |World, 1996|
|adegenet::glPCA|Missing values are replaced by the dosage (a factor homozygous ref = 0, heterozygous = 1, homozygous alternate = 2) mean of the column representing the variant |**fill in**|
|SNPRelate::SNPRelatePCA|Missing values are imputed by the dosage mean|**fill in**|
|PCAngsd/EMU|Missing data is imputed based on population structure inferred from the posterior genotype dosages from the top K inferred PCs. The method to choose the top PCs to represent population structure and  estimate the individual allele frequencies follows Velicier’s minimum average partial (MAP) (Shriner, 2011).|(Meisner, Liu, et al. 2021;Meisner & Albrechtsen, 2018)|
|dudi.pca()|Does not work with missing genotypes. Missingness can manually be replaced or removed|**fill in**|
|BaseR, prcomp()|Does not work with missing genotypes. Missingness can manually be replaced or removed|**fill in**|
|vegan::RDA()|This function performs a PCA when no conditions are added. Does not work with missing genotypes. Missingness can manually be replaced, usually by the dosage mean.|**fill in**|
|Plink PCA|Can implement different computations: The standard computation computes the covariance matrix between pairs of individuals, using only variables (loci) that are available for both individuals (but you can add the modifier “meanimpute” for mean imputation). The randomisation algorithm (aka. FASTPCA) (--pca approx) performs mean-imputation of missing genotypes |//plink handbook. Randomisation algorithm (aka see)|
|bigSNPR:, big_SVD() and big_randomSVD()|Does not work with missing data, but offers various  functions to impute missingness based on plink, beagle or methods outlined in (Wang et al., 2012)|**fill in**|
|smartPCA|Based on Randomized Singular Value Decomposition (RSVD) (Halko et al. 2011), performs full eigendecomposition on the GRM with mean imputation|**fill in**|

### 2023-11-08 DRAFT_PCA.Rmd
added .bib, made a bunch of links and added some refs

@@
 title: "PCA"
+bibliography: DRAFT_PCA/PCA_marineomics.bib
 authors: "DL Davenport, KE Lotterhos"
@@
 Importantly, there are many existing resources for navigating PCA from both the primary literature and online tutorials, several of which we reference throughout this page.
-These useful resources can be found in the [References](link to section) section at the bottom of this page!
+These useful resources can be found in the [References](#refs) section at the bottom of this page!
 
@@
     + Horseshoe and wave patterns
-* PCA to inform filtering for other downstream analyses **keep?**
 * Comparison to other common approaches to investigating population genetic structure
     + PCA  vs ADMIXTURE/STRUCTURE
-* PCA in outlier analysis (i.e. genotype-environment analysis) **keep?**
 
@@
 
-The output of PCA are the eigenvalues and eigenvectors (eigenvalues = principal components (PCs), eigenvectors = principal axes) (see here for video explanation) which can be represented as graphical outputs that summarise the information of a large number of variables by a smaller number of dimensions (PCs).
+The output of PCA are the eigenvalues and eigenvectors (eigenvalues = principal components (PCs), eigenvectors = principal axes) (see [here](https://www.youtube.com/watch?v=PFDu9oVAE-g) for video explanation) which can be represented as graphical outputs that summarise the information of a large number of variables by a smaller number of dimensions (PCs).
 When applied to very large genetic datasets, the eigenvectors and eigenvalues are usually determined by a method called singular value decomposition (SVD) or some variation of it, a matrix factorisation method recommended when the number of variables is very large  (i.e.
-often more variables than observed entities, n << m, see this youtube playlist for details).
+often more variables than observed entities, n << m, see this [YouTube playlist](https://www.youtube.com/playlist?list=PLMrJAkhIeNNSVjnsviglFoY2nXildDCcv) for details).
 Other methods use the covariance (or correlation) matrix.
-The data are usually centered (so it becomes zero mean gaussian, center = TRUE ), and scaled ( scale = TRUE ), where scaling is recommended when the variance is not homogeneous across variables (REF Jolliffe, I.
-(2005).
-Principal component analysis.
-Wiley Online Library).
+The data are usually centered (so it becomes zero mean gaussian, center = TRUE ), and scaled ( scale = TRUE ), where scaling is recommended when the variance is not homogeneous across variables (@Jolliffe2005-hw).
 
@@
 In this respect, PCA is a useful tool to apply to your data, both prior to broader analysis, and as an analysis tool.
-Below we show how PCA can be applied including how it can be used to check for issues in your data during the filtering/cleaning stage (Batch effects, missing data, linkage disequilibrium, hybridization), how it can be used to highlight informative loci/samples for further exploration (i.e.
+Below we show how PCA can be applied including how it can be used to check for issues in your data during the filtering/cleaning stage ([batch effects](#batcheffects), [missing data](#missingdata), [linkage disequilibrium](#ld), [hybridization](#hybrid)), how it can be used to highlight informative loci/samples for further exploration (i.e.
 structural variation, linkage-disequilibrium, sex-linked markers, hybrids), and how it it is commonly used to visualise population structure in a set of unlinked genetic markers.
 
-### Batch effects
+### Batch effects {#batcheffects}
 Batch effects are caused by technical differences among batches (i.e.
@@
 
-### LD and Inversion effects
+### LD and Inversion effects {#ld}
 Linkage disequilibrium (LD) is the non-random association between alleles and can arise from physical proximity on a chromosome, demography, or selection.
@@
 
-### Missing Data in PCA and Imputation
+### Missing Data in PCA and Imputation {#missingdata}
 
@@
 This can be problematic if you have a lot of missing data in your dataset, especially if that missing data is biased (not random).
-Yi and Latch (2021) demonstrated this effect using different simulated populations (no migration.
+Yi and Latch (2022) demonstrated this effect using different simulated populations (no migration.
 5% migration, isolation-by-distance-cline) with 1%, 10% and 20% missing data, either randomly distributed among individuals, or biased among some individuals and some populations.
@@
 ```{r, echo = FALSE, include = TRUE, out.width = "80%", fig.align = "center", fig.cap="Figure 9.
-Figure 2(b) from Yi and Latch (2021) shows PCA on the individual-biased missing data (1%, 10% or 20%) introduced to a simulated population experiencing an isolation-by-distance cline, where one population (represented as triangles) is an admixture of the remaining two (circle and squares).
+Figure 2(b) from Yi and Latch (2022) shows PCA on the individual-biased missing data (1%, 10% or 20%) introduced to a simulated population experiencing an isolation-by-distance cline, where one population (represented as triangles) is an admixture of the remaining two (circle and squares).
 Individual colors represent their amounts of missing data with relatively higher missingness shown in lighter blue in each plot (legend, notice changing scale from left to right).
@@
 ```{r, echo = FALSE, include = TRUE, out.width = "80%", fig.align = "center", fig.cap="Figure 10.
-Figure 4 from Yi and Latch (2021) shows PCA on the cline model with missing data condensed in either the (a) admixed population and (b) one end population.
+Figure 4 from Yi and Latch (2022) shows PCA on the cline model with missing data condensed in either the (a) admixed population and (b) one end population.
 The admixed population (pop2) has high migration with both end populations (pop1 and pop3) while no migration occurs between the end populations."}
@@
 
-### Hybridization and inbreeding effects
+### Hybridization and inbreeding effects {#hybrid}
 Genetic differences between two different parental species typically constitute the dominant axis of genetic variation in a PCA, with hybrids mapping in PC space between the parental species (https://onlinelibrary.wiley.com/doi/full/10.1111/eva.12380 ).
@@
 In cases of simple two- or three-way admixture, where close relatives of the source populations can be identified and sampled, estimation of admixture proportions can be achieved from projecting samples onto the PCs identified from the source populations.
-Figure 5 from McVean 2009 below shows an example for human populations from the HapMap3 project.
+Figure 5 from McVean 2009 below shows an example for human populations from the [HapMap3 project](https://www.sanger.ac.uk/science/tools/#).
 Samples having African ancestry in Southwest USA (ASW) and two groups of samples representing the “source” populations for ASW: (1) Yoruba in Ibadan, Nigeria (YRI) and (2) Utah USA with Northern and Western European ancestry (CEU).
@@
 | Method or Function, command call | Treatment of Missingness | Ref |
-|:---------------------------------|:-------------------------|:----|
+|:----------|:------------------------------------------------|:----|
 |pcadapt|Missing data are coded as ‘9’ in the input.
@@
 The updated pcadapt (v.4.x)  is very fast, using a truncated SVD algorithm and accounts for missing data using a function which computes the product and cross-product of the (scaled) genotype matrix with a given vector (see Privé et al., 2020 supplementary materials).
-If you have A LOT of missing data, this method can affect the finding of ‘outlier loci’ (significance of loadings).|(Privé et al., 2020; Luu, et al.,  2016)|
+If you have A LOT of missing data, this method can affect the finding of ‘outlier loci’ (significance of loadings).|@privePerformingHighlyEfficient2020; @luuPcadaptPackagePerform2017|
 |NIPALS (Nonlinear Iterative Partial Least Squares)|Missing values are initially set to the row averages, and SVD of the SNP matrix is used to create orthogonal PCs.
 The PCs which correspond to the largest eigenvalues are then used to reconstruct the missing SNP genotypes in the SNP matrix (REF).
-|World, 1996|
-|adegenet::glPCA|Missing values are replaced by the dosage (a factor homozygous ref = 0, heterozygous = 1, homozygous alternate = 2) mean of the column representing the variant |**fill in**|
-|SNPRelate::SNPRelatePCA|Missing values are imputed by the dosage mean|**fill in**|
+|@wold1975path|
+|adegenet::glPCA|Missing values are replaced by the dosage (a factor homozygous ref = 0, heterozygous = 1, homozygous alternate = 2) mean of the column representing the variant |@jombart2008adegenet|
+|SNPRelate::SNPRelatePCA|Missing values are imputed by the dosage mean|@zhengHighperformanceComputingToolset2012a|
 |PCAngsd/EMU|Missing data is imputed based on population structure inferred from the posterior genotype dosages from the top K inferred PCs.
-The method to choose the top PCs to represent population structure and  estimate the individual allele frequencies follows Velicier’s minimum average partial (MAP) (Shriner, 2011).|(Meisner, Liu, et al.
-2021;Meisner & Albrechtsen, 2018)|
-|dudi.pca()|Does not work with missing genotypes. Missingness can manually be replaced or removed|**fill in**|
-|BaseR, prcomp()|Does not work with missing genotypes. Missingness can manually be replaced or removed|**fill in**|
-|vegan::RDA()|This function performs a PCA when no conditions are added. Does not work with missing genotypes. Missingness can manually be replaced, usually by the dosage mean.|**fill in**|
+The method to choose the top PCs to represent population structure and  estimate the individual allele frequencies follows Velicier’s minimum average partial (MAP) (@shrinerInvestigatingPopulationStratification2011).|@meisnerLargescaleInferencePopulation2021; @meisnerInferringPopulationStructure2018|
+|dudi.pca()|Does not work with missing genotypes. Missingness can manually be replaced or removed|@dray2007ade4|
+|BaseR, prcomp()|Does not work with missing genotypes. Missingness can manually be replaced or removed|@rcoreteamLanguageEnvironmentStatistical2013|
+|vegan::RDA()|This function performs a PCA when no conditions are added.
+Does not work with missing genotypes.
+Missingness can manually be replaced, usually by the dosage mean.|@oksanenVeganCommunityEcology2019|
 |Plink PCA|Can implement different computations: The standard computation computes the covariance matrix between pairs of individuals, using only variables (loci) that are available for both individuals (but you can add the modifier “meanimpute” for mean imputation).
 The randomisation algorithm (aka.
-FASTPCA) (--pca approx) performs mean-imputation of missing genotypes |//plink handbook.
-Randomisation algorithm (aka see)|
-|bigSNPR:, big_SVD() and big_randomSVD()|Does not work with missing data, but offers various  functions to impute missingness based on plink, beagle or methods outlined in (Wang et al., 2012)|**fill in**|
-|smartPCA|Based on Randomized Singular Value Decomposition (RSVD) (Halko et al. 2011), performs full eigendecomposition on the GRM with mean imputation|**fill in**|
+FASTPCA) (--pca approx) performs mean-imputation of missing genotypes (see @galinskyFastPrincipalComponentAnalysis2016a)|
+ |@purcell2007plink|
+|bigSNPR:, big_SVD() and big_randomSVD()|Does not work with missing data, but offers various  functions to impute missingness based on plink, beagle or methods outlined in @wang2012fast|@prive2018efficient; @Prive2020-fp|
+
+## References {#refs}

### 2023-11-11 DRAFT_PCA.Rmd
fixed more references and links for PCA page. full draft ready for review

@@
 Principal component analysis (PCA) is a multivariate method often used for summarising and visualising genetic data.
-In MarineOmics, PCA is commonly used to explore [population structure](/lnk to population structure section) and to highlight data issues such as batch effects, outliers, or variation to be explored (i.e.
+In MarineOmics, PCA is commonly used to explore population structure and to highlight data issues such as batch effects, outliers, or variation to be explored (i.e.
 structural variation).
@@
 Because we cannot visualise 500,000 dimensions, PCA helps us distill these down to increase the interpretability of data while preserving the maximum amount of information (aka variance) .
-It does this by creating new, uncorrelated variables that maximise variance called “principal components” (PCs, see  for the OG description).
-Like other multivariate methods, PCA involves a series of geometric operations and associated computations, usually achieved by matrix factorization methods (for details see REF).
+It does this by creating new, uncorrelated variables that maximise variance called “principal components” (PCs, see @Pearson1901-xe).
+Like other multivariate methods, PCA involves a series of geometric operations and associated computations, usually achieved by matrix factorization methods (for details see @Thioulouse2018-bj).
 In PCA, the first few PCs will account for a substantial proportion of the variation in the original variables, and can consequently be used to provide a convenient lower-dimensional summary of these variables.
@@
 'Screeplot' of PCA eigenvalues, where there is one eigenvalue for every eigenvector.
-From Spies et al.
-2022 (Supp.
+From @Spies2022-xj (Supp.
 Fig.
@@
 ```{r, echo = FALSE, include = TRUE, out.width = "60%", fig.align = "center", fig.cap="Figure 2.
-PCA plots from two studies (A) from Figure 2 in x, showing population structure results for 19 Ostrea lurida populations Plots of individual admixture determined using the program STRUCTURE (top) and PCA plots for PCs 1–5.
-PC1 is plotted against the latitude of the sampling site, then PC2 versus PC3 and PC4 versus PC5 and (B) from Figure 2 in x, showing PCA scores of 734 individuals of Atlantic Halibut (Hippoglossus hippoglossus) from samling sites on the North West Atlantic on PC1 and PC3 axes from PCA on 86 043 SNPs, highlighting population structure (PC3) and a putative inversion (PC1).
+PCA plots from two studies (A) from Figure 2 in @sillimanPopulationStructureGenetic2019, showing population structure results for 19 Ostrea lurida populations Plots of individual admixture determined using the program STRUCTURE (top) and PCA plots for PCs 1–5.
+PC1 is plotted against the latitude of the sampling site, then PC2 versus PC3 and PC4 versus PC5 and (B) from Figure 2 in @kessPutativeStructuralVariant2021, showing PCA scores of 734 individuals of Atlantic Halibut (Hippoglossus hippoglossus) from samling sites on the North West Atlantic on PC1 and PC3 axes from PCA on 86 043 SNPs, highlighting population structure (PC3) and a putative inversion (PC1).
 In both plots, colors refer to the phylogeographic regions of each sample."}
@@
 
-* Jolliffe, I. (2005). Principal component analysis. Wiley Online Library.
-https://doi.org/10.1098/rsta.2015.0202
-* Pearson K (1901) On lines and planes of closest fit to systems of points in space. Philos Mag
-2:559–572
-* Hotelling H (1936) Relations between two sets of variates. Biometrika 28:321–377
-* // add good reviews here: task Katherine Silliman and others
-* Fentaw Abegaz et al.,2019. Principals about principal components in statistical genetics, Briefings in Bioinformatics. DOI: https://doi.org/10.1093/bib/bby08
-* Josse J., Husson F. (2012) Handling missing values in exploratory multivariate data analysis methods. J. Soc. Française Stat., 153, 79–99.
+* [Jolliffe, I. & Cadima, J. (2016) Principal component analysis: a review and recent developments](https://doi.org/10.1098/rsta.2015.0202)
+* [Pearson K. (1901) On lines and planes of closest fit to systems of points in space. Philos Mag 2:559–572](https://doi.org/10.1080/14786440109462720)
+* [Hotelling H (1936) Relations between two sets of variates. Biometrika 28:321–377](https://doi.org/10.1007/978-1-4612-4380-9_14)
+* [Fentaw Abegaz et al. (2019) Principals about principal components in statistical genetics, Briefings in Bioinformatics](https://doi.org/10.1093/bib/bby08)
+* [Josse J., Husson F. (2012) Handling missing values in exploratory multivariate data analysis methods. J. Soc. Française Stat., 153, 79–99.](http://www.numdam.org/item/JSFS_2012__153_2_79_0/)
 
@@
 
-This study by [Gompert & Buerkle, 2021](https://onlinelibrary.wiley.com/doi/full/10.1111/eva.12380) simulated a metapopulation along a 1-D stepping stone model with 50 patches (patch - a location in space), where dispersal was allowed only between adjacent patches, leading to isolation by distance (https://onlinelibrary.wiley.com/doi/full/10.1111/eva.12380).
+This study by @Gompert2016-wa simulated a metapopulation along a 1-D stepping stone model with 50 patches (patch - a location in space), where dispersal was allowed only between adjacent patches, leading to isolation by distance (@Gompert2016-wa).
 
@@
 In this case, sampling only the central patches also exaggerated the level of population structure.
-This horseshoe pattern in a PCA is a consequence of distance metrics that saturate (https://journals.asm.org/doi/10.1128/mSystems.00166-16).
+This horseshoe pattern in a PCA is a consequence of distance metrics that saturate (@Morton2017-xy).
 This saturation property arises in the case of  isolation by distance.
@@
 
-```{r, echo = FALSE, include = TRUE, out.width = "70%", fig.align = "center", fig.cap="Figure 3. Figure 4 from https://onlinelibrary.wiley.com/doi/full/10.1111/eva.12380"}
+```{r, echo = FALSE, include = TRUE, out.width = "70%", fig.align = "center", fig.cap="Figure 3. Figure 4 from @Gompert2016-wa"}
 
@@
 FST) among populations decreases, the number of SNPs needed to separate the signal from the noise increases.
-The paper makes useful recommendations on how many SNPs would be needed to resolve population structure in PCA space for a given FST for a two population model (see  McVean //add ref):
+The paper makes useful recommendations on how many SNPs would be needed to resolve population structure in PCA space for a given FST for a two population model (see @McVean2009-ms):
 
@@
 Batch effects are caused by technical differences among batches (i.e.
-groups) of samples in a dataset and may reflect differences in DNA quality, library preparation method, depth of sequencing, sequencing platform, read type (single vs paired-end), and/or read length (https://onlinelibrary.wiley.com/doi/full/10.1111/1755-0998.13559).
+groups) of samples in a dataset and may reflect differences in DNA quality, library preparation method, depth of sequencing, sequencing platform, read type (single vs paired-end), and/or read length (@Lou2022-lc).
 These technical differences can result in differences in missing data, genotype error rates, allele frequency estimates, or SNP coverage among batches, causing different batches to appear as unique clusters in PC space.
-Batch effects can systematically bias genetic diversity estimates, population structure inference and selection scans (https://onlinelibrary.wiley.com/doi/full/10.1111/1755-0998.13559).
-Batch effects can be removed with careful read trimming and filtering (see https://onlinelibrary.wiley.com/doi/full/10.1111/1755-0998.13559 and references therein).
+Batch effects can systematically bias genetic diversity estimates, population structure inference and selection scans (@Lou2022-lc).
+Batch effects can be removed with careful read trimming and filtering (see @Lou2022-lc and references therein).
 
-As an example, let us consider simulated results from Lou & Therkildsen (2022; https://onlinelibrary.wiley.com/doi/full/10.1111/1755-0998.13559) depicting batch effects related to differences in sequencing depth in a low-coverage whole genome dataset.
-Simulations consisted of nine populations on a 3x3 grid connected to neighbors in a low migration scenario (see Supporting Information at https://onlinelibrary.wiley.com/doi/full/10.1111/1755-0998.13559 for details).
+As an example, let us consider simulated results from @Lou2022-lc depicting batch effects related to differences in sequencing depth in a low-coverage whole genome dataset.  
+Simulations consisted of nine populations on a 3x3 grid connected to neighbors in a low migration scenario (see Supporting Information at @Lou2022-lc for details).
 In this case, the PCA based on “true genotypes” should look similar to the figure below in which each of the populations are able to be distinguished as separate clusters in PC space.
@@
 ```{r, echo = FALSE, include = TRUE, out.width = "30%", fig.align = "center", fig.cap="Figure 6.
-Example figure above modified from Figure 4 in https://onlinelibrary.wiley.com/doi/abs/10.1111/mec.16077.
+Example figure above modified from Figure 4 in @Lou2021-at.
 This figure depicts patterns of spatial population structure inferred through principal component analysis (PCA) with lcWGS data.
@@
 Color represents sequencing depth: yellow points mark individuals that were simulated sequenced at 0.125×; grey points for 4×.
-Figure 6 from https://onlinelibrary.wiley.com/doi/full/10.1111/1755-0998.13559."}
+Figure 6 from @Lou2022-lc."}
 
@@
 
-The following Figure 3 from Lotterhos 2019 (https://academic.oup.com/g3journal/article/9/6/1851/6028019) illustrates what happens in a PCA on landscape genomic data when there is recombination variation.
+The following Figure 3 from Lotterhos 2019 (@Lotterhos2019-yz) illustrates what happens in a PCA on landscape genomic data when there is recombination variation.
 She simulated a 2-D continuous landscape with local adaptation to an environmental cline, with a genome that consisted of linkage groups with quantitative trait loci that contributed to adaptation, a large neutral inversion (not involved in adaptation), and a region of low recombination (also not involved in adaptation).
@@
 
-When SNPs were thinned for LD and then a PCA was run on the data, the PCA showed a pattern of isolation-by-environment, which more accurately captured the population structure in the simulation (Figure B below - https://academic.oup.com/g3journal/article/9/6/1851/6028019).
+When SNPs were thinned for LD and then a PCA was run on the data, the PCA showed a pattern of isolation-by-environment, which more accurately captured the population structure in the simulation (Figure B below @Lotterhos2019-yz).
 
-```{r, echo = FALSE, include = TRUE, out.width = "50%", fig.align = "center", fig.cap="Figure 8. Figure 3 from https://academic.oup.com/g3journal/article/9/6/1851/6028019."}
+```{r, echo = FALSE, include = TRUE, out.width = "50%", fig.align = "center", fig.cap="Figure 8. Figure 3 from @Lotterhos2019-yz."}
 
@@
 Similar patterns have been observed for sex linked-markers.
-For example, https://onlinelibrary.wiley.com/doi/10.1111/mec.14217  found genetic clustering in a PCA by sex instead of by population structure, which was driven by as few as 12 sex-linked markers in the data.
+For example, @Benestan2017-oi found genetic clustering in a PCA by sex instead of by population structure, which was driven by as few as 12 sex-linked markers in the data.
 Removing the sex-linked markers led to nonsignificant genetic structure in one species and a more accurate estimation of FST in the second species.
@@
 Imputation is the process of replacing a missing value with a numerical value via inference.
-Imputation of missing data can be done by sophisticated methods such as in genomic association studies, linked haplotypes, and strong genotype-phenotype correlations (see Yi & Latch, 2021).
+Imputation of missing data can be done by sophisticated methods such as in genomic association studies, linked haplotypes, and strong genotype-phenotype correlations (see @Yi2022-bi).
 However, in population genomic studies for species of conservation concern which may have few genomic resources, simple imputations such as mean imputation (replace missing with column mean) are commonly encountered.
@@
 This can be problematic if you have a lot of missing data in your dataset, especially if that missing data is biased (not random).
-Yi and Latch (2022) demonstrated this effect using different simulated populations (no migration.
+@Yi2022-bi demonstrated this effect using different simulated populations (no migration.
 5% migration, isolation-by-distance-cline) with 1%, 10% and 20% missing data, either randomly distributed among individuals, or biased among some individuals and some populations.
@@
 ```{r, echo = FALSE, include = TRUE, out.width = "80%", fig.align = "center", fig.cap="Figure 9.
-Figure 2(b) from Yi and Latch (2022) shows PCA on the individual-biased missing data (1%, 10% or 20%) introduced to a simulated population experiencing an isolation-by-distance cline, where one population (represented as triangles) is an admixture of the remaining two (circle and squares).
+Figure 2(b) from @Yi2022-bi shows PCA on the individual-biased missing data (1%, 10% or 20%) introduced to a simulated population experiencing an isolation-by-distance cline, where one population (represented as triangles) is an admixture of the remaining two (circle and squares).
 Individual colors represent their amounts of missing data with relatively higher missingness shown in lighter blue in each plot (legend, notice changing scale from left to right).
@@
 This is because this method of imputation would make samples with missing data look more similar to the samples used to generate imputation values.
-This method is not recommended (see Yi & Latch 2021), since within-group mean imputation depends largely on the a priori population designation and can easily bring artificial biases.
+This method is not recommended (see @Yi2022-bi), since within-group mean imputation depends largely on the a priori population designation and can easily bring artificial biases.
 
 ```{r, echo = FALSE, include = TRUE, out.width = "80%", fig.align = "center", fig.cap="Figure 10.
-Figure 4 from Yi and Latch (2022) shows PCA on the cline model with missing data condensed in either the (a) admixed population and (b) one end population.
+Figure 4 from @Yi2022-bi shows PCA on the cline model with missing data condensed in either the (a) admixed population and (b) one end population.
 The admixed population (pop2) has high migration with both end populations (pop1 and pop3) while no migration occurs between the end populations."}
@@
 ### Hybridization and inbreeding effects {#hybrid}
-Genetic differences between two different parental species typically constitute the dominant axis of genetic variation in a PCA, with hybrids mapping in PC space between the parental species (https://onlinelibrary.wiley.com/doi/full/10.1111/eva.12380 ).
+Genetic differences between two different parental species typically constitute the dominant axis of genetic variation in a PCA, with hybrids mapping in PC space between the parental species (@Gompert2016-wa).
 However, PC axes can also pick up on differences in homozygosity caused by many generations of inbreeding.
@@
 
-Below is Figure 5 from https://onlinelibrary.wiley.com/doi/full/10.1111/eva.12380, with two parental species shown in dark red or dark blue, the F1 hybrids in yellow circles, and the F1-parental backcrosses in the lighter colors.
+Below is Figure 5 from @Gompert2016-wa, with two parental species shown in dark red or dark blue, the F1 hybrids in yellow circles, and the F1-parental backcrosses in the lighter colors.
 Subsequent crosses between hybrids are shown in yellow with different symbols (F2 squares, F5 upward triangles, F20 downward triangles).
@@
 
-```{r, echo = FALSE, include = TRUE, out.width = "50%", fig.align = "center", fig.cap="Figure 11. Figure 5 from https://onlinelibrary.wiley.com/doi/full/10.1111/eva.12380."}
+```{r, echo = FALSE, include = TRUE, out.width = "50%", fig.align = "center", fig.cap="Figure 11. Figure 5 from @Gompert2016-wa."}
 
@@
 Horseshoe patterns arise in population genetic data that has isolation by distance structure, by which genetic differentiation among locations increases with the distance between them.
-The result when this type of data is put into a PCA is that the pattern looks like a horseshoe (Figure 1B from Morton et al below).
+The result when this type of data is put into a PCA is that the pattern looks like a horseshoe (Figure 1B from @Morton2017-xy below).
 In Figure 1B, Sample 0 is most distantly related from Sample 20, but they have the same loading on the PC1 axis.
-Horseshoe patterns also arise in microbial ecology, where they were originally misinterpreted as having an ecological explanation before they were proven to be a statistical artifact (https://journals.asm.org/doi/10.1128/mSystems.00166-16).
+Horseshoe patterns also arise in microbial ecology, where they were originally misinterpreted as having an ecological explanation before they were proven to be a statistical artifact (@Morton2017-xy).
 The pattern arises in microbial data or genetic data as a consequence of distance metrics that saturate (Figure 1A and C below), typically when distance metrics as used in PCA cannot discriminate between samples that do not share any common features.
@@
 ```{r, echo = FALSE, include = TRUE, out.width = "70%", fig.align = "center", fig.cap="Figure 12.
-Figure 1 from https://journals.asm.org/doi/10.1128/mSystems.00166-16.
+Figure 1 from @Morton2017-xy.
 (a) A band table for haploids where SNPs are in rows, samples are in columns, and the color is the allele (0 or 1).
@@
 
-The saturation property of PCA can also create “wave” patterns in a PCA of genetic data, which were initially misinterpreted to be signatures of specific migration events before they were shown to be statistical artifacts (Novembre & Stephens).
-Figure 2 from Novembre & Stephens below shows the pattern arising from simulating a 1-D stepping stone model, which is analogous to marine samples that are collected along a coastline.
+The saturation property of PCA can also create “wave” patterns in a PCA of genetic data, which were initially misinterpreted to be signatures of specific migration events before they were shown to be statistical artifacts (@Novembre2008-gs).
+Figure 2 from @Novembre2008-gs below shows the pattern arising from simulating a 1-D stepping stone model, which is analogous to marine samples that are collected along a coastline.
 The horseshoe pattern arises in the first 2PC axes (Figure 2C), but this example also shows the wave patterns that arise on subsequent PC axes.
@@
 ```{r, echo = FALSE, include = TRUE, out.width = "70%", fig.align = "center", fig.cap="Figure 13.
-Figure 2 from Novembre & Stephens.
+Figure 2 from @Novembre2008-gs.
 (a) Schematic of the one-dimensional habitat, with circles marking sampling locations and shades of blue marking order along the line.
@@
 
-In MacVean (2009) Figure 5A below, PCA is carried out only on the haplotypes from CEU and YRI and all samples are subsequently projected onto the first PC identified from this analysis.
+In @McVean2009-ms Figure 5A below, PCA is carried out only on the haplotypes from CEU and YRI and all samples are subsequently projected onto the first PC identified from this analysis.
 Figure 5A shows how the loadings of individuals map onto PC1 each chromosome (in rows), and note the uniformity of the loadings for the source populations.
@@
 
-```{r, echo = FALSE, include = TRUE, out.width = "50%", fig.align = "center", fig.cap="Figure 14.
-Figure 5 from https://journals.plos.org/plosgenetics/article?id=10.1371/journal.pgen.1000686.
-See text above for explanation."}
+```{r, echo = FALSE, include = TRUE, out.width = "50%", fig.align = "center", fig.cap="Figure 14. Figure 5 from @McVean2009-ms. See text above for explanation."}
 
@@
 |pcadapt|Missing data are coded as ‘9’ in the input.
-Applies a covariance approach to missing data, outlined in (Dray & Josse, 2015); this approach deals with missing data by only computing the covariance matrix (also called the genomic relationship matrix (GRM)) between each pair of individuals using the variables (loci) that are available for both individuals (at least until version < 3.x).
-The updated pcadapt (v.4.x)  is very fast, using a truncated SVD algorithm and accounts for missing data using a function which computes the product and cross-product of the (scaled) genotype matrix with a given vector (see Privé et al., 2020 supplementary materials).
+Applies a covariance approach to missing data, outlined in @drayPrincipalComponentAnalysis2015; this approach deals with missing data by only computing the covariance matrix (also called the genomic relationship matrix (GRM)) between each pair of individuals using the variables (loci) that are available for both individuals (at least until version < 3.x).
+The updated pcadapt (v.4.x)  is very fast, using a truncated SVD algorithm and accounts for missing data using a function which computes the product and cross-product of the (scaled) genotype matrix with a given vector (see @privePerformingHighlyEfficient2020 supplementary materials).
 If you have A LOT of missing data, this method can affect the finding of ‘outlier loci’ (significance of loadings).|@privePerformingHighlyEfficient2020; @luuPcadaptPackagePerform2017|

### 2023-11-21 DRAFT_PCA.Rmd
implemented Dani's changes to PCA page, minor edits

@@
 ```
+Authors: KE Lotterhos, DL Davenport, S Truskey <br>
+Acknowledgements/ Page Reviewers: Jason Johns
 
@@
 
-On this page we outline and describe many aspects of PCA:
+On this page we outline and describe many aspects of PCA as it relates to population genomics:
 
-* What is PCA and what are its common uses in population genetics?
+* [What is PCA and what are its common uses in population genetics?](#whatspca)
 * Considerations before sampling:
-    + How one chooses to sample from a landscape
-    + Sample sizes
+    + [How one chooses to sample from a landscape](#sampling)
+    + [Sample sizes](#samplesize)
 * Considerations while processing data (aka 'weird stuff that happens in PCA'): 
-    + Batch effects
-    + LD and inversion effects
-    + Missing data and imputation
-    + Hybridization and inbreeding effects
-    + Horseshoe and wave patterns
+    + [Batch effects](#batcheffects)
+    + [LD and inversion effects](#ld)
+    + [Missing data and imputation](#missingdata)
+    + [Hybridization and inbreeding effects](#hybrid)
+    + [Horseshoe and wave patterns](#horseshoe)
 * Comparison to other common approaches to investigating population genetic structure
-    + PCA  vs ADMIXTURE/STRUCTURE
+    + [PCA  vs ADMIXTURE/STRUCTURE](#pca_admix)
 
@@
 
-## What is PCA and what are its common uses in population genetics?
+## What is PCA and what are its common uses in population genetics?{#whatspca}
 
@@
 The eigenvectors are represented as a standard scatter plot, where points are samples/individuals represented on the new system of axes (PCs).
-Dimension reduction to eigenvectrors (PCs) means that the 1st PC is the linear combination of the original variables that explains the most/greatest amount of variation, while the 2nd PC accounts for the greatest amount of the remaining variation being orthogonal (uncorrelated) to the 1st PC, the 3rd PC accounts for the greatest amount of the remaining variation being orthogonal (uncorrelated) to the 1st and 2nd PC so on and so on.
+Dimension reduction to eigenvectors (PCs) means that the 1st PC is the linear combination of the original variables that explains the most/greatest amount of variation, while the 2nd PC accounts for the greatest amount of the remaining variation being orthogonal (uncorrelated) to the 1st PC, the 3rd PC accounts for the greatest amount of the remaining variation being orthogonal (uncorrelated) to the 1st and 2nd PC so on and so on.
 In population genomics, typically the 1st PC will capture broad scale population structure (because broadly distributed populations are usually the most differentiated) while subsequent axes capture more regional population structure (usually less differentiated at smaller geographic scales, see Figure 2A).
-However, other cool examples using PCA find the greatest amount of variation is between individuals, particularly in the case of inversions, structural variation or regions associated with a specific phenotype or adaption (see Figure 2B).
+However, other cool examples using PCA find the greatest amount of variation is between individuals, particularly in the case of inversions, structural variation or regions associated with a specific phenotype or adaptation (see Figure 2B).
 
@@
 * [Josse J., Husson F. (2012) Handling missing values in exploratory multivariate data analysis methods. J. Soc. Française Stat., 153, 79–99.](http://www.numdam.org/item/JSFS_2012__153_2_79_0/)
+* [François O, Gain C (2021) A spectral theory for Wright’s inbreeding coefficients and related quantities.
+PLoS Genet 17: e1009665](https://journals.plos.org/plosgenetics/article?id=10.1371/journal.pgen.1009665)
+* [Patterson N, Price AL, Reich D (2006) Population structure and eigenanalysis. PLoS Genet 2: e190](https://journals.plos.org/plosgenetics/article?id=10.1371/journal.pgen.0020190)
+* [Bakker JD (2023) PCA. Applied Multivariate Statistics in R. University of Washington](https://uw.pressbooks.pub/appliedmultivariatestatistics/chapter/pca/)
 
@@
 ## Considerations before sampling
-### How one chooses to sample from a landscape
+### How one chooses to sample from a landscape{#sampling}
 How samples are selected from a metapopulation also affects the visual pattern in a PCA.
@@
 
-### Sample sizes
+### Sample sizes{#samplesize}
 Uneven sampling between groups, the number of samples and/or the number of markers/SNPs used in PCA can change the PCA projection space, and thus interpretation of results, relative to the true demographic history of the sampled groups.
@@
 low coverage NGS, microsatellite null alleles, mutation at RAD cut sites).
-In general, if missing data reflects a true population signal (e.g.
-null alleles in one population but not others) then it requires careful consideration during analysis.
+In general, if missing data reflects a true population signal then it requires careful consideration during analysis.
 
@@
 
-Implementations of PCA in genomics typically (a) avoid missing values (i.e pcadapt, plink) usually by first determining the covariance-matrix or genomic relationship matrix (GRM)) between each pair of individuals using the variables (loci) that are available for both samples or by (b) impute missing values.
+Implementations of PCA in genomics typically (a) avoid missing values (i.e pcadapt, plink) usually by first determining the covariance-matrix or genomic relationship matrix (GRM)) between each pair of individuals using the variables (loci) that are available for both samples or by (b) imputing missing values.
 Imputation is the process of replacing a missing value with a numerical value via inference.
-Imputation of missing data can be done by sophisticated methods such as in genomic association studies, linked haplotypes, and strong genotype-phenotype correlations (see @Yi2022-bi).
-However, in population genomic studies for species of conservation concern which may have few genomic resources, simple imputations such as mean imputation (replace missing with column mean) are commonly encountered.
+Imputation of missing data can be done by sophisticated methods often seen in genomic association studies, or if you have linked haplotypes or strong genotype-phenotype correlations (see @Yi2022-bi).
+However, in population genomic studies for species of conservation concern which may have few genomic resources, simple imputations such as mean imputation (replace missing with column mean) is commonly encountered.
 
@@
 In this case, missing values for all individuals at a locus are replaced with the mean for that locus (also called a mean-imputed matrix).
-Importantly, when the mean-imputed matrix is centered and scaled, the original missing become non-informative in PCA (missing values are placed at the origin).
+Importantly, when the mean-imputed matrix is centered and scaled, the original missing genotypes become non-informative in PCA (missing values are placed at the origin).
 This can be problematic if you have a lot of missing data in your dataset, especially if that missing data is biased (not random).
@@
 
-**Effects of different approaches to imputation on PCA.**
+In summary, the choice of imputation can have drastic effects on inference if there is a lot of missing data.
+When there is population structure and missing genotypes are imputed with the mean genotype across all populations, the genetic differentiation among populations may be artificially reduced and the metapopulation would appear to be panmictic.
+When there is no population structure and missing genotypes are imputed with the mean genotype within each sampling location, the genetic differentiation among populations may be artificially inflated and it could appear that there is genetic structure when there is not.
 
-The choice of imputation can have drastic effects on inference if there is a lot of missing data.
-When there is population structure and missing genotypes are imputed with the mean genotype across all populations, the genetic differentiation among populations may be artificially reduced and the metapopulation would appear to be panmictic.
-When there is no population structure and missing genotypes are imputed with the mean genotype within each sampling location, the genetic differentiation among populations may be artificially inflated and it could appear that their genetic structure when there is in fact none.
-As such, it is really important to understand the type of imputation a PCA method might be implementing  and its effect  on population clustering and final inferences.
+Since most genomic data has missingness, it may be preferable to implement a method of PCA which explicitly accounts for it (i.e pcadapt, PCAngsd).
 
-Since most genomic data has missingness, it may be preferable to implement a method of PCA which explicitly accepts it (i.e pcadapt, PCAngsd).
+Table 1.
+Common PCA methods implemented in population genomics studies, a description of how they work, and how they treat missing data.
+Abbreviations include Singular Value Decomposition (SVD), Single Nucleotide Polymorphism (SNP), genomic relationship matrix (GRM), minor allele frequency (MAF)
+
+| Method or Function, command call | Treatment of Missingness | Ref |
+|:----------|:------------------------------------------------|:----|
+|pcadapt|Missing data are coded as ‘9’ in the input.
+Applies a covariance approach to missing data, outlined in @drayPrincipalComponentAnalysis2015; this approach deals with missing data by only computing the covariance matrix (also called the genomic relationship matrix (GRM)) between each pair of individuals using the variables (loci) that are available for both individuals (at least until version < 3.x).
+The updated pcadapt (v.4.x)  is very fast, using a truncated SVD algorithm and accounts for missing data using a function which computes the product and cross-product of the (scaled) genotype matrix with a given vector (see @privePerformingHighlyEfficient2020 supplementary materials).
+If you have A LOT of missing data, this method can affect the finding of ‘outlier loci’ (significance of loadings).|@privePerformingHighlyEfficient2020; @luuPcadaptPackagePerform2017|
+|NIPALS (Nonlinear Iterative Partial Least Squares)|Missing values are initially set to the row averages, and SVD of the SNP matrix is used to create orthogonal PCs.
+The PCs which correspond to the largest eigenvalues are then used to reconstruct the missing SNP genotypes in the SNP matrix (REF).
+|@wold1975path|
+|adegenet::glPCA|Missing values are replaced by the dosage (a factor homozygous ref = 0, heterozygous = 1, homozygous alternate = 2) mean of the column representing the variant |@jombart2008adegenet|
+|SNPRelate::SNPRelatePCA|Missing values are imputed by the dosage mean|@zhengHighperformanceComputingToolset2012a|
+|PCAngsd/EMU|Missing data is imputed based on population structure inferred from the posterior genotype dosages from the top K inferred PCs.
+The method to choose the top PCs to represent population structure and  estimate the individual allele frequencies follows Velicier’s minimum average partial (MAP) (@shrinerInvestigatingPopulationStratification2011).|@meisnerLargescaleInferencePopulation2021; @meisnerInferringPopulationStructure2018|
+|dudi.pca()|Does not work with missing genotypes. Missingness can manually be replaced or removed|@dray2007ade4|
+|BaseR, prcomp()|Does not work with missing genotypes. Missingness can manually be replaced or removed|@rcoreteamLanguageEnvironmentStatistical2013|
+|vegan::RDA()|This function performs a PCA when no conditions are added.
+Does not work with missing genotypes.
+Missingness can manually be replaced, usually by the dosage mean.|@oksanenVeganCommunityEcology2019|
+|Plink PCA|Can implement different computations: The standard computation computes the covariance matrix between pairs of individuals, using only variables (loci) that are available for both individuals (but you can add the modifier “meanimpute” for mean imputation).
+The randomisation algorithm (aka.
+FASTPCA) (--pca approx) performs mean-imputation of missing genotypes (see @galinskyFastPrincipalComponentAnalysis2016a)|
+ |@purcell2007plink|
+|bigSNPR:, big_SVD() and big_randomSVD()|Does not work with missing data, but offers various  functions to impute missingness based on plink, beagle or methods outlined in @wang2012fast|@prive2018efficient; @Prive2020-fp|
 
@@
 
-### Horseshoe and Wave patterns
+### Horseshoe and Wave patterns{#horseshoe}
 Horseshoe patterns arise in population genetic data that has isolation by distance structure, by which genetic differentiation among locations increases with the distance between them.
@@
 
-## Relationship between PCA and Admixture
+## Relationship between PCA and Admixture{#pca_admix}
 
@@
 
-Table 1.
-Common PCA methods implemented in population genomics studies, a description of how they work, and how they treat missing data.
-Abbreviations include Singular Value Decomposition (SVD), Single Nucleotide Polymorphism (SNP), genomic relationship matrix (GRM), minor allele frequency (MAF)
-
-| Method or Function, command call | Treatment of Missingness | Ref |
-|:----------|:------------------------------------------------|:----|
-|pcadapt|Missing data are coded as ‘9’ in the input.
-Applies a covariance approach to missing data, outlined in @drayPrincipalComponentAnalysis2015; this approach deals with missing data by only computing the covariance matrix (also called the genomic relationship matrix (GRM)) between each pair of individuals using the variables (loci) that are available for both individuals (at least until version < 3.x).
-The updated pcadapt (v.4.x)  is very fast, using a truncated SVD algorithm and accounts for missing data using a function which computes the product and cross-product of the (scaled) genotype matrix with a given vector (see @privePerformingHighlyEfficient2020 supplementary materials).
-If you have A LOT of missing data, this method can affect the finding of ‘outlier loci’ (significance of loadings).|@privePerformingHighlyEfficient2020; @luuPcadaptPackagePerform2017|
-|NIPALS (Nonlinear Iterative Partial Least Squares)|Missing values are initially set to the row averages, and SVD of the SNP matrix is used to create orthogonal PCs.
-The PCs which correspond to the largest eigenvalues are then used to reconstruct the missing SNP genotypes in the SNP matrix (REF).
-|@wold1975path|
-|adegenet::glPCA|Missing values are replaced by the dosage (a factor homozygous ref = 0, heterozygous = 1, homozygous alternate = 2) mean of the column representing the variant |@jombart2008adegenet|
-|SNPRelate::SNPRelatePCA|Missing values are imputed by the dosage mean|@zhengHighperformanceComputingToolset2012a|
-|PCAngsd/EMU|Missing data is imputed based on population structure inferred from the posterior genotype dosages from the top K inferred PCs.
-The method to choose the top PCs to represent population structure and  estimate the individual allele frequencies follows Velicier’s minimum average partial (MAP) (@shrinerInvestigatingPopulationStratification2011).|@meisnerLargescaleInferencePopulation2021; @meisnerInferringPopulationStructure2018|
-|dudi.pca()|Does not work with missing genotypes. Missingness can manually be replaced or removed|@dray2007ade4|
-|BaseR, prcomp()|Does not work with missing genotypes. Missingness can manually be replaced or removed|@rcoreteamLanguageEnvironmentStatistical2013|
-|vegan::RDA()|This function performs a PCA when no conditions are added.
-Does not work with missing genotypes.
-Missingness can manually be replaced, usually by the dosage mean.|@oksanenVeganCommunityEcology2019|
-|Plink PCA|Can implement different computations: The standard computation computes the covariance matrix between pairs of individuals, using only variables (loci) that are available for both individuals (but you can add the modifier “meanimpute” for mean imputation).
-The randomisation algorithm (aka.
-FASTPCA) (--pca approx) performs mean-imputation of missing genotypes (see @galinskyFastPrincipalComponentAnalysis2016a)|
- |@purcell2007plink|
-|bigSNPR:, big_SVD() and big_randomSVD()|Does not work with missing data, but offers various  functions to impute missingness based on plink, beagle or methods outlined in @wang2012fast|@prive2018efficient; @Prive2020-fp|
-
 ## References {#refs}

### 2023-11-28 DRAFT_PCA.Rmd
Formatting updates to PCA draft, including equation setting

@@
 Authors: KE Lotterhos, DL Davenport, S Truskey <br>
-Acknowledgements/ Page Reviewers: Jason Johns
+Acknowledgements/ Page Reviewers: Jason Johns, Katherine Silliman
 
@@
 * [What is PCA and what are its common uses in population genetics?](#whatspca)
+    + [Recommended reading](#recreading)
 * Considerations before sampling:
@@
 
-Because we cannot visualise 500,000 dimensions, PCA helps us distill these down to increase the interpretability of data while preserving the maximum amount of information (aka variance) .
+Because we cannot visualise 500,000 dimensions, PCA helps us distill these down to increase the interpretability of data while preserving the maximum amount of information (aka **variance**) .
 It does this by creating new, uncorrelated variables that maximise variance called “principal components” (PCs, see @Pearson1901-xe).
@@
 
-The output of PCA are the eigenvalues and eigenvectors (eigenvalues = principal components (PCs), eigenvectors = principal axes) (see [here](https://www.youtube.com/watch?v=PFDu9oVAE-g) for video explanation) which can be represented as graphical outputs that summarise the information of a large number of variables by a smaller number of dimensions (PCs).
-When applied to very large genetic datasets, the eigenvectors and eigenvalues are usually determined by a method called singular value decomposition (SVD) or some variation of it, a matrix factorisation method recommended when the number of variables is very large  (i.e.
+The output of PCA are the eigenvalues and eigenvectors (**eigenvalues** = principal components (PCs), **eigenvectors** = principal axes) (see [here](https://www.youtube.com/watch?v=PFDu9oVAE-g) for video explanation) which can be represented as graphical outputs that summarise the information of a large number of variables by a smaller number of dimensions (PCs).
+When applied to very large genetic datasets, the eigenvectors and eigenvalues are usually determined by a method called **singular value decomposition (SVD)** or some variation of it, a matrix factorisation method recommended when the number of variables is very large  (i.e.
 often more variables than observed entities, n << m, see this [YouTube playlist](https://www.youtube.com/playlist?list=PLMrJAkhIeNNSVjnsviglFoY2nXildDCcv) for details).
 Other methods use the covariance (or correlation) matrix.
-The data are usually centered (so it becomes zero mean gaussian, center = TRUE ), and scaled ( scale = TRUE ), where scaling is recommended when the variance is not homogeneous across variables (@Jolliffe2005-hw).
+The data are usually **centered** (so it becomes zero mean gaussian, center = TRUE ), and **scaled** ( scale = TRUE ), where scaling is recommended when the variance is not homogeneous across variables (@Jolliffe2005-hw).
 
-The eigenvalues represent the amount of variance explained per component and are typically represented as a barplot, in descending order (Figure 1) (also called variable loadings).
+The eigenvalues represent the amount of variance explained per component and are typically represented as a barplot, in descending order (Figure 1) (also called **variable loadings**).
 The percentage of explained variance by each eigenvector will vary between datasets since it is highly dependent on the size of the data and the correlation structure between variables.
@@
 
+### Recommended reading{#recreading} 
 To get a better understanding of PCA, we recommend reading these useful reviews and studies of PCA:
@@
 FST) among populations decreases, the number of SNPs needed to separate the signal from the noise increases.
-The paper makes useful recommendations on how many SNPs would be needed to resolve population structure in PCA space for a given FST for a two population model (see @McVean2009-ms):
+The paper makes useful recommendations on how many SNPs would be needed to resolve population structure in PCA space for a given FST for a two population model (see @McVean2009-ms).
+Note that number of samples below represents haploid samples, so double the number for diploid samples:
 
-FST > 1/sqrt(number_of_SNPs * number_of_samples_per_group) 		Eq1
+**Eq. 1**  
+$$FST > \frac{1}{\sqrt{(number\_of\_SNPs * number\_of\_samples\_per\_group)}}$$ 	
 Rearranging, if we know FST and the number of samples per group, we can calculate how many SNPs we would need to sequence for our PCA to be above the signal-to-noise ratio; or if we know FST and the number of SNPs, we can calculate how many samples per group we would need to sequence for our PCA to be above the signal-to-noise ratio:
 
-Number_of_samples_per_group <- 1/(number_of_SNPs * FST^2)		Eq2
+**Eq. 2**
+$$number\_of\_samples\_per\_group > \frac{1}{(number\_of\_SNPs * FST^2)}$$
+**Eq. 3**
+$$number\_of\_snps> \frac{1}{(number\_of\_samples\_per\_group * FST^2)}$$
 
-Number_of_SNPs <- 1/(Number_of_samples_per_group * FST^2)		Eq3
-
-For instance, if FST = 0.01 and there is 100 samples per group, then you would need 1/(100 * 0.01^2) = 100 SNPs .
+For instance, if FST = 0.01 and there is 100 samples per group, then you would need at least 1/(100 * 0.01^2) = 100 SNPs .
 If FST = 0.001 and there is 20 samples per group (not out of the question for a marine species), then you would need 1/(20 * 0.001^2) = 50,000 SNPs.
@@
 ```{r, echo = FALSE, include = TRUE, out.width = "60%", fig.align = "center", fig.cap="Figure 4.
-Figure 2 from MacVean - Simulated a population (Fig 2A), where the expected locations for the two populations on the first PC are defined by the time-since-divergence (the lines in Fig 2C).
+Figure 2 from MacVean (2009) - Simulated a population (Fig 2A), where the expected locations for the two populations on the first PC are defined by the time-since-divergence (the lines in Fig 2C).
 The signal-to-noise threshold for a sample size of 100 and FST=0.01 is 100 SNPs (see Eq3 above).
@@
 
-MacVean (2009) used simulations of two populations of equal size to show how the number of individuals sampled per group influenced PC projections (Figure 2a, MacVean, 2009).
-Figure 2B (MacVean, 2009) shows how groups with larger sample sizes will lie closer to the origin of the first PC axis, while Figure 2c shows the effect of using fewer SNPs on the inferring the true population configuration, where lines indicate the expectation of population scores on the PC-axis.
+MacVean (2009) used simulations of two populations of equal size to show how the number of individuals sampled per group influenced PC projections (Figure 4A).
+Figure 4B shows how groups with larger sample sizes will lie closer to the origin of the first PC axis, while Figure 4C shows the effect of using fewer SNPs on the inferring the true population configuration, where lines indicate the expectation of population scores on the PC-axis.
 
 The author extended this analysis to a 9-population stepping stone lattice.
-Figure 3 from the paper shows how differences in the number of samples per population can warp the projection space of PCA, even when migration rates and effective population size in each deme are the same.
+Figure 3 from the paper shows how differences in the number of samples per population can warp the projection space of PCA, even when migration rates and effective population size in each deme are the same (Figure 5).
 Note that FST among demes would be similar in all panels because it is based on an allele frequency in each deme (although small differences would occur due to sampling error when fewer individuals within a deme are sampled).
@@
 
-Imputation via the mean genotype in the metapopulation.
+**Imputation via the mean genotype in the metapopulation.**
+
 In this case, missing values for all individuals at a locus are replaced with the mean for that locus (also called a mean-imputed matrix).
@@
 They conducted PCA using the adegenet::glPca() implementation of PCA with default mean imputation and standardization, retaining all PCA axes.
-They found that individuals biased with missing data would be dragged away from their real population clusters to the origin of PCA plots, making them indistinguishable from known admixed individuals, potentially leading to misinterpreted population structure (Figure 11).
+They found that individuals biased with missing data would be dragged away from their real population clusters to the origin of PCA plots, making them indistinguishable from known admixed individuals, potentially leading to misinterpreted population structure (Figure 9).
 
@@
 
-Imputation via the mean genotype within a site, sampling location, or population. 
+**Imputation via the mean genotype within a site, sampling location, or population.**
+
 In this case, missing values for each individual sample at each locus is replaced with the mean genotype of other individuals from the same sampling location or population.
-The effect of this imputation in a PCA is that the mapping for that imputed sample will move closer to the center of the population cluster, and will make the PCA appear to be more clustered (Figure 12).
+The effect of this imputation in a PCA is that the mapping for that imputed sample will move closer to the center of the population cluster, and will make the PCA appear to be more clustered (Figure 10).
 This is because this method of imputation would make samples with missing data look more similar to the samples used to generate imputation values.
@@
 The randomisation algorithm (aka.
-FASTPCA) (--pca approx) performs mean-imputation of missing genotypes (see @galinskyFastPrincipalComponentAnalysis2016a)|
- |@purcell2007plink|
+FASTPCA) (--pca approx) performs mean-imputation of missing genotypes (see @galinskyFastPrincipalComponentAnalysis2016a)|@purcell2007plink|
 |bigSNPR:, big_SVD() and big_randomSVD()|Does not work with missing data, but offers various  functions to impute missingness based on plink, beagle or methods outlined in @wang2012fast|@prive2018efficient; @Prive2020-fp|

### 2023-12-05 POP_08_PCA.Rmd
publishing PCA page with all other pages re-knitted so links to PCA page work

--- 2023-11-28 DRAFT_PCA.Rmd
@@
 title: "PCA"
-bibliography: DRAFT_PCA/PCA_marineomics.bib
+bibliography: POP_08_PCA_files/PCA_marineomics.bib
 authors: "DL Davenport, KE Lotterhos"
-date: "2023-11-07"
+date: "2023-12-05"
 output:
@@
 
-knitr::include_graphics("DRAFT_PCA/Fig.1_Spies_2022_supfig2.png")
+knitr::include_graphics("POP_08_PCA_files/Fig.1_Spies_2022_supfig2.png")
 
@@
 
-knitr::include_graphics("DRAFT_PCA/Fig.2_PCAs.png")
+knitr::include_graphics("POP_08_PCA_files/Fig.2_PCAs.png")
 
@@
 
-knitr::include_graphics("DRAFT_PCA/Fig.3_Gompert_Buerkle_fig4.png")
+knitr::include_graphics("POP_08_PCA_files/Fig.3_Gompert_Buerkle_fig4.png")
 
@@
 
-knitr::include_graphics("DRAFT_PCA/Fig.4_McVean_fig2.png")
+knitr::include_graphics("POP_08_PCA_files/Fig.4_McVean_fig2.png")
 
@@
 
-knitr::include_graphics("DRAFT_PCA/Fig.5_McVean_fig3.png")
+knitr::include_graphics("POP_08_PCA_files/Fig.5_McVean_fig3.png")
 
@@
 
-knitr::include_graphics("DRAFT_PCA/Fig.6_Lou_Thilk_fig4.png")
+knitr::include_graphics("POP_08_PCA_files/Fig.6_Lou_Thilk_fig4.png")
 
@@
 
-knitr::include_graphics("DRAFT_PCA/Fig.7_Lou_Thilk_fig6.png")
+knitr::include_graphics("POP_08_PCA_files/Fig.7_Lou_Thilk_fig6.png")
 
@@
 
-knitr::include_graphics("DRAFT_PCA/Fig.8_Lotterhos2019_fig3.png")
+knitr::include_graphics("POP_08_PCA_files/Fig.8_Lotterhos2019_fig3.png")
 
@@
 
-knitr::include_graphics("DRAFT_PCA/Fig.9_Yi_Latch_fig2.png")
+knitr::include_graphics("POP_08_PCA_files/Fig.9_Yi_Latch_fig2.png")
 
@@
 
-knitr::include_graphics("DRAFT_PCA/Fig.10_Yi_Latch_fig4.png")
+knitr::include_graphics("POP_08_PCA_files/Fig.10_Yi_Latch_fig4.png")
 
@@
 
-knitr::include_graphics("DRAFT_PCA/Fig.11_Gompert_Buerkle_fig5.png")
+knitr::include_graphics("POP_08_PCA_files/Fig.11_Gompert_Buerkle_fig5.png")
 
@@
 
-knitr::include_graphics("DRAFT_PCA/Fig.12_Morton_fig1.png")
+knitr::include_graphics("POP_08_PCA_files/Fig.12_Morton_fig1.png")
 
@@
 
-knitr::include_graphics("DRAFT_PCA/Fig.13_Novembre_Stephens_fig2.png")
+knitr::include_graphics("POP_08_PCA_files/Fig.13_Novembre_Stephens_fig2.png")
 
@@
 
-knitr::include_graphics("DRAFT_PCA/Fig.14_McVean_fig5.png")
+knitr::include_graphics("POP_08_PCA_files/Fig.14_McVean_fig5.png")
 

### 2023-12-05 index.Rmd
publishing PCA page with all other pages re-knitted so links to PCA page work

@@
 * [Redundancy Analysis (RDA) Trait Prediction](POP_05_RDAtraitPredictionTutorial.html)
+* [PCA](POP_08_PCA.html)
 

### 2023-12-13 DRAFT_DNA_methylation.Rmd
Update DRAFT_DNA_methylation.Rmd

@@
 
-The purpose of this tutorial is to provide an example of identifying loci- and gene-level methylation in bisulfite-treated DNA sequence data.
+Thanks to advances in sample preparation and sequencing methods, there is an influx of research examining epigenomics in non-model systems.
+Broadly, epigenetics can be defined as changes to gene expression that do not arise from changes in the DNA sequence.
+DNA methylation, or addition of a methyl (CH<sub>3</sub>) group to a cytosine base adjacent to a guanine (CpG) is one of the more commonly studies epigenetic mechanisms, partly because of its environmental sensitivity and potential role in phenotypic plasticity.
+For a good review of epigenetic mechanisms including DNA methylation in marine organisms see @doi:10.1146/annurev-marine-010318-095114.
+
+The purpose of this tutorial is to provide an example of aligning bisulfite-treated and enzymatically converted DNA sequence data and discuss needs for potential downstream applications.
 The principle behind creating bisulfite-treated DNA libraries is that when cytosines are not methylated, then they will be converted to uracil, which ultimately result in thymine nucleotides in the sequence data.
-When aligning to a reference genome the percent methylation at a given CpG loci is determined by examining the ratio of thymines to cytosines.
+@https://doi.org/10.1111/1755-0998.13542 offers a comparison of three methods for quantifying DNA methylation at single base-pair resolution using whole genome bisulfite sequencing (WGBS), reduced representation bisulfite sequencing (RRBS), and methyl-CpG binding domain bisulfite sequencing (MBDBS).
+When aligning bisulfite-converted data to a reference genome the percent methylation at a given CpG loci is determined by examining the ratio of thymines to cytosines.
 For example if there is a given cytosine adjacent to a guanine (ie CpG locus) with 30% of the reads containing thymine, one would consider this cytosine loci (CpG) to be 70% methylated.
+Since bisulfite treatment can damage DNA, newer library preparation and sequencing methods (ex.
+[NEBNext® Enzymatic Methyl-seq Kit; EM-seq](https://www.neb.com/en-us/products/e7120-nebnext-enzymatic-methyl-seq-kit#Protocols,%20Manuals%20&%20Usage) are starting to gain popularity.
+As EM-Seq produces loci-level data, alignment of EM-Seq data is similar to WGBS.
 
-For a good review of epigenetic mechanisms including DNA methylation in marine organisms see @doi:10.1146/annurev-marine-010318-095114.
-@https://doi.org/10.1111/1755-0998.13542 offers a comparison of three methods for quantifying DNA methylation at single base-pair resolution using whole genome bisulfite sequencing (WGBS), reduced representation bisulfite sequencing (RRBS), and methyl-CpG binding domain bisulfite sequencing (MBDBS).
-The tutorial below is based on WGBS, though the general workflow would be consistent.
+The tutorial will cover the following:
+
+- Sequence quality assessment
+- Read alignment
+- Methylation quantification
+- File conversions
+- Potential downstream applications. Although this tutorial will not provide code for downstream applications, it will link to other open-access resources and examples.
+
+The tutorial below is based on WGBS from a species of marine intertebrate (*Montipora capitata* coral) examined in @https://doi.org/10.1111/1755-0998.13542, and all raw data can be accessed under NCBI Bioproject PRJNA691891 if you would like to follow the tutorial with this dataset.
+The general workflow would be consistent for other library preparation and sequencing methods that produce loci-level data.
+The tutorial indicates places where special analytical consideration should be given to other data types.
 
@@
 
+# Potential downstream applications
+
+- Gene-level methylation analysis: people do this, but worth considering the utility of converting loci-level resolution data to gene-level resolution data
+- methylKit
+- IGV
+- Analysis with other datasets (ex. pop gen, RNA-Seq)
+
 # References

### 2023-12-13 DRAFT_DNA_methylation.Rmd
adding expectations to intro
Update DRAFT_DNA_methylation.Rmd

@@
 - File conversions
-- Potential downstream applications. Although this tutorial will not provide code for downstream applications, it will link to other open-access resources and examples.
+- Potential downstream applications.
+Although this tutorial will not provide code for downstream applications such as differential methylation analysis, it will link to other open-access resources and examples.
 
@@
 An example of what raw data might look like from a marine invertebrate is shown.
-*Note that DNA methylation patters are diverse across taxa with most marine invertebrates demonstrating a mosiac pattern and global methylation at CpG loci is approximately 15%.*
+*Note that DNA methylation patters are diverse across taxa with most invertebrates demonstrating a mosiac pattern and global methylation at CpG loci is approximately 15%.*
 
-Following trimming one would expect to see something similar to this. This particular data set was trimmed with the following parameters:
+Following trimming one would expect to see FastQC per base sequencing content similar to the plot shown below. This particular data set was trimmed with the following parameters:
 
@@
 
-The software used as part of this tutorial is **Bismark Bisulfite Mapper**.
+The software used as part of this tutorial is **Bismark Bisulfite Mapper** (@http://dx.doi.org/10.1093/bioinformatics/btr167).
 As with any software it is best to be famililar with the [User Manual](https://rawgit.com/FelixKrueger/Bismark/master/Docs/Bismark_User_Guide.html).
-Thus content below will provided with the assumption that the reader has read the manual and is meant to serve as guidance based on experience working with marine invertebrates.
+The content below will provided with the assumption that the reader has read the manual and is meant to serve as guidance based on experience working with marine invertebrates.
 

### 2023-12-13 DRAFT_DNA_methylation.Rmd
Update DRAFT_DNA_methylation.Rmd
showing kathryn how pull requests work

--- 2023-10-12 DRAFT_DNA_methylation.Rmd
@@
 
+- Sequence quality assessment
+- Read alignment
+- Methylation quantification
+- File conversions
+- Potential downstream applications.
+Although this tutorial will not provide code for downstream applications such as differential methylation analysis, it will link to other open-access resources and examples.
+
+## What to expect
+DNA methylation at CpG locus can vary to 0-100%. Like wise methyation at the level Generally speaking in
+
 # Sequence Quality

### 2023-12-13 DRAFT_DNA_methylation.Rmd
added methylation figure

@@
 ## What to expect
-DNA methylation at CpG locus can vary to 0-100%. Like wise methyation at the level Generally speaking in
+DNA methylation at CpG locus can vary to 0-100%.
+Like wise methylation at the gene level can vary and is commonly established as mean methylation level of CpG across gene, assuming some majority degree of gene coverage.
+Similarly with considering the methylation level of a given loci often time some threshold might be set as seen in code below.
+From an taxa perspective it can vary WIDELY on what to expect.
+The figure below shows
+
+```{r meth-bp, echo = FALSE, out.width = "70%", fig.align = "center"}
+knitr::include_graphics("DRAFT_DNA_methylation/meth_ratio.png")
+```
 

### 2023-12-13 DRAFT_DNA_methylation.Rmd
meth what to expect
add YV methylation text back in

--- 2023-12-13 DRAFT_DNA_methylation.Rmd
@@
 
-The purpose of this tutorial is to provide an example of aligning bisulfite-treated and enzymatically converted DNA sequence data and discuss needs for potential downstream applications.
+The purpose of this tutorial is to provide an example of identifying loci- and gene-level methylation in bisulfite-treated DNA sequence data.
+The principle behind creating bisulfite-treated DNA libraries is that when cytosines are not methylated, then they will be converted to uracil, which ultimately result in thymine nucleotides in the sequence data.
+When aligning to a reference genome the percent methylation at a given CpG loci is determined by examining the ratio of thymines to cytosines.
+For example if there is a given cytosine adjacent to a guanine (ie CpG locus) with 30% of the reads containing thymine, one would consider this cytosine loci (CpG) to be 70% methylated.
+
+e purpose of this tutorial is to provide an example of aligning bisulfite-treated and enzymatically converted DNA sequence data and discuss needs for potential downstream applications.
 The principle behind creating bisulfite-treated DNA libraries is that when cytosines are not methylated, then they will be converted to uracil, which ultimately result in thymine nucleotides in the sequence data.
@@
 
+For a good review of epigenetic mechanisms including DNA methylation in marine organisms see @doi:10.1146/annurev-marine-010318-095114.
+@https://doi.org/10.1111/1755-0998.13542 offers a comparison of three methods for quantifying DNA methylation at single base-pair resolution using whole genome bisulfite sequencing (WGBS), reduced representation bisulfite sequencing (RRBS), and methyl-CpG binding domain bisulfite sequencing (MBDBS).
+The tutorial below is based on WGBS, though the general workflow would be consistent.
+
 The tutorial will cover the following:
-
 - Sequence quality assessment
@@
 The tutorial indicates places where special analytical consideration should be given to other data types.
+
+## What to expect
+DNA methylation at CpG locus can vary to 0-100%.
+Like wise methylation at the gene level can vary and is commonly established as mean methylation level of CpG across gene, assuming some majority degree of gene coverage.
+Similarly with considering the methylation level of a given loci often time some threshold might be set as seen in code below.
+From an taxa perspective it can vary WIDELY on what to expect.
+The figure below shows the mehthylation ratio of CpGs across the oyster genome.
+In some marine invertebrates we see ~15% methylation of CpGs (defined as > 75% methylated) across the genome with most CpGs unmethylated.
+In these taxa methylation usually occurs in gene bodes.
+
+```{r meth-bp, echo = FALSE, out.width = "70%", fig.align = "center"}
+knitr::include_graphics("DRAFT_DNA_methylation/meth_ratio.png")
+```
 

### 2023-12-14 DRAFT_DNA_methylation.Rmd
Update DRAFT_DNA_methylation.Rmd
meth - removed duplicate Intro section

@@
 
-The purpose of this tutorial is to provide an example of identifying loci- and gene-level methylation in bisulfite-treated DNA sequence data.
-The principle behind creating bisulfite-treated DNA libraries is that when cytosines are not methylated, then they will be converted to uracil, which ultimately result in thymine nucleotides in the sequence data.
-When aligning to a reference genome the percent methylation at a given CpG loci is determined by examining the ratio of thymines to cytosines.
-For example if there is a given cytosine adjacent to a guanine (ie CpG locus) with 30% of the reads containing thymine, one would consider this cytosine loci (CpG) to be 70% methylated.
-
-e purpose of this tutorial is to provide an example of aligning bisulfite-treated and enzymatically converted DNA sequence data and discuss needs for potential downstream applications.
+The purpose of this tutorial is to provide an example of aligning bisulfite-treated and enzymatically converted DNA sequence data and discuss needs for potential downstream applications.
 The principle behind creating bisulfite-treated DNA libraries is that when cytosines are not methylated, then they will be converted to uracil, which ultimately result in thymine nucleotides in the sequence data.
@@
 ## What to expect
+
 DNA methylation at CpG locus can vary to 0-100%.

### 2023-12-14 DRAFT_DNA_methylation.Rmd
added figure legend - meth ratio

@@
 
-```{r meth-bp, echo = FALSE, out.width = "70%", fig.align = "center"}
+```{r meth-ratio, echo = FALSE, out.width = "70%", fig.align = "center"}
 knitr::include_graphics("DRAFT_DNA_methylation/meth_ratio.png")
 ```
+_Frequency distribution of methylation ratios for CpG dinucleotides in oyster gill tissue._ DOI: [10.7717/peerj.215/fig-1](https://doi.org/10.7717/peerj.215/fig-1)
 

### 2023-12-14 DRAFT_DNA_methylation.Rmd
moved text to alignment - added code explanation

--- 2023-12-14 DRAFT_DNA_methylation.Rmd
@@
 ---
-Initial publication year: TBD <br>
-[How to cite](https://marineomics.github.io/#How_to_Cite)
+
+Initial publication year: TBD <br> [How to cite](https://marineomics.github.io/#How_to_Cite)
 
@@
 
-The tutorial will cover the following:
-- Sequence quality assessment
-- Read alignment
-- Methylation quantification
-- File conversions
+The tutorial will cover the following: - Sequence quality assessment\
+- Read alignment\
+- Methylation quantification\
+- File conversions\
 - Potential downstream applications.
@@
 
-## What to expect
-
-DNA methylation at CpG locus can vary to 0-100%.
-Like wise methylation at the gene level can vary and is commonly established as mean methylation level of CpG across gene, assuming some majority degree of gene coverage.
-Similarly with considering the methylation level of a given loci often time some threshold might be set as seen in code below.
-From an taxa perspective it can vary WIDELY on what to expect.
-The figure below shows the mehthylation ratio of CpGs across the oyster genome.
-In some marine invertebrates we see ~15% methylation of CpGs (defined as > 75% methylated) across the genome with most CpGs unmethylated.
-In these taxa methylation usually occurs in gene bodes.
-
-```{r meth-bp, echo = FALSE, out.width = "70%", fig.align = "center"}
-knitr::include_graphics("DRAFT_DNA_methylation/meth_ratio.png")
-```
+*Frequency distribution of methylation ratios for CpG dinucleotides in oyster gill tissue.* DOI: [10.7717/peerj.215/fig-1](https://doi.org/10.7717/peerj.215/fig-1)
 
@@
 
+```         
 /gscratch/srlab/programs/TrimGalore-0.4.5/trim_galore \
@@
 /gscratch/scrubbed/sr320/froger-raw/00_fastq/Meth17_R1_001.fastq.gz \
+```
 
@@
 
+## What to expect
+
+DNA methylation at CpG locus can vary to 0-100%.
+Like wise methylation at the gene level can vary and is commonly established as mean methylation level of CpG across gene, assuming some majority degree of gene coverage.
+Similarly with considering the methylation level of a given loci often time some threshold might be set as seen in code below.
+From an taxa perspective it can vary WIDELY on what to expect.
+The figure below shows the mehthylation ratio of CpGs across the oyster genome.
+In some marine invertebrates we see \~15% methylation of CpGs (defined as \> 75% methylated) across the genome with most CpGs unmethylated.
+In these taxa methylation usually occurs in gene bodies.
+
+```{r meth-ratio, echo = FALSE, out.width = "70%", fig.align = "center"}
+knitr::include_graphics("DRAFT_DNA_methylation/meth_ratio.png")
+```
+
 The software used as part of this tutorial is **Bismark Bisulfite Mapper** (@http://dx.doi.org/10.1093/bioinformatics/btr167).
@@
 
+```         
 # Directories and programs
@@
 ${genome_folder}
+```
 
-You should expect to prepared genome with new directories including
+1.  **Bismark Genome Preparation Command**:
 
+    -   **`${bismark_dir}/bismark_genome_preparation`**: This is the command to run the genome preparation part of Bismark.
+The **`${bismark_dir}`** variable is expanded to the path where Bismark is installed, so the script knows where to find the **`bismark_genome_preparation`** program.
+
+    -   **`--verbose`**: This flag makes the program output more detailed information about what it is doing, which is helpful for debugging or understanding the process.
+
+    -   **`--parallel 28`**: This option tells Bismark to use 28 threads in parallel to speed up the process.
+
+    -   **`--path_to_aligner ${bowtie2_dir}`**: This specifies the path to the aligner (Bowtie2) that Bismark will use.
+The **`${bowtie2_dir}`** variable is expanded to the path where Bowtie2 is installed.
+
+    -   **`${genome_folder}`**: Finally, this specifies the location of the genome files. The script uses the **`${genome_folder}`** variable, which holds the path to these files.
+
+This will result in a prepared genome with associated new directories including
+
+```         
 ./Bisulfite_Genome
@@
 ./Bisulfite_Genome/CT_conversion
+```
 
@@
 
+```         
 find ${reads_dir}*_R1_001_val_1.fq.gz \
@@
 -o Mcap_tg
+```
 
@@
 
+```         
 find *.bam | \
@@
 {}.bam
+```
 
@@
 
+```         
 ${bismark_dir}/bismark_methylation_extractor \
@@
 *deduplicated.bam
+```
 
@@
 
+```         
 NC_035784.1 141 141 37.5    3   5
@@
 NC_035784.1 611 611 0   0   4
+```
 
@@
 
+```         
 <chromosome> <start position> <end position> <methylation percentage> <count methylated> <count unmethylated>
+```
 
@@
 
+```         
 find *deduplicated.bismark.cov.gz \
@@
 {}_trimmed_bismark_bt2.deduplicated.bismark.cov.gz
+```
 
@@
 
+```         
 NC_035785.1 217 219 100.000000  17  0
@@
 NC_035785.1 2054    2056    93.333333   14  1
+```
 
@@
 
+```         
 for f in *merged_CpG_evidence.cov
@@
 done
+```
 
@@
 
+```         
 # Sort files for methylkit and IGV
@@
 index -@ 28 {}.sorted.bam
+```
 

### 2023-12-14 DRAFT_DNA_methylation.Rmd
add quant code explanation

@@
 
-## What to expect
-
-DNA methylation at CpG locus can vary to 0-100%.
-Like wise methylation at the gene level can vary and is commonly established as mean methylation level of CpG across gene, assuming some majority degree of gene coverage.
-Similarly with considering the methylation level of a given loci often time some threshold might be set as seen in code below.
-From an taxa perspective it can vary WIDELY on what to expect.
-The figure below shows the mehthylation ratio of CpGs across the oyster genome.
-In some marine invertebrates we see \~15% methylation of CpGs (defined as \> 75% methylated) across the genome with most CpGs unmethylated.
-In these taxa methylation usually occurs in gene bodies.
-
-```{r meth-ratio, echo = FALSE, out.width = "70%", fig.align = "center"}
-knitr::include_graphics("DRAFT_DNA_methylation/meth_ratio.png")
-```
-
 The software used as part of this tutorial is **Bismark Bisulfite Mapper** (@http://dx.doi.org/10.1093/bioinformatics/btr167).
@@
 
+Specifically this is what the code chunk does:
+
+1.  **Finding Files and Preparing Filenames**:
+
+    -   **`find ${reads_dir}*_R1_001_val_1.fq.gz`**: This command uses the **`find`** utility to search for files in the directory specified by the variable **`reads_dir`**.
+The files it looks for have a naming pattern ending with **`_R1_001_val_1.fq.gz`**, which typically denotes the first file of a pair in paired-end sequencing data.
+
+    -   **`| xargs basename -s _R1_001_val_1.fq.gz`**: The **`xargs`** command takes the list of files found and passes them to **`basename`**, which strips the suffix **`_R1_001_val_1.fq.gz`** from each filename.
+This step extracts the base name of each file, which is used to pair the R1 and R2 files correctly in the following steps.
+
+2.  **Running Bismark for Sequence Alignment**:
+
+    -   **`| xargs -I{} ${bismark_dir}/bismark`**: The **`xargs`** command here is used to pass the base filenames to Bismark for alignment. **`{}`** is a placeholder for each base filename.
+
+    -   **`--path_to_bowtie ${bowtie2_dir}`**: This option specifies the path to the Bowtie2 aligner, which Bismark uses for the alignment process.
+
+    -   **`-genome ${genome_folder}`**: Specifies the directory where the genome files, prepared in advance, are located.
+
+    -   **`-p 4`**: Sets the number of processing threads to 4, allowing Bismark to perform the alignment in parallel, which can speed up the process.
+
+    -   **`-score_min L,0,-0.6`**: This sets the scoring threshold for alignments, impacting the sensitivity and specificity of the alignment.
+
+    -   **`--non_directional`**: This flag indicates the sequencing data is non-directional, which is important for correctly interpreting the methylation status in bisulfite sequencing.
+
+    -   **`-1 ${reads_dir}{}_R1_001_val_1.fq.gz -2 ${reads_dir}{}_R2_001_val_2.fq.gz`**: Specifies the files for paired-end reads. R1 and R2 files are paired based on their base filename.
+
+    -   **`-o Mcap_tg`**: Sets the output directory for the alignment results to **`Mcap_tg`**.
+
 Files are then deduplicated if they are whole genome bisulfite treated samples.
@@
 
+## What to expect
+
+DNA methylation at CpG locus can vary to 0-100%.
+Like wise methylation at the gene level can vary and is commonly established as mean methylation level of CpG across gene, assuming some majority degree of gene coverage.
+Similarly with considering the methylation level of a given loci often time some threshold might be set as seen in code below.
+From an taxa perspective it can vary WIDELY on what to expect.
+The figure below shows the mehthylation ratio of CpGs across the oyster genome.
+In some marine invertebrates we see \~15% methylation of CpGs (defined as \> 75% methylated) across the genome with most CpGs unmethylated.
+In these taxa methylation usually occurs in gene bodies.
+
+```{r meth-ratio, echo = FALSE, out.width = "70%", fig.align = "center"}
+knitr::include_graphics("DRAFT_DNA_methylation/meth_ratio.png")
+```
+
 Methylation levels are then extracted using the `bismark_methylation_extractor`. For example:
@@
 ```
+
+This particular command is used to extract and process methylation data from BAM files (generated after sequencing data alignment). Let's break down the command and its options:
+
+1.  **Command Base**:
+
+    -   **`${bismark_dir}/bismark_methylation_extractor`**: This specifies the path to the **`bismark_methylation_extractor`** executable within the Bismark software directory.
+The variable **`${bismark_dir}`** holds the path to the Bismark installation directory.
+
+2.  **Options**:
+
+    -   **`--bedGraph`**: Generates a BedGraph file output, which is a format that allows visualization of the methylation levels across the genome in genome browsers.
+
+    -   **`--counts`**: Outputs the count of methylated and unmethylated calls for each cytosine in the genome.
+
+    -   **`--comprehensive`**: This option makes the methylation extractor process all cytosines in the genome, not just those in a CpG context.
+It combines all three contexts (CpG, CHG, and CHH) into one output file.
+
+    -   **`--merge_non_CpG`**: Merges the methylation information for CHG and CHH contexts (non-CpG) into a single file.
+
+    -   **`--multicore 28`**: Specifies that the program should use 28 cores to process the data in parallel, which can significantly speed up the analysis.
+
+    -   **`--buffer_size 75%`**: This option sets the buffer size for reading in BAM files to 75% of the available system memory.
+It helps optimize the performance of the tool by adjusting how much memory is used for processing.
+
+    -   **`*deduplicated.bam`**: This is the input file pattern for the command.
+It indicates that the methylation extractor should process all BAM files in the current directory that have names ending with "deduplicated.bam".
+These files are typically the output of a deduplication process applied to aligned bisulfite sequencing data.
 
@@
 
-Generating a file `.CpG_report.merged_CpG_evidence.cov` (merged bedGraph file)
+The script converts coverage files into a format that details cytosine methylation levels across the genome. Here's a breakdown of each step:
+
+1.  **Finding Files**:
+
+    -   **`find *deduplicated.bismark.cov.gz`**: This command is used to search for all files in the current directory that end with **`deduplicated.bismark.cov.gz`**.
+These files are coverage files produced by Bismark, typically after aligning bisulfite-treated DNA sequences.
+They contain information about methylation at different sites in the genome.
+
+2.  **Extracting Base File Names**:
+
+    -   **`| xargs basename -s _trimmed_bismark_bt2.deduplicated.bismark.cov.gz`**: This pipeline uses **`xargs`** to pass each found file to **`basename`**, which strips the suffix **`_trimmed_bismark_bt2.deduplicated.bismark.cov.gz`** from each file name.
+This step extracts the base names of the files, which will be used to generate output filenames in the following steps.
+
+3.  **Converting Coverage to Cytosine Methylation Levels**:
+
+    -   **`| xargs -I{} ${bismark_dir}/coverage2cytosine`**: This pipeline uses **`xargs`** again to pass each base filename to the **`coverage2cytosine`** tool in the Bismark suite.
+The placeholder **`{}`** is replaced by the base filename.
+
+    -   **`--genome_folder ${genome_folder}`**: This option specifies the location of the genome folder.
+The **`${genome_folder}`** variable holds the path to this folder, which contains reference genome data necessary for the analysis.
+
+    -   **`-o {}`**: The **`-o`** option specifies the output file name, using the base filename **`{}`** obtained from the previous steps.
+
+    -   **`--merge_CpG`**: This flag instructs the program to merge the methylation information of cytosines in CpG context, providing a single methylation value for each CpG site.
+
+    -   **`--zero_based`**: This option indicates that the output file should use zero-based coordinates, which is a common format in bioinformatics where the first base of a sequence is numbered 0 instead of 1.
+
+    -   **`{}_trimmed_bismark_bt2.deduplicated.bismark.cov.gz`**: Finally, this specifies the input file for each iteration, where **`{}`** is replaced by the base filename obtained earlier.
+These are the coverage files previously identified by the **`find`** command.
+
+Output file format `.CpG_report.merged_CpG_evidence.cov` (merged bedGraph file)
 

### 2023-12-14 DRAFT_DNA_methylation.Rmd
more meth tutorial edits (align)

@@
 title: 'DNA Methylation Analyses'
-subtitle: '*Steven Roberts*, *Sam White*, and *Yaamini Venkataraman*'
+subtitle: '*Yaamini Venkataraman*, *Sam White*, and *Steven Roberts*'
 bibliography: DRAFT_DNA_methylation/DNA-Methylation.bib
@@
 
-*Frequency distribution of methylation ratios for CpG dinucleotides in oyster gill tissue.* DOI: [10.7717/peerj.215/fig-1](https://doi.org/10.7717/peerj.215/fig-1)
-
 # Sequence Quality
@@
 # Read Alignment
+
+## Considerations
 
@@
 
-This will result in a prepared genome with associated new directories including
+Bismark will create the following subdirectories as outputs of the above code for genome preparation
 
@@
 
-This will create bam files (sequence alignment files)
+This will create BAM files (sequence alignment files)
 
@@
 
-This particular command is used to extract and process methylation data from BAM files (generated after sequencing data alignment). Let's break down the command and its options:
+This particular command is used to extract and process methylation data from BAM files (generated after sequencing data alignment). Let's break down the options:
 
-1.  **Command Base**:
-
-    -   **`${bismark_dir}/bismark_methylation_extractor`**: This specifies the path to the **`bismark_methylation_extractor`** executable within the Bismark software directory.
-The variable **`${bismark_dir}`** holds the path to the Bismark installation directory.
-
-2.  **Options**:
+1.  **Options**:
 

### 2023-12-14 DRAFT_DNA_methylation.Rmd
more on bismark genome prep

@@
 
-Bismark will create the following subdirectories as outputs of the above code for genome preparation
+Bismark will create two individual folders within this directory, one for a C->T converted genome and the other one for the G->A converted genome.
+After creating C->T and G->A versions of the genome they will be indexed in parallel usingthe indexer.
+Bismark will create the following subdirectories as outputs.
 

### 2023-12-14 DRAFT_DNA_methylation.Rmd
added ref

--- 2023-12-14 DRAFT_DNA_methylation.Rmd
@@
 ---
-title: 'DNA Methylation Analyses'
+title: 'DNA Methylation Assessment'
 subtitle: '*Yaamini Venkataraman*, *Sam White*, and *Steven Roberts*'
@@
 
-The tutorial will cover the following: - Sequence quality assessment\
+The tutorial will cover the following: \
+- Sequence quality assessment\
 - Read alignment\
@@
 # Sequence Quality
-
-An example of what raw data might look like from a marine invertebrate is shown.
-*Note that DNA methylation patters are diverse across taxa with most invertebrates demonstrating a mosiac pattern and global methylation at CpG loci is approximately 15%.*
 
@@
 
-Bismark will create the following subdirectories as outputs of the above code for genome preparation
+Bismark will create two individual folders within this directory, one for a C-\>T converted genome and the other one for the G-\>A converted genome.
+After creating C-\>T and G-\>A versions of the genome they will be indexed in parallel usingthe indexer.
+Bismark will create the following subdirectories as outputs.
 
@@
 ```
+
+*Frequency distribution of methylation ratios for CpG dinucleotides in oyster gill tissue.* @10.7717/peerj.215
 
@@
 
-# Potential downstream applications
+# Downstream applications
 

### 2023-12-14 DRAFT_DNA_methylation.Rmd
add igv screenshot meth

@@
 
+Location of methylation across the genome varies across taxa. Here is what you might expect to see in a marine invertebrate.
+
+```{r meth-igv, echo = FALSE, out.width = "70%", fig.align = "center"}
+knitr::include_graphics("DRAFT_DNA_methylation/igv.png")
+```
+*Screenshot from IGV. Bedgraph files are displayed. See below for code to create these file (File Conversion)*
+
 Methylation levels are then extracted using the `bismark_methylation_extractor`. For example:

### 2023-12-14 DRAFT_DNA_methylation.Rmd
add bismark citation in bib (meth)

@@
 
-The tutorial will cover the following: \
+The tutorial will cover the following:\
 - Sequence quality assessment\
@@
 
-The software used as part of this tutorial is **Bismark Bisulfite Mapper** (@http://dx.doi.org/10.1093/bioinformatics/btr167).
+The software used as part of this tutorial is **Bismark Bisulfite Mapper** @10.1093/bioinformatics/btr167.
 As with any software it is best to be famililar with the [User Manual](https://rawgit.com/FelixKrueger/Bismark/master/Docs/Bismark_User_Guide.html).
@@
 ```
+
 *Screenshot from IGV. Bedgraph files are displayed. See below for code to create these file (File Conversion)*

### 2023-12-14 DRAFT_DNA_methylation.Rmd
add bsnpr meth

@@
 
-The software used as part of this tutorial is **Bismark Bisulfite Mapper** @10.1093/bioinformatics/btr167.
+The software used as part of this tutorial is **Bismark Bisulfite Mapper** (@10.1093/bioinformatics/btr167).
 As with any software it is best to be famililar with the [User Manual](https://rawgit.com/FelixKrueger/Bismark/master/Docs/Bismark_User_Guide.html).
@@
 
-# Downstream applications
+# Next Steps & Downstream applications
+
+## CT SNPs
+
+Masking cytosine (C) to thymine (T) single nucleotide polymorphisms (SNPs) in bisulfite sequencing analysis could be something to considerfor a few reasons:
+
+1.
+**Bisulfite Conversion Process**: In bisulfite sequencing, unmethylated cytosines are converted to uracil, which is then read as thymine in sequencing.
+This conversion is key to identifying methylated cytosines.
+However, if a genomic DNA sequence naturally has a C to T SNP, it can be falsely interpreted as a converted cytosine, leading to incorrect conclusions about methylation.
+
+2.
+**Reduction of False Positives**: By masking C to T SNPs, one can reduce false positives in methylation analysis.
+Without masking, these SNPs might be misinterpreted as unmethylated cytosines that have undergone conversion, skewing the data towards an apparent increase in methylation.
+
+3.
+**Accurate Methylation Mapping**: For precise mapping of methylation patterns, it's crucial to distinguish between true methylation signals and genetic variation (SNPs).
+Masking C to T SNPs allows for a more accurate representation of the methylation status of the genome.
+
+4.
+**Analysis Consistency**: In comparative studies, where multiple samples or conditions are analyzed, masking C to T SNPs ensures consistency across samples.
+This is especially important in cases where SNP profiles may vary between samples, which could otherwise lead to inconsistent methylation data.
+
+5.
+**Enhanced Data Interpretation**: By eliminating these potential sources of error, researchers can have more confidence in their interpretation of the bisulfite sequencing data, leading to more reliable and valid conclusions about DNA methylation patterns.
+
+One way to do this is using BS-Snper (@10.1093/bioinformatics/btv507).
+<https://github.com/hellbelly/BS-Snper>.
+In short one could merge all BAM files from experiment run BS-Snper, filter out CT SNPS then use this to remove loci.
+
+```         
+samtools merge \
+Pg_merged.bam \
+*.sorted.bam
+
+perl /gscratch/srlab/programs/BS-Snper-master/BS-Snper.pl \
+Pg_merged.bam \
+--fa Panopea-generosa-v1.0.fa \
+--output snp.candidate.out \
+--methcg meth.cg \
+--methchg meth.chg \
+--methchh meth.chh \
+--minhetfreq 0.1 \
+--minhomfreq 0.85 \
+--minquali 15 \
+--mincover 10 \
+--maxcover 1000 \
+--minread2 2 \
+--errorate 0.02 \
+--mapvalue 20 \
+>SNP.vcf 2>SNP.log
+
+grep $'C\tG' output/SNP.vcf > output/CT-SNP.vcf
+wc -l output/CT-SNP.vcf
+```
+
+Let's break down the command and its arguments:
+
+1.
+**`perl /gscratch/srlab/programs/BS-Snper-master/BS-Snper.pl`**: This part of the command calls the Perl interpreter to execute the **`BS-Snper.pl`** script.
+The script is located in the **`/gscratch/srlab/programs/BS-Snper-master/`** directory.
+
+2.  **`Pg_merged.bam`**: This is the input file for the script, likely a BAM file (Binary Alignment/Map format) which contains alignment information of bisulfite-treated sequencing reads.
+
+3.  **`--fa Panopea-generosa-v1.0.fa`**: This option specifies the reference genome file in FASTA format against which the BAM file will be compared.
+
+4.  **`--output snp.candidate.out`**: This argument sets the name of the output file where the SNP candidates will be written.
+
+5.
+**`--methcg meth.cg`, `--methchg meth.chg`, `--methchh meth.chh`**: These options specify the output files for methylation analysis for CG, CHG, and CHH contexts, respectively (where H can be A, C, or T).
+
+6.  **`--minhetfreq 0.1`**: Sets the minimum frequency for heterozygous bases to be considered in SNP calling (here, 10%).
+
+7.  **`--minhomfreq 0.85`**: Sets the minimum frequency for homozygous bases to be considered in SNP calling (here, 85%).
+
+8.  **`--minquali 15`**: Sets the minimum base quality score.
+
+9.  **`--mincover 10`, `--maxcover 1000`**: These set the minimum and maximum coverage (number of reads) for a position to be considered in the analysis.
+
+10. **`--minread2 2`**: The minimum number of reads required to support a SNP candidate.
+
+11. **`--errorate 0.02`**: Sets the error rate of sequencing (here, 2%).
+
+12. **`--mapvalue 20`**: Specifies the mapping quality value, a threshold for considering reads in the analysis.
+
+13. **`>SNP.vcf`**: Redirects the standard output to a file named **`SNP.vcf`**. This file will contain the SNP calls in VCF (Variant Call Format) format.
+
+14. **`2>SNP.log`**: Redirects the standard error output to a file named **`SNP.log`**. This file will contain error messages and possibly additional log information.
+
+If for example in downstream analysis you would might be using a _5x.tab file described above the following is a means in R to remove these SNPs.
+
+```{r}
+# Read in CT SNP file
+ct <- read.csv("../output/CT-SNP.vcf", header = FALSE, sep = "\t") %>%
+  mutate(loci = paste0(V1, "_", V2))
+
+# 1. List all files with _5x.tab suffix
+files <- list.files(path = "../data/", pattern = "_5x.tab$", full.names = TRUE)
+
+# 2. Iterate over each file
+for(file in files) {
+
+  # Extract base filename without the directory for naming purposes
+  base_name <- basename(file)
+
+  # Read the file
+  data <- read.csv(file, header = FALSE, sep = "\t")
+
+  # Modify the data
+  modified_data <- data %>%
+    mutate(loci = paste0(V1, "_", V2)) %>%
+    anti_join(ct, by = "loci") %>%
+    select(-loci)
+
+  # Write the modified data to an output file
+  output_file <- paste0("../output/f", base_name)
+  write.table(modified_data, file = output_file, sep = "\t", row.names = FALSE, quote = FALSE, col.names = FALSE)
+}
+```
+
+## Other application
 

### 2023-12-14 DRAFT_DNA_methylation.Rmd
added secion on lambda phage spiking

@@
 
+## Background
+
 Thanks to advances in sample preparation and sequencing methods, there is an influx of research examining epigenomics in non-model systems.
@@
 As EM-Seq produces loci-level data, alignment of EM-Seq data is similar to WGBS.
+Also, technologies like Nanopore and PacBio sequencing can directly detect DNA methylation without bisulfite conversion.
+As DNA passes through a nanopore or is sequenced by single-molecule real-time (SMRT) technology, changes in electrical current or fluorescence patterns can indicate the presence of methylated bases.
+
+## Experimental Considerations
+
+### Sample Size
+
+something about samples size.
+
+### Bisulfite Conversion Efficiency
+
+Bisulfite Conversion Efficiency is important to assess.
+One common step during library preparation us spiking library with Lambda phage DNA.
+Lambda phage DNA lacks cytosine methylation, making it an ideal control for bisulfite treatment.
+Here's how it works and why it's useful:
+
+1.
+**Background on Lambda Phage DNA**: Lambda phage DNA is a widely used control in bisulfite sequencing because it is naturally unmethylated.
+By including this DNA in your sample, you create a known unmethylated reference.
+
+2.
+**Preparation and Spiking**: You prepare the lambda phage DNA and spike it into your DNA samples before bisulfite treatment.
+The proportion of lambda DNA added should be enough to be detected post-sequencing, but not so much that it overwhelms the sample.
+A typical proportion might be around 0.1% to 1% of the total DNA.
+
+3.
+**Bisulfite Treatment and Sequencing**: After spiking, you proceed with bisulfite treatment and sequencing of the entire sample, which now includes your genomic DNA of interest and the spiked lambda DNA.
+
+4.
+**Assessing Conversion Efficiency**: Once sequencing is complete, you analyze the lambda phage DNA sequences.
+Since this DNA is unmethylated, all cytosines should be converted to thymines after bisulfite treatment and subsequent PCR.
+By calculating the percentage of cytosines that were successfully converted to thymines in the lambda DNA, you can determine the bisulfite conversion efficiency.
+An efficient conversion typically results in over 99% of cytosines being converted.
+
+## Tutorial Details
 
@@
 
-If for example in downstream analysis you would might be using a _5x.tab file described above the following is a means in R to remove these SNPs.
+If for example in downstream analysis you would might be using a \_5x.tab file described above the following is a means in R to remove these SNPs.
 
-```{r}
+```{r, eval=FALSE}
 # Read in CT SNP file

### 2023-12-14 DRAFT_DNA_methylation.Rmd
methy - sample size

@@
 
-something about samples size.
+Sample size is a critical factor when conducting DNA methylation analysis, particularly because of the significant variation in methylation patterns both between individuals and across different tissues.
+This variation, which is not yet fully understood, can significantly impact the outcomes and interpretations of methylation studies.
+
+When selecting samples for DNA methylation analysis, the key principle is that larger sample sizes generally provide more reliable and generalizable results.
+A larger number of samples allows for a more accurate representation of the population, accounting for individual and tissue-level differences.
+This is essential for identifying true biological patterns rather than artifacts of small, non-representative samples.
+
+However, the practicality of obtaining a large number of samples must be balanced with resource constraints.
+In this context, a sample size of around eight can be considered a decent starting point.
+This number is often sufficient to begin observing trends and patterns in methylation across individuals and tissues, providing a foundational understanding which can be built upon with further research.
+
+It's crucial to remember that while more samples can enhance the reliability of the results, the quality of the samples and the robustness of the methylation analysis techniques are equally important.
+Therefore, researchers must strive for a balance between quantity and quality in their sample selection to ensure meaningful and accurate insights into DNA methylation.
 

### 2023-12-14 DRAFT_DNA_methylation.Rmd
Update DRAFT_DNA_methylation.Rmd

@@
 Therefore, researchers must strive for a balance between quantity and quality in their sample selection to ensure meaningful and accurate insights into DNA methylation.
+
+### Sequencing Depth
 
@@
 
-The software used as part of this tutorial is **Bismark Bisulfite Mapper** (@10.1093/bioinformatics/btr167).
+The software used as part of this tutorial is **Bismark Bisulfite Mapper** (@10.1093/bioinformatics/btr167), which is one of the more widely-used alignment softwares for methylation data.
+Another commonly-used options include bwa-meth (@https://doi.org/10.48550/arXiv.1401.1129) and BSMAP (@https://doi.org/10.1186/1471-2105-10-232) (see @https://doi.org/10.1093/bib/bbab021 and @https://doi.org/10.1016/j.csbj.2022.08.051 for comparisons of these software for plant and mammal species, respectively).
 As with any software it is best to be famililar with the [User Manual](https://rawgit.com/FelixKrueger/Bismark/master/Docs/Bismark_User_Guide.html).

### 2023-12-14 DRAFT_DNA_methylation.Rmd
downstream

--- 2023-12-14 DRAFT_DNA_methylation.Rmd
@@
 
-## Other application
+## Downstream application
 
--   Gene-level methylation analysis: people do this, but worth considering the utility of converting loci-level resolution data to gene-level resolution data
--   methylKit
--   IGV
--   Analysis with other datasets (ex. pop gen, RNA-Seq)
+-   Gene-level methylation analysis. It is worth considering the utility of converting loci-level resolution data to gene-level resolution data
+
+-   Differential loci or region analyses. methylKit is a popular package to conduct this analysis
+
+-   Feature overlap. Bedtools for quantitative assessment and IGV for visual assessment are valuable for thisl
 

### 2023-12-14 DRAFT_DNA_methylation.Rmd
Update DRAFT_DNA_methylation.Rmd
added more context for bowtie

--- 2023-12-14 DRAFT_DNA_methylation.Rmd
@@
 
-The purpose of this tutorial is to provide an example of aligning bisulfite-treated and enzymatically converted DNA sequence data and discuss needs for potential downstream applications.
+The purpose of this tutorial is to provide an example of aligning bisulfite-treated and enzymatically converted DNA sequence data to an existing genome, and discuss needs for potential downstream applications.
 The principle behind creating bisulfite-treated DNA libraries is that when cytosines are not methylated, then they will be converted to uracil, which ultimately result in thymine nucleotides in the sequence data.
@@
 
-## Considerations
+## Software Considerations
 
-The software used as part of this tutorial is **Bismark Bisulfite Mapper** (@10.1093/bioinformatics/btr167), which is one of the more widely-used alignment softwares for methylation data.
+The software used as part of this tutorial is **Bismark Bisulfite Mapper** (@10.1093/bioinformatics/btr167), which is one of the more widely-used softwares to align methylation data to a reference genome.
 Another commonly-used options include bwa-meth (@https://doi.org/10.48550/arXiv.1401.1129) and BSMAP (@https://doi.org/10.1186/1471-2105-10-232) (see @https://doi.org/10.1093/bib/bbab021 and @https://doi.org/10.1016/j.csbj.2022.08.051 for comparisons of these software for plant and mammal species, respectively).
-As with any software it is best to be famililar with the [User Manual](https://rawgit.com/FelixKrueger/Bismark/master/Docs/Bismark_User_Guide.html).
+As with any software it is best to be famililar with the [User Manual](https://felixkrueger.github.io/Bismark/).
+
+Within Bismark, the tutorial uses Bowtie 2 as the alignment software.
+
 The content below will provided with the assumption that the reader has read the manual and is meant to serve as guidance based on experience working with marine invertebrates.
 
-The first step in the process in preparing the genome. Example code:
+## Genome Preparation
+
+The first step in the process in preparing the reference genome by performing an in silico bisulfite conversion to allow Bowtie alignments. Example code:
 
@@
 
-    -   **`--parallel 28`**: This option tells Bismark to use 28 threads in parallel to speed up the process.
+    -   **`--parallel 28`**: This option tells Bismark to use 28 threads in parallel to speed up the process. You can adjust this based on your computing power.
 
@@
 ```
+
+## Aligning reads
 

### 2023-12-14 DRAFT_DNA_methylation.Rmd
Update DRAFT_DNA_methylation.Rmd
added information on score_min

@@
 
-Within Bismark, the tutorial uses Bowtie 2 as the alignment software.
+Within Bismark, the tutorial uses Bowtie 2 (@https://doi.org/10.1038/nmeth.1923) as the alignment software.
+Although the Bismark user manual covers the necessary Bowtie 2 parameters, it may be useful to refer to the [Bowtie 2 manual](https://bowtie-bio.sourceforge.net/bowtie2/manual.shtml).
 
@@
 
+This preparation step only needs to happen once for a given genome.
+
 ## Aligning reads
 
-For aligning the trimmed reads to the genome the following code structure is used
+Once the genome is prepared, reads can be aligned. For aligning the trimmed reads to the genome the following code structure is used:
 
@@
 
-This will create BAM files (sequence alignment files)
+This will create BAM files (sequence alignment files) and txt files (mapping reports). The reports contain information such as mapping efficiency.
 
@@
 
-    -   **`--path_to_bowtie ${bowtie2_dir}`**: This option specifies the path to the Bowtie2 aligner, which Bismark uses for the alignment process.
+    -   **`--path_to_bowtie ${bowtie2_dir}`**: This option specifies the path to the Bowtie 2 aligner, which Bismark uses for the alignment process.
 
@@
     -   **`-p 4`**: Sets the number of processing threads to 4, allowing Bismark to perform the alignment in parallel, which can speed up the process.
+This parameter can be changed based on your computing power.
 
     -   **`-score_min L,0,-0.6`**: This sets the scoring threshold for alignments, impacting the sensitivity and specificity of the alignment.
+The default scoring threshold used by Bismark is L,0,-0.2.
 
@@
     -   **`-o Mcap_tg`**: Sets the output directory for the alignment results to **`Mcap_tg`**.
+
+### Mapping reports
+
+EXAMPLE MAPPING REPORT FROM BISMARK AND HOW TO VISUALIZE IT
+
+### Optimizing alignment sensitivity
+
+The code above uses `score_min L,0,-0.6` as the scoring threshold for alignment sensitivity and specificity.
+This is different than the Bismark default, `score_min L,0,-0.2`.
+Reducing the sensitivity of the alignment can increase the percentage of reads aligned to the reference genome, but of course the quality of the alignment may decrease.
+A good practice is to align a subset of your data to the reference genome using different `score_min` parameters, then comparing mapping efficiency from the mapping reports.
+Generally, you want to choose a parameter that maximizes percent alignment and specificity.
+Example code
+
+```         
+find ${reads_dir}*_R1_001_val_1.fq.gz \
+| xargs basename -s _R1_001_val_1.fq.gz | xargs -I{} ${bismark_dir}/bismark \
+--path_to_bowtie ${bowtie2_dir} \
+-genome ${genome_folder} \
+-p 4 \
+- u 10000 \
+-score_min L,0,-0.6 \
+--non_directional \
+-1 ${reads_dir}{}_R1_001_val_1.fq.gz \
+-2 ${reads_dir}{}_R2_001_val_2.fq.gz \
+-o Mcap_tg
+```
+
+This is the same code as above, except `-u 10000` is used to run the alignment on a 10,000 bp subset of the data in each input file.
+
+## Deduplication
 

### 2023-12-14 DRAFT_DNA_methylation.Rmd
Update DRAFT_DNA_methylation.Rmd

@@
 DNA methylation at CpG locus can vary to 0-100%.
-Like wise methylation at the gene level can vary and is commonly established as mean methylation level of CpG across gene, assuming some majority degree of gene coverage.
-Similarly with considering the methylation level of a given loci often time some threshold might be set as seen in code below.
-From an taxa perspective it can vary WIDELY on what to expect.
-The figure below shows the mehthylation ratio of CpGs across the oyster genome.
+How methylated CpGs are across a genome can vary WIDELY depending on your focal taxa.
+Generally, invertebrate genomes are predominantly unmethylated, with smaller regions of higher methylation in a mosaic pattern.
 In some marine invertebrates we see \~15% methylation of CpGs (defined as \> 75% methylated) across the genome with most CpGs unmethylated.
 In these taxa methylation usually occurs in gene bodies.
+
+The figure below shows the methylation ratio of CpGs across the Pacific oyster (*Crassostrea gigas*) genome, as well as the location of methylation across the genome:
 
@@
 
-Location of methylation across the genome varies across taxa. Here is what you might expect to see in a marine invertebrate.
-
 ```{r meth-igv, echo = FALSE, out.width = "70%", fig.align = "center"}
@@
 *Screenshot from IGV. Bedgraph files are displayed. See below for code to create these file (File Conversion)*
+
+Vertebrate genomes, on the other hand, have higher overall levels of methylation, with methylation occuring in CpG islands.
+
+[ placeholder for Kathryn to add some stuff! ]
+
+## Methylation extraction
 

### 2023-12-14 DRAFT_DNA_methylation.Rmd
Update DRAFT_DNA_methylation.Rmd

@@
 title: 'DNA Methylation Assessment'
-subtitle: '*Yaamini Venkataraman*, *Sam White*, and *Steven Roberts*'
+subtitle: '*Yaamini Venkataraman*, *Kathryn Greiner-Ferris*, *Sam White*, and *Steven Roberts*'
 bibliography: DRAFT_DNA_methylation/DNA-Methylation.bib
@@
     -   **`--multicore 28`**: Specifies that the program should use 28 cores to process the data in parallel, which can significantly speed up the analysis.
+This can be adjusted depending on your computing power.
 
@@
 It helps optimize the performance of the tool by adjusting how much memory is used for processing.
+This can be adjusted depending on your computing power.
 
@@
 
-This will create a file with the suffix `deduplicated.bismark.cov.gz` (example shown below in uncompressed format).
-The resultant bedGraph file contains information from *unmerged* strands, meaning reads aligning to forward and reverse strands in the same position are not combined.
-Later in this workflow, a bedGraph with *merged* information will be created.
+This will create a file with the suffix `deduplicated.bismark.cov.gz` (example shown below in uncompressed format):
 
@@
 
-A genome-wide cytosine report is generated (including merged bedGraph file), with this code as an example:
+The resultant bedGraph file contains information from *unmerged* strands, meaning reads aligning to forward and reverse strands in the same position are not combined.
+It is important to merge the reads from the forward and reserve strands for more accurate methylation quantification.
+A genome-wide cytosine report and merged bedGraph file are generated (including merged bedGraph file) from the `deduplicated.bismark.cov.gz` generated above in this example code:
 
@@
 
+column organization of the file
+
+```         
+<chromosome> <start position> <end position> <methylation percentage> <count methylated> <count unmethylated>
+```
+
+# Summary report
+
 # File Conversions
 
-From here it could be useful to do some simple file reorganizaion to obtain bedGraph or tab files for downstream analysis (eg bedtools, GLMs).
+From here it could be useful to do some file reorganizaion to obtain bedGraph or tab files for downstream analysis (eg `bedtools`, GLMs).
+Specifically, it may be important to filter the CpG methylation data to a specific coverage threshold (ex.
+5x or 10x).
+The more read coverage a CpG has, the higher the confidence in methylation quantification.
 
@@
 ```         
-for f in *merged_CpG_evidence.cov
-do
-  STEM=$(basename "${f}" .CpG_report.merged_CpG_evidence.cov)
-  cat "${f}" | awk -F $'\t' 'BEGIN {OFS = FS} {if ($5+$6 >= 10) {print $1, $2, $3, $4}}' \
-  > "${STEM}"_10x.bedgraph
-done
-
 for f in *merged_CpG_evidence.cov
@@
 ```         
-for f in *merged_CpG_evidence.cov
-do
-  STEM=$(basename "${f}" .CpG_report.merged_CpG_evidence.cov)
-  cat "${f}" | awk -F $'\t' 'BEGIN {OFS = FS} {if ($5+$6 >= 10) {print $1, $2, $3, $4, $5, $6}}' \
-  > "${STEM}"_10x.tab
-done
-
 for f in *merged_CpG_evidence.cov
@@
 
-# Next Steps & Downstream applications
+# Next Steps
 
-## CT SNPs
+So you've aligned your trimmed reads to your reference genome and have methylation data for CpGs. Now what?
 
-Masking cytosine (C) to thymine (T) single nucleotide polymorphisms (SNPs) in bisulfite sequencing analysis could be something to considerfor a few reasons:
+## Removing C to T (C->T) SNPs
+
+Removing cytosine (C) to thymine (T) single nucleotide polymorphisms (SNPs) in bisulfite sequencing analysis could be something to consider for a few reasons:
 
@@
 This conversion is key to identifying methylated cytosines.
-However, if a genomic DNA sequence naturally has a C to T SNP, it can be falsely interpreted as a converted cytosine, leading to incorrect conclusions about methylation.
+However, if a genomic DNA sequence naturally has a C to T (C->T) SNP, it can be falsely interpreted as a converted cytosine, leading to incorrect conclusions about methylation.
 
 2.
-**Reduction of False Positives**: By masking C to T SNPs, one can reduce false positives in methylation analysis.
-Without masking, these SNPs might be misinterpreted as unmethylated cytosines that have undergone conversion, skewing the data towards an apparent increase in methylation.
+**Reduction of False Positives**: By removing C->T SNPs, one can reduce false positives in methylation analysis.
+Without removal, these SNPs might be misinterpreted as unmethylated cytosines that have undergone conversion, skewing the data towards an apparent increase in methylation.
 
@@
 **Accurate Methylation Mapping**: For precise mapping of methylation patterns, it's crucial to distinguish between true methylation signals and genetic variation (SNPs).
-Masking C to T SNPs allows for a more accurate representation of the methylation status of the genome.
+Removing C->T SNPs allows for a more accurate representation of the methylation status of the genome.
 
 4.
-**Analysis Consistency**: In comparative studies, where multiple samples or conditions are analyzed, masking C to T SNPs ensures consistency across samples.
+**Analysis Consistency**: In comparative studies, where multiple samples or conditions are analyzed, removing C->T SNPs ensures consistency across samples.
 This is especially important in cases where SNP profiles may vary between samples, which could otherwise lead to inconsistent methylation data.
 
-5.
-**Enhanced Data Interpretation**: By eliminating these potential sources of error, researchers can have more confidence in their interpretation of the bisulfite sequencing data, leading to more reliable and valid conclusions about DNA methylation patterns.
+By eliminating these potential sources of error, researchers can have more confidence in their interpretation of the bisulfite sequencing data, leading to more reliable and valid conclusions about DNA methylation patterns.
 
@@
 <https://github.com/hellbelly/BS-Snper>.
-In short one could merge all BAM files from experiment run BS-Snper, filter out CT SNPS then use this to remove loci.
+In short one could merge all BAM files from experiment run BS-Snper, filter out C->T SNPS then use this to remove loci.
 
@@
 
-10. **`--minread2 2`**: The minimum number of reads required to support a SNP candidate.
+10.
+**`--minread2 2`**: The minimum number of reads required to support a SNP candidate.
+It could be useful to use the same values for the minimum read threshold for SNP identification and for minimum coverage in the methylation data.
 

### 2023-12-14 DRAFT_DNA_methylation.Rmd
Update DRAFT_DNA_methylation.Rmd
added bismark summary report code

@@
 
-### Mapping reports
-
-EXAMPLE MAPPING REPORT FROM BISMARK AND HOW TO VISUALIZE IT
-
 ### Optimizing alignment sensitivity
@@
 
-# Summary report
+# Summary Report
+
+Many steps of the Bismark workflow produce reports (.txt files) with useful information (ex.
+mapping efficiency, percent cytosine methylation, etc.).
+These files can be collated into processing reports for each individual sample, which can then be further collated into an [HTML summary report](https://gannet.fish.washington.edu/seashell/bu-mox/scrubbed/032120-fds/bismark_summary_report.html):
+
+```
+#HTML Processing Report
+
+${bismark_dir}/bismark2report
+
+#Summary Report
+
+${bismark_dir}/bismark2summary
+```
+
+These commands should be run in the same directory where all output files are stored.
+
+```{r meth-igv, echo = FALSE, out.width = "70%", fig.align = "center"}
+knitr::include_graphics("DRAFT_DNA_methylation/bismark-summary-report.png")
+```
 

### 2023-12-15 DRAFT_DNA_methylation.Rmd
adding explanation of BSsnper

@@
 <https://github.com/hellbelly/BS-Snper>.
-In short one could merge all BAM files from experiment run BS-Snper, filter out C->T SNPS then use this to remove loci.
+This program finds SNPs in Bisulfite treated alignment files.
+The bash code below represents steps to identify SNPS and create a list of all C->T SNPS.
 
@@
 
+The resulting could be used to mask the genome and the entire process described abo
+
 If for example in downstream analysis you would might be using a \_5x.tab file described above the following is a means in R to remove these SNPs.

### 2023-12-15 DRAFT_DNA_methylation.Rmd
meth - mod summary image chunk

@@
 
-```{r meth-igv, echo = FALSE, out.width = "70%", fig.align = "center"}
+```{r meth-summary, echo = FALSE, out.width = "70%", fig.align = "center"}
 knitr::include_graphics("DRAFT_DNA_methylation/bismark-summary-report.png")
@@
 
-The resulting could be used to mask the genome and the entire process described abo
+The resulting could be used to mask the genome and the entire process described above could be repeated (ie Genome Preparation, Alignment, Quantification)
 

### 2023-12-15 DRAFT_DNA_methylation.Rmd
meth - added explanation of R code post SNP

@@
 
-## Removing C to T (C->T) SNPs
+## Removing C to T (C-\>T) SNPs
 
@@
 This conversion is key to identifying methylated cytosines.
-However, if a genomic DNA sequence naturally has a C to T (C->T) SNP, it can be falsely interpreted as a converted cytosine, leading to incorrect conclusions about methylation.
+However, if a genomic DNA sequence naturally has a C to T (C-\>T) SNP, it can be falsely interpreted as a converted cytosine, leading to incorrect conclusions about methylation.
 
 2.
-**Reduction of False Positives**: By removing C->T SNPs, one can reduce false positives in methylation analysis.
+**Reduction of False Positives**: By removing C-\>T SNPs, one can reduce false positives in methylation analysis.
 Without removal, these SNPs might be misinterpreted as unmethylated cytosines that have undergone conversion, skewing the data towards an apparent increase in methylation.
@@
 **Accurate Methylation Mapping**: For precise mapping of methylation patterns, it's crucial to distinguish between true methylation signals and genetic variation (SNPs).
-Removing C->T SNPs allows for a more accurate representation of the methylation status of the genome.
+Removing C-\>T SNPs allows for a more accurate representation of the methylation status of the genome.
 
 4.
-**Analysis Consistency**: In comparative studies, where multiple samples or conditions are analyzed, removing C->T SNPs ensures consistency across samples.
+**Analysis Consistency**: In comparative studies, where multiple samples or conditions are analyzed, removing C-\>T SNPs ensures consistency across samples.
 This is especially important in cases where SNP profiles may vary between samples, which could otherwise lead to inconsistent methylation data.
@@
 This program finds SNPs in Bisulfite treated alignment files.
-The bash code below represents steps to identify SNPS and create a list of all C->T SNPS.
+The bash code below represents steps to identify SNPS and create a list of all C-\>T SNPS.
 
@@
 
-The resulting could be used to mask the genome and the entire process described above could be repeated (ie Genome Preparation, Alignment, Quantification)
-
-If for example in downstream analysis you would might be using a \_5x.tab file described above the following is a means in R to remove these SNPs.
+The resulting could be used to mask the genome and the entire process described above could be repeated (ie Genome Preparation, Alignment, Quantification).
+An alternative and common practice is to remove cytosines from downstream analysis by removing loci in intermediate files (ie tab-delimited data files).
+Below is an example using `antijoin` to remove loci that correspond to CT SNPs from 5x coverage methyation quantification files.
 
@@
 
+Let's break down what each part of the script does:
+
+1.  **Reading in a CT SNP file**:
+
+    -   **`ct <- read.csv("../output/CT-SNP.vcf", header = FALSE, sep = "\t")`**: This line reads a file named **`CT-SNP.vcf`** from the specified directory.
+The file is expected to be in a tab-separated format (**`sep = "\t"`**), and it doesn't have a header row (**`header = FALSE`**).
+
+    -   **`mutate(loci = paste0(V1, "_", V2))`**: The **`mutate`** function from the **`dplyr`** package is used to add a new column (**`loci`**) to the **`ct`** data frame.
+This column is created by concatenating the values in the first (**`V1`**) and second columns (**`V2`**) of each row, separated by an underscore.
+
+2.  **Listing files with a specific suffix**:
+
+    -   **`files <- list.files(path = "../data/", pattern = "_5x.tab$", full.names = TRUE)`**: This line generates a list of file paths in the **`../data/`** directory.
+It only includes files that end with **`_5x.tab`** (as indicated by the **`pattern`** argument).
+
+3.  **Processing each file**:
+
+    -   The **`for(file in files)`** loop iterates over each file found in the previous step.
+
+        -   **`base_name <- basename(file)`**: Extracts the file name without the directory path.
+
+        -   **`data <- read.csv(file, header = FALSE, sep = "\t")`**: Reads the current file in the loop. Similar to the first **`read.csv`** call, it expects a tab-separated file with no header row.
+
+        -   The **`mutate`** and **`anti_join`** functions are used to modify **`data`**:
+
+            -   **`mutate(loci = paste0(V1, "_", V2))`**: Adds a **`loci`** column similar to the **`ct`** data frame.
+
+            -   **`anti_join(ct, by = "loci")`**: Removes rows from **`data`** that have matching **`loci`** values in the **`ct`** data frame.
+
+            -   **`select(-loci)`**: Removes the **`loci`** column from the data.
+
+        -   The modified data is then written to an output file:
+
+            -   **`output_file <- paste0("../output/f", base_name)`**: Constructs the output file path.
+
+            -   **`write.table(modified_data, file = output_file, sep = "\t", row.names = FALSE, quote = FALSE, col.names = FALSE)`**: Writes the **`modified_data`** to the specified file.
+The output is tab-separated, and it doesn't include row names, quotes, or column names.
+
 ## Downstream application

### 2023-12-15 DRAFT_DNA_methylation.Rmd
knit with bssnp R code
meth - added more bibtex - citations

@@
 The software used as part of this tutorial is **Bismark Bisulfite Mapper** (@10.1093/bioinformatics/btr167), which is one of the more widely-used softwares to align methylation data to a reference genome.
-Another commonly-used options include bwa-meth (@https://doi.org/10.48550/arXiv.1401.1129) and BSMAP (@https://doi.org/10.1186/1471-2105-10-232) (see @https://doi.org/10.1093/bib/bbab021 and @https://doi.org/10.1016/j.csbj.2022.08.051 for comparisons of these software for plant and mammal species, respectively).
+Another commonly-used options include bwa-meth (@pedersen2014fast) and BSMAP (@10.1186/1471-2105-10-232) (see @10.1093/bib/bbab021 and @10.1016/j.csbj.2022.08.051) for comparisons of these software for plant and mammal species, respectively).
 As with any software it is best to be famililar with the [User Manual](https://felixkrueger.github.io/Bismark/).
 
-Within Bismark, the tutorial uses Bowtie 2 (@https://doi.org/10.1038/nmeth.1923) as the alignment software.
+Within Bismark, the tutorial uses Bowtie 2 (@10.1038/nmeth.1923) as the alignment software.
 Although the Bismark user manual covers the necessary Bowtie 2 parameters, it may be useful to refer to the [Bowtie 2 manual](https://bowtie-bio.sourceforge.net/bowtie2/manual.shtml).
@@
 
--   Feature overlap. Bedtools for quantitative assessment and IGV for visual assessment are valuable for thisl
+-   Feature overlap. Bedtools for quantitative assessment and IGV for visual assessment are valuable for this.
 

### 2023-12-15 DRAFT_DNA_methylation.Rmd
image formatting - meth

@@
                cache = FALSE)
-cite_options(citation_format = "pandoc", max.names = 3, style = "html", 
-             hyperlink = "to.doc")
+cite_options(citation_format = "pandoc", max.names = 3, style = "html", hyperlink = "to.html")
 ```
@@
 
-The software used as part of this tutorial is **Bismark Bisulfite Mapper** (@10.1093/bioinformatics/btr167), which is one of the more widely-used softwares to align methylation data to a reference genome.
-Another commonly-used options include bwa-meth (@pedersen2014fast) and BSMAP (@10.1186/1471-2105-10-232) (see @10.1093/bib/bbab021 and @10.1016/j.csbj.2022.08.051) for comparisons of these software for plant and mammal species, respectively).
+The software used as part of this tutorial is **Bismark Bisulfite Mapper** [@10.1093/bioinformatics/btr167], which is one of the more widely-used softwares to align methylation data to a reference genome.
+Another commonly-used options include bwa-meth [@pedersen2014fast] and BSMAP [@10.1186/1471-2105-10-232] (see [@10.1093/bib/bbab021] and [@10.1016/j.csbj.2022.08.051] for comparisons of these software for plant and mammal species, respectively).
 As with any software it is best to be famililar with the [User Manual](https://felixkrueger.github.io/Bismark/).
 
-Within Bismark, the tutorial uses Bowtie 2 (@10.1038/nmeth.1923) as the alignment software.
+Within Bismark, the tutorial uses Bowtie 2 [@10.1038/nmeth.1923] as the alignment software.
 Although the Bismark user manual covers the necessary Bowtie 2 parameters, it may be useful to refer to the [Bowtie 2 manual](https://bowtie-bio.sourceforge.net/bowtie2/manual.shtml).
@@
 
-```{r meth-ratio, echo = FALSE, out.width = "70%", fig.align = "center"}
+```{r meth-ratio, echo = FALSE, out.width = "70%", fig.align = "center", fig.cap = "Frequency distribution of methylation ratios for CpG dinucleotides in oyster gill tissue. @10.7717/peerj.215"}
 knitr::include_graphics("DRAFT_DNA_methylation/meth_ratio.png")
@@
 
-*Frequency distribution of methylation ratios for CpG dinucleotides in oyster gill tissue.* @10.7717/peerj.215
-
-```{r meth-igv, echo = FALSE, out.width = "70%", fig.align = "center"}
+```{r meth-igv, echo = FALSE, out.width = "100%", fig.align = "center", fig.cap = "Screenshot from IGV. Bedgraph files are displayed. See below for code to create these file (File Conversion)"}
 knitr::include_graphics("DRAFT_DNA_methylation/igv.png")
@@
 
-*Screenshot from IGV. Bedgraph files are displayed. See below for code to create these file (File Conversion)*
+---
 
@@
 
-```{r meth-summary, echo = FALSE, out.width = "70%", fig.align = "center"}
+```{r meth-summary, echo = FALSE, out.width = "100%", fig.align = "center"}
 knitr::include_graphics("DRAFT_DNA_methylation/bismark-summary-report.png")

### 2023-12-15 DRAFT_DNA_methylation.Rmd
meth - refes in downstream section

@@
 DNA methylation, or addition of a methyl (CH<sub>3</sub>) group to a cytosine base adjacent to a guanine (CpG) is one of the more commonly studies epigenetic mechanisms, partly because of its environmental sensitivity and potential role in phenotypic plasticity.
-For a good review of epigenetic mechanisms including DNA methylation in marine organisms see @doi:10.1146/annurev-marine-010318-095114.
+For a good review of epigenetic mechanisms including DNA methylation in marine organisms see [@doi:10.1146/annurev-marine-010318-095114](#refs).
 
@@
 The principle behind creating bisulfite-treated DNA libraries is that when cytosines are not methylated, then they will be converted to uracil, which ultimately result in thymine nucleotides in the sequence data.
-@https://doi.org/10.1111/1755-0998.13542 offers a comparison of three methods for quantifying DNA methylation at single base-pair resolution using whole genome bisulfite sequencing (WGBS), reduced representation bisulfite sequencing (RRBS), and methyl-CpG binding domain bisulfite sequencing (MBDBS).
+[@https://doi.org/10.1111/1755-0998.13542](#refs) offers a comparison of three methods for quantifying DNA methylation at single base-pair resolution using whole genome bisulfite sequencing (WGBS), reduced representation bisulfite sequencing (RRBS), and methyl-CpG binding domain bisulfite sequencing (MBDBS).
 When aligning bisulfite-converted data to a reference genome the percent methylation at a given CpG loci is determined by examining the ratio of thymines to cytosines.
@@
 
-For a good review of epigenetic mechanisms including DNA methylation in marine organisms see @doi:10.1146/annurev-marine-010318-095114.
-@https://doi.org/10.1111/1755-0998.13542 offers a comparison of three methods for quantifying DNA methylation at single base-pair resolution using whole genome bisulfite sequencing (WGBS), reduced representation bisulfite sequencing (RRBS), and methyl-CpG binding domain bisulfite sequencing (MBDBS).
+For a good review of epigenetic mechanisms including DNA methylation in marine organisms see [@doi:10.1146/annurev-marine-010318-095114](#refs).
+[@https://doi.org/10.1111/1755-0998.13542](#refs) offers a comparison of three methods for quantifying DNA methylation at single base-pair resolution using whole genome bisulfite sequencing (WGBS), reduced representation bisulfite sequencing (RRBS), and methyl-CpG binding domain bisulfite sequencing (MBDBS).
 The tutorial below is based on WGBS, though the general workflow would be consistent.
@@
 
-The tutorial below is based on WGBS from a species of marine intertebrate (*Montipora capitata* coral) examined in @https://doi.org/10.1111/1755-0998.13542, and all raw data can be accessed under NCBI Bioproject PRJNA691891 if you would like to follow the tutorial with this dataset.
+The tutorial below is based on WGBS from a species of marine intertebrate (*Montipora capitata* coral) examined in [@https://doi.org/10.1111/1755-0998.13542](#refs), and all raw data can be accessed under NCBI Bioproject PRJNA691891 if you would like to follow the tutorial with this dataset.
 The general workflow would be consistent for other library preparation and sequencing methods that produce loci-level data.
@@
 
-The software used as part of this tutorial is **Bismark Bisulfite Mapper** [@10.1093/bioinformatics/btr167], which is one of the more widely-used softwares to align methylation data to a reference genome.
-Another commonly-used options include bwa-meth [@pedersen2014fast] and BSMAP [@10.1186/1471-2105-10-232] (see [@10.1093/bib/bbab021] and [@10.1016/j.csbj.2022.08.051] for comparisons of these software for plant and mammal species, respectively).
+The software used as part of this tutorial is **Bismark Bisulfite Mapper** [@10.1093/bioinformatics/btr167](#refs), which is one of the more widely-used softwares to align methylation data to a reference genome.
+Another commonly-used options include bwa-meth [@pedersen2014fast](#refs) and BSMAP [@10.1186/1471-2105-10-232](#refs) (see [@10.1093/bib/bbab021](#refs) and [@10.1016/j.csbj.2022.08.051] for comparisons of these software for plant and mammal species, respectively).
 As with any software it is best to be famililar with the [User Manual](https://felixkrueger.github.io/Bismark/).
 
-Within Bismark, the tutorial uses Bowtie 2 [@10.1038/nmeth.1923] as the alignment software.
+Within Bismark, the tutorial uses Bowtie 2 [@10.1038/nmeth.1923](#refs) as the alignment software.
 Although the Bismark user manual covers the necessary Bowtie 2 parameters, it may be useful to refer to the [Bowtie 2 manual](https://bowtie-bio.sourceforge.net/bowtie2/manual.shtml).
@@
 
-```{r meth-ratio, echo = FALSE, out.width = "70%", fig.align = "center", fig.cap = "Frequency distribution of methylation ratios for CpG dinucleotides in oyster gill tissue. @10.7717/peerj.215"}
+```{r meth-ratio, echo = FALSE, out.width = "70%", fig.align = "center", fig.cap = "Frequency distribution of methylation ratios for CpG dinucleotides in oyster gill tissue.
+[@10.7717/peerj.215](#refs)"}
 knitr::include_graphics("DRAFT_DNA_methylation/meth_ratio.png")
@@
 
--   Gene-level methylation analysis. It is worth considering the utility of converting loci-level resolution data to gene-level resolution data
+-   Integrating with population genetic data. example: [@10.1093/gbe/evad013](#refs)
 
--   Differential loci or region analyses. methylKit is a popular package to conduct this analysis
+-   Gene-level methylation analysis: see [@. It is worth considering the utility of converting loci-level resolution data to gene-level resolution data.
 
--   Feature overlap. Bedtools for quantitative assessment and IGV for visual assessment are valuable for this.
+-   Differential loci methylation and/or region analyses. example: [@10.1186/s12864-022-08781-5](#refs). methylKit is a popular package to conduct this analysis differential methylation anlaysis
+-   Feature overlap. see [@10.1186/s12864-022-08781-5](#refs) Bedtools for quantitative assessment and IGV for visual assessment are valuable for this.
 
-# References
+# References {#refs}

### 2023-12-15 DRAFT_DNA_methylation.Rmd
Update DRAFT_DNA_methylation.Rmd
added vert methylation distribution image
meth - ref formatting

@@
 Also, technologies like Nanopore and PacBio sequencing can directly detect DNA methylation without bisulfite conversion.
+See [@10.1093/g3journal/jkab148](#refs) as example of using Nanopore sequencing.
 As DNA passes through a nanopore or is sequenced by single-molecule real-time (SMRT) technology, changes in electrical current or fluorescence patterns can indicate the presence of methylated bases.
@@
 For WGBS and MBDBS data, it is general practice to hard-trim 10 bp from the beginning and end of each read.
-[RRBS data often needs different trimming specifications](https://rawgit.com/FelixKrueger/Bismark/master/Docs/Bismark_User_Guide.html#ix-notes-about-different-library-types-and-commercial-kits).
+RRBS data often needs different trimming specifications [see Manual for details](https://felixkrueger.github.io/Bismark/)
 
@@
 
--   Gene-level methylation analysis: see [@. It is worth considering the utility of converting loci-level resolution data to gene-level resolution data.
+-   Gene-level methylation analysis: example [@10.1101/2022.06.24.497506](#refs). It is worth considering the utility of converting loci-level resolution data to gene-level resolution data.
 
 -   Differential loci methylation and/or region analyses. example: [@10.1186/s12864-022-08781-5](#refs). methylKit is a popular package to conduct this analysis differential methylation anlaysis
+
 -   Feature overlap. see [@10.1186/s12864-022-08781-5](#refs) Bedtools for quantitative assessment and IGV for visual assessment are valuable for this.

### 2023-12-15 DRAFT_DNA_methylation.Rmd
Update DRAFT_DNA_methylation.Rmd
added caption to vert image

--- 2023-12-15 DRAFT_DNA_methylation.Rmd
@@
 
+```{r meth-vert, echo = FALSE, out.width = "70%", fig.align = "center", fig,cap= "Genome screenshots of methylation quantified in 2kb running windows over the first 3Mb of chromosome 1 zebrafish and mouse, and of the largest scaffold (NW_006890054.1) in elephant shark."}
+knitr::include_graphics("DRAFT_DNA_methylation/vert_methyl_distr.png")
+```
 [ placeholder for Kathryn to add some stuff! ]

### 2023-12-15 DRAFT_DNA_methylation.Rmd
Update DRAFT_DNA_methylation.Rmd
added citation to vert image

--- 2023-12-15 DRAFT_DNA_methylation.Rmd
@@
 
+```{r meth-vert, echo = FALSE, out.width = "70%", fig.align = "center", fig,cap= "Genome screenshots of methylation quantified in 2kb running windows over the first 3Mb of chromosome 1 zebrafish and mouse, and of the largest scaffold (NW_006890054.1) in elephant shark (doi: 10.12688/f1000research.11281.1)"}
+knitr::include_graphics("DRAFT_DNA_methylation/vert_methyl_distr.png")
+```
 [ placeholder for Kathryn to add some stuff! ]

### 2023-12-15 DRAFT_lncRNA.Rmd
started Draft lncRNA discovery page

--- 2023-12-15 DRAFT_DNA_methylation.Rmd
@@
 ---
-title: 'DNA Methylation Assessment'
-subtitle: '*Yaamini Venkataraman*, *Kathryn Greiner-Ferris*, *Sam White*, and *Steven Roberts*'
-bibliography: DRAFT_DNA_methylation/DNA-Methylation.bib
+title: 'Long non-coding RNA Discovery'
+subtitle: '*Steven Roberts* and  *Zach Bengston*'
+bibliography: DRAFT_lncRNA/DRAFT_lncRNA.bib
 ---
@@
 
-## Background
-
-Thanks to advances in sample preparation and sequencing methods, there is an influx of research examining epigenomics in non-model systems.
-Broadly, epigenetics can be defined as changes to gene expression that do not arise from changes in the DNA sequence.
-DNA methylation, or addition of a methyl (CH<sub>3</sub>) group to a cytosine base adjacent to a guanine (CpG) is one of the more commonly studies epigenetic mechanisms, partly because of its environmental sensitivity and potential role in phenotypic plasticity.
-For a good review of epigenetic mechanisms including DNA methylation in marine organisms see [@doi:10.1146/annurev-marine-010318-095114](#refs).
-
-The purpose of this tutorial is to provide an example of aligning bisulfite-treated and enzymatically converted DNA sequence data to an existing genome, and discuss needs for potential downstream applications.
-The principle behind creating bisulfite-treated DNA libraries is that when cytosines are not methylated, then they will be converted to uracil, which ultimately result in thymine nucleotides in the sequence data.
-[@https://doi.org/10.1111/1755-0998.13542](#refs) offers a comparison of three methods for quantifying DNA methylation at single base-pair resolution using whole genome bisulfite sequencing (WGBS), reduced representation bisulfite sequencing (RRBS), and methyl-CpG binding domain bisulfite sequencing (MBDBS).
-When aligning bisulfite-converted data to a reference genome the percent methylation at a given CpG loci is determined by examining the ratio of thymines to cytosines.
-For example if there is a given cytosine adjacent to a guanine (ie CpG locus) with 30% of the reads containing thymine, one would consider this cytosine loci (CpG) to be 70% methylated.
-Since bisulfite treatment can damage DNA, newer library preparation and sequencing methods (ex.
-[NEBNext® Enzymatic Methyl-seq Kit; EM-seq](https://www.neb.com/en-us/products/e7120-nebnext-enzymatic-methyl-seq-kit#Protocols,%20Manuals%20&%20Usage) are starting to gain popularity.
-As EM-Seq produces loci-level data, alignment of EM-Seq data is similar to WGBS.
-Also, technologies like Nanopore and PacBio sequencing can directly detect DNA methylation without bisulfite conversion.
-See [@10.1093/g3journal/jkab148](#refs) as example of using Nanopore sequencing.
-As DNA passes through a nanopore or is sequenced by single-molecule real-time (SMRT) technology, changes in electrical current or fluorescence patterns can indicate the presence of methylated bases.
-
-## Experimental Considerations
-
-### Sample Size
-
-Sample size is a critical factor when conducting DNA methylation analysis, particularly because of the significant variation in methylation patterns both between individuals and across different tissues.
-This variation, which is not yet fully understood, can significantly impact the outcomes and interpretations of methylation studies.
-
-When selecting samples for DNA methylation analysis, the key principle is that larger sample sizes generally provide more reliable and generalizable results.
-A larger number of samples allows for a more accurate representation of the population, accounting for individual and tissue-level differences.
-This is essential for identifying true biological patterns rather than artifacts of small, non-representative samples.
-
-However, the practicality of obtaining a large number of samples must be balanced with resource constraints.
-In this context, a sample size of around eight can be considered a decent starting point.
-This number is often sufficient to begin observing trends and patterns in methylation across individuals and tissues, providing a foundational understanding which can be built upon with further research.
-
-It's crucial to remember that while more samples can enhance the reliability of the results, the quality of the samples and the robustness of the methylation analysis techniques are equally important.
-Therefore, researchers must strive for a balance between quantity and quality in their sample selection to ensure meaningful and accurate insights into DNA methylation.
-
-### Sequencing Depth
-
-### Bisulfite Conversion Efficiency
-
-Bisulfite Conversion Efficiency is important to assess.
-One common step during library preparation us spiking library with Lambda phage DNA.
-Lambda phage DNA lacks cytosine methylation, making it an ideal control for bisulfite treatment.
-Here's how it works and why it's useful:
-
-1.
-**Background on Lambda Phage DNA**: Lambda phage DNA is a widely used control in bisulfite sequencing because it is naturally unmethylated.
-By including this DNA in your sample, you create a known unmethylated reference.
-
-2.
-**Preparation and Spiking**: You prepare the lambda phage DNA and spike it into your DNA samples before bisulfite treatment.
-The proportion of lambda DNA added should be enough to be detected post-sequencing, but not so much that it overwhelms the sample.
-A typical proportion might be around 0.1% to 1% of the total DNA.
-
-3.
-**Bisulfite Treatment and Sequencing**: After spiking, you proceed with bisulfite treatment and sequencing of the entire sample, which now includes your genomic DNA of interest and the spiked lambda DNA.
-
-4.
-**Assessing Conversion Efficiency**: Once sequencing is complete, you analyze the lambda phage DNA sequences.
-Since this DNA is unmethylated, all cytosines should be converted to thymines after bisulfite treatment and subsequent PCR.
-By calculating the percentage of cytosines that were successfully converted to thymines in the lambda DNA, you can determine the bisulfite conversion efficiency.
-An efficient conversion typically results in over 99% of cytosines being converted.
-
-## Tutorial Details
-
-For a good review of epigenetic mechanisms including DNA methylation in marine organisms see [@doi:10.1146/annurev-marine-010318-095114](#refs).
-[@https://doi.org/10.1111/1755-0998.13542](#refs) offers a comparison of three methods for quantifying DNA methylation at single base-pair resolution using whole genome bisulfite sequencing (WGBS), reduced representation bisulfite sequencing (RRBS), and methyl-CpG binding domain bisulfite sequencing (MBDBS).
-The tutorial below is based on WGBS, though the general workflow would be consistent.
-
-The tutorial will cover the following:\
-- Sequence quality assessment\
-- Read alignment\
-- Methylation quantification\
-- File conversions\
-- Potential downstream applications.
-Although this tutorial will not provide code for downstream applications such as differential methylation analysis, it will link to other open-access resources and examples.
-
-The tutorial below is based on WGBS from a species of marine intertebrate (*Montipora capitata* coral) examined in [@https://doi.org/10.1111/1755-0998.13542](#refs), and all raw data can be accessed under NCBI Bioproject PRJNA691891 if you would like to follow the tutorial with this dataset.
-The general workflow would be consistent for other library preparation and sequencing methods that produce loci-level data.
-The tutorial indicates places where special analytical consideration should be given to other data types.
-
-# Sequence Quality
-
-Following trimming one would expect to see FastQC per base sequencing content similar to the plot shown below. This particular data set was trimmed with the following parameters:
-
-```         
-/gscratch/srlab/programs/TrimGalore-0.4.5/trim_galore \
---output_dir /gscratch/scrubbed/strigg/analyses/20200311/WGBS_MBD \
---paired \
---fastqc_args \
---outdir /gscratch/scrubbed/strigg/analyses/20200311/WGBS_MBD/FASTQC \
---threads 28 \
---illumina \
---clip_R1 10 \
---clip_R2 10 \
---three_prime_clip_R1 10 \
---three_prime_clip_R2 10 \
---path_to_cutadapt /gscratch/srlab/programs/miniconda3/bin/cutadapt \
-/gscratch/scrubbed/sr320/froger-raw/00_fastq/Meth17_R1_001.fastq.gz \
-```
-
-```{r meth-bp, echo = FALSE, out.width = "70%", fig.align = "center"}
-knitr::include_graphics("DRAFT_DNA_methylation/meth-bp-chart.png")
-```
-
-For WGBS and MBDBS data, it is general practice to hard-trim 10 bp from the beginning and end of each read.
-RRBS data often needs different trimming specifications [see Manual for details](https://felixkrueger.github.io/Bismark/)
-
-# Read Alignment
-
-## Software Considerations
-
-The software used as part of this tutorial is **Bismark Bisulfite Mapper** [@10.1093/bioinformatics/btr167](#refs), which is one of the more widely-used softwares to align methylation data to a reference genome.
-Another commonly-used options include bwa-meth [@pedersen2014fast](#refs) and BSMAP [@10.1186/1471-2105-10-232](#refs) (see [@10.1093/bib/bbab021](#refs) and [@10.1016/j.csbj.2022.08.051] for comparisons of these software for plant and mammal species, respectively).
-As with any software it is best to be famililar with the [User Manual](https://felixkrueger.github.io/Bismark/).
-
-Within Bismark, the tutorial uses Bowtie 2 [@10.1038/nmeth.1923](#refs) as the alignment software.
-Although the Bismark user manual covers the necessary Bowtie 2 parameters, it may be useful to refer to the [Bowtie 2 manual](https://bowtie-bio.sourceforge.net/bowtie2/manual.shtml).
-
-The content below will provided with the assumption that the reader has read the manual and is meant to serve as guidance based on experience working with marine invertebrates.
-
-## Genome Preparation
-
-The first step in the process in preparing the reference genome by performing an in silico bisulfite conversion to allow Bowtie alignments. Example code:
-
-```         
-# Directories and programs
-bismark_dir="/programs/Bismark-0.21.0"
-bowtie2_dir="/programs/bowtie2-2.3.4.1-linux-x86_64/"
-genome_folder="/where/the/fastafile/lives/"
-
-${bismark_dir}/bismark_genome_preparation \
---verbose \
---parallel 28 \
---path_to_aligner ${bowtie2_dir} \
-${genome_folder}
-```
-
-1.  **Bismark Genome Preparation Command**:
-
-    -   **`${bismark_dir}/bismark_genome_preparation`**: This is the command to run the genome preparation part of Bismark.
-The **`${bismark_dir}`** variable is expanded to the path where Bismark is installed, so the script knows where to find the **`bismark_genome_preparation`** program.
-
-    -   **`--verbose`**: This flag makes the program output more detailed information about what it is doing, which is helpful for debugging or understanding the process.
-
-    -   **`--parallel 28`**: This option tells Bismark to use 28 threads in parallel to speed up the process. You can adjust this based on your computing power.
-
-    -   **`--path_to_aligner ${bowtie2_dir}`**: This specifies the path to the aligner (Bowtie2) that Bismark will use.
-The **`${bowtie2_dir}`** variable is expanded to the path where Bowtie2 is installed.
-
-    -   **`${genome_folder}`**: Finally, this specifies the location of the genome files. The script uses the **`${genome_folder}`** variable, which holds the path to these files.
-
-Bismark will create two individual folders within this directory, one for a C-\>T converted genome and the other one for the G-\>A converted genome.
-After creating C-\>T and G-\>A versions of the genome they will be indexed in parallel usingthe indexer.
-Bismark will create the following subdirectories as outputs.
-
-```         
-./Bisulfite_Genome
-./Bisulfite_Genome/GA_conversion
-./Bisulfite_Genome/CT_conversion
-```
-
-This preparation step only needs to happen once for a given genome.
-
-## Aligning reads
-
-Once the genome is prepared, reads can be aligned. For aligning the trimmed reads to the genome the following code structure is used:
-
-```         
-find ${reads_dir}*_R1_001_val_1.fq.gz \
-| xargs basename -s _R1_001_val_1.fq.gz | xargs -I{} ${bismark_dir}/bismark \
---path_to_bowtie ${bowtie2_dir} \
--genome ${genome_folder} \
--p 4 \
--score_min L,0,-0.6 \
---non_directional \
--1 ${reads_dir}{}_R1_001_val_1.fq.gz \
--2 ${reads_dir}{}_R2_001_val_2.fq.gz \
--o Mcap_tg
-```
-
-This will create BAM files (sequence alignment files) and txt files (mapping reports). The reports contain information such as mapping efficiency.
-
-Specifically this is what the code chunk does:
-
-1.  **Finding Files and Preparing Filenames**:
-
-    -   **`find ${reads_dir}*_R1_001_val_1.fq.gz`**: This command uses the **`find`** utility to search for files in the directory specified by the variable **`reads_dir`**.
-The files it looks for have a naming pattern ending with **`_R1_001_val_1.fq.gz`**, which typically denotes the first file of a pair in paired-end sequencing data.
-
-    -   **`| xargs basename -s _R1_001_val_1.fq.gz`**: The **`xargs`** command takes the list of files found and passes them to **`basename`**, which strips the suffix **`_R1_001_val_1.fq.gz`** from each filename.
-This step extracts the base name of each file, which is used to pair the R1 and R2 files correctly in the following steps.
-
-2.  **Running Bismark for Sequence Alignment**:
-
-    -   **`| xargs -I{} ${bismark_dir}/bismark`**: The **`xargs`** command here is used to pass the base filenames to Bismark for alignment. **`{}`** is a placeholder for each base filename.
-
-    -   **`--path_to_bowtie ${bowtie2_dir}`**: This option specifies the path to the Bowtie 2 aligner, which Bismark uses for the alignment process.
-
-    -   **`-genome ${genome_folder}`**: Specifies the directory where the genome files, prepared in advance, are located.
-
-    -   **`-p 4`**: Sets the number of processing threads to 4, allowing Bismark to perform the alignment in parallel, which can speed up the process.
-This parameter can be changed based on your computing power.
-
-    -   **`-score_min L,0,-0.6`**: This sets the scoring threshold for alignments, impacting the sensitivity and specificity of the alignment.
-The default scoring threshold used by Bismark is L,0,-0.2.
-
-    -   **`--non_directional`**: This flag indicates the sequencing data is non-directional, which is important for correctly interpreting the methylation status in bisulfite sequencing.
-
-    -   **`-1 ${reads_dir}{}_R1_001_val_1.fq.gz -2 ${reads_dir}{}_R2_001_val_2.fq.gz`**: Specifies the files for paired-end reads. R1 and R2 files are paired based on their base filename.
-
-    -   **`-o Mcap_tg`**: Sets the output directory for the alignment results to **`Mcap_tg`**.
-
-### Optimizing alignment sensitivity
-
-The code above uses `score_min L,0,-0.6` as the scoring threshold for alignment sensitivity and specificity.
-This is different than the Bismark default, `score_min L,0,-0.2`.
-Reducing the sensitivity of the alignment can increase the percentage of reads aligned to the reference genome, but of course the quality of the alignment may decrease.
-A good practice is to align a subset of your data to the reference genome using different `score_min` parameters, then comparing mapping efficiency from the mapping reports.
-Generally, you want to choose a parameter that maximizes percent alignment and specificity.
-Example code
-
-```         
-find ${reads_dir}*_R1_001_val_1.fq.gz \
-| xargs basename -s _R1_001_val_1.fq.gz | xargs -I{} ${bismark_dir}/bismark \
---path_to_bowtie ${bowtie2_dir} \
--genome ${genome_folder} \
--p 4 \
-- u 10000 \
--score_min L,0,-0.6 \
---non_directional \
--1 ${reads_dir}{}_R1_001_val_1.fq.gz \
--2 ${reads_dir}{}_R2_001_val_2.fq.gz \
--o Mcap_tg
-```
-
-This is the same code as above, except `-u 10000` is used to run the alignment on a 10,000 bp subset of the data in each input file.
-
-## Deduplication
-
-Files are then deduplicated if they are whole genome bisulfite treated samples.
-This command will deduplicate the Bismark alignment BAM file and remove all reads that align to the the very same position and in the same orientation except for one.
-This step is recommended for whole-genome bisulfite samples, but **should not** be used for reduced representation libraries such as RRBS, amplicon, or target enrichment libraries.
-The [`bismark` manual](https://rawgit.com/FelixKrueger/Bismark/master/Docs/Bismark_User_Guide.html#ix-notes-about-different-library-types-and-commercial-kits) provides additional information.
-
-```         
-find *.bam | \
-xargs basename -s .bam | \
-xargs -I{} ${bismark_dir}/deduplicate_bismark \
---bam \
---paired \
-{}.bam
-```
-
-# Methylation Quantification
-
-## What to expect
-
-DNA methylation at CpG locus can vary to 0-100%.
-How methylated CpGs are across a genome can vary WIDELY depending on your focal taxa.
-Generally, invertebrate genomes are predominantly unmethylated, with smaller regions of higher methylation in a mosaic pattern.
-In some marine invertebrates we see \~15% methylation of CpGs (defined as \> 75% methylated) across the genome with most CpGs unmethylated.
-In these taxa methylation usually occurs in gene bodies.
-
-The figure below shows the methylation ratio of CpGs across the Pacific oyster (*Crassostrea gigas*) genome, as well as the location of methylation across the genome:
-
-```{r meth-ratio, echo = FALSE, out.width = "70%", fig.align = "center", fig.cap = "Frequency distribution of methylation ratios for CpG dinucleotides in oyster gill tissue.
-[@10.7717/peerj.215](#refs)"}
-knitr::include_graphics("DRAFT_DNA_methylation/meth_ratio.png")
-```
-
-```{r meth-igv, echo = FALSE, out.width = "100%", fig.align = "center", fig.cap = "Screenshot from IGV. Bedgraph files are displayed. See below for code to create these file (File Conversion)"}
-knitr::include_graphics("DRAFT_DNA_methylation/igv.png")
-```
-
----
-
-Vertebrate genomes, on the other hand, have higher overall levels of methylation, with methylation occuring in CpG islands.
-
-[ placeholder for Kathryn to add some stuff! ]
-
-## Methylation extraction
-
-Methylation levels are then extracted using the `bismark_methylation_extractor`. For example:
-
-```         
-${bismark_dir}/bismark_methylation_extractor \
---bedGraph \
---counts \
---comprehensive \
---merge_non_CpG \
---multicore 28 \
---buffer_size 75% \
-*deduplicated.bam
-```
-
-This particular command is used to extract and process methylation data from BAM files (generated after sequencing data alignment). Let's break down the options:
-
-1.  **Options**:
-
-    -   **`--bedGraph`**: Generates a BedGraph file output, which is a format that allows visualization of the methylation levels across the genome in genome browsers.
-
-    -   **`--counts`**: Outputs the count of methylated and unmethylated calls for each cytosine in the genome.
-
-    -   **`--comprehensive`**: This option makes the methylation extractor process all cytosines in the genome, not just those in a CpG context.
-It combines all three contexts (CpG, CHG, and CHH) into one output file.
-
-    -   **`--merge_non_CpG`**: Merges the methylation information for CHG and CHH contexts (non-CpG) into a single file.
-
-    -   **`--multicore 28`**: Specifies that the program should use 28 cores to process the data in parallel, which can significantly speed up the analysis.
-This can be adjusted depending on your computing power.
-
-    -   **`--buffer_size 75%`**: This option sets the buffer size for reading in BAM files to 75% of the available system memory.
-It helps optimize the performance of the tool by adjusting how much memory is used for processing.
-This can be adjusted depending on your computing power.
-
-    -   **`*deduplicated.bam`**: This is the input file pattern for the command.
-It indicates that the methylation extractor should process all BAM files in the current directory that have names ending with "deduplicated.bam".
-These files are typically the output of a deduplication process applied to aligned bisulfite sequencing data.
-
-This will create a file with the suffix `deduplicated.bismark.cov.gz` (example shown below in uncompressed format):
-
-```         
-NC_035784.1 141 141 37.5    3   5
-NC_035784.1 142 142 100 2   0
-NC_035784.1 155 155 70  7   3
-NC_035784.1 156 156 100 2   0
-NC_035784.1 291 291 0   0   2
-NC_035784.1 292 292 0   0   3
-NC_035784.1 313 313 0   0   1
-NC_035784.1 314 314 66.6666666666667    2   1
-NC_035784.1 470 470 66.6666666666667    4   2
-NC_035784.1 611 611 0   0   4
-```
-
-column organization of the file
-
-```         
-<chromosome> <start position> <end position> <methylation percentage> <count methylated> <count unmethylated>
-```
-
-The resultant bedGraph file contains information from *unmerged* strands, meaning reads aligning to forward and reverse strands in the same position are not combined.
-It is important to merge the reads from the forward and reserve strands for more accurate methylation quantification.
-A genome-wide cytosine report and merged bedGraph file are generated (including merged bedGraph file) from the `deduplicated.bismark.cov.gz` generated above in this example code:
-
-```         
-find *deduplicated.bismark.cov.gz \
-| xargs basename -s _trimmed_bismark_bt2.deduplicated.bismark.cov.gz \
-| xargs -I{} ${bismark_dir}/coverage2cytosine \
---genome_folder ${genome_folder} \
--o {} \
---merge_CpG \
---zero_based \
-{}_trimmed_bismark_bt2.deduplicated.bismark.cov.gz
-```
-
-The script converts coverage files into a format that details cytosine methylation levels across the genome. Here's a breakdown of each step:
-
-1.  **Finding Files**:
-
-    -   **`find *deduplicated.bismark.cov.gz`**: This command is used to search for all files in the current directory that end with **`deduplicated.bismark.cov.gz`**.
-These files are coverage files produced by Bismark, typically after aligning bisulfite-treated DNA sequences.
-They contain information about methylation at different sites in the genome.
-
-2.  **Extracting Base File Names**:
-
-    -   **`| xargs basename -s _trimmed_bismark_bt2.deduplicated.bismark.cov.gz`**: This pipeline uses **`xargs`** to pass each found file to **`basename`**, which strips the suffix **`_trimmed_bismark_bt2.deduplicated.bismark.cov.gz`** from each file name.
-This step extracts the base names of the files, which will be used to generate output filenames in the following steps.
-
-3.  **Converting Coverage to Cytosine Methylation Levels**:
-
-    -   **`| xargs -I{} ${bismark_dir}/coverage2cytosine`**: This pipeline uses **`xargs`** again to pass each base filename to the **`coverage2cytosine`** tool in the Bismark suite.
-The placeholder **`{}`** is replaced by the base filename.
-
-    -   **`--genome_folder ${genome_folder}`**: This option specifies the location of the genome folder.
-The **`${genome_folder}`** variable holds the path to this folder, which contains reference genome data necessary for the analysis.
-
-    -   **`-o {}`**: The **`-o`** option specifies the output file name, using the base filename **`{}`** obtained from the previous steps.
-
-    -   **`--merge_CpG`**: This flag instructs the program to merge the methylation information of cytosines in CpG context, providing a single methylation value for each CpG site.
-
-    -   **`--zero_based`**: This option indicates that the output file should use zero-based coordinates, which is a common format in bioinformatics where the first base of a sequence is numbered 0 instead of 1.
-
-    -   **`{}_trimmed_bismark_bt2.deduplicated.bismark.cov.gz`**: Finally, this specifies the input file for each iteration, where **`{}`** is replaced by the base filename obtained earlier.
-These are the coverage files previously identified by the **`find`** command.
-
-Output file format `.CpG_report.merged_CpG_evidence.cov` (merged bedGraph file)
-
-```         
-NC_035785.1 217 219 100.000000  17  0
-NC_035785.1 523 525 87.500000   7   1
-NC_035785.1 556 558 50.000000   5   5
-NC_035785.1 727 729 100.000000  16  0
-NC_035785.1 1330    1332    0.000000    0   2
-NC_035785.1 1403    1405    0.000000    0   2
-NC_035785.1 1494    1496    66.666667   2   1
-NC_035785.1 1747    1749    100.000000  8   0
-NC_035785.1 2024    2026    100.000000  24  0
-NC_035785.1 2054    2056    93.333333   14  1
-```
-
-column organization of the file
-
-```         
-<chromosome> <start position> <end position> <methylation percentage> <count methylated> <count unmethylated>
-```
-
-# Summary Report
-
-Many steps of the Bismark workflow produce reports (.txt files) with useful information (ex.
-mapping efficiency, percent cytosine methylation, etc.).
-These files can be collated into processing reports for each individual sample, which can then be further collated into an [HTML summary report](https://gannet.fish.washington.edu/seashell/bu-mox/scrubbed/032120-fds/bismark_summary_report.html):
-
-```         
-#HTML Processing Report
-
-${bismark_dir}/bismark2report
-
-#Summary Report
-
-${bismark_dir}/bismark2summary
-```
-
-These commands should be run in the same directory where all output files are stored.
-
-```{r meth-summary, echo = FALSE, out.width = "100%", fig.align = "center"}
-knitr::include_graphics("DRAFT_DNA_methylation/bismark-summary-report.png")
-```
-
-# File Conversions
-
-From here it could be useful to do some file reorganizaion to obtain bedGraph or tab files for downstream analysis (eg `bedtools`, GLMs).
-Specifically, it may be important to filter the CpG methylation data to a specific coverage threshold (ex.
-5x or 10x).
-The more read coverage a CpG has, the higher the confidence in methylation quantification.
-
-Creating bedGraphs post-merge at a specific coverage threshold (ex. 10x or 5x):
-
-```         
-for f in *merged_CpG_evidence.cov
-do
-  STEM=$(basename "${f}" .CpG_report.merged_CpG_evidence.cov)
-  cat "${f}" | awk -F $'\t' 'BEGIN {OFS = FS} {if ($5+$6 >= 5) {print $1, $2, $3, $4}}' \
-  > "${STEM}"_5x.bedgraph
-done
-```
-
-Creating tab files with raw count data for GLMs at 10x and 5x coverage:
-
-```         
-for f in *merged_CpG_evidence.cov
-do
-  STEM=$(basename "${f}" .CpG_report.merged_CpG_evidence.cov)
-  cat "${f}" | awk -F $'\t' 'BEGIN {OFS = FS} {if ($5+$6 >= 5) {print $1, $2, $3, $4, $5, $6}}' \
-  > "${STEM}"_5x.tab
-done
-```
-
-It is sometimes useful to sort BAM files for downstream analysis (eg methylkit, IGV)
-
-```         
-# Sort files for methylkit and IGV
-find *.bam | \
-xargs basename -s .bam | \
-xargs -I{} ${samtools} \
-sort --threads 28 {}.bam \
--o {}.sorted.bam
-
-# Index sorted files for IGV
-find *.sorted.bam | \
-xargs basename -s .sorted.bam | \
-xargs -I{} ${samtools} \
-index -@ 28 {}.sorted.bam
-```
-
-# Next Steps
-
-So you've aligned your trimmed reads to your reference genome and have methylation data for CpGs. Now what?
-
-## Removing C to T (C-\>T) SNPs
-
-Removing cytosine (C) to thymine (T) single nucleotide polymorphisms (SNPs) in bisulfite sequencing analysis could be something to consider for a few reasons:
-
-1.
-**Bisulfite Conversion Process**: In bisulfite sequencing, unmethylated cytosines are converted to uracil, which is then read as thymine in sequencing.
-This conversion is key to identifying methylated cytosines.
-However, if a genomic DNA sequence naturally has a C to T (C-\>T) SNP, it can be falsely interpreted as a converted cytosine, leading to incorrect conclusions about methylation.
-
-2.
-**Reduction of False Positives**: By removing C-\>T SNPs, one can reduce false positives in methylation analysis.
-Without removal, these SNPs might be misinterpreted as unmethylated cytosines that have undergone conversion, skewing the data towards an apparent increase in methylation.
-
-3.
-**Accurate Methylation Mapping**: For precise mapping of methylation patterns, it's crucial to distinguish between true methylation signals and genetic variation (SNPs).
-Removing C-\>T SNPs allows for a more accurate representation of the methylation status of the genome.
-
-4.
-**Analysis Consistency**: In comparative studies, where multiple samples or conditions are analyzed, removing C-\>T SNPs ensures consistency across samples.
-This is especially important in cases where SNP profiles may vary between samples, which could otherwise lead to inconsistent methylation data.
-
-By eliminating these potential sources of error, researchers can have more confidence in their interpretation of the bisulfite sequencing data, leading to more reliable and valid conclusions about DNA methylation patterns.
-
-One way to do this is using BS-Snper (@10.1093/bioinformatics/btv507).
-<https://github.com/hellbelly/BS-Snper>.
-This program finds SNPs in Bisulfite treated alignment files.
-The bash code below represents steps to identify SNPS and create a list of all C-\>T SNPS.
-
-```         
-samtools merge \
-Pg_merged.bam \
-*.sorted.bam
-
-perl /gscratch/srlab/programs/BS-Snper-master/BS-Snper.pl \
-Pg_merged.bam \
---fa Panopea-generosa-v1.0.fa \
---output snp.candidate.out \
---methcg meth.cg \
---methchg meth.chg \
---methchh meth.chh \
---minhetfreq 0.1 \
---minhomfreq 0.85 \
---minquali 15 \
---mincover 10 \
---maxcover 1000 \
---minread2 2 \
---errorate 0.02 \
---mapvalue 20 \
->SNP.vcf 2>SNP.log
-
-grep $'C\tG' output/SNP.vcf > output/CT-SNP.vcf
-wc -l output/CT-SNP.vcf
-```
-
-Let's break down the command and its arguments:
-
-1.
-**`perl /gscratch/srlab/programs/BS-Snper-master/BS-Snper.pl`**: This part of the command calls the Perl interpreter to execute the **`BS-Snper.pl`** script.
-The script is located in the **`/gscratch/srlab/programs/BS-Snper-master/`** directory.
-
-2.  **`Pg_merged.bam`**: This is the input file for the script, likely a BAM file (Binary Alignment/Map format) which contains alignment information of bisulfite-treated sequencing reads.
-
-3.  **`--fa Panopea-generosa-v1.0.fa`**: This option specifies the reference genome file in FASTA format against which the BAM file will be compared.
-
-4.  **`--output snp.candidate.out`**: This argument sets the name of the output file where the SNP candidates will be written.
-
-5.
-**`--methcg meth.cg`, `--methchg meth.chg`, `--methchh meth.chh`**: These options specify the output files for methylation analysis for CG, CHG, and CHH contexts, respectively (where H can be A, C, or T).
-
-6.  **`--minhetfreq 0.1`**: Sets the minimum frequency for heterozygous bases to be considered in SNP calling (here, 10%).
-
-7.  **`--minhomfreq 0.85`**: Sets the minimum frequency for homozygous bases to be considered in SNP calling (here, 85%).
-
-8.  **`--minquali 15`**: Sets the minimum base quality score.
-
-9.  **`--mincover 10`, `--maxcover 1000`**: These set the minimum and maximum coverage (number of reads) for a position to be considered in the analysis.
-
-10.
-**`--minread2 2`**: The minimum number of reads required to support a SNP candidate.
-It could be useful to use the same values for the minimum read threshold for SNP identification and for minimum coverage in the methylation data.
-
-11. **`--errorate 0.02`**: Sets the error rate of sequencing (here, 2%).
-
-12. **`--mapvalue 20`**: Specifies the mapping quality value, a threshold for considering reads in the analysis.
-
-13. **`>SNP.vcf`**: Redirects the standard output to a file named **`SNP.vcf`**. This file will contain the SNP calls in VCF (Variant Call Format) format.
-
-14. **`2>SNP.log`**: Redirects the standard error output to a file named **`SNP.log`**. This file will contain error messages and possibly additional log information.
-
-The resulting could be used to mask the genome and the entire process described above could be repeated (ie Genome Preparation, Alignment, Quantification).
-An alternative and common practice is to remove cytosines from downstream analysis by removing loci in intermediate files (ie tab-delimited data files).
-Below is an example using `antijoin` to remove loci that correspond to CT SNPs from 5x coverage methyation quantification files.
-
-```{r, eval=FALSE}
-# Read in CT SNP file
-ct <- read.csv("../output/CT-SNP.vcf", header = FALSE, sep = "\t") %>%
-  mutate(loci = paste0(V1, "_", V2))
-
-# 1. List all files with _5x.tab suffix
-files <- list.files(path = "../data/", pattern = "_5x.tab$", full.names = TRUE)
-
-# 2. Iterate over each file
-for(file in files) {
-
-  # Extract base filename without the directory for naming purposes
-  base_name <- basename(file)
-
-  # Read the file
-  data <- read.csv(file, header = FALSE, sep = "\t")
-
-  # Modify the data
-  modified_data <- data %>%
-    mutate(loci = paste0(V1, "_", V2)) %>%
-    anti_join(ct, by = "loci") %>%
-    select(-loci)
-
-  # Write the modified data to an output file
-  output_file <- paste0("../output/f", base_name)
-  write.table(modified_data, file = output_file, sep = "\t", row.names = FALSE, quote = FALSE, col.names = FALSE)
-}
-```
-
-Let's break down what each part of the script does:
-
-1.  **Reading in a CT SNP file**:
-
-    -   **`ct <- read.csv("../output/CT-SNP.vcf", header = FALSE, sep = "\t")`**: This line reads a file named **`CT-SNP.vcf`** from the specified directory.
-The file is expected to be in a tab-separated format (**`sep = "\t"`**), and it doesn't have a header row (**`header = FALSE`**).
-
-    -   **`mutate(loci = paste0(V1, "_", V2))`**: The **`mutate`** function from the **`dplyr`** package is used to add a new column (**`loci`**) to the **`ct`** data frame.
-This column is created by concatenating the values in the first (**`V1`**) and second columns (**`V2`**) of each row, separated by an underscore.
-
-2.  **Listing files with a specific suffix**:
-
-    -   **`files <- list.files(path = "../data/", pattern = "_5x.tab$", full.names = TRUE)`**: This line generates a list of file paths in the **`../data/`** directory.
-It only includes files that end with **`_5x.tab`** (as indicated by the **`pattern`** argument).
-
-3.  **Processing each file**:
-
-    -   The **`for(file in files)`** loop iterates over each file found in the previous step.
-
-        -   **`base_name <- basename(file)`**: Extracts the file name without the directory path.
-
-        -   **`data <- read.csv(file, header = FALSE, sep = "\t")`**: Reads the current file in the loop. Similar to the first **`read.csv`** call, it expects a tab-separated file with no header row.
-
-        -   The **`mutate`** and **`anti_join`** functions are used to modify **`data`**:
-
-            -   **`mutate(loci = paste0(V1, "_", V2))`**: Adds a **`loci`** column similar to the **`ct`** data frame.
-
-            -   **`anti_join(ct, by = "loci")`**: Removes rows from **`data`** that have matching **`loci`** values in the **`ct`** data frame.
-
-            -   **`select(-loci)`**: Removes the **`loci`** column from the data.
-
-        -   The modified data is then written to an output file:
-
-            -   **`output_file <- paste0("../output/f", base_name)`**: Constructs the output file path.
-
-            -   **`write.table(modified_data, file = output_file, sep = "\t", row.names = FALSE, quote = FALSE, col.names = FALSE)`**: Writes the **`modified_data`** to the specified file.
-The output is tab-separated, and it doesn't include row names, quotes, or column names.
-
-## Downstream application
-
--   Integrating with population genetic data. example: [@10.1093/gbe/evad013](#refs)
-
--   Gene-level methylation analysis: example [@10.1101/2022.06.24.497506](#refs). It is worth considering the utility of converting loci-level resolution data to gene-level resolution data.
-
--   Differential loci methylation and/or region analyses. example: [@10.1186/s12864-022-08781-5](#refs). methylKit is a popular package to conduct this analysis differential methylation anlaysis
-
--   Feature overlap. see [@10.1186/s12864-022-08781-5](#refs) Bedtools for quantitative assessment and IGV for visual assessment are valuable for this.
-
 # References {#refs}

### 2023-12-15 DRAFT_DNA_methylation.Rmd
meth -adding references

--- 2023-12-15 DRAFT_DNA_methylation.Rmd
@@
 
+```{r meth-vert, echo = FALSE, out.width = "70%", fig.align = "center"}
+knitr::include_graphics("DRAFT_DNA_methylation/vert_methyl_distr.png")
+```
 [ placeholder for Kathryn to add some stuff! ]
@@
 
--   Feature overlap. see [@10.1186/s12864-022-08781-5](#refs) Bedtools for quantitative assessment and IGV for visual assessment are valuable for this.
+-   Feature overlap. example: [@10.3389/fmars.2020.00225](#refs) Bedtools is excellent for quantitative assessment of feature overlap and IGV is good for visual assessment.
 

### 2023-12-15 DRAFT_DNA_methylation.Rmd

@@
 
-```{r meth-vert, echo = FALSE, out.width = "70%", fig.align = "center"}
+```{r meth-vert, echo = FALSE, out.width = "70%", fig.align = "center", fig,cap= "Genome screenshots of methylation quantified in 2kb running windows over the first 3Mb of chromosome 1 zebrafish and mouse, and of the largest scaffold (NW_006890054.1) in elephant shark (doi: 10.12688/f1000research.11281.1)"}
 knitr::include_graphics("DRAFT_DNA_methylation/vert_methyl_distr.png")

### 2023-12-15 DRAFT_DNA_methylation.Rmd
Update DRAFT_DNA_methylation.Rmd
updated info on vert methylation

--- 2023-12-15 DRAFT_DNA_methylation.Rmd
@@
 Vertebrate genomes, on the other hand, have higher overall levels of methylation, with methylation occuring in CpG islands.
+An example analysis of methylation at individual CG dinucleotides in the elephant shark, mice, and zebra fish is shown, below.
+Results reavel comparable levels of global methyaltion patterns with the majority of sites being highly methylated (≥ 80%), and fewer than one tenth being unmethylated (doi: 10.12688/f1000research.11281.1).
 
-```{r meth-vert, echo = FALSE, out.width = "70%", fig.align = "center"}
+```{r meth-vert, echo = FALSE, out.width = "70%", fig.align = "center", fig,cap= "Genome screenshots of methylation quantified in 2kb running windows over the first 3Mb of chromosome 1 zebrafish and mouse, and of the largest scaffold (NW_006890054.1) in elephant shark (doi: 10.12688/f1000research.11281.1)"}
 knitr::include_graphics("DRAFT_DNA_methylation/vert_methyl_distr.png")
 ```
-[ placeholder for Kathryn to add some stuff! ]
 

### 2023-12-15 DRAFT_lncRNA.Rmd
meeth - hpc comments

@@
 opts_chunk$set(fig.width = 10,
+               eval = FALSE,         # Evaluate code chunks
                fig.height = 5,
@@
 
+## Notes on running code. This workflow is written such that a user an fully reproduce the code. Th
+
 # References {#refs}

### 2023-12-15 DRAFT_DNA_methylation.Rmd
meeth - hpc comments
fix legend

@@
 DNA methylation, or addition of a methyl (CH<sub>3</sub>) group to a cytosine base adjacent to a guanine (CpG) is one of the more commonly studies epigenetic mechanisms, partly because of its environmental sensitivity and potential role in phenotypic plasticity.
-For a good review of epigenetic mechanisms including DNA methylation in marine organisms see [@doi:10.1146/annurev-marine-010318-095114](#refs).
+For a good review of epigenetic mechanisms including DNA methylation in marine organisms see [\@doi:10.1146/annurev-marine-010318-095114](#refs).
 
@@
 The principle behind creating bisulfite-treated DNA libraries is that when cytosines are not methylated, then they will be converted to uracil, which ultimately result in thymine nucleotides in the sequence data.
-[@https://doi.org/10.1111/1755-0998.13542](#refs) offers a comparison of three methods for quantifying DNA methylation at single base-pair resolution using whole genome bisulfite sequencing (WGBS), reduced representation bisulfite sequencing (RRBS), and methyl-CpG binding domain bisulfite sequencing (MBDBS).
+[\@https://doi.org/10.1111/1755-0998.13542](#refs) offers a comparison of three methods for quantifying DNA methylation at single base-pair resolution using whole genome bisulfite sequencing (WGBS), reduced representation bisulfite sequencing (RRBS), and methyl-CpG binding domain bisulfite sequencing (MBDBS).
 When aligning bisulfite-converted data to a reference genome the percent methylation at a given CpG loci is determined by examining the ratio of thymines to cytosines.
@@
 Also, technologies like Nanopore and PacBio sequencing can directly detect DNA methylation without bisulfite conversion.
-See [@10.1093/g3journal/jkab148](#refs) as example of using Nanopore sequencing.
+See [\@10.1093/g3journal/jkab148](#refs) as example of using Nanopore sequencing.
 As DNA passes through a nanopore or is sequenced by single-molecule real-time (SMRT) technology, changes in electrical current or fluorescence patterns can indicate the presence of methylated bases.
@@
 
-For a good review of epigenetic mechanisms including DNA methylation in marine organisms see [@doi:10.1146/annurev-marine-010318-095114](#refs).
-[@https://doi.org/10.1111/1755-0998.13542](#refs) offers a comparison of three methods for quantifying DNA methylation at single base-pair resolution using whole genome bisulfite sequencing (WGBS), reduced representation bisulfite sequencing (RRBS), and methyl-CpG binding domain bisulfite sequencing (MBDBS).
+For a good review of epigenetic mechanisms including DNA methylation in marine organisms see [\@doi:10.1146/annurev-marine-010318-095114](#refs).
+[\@https://doi.org/10.1111/1755-0998.13542](#refs) offers a comparison of three methods for quantifying DNA methylation at single base-pair resolution using whole genome bisulfite sequencing (WGBS), reduced representation bisulfite sequencing (RRBS), and methyl-CpG binding domain bisulfite sequencing (MBDBS).
 The tutorial below is based on WGBS, though the general workflow would be consistent.
@@
 
-The tutorial below is based on WGBS from a species of marine intertebrate (*Montipora capitata* coral) examined in [@https://doi.org/10.1111/1755-0998.13542](#refs), and all raw data can be accessed under NCBI Bioproject PRJNA691891 if you would like to follow the tutorial with this dataset.
+The tutorial below is based on WGBS from a species of marine intertebrate (*Montipora capitata* coral) examined in [\@https://doi.org/10.1111/1755-0998.13542](#refs), and all raw data can be accessed under NCBI Bioproject PRJNA691891 if you would like to follow the tutorial with this dataset.
 The general workflow would be consistent for other library preparation and sequencing methods that produce loci-level data.
 The tutorial indicates places where special analytical consideration should be given to other data types.
+Given the size of the data files and the computational resources necessary this is not developed to be actionable and reproducible.
+For context this process outlined below took weeks on a university HPC system.
 
@@
 
-The software used as part of this tutorial is **Bismark Bisulfite Mapper** [@10.1093/bioinformatics/btr167](#refs), which is one of the more widely-used softwares to align methylation data to a reference genome.
-Another commonly-used options include bwa-meth [@pedersen2014fast](#refs) and BSMAP [@10.1186/1471-2105-10-232](#refs) (see [@10.1093/bib/bbab021](#refs) and [@10.1016/j.csbj.2022.08.051] for comparisons of these software for plant and mammal species, respectively).
+The software used as part of this tutorial is **Bismark Bisulfite Mapper** [\@10.1093/bioinformatics/btr167](#refs), which is one of the more widely-used softwares to align methylation data to a reference genome.
+Another commonly-used options include bwa-meth [\@pedersen2014fast](#refs) and BSMAP [\@10.1186/1471-2105-10-232](#refs) (see [\@10.1093/bib/bbab021](#refs) and [@10.1016/j.csbj.2022.08.051] for comparisons of these software for plant and mammal species, respectively).
 As with any software it is best to be famililar with the [User Manual](https://felixkrueger.github.io/Bismark/).
 
-Within Bismark, the tutorial uses Bowtie 2 [@10.1038/nmeth.1923](#refs) as the alignment software.
+Within Bismark, the tutorial uses Bowtie 2 [\@10.1038/nmeth.1923](#refs) as the alignment software.
 Although the Bismark user manual covers the necessary Bowtie 2 parameters, it may be useful to refer to the [Bowtie 2 manual](https://bowtie-bio.sourceforge.net/bowtie2/manual.shtml).
@@
 
----
+------------------------------------------------------------------------
 
@@
 
-```{r meth-vert, echo = FALSE, out.width = "70%", fig.align = "center", fig,cap= "Genome screenshots of methylation quantified in 2kb running windows over the first 3Mb of chromosome 1 zebrafish and mouse, and of the largest scaffold (NW_006890054.1) in elephant shark (doi: 10.12688/f1000research.11281.1)"}
+```{r meth-vert, echo = FALSE, out.width = "70%", fig.align = "center", fig.cap= "Genome screenshots of methylation quantified in 2kb running windows over the first 3Mb of chromosome 1 zebrafish and mouse, and of the largest scaffold (NW_006890054.1) in elephant shark (doi: 10.12688/f1000research.11281.1)"}
 knitr::include_graphics("DRAFT_DNA_methylation/vert_methyl_distr.png")
@@
 
--   Integrating with population genetic data. example: [@10.1093/gbe/evad013](#refs)
+-   Integrating with population genetic data. example: [\@10.1093/gbe/evad013](#refs)
 
--   Gene-level methylation analysis: example [@10.1101/2022.06.24.497506](#refs). It is worth considering the utility of converting loci-level resolution data to gene-level resolution data.
+-   Gene-level methylation analysis: example [\@10.1101/2022.06.24.497506](#refs). It is worth considering the utility of converting loci-level resolution data to gene-level resolution data.
 
--   Differential loci methylation and/or region analyses. example: [@10.1186/s12864-022-08781-5](#refs). methylKit is a popular package to conduct this analysis differential methylation anlaysis
+-   Differential loci methylation and/or region analyses. example: [\@10.1186/s12864-022-08781-5](#refs). methylKit is a popular package to conduct this analysis differential methylation anlaysis
 
--   Feature overlap. example: [@10.3389/fmars.2020.00225](#refs) Bedtools is excellent for quantitative assessment of feature overlap and IGV is good for visual assessment.
+-   Feature overlap. example: [\@10.3389/fmars.2020.00225](#refs) Bedtools is excellent for quantitative assessment of feature overlap and IGV is good for visual assessment.
 

### 2023-12-15 DRAFT_DNA_methylation.Rmd
Update DRAFT_DNA_methylation.Rmd

--- 2023-12-15 DRAFT_DNA_methylation.Rmd
@@
 Vertebrate genomes, on the other hand, have higher overall levels of methylation, with methylation occuring in CpG islands.
-An example analysis of methylation at individual CG dinucleotides in the elephant shark, mice, and zebra fish is shown, below.
+An example analysis of methylation at individual CG dinucleotides in an elephant shark, mouse, and zebra fish is shown, below.
 Results reavel comparable levels of global methyaltion patterns with the majority of sites being highly methylated (≥ 80%), and fewer than one tenth being unmethylated (doi: 10.12688/f1000research.11281.1).
 
-```{r meth-vert, echo = FALSE, out.width = "70%", fig.align = "center", fig,cap= "Genome screenshots of methylation quantified in 2kb running windows over the first 3Mb of chromosome 1 zebrafish and mouse, and of the largest scaffold (NW_006890054.1) in elephant shark (doi: 10.12688/f1000research.11281.1)"}
+```{r meth-vert, echo = FALSE, out.width = "70%", fig.align = "center", fig.cap= "Genome screenshots of methylation quantified in 2kb running windows over the first 3Mb of chromosome 1 zebrafish and mouse, and of the largest scaffold (NW_006890054.1) in elephant shark (doi: 10.12688/f1000research.11281.1)"}
 knitr::include_graphics("DRAFT_DNA_methylation/vert_methyl_distr.png")

### 2023-12-15 DRAFT_DNA_methylation.Rmd
fixes reference formatting

@@
 The tutorial indicates places where special analytical consideration should be given to other data types.
+Given the size of the data files and the computational resources necessary this is not developed to be actionable and reproducible.
+For context this process outlined below took weeks on a university HPC system.
 
@@
 
----
+------------------------------------------------------------------------
 
 Vertebrate genomes, on the other hand, have higher overall levels of methylation, with methylation occuring in CpG islands.
-An example analysis of methylation at individual CG dinucleotides in an elephant shark, mouse, and zebra fish is shown, below.
+An example analysis of methylation at individual CG dinucleotides in the elephant shark, mice, and zebra fish is shown, below.
 Results reavel comparable levels of global methyaltion patterns with the majority of sites being highly methylated (≥ 80%), and fewer than one tenth being unmethylated (doi: 10.12688/f1000research.11281.1).

### 2023-12-15 DRAFT_QTL_tutorial.Rmd
progress on QTL page

@@
 ---
-
-Initial publication year: 2023 <br>
-[How to cite](https://marineomics.github.io/#How_to_Cite)
 
@@
 
-*I owe most of this code and most of what I know about QTL analysis to Dr. Evangeline Ballerini.
+*All of this code and most of what I know about QTL analysis to [Prof. Evangeline Ballerini](https://www.ballerinilab.com/). Add something about Nathan. Any mistakes or lack of clarity are mine.
 
@@
 
-## Tutorial
+# Tutorial
+
 This tutorial will walk you through the essentials of the particular analysis that I did, but my hope is that it will be useful even for those who take different approaches.
@@
 
-I used whole genome skim sequencing due to the ease of a multiplex library prep, making costs and wet lab time comparable to reduced representation methods.
+## Genetic map construction
+The first step in QTL mapping is making a genetic map, where we have the determined the genotype of each F2 within recombination windows.
+
+There are many different approaches for doing this, but below I describe our lab's method of using whole genome skim sequencing and custom scripts.
+I used a multiplexed library prep kit (formerly iGenomx, now Twist 96-plex kit), making costs and wet lab time comparable to reduced representation methods.
 
@@
 
-### We're going to start by defining all of our filepaths, variable names, and tools
+### defining all of our filepaths, variable names, and tools
 ```{bash, eval = F}
@@
 reference="/Volumes/HD/genomes.reference/Aquilegia/V3/sequences/Aquilegia_coerulea.main_genome.scaffolds.fasta"
-f1bam="$bamdir/f2.merged.bam"
+f1bam="$genodir/f2.merged.bam"
 pabam="/Volumes/HD/projects/genomes.species/V3/alignments/jonesii.sorted.bam"
@@
 
-### To begin our genotyping, let's determine the useful sites for genotyping the F2s
+### determining the useful sites for genotyping the F2s
 ####  Generate a genotype depth sequence file for the F1
@@
 # make a list file with all of the file paths of our .bams for samtools to merge
-ls -d -1 $bamdir/*.* > $bamdir/f2.paths.list
+ls -d -1 $bamdir/*.* | grep -v bai > $infodir/f2.paths.list
 
 # merge bam files
-samtools merge $bamdir/f2.merged.bam -b $bamdir/f2.paths.list
+samtools merge $genodir/f2.merged.bam -b $genodir/f2.paths.list
 
 # index bam files
-samtools index $bamdir/f2.merged.bam
+samtools index $genodir/f2.merged.bam
 ```
@@
     * snp.list is our final list of usable SNPs, of which there were ~1,000,000 in this case
-    * Now we'll call genotypes for both the F1 and Pa at these sites using a combination of [samtools mpileup](http://www.htslib.org/doc/samtools-mpileup.html), [bcftools call](https://samtools.github.io/bcftools/bcftools.html#call), and [bcftools query](https://samtools.github.io/bcftools/bcftools.html#query).
+    * Now we'll call genotypes for Pa at these sites using a combination of [samtools mpileup](http://www.htslib.org/doc/samtools-mpileup.html), [bcftools call](https://samtools.github.io/bcftools/bcftools.html#call), and [bcftools query](https://samtools.github.io/bcftools/bcftools.html#query).
 This is a simple text file with 4 columns:
-      * chromosome, position, reference allele, and the alternate allele:
+      * chromosome, position, reference allele (which is Pb's allele), the alternate allele (which is Pa's allele), and Pa's genotype, which in this case will always be 1/1 for alternate/alternate:
       ```{bash, eval = F}
-      samtools mpileup -l snp.list -q 30 -u -B -f $reference $f1 $pabam | bcftools call -vm -f GQ | awk ' ($8 !~ /INDEL/) {print $0} ' | bcftools query -f        '%CHROM\:%POS\t%REF\t%ALT[\t%GT]\n' > f1.pa.genotypes
+      samtools mpileup -l snp.list -q 30 -u -B -f $reference $f1 $pabam | bcftools call -vm -f GQ | awk ' ($8 !~ /INDEL/) {print $0} ' | bcftools query -f        '%CHROM\:%POS\t%REF\t%ALT[\t%GT]\n' > pa.genotypes
       ```
-    * Finally, we'll print a file with the sites, the Pa allele, and the Pb allele
+    * Finally, we'll print a file with the site, the Pa allele, and the Pb allele.
     ```{bash, eval = F}
-    awk ' { if ( $5 ~ /0\/0/ ) print $1 "\t" $2 "\t" $3; else print $1 "\t" $3 "\t" $2; } ' f1.pa.genotypes > pos.pa.pb.txt    
+    awk ' { if ( $5 ~ /0\/0/ ) print $1 "\t" $2 "\t" $3; else print $1 "\t" $3 "\t" $2; } ' pa.genotypes > pos.pa.pb.txt    
     ```
 
-### Now we can actually genotype the F2s
+### genotyping the F2s
+The first step is to go through each F2 individual and count the number of reads of each allele.
+We'll do this using a combination of samtools mpileup and a custom perl script written by Prof.
+Anji Ballerini (make sure true).
+The mpileup command goes through our F2 .bam files and counts alleles for each individual at our snp list.
+The output has 6 columns:
+  * 1: chromosome/scaffold
+  * 2: position
+  * 3: reference allele
+  * 4: read counts at that locus
+  * 5: allele
+    * . matches reference on forward strand
+    * , matches reference on reverse strand
+    * uppercase letter is identity of alternate allele read on forward strand and lowercase is the same but on the reverse strand
+  * 6: ASCII base quality
+
+I won't pretend to speak perl, but I know this script takes the mpileup as input and outputs a 'counts' file with 5 columns:
+  * 1: chromosome/scaffold:position
+  * 2: # A reads
+  * 3: # T reads
+  * 4: # C reads
+  * 5: # G reads
+
+Below is the for loop that goes through each F2, calling on mpileup and the perl script to output count files.
+
+```{bash, eval = F}
+# make the snp.list a variable that we can easily call
+snplist="$idallelesdir/snp.list"
+
+# make the filepath to the perl script a variable
+MAYBE LINK TO A NEW HTML WITH THE FILE
+
+# make a list of the F2s from the .bamdir
+ls $bamdir | grep -v bai | awk '{ gsub(/.nodup.bam/, "") ; print $0 }' > $infodir/f2.list
+
+#define a list of variables that will be known as "samples" using our f2.list file
+samples=($(cat $infodir/f2.list))
+
+# make a for loop that will cycle through the list of F2 samples one at a time and genotype each sample at our SNP.list positions
+for (( k = 0 ; k < ${#samples[@]} ; k++ ))
+do
+
+samtools mpileup -Q 30 -q 60 -B -f $reference -l $snplist $bamdir/${samples[$k]}.nodup.bam | perl $countscript | awk ' { print $1":"$2 "\t" $3 "\t" $4 "\t" $5 "\t" $6 } ' | sort -k 1,1 > ${samples[$k]}.counts
+
+done
+```
+
+Now that we have small and manageable counts files for each F2, we can leave the world of bash and head to R to finish up our genotyping and genetic map construction.
+Because we're calling genotypes in recombination bins, we need to determine our bin sizes to accurately capture our recombination points.
+We used anywhere from 0.1 - 1Mb bins, depending on where on the chromosome we were.
+Generally we used 0.5Mb bins on the ends of chromosome arms and 1Mb bins to account for spatially differing recombination rates.
+0.1Mb bins were used in regions that were known to have discordance between the genetic and physical maps based om previous studies in *Aquilegia*.
+
  * determine useful window sizes depending on recombination rate

### 2023-12-15 DRAFT_DNA_methylation.Rmd

--- 2023-12-15 DRAFT_DNA_methylation.Rmd
@@
 
----
-
+----------------------------------------------------------
 Vertebrate genomes, on the other hand, have higher overall levels of methylation, with methylation occuring in CpG islands.

### 2023-12-15 ADMIN_01_submissions_instructions.Rmd
added a bit to submisisons instructions page about makign sure to compile a list of dependencies to render .Rmd for our conda environment that I'm starting to build

@@
 3. Review: Full drafts of content will be reviewed by the website manager, Katherine Silliman and/or Sam Bogan, and at least one other active MarineOmics member assigned to the project.
+4.
+Publishing: Final versions of content will be published with the help of the website manager.
+Please have a list of all software dependencies necessary for your .Rmd to be rendered so we can add it to our MarineOmics conda environment.
 
@@
 
-After following these step, you can cite them in the RMarkdown document.
+After following these steps, you can cite them in the RMarkdown document.
 Let's say you have a paper you want to cite by LastName et al. 2023 and
@@
 sentence. This will produce a reference such as the one below for Tripp
-et al. 2017.
-
+et al. 2017. <br>
+<br>
 *syntax in .Rmd text **(not in a code chunk)***:
@@
 <br>
+
+You can also add hyperlinks to each in text citation that direct to the References section at the bottom of the page.
+The in text citation keeps its normal format, but with brackets added around the parentheses.
+Directly after the citation (also in parentheses) is the '#' link to the References section: (#refs) in this case.<br>
+<br>
+**.Rmd**: [(\@Tripp2017-cp)]\(#refs) <br>
+**rendered html**: [(@Tripp2017-cp)](#refs) <br>
+<br>
+At the bottom of the .Rmd, where the '## References' heading is, we have a '{#refs}', which creates the target for all links with '(#refs)'. It looks like this on the .Rmd:<br>
+## References {#refs}
 
@@
 
-## References
+## References {#refs}

### 2023-12-15 DRAFT_lncRNA.Rmd
lncRNA download

@@
 
-## Notes on running code. This workflow is written such that a user an fully reproduce the code. Th
+## Notes on running code. This workflow is written such that a user an fully reproduce the code.
+
+# Downloading data
+
+The data used for this workflow is from Acropora pulcra, and part of the E5 (https://e5coral.org) project. see https://github.com/urol-e5/deep-dive for more details.
+
+```{r, engine='bash'}
+mkdir ../data/lncRNA
+
+wget -r \
+--no-directories --no-parent \
+-P ../data/lncRNA \
+-A "*fastq.gz" https://gannet.fish.washington.edu/Atumefaciens/20230519-E5_coral-fastqc-fastp-multiqc-RNAseq/A_pulchra/trimmed/
+
+```
 

### 2023-12-15 DRAFT_lncRNA.Rmd
lncRNA code

@@
                fig.height = 5,
-               cache = FALSE)
+               cache = FALSE,
+               message = FALSE,
+               warning = FALSE)
 cite_options(citation_format = "pandoc", max.names = 3, style = "html", hyperlink = "to.html")
@@
 
-## Notes on running code. This workflow is written such that a user an fully reproduce the code.
+Notes on running code. This workflow is written such that a user an fully reproduce the code.
 
@@
 ```{r, engine='bash'}
+# Make directories - these need to have a lot of space.
 mkdir ../data/lncRNA
+mkdir ../output/lncRNA
+```
+
+```{r, engine='bash'}
 
@@
 
+## Genome
+
+For this we will be using the closly related genome on Acropora millipora.
+
+```{r, engine='bash'}
+cd ../data/lncRNA
+
+curl -O https://gannet.fish.washington.edu/seashell/bu-github/deep-dive/D-Apul/data/Amil/ncbi_dataset/data/GCF_013753865.1/GCF_013753865.1_Amil_v2.1_genomic.fna
+```
+
+```{r, engine='bash'}
+cd ../data/lncRNA
+
+curl -O https://gannet.fish.washington.edu/seashell/bu-github/deep-dive/D-Apul/data/Amil/ncbi_dataset/data/GCF_013753865.1/genomic.gtf
+```
+
+```{r, engine='bash'}
+cd ../data/lncRNA
+
+curl -O https://gannet.fish.washington.edu/seashell/bu-github/deep-dive/D-Apul/data/Amil/ncbi_dataset/data/GCF_013753865.1/genomic.gff
+```
+
+# Alignment
+
+```{r, engine='bash'}
+/home/shared/hisat2-2.2.1/hisat2-build \
+../data/lncRNA/GCF_013753865.1_Amil_v2.1_genomic.fna \
+../output/lncRNA/GCF_013753865.1_Amil_v2.1.index \
+-p 48 \
+../data/lncRNA/genomic.gff \
+2> ../output/lncRNA/hisat2-build_stats.txt
+```
+
+```{r, engine='bash'}
+find ../data/lncRNA/*R2_001.fastp-trim.20230519.fastq.gz \
+| xargs basename -s -S1-TP2_R2_001.fastp-trim.20230519.fastq.gz | xargs -I{} \
+/home/shared/hisat2-2.2.1/hisat2 \
+-x ../output/lncRNA/GCF_013753865.1_Amil_v2.1.index \
+-p 48 \
+-1 ../data/lncRNA/{}-S1-TP2_R1_001.fastp-trim.20230519.fastq.gz \
+-2 ../data/lncRNA/{}-S1-TP2_R2_001.fastp-trim.20230519.fastq.gz \
+-S ../output/lncRNA/{}.sam \
+2> ../output/lncRNA/hisat.out
+```
+
+# Convert BAMs to SAMs
+
+```{r, engine='bash'}
+for samfile in ../output/lncRNA/*.sam; do
+  bamfile="${samfile%.sam}.bam"
+  sorted_bamfile="${samfile%.sam}.sorted.bam"
+
+  # Convert SAM to BAM
+  /home/shared/samtools-1.12/samtools view -bS -@ 40 "$samfile" > "$bamfile"
+
+  # Sort BAM
+  /home/shared/samtools-1.12/samtools sort -@ 40 "$bamfile" -o "$sorted_bamfile"
+
+  # Index sorted BAM
+  /home/shared/samtools-1.12/samtools index -@ 40 "$sorted_bamfile"
+
+  # remove sam files, no longer need
+  rm *.sam
+
+done
+
+```
+
+# Stringtie
+
+```{r, engine='bash'}
+find ../output/05.33-lncRNA-discovery/*sorted.bam \
+| xargs basename -s .sorted.bam | xargs -I{} \
+/home/shared/stringtie-2.2.1.Linux_x86_64/stringtie \
+-p 16 \
+-G ../data/Amil/ncbi_dataset/data/GCF_013753865.1/genomic.gff \
+-o ../output/05.33-lncRNA-discovery/{}.gtf \
+../output/05.33-lncRNA-discovery/{}.sorted.bam
+```
+
 # References {#refs}

### 2023-12-15 DRAFT_lncRNA.Rmd
lncRNA - stringtie

@@
 ```{r, engine='bash'}
-
 wget -r \
@@
 curl -O https://gannet.fish.washington.edu/seashell/bu-github/deep-dive/D-Apul/data/Amil/ncbi_dataset/data/GCF_013753865.1/GCF_013753865.1_Amil_v2.1_genomic.fna
-```
-
-```{r, engine='bash'}
-cd ../data/lncRNA
-
-curl -O https://gannet.fish.washington.edu/seashell/bu-github/deep-dive/D-Apul/data/Amil/ncbi_dataset/data/GCF_013753865.1/genomic.gtf
 ```
@@
 ```{r, engine='bash'}
-find ../output/05.33-lncRNA-discovery/*sorted.bam \
+find ../output/lncRNA/*sorted.bam \
 | xargs basename -s .sorted.bam | xargs -I{} \
@@
 -p 16 \
--G ../data/Amil/ncbi_dataset/data/GCF_013753865.1/genomic.gff \
--o ../output/05.33-lncRNA-discovery/{}.gtf \
-../output/05.33-lncRNA-discovery/{}.sorted.bam
+-G ../lncRNA/genomic.gff \
+-o ../output/lncRNA/{}.gtf \
+../output/lncRNA/{}.sorted.bam
 ```

### 2023-12-16 DRAFT_lncRNA.Rmd
draft lncRNA multiqc

@@
 
+# Requirements to run code
+
+Need to following software / programs
+
+```
+wget
+curl
+HiSat
+samtools
+Stringtie
+```
+Hardware wise need at least 500 GB disk space, and multiple CPUs otherwise would take extrordinary amoutn of time. Code below was run on machine with 48 CPUs.
+
 # Downloading data
@@
 -A "*fastq.gz" https://gannet.fish.washington.edu/Atumefaciens/20230519-E5_coral-fastqc-fastp-multiqc-RNAseq/A_pulchra/trimmed/
+
+```
+
+Run fastqc
+
+```{r, engine='bash'}
+
+/home/shared/FastQC-0.12.1/fastqc \
+-t 40 \
+-o ../output/lncRNA \
+../data/lncRNA/*fastq.gz
 
@@
 
+multiqc
+```{r, engine='bash'}
+cd ../output/lncRNA
+
+eval "$(/opt/anaconda/anaconda3/bin/conda shell.bash hook)"
+conda activate
+which multiqc
+multiqc .
+```
+
 # References {#refs}

### 2023-12-16 DRAFT_lncRNA.Rmd
lncRNA stringtie

@@
 -p 48 \
+--new-summary \
 -1 ../data/lncRNA/{}-S1-TP2_R1_001.fastp-trim.20230519.fastq.gz \
@@
 -p 16 \
--G ../lncRNA/genomic.gff \
+-G ../data/lncRNA/genomic.gff \
 -o ../output/lncRNA/{}.gtf \
@@
 
+```{r, engine='bash'}
+/home/shared/stringtie-2.2.1.Linux_x86_64/stringtie \
+--merge \
+-p 16 \
+-G ../data/lncRNA/genomic.gff \
+-o ../output/lncRNA/stringtie_merged.gtf \
+../output/lncRNA/*.gtf
+```
+
+# GFF compare
+
+```{r, engine='bash'}
+/home/shared/gffcompare-0.12.6.Linux_x86_64/gffcompare \
+-r ../data/lncRNA/genomic.gff \
+-o ../output/lncRNA/gffcompare_merged \
+../output/lncRNA/stringtie_merged.gtf
+```
+
 multiqc
 ```{r, engine='bash'}
+
 cd ../output/lncRNA

### 2024-02-22 index.Rmd
fix index page for DNA methylation page

@@
 * [Multifactorial RNAseq analyses](DGE_comparison_v2.html) 
+* [DNA Methylation Assessment](FUN_02_DNA_methylation.html) 
 * Align and count RNAseq reads (coming soon!)

### 2024-02-22 FUN_02_DNA_methylation.Rmd
fixing references to photos

--- 2023-12-15 DRAFT_DNA_methylation.Rmd
@@
 subtitle: '*Yaamini Venkataraman*, *Kathryn Greiner-Ferris*, *Sam White*, and *Steven Roberts*'
-bibliography: DRAFT_DNA_methylation/DNA-Methylation.bib
+bibliography: FUN_02_DNA_methylation/DNA-Methylation.bib
 ---
@@
 ```{r meth-bp, echo = FALSE, out.width = "70%", fig.align = "center"}
-knitr::include_graphics("DRAFT_DNA_methylation/meth-bp-chart.png")
+knitr::include_graphics("FUN_02_DNA_methylation/meth-bp-chart.png")
 ```
@@
 [@10.7717/peerj.215](#refs)"}
-knitr::include_graphics("DRAFT_DNA_methylation/meth_ratio.png")
+knitr::include_graphics("FUN_02_DNA_methylation/meth_ratio.png")
 ```
@@
 ```{r meth-igv, echo = FALSE, out.width = "100%", fig.align = "center", fig.cap = "Screenshot from IGV. Bedgraph files are displayed. See below for code to create these file (File Conversion)"}
-knitr::include_graphics("DRAFT_DNA_methylation/igv.png")
+knitr::include_graphics("FUN_02_DNA_methylation/igv.png")
 ```
@@
 ```{r meth-vert, echo = FALSE, out.width = "70%", fig.align = "center", fig.cap= "Genome screenshots of methylation quantified in 2kb running windows over the first 3Mb of chromosome 1 zebrafish and mouse, and of the largest scaffold (NW_006890054.1) in elephant shark (doi: 10.12688/f1000research.11281.1)"}
-knitr::include_graphics("DRAFT_DNA_methylation/vert_methyl_distr.png")
+knitr::include_graphics("FUN_02_DNA_methylation/vert_methyl_distr.png")
 ```
@@
 ```{r meth-summary, echo = FALSE, out.width = "100%", fig.align = "center"}
-knitr::include_graphics("DRAFT_DNA_methylation/bismark-summary-report.png")
+knitr::include_graphics("FUN_02_DNA_methylation/bismark-summary-report.png")
 ```
@@
 
-```         
+```{r, engine='perl', eval=FALSE}         
 samtools merge \

### 2024-08-16 RDAtraitPredictionTutorial.Rmd
changed file path reference to first multivar figure
changed code so loading library code is visible
re-rendering site to have correct html for Katie's RDA page, since she published the link in PNAS

@@
 
-This tutorial accompanies the paper "The paradox of adaptive trait clines with non-clinal patterns in the underlying genes" published in PNAS [add link].
+[How to cite](https://marineomics.github.io/#How_to_Cite)
+
+This tutorial accompanies the paper ["The paradox of adaptive trait clines with non-clinal patterns in the underlying genes"](https://doi.org/10.1073/pnas.2220313120) published in PNAS.
 

### 2024-08-16 index.Rmd
re-rendering site to have correct html for Katie's RDA page, since she published the link in PNAS

@@
 * [Poolseq](POP_03_poolseq.html)
-* [Redundancy Analysis (RDA) Trait Prediction](POP_05_RDAtraitPredictionTutorial.html)
+* [Redundancy Analysis (RDA) Trait Prediction](RDAtraitPredictionTutorial.html)
 * [PCA](POP_08_PCA.html)

### 2024-09-18 POP_09_aDNA.Rmd
adding aDNA page - not quite done yet, need bib

---
title: "Ancient and Degraded DNA"
output: html_document
bibliography: ../aDNA_Part1_ReviewDoc_Refs.bib
---

```{r setup, echo=F, warning=F}
library(knitr)
knitr::opts_chunk$set(
	fig.height = 5,
	fig.width = 10,
	cache = TRUE
)
library(knitcitations)
library(kableExtra)
opts_chunk$set(fig.width = 10,
               fig.height = 5,
               cache = TRUE)
cite_options(citation_format = "pandoc", max.names = 3, style = "html", 
             hyperlink = "to.doc", cite.style = "author")
```

Authors:  Madeline Eppley\*, Sara M. Schaal\*, John Whalen,  Shrushti Modi, Kristen Rayfield, Jem Baldisimo, Evan Ho, Stephen Gaughran   
\* These authors contributed equally to this page.  
Acknowledgements: Thank you to our two reviewers Ram&oacute;n Gallego, and Ally Swank for providing helpful feedback. Also to Jason Johns for helping to facilitate publication of our page.

[How to cite](https://marineomics.github.io/#How_to_Cite)

## Introduction

This page is an introduction to genomic analyses (sample preparation, DNA extractions, library prep, sequencing, and data analysis) of degraded or damaged DNA, including DNA from museum specimens, ancient samples, and poorly preserved modern samples. We assume that you have some experience in analyzing high-throughput sequencing data. If not, we suggest you review the [Population Genomics tab on the MarineOmics page](https://marineomics.github.io/#Population_Genomics) to become familiar with the basics of genome sequencing and data analysis pipelines. The information and tutorials on this page will focus on genomic analyses that are specific to ancient, historical, and/or museum DNA. Because many of the technical themes revolve around DNA degradation and damage inherent to ancient DNA, this page may also be useful to those working with highly degraded modern DNA.

### Ancient DNA  
In the 1980s, scientists discovered that DNA was preserved in museum and ancient specimens, from which it could be extracted and sequenced. While this gave rise to Michael Crichton’s science fiction novel Jurassic Park, it also led to a flourishing field of evolutionary research. Marine biologists have lagged somewhat behind in this field, driven in part by the difficulty of recovering ancient marine samples compared to terrestrial taxa.

Ancient DNA (aDNA) refers to DNA extracted from non-modern samples. These specimens can range in age from a few decades to millions of years old. Sources of aDNA include teeth, bone, herbarium specimens, preserved or mummified tissue, paleofeces, dental calculus, and sediment, among others (@Hofman2019-ur). The context of these sources (ancient, historical, or museum DNA) are sometimes used interchangeably, and while distinct, share several characteristics and should be handled with the same laboratory procedures.  They are all affected by DNA degradation, damage, and contamination. The average aDNA fragment length is around 50 base pairs, but this can be highly variable depending on the age of the samples and preservation (see: @Heintzman2014-io). Samples generally contain low quantities of the original (endogenous) DNA, which are often contaminated with DNA from other sources (exogenous; @Hofman2019-ur). This page will focus on genomic analyses that take into account these challenges, which can also be used as a guide to processing highly degraded modern DNA. Throughout this page we will refer to all non-modern samples as “aDNA” while highlighting some of the complexities in sample preparation and extraction methods due to the source and/or context of the sample (i.e., museum curation impacts on biomolecular yields and downstream data interpretation).

### aDNA Study Design

For the last several decades, developments have been made in extracting and sequencing DNA from degraded samples. In recent years, high-throughput and short-read sequencing technologies have especially advanced the aDNA field. These advances have made it possible to explore the genetics of populations in the recent and distant past. Although there are still many obstacles to aDNA projects, we are now at a point where population-level ancient sequencing projects are feasible for non-model and marine organisms. Some questions that museum and ancient samples can address are:

 - What were past baselines of genetic diversity and how do they compare to modern counterparts?  
 - How have historical events (e.g., environmental change, anthropogenic impacts, admixture, or selection events) affected changes in genetic diversity and allele frequency?  
 - How are extinct species or populations related to extant ones?  
 - What were past migratory routes, range expansions, or range contractions?  
 - What did past ecosystems look like (e.g., aDNA + stable isotope analysis)?  
 - How well do simulations capture past genetic changes, and can we validate them with empirical aDNA data?  
 - What was genetic load like in the past?  
 - What pathogens existed in ancient populations, and how did they evolve?  
 - What is the genetic basis of ancient traits or diseases?  
 - What were established levels of genetic diversity pre-environmental degradation or pre-exploitation?  
 - What were the effects of prior range expansions or contractions during periods of climatic change?  
 - Can populations show evidence of rapid adaptation to environmental stress (abiotic or biotic) over short periods of time?

Designing a study incorporating ancient or historical specimens allows you to ask many new questions but also requires several important considerations. Below are examples of how to form those questions while considering the limitations of your sample quantity and quality.

#### Sampling design: statistical power vs limited availability of specimens

Often, ancient and historical samples are extremely limited in number. When used for genomic analysis, a portion or all of a given sample is completely destroyed. Because of this, researchers need to think critically about their sampling design and the resulting statistical power in their analyses before beginning an aDNA study. Without this consideration, aDNA studies may face power issues when detecting demographic events or changes with population genetic analyses. Ideally, temporal genomics studies will sample individuals serially across multiple points in time over the study period, which adds significant power to analyses of selection, demography, and migration (@Clark2023-vc). However, when samples do not permit serial sampling, two sampling periods - a baseline “historic” genotype, and a more recent “post-event” - can still be informative about change over time (@Clark2023-vc) and have been effective for detecting selective sweeps (@Whitehouse2023-ue).

Depending on the analysis, even single samples can provide enough data for drawing conclusions. For example, genome-wide heterozygosity and runs of homozygosity can be estimated from single samples and can reflect population-level phenomena (e.g., @Beichman2023-eg). In addition, analyses such as the Pairwise Sequentially Markovian Coalescent can take single-sample input to reconstruct demographic history of the distant past (e.g., @Sharko2021-jo). Phylogenetic and phylogeographic reconstructions can also be performed using single samples per population or species (e.g., @Scarsbrook2022-sq).

#### Unequal sample sizes

In most cases, modern DNA samples are more easily obtained than aDNA samples. In the case of temporal comparisons between past and present populations, inclusion of modern samples may need to be scaled down to match the available aDNA sample sizes. In a study of sea otter demographic history, (@Beichman2023-eg) analyzed three aDNA samples from ~1500 and ~200 years ago with seascape population genomic data of 107 contemporary samples from five distinct populations across the species’ range. Modern populations were downsampled to three representative individuals to equal the sample size of the three historical samples for principal components analysis (PCA) and ADMIXTURE analyses. On the other hand, large numbers of modern samples can be leveraged to impute genotypes in low-coverage ancient genomes (@Ausmees2022-fa).

#### Potential downstream population genetics and genomics analyses

In some species, the number of available aDNA samples may be large enough to allow researchers to carry out population genetics and in some cases genomics studies (@Parks2015-nk). In the past, technological limitations have hindered researchers' ability to analyze a large number of markers across the genome and across many samples due to limitations such as quantity of isolated target DNA (see below in challenges with aDNA: exogenous DNA), cost of sequencing to a great enough depth, and a lack of methods that facilitate the evaluation of genomic level marker sets. However, these limitations have recently been alleviated with better extraction methods, reduced costs of sequencing, and the availability of whole-genome sequence capture (@Carpenter2013-py), thus providing genomic level data for aDNA studies. These advances have opened the door for researchers to evaluate evolutionary changes over time and conduct population genomic analyses. Similarly to a contemporary population genomics approach, aDNA can be processed for potential downstream analyses such as PCA to evaluate population genetic structure (@Louis2023-la), effective population size (*N~e~*) to assess demographic changes over time (@Fournier2022-cd), the fixation index (*F*~ST~) to identify regions in the genome with signatures of selection, or allele frequency changes over time. Uniquely, aDNA can also be used for reconstructing ancient genomes, provide context for phylogenetic relationships, reveal the effects of past selection events, or detail historic levels of genetic variation in populations. Additionally, temporal sampling of populations allows direct analysis of past populations rather than relying on population genetic models to infer past states. This adds significant power to evolutionary analyses of selection, demography, and migration.

#### Conservation genetics applications

aDNA can be useful for understanding how past organism-environment interactions proceeded. Faced with anthropogenic climate change, forecasting how biodiversity may respond to future environments is a key conservation challenge (@Hofman2015-ct). Genomic studies using aDNA can provide unique insight for successful conservation and management approaches (@Nakahama2021-su; @Leonard2008-bt). Notably, the following informative metrics can be generated from aDNA studies:

- Estimates of temporal changes in *N~e~* (e.g., @Bolton2018-ip).  
- Recent and historical connectivity of populations (e.g., @Pinsky2010-qa).  
- Changes in allele frequencies due to natural selection from environmental change, disease outbreaks, or natural disasters (e.g., @Lonsinger2018-ws).  
- Changes in allele frequencies due to population collapses or bottlenecks (i.e., genetic drift and relaxed natural selection) (e.g., @Weber2004-ba; @Garcia-Navas2015-rm).  
- Temporal levels of genetic diversity or inbreeding in populations and the timescales on which diversity is lost or gained (e.g., @Baillie2016-kh).   
- Temporal relatedness of isolated populations for genetic rescue applications (e.g., @Hogg2006-jv; @Miller2012-cp).   
- Determining if small populations of genetically distinct individuals or rare haplotypes differ from historic populations to delineate species management units (e.g., @Perrine2007-fd; @Goldstein2003-tf).

### Select Case Studies of aDNA in Conservation and Environment

#### i. Steller’s sea cow:   
Nuclear de novo assembly of Steller's sea cow using museum samples determined that population declines began considerably earlier than the arrival of Western Europeans in the North Pacific, contrary to the previous estimate. The study shows a considerable drop in population thousands of years prior to European description, emphasizing the importance of environmental changes in their extinction (@Sharko2021-jo). Another study demonstrated convergent evolution between Steller's sea cow and cetaceans, highlighting genes involved with cold aquatic adaptation (@Duc2022-dx). These studies highlight the importance of aDNA as a tool for historical reconstruction, providing insights into the impact of anthropogenic activity and environmental shifts on species extinction, such as the Steller’s sea cow.

#### ii. Corals:   
Coral holobiont evolution was  studied from the fragments of millenia-old *Acropora palmata* coral (@Scott2022-xz). In this  study, the authors sequenced aDNA and discovered a  closely related yet distinct genetic relationship between ancient and modern *A. palmata*. Metagenome assemblies showed the stability of millennia-old holobionts that can be used for studying the impacts of environmental stress and evolutionary constraints over time (@Scott2022-xz). The use of aDNA in marine cnidarians can prove instrumental in studying long-term reef evolution, demographic trends, and the impact of anthropogenic disturbances on reef ecosystems.

#### iii. Baltic herring:  
Historic whole-genome sequences from 40 archaeological samples of Atlantic and Baltic herring were used to reveal human impacts on fishery stocks over the past 800 years. In this study, four stocks of herring were analyzed for population structure and temporal fluctuations in Ne over 200 generations. This research illustrates the value aDNA can contribute to modern fisheries management of an economically-important species in a time of anthropogenic impacts on fisheries (@Atmore2022-aj).

#### iv. Sea otter:  
@Wellman2020-pt address the complex historical ecology of sea otters (*Enhydra lutris*) and their failed reintroduction to coastal Oregon in the 1970s. Integrating archaeological and historical museum specimens, they compare the mitochondrial genomes of pre-extirpation Oregon sea otters to extant and historical populations. Complete ancient mitochondrial genomes were sequenced from archaeological Oregon sea otters (*Enhydra lutris*) dentine (n=20) and historical sea otter dental calculus (n=21; @Wellman2020-pt). From their analysis they reveal northern extant populations to be an appropriate population for future reintroduction efforts. Of importance, this study highlights the promising use of dental calculus for ancient population studies as an alternative sample source for research questions of conservation interest.

#### v. Ancient Environmental DNA (eDNA): 
Ancient eDNA can be used as a direct survey of organisms that occurred in  past environments as well as a proxy for ecosystem change. Generally, there are three different types of ancient eDNA: Sedimentary DNA (sedaDNA), ice cores, and cave deposits. sedaDNA refers to DNA from environmental and sedimentary core–samples. Sediment layers contain DNA from organisms that lived in the water or sediment during the time that the layer was deposited. Much of the recovered DNA is microbial, but studies have successfully recovered animal and plant DNA as well (@De_Schepper2019-cy; @Armbrecht2019-be). Ice cores can contain well-preserved DNA from organisms buried deep in the fossil record covered by ice sheets, providing valuable information about past ecosystems (@Willerslev2007-bn; @Reiss2006-pt; @Hansen2002-to; @Zhong2021-cl). Cave deposits are accumulated DNA from organisms that inhabited or passed through caves. These include cave sediments from ancient humans (@Vernot2021-hy; @Sarhan2021-eh), speleothems (@Marchesini2023-ny; @Lipar2020-as), and guano (@Bogdanowicz2020-ak; @McFarlane2024-kq; @Massilani2022-al; @Jenkins2013-ga; @Haidau2022-wl). Taken together, all of these sources of ancient eDNA provide a valuable tool in reconstructing and understanding past environments for assessing global environmental change.

### Challenges with aDNA
#### i. DNA damage:

DNA damage and degradation is the uniting theme of this page, and perhaps the most significant challenge in working with historical, ancient, or poorly preserved samples. DNA damage comes in two main flavors: DNA fragmentation and chemical modifications. **Fragmentation** results from the breaking of chemical bonds between bases, which can cause double-strand breaks, missing bases (“nicks”), and single-strandedness (@Briggs2007-kw). Fragmentation is expected to accumulate over time, resulting in older DNA being more degraded. However, environmental factors also affect degradation rates, including temperature, humidity, local environment/substrate, and storage conditions (e.g., Formalin-Fixed, Paraffin-Embedded). **Cytosine deamination** is the primary chemical modification in aDNA, in which cytosine is chemically converted to uracil. During library prep, standard polymerases will incorporate adenine (instead of guanine) across from the uracil, resulting in erroneous C-to-T or G-to-A transitions (@Sawyer2012-pq). Although they are sources of error, these predictable patterns of DNA damage and degradation can be used to verify that genuine aDNA was sequenced.

#### ii. Exogenous DNA contamination:

When working with ancient or historical DNA samples, contamination of DNA from organisms and/or species that are not the target (exogenous DNA) is expected. Exogenous DNA contamination includes modern DNA (typically human DNA from handling samples or from other modern species) and microbial DNA (typically from environmental contamination where the samples were collected). Since modern DNA is a common source of exogenous DNA contamination, it is critical to work in a lab space designed for only working with ancient DNA. This should be a clean room where no modern DNA work has previously been done. There should be contamination control workflows including protective clothing (e.g., tyvek suits, double layered gloves), sterilization requirements (e.g., bleach, UV-sterilization), and HEPA-filtered laminar flow hoods. In addition to working in a lab specifically set up for working with ancient DNA, there are steps that can be taken in laboratory protocols to try to reduce the amount of exogenous DNA contamination during sample preparation, DNA extraction, and library preparation. Although mitigation strategies can help to reduce the amount of exogenous DNA contamination in laboratory protocols, some amount of exogenous DNA will still be present in the sample and therefore in the sequencing data. There are a number of ways that researchers can deal with this bioinformatically.

#### iii. Chemical contaminants/PCR inhibitors:

Historic museum samples are often chemically treated to enhance preservation with chemicals (e.g., ethanol, isopropyl alcohol, formaldehyde, formalin, paraffin). These chemical methods of preservation inherently damage and degrade DNA. Due to contamination from environmental preservation, various chemical methods often need to be used to decontaminate ancient samples (e.g., remove physical surface contaminants or exogenous DNA) before extracting DNA (@Llamas2017-zo). These decontamination treatments can damage endogenous DNA and result in additional sample degradation (@Orlando2021-zd). After DNA extraction of either historic or ancient samples, polymerase chain reaction (PCR) can be used to amplify small amounts of DNA to a quantity suitable for sequencing. However, PCR inhibitors that are co-extracted alongside DNA are common and may make PCR amplification impossible (@Kemp2014-qf).

#### iv. Limited sample availability:

The availability and preservation of samples for aDNA research can be difficult and frequently results in sample size limitations, particularly when the species being studied is not well-represented in the archaeological or paleoecological record (@Schwarz2009-df). This makes it harder to get a thorough understanding of past migrations and demographics. Geographic limitations may also exist for aDNA studies in regions with a low concentration of sites or poorly conserved remnants. It might be difficult to extract high-quality DNA from samples because of variables like temperature and humidity that cause sample degradation. Ethical and legal issues, such as regulations pertaining to excavation and examination, further hinder access to historical materials (@Licata2020-ap). Because aDNA research involves the destruction of valuable materials, ethical handling of the limited resources is imperative (see [aDNA Ethics] section below).

#### v. Data analysis and interpretation difficulties:

Even when specialized methods are used to address the above challenges at the stages of DNA extraction and sequencing, ancient and degraded DNA inevitably produces lower quality data and less-than-ideal sample sets compared to most modern DNA projects. Fortunately, since the patterns produced from DNA damage and degradation are somewhat predictable, special analytical tools have been developed that reduce the noise that comes from low-quality data. These include short-read mapping approaches that are especially sensitive to mapping reads that are < 50 bp (@Oliva2021-yi; @Xu2021-iv), methods for genotype imputation from extremely low-coverage data (@Rubinacci2021-wv; @Howie2009-uv; @Rubinacci2020-ux; @Davies2016-hk), and variant callers that explicitly account for patterns of DNA damage (@Prufer2018-ki; @Kawash2018-og).

### Sample Acquisition

#### Traits Affecting Sample Quality

Given recent advances in aDNA sequencing and analysis, the greatest challenge to aDNA projects may now be finding appropriate samples to sequence. Unlike fresh tissue samples collected for genomic analyses, samples used in an aDNA project were likely not preserved with the intention of future DNA extraction and sequencing. Many samples are prehistoric or archaeological, and were exposed to the elements for hundreds of thousands of years. Other samples may have been prepared for museum collections but treated with preservative chemicals that degrade DNA. No matter how good the preservation conditions, DNA will degrade at room temperature in dead tissue, resulting in low quality and quantity DNA in any museum or ancient samples. For these reasons, aDNA projects must often be designed in an iterative process, with the available samples dictating the questions that can be asked, and the initial sequencing results dictating which samples are usable, which determines the power of downstream analyses. Some traits that may be correlated with sample quality are:

##### i. Age

Just like radioactive elements, DNA degrades in a clock-like way. Unlike elemental degradation, however, the rate of decay depends on many factors besides age. Still, age can generally be used as a first order proxy to estimate the quality of a sample. Most degradation happens rapidly, within the first days and weeks after tissue death. Still, degradation continues after, with older (i.e., prehistoric) samples generally yielding lower quantity and quality DNA than younger (i.e., historic) samples. However, post-mortem alterations due to the environment, climate, or museum preservation efforts can challenge this assumption (@Austin2019-nc).

@Allentoft2012-om calculated an average DNA half-life of fossil samples from the extinct New Zealand moa to be 521 years for a 242 bp mtDNA sequence, which corresponds to a per nucleotide fragmentation rate (*k*) of 5.5010 - 6 per year. The authors also identified that nuclear DNA degraded at least twice as fast as mtDNA based on Illumina HiSeq data. They argue that Equation 1 represents the best available approximation of the rate of mtDNA decay in fossil bone, in which (*k*) is the average decay rate per site per year for mtDNA in moa bone and (*T*) is temperature.

$$ln \space k = 41.2 - 15267.6 \times \frac{1}{T} \space \space \space \space \space	\mathrm{Eq.1}  $$

##### ii. Climate

The immediate environmental conditions where a specimen has been for most of its post-mortem existence is a strong factor in DNA quality and state of degradation. In general, cold and static environments (e.g., permafrost) will preserve DNA best; hot, variable environments (e.g., tropics) will accelerate DNA degradation. Likewise, stable museum cabinets also provide relatively good preservation of DNA. Even so, new methods are working to improve DNA sequencing from ancient specimens in the tropics.

##### iii. Exposure/environment

In addition to climate factors like temperature and humidity, the microclimate and exposure of a specimen also affects DNA quality. For example, a bone in a tropical cave may yield better quality DNA than a bone exposed to the sun just outside the cave for an equal amount of time. These factors can also be complex: buried specimens may have better preserved DNA but also more contamination from soil microbes invading the tissue matrix.

##### iv. Tissue type

Just as with modern DNA, different types of tissue can yield different amounts of DNA. Unlike modern tissue, though, the type of tissue also factors into DNA degradation rates. For example, soft tissues may yield more DNA in some circumstances, but bone and teeth in vertebrates often preserve DNA better than softer tissue. Even within a tissue type, there can be variability. In general, dense bones (i.e., mammalian petrosal bone) preserve DNA much better than spongy, porous bones.

##### v. Species

The species itself interacts with all of the above factors: the climatic range of the species, the type of tissue making up that organism, the density of its bone (if it has any). For example, it is possible to predict specimen quality for ancient marine mammals depending on some species-specific characteristics, see @Keighley2021-jh.

##### vi. Chance!

Most importantly, there seems to be a huge amount of chance in specimen quality when it comes to aDNA. This is true both because of inherent qualities related to the species of interest and because of the specimens that may be available. Although we advocate for careful consideration of whether there are enough samples and then which samples to use before attempting a study, there may sometimes be situations where you will not know the quality of the sample until you evaluate DNA quantity and quality.  If sample size is limited, a pilot study evaluating these characteristics may be needed before attempting destructive analysis on all available specimens. Otherwise, researchers risk destroying precious samples and then having no or limited usable data from the samples.

#### Locating Specimens in Museum Collections and Repositories

Specimens suitable for DNA extraction and analysis can be located in many different museums, collections, and repositories. Below are links to collections that are available to be browsed online.

##### i. Collection Databases

- [Symbiota](https://symbiota.org/)  
- [Integrated Digitized Biocollections (iDigBio)](https://www.idigbio.org/)  
- [Smithsonian Institution](https://collections.si.edu/search/)  
- [VertNet](http://www.vertnet.org/index.html)

##### ii. Tissue Repositories

- [NEON Biorepository Data Portal](https://biorepo.neonscience.org/portal/)  
- [Global Biodiversity Information Facility (GBIF)](https://www.gbif.org/)  
- [Arctos Collaborative Collection Management Solution](https://arctos.database.museum/home.cfm)  
- [Ocean Genome Legacy at Northeastern University](https://ogl.northeastern.edu/)

##### iii. Sequence and Metadata Repositories

- [NCBI](https://www.ncbi.nlm.nih.gov/)  
- [BOLD](https://boldsystems.org/index.php/IDS_OpenIdEngine)  
- [NEON](https://www.neonscience.org/data-samples/data-collection/observational-sampling/observation-types/dna-sequences)  
- [Genomic Observatories MetaDatabase (GEOME)](https://geome-db.org/)  
- [European Nucleotide Archive](https://www.ebi.ac.uk/ena/browser/home)  
- [Diversity of the Indo-Pacific Network](http://diversityindopacific.net/)  
- [Ocean Biodiversity Information System](https://obis.org/)  
- [SRA-Explorer](https://sra-explorer.info/)

#### A Note on Sampling Bias in Museum Collections

It’s important to note that preserved specimens in natural history museums probably do not represent the full diversity of wild genotypes. Natural history museums often preserve specimens of unique phenotypic interest, for a particular comparative project, or in a localized area, resulting in substantial sampling bias. For example, bird collections in large natural history museums were found to have an overall sex bias (only 40% female; @Cooper2019-nn). This bias was more distinct in species with sexual dimorphism and showy male traits (e.g., colorful plumage), and even more distinct in type specimens (only 25% female; @Cooper2019-nn). More generally, collection biases are apparent in the overrepresentation of mammals and birds in natural history museums, the preference for collecting larger specimens (e.g., of antlers, skulls, horns, tusks etc.) where possible, and the geographic limits from where specimens are able to be collected (@Cooper2019-nn; @Dehasque2020-tp). Of course, the most ideal collection of specimens would be serially sampled over regular intervals, collected from the same geographic locations at each sampling point, and collected without bias to phenotype, but rarely are all of these conditions met. Given this, when designing temporal genomics studies using preserved specimens, be sure to evaluate the collection for sampling bias and be aware of how sex, phenotype, or geographic sampling bias could impact your analysis.

### aDNA Ethics

While every field has their own guidelines and standards for ethical research, we highlight the ongoing discussions of ethics involved in aDNA. Most prominently these discussions have revolved around human aDNA, focusing on sample collection, collaborative and inclusive research that ensures sensitivity to diverse stakeholders, data sovereignty, minimizing the misuse of research results, and respect for the remains and those connected to ancient individuals (@Alpaslan-Roodenberg2021-cg; @Claw2018-da; @Carroll2020-xj; @Wagner2020-xb; @Tsosie2021-mn). Given the scope of this page, here we focus on the importance of extending these discussions of aDNA ethical research to study marine systems and some of the ways it has been implemented.

Ethical issues were first addressed by Article 15 of *The Convention on Biological Diversity*, published in 1992. It established the foundational principles regarding access to genetic resources and the fair and equitable sharing of benefits arising from their utilization. It emphasizes the sovereign rights of states over their natural resources, which extends to genetic resources.

The 2010 *Nagoya Protocol* further elaborated on these principles by aiming to ensure the benefits arising from the use of genetic resources, including aDNA, are shared fairly and equitably with the country of origin. This is particularly significant for aDNA studies, where genetic material may come from indigenous territory or developing nations. The protocol calls for clear agreements on the terms of access and the specific nature of benefit-sharing, which can include sharing of research results, joint ventures, the development of local research capabilities, and monetary compensation. @Lin2023-da exemplify inclusive and integrated research by working with the Coast Salish community and incorporating Coast Salish Indigenous knowledge in their study, which sequenced the only known ancient specimen of the Coast Salish Woolly dog.

The paper *Who Owns the Ocean? Policy Issues Surrounding Marine Genetic Resources* (@Vierros2016-as) addresses the complex regulatory and ethical issues related to the ownership and use of marine genetic resources. The study discusses the debate over the governance of international waters and the genetic materials they contain, highlighting the lack of clear international legal frameworks to address the rights of these resources. @Vierros2016-as examine the implications of the *Convention on Biological Diversity* and the *Nagoya Protocol* on marine biodiversity and how these agreements could be extended or adapted to cover areas beyond national jurisdiction. This serves as a crucial resource for researchers and policymakers about the challenges of ensuring the fair and sustainable use of marine genetic resources.

In addition, national and international laws govern the use of many ancient specimens. For example, many countries require certificates or permits for museum and archaeological specimens that are listed under the Convention on International Trade in Endangered Species (CITES). In the United States of America, the Native American Graves Protection and Repatriation Act (NAGPRA) applies to any items of cultural significance to Native Americans, including archaeological and anthropological specimens of non-human origin. Cultural significance is determined by members of the Tribe or descendant community, and cannot be presumed by non-members based on archaeological context (e.g., discovery of an item in a midden). If you are working with cultural or archaeological specimens–even those accessioned into a museum–be sure to follow all applicable laws and seek guidance on best ethical practices.

We have listed a few questions to consider in acknowledgement of ancient sampling ethics, and separated them into the following categories:

#### i. Sampling

- How are samples collected without disturbing natural/preserved environments?  
- Does the research imply the destruction of preserved or historical specimens, such as museum samples, archaeological specimens, or holobiont cores?  
- Will unused DNA be preserved for future studies in ways that other researchers can access? Will there be leftover material that can be returned to the museum and community?   
- Will there be 3D or high resolution images taken of samples before destruction?

#### ii. Data Sharing

- To what extent is the research reproducible?  
- Are all scripts publicly available?  
- Can future collaborators contact your research team for questions/further analyses?

#### iii. Community Participation

- What is the opinion of the local community in terms of the research? Does the project have potential cultural implications?  
- What communities are impacted, both short and long term, by the findings of the research? Were the project and potential impacts discussed in clear terms with members of the community?   
- How does the research benefit local communities, especially those that are often marginalized? How will the researchers share findings and engage with the community about the findings?  
- If local community knowledge was used, was the opportunity for authorship extended?

### aDNA Research Societies

Modern advances in DNA sequencing techniques have expanded our capacity to analyze degraded and low quantity DNA. This has incredible implications for practically any field involving genomic work, but especially for analyzing ancient DNA. Because this field is still relatively new, many communities have formed to discuss best principles when analyzing aDNA. Here you will find a brief, nonexhaustive list of communities, beyond our MarineOmics group, that are undergoing the challenging (yet exciting!) task of establishing sustainable protocols and guidelines.

- [SPAAM](https://www.spaam-community.org/) – Standards, Precautions, and Advances in Ancient Metagenomics. Mission statements are to provide scientific advancement through community-driven projects, training and support, and networking for the ancient metagenomics community.  
- [sedaDNA Scientific Society](https://sedadna.github.io/) – Their mission is to transmit information about current sedaDNA research, promote best practices, and increase collaborations between research groups. They have numerous ongoing collaborative projects, including a fish sedaDNA working group that is developing methods to analyze eDNA preserved in aquatic sediments.  
- [Southern eDNA Society](https://sednasociety.com/) – Based in Australia and New Zealand, this society aims to establish eDNA best practices across government, private, and community sectors. They have released several guidelines and manuals ranging from biomonitoring, DNA test validations, and biodiversity assessments.  
- [Japanese eDNA Society](https://ednasociety.org/en/) – This society aims to foster and develop eDNA science as a discipline. They hold annual meetings, eDNA technical seminars, and public symposia each year.  
- [Animal aDNA Research Community](https://animal-adna-research-community.github.io/AaRC.github.io/) - An international and open community of animal ancient DNA researchers.

### References

### 2024-09-18 index.Rmd
adding aDNA page - not quite done yet, need bib

@@
 * [PCA](POP_08_PCA.html)
+* [Ancient & Degraded DNA](POP_09_aDNA.html)
 

### 2025-03-06 DRAFT_Signatures_Selection/DRAFT_Signatures_of_Selection.Rmd
Adding Signatures of Selection material

---
title: "A MarineOmics overview of methods to detect signatures of selection"
author: "Joanna Griffiths, Danielle Davenport, Madison Armstrong, Camille Rumberger"
date: "2025-02-24"
output:
  html_document:
    toc: yes
bibliography: "common-bib_01.bib"
---

*Created using [HackMD](https://https://hackmd.io/) and [tablesgenerator](https://www.tablesgenerator.com/markdown_tables)*

## Target audience
This page is designed for beginners and provides a broad introduction to different methods commonly used to study natural selection in wild populations at a microevolutionary scale. On this page we will explore how techniques like genotype-environment analysis (GEA) and genome-wide association studies (GWAS) can help us study natural selection. To learn more about specific methods, read through the suggested readings in each section and visit the tutorials where links are provided.

## Introduction
This page provides an overview of different methods for detecting microevolutionary change, specifically identifying genetic loci that may be subject to selection. Natural selection is the differential survival and reproduction of individuals based on their inherited traits. Selection acts on observable characteristics (phenotypes), changing the prevalence of different gene variants (alleles) within a population over time. This leads to changes in allele frequencies over time (generations) and adaptation to spatially heterogeneous environments which may help a population better survive and reproduce in a specific environment. This is particularly relevant in the field of marine ‘omics, where understanding a population's ability to adapt to changing environments, such as climate change or habitat loss, is crucial for preventing extinction.

The methods discussed on this page focus on the microevolutionary scale and examine changes that occur within populations over a relatively short period. This is in contrast to the study of macroevolution, which examines large-scale evolutionary patterns over much longer timescales (i.e. variation in conservation or non-synonymous mutations). The methods described here are used to identify loci which may be subject to selection and are typically applicable to use for any type of population or individual genomic data (e.g. Whole Genome Sequencing, RADseq, etc). Some of the methods can be applied to datasets where you have phenotypic or fitness data (i.e. GWAS for growth, survival, phenotype, etc.), while other methods do not require any data related to either the environment or phenotypes at all!

Please note, if temporal samples are available, other approaches can be used to track allele frequency changes over time. For instance, [(@buffalo2020estimating)](#refs) describe methods for estimating the genome-wide contribution of selection to temporal allele frequency change, providing a framework for leveraging time-series data to detect selection more directly. These temporal methods allow for a more explicit examination of how selection shapes genetic variation across generations, which is not possible with single-time point datasets [(@brennan2022experimental)](#refs).

<ins>On this page we describe three classes of methods to identify loci under selection at a microevolutionary scale using samples collected from a single time point across space/environments. .</ins>
1. **Outlier Analyses** identify extreme deviations from what would be expected under neutral evolution. 
2. **Genotype-Environment-Analysis (GEA)** links genomic variation to environmental gradients.
3. **Genome-Wide Association Studies (GWAS)** examine genetic variants and identify those statistically associated with a particular trait. When a specific phenotype shows a significant association with a particular loci, gene or region, it suggests that this genetic variation may be adaptive.

## Which method should I use to find loci under selection?
Briefly, the biology of your system is always important to consider before conducting specific analyses, and no matter which method(s) you use, it is recommended that comparisons of outlier vs non-outliers (neutral loci) are performed [(@riginos2016navigating)](#refs) as a kind-of “validation” step, where we would expect loci under selection (i.e. outliers) to demonstrate a stronger signal of environmental dependence than non-outliers. However, a common but conservative approach taken by researchers performing outlier detection is to apply multiple test methods and identify outlier SNPs that are overlapping across them. For example, some studies perform a differentiation outlier test first (see [Outlier Analyses Section](#outlier-analyses)) and then use only the resulting outliers as input for the Genome-Environment Associations  [(@rellstab2015practical)](#refs).

Importantly, to conduct outlier analyses you do not necessarily need to have neutral genetic differentiation (genetic variation among populations  primarily driven by random processes like genetic drift and migration) in your data. For example, marine invertebrates with pelagic larval dispersal and high gene flow which show very low to no detectable population structure at neutral alleles across their range (i.e using STRUCTURE, ADMIXTURE, PCA) still show signals of adaptation. Other examples include Pacific purple sea urchins ([(@petak2023whole-cp)](#refs);[(@rumberger2023selection)](#refs)) or scallops [(@van2017identifying-cp)](#refs).

# Outlier Analyses

## What are Outlier Analyses?

Genetic differences among groups may be driven by a combination of genetic drift, demography and/ or selection. Each of these processes can directly change the allele frequency of a population. However, alleles that are under selection and influence fitness are expected to occur at  frequencies different to those affected only by neutral processes. For example, loci under divergent selection (and those physically linked to them) exhibit stronger differentiation than neutral regions and so we might expect these loci to be more differentiated between populations/groups.  We can identify these differences in allele frequencies among different populations/environments using outlier tests. Outlier tests for selection assume a null model of neutrality, where allele frequencies drift randomly without selection pressures. Loci with significantly higher or lower differentiation (i.e. Fst) between populations/groups than expected under a neutral model are flagged as potential targets of selection. Specifically, while selectively neutral loci are expected to form a normal frequency distribution,  when alleles increase fitness (positive selection) they can be expected to be observed at higher frequencies and where they decrease fitness (negative selection) they can be expected to be observed at at lower frequencies [(@hoban2021genetic-cp)](#refs).  These are  “outliers” or “putatively adaptive loci” which may be candidates of selection.

There are two commonly applied approaches to outlier analysis; FST-based analysis (Figure 1) and PCA-based analysis (Figure 2). Both are used to identify loci under selection, but they differ in their underlying methods. FST is a standardized measure of the variance of allele frequencies among populations [(@wright1949population)](#refs). By directly comparing genetic differentiation between populations/groups to neutral expectations, FST outlier tests can identify loci which may be under selection (Figure 1). In contrast, PCA-based methods use Principal Component Analysis (PCA) to capture population structure and can identify loci with extreme loadings on principal components as potential outliers (Figure 2).

```{r, echo=FALSE, out.width="75%", fig.align="center", fig.cap="left-aligned", fig.cap="Figure 1: Example of an FST-based analysis using OutFLANK. In this dataset, 1000 individuals were collected from across 39 populations  spanning an environmental gradient. Each dot is a SNP and its location in the genome plotted on the X-axis. The Y-axis is the FST for each SNP. Higher FST values are indicative of larger genetic differentiation and candidate SNPs under selection are highlighted in pink. [Source](https://htmlpreview.github.io/?https://github.com/whitlock/OutFLANK/blob/master/inst/doc/OutFLANKAnalysis.html)"}
knitr::include_graphics("images/Fig1.png")

```

```{r, echo=FALSE, out.width="75%", fig.align="center", fig.cap="left-aligned", fig.cap="Figure 2: Example of a PCA-based analysis using pcadapt. In this dataset, a total of 150 individuals from three different populations were genotyped at 1,500 markers. Population structure was identified for three discrete groups of individuals, matching population assignment expectations (A). SNPs that are significantly associated with PC1 and PC2 are given higher weightings and are used as input for a Manhattan plot, where the SNP location in the genome is plotted on the X-axis and the weightings (or -log pvalue) are plotted on the Y-axis. SNPs that have higher weightings will have higher -log pvalues and are candidate SNPs under selection among populations (B). [Source](https://bcm-uga.github.io/pcadapt/articles/pcadapt.html)"}
knitr::include_graphics("images/Fig2.png")

```

## General Method Overview:

### Step 1 - Filtering your dataset

Nonselective evolutionary forces can affect the distribution of outlier loci from loci across a genome [(@lotterhos2014evaluation)](#refs), therefore, the first step is to identify the  underlying distribution that is expected under neutral processes by answering the question;  how much variation in among loci would be expected (i.e. the null distribution of FST) in the absence of selection?  One must first create a null distribution using a set of SNPs (single nucleotide polymorphisms)  that accurately model the distribution expected under neutrality. An accurate null distribution typically requires some filtering of a set of SNPs, and in many cases, the best set of SNPs for neutral parameterization is not the entire set of SNPs that you want to evaluate for outliers. Common filtering steps could include a minor allele frequency/count, removing missing data and related individuals, genotype depth and quality, and removing SNPs in linkage disequilibrium (LD). As a general guide, we recommend the following [SNP filtering tutorial](https://ddocent.com//filtering/). Deciding on how much filtering is required can be tricky and can feel subjective. We recommend visualizing the filtering steps using a PCA, which will elucidate many common biases in the dataset (see the [MarineOmics page on PCAs](https://https://marineomics.github.io/POP_08_PCA.html)). We recommend using the same set of SNPs that you use for estimating population structure for neutral parameterization (which is often a set of SNPs not in LD). Once neutral parameterization is performed on one set of SNPs, for most programs that model can be applied to the larger filtered SNP dataset (that across the genome for evaluating their outlier status.

### Step 2 for detecting outlier loci - choose an outlier method based on the biology of your system!

When choosing an outlier method, it is important to understand what biological assumptions are being ignored/violated and whether the model accounts for the demographic history when estimating divergence among populations (see section “Considerations” above). For example, early models such as fdist and fdist2 [(@beaumont1996evaluating)](#refs) assume populations are all equal in size and there is equal probability of migration and gene flow among all populations (even if some populations are located further away than others). This method produced a lot of false positive outliers because more realistic scenarios violated these simple assumptions [(@lotterhos2014evaluation)](#refs). An improvement over this approach was implemented in the program BayeScan where each population has its own neutral model [(@foll2008genome)](#refs), but like fdist/fdist2, BayeScan also  suffers from a large false positive rate due to its violated assumptions of non-independence among populations. Newer outlier methods attempt to account for population demographic history and independence in different ways. For example,  pcadapt uses PC loadings (insead of Fst distributions) which are meant to be less reliant on assumptions about the underlying demography [(@duforet2015haplopop)](#refs). However, PC loadings can be influenced by recombination variation, so it is important to use a set of SNPs thinned for linkage disequilibrium for neutral parameterization [(@lotterhos2019effect)](#refs). These estimates of population structure are included as covariates in the final Outlier Analysis model. See Table 1 below for more information on how each method accounts for population demographic history.

Population structure is also important to consider. For example, if you discover only one genetic cluster (K=1 in STRUCTURE or Admixture plot), perhaps any of the methods in Table 1 would work well for your data. On the other hand, if there is a lot of population structure (i.e., K>1), then a method that accounts for demographic history should be chosen. Additionally, consider what assumptions are made in the tutorials that you use as a resource. Does the tutorial you are using assume that your data is normally distributed? Multiple tutorials have been added below to help with comparing between different assumptions made for different kinds of data.

Many researchers will identify outliers using two or three of these methods and look for overlap between them [(@silliman2019population)](#refs). This is a conservative approach, and researchers should keep in mind that they may not find much overlap among different analyses because of the different assumptions made by each method.

<ins>Further Recommended Reading:</ins>
* [(@whitlock2015reliable)](#refs) (Outflank)--method comparison
* [(@silliman2019population)](#refs) --empirical study with method comparisons 
* [(@lotterhos2014evaluation)](#refs)(Fst outlier methods)--method comparison
* [(@lotterhos2019effect)](#refs)(pcadapt, Outflank, Baypass)--method comparison
* [(@lotterhos2022simulation)](#refs)–method validation
* [(@capblancq2018evaluation)](#refs) (GEA & pcadapt)--method comparison
* [(@luu2017pcadapt)](#refs) (pcadapt)--method details
* [(@hoban2016finding)](#refs)–helpful review and suggested best practices for statistical analyses
* [(@robertson1975gene)](#refs) – theory
* [(@ahrens2018search)](#refs) –review, summary of findings from outlier scans, and suggested best practices for experimental design

### TABLE 1. Common methods for identifying outliers and the associated approach for calculating divergence, the biological assumptions, and how demographic history is accounted for in the model. We also provide a link to public tutorials for each method.
| Model & Approach                                                                                                                                                                                                                                                                                | Demographic History                             | Assumptions/ Known issues                                                                                                                                              | False Positive Rate                                                                                                                                     | Tutorial                                                                                                                                                                                                                                                   | Citation                                                            |
| -------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------- | -------------------------------------------- | ---------------------------------------------------------------------------------------------------------------------------------------------------------------------- | ------------------------------------------------------------------------------------------------------------------------------------------------------- | ---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------- | ------------------------------------------------------------------- |
| fdist & fdist2 (Fst; using Arlequin software)                                                                                                                                                                                                 | * accounts for loci with different heterozygosities and with finite sample sizes per subpopulation  | * assumes populations are equal in size and migration rates <br>* non-independence among populations                                                                   | High [(@lotterhos2014evaluation)](#refs) [(@narum2011comparison)](#refs) [(@perez2010comparing)](#refs), [(@whitlock2015reliable)](#refs)               | Outdated due to high false positive rates                                                                                                                                                                                                                  | [(@beaumont1996evaluating)](#refs)                                  |
| Bayescan (Fst)                                                                                                                                                              | * populations defined ahead of time <br>* Dirichlet-multinomial distribution: each population has its own parameter for neutral model                                  | * does not allow for the non-random correlations among pairs of populations that cause the non-independence of sampling <br>* slow run time [(@luu2017pcadapt)](#refs) | Low to high depending on the demography                                                                                                                 | [Benestan tutorial (2020)](https://rpubs.com/lbenestan/outlier)                                                                                                                                                                                            | [(@foll2008genome)](#refs)                                          |
| PCAdapt (PC loadings) | * Latent factors to estimate values of K <br>* uses PC loadings (less reliant on assumptions about the underlying demography) [(@duforet2015haplopop)](#refs) <br>* Important to use a set of SNPs thinned for linkage disequilibrium for neutral parameterization [(@lotterhos2019effect)](#refs)                           | * individuals not sorted into populations ahead of time <br>* Performs well in populations with high admixture                                                         | Low to medium [(@capblancq2018evaluation)](#refs), [(@luu2017pcadapt)](#refs) Might not be suitable for populations with complex demographic histories. | [Luu, Privé and Blum tutorial](https://https://bcm-uga.github.io/pcadapt/articles/pcadapt.html), [UC Davis Marine Genomics course tutorial ](https://baylab.github.io/MarineGenomics/week-7--fst-and-outlier-analysis.html#finding-outliers-using-pcadapt) | [(@duforet2015haplopop)](#refs)  [(@silliman2019population)](#refs) |
| Outflank (Weir and Cockerham’s Fst)                                                                                                                                                                                                                                                                | * populations defined ahead of time             | * estimates the number of K genetic clusters to inform a null model of Fst distributions <br>* Performs poorly in populations with admixture and pooledseq samples     |                                                                                                                                                         |                                                                                                                                                                                                                                                            |                                                                     |* finite sample sizes in pop                                                                                                                                                                                                                                                                                                                                        | Low, but suffers from low power                                                                                                                         | [Lotterhos Tutorial](https://github.com/whitlock/OutFLANK/blob/master/inst/doc/OutFLANKAnalysis.html) (2018) [UC Davis Marine Genomics course tutorial](https://baylab.github.io/MarineGenomics/week-7--fst-and-outlier-analysis.html#using-fst-to-find-outliers)  | [(@lotterhos2014evaluation)](#refs) [(@schlotterer2014sequencing)](#refs) |
| FLK TF-LK                                    | Sensitive to complex patterns of migration and sharp differences in the migration rate among populations [(@bonhomme2010detecting)](#refs)                                                            | * The program FLK uses a population tree (without migration) to describe the covariance among populations and build a null model of Fst distributions.                                                                                                                   | Low [(@bonhomme2010detecting)](#refs)                                                                                                                   | [UoF Bioinformatics Hub Tutorial](https://uofabioinformaticshub.github.io/GumCreek2012/scripts/S1_FLK )                                                                                                                                                                                                                        | See supplementary file 1 from [(@bonhomme2010detecting)](#refs)           |
| BayPass XTX statistics                       | Assumes HWE for PoolSeq data [(@gautier2015genome)](#refs)                                                                                                                                            | * Uses a covariance matrix of allele frequencies among populations  [(@pickrell2012inference)](#refs) * can be used with or without environmental data                                                                                                                                                                           | low                                                                                                                                                     |[Barley-Agroclimatic-Association Tutorial](https://eead-csic-compbio.github.io/barley-agroclimatic-association/HOWTOXtX_BAYPASS.html), [Erica Nielson Tutorial](https://https://esnielsen.github.io/post/pool-seq-analyses-poolfstat-baypass/)                                                                                                                            | [(@gautier2015genome)](#refs)                                             |

# Genome Environment Association (GEA)
## What is a Genome-Environment Association?

The distribution of genetic diversity within natural populations is significantly influenced by environmental conditions. Along environmental gradients, many marine species demonstrate genetic variation for adaptive traits [(@pinsky2021diversification)](#refs). Genotype environment association (GEA) analysis is a statistical framework that allows us to investigate these genotype-environment relationships and identify specific genetic markers (loci) associated with environmental variation across populations [(@rellstab2015practical)](#refs). In a heterogeneous environment, environmental factors may act as selective pressure for local adaptation. In this case, allele frequencies at loci underlying adaptive phenotypic traits should be associated with environmental factors that act as proxies for the unobserved selective pressures [(@frichot2015lea)](#refs). GEAs are distinct from  outlier detection methods previously described as they do not rely on an allele frequency distribution, but instead use the correlation between genotypes and environment for identifying both loci under selection and key climatic factors driving local adaptation. GEA analyses usually control for isolation by distance in addition to demography (i.e. that associations between genotype and environment are not attributed to environments being far away from one another).

## General Method Overview: GEA Methods 
GEA broadly refers to any methodology that assesses the association between allele frequencies and environmental variables. Therefore, you will need four things before getting into GEAs - a good sampling regime, genetic data, geo-referenced environmental variables (see Section x). GEA methods commonly cited in the literature include: Redundancy Analysis (RDA) Gradient Forest (GF), Bayesian methods (i.e. Bayenv) and Latent Factor Mixed Models (LFMM).

**Redundancy Analysis (RDA)** is a multivariate statistical technique used to explain the variation in a response variable (e.g., genotype) by a set of explanatory variables (e.g., environmental factors).  RDA is an extension of Principal Component Analysis (PCA) that relates two data matrices. It identifies linear relationships between the response variable and explanatory variables (Figure 3).

```{r, echo=FALSE, out.width="75%", fig.align="center", fig.cap="left-aligned", fig.cap="Figure 3: Example of a RDA plot. The blue vectors are the environmental predictors. Individuals are color coded by location and plotted in the ordination space based on their relationship with the axes (which are linear combinations of the predictor variables). SNPs are plotted gray in the center of the plot. Note that individual genotypes from wet and temperate British Columbia (green) are positively related to high annual precipitation (AP) and low temperature seasonality (sdT). Based on RDA results, SNPs are strongly correlated with these two precipitation values and these candidate SNPs are plotted along the AP vector out from the center of the plot. [Source](https://popgen.nescent.org/2018-03-27_RDA_GEA.html)"}
knitr::include_graphics("images/Fig3.png")

```

**Gradient forest (GF)** [(@ellis2012gradient)](#refs) is a machine learning method that is a multidimensional extension of the random forest algorithm [(@breiman2001random)](#refs). GF uses cross-validation to identify loci under selection, compute the importance of predictors and can identify important thresholds or change points in genetic variation along the environmental variable (Figure 4). GF can also handle non-linear relationships and correlated predictors.

```{r, echo=FALSE, out.width="75%", fig.align="center", fig.cap="left-aligned", fig.cap="Figure 4. In a study by [(@nielsen2021distinct)](#refs), GF modeling was used to infer genomic variation–environment relationships. Models were fitted to the allele frequencies of putatively neutral and outlier SNP datasets (detected with other methods) for three rocky shore species and environmental data from WorldClim and principal coordinates of neighbor matrices (PCNMs) were used as predictors in the model to account for spatial autocorrelation and unmeasured environmental variation. The figure shows cumulative importance curves based on the turnover functions for each environmental variable for the urchin (left) where the curve shape indicates the rate of change in allele frequencies along the environmental gradient, and the maximum height indicates the total turnover in allele frequencies. The authors then used the turnover functions from the GF models to transform the environmental variables into a common biological importance scale (i.e [(@fitzpatrick2015ecological)](#refs) (left). We can see neutral (a, c, e) or outlier (b, d, f) loci where the coloured map cells indicate similar allele frequencies and show that the genomic turnover patterns of putative neutral and outlier loci generally followed the biogeographical breaks of the marine environment in the region."}
knitr::include_graphics("images/Fig4.png")

```

**Bayenv** is a Bayesian hierarchical model that estimates the posterior probability of a locus being under selection. It models allele frequencies as random variables and estimates the probability of selection for each locus (Figure 5).

```{r, echo=FALSE, out.width="75%", fig.align="center", fig.cap="left-aligned", fig.cap="Figure 5:  An example of a SNP where a lower allele frequency was found to be significantly associated with warmer minimum winter temperatures using Bayenv. This SNP was characterized by a high Bayes factors and extreme allele frequencies. (A) Allele frequencies and standardized minimum winter temperatures of the SNP, rs6001912 (B) shows the geographical distribution of SNP rs6001912 [(@fitzpatrick2021experimental)](#refs)"}
knitr::include_graphics("images/Fig5.png")

```

**Latent factor mixed models** (LFMM) is a univariate regression model that includes unobserved variables (latent factors) that act to correct the model for confounding effects (i.e.population structure), which can improve power of the model when environment and demography are correlated (Figure 6) ([(@frichot2013testing)](#refs); [(@frichot2015lea)](#refs)).

```{r, echo=FALSE, out.width="75%", fig.align="center", fig.cap="left-aligned", fig.cap="Figure 6: In a study on red abalone population genomics by [(@griffiths2025seascape)](#refs), lfmm2 modeling was used to infer genomic variation–environment relationships with maximum sea surface temperature (SST) (A), Coastal Upwelling Transport Index (CUTI) (B), Biologically Effective Upwelling Transport Index (BEUTI) (C) and range BEUTI (D). The horizontal line on the manhattan plots represents the significance threshold after Bonferroni correction of 0.05 (red). Very few SNPs passed the significance threshold and they are spread across multiple scaffolds, suggesting there is weak evidence for local adaptation to these environmental gradients."}
knitr::include_graphics("images/Fig6.png")

```

## So you want to do a GEA?
So you want to perform GEA? You will need five things before getting started...

1. A good sampling regime
2. Genetic data
3. An understanding of population structure/ and/or isolation by distance (IBD) in your system 
4. Geo-referenced environmental variables and a GEA method (or two)!

## Experimental Design Conditions for GEA
### 1. A good sampling regime 
Sampling design has implications for minimizing false-discoveries and for detecting true positives (statistical power) in relation to environmental selection. Compared with random or opportunistic sampling approaches, informed sampling designs based on the characteristics of the study area and species life history tend to improve the power of GEAs [(@whitlock2015reliable)](#refs) . The general consensus suggests “stratified sampling” [(@dauphin2023re)](#refs) - a sampling regime that is informed by a priori knowledge of neutral genetic differentiation (or geographic variation) and/or environmental representativeness to capture genetic diversity and population structure. To action a stratified sampling method you would collect an equal number of samples from each group, where each group is identified from  environmental and population structure information  ([(@dauphin2023re)](#refs), Figure 7). If you  are unsure of the genetic structure of your species, it is recommended that you conduct a PCA or build a map of the environmental variables across your potential study area, with the aim to visualize the environmental variation across geographical sites before sampling (see section 4 on Environmental data below). In this case, sampling locations that are environmentally different, but replicated within environments can help distinguish selection from neutral population structure. A key problem in this field is when unobserved variables  correlate with both the primary environmental variables and the genomic data. For example, GEAs tend to have low power and high rates of false positives when environmental gradients are correlated with neutral population structure ([(@whitlock2015reliable)](#refs); [(@caye2019lfmm)](#refs)). Several methods have been specifically developed to account for this issue (see Table 2).

```{r, echo=FALSE, out.width="75%", fig.align="center", fig.cap="left-aligned", fig.cap="Figure 7 (adapted from Figure 1 from [(@dauphin2023re)](#refs)): The green boundary represents the species regional range limit and the colors reflect topography. There are three main sampling design strategies that can be used in landscape genetic studies: (A) a sampling design that doesn’t rely on environmental or genetic data, (B) a sampling design that considers environmental data only, and (C) a sampling design that takes into account genetic and environmental information (Circles in box C represent sampling locations, with shadings referring to a priori known neutral genetic differences)."}
knitr::include_graphics("images/Fig7.jpg")

```

### 2. Genetic data
Any type of genetic data can be used for GEA. Whole genome data (WGS) is typical in GEA studies, though reduced representation data (i.e. RADSeq, [(@xuereb2018asymmetric)](#refs) is also commonly used. It is important that the genetic dataset is filtered for potential biases, such as applying minor allele frequency/count cut-offs, removing missing data and related individuals, and minimum genotype depth and quality. The same set of filtering guidelines we recommended above for Outlier Analyses, also applies here to GEA (please see Step 1 in the [Outlier Analyses Section](#outlier-analyses) more details and recommended tutorials).

<ins>Some examples of different genomic data applying GEA methods available in the literature include:</ins> 
* [(@xuereb2018asymmetric)](#refs)--RadSeq
* [(@griffiths2025seascape)](#refs); [(@rumberger2023selection)](#refs); [(@wooldridge2024limited)](#refs)-- WGS
* [(@forester2025population)](#refs)--Exome Capture

### 3. Understanding population structure in your data 
Just like the outlier methods previously described, understanding *Isolation by Distance (IBD)* and/or *population structure* is critical before performing GEA analysis. IBD refers to segments of the genome shared by individuals due to common ancestry, and the degree of relatedness which can vary with distance whereby genetic similarities might be influenced by shared ancestry rather than true genotype-environment associations. Population structure, which reflects how genetic variation is distributed across different groups, can also confound results by creating artificial associations between genotypes and phenotypes. Correcting for IBD and population structure ensures that the identified associations in GEA studies are reflective of actual biological relationships, improving the reliability and validity of the findings.

<ins>Some common approaches for controlling for IBD or population structure in GEA include:</ins>

**Principal Component Analysis (PCA)**: PCA is often used to detect and account for population structure. It reduces the dimensionality of genetic data and captures the main sources of variation. The first few principal components (PCs) usually correspond to major population substructures. By including these PCs as covariates in statistical models (such as GEA or GWAS), researchers can control for population structure and avoid spurious associations due to genetic differentiation between subpopulations.

**Ancestry**: Ancestry scores (like outputs from NGSadmix, STRUCTURE, ADMIXTURE) can quantify the proportion of an individual’s ancestry from different populations. These scores can be used to adjust for population structure in association analyses by including them as covariates, helping to differentiate genetic associations that arise from population demography vs actual genotype-environment relationships.

**Geographical Coordinates (Latitude/Longitude)**: In some studies, particularly those involving environmental associations, researchers may use the geographical location of individuals (latitude/longitude) as a covariate. This helps to account for geographic gradients in environmental variables that may correlate with genetic variation.

**Distance-Based Moran's Eigenvector Maps (dbMEM)**: Distance-Based Moran's Eigenvector Maps (dbMEMS) are  derived from distance matrices, which reflect genetic or geographical distances between individuals. dbMEMs can be used to model spatial structure by decomposing distance matrices into eigenvectors that represent the underlying spatial patterns in genetic data. These eigenvectors are then incorporated into the model to control for spatial autocorrelation.

### 4. Environmental data
Lastly, you will need geo-referenced environmental data (see Table 3). There are two key types of environmental data that are used; in situ or spatially interpolated data:

1. In situ data, are on-site measurements of environmental conditions usually collected directly from the field or from remote sensing devices. 
2. Spatially interpolated data (e.g. oceanographic models) using different models and processes to remove missing data and errors, extend the evaluation of environmental conditions to periods and locations not covered by in-situ measurements.

Spatially interpolated data are commonly derived from the assimilation of atmospheric and oceanographic data from models, satellites and in situ observations. These ocean models usually provide accurate estimates of the state of the system, and are commonly used for GEAs because they have broad spatial and temporal resolution, and are usually publicly available and accessible. Some of the useful databases hosting past, present and future climate information  can be found in Table 3.

The most important consideration when selecting a database is the resolution and grain of the environmental dataset. For example, due to the small spatial scale of some key regionally specific processes (e.g. coastal processes), you should aim to identify and use an ocean model with a sufficiently high spatial and temporal resolution to resolve those processes (e.g. upwelling) which may be important for selection/adaptation in your system. In this case, you might choose to use a regional model (downscaled model) which uses data from a global climate model as input into a fine-scale, regional atmospheric model with detailed bathymetry, coastal boundaries, etc.

<ins>Resources for Environmental Data:</ins>
There are plenty of existing resources for processing this environmental data

* https://ropensci.org/blog/2019/11/05/tidync/

* https://ouicodedata.com//posts/worldclim-with-r/

* https://jorgemfa.medium.com/easy-to-use-marine-climate-layers-for-ecological-modelling-83950cd7a75f

* https://github.com/bio-oracle/biooracler

* https://www.bio-oracle.org/

* [(@assis2018bio)](#refs)– paper on bioclimatic modelling

<ins>Further recommended reading:</ins>
* [(@rellstab2015practical)](#refs) –method comparison
* [(@lotterhos2015relative)](#refs)–method comparison
* [(@franccois2016controlling)](#refs)–method comparison for false discovery rates
* [(@forester2016detecting)](#refs) (GEA)-- method comparison
* [(@dauphin2023re)](#refs)–discussion of how different types/scales of of environmental data can be used to answer different questions
* [(@capblancq2018evaluation)](#refs) -- method comparison (GEA & pcadapt)
* [(@alvarado2022genotype)](#refs) - example paper that uses gradientForest
* [(@riginos2016navigating)](#refs) – spatial scale in landscape genomics
* [(@capblancq2021redundancy)](#refs) –RDA
* [(@selmoni2020seascape)](#refs)– example landscape genomics study

### TABLE 2: Common methods for identifying candidate SNPs associated with environmental data (GEA), the associated approach for calculating divergence, the biological assumptions, and how demographic history is accounted for in the model. We also provide a link to public tutorials for each method. 
| Model & Package   | Approach/ Statistics    | Demographic History    | False Positive Rate    | Tutorial  |
|-----------------------------------------------------------------------------------------|---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------|----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------|----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------|------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------|
| Bayenv & Bayenv2 Command Line                                                           | * XtX statistics <br>* Univariate, correlation between an environmental variable and allele frequencies                              |   uses a pairwise covariance matrix derived from information on all loci to control for demographic history                                                                                                                                                                      | Inflated type-1 error (false positives), can use PCs of environment or correct for multiple testing (i.e bonferroni correction )                                                                                                                         | [2017 Maier tutorial](https://www.mountainmanmaier.com/software/pop_genom/#search-for-outliers-with-bayenv2)                                                                                                                          |
| [BayPass Command Line](https://www1.montpellier.inrae.fr/CBGP/software/baypass/index.html)  | XtX statistics       | uses a pairwise covariance matrix derived from information on all loci to control for demographic history                                                                                                                                                                       |                                                                                                                                                                                                                                                          |  [Tutorial with no environmental data](https://eead-csic-compbio.github.io/barley-agroclimatic-association/HOWTOXtX_BAYPASS.html), [Tutorial with environmental data](https://esnielsen.github.io/post/pool-seq-analyses-poolfstat-baypass/)                             |
| BayeScEnv Command Line                                                                  | Bayesian; Fst        | * like Bayescan, populations defined ahead of time <br>* Dirichlet-multinomial distribution: each population has its own parameter for neutral model                                                                                                                                   |                                                                                                                                                                                                                                                          | [Tutorial](https://github.com/devillemereuil/bayescenv/wiki), [BayeScan PDF](chrome-extension://efaidnbmnnnibpcajpcglclefindmkaj/http://evomics.org/wp-content/uploads/2016/01/BayeScan_BayeScEnv_exercises.pdf)                                                |
| RDA R(Vegan Package)                                                                    | * Multiple regression followed by PCA on resulting matrix <br>* Q-value (FDR adjusted p-value) determined from Manhanobis distances                                                                                                                                                                         | partial RDA (pRDA): RDA is first performed with covariates (demography), and the output residuals are the response variable in the second RDA   |                                                                                                                                                                                                                                                          | [Tutorial by KE Lotterhos ](https://marineomics.github.io/POP_05_RDAtraitPredictionTutorial.html]), [Partial RDA tutorial](https://r.qcbs.ca/workshop10/book-en/partial-redundancy-analysis.html), [RDA GEA tutorial](https://popgen.nescent.org/2018-03-27_RDA_GEA.html)|
| Lfmm, lfmm2 R (LEA package)                                                             | Univariate, Latent factor mixed model Lfmm: MCMC Lfmm2: least-squares estimates (for large SNP datasets)                                                                                                                    | Low, but it is more likely to have higher rates of false positives in datasets with complex population structure and with polygenic traits–controlling for neutral structure and using corrections for multiple testing (i.e FDR, Bonferroni correction) | [Bay Lab GEA Tutorial](https://baylab.github.io/MarineGenomicsSemester/genome-environment-association.html)                                                                                                                                                |
| Gradient Forest/ Random Forest  (GF) ‘gradientForest’                                   | Random forest, multivariate, machine learning approach                                                                                                                                                                                  | Typically done prior to fitting GF (i.e with BAYENV, or include spatial predictors (i.e. Moran’s eigenvector Maps (MEM).                                                                                                                                                         | Thought to be less prone to the multiple testing problem inherent, but further research needed                                                                                                                                                           | [Landscape Genomics Tutorial](https://github.com/pgugger/LandscapeGenomics/blob/master/2018_China/Exercise3.md)                                                                                                                                                  |
| Generalized dissimilarity modeling (GDM)                                                | Distance-based multivariate approach that can explores nonlinear relationships    | can accommodate a measure of population structure, but must be provided by researcher                                                                                                                                                                                           | As above                                                                                                                                                                                                                                                 |                                                                                                                                                                                                                                    |
| Weighted-Z Analysis (WZA)                                                               | WZA is a method for performing GEA using analysis windows rather than individual genetic markers (e.g. SNPs).                                                                                                                                                                                                                     | The method does not model population structure explicitly, rather it identifies genomic regions with a pattern of GEA statistics that deviate from the genome-wide average                                                                                                       | Reported to be less than baypass and lfmm-lea                                                                                                                                                                                                            | [WZA Tutorial](https://github.com/CoAdapTree/WZA)                                                                                                                                                                                                  |

### TABLE 3A. A summary (not an exhaustive list) of freely available oceanographic databases to find environmental data for your GEAs

| Database                                                            | Description                                                                                                                                                                                                           | Link/ Reference                                          |
|---------------------------------------------------------------------|-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------|----------------------------------------------------------|
| WorldClim                                                           | A set of global climate layers (gridded climate data in GeoTiff format) that can be used for mapping and spatial modeling.                                                                                            | [WorldClim Website](https://www.worldclim.org/)                               |
| World Ocean Database (WOD)                                          | Includes temperature, salinity, sea surface topography, chemistry/biology optics                                                                                                                                      | [WOD Website](www.nodc.noaa.gov/OC5/WOD/pr_wod.html)                    |
| NASA’s Earth Observing System Data and Information System (EOSDIS)  | Includes temperature, salinity, sea surface topography, chemistry/biology optics                                                                                                                                      | [EOSDIS Website](https://earthdata.nasa.gov/discipline/ocean)              |
| Australian Ocean Data Network (AODN)                                | Online network of marine and climate data resources, includes a wide range of parameters in different ocean environments collected from ocean-going ships, autonomous vehicles, moorings and other platforms.         | [AODN Website](https://imos.org.au/facilities/aodn)                      |
| NOAA Data Buoy Center (NDBC)                                        | National Oceanic and Atmospheric Administration’s (NOAA) National Weather Service (NWS) which operates, and maintains a network of data collecting buoys and coastal stations.  Can be accessed using R package rnoaa |  [NDBC Website](www.ndbc.noaa.gov)                                       |
| ICES Data Centre - Ocean (North Atlantic)                           | Oceanographic data which includes includes temperature, salinity, oxygen, chlorophyll a, and nutrients                                                                                                                | [ICES Website](http://ices.dk/marine-data/data-portals/Pages/ocean.aspx) |
| Baltic Environmental Database (BED) (Baltic Sea)                    | Bathymetry, hydrography, nutrients, sediment data, atmospheric nutrient input, and runoff data for the Baltic Sea.                                                                                                    | [BED Website](www.balticnest.org/bed)                                  |

### TABLE 3B: Environmental data reanalyses
Climate reanalyses combine past observations with models and so they tend to have wide coverage and can provide data for variables and locations where we don't have observations (i.e [here](https://www.youtube.com/watch?v=FAGobvUGl24))

| Database                                            | Description                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                 | Link/ Reference                                                           |
|-----------------------------------------------------|-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------|---------------------------------------------------------------------------|
| Copernicus Marine                                   | A database hosting various re-analysis products. Includes For example Global Ocean Physics Reanalysis (GLORYS12V1 [(@lellouche2018recent)](#refs)) includes GLOBAL_MULTIYEAR_PHY_001_030 which specifically provides an eddy resolving (1/12°) global ocean simulation, covering the recent period,  at 1/12 degree (approx. 8km) and on 50 standard levels, in four dimensions.  ERA5 provides a snapshot of the atmosphere, land surface and ocean waves for each hour from 1940 onwards (i.e  reanalysis ERA5)  Many code examples for accessing data using your fave language (R, python, Julia) here.  |[Copernicus Link](https://data.marine.copernicus.eu/products) [(@jean2021copernicus)](#refs) |
| Bluelink ReANalysis (BRAN) (Australsia region)      | A series of ocean reanalyses; the most recent being BRAN2020 [(@chamberlain2021multiscale)](#refs)  BRAN is closely related to OceanMAPS, an operational ocean reanalysis system used by the Australian Bureau of Meteorology [(@brassington2007bluelink)](#refs)                                                                                                                                                                                                                                                                                                                                           | [(@chamberlain2021multiscale)](#refs)                                     |
| European Centre for Medium-Range Weather Forecasts  | ERA5 provides a snapshot of the atmosphere, land surface and ocean waves for each hour from 1940 onwards (i.e  reanalysis ERA5)                                                                                                                                                                                                                                                                                                                                                                                                                                                                             | [ECMWF Link](https://www.ecmwf.int/en/research/climate-reanalysis)                      |

### TABLE 3C: Regional projections/ Future Climate Scenarios
Examples of some present and future climate change predictions (regional, typically modeled with reanalysis products).

| Database                                            | Description                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                 | Link/ Reference                                                           |
|-----------------------------------------------------|-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------|---------------------------------------------------------------------------|
| Copernicus Marine                                   | A database hosting various re-analysis products. Includes For example Global Ocean Physics Reanalysis (GLORYS12V1 [(@lellouche2018recent)](#refs)) includes GLOBAL_MULTIYEAR_PHY_001_030 which specifically provides an eddy resolving (1/12°) global ocean simulation, covering the recent period,  at 1/12 degree (approx. 8km) and on 50 standard levels, in four dimensions.  ERA5 provides a snapshot of the atmosphere, land surface and ocean waves for each hour from 1940 onwards (i.e  reanalysis ERA5)  Many code examples for accessing data using your fave language (R, python, Julia) here.  | [Copernicus Link](https://data.marine.copernicus.eu/products) [(@jean2021copernicus)](#refs) |
| Bluelink ReANalysis (BRAN) (Australsia region)      | A series of ocean reanalyses; the most recent being BRAN2020 [(@chamberlain2021multiscale)](#refs)  BRAN is closely related to OceanMAPS, an operational ocean reanalysis system used by the Australian Bureau of Meteorology [(@brassington2007bluelink)](#refs)                                                                                                                                                                                                                                                                                                                                           | [(@chamberlain2021multiscale)](#refs)                                     |
| European Centre for Medium-Range Weather Forecasts  | ERA5 provides a snapshot of the atmosphere, land surface and ocean waves for each hour from 1940 onwards (i.e  reanalysis ERA5)                                                                                                                                                                                                                                                                                                                                                                                                                                                                             | [ECMWF Link](https://www.ecmwf.int/en/research/climate-reanalysis)                      |

# Genome-Wide Association Study (GWAS)

## What is a Genome-Wide Association Study?
A GWAS is used to associate some phenotypic trait with putatively causative loci/SNPs. The methods used are similar to GEA, but instead of environmental data as the explanatory variable, phenotypic traits are used. Importantly, the phenotypic and genotypic data must be tied to each individual in the study. This has primarily been used in humans to identify genetic variants associated with disease [(@uffelmann2021genome)](#refs), but it is becoming more prominent for use in non-model organisms for how a particular trait related to climate adaptation (e.g., physiological, morphological, and life history) is under selection.  For example, identifying loci associated with coral bleaching (Figure 8) [(@fuller2020population)](#refs) and beak size/morphology in Galapagos finches [(@enbody2023community)](#refs). Effect sizes of loci can also be determined based on how strongly associated variants are with phenotypic variation.

In contrast to GEA, GWAS often requires a much larger sample size (and even more so for polygenic traits). This is why many GWAS studies are now emerging that employ a low-coverage sequencing approach combined with genotype imputation ([(@lou2021beginner)](#refs); [(@jorsboe2022efficient)](#refs)),where a larger number of individuals are sequenced, increasing the power to detect significant associations. Additionally, the inclusion of a larger sampling strategy will encompass a wide range of genomic diversity to identify all of the SNPs associated with the trait, such as rare SNPs.

Similar to outlier and GEA studies, GWAS are also subject to genotyping bias issues and false positives resulting from demographic history [(@tam2019benefits)](#refs). We recommend the researcher perform some combination of filtering on their dataset, which could include removing rare SNPs, SNPs not in Hardy-Weinberg equilibrium, and SNPs that are missing from a fraction of the individuals genotypes. Go back to our "Outlier Analyses" section for more details and recommended tutorials for filtering.

```{r, echo=FALSE, out.width="75%", fig.align="center", fig.cap="left-aligned", fig.cap="Figure 8: Example of GWAS implemented using a linear regression in the program PLINK. 237 coral samples were collected across 12 reefs from the Great Barrier reef. Low-coverage WGS was performed and bleaching susceptibility was measured in each individual. Each dot is a SNP and it is plotted by its location in the genome on the X-axis. The Y-axis is the -log p value for the linear regression of each SNP and its association with bleaching susceptibility. Larger values demonstrate stronger associations. The authors identified two genes with significant loci that were associated with bleaching susceptibility. Source: [(@fuller2020population)](#refs)"}
knitr::include_graphics("images/Fig8.png")

```

## Different GWAS Sampling Strategies Will Require Different Statistical Model Designs
GWAS studies could take two approaches where traits and genotype information are collected from wild, unrelated, individuals (a population-based GWAS), or from related individuals (a family-based GWAS). In addition, these two types of GWAS could include a GxE interaction where traits are measured under different experimental treatment exposures and genotype information is collected at the end of the study. While both approaches employ similar statistical analyses, such as a linear or logistic regression depending on whether the trait is continuous (e.g., body size), or binary (alive/dead) (See Table 4 below), the model matrix design may require a different set up. For example, if sampling spans across a large and variable landscape, significant genotype-phenotype associations may actually be false positives due to population structure [(@tam2019benefits)](#refs). Demographic history should be accounted for as a covariate in the model. Different methods outlined in Table 4 below will estimate population structure differently and the reader may want to choose a method that is best suited to the biology of their system.

Accounting for sample relatedness and ancestry is incredibly important for family-based GWAS approaches and may otherwise result in over-estimates of SNP-based heritability for the trait. In outlier and GEA, you may want to remove related individuals from the analysis, but for GWAS approaches, related individuals may increase your power to detect significant and heritable SNPs! Some GWAS designs specifically use a breeding or pedigree based approach for this reason, but related individuals must be identified first and pedigree information provided as a covariate in the model design to GWAS.

Other model matrix designs may include the addition of a fixed effect based on treatment exposures, specficially if it's a part of the experimental design. There may be other important covariates to include in the model as random effect terms, such as the age or sex of individuals depending on the biology of the system, or independent replicant treatment exposures depending on the experimental design.

## General Method Overview:
There are a variety of methods for GWAS, many that vary based on the type of sequencing or experimental design. For example, ANGSD-asso and SNPTEST are designed to work well with genotype probabilities from low-coverage data or data that has been imputed first. PLINK/PLINK2 is a widely used tool with lots of documentation. Other methods differ on the types of statistical associations employed. For example, the programs GEMMA and fastGWA were designed to efficiently (and quickly) test for associations for large numbers of individuals using linear mixed models. Methods also differ widely for how they estimate population structure and genetic relatedness among individuals. Some methods use a PCA based approach for estimating population structure (GEMMA, fastGWA, and SNPTEST). In a slightly similar approach, PLINK/PLINK2 uses classical Multidimensional scaling (MDS) and LEA uses latent factors to estimate values of K. Other programs, such as ANGSD-asso, require external programs for estimating population structure, but accept multiple inputs from PCAngsd, ADMIXTURE, or NGSadmix. For genetic relatedness, some approaches, EMMAX and GEMMA generate a kinship matrix or genetic matrix that is used as a covariate in the GWAS model. Tutorials for ANGSD-asso, PLINK/PLINK2, LEA, and SNPTEST don't specifically touch on whether genetic relatedness is accounted for, thus these methods are not recommended for family-based GWAS approaches. Even if researchers were to use this method for population-based GWAS, we recommend researchers use external programs to estimate genetic relatedness (i.e., COLONY, CERVUS, AlpaAssign, King’s relatedness matrix) to identify related individuals to remove them from the dataset before proceeding with GWAS. For more details, see Table 4 and our further recommended reading below.

<ins>Further recommended reading:</ins>
* [(@uffelmann2021genome)](#refs)–overview of different programs available
* [(@fuller2020population)](#refs)– GWAS example study
* [(@enbody2023community)](#refs)–GWAS example study
* [(@tam2019benefits)](#refs)–benefits and limitations of GWAS
* [(@yanez2023genome)](#refs)–Applied uses of GWAS
* [(@kijas2018evolution)](#refs)–example study

### TABLE 4: Common methods for identifying candidate SNPs associated with phenotypic data (GWAS), the associated approach for calculating divergence, the biological assumptions, and how demographic history is accounted for in the model. We also provide a link to public tutorials for each method. 
| Program                              | Approach/ Method                                                                                                                                                                                                                               | Demographic History                                                                                                                                                                                                                                                                              | Tutorial                                                                                                                                                                                         | Citation                                                                                       |
|--------------------------------------|------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------|--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------|--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------|------------------------------------------------------------------------------------------------|
| BEAGLE + ANGSD-asso                  | Impute missing genotypes with Beagle, and use beagle output as input for ANGSD -latent model (using GLM framework) that can account for genotype probabilities for low depth seq data                                                          | * Pop. structure detected with NGSadmix or ADMIXTURE or PCAngsd programs -changes allele freq prior if pop structure is present <br>* does not account for  closely related individuals–our suggestion would be to remove these individuals from the dataset                                            | [Bay Lab Tutorial](https://baylab.github.io/MarineGenomicsSemester/week-10-genome-wide-association-study-gwas.html) <br>[ANSG Tutorial](https://www.popgen.dk/angsd/index.php/Association)                                              | [(@jorsboe2022efficient)](#refs)                                                               |
| GEMMA (command line)                 | Univariate and multivariate Linear mixed models (LMMs)                                                                                                                                                                                         | * user accounts for pop structure independently by calculating eigenvalues from user-generated PCA as covariate option <br>* relatedness matrices calculated using either the centered or standardized genotype methods                                                                                | [U of Chicago Workshop Tutorial](https://github.com/rcc-uchicago/genetic-data-analysis-2) <br>[Northwestern Tutorial](http://htmlpreview.github.io/?https://github.com/northeastern-rc/training_GEMMA/blob/main/GWAS_2023.html)                                  | [(@zhou2012genome)](#refs) [(@zhou2014efficient)](#refs)                                       |
| PLINK/PLINK2                         | Multiple methods:  Cochran-Armitage trend test, Fisher’s exact test, genotypic tests (general, dominant, and recessive models), and Cochran-Mantel-Haenszeltests for stratified tables and standard linear regressions for quantitative traits | * Calculates identity by descent; calculates  complete-linkage hierarchical clustering to assess population stratification, OR can use a data-reduction technique of classical MDS to produce a k-dimensional representation of any substructure. <br>* You need to remove closely related individuals  | [Plink Tutorial](https://www.cog-genomics.org/plink/1.9/assoc) <br>[Plink2 Tutorial](https://www.cog-genomics.org/plink/2.0/assoc)                                                                                         | [(@purcell2007plink)](#refs)                                                                   |
| LEA (R package)                      | Latent factor mixed model Lfmm: MCMC Lfmm2: least-squares estimates (for large SNP datasets)                                                                                                                                                   | * Uses latent factors to estimate values of K <br>* does not account for closely related individuals–our suggestion would be to remove these individuals from the dataset                                                                                                                             | [Bay Lab Tutorial](https://baylab.github.io/MarineGenomicsSemester/genome-environment-association.html) <br>*Note that phenotypic traits must be in the env format (substitute environmental data for phenotypic traits) | [(@frichot2013testing)](#refs) [(@frichot2015lea)](#refs) [(@franccois2016controlling)](#refs) |
| fastGWA                              | mixed linear model                                                                                                                                                                                                                             | * uses SNP-derived PCs to correct for the effect due to population stratification <br>* estimates relatedness using a genetic relationship matrix (GRM) between pairs of individuals from a set of SNPs                                                                                                | [Yang Lab Tutorial](https://yanglab.westlake.edu.cn/software/gcta/#fastGWA)                                                                                                                                           | [(@jiang2019resource)](#refs)                                                                  |
| SNPTEST (improved imputation method) | Options for both frequentist linear models or bayesian models                                                                                                                                                                                  | * uses eigenvectors from a PCA analysis to code for unobserved population structure and included as a covariate in the model <br>* does not account for closely related individuals–our suggestion would be to remove these individuals from the dataset                                               | [SNPTest Tutorial](https://www.chg.ox.ac.uk/~gav/snptest/#introduction )                                                                                                                                             | [(@marchini2007new)](#refs)                                                                    |
| EMMAX                                | restricted maximum likelihoods (REML)                                                                                                                                                                                                         | * generates kinship matrix (IBS or BN, BN is preferred) using emmax-kin flag <br>* used for both population stratification and closely related individuals                                                                                                                                             | [EMMAX Tutorial](https://genome.sph.umich.edu/wiki/EMMAX)                                                                                                                                                         | [(@kang2010variance)](#refs)                                                                   |

# So you found some candidate SNPs…What next?
Great! There are many possible follow-up steps and post-hoc validation to choose from!

You might want to investigate whether loci under selection overlap any genes or are in linkage disequilibrium (LD) with genes, consider using the program LDannot if you have adequate genomic resources [(@prunier2019ld)](#refs). You need an annotated genome, but hot tip: If you don’t have an annotated genome, you can email NCBI to annotate it for you, or you can do it yourself with a program such as [Liftoff](https://https://github.com/agshumate/Liftoff) (more details in see Table 5). The program calculates linkage disequilibrium information in your VCF file using known gene regions from your annotation file to identify which loci under selection are in LD with genes.

If you have identified a long list of candidate SNPs that overlap with genes, you can perform a functional enrichment test using Gene Ontology terms (GO; a standardized database of the biological process, molecular function and cellular component of a gene) using R packages such as REVIGO or TopGO. Alternatively, if you have identified a small number of SNPs that overlap with genes (>10 or less; note that this exact number will be subjective), you can look at the functions of just those set of genes, either from the annotated genome itself, or blasting the sequence on NCBI.

Other types of questions that can be answered are the rate of genetic load in candidate SNPs. This can be explored by investigating the occurrence of nonsynonymous SNPs (a variant that changes amino acid in the protein) and the ratio of nonsynonymous vs synonymous substitutions (dN/dS). The increased ratio of dN/dS substitutions are hypothesized to be increased evidence for their relevance to adaptation [(@rellstab2015practical)](#refs).

Another growing area of research is to investigate the genomic vulnerability of the species, which uses identified outlier loci to estimate the magnitude of adaptation required to keep up with predicted changes in climate [(@capblancq2020genomic)](#refs). The R package LEA can compute values of genetic offset statistics based on new or predicted environments (see the functions genetic.gap, genetic.offset).

### TABLE 5: A short list of a few resources for annotating your candidate SNPs under selection. This is not an exhaustive list!. We also provide a link to public tutorials for each method.

| Program       | Program Function                | Tutorial                                                                                                                                | Reference                      |
|---------------|---------------------------------|-----------------------------------------------------------------------------------------------------------------------------------------|--------------------------------|
| TopGO (topGO) | Functional Enrichment           | [TopGo Tutorial](https://datacatz.wordpress.com/2018/01/19/gene-set-enrichment-analysis-with-topgo-part-1/)                                               | [(@sayols2023rrvgo)](#refs)    |
| REVIGO (rrgo) | Functional Enrichment           | [GUI web app](http://revigo.irb.hr/), [rrvgo manual](https://www.bioconductor.org/packages/release/bioc/vignettes/rrvgo/inst/doc/rrvgo.html) |  [(@supek2011revigo)](#refs)   |
| LD-annot      | Find overlap of SNPs with genes | [GitHub Link](https://github.com/ArnaudDroitLab/LD-annot)                                                                                      | [(@prunier2019ld)](#refs)      |
| Liftoff       | Annotate your genome            | [GitHub Link](https://github.com/agshumate/Liftoff)                                                                                            | [(@shumate2021liftoff)](#refs) |

# References{#refs}

### 2025-04-15 DRAFT_Signatures_Selection/DRAFT_Signatures_of_Selection.Rmd
updates to Sig Sel page

@@
     toc: yes
+    toc_float:
+      collapsed: true
 bibliography: "common-bib_01.bib"
@@
 
-*Created using [HackMD](https://https://hackmd.io/) and [tablesgenerator](https://www.tablesgenerator.com/markdown_tables)*
+*Created using [HackMD](https://hackmd.io/) and [tablesgenerator](https://www.tablesgenerator.com/markdown_tables)*
 
 ## Target audience
-This page is designed for beginners and provides a broad introduction to different methods commonly used to study natural selection in wild populations at a microevolutionary scale.
-On this page we will explore how techniques like genotype-environment analysis (GEA) and genome-wide association studies (GWAS) can help us study natural selection.
+This page is designed for beginners (with some basic undergraduate-level of understanding of evolutionary biology) and provides a broad introduction to different methods commonly used to study natural selection in wild populations at a microevolutionary scale.
+On this page we will explore how techniques like outlier analyses, genotype-environment analysis (GEA) and genome-wide association studies (GWAS) can help us study natural selection.
 To learn more about specific methods, read through the suggested readings in each section and visit the tutorials where links are provided.
@@
 **Genome-Wide Association Studies (GWAS)** examine genetic variants and identify those statistically associated with a particular trait.
-When a specific phenotype shows a significant association with a particular loci, gene or region, it suggests that this genetic variation may be adaptive.
+When a specific phenotype shows a significant association with a particular loci, gene or region, it suggests that this genetic variation may be under selection.
 
 ## Which method should I use to find loci under selection?
-Briefly, the biology of your system is always important to consider before conducting specific analyses, and no matter which method(s) you use, it is recommended that comparisons of outlier vs non-outliers (neutral loci) are performed [(@riginos2016navigating)](#refs) as a kind-of “validation” step, where we would expect loci under selection (i.e.
+The method you choose will partly depend on the data that you have available: Outlier analyses require only genetic data, but GEA requires environmental data across sampled populations, and GWAS requires trait data for each genotyped individual.
+Addtionally, the biology of your system is always important to consider before conducting specific analyses, and no matter which method(s) you use, it is recommended that comparisons of outlier vs non-outliers (neutral loci) are performed [(@riginos2016navigating)](#refs) as a kind-of “validation” step, where we would expect loci under selection (i.e.
 outliers) to demonstrate a stronger signal of environmental dependence than non-outliers.
@@
 For example, marine invertebrates with pelagic larval dispersal and high gene flow which show very low to no detectable population structure at neutral alleles across their range (i.e using STRUCTURE, ADMIXTURE, PCA) still show signals of adaptation.
-Other examples include Pacific purple sea urchins ([(@petak2023whole-cp)](#refs);[(@rumberger2023selection)](#refs)) or scallops [(@van2017identifying-cp)](#refs).
+Other examples include Pacific purple sea urchins ([(@petak2023whole-cp)](#refs);[(@rumberger2023selection)](#refs)) or scallops [(@van2017identifying)](#refs).
 
@@
 Fst) between populations/groups than expected under a neutral model are flagged as potential targets of selection.
-Specifically, while selectively neutral loci are expected to form a normal frequency distribution,  when alleles increase fitness (positive selection) they can be expected to be observed at higher frequencies and where they decrease fitness (negative selection) they can be expected to be observed at at lower frequencies [(@hoban2021genetic-cp)](#refs).
+Specifically, while selectively neutral loci are expected to form a normal frequency distribution,  when alleles increase fitness (positive selection) they can be expected to be observed at higher frequencies and where they decrease fitness (negative selection) they can be expected to be observed at at lower frequencies [(@hoban2021genetic)](#refs).
 These are  “outliers” or “putatively adaptive loci” which may be candidates of selection.
 
-There are two commonly applied approaches to outlier analysis; FST-based analysis (Figure 1) and PCA-based analysis (Figure 2).
+There are two commonly applied approaches to outlier analysis; Fst-based analysis (Figure 1) and PCA-based analysis (Figure 2).
 Both are used to identify loci under selection, but they differ in their underlying methods.
-FST is a standardized measure of the variance of allele frequencies among populations [(@wright1949population)](#refs).
-By directly comparing genetic differentiation between populations/groups to neutral expectations, FST outlier tests can identify loci which may be under selection (Figure 1).
+Fst is a standardized measure of the variance of allele frequencies among populations [(@wright1949population)](#refs).
+By directly comparing genetic differentiation between populations/groups to neutral expectations, Fst outlier tests can identify loci which may be under selection (Figure 1).
 In contrast, PCA-based methods use Principal Component Analysis (PCA) to capture population structure and can identify loci with extreme loadings on principal components as potential outliers (Figure 2).
 
-```{r, echo=FALSE, out.width="75%", fig.align="center", fig.cap="left-aligned", fig.cap="Figure 1: Example of an FST-based analysis using OutFLANK.
+```{r, echo=FALSE, out.width="75%", fig.align="center", fig.cap="left-aligned", fig.cap="Figure 1: Example of an Fst-based analysis using [OutFLANK](https://rstudio-pubs-static.s3.amazonaws.com/305384_9aee1c1046394fb9bd8e449453d72847.html).
 In this dataset, 1000 individuals were collected from across 39 populations  spanning an environmental gradient.
 Each dot is a SNP and its location in the genome plotted on the X-axis.
-The Y-axis is the FST for each SNP.
-Higher FST values are indicative of larger genetic differentiation and candidate SNPs under selection are highlighted in pink.
+The Y-axis is the Fst for each SNP.
+Higher Fst values are indicative of larger genetic differentiation and candidate SNPs under selection are highlighted in pink.
 [Source](https://htmlpreview.github.io/?https://github.com/whitlock/OutFLANK/blob/master/inst/doc/OutFLANKAnalysis.html)"}
@@
 
-```{r, echo=FALSE, out.width="75%", fig.align="center", fig.cap="left-aligned", fig.cap="Figure 2: Example of a PCA-based analysis using pcadapt.
+```{r, echo=FALSE, out.width="75%", fig.align="center", fig.cap="left-aligned", fig.cap="Figure 2: Example of a PCA-based analysis using [pcadapt](https://bcm-uga.github.io/pcadapt/articles/pcadapt.html).
 In this dataset, a total of 150 individuals from three different populations were genotyped at 1,500 markers.
@@
 
-Nonselective evolutionary forces can affect the distribution of outlier loci from loci across a genome [(@lotterhos2014evaluation)](#refs), therefore, the first step is to identify the  underlying distribution that is expected under neutral processes by answering the question;  how much variation in among loci would be expected (i.e.
-the null distribution of FST) in the absence of selection?
+Non-selective evolutionary forces can affect the distribution of outlier loci from loci across a genome [(@lotterhos2014evaluation)](#refs), therefore, the first step is to identify the  underlying distribution that is expected under neutral processes by answering the question;  how much variation in among loci would be expected (i.e.
+the null distribution of Fst) in the absence of selection?
 One must first create a null distribution using a set of SNPs (single nucleotide polymorphisms)  that accurately model the distribution expected under neutrality.
@@
 Deciding on how much filtering is required can be tricky and can feel subjective.
-We recommend visualizing the filtering steps using a PCA, which will elucidate many common biases in the dataset (see the [MarineOmics page on PCAs](https://https://marineomics.github.io/POP_08_PCA.html)).
+We recommend visualizing the filtering steps using a PCA, which will elucidate many common biases in the dataset (see the [MarineOmics page on PCAs](https://marineomics.github.io/POP_08_PCA.html)).
 We recommend using the same set of SNPs that you use for estimating population structure for neutral parameterization (which is often a set of SNPs not in LD).
-Once neutral parameterization is performed on one set of SNPs, for most programs that model can be applied to the larger filtered SNP dataset (that across the genome for evaluating their outlier status.
+Once neutral parameterization is performed on one set of SNPs, for most programs that model can be applied to the larger filtered SNP dataset.
 
-### Step 2 for detecting outlier loci - choose an outlier method based on the biology of your system!
+### Step 2 - choose an outlier method based on the biology of your system!
 
-When choosing an outlier method, it is important to understand what biological assumptions are being ignored/violated and whether the model accounts for the demographic history when estimating divergence among populations (see section “Considerations” above).
+When choosing an outlier method, it is important to understand what biological assumptions are being ignored/violated and whether the model accounts for the demographic history when estimating divergence among populations (see [Table 1](#Table1)).
 For example, early models such as fdist and fdist2 [(@beaumont1996evaluating)](#refs) assume populations are all equal in size and there is equal probability of migration and gene flow among all populations (even if some populations are located further away than others).
@@
 These estimates of population structure are included as covariates in the final Outlier Analysis model.
-See Table 1 below for more information on how each method accounts for population demographic history.
+See [Table 1](#Table1) for more information on how each method accounts for population demographic history.
 
 Population structure is also important to consider.
-For example, if you discover only one genetic cluster (K=1 in STRUCTURE or Admixture plot), perhaps any of the methods in Table 1 would work well for your data.
+For example, if you discover only one genetic cluster (K=1 in STRUCTURE or Admixture plot), perhaps any of the methods in [Table 1](#Table1) would work well for your data.
 On the other hand, if there is a lot of population structure (i.e., K>1), then a method that accounts for demographic history should be chosen.
@@
 
-### TABLE 1.
+### TABLE 1. Common methods for identifying outliers {#Table1}
 Common methods for identifying outliers and the associated approach for calculating divergence, the biological assumptions, and how demographic history is accounted for in the model.
 We also provide a link to public tutorials for each method.
-| Model & Approach                                                                                                                                                                                                                                                                                | Demographic History                             | Assumptions/ Known issues                                                                                                                                              | False Positive Rate                                                                                                                                     | Tutorial                                                                                                                                                                                                                                                   | Citation                                                            |
-| -------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------- | -------------------------------------------- | ---------------------------------------------------------------------------------------------------------------------------------------------------------------------- | ------------------------------------------------------------------------------------------------------------------------------------------------------- | ---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------- | ------------------------------------------------------------------- |
+
+| Model & Approach | Demographic History | Assumptions / Known issues | False Positive Rate | Tutorial | Citation |
+|------------------|---------------------|-----------------------------|----------------------|----------|----------|
 | fdist & fdist2 (Fst; using Arlequin software)       | * accounts for loci with different heterozygosities and with finite sample sizes per subpopulation  | * assumes populations are equal in size and migration rates <br>* non-independence among populations                                                                   | High [(@lotterhos2014evaluation)](#refs) [(@narum2011comparison)](#refs) [(@perez2010comparing)](#refs), [(@whitlock2015reliable)](#refs)               | Outdated due to high false positive rates                                                                                                                                                                                                                  | [(@beaumont1996evaluating)](#refs)                                  |
@@
 | PCAdapt (PC loadings) | * Latent factors to estimate values of K <br>* uses PC loadings (less reliant on assumptions about the underlying demography) [(@duforet2015haplopop)](#refs) <br>* Important to use a set of SNPs thinned for linkage disequilibrium for neutral parameterization [(@lotterhos2019effect)](#refs)                           | * individuals not sorted into populations ahead of time <br>* Performs well in populations with high admixture                                                         | Low to medium [(@capblancq2018evaluation)](#refs), [(@luu2017pcadapt)](#refs) Might not be suitable for populations with complex demographic histories.
-| [Luu, Privé and Blum tutorial](https://https://bcm-uga.github.io/pcadapt/articles/pcadapt.html), [UC Davis Marine Genomics course tutorial ](https://baylab.github.io/MarineGenomics/week-7--fst-and-outlier-analysis.html#finding-outliers-using-pcadapt) | [(@duforet2015haplopop)](#refs)  [(@silliman2019population)](#refs) |
-| Outflank (Weir and Cockerham’s Fst)                                                                                                                                                                                                                                                                | * populations defined ahead of time             | * estimates the number of K genetic clusters to inform a null model of Fst distributions <br>* Performs poorly in populations with admixture and pooledseq samples     |                                                                                                                                                         |                                                                                                                                                                                                                                                            |                                                                     |* finite sample sizes in pop                                                                                                                                                                                                                                                                                                                                        | Low, but suffers from low power                                                                                                                         | [Lotterhos Tutorial](https://github.com/whitlock/OutFLANK/blob/master/inst/doc/OutFLANKAnalysis.html) (2018) [UC Davis Marine Genomics course tutorial](https://baylab.github.io/MarineGenomics/week-7--fst-and-outlier-analysis.html#using-fst-to-find-outliers)  | [(@lotterhos2014evaluation)](#refs) [(@schlotterer2014sequencing)](#refs) |
+| [Luu, Privé and Blum tutorial](https://bcm-uga.github.io/pcadapt/articles/pcadapt.html), [UC Davis Marine Genomics course tutorial ](https://baylab.github.io/MarineGenomics/week-7--fst-and-outlier-analysis.html#finding-outliers-using-pcadapt) | [(@duforet2015haplopop)](#refs)  [(@silliman2019population)](#refs) |
+| Outflank (Weir and Cockerham’s Fst) | * populations defined ahead of time | * Estimates the number of K genetic clusters to inform a null model of Fst distributions <br>* Performs poorly in populations with admixture and PoolSeq samples <br>* Affected by finite sample sizes | Low, but suffers from low power | [Lotterhos Tutorial](https://github.com/whitlock/OutFLANK/blob/master/inst/doc/OutFLANKAnalysis.html) (2018), [UC Davis Marine Genomics course tutorial](https://baylab.github.io/MarineGenomics/week-7--fst-and-outlier-analysis.html#using-fst-to-find-outliers) | [(@lotterhos2014evaluation)](#refs), [(@schlotterer2014sequencing)](#refs) |
 | FLK TF-LK                                    | Sensitive to complex patterns of migration and sharp differences in the migration rate among populations [(@bonhomme2010detecting)](#refs)     | * The program FLK uses a population tree (without migration) to describe the covariance among populations and build a null model of Fst distributions.
 | Low [(@bonhomme2010detecting)](#refs)   | [UoF Bioinformatics Hub Tutorial](https://uofabioinformaticshub.github.io/GumCreek2012/scripts/S1_FLK )    | See supplementary file 1 from [(@bonhomme2010detecting)](#refs)    |
-| BayPass XTX statistics                       | Assumes HWE for PoolSeq data [(@gautier2015genome)](#refs)                                                                                                                                            | * Uses a covariance matrix of allele frequencies among populations  [(@pickrell2012inference)](#refs) * can be used with or without environmental data                                                                                                                                                                           | low                                                                                                                                                     |[Barley-Agroclimatic-Association Tutorial](https://eead-csic-compbio.github.io/barley-agroclimatic-association/HOWTOXtX_BAYPASS.html), [Erica Nielson Tutorial](https://https://esnielsen.github.io/post/pool-seq-analyses-poolfstat-baypass/)                                                                                                                            | [(@gautier2015genome)](#refs)                                             |
+| BayPass XTX statistics | Assumes HWE for PoolSeq data [(@gautier2015genome)](#refs) | * Uses a covariance matrix of allele frequencies among populations [(@pickrell2012inference)](#refs) <br> * Can be used with or without environmental data | Low | [Barley-Agroclimatic-Association Tutorial](https://eead-csic-compbio.github.io/barley-agroclimatic-association/HOWTOXtX_BAYPASS.html), [Erica Nielson Tutorial](https://esnielsen.github.io/post/pool-seq-analyses-poolfstat-baypass/) | [(@gautier2015genome)](#refs) |
+
+| Model & Approach | Demographic History | Assumptions / Known issues | False Positive Rate | Tutorial | Citation |
+|------------------|---------------------|-----------------------------|----------------------|----------|----------|
+| fdist & fdist2 (Fst; using Arlequin software) | * Accounts for loci with different heterozygosities and with finite sample sizes per subpopulation | * Assumes populations are equal in size and migration rates  <br> * Non-independence among populations | High [@lotterhos2014evaluation](#refs); [@narum2011comparison](#refs); [@perez2010comparing](#refs); [@whitlock2015reliable](#refs) | Outdated due to high false positive rates | [@beaumont1996evaluating](#refs) |
+| Bayescan (Fst) | * Populations defined ahead of time  <br> * Dirichlet-multinomial distribution: each population has its own parameter for neutral model | * Does not allow for non-random correlations among pairs of populations <br> * Slow run time [@luu2017pcadapt](#refs) | Low to high depending on the demography | [Benestan tutorial (2020)](https://rpubs.com/lbenestan/outlier) | [@foll2008genome](#refs) |
+| PCAdapt (PC loadings) | * Latent factors estimate values of K <br> * Uses PC loadings; less reliant on assumptions about the underlying demography [@duforet2015haplopop](#refs) <br> * Important to use a set of SNPs thinned for linkage disequilibrium [@lotterhos2019effect](#refs) | * Individuals not sorted into populations ahead of time <br> * Performs well in populations with high admixture | Low to medium [@capblancq2018evaluation](#refs); [@luu2017pcadapt](#refs) <br> Might not be suitable for populations with complex demographic histories | [Luu, Privé and Blum tutorial](https://bcm-uga.github.io/pcadapt/articles/pcadapt.html), [UC Davis Marine Genomics tutorial](https://baylab.github.io/MarineGenomics/week-7--fst-and-outlier-analysis.html#finding-outliers-using-pcadapt) | [@duforet2015haplopop](#refs); [@silliman2019population](#refs) |
+| Outflank (Weir & Cockerham’s Fst) | * Populations defined ahead of time | * Estimates the number of K clusters to inform a null Fst model <br> * Performs poorly in populations with admixture and PoolSeq <br> * Affected by finite sample sizes | Low, but suffers from low power | [Lotterhos Tutorial](https://github.com/whitlock/OutFLANK/blob/master/inst/doc/OutFLANKAnalysis.html), [UC Davis tutorial](https://baylab.github.io/MarineGenomics/week-7--fst-and-outlier-analysis.html#using-fst-to-find-outliers) | [@lotterhos2014evaluation](#refs); [@schlotterer2014sequencing](#refs) |
+| FLK / TF-LK | * Sensitive to complex migration patterns and sharp differences in migration rates [@bonhomme2010detecting](#refs) | * Uses a population tree (without migration) to describe the covariance among populations to build a null Fst model | Low [@bonhomme2010detecting](#refs) | [UoF Bioinformatics Hub Tutorial](https://uofabioinformaticshub.github.io/GumCreek2012/scripts/S1_FLK) | See Supplementary File 1 from [@bonhomme2010detecting](#refs) |
+| BayPass (XTX statistics) | * Assumes HWE for PoolSeq data [@gautier2015genome](#refs) | * Uses a covariance matrix of allele frequencies among populations [@pickrell2012inference](#refs) <br> * Can be used with or without environmental data | Low | [Barley Agroclimatic Tutorial](https://eead-csic-compbio.github.io/barley-agroclimatic-association/HOWTOXtX_BAYPASS.html), [Erica Nielsen Tutorial](https://esnielsen.github.io/post/pool-seq-analyses-poolfstat-baypass/) | [@gautier2015genome](#refs) |
 
@@
 GEA analyses usually control for isolation by distance in addition to demography (i.e.
-that associations between genotype and environment are not attributed to environments being far away from one another).
+ensuring that associations between genotype and environment are not driven by physical distance between environments).
 
@@
 GEA broadly refers to any methodology that assesses the association between allele frequencies and environmental variables.
-Therefore, you will need four things before getting into GEAs - a good sampling regime, genetic data, geo-referenced environmental variables (see Section x).
-GEA methods commonly cited in the literature include: Redundancy Analysis (RDA) Gradient Forest (GF), Bayesian methods (i.e.
-Bayenv) and Latent Factor Mixed Models (LFMM).
+Therefore, you will need four things before getting into GEAs - a good sampling regime, genetic data, geo-referenced environmental variables (see [Section 4.
+Environmental data](#env-data)).
+GEA methods commonly cited in the literature include: Redundancy Analysis (RDA), Gradient Forest (GF), Bayesian methods (i.e.
+Bayenv), and Latent Factor Mixed Models (LFMM).
 
@@
 ## So you want to do a GEA?
-So you want to perform GEA? You will need five things before getting started...
+So you want to perform GEA? You will need four things before getting started...
 
@@
 The general consensus suggests “stratified sampling” [(@dauphin2023re)](#refs) - a sampling regime that is informed by a priori knowledge of neutral genetic differentiation (or geographic variation) and/or environmental representativeness to capture genetic diversity and population structure.
-To action a stratified sampling method you would collect an equal number of samples from each group, where each group is identified from  environmental and population structure information  ([(@dauphin2023re)](#refs), Figure 7).
-If you  are unsure of the genetic structure of your species, it is recommended that you conduct a PCA or build a map of the environmental variables across your potential study area, with the aim to visualize the environmental variation across geographical sites before sampling (see section 4 on Environmental data below).
+To conduct a stratified sampling method you would collect an equal number of samples from each group, where each group is identified from  environmental and population structure information  ([(@dauphin2023re)](#refs), Figure 7).
+If you  are unsure of the genetic structure of your species, it is recommended that you conduct a PCA or build a map of the environmental variables across your potential study area, with the aim to visualize the environmental variation across geographical sites before sampling (see [section 4 on Environmental data](#env-data)).
 In this case, sampling locations that are environmentally different, but replicated within environments can help distinguish selection from neutral population structure.
@@
 For example, GEAs tend to have low power and high rates of false positives when environmental gradients are correlated with neutral population structure ([(@whitlock2015reliable)](#refs); [(@caye2019lfmm)](#refs)).
-Several methods have been specifically developed to account for this issue (see Table 2).
+Several methods have been specifically developed to account for this issue (see [Table 2](#Table2)).
 
@@
 
-### 4. Environmental data
-Lastly, you will need geo-referenced environmental data (see Table 3). There are two key types of environmental data that are used; in situ or spatially interpolated data:
+### 4. Environmental data {#env-data}
+Lastly, you will need geo-referenced environmental data (see [Table 3A](#Table3A) and [Table 3B](#Table3B)).
+There are two key types of environmental data that are used; in situ or spatially interpolated data:
 
@@
 These ocean models usually provide accurate estimates of the state of the system, and are commonly used for GEAs because they have broad spatial and temporal resolution, and are usually publicly available and accessible.
-Some of the useful databases hosting past, present and future climate information  can be found in Table 3.
+Some of the useful databases hosting past, present and future climate information  can be found in [Table 3A](#Table3A), [Table 3B](#Table3B), and [Table 3C](#Table3C).
 
@@
 
-* https://ropensci.org/blog/2019/11/05/tidync/
+* [Data extraction from the NetCDF databases](https://ropensci.org/blog/2019/11/05/tidync/)
 
-* https://ouicodedata.com//posts/worldclim-with-r/
+* [How to extract and plot data from Worldclim](https://ouicodedata.com//posts/worldclim-with-r/)
 
-* https://jorgemfa.medium.com/easy-to-use-marine-climate-layers-for-ecological-modelling-83950cd7a75f
+* [Data extraction from the Bio-ORACLE databse](https://jorgemfa.medium.com/easy-to-use-marine-climate-layers-for-ecological-modelling-83950cd7a75f)
 
-* https://github.com/bio-oracle/biooracler
-
-* https://www.bio-oracle.org/
-
-* [(@assis2018bio)](#refs)– paper on bioclimatic modelling
+* [Another data extraction method from the Bio-ORACLE database](https://github.com/bio-oracle/biooracler)
 
@@
 * [(@selmoni2020seascape)](#refs)– example landscape genomics study
+* [(@assis2018bio)](#refs)– paper on bioclimatic modelling
 
-### TABLE 2: Common methods for identifying candidate SNPs associated with environmental data (GEA), the associated approach for calculating divergence, the biological assumptions, and how demographic history is accounted for in the model.
+### TABLE 2: Common methods for identifying candidate SNPs associated with environmental data (GEA) {#Table2}
+Common methods for identifying candidate SNPs associated with environmental data (GEA), the associated approach for calculating divergence, the biological assumptions, and how demographic history is accounted for in the model.
 We also provide a link to public tutorials for each method.
-| Model & Package   | Approach/ Statistics    | Demographic History    | False Positive Rate    | Tutorial  |
-|-----------------------------------------------------------------------------------------|---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------|----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------|----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------|------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------|
-| Bayenv & Bayenv2 Command Line                                                           | * XtX statistics <br>* Univariate, correlation between an environmental variable and allele frequencies                              |   uses a pairwise covariance matrix derived from information on all loci to control for demographic history                                                                                                                                                                      | Inflated type-1 error (false positives), can use PCs of environment or correct for multiple testing (i.e bonferroni correction )                                                                                                                         | [2017 Maier tutorial](https://www.mountainmanmaier.com/software/pop_genom/#search-for-outliers-with-bayenv2)                                                                                                                          |
-| [BayPass Command Line](https://www1.montpellier.inrae.fr/CBGP/software/baypass/index.html)  | XtX statistics       | uses a pairwise covariance matrix derived from information on all loci to control for demographic history                                                                                                                                                                       |                                                                                                                                                                                                                                                          |  [Tutorial with no environmental data](https://eead-csic-compbio.github.io/barley-agroclimatic-association/HOWTOXtX_BAYPASS.html), [Tutorial with environmental data](https://esnielsen.github.io/post/pool-seq-analyses-poolfstat-baypass/)                             |
-| BayeScEnv Command Line                                                                  | Bayesian; Fst        | * like Bayescan, populations defined ahead of time <br>* Dirichlet-multinomial distribution: each population has its own parameter for neutral model                                                                                                                                   |                                                                                                                                                                                                                                                          | [Tutorial](https://github.com/devillemereuil/bayescenv/wiki), [BayeScan PDF](chrome-extension://efaidnbmnnnibpcajpcglclefindmkaj/http://evomics.org/wp-content/uploads/2016/01/BayeScan_BayeScEnv_exercises.pdf)                                                |
-| RDA R(Vegan Package)                                                                    | * Multiple regression followed by PCA on resulting matrix <br>* Q-value (FDR adjusted p-value) determined from Manhanobis distances                                                                                                                                                                         | partial RDA (pRDA): RDA is first performed with covariates (demography), and the output residuals are the response variable in the second RDA   |                                                                                                                                                                                                                                                          | [Tutorial by KE Lotterhos ](https://marineomics.github.io/POP_05_RDAtraitPredictionTutorial.html]), [Partial RDA tutorial](https://r.qcbs.ca/workshop10/book-en/partial-redundancy-analysis.html), [RDA GEA tutorial](https://popgen.nescent.org/2018-03-27_RDA_GEA.html)|
-| Lfmm, lfmm2 R (LEA package)                                                             | Univariate, Latent factor mixed model Lfmm: MCMC Lfmm2: least-squares estimates (for large SNP datasets)                                                                                                                    | Low, but it is more likely to have higher rates of false positives in datasets with complex population structure and with polygenic traits–controlling for neutral structure and using corrections for multiple testing (i.e FDR, Bonferroni correction) | [Bay Lab GEA Tutorial](https://baylab.github.io/MarineGenomicsSemester/genome-environment-association.html)                                                                                                                                                |
-| Gradient Forest/ Random Forest  (GF) ‘gradientForest’                                   | Random forest, multivariate, machine learning approach                                                                                                                                                                                  | Typically done prior to fitting GF (i.e with BAYENV, or include spatial predictors (i.e.
-Moran’s eigenvector Maps (MEM).
-| Thought to be less prone to the multiple testing problem inherent, but further research needed                                                                                                                                                           | [Landscape Genomics Tutorial](https://github.com/pgugger/LandscapeGenomics/blob/master/2018_China/Exercise3.md)                                                                                                                                                  |
-| Generalized dissimilarity modeling (GDM)                                                | Distance-based multivariate approach that can explores nonlinear relationships    | can accommodate a measure of population structure, but must be provided by researcher                                                                                                                                                                                           | As above                                                                                                                                                                                                                                                 |                                                                                                                                                                                                                                    |
-| Weighted-Z Analysis (WZA)                                                               | WZA is a method for performing GEA using analysis windows rather than individual genetic markers (e.g.
-SNPs).
-| The method does not model population structure explicitly, rather it identifies genomic regions with a pattern of GEA statistics that deviate from the genome-wide average                                                                                                       | Reported to be less than baypass and lfmm-lea                                                                                                                                                                                                            | [WZA Tutorial](https://github.com/CoAdapTree/WZA)                                                                                                                                                                                                  |
 
-### TABLE 3A. A summary (not an exhaustive list) of freely available oceanographic databases to find environmental data for your GEAs
+| Model & Package | Approach / Statistics | Demographic History | False Positive Rate | Tutorial / Citation |
+|-----------------|-----------------------|----------------------|---------------------|----------------------|
+|   Bayenv / Bayenv2 (Command Line)   | XtX statistics; Univariate correlation between environment and allele frequencies | Covariance matrix from genome-wide data controls for demography | Inflated; use PCs or Bonferroni correction | [Maier 2017 Tutorial](https://www.mountainmanmaier.com/software/pop_genom/#search-for-outliers-with-bayenv2) [(@coop2010using)](#refs) |
+|   [BayPass (Command Line)](https://www1.montpellier.inrae.fr/CBGP/software/baypass/index.html)   | XtX statistics | Pairwise covariance matrix controls for demography | Similar to Bayenv; adjust for multiple testing | [XtX only](https://eead-csic-compbio.github.io/barley-agroclimatic-association/HOWTOXtX_BAYPASS.html), [With env data](https://esnielsen.github.io/post/pool-seq-analyses-poolfstat-baypass/), [(@gautier2015genome)](#refs) |
+|   BayeScEnv (Command Line)   | Bayesian Fst; like BayeScan, populations defined a priori | Dirichlet-multinomial model with pop-specific parameters | Not well documented | [GitHub Tutorial](https://github.com/devillemereuil/bayescenv/wiki), [BayeScan PDF](http://evomics.org/wp-content/uploads/2016/01/BayeScan_BayeScEnv_exercises.pdf), [(@de2015new)](#refs) |
+|   RDA (R, Vegan Package)   | Multiple regression + PCA; Mahalanobis distance (Q-values) | Partial RDA includes demographic covariates | Generally low; may be sensitive to multicollinearity | [KE Lotterhos Tutorial](https://marineomics.github.io/POP_05_RDAtraitPredictionTutorial.html), [Partial RDA](https://r.qcbs.ca/workshop10/book-en/partial-redundancy-analysis.html), [RDA GEA](https://popgen.nescent.org/2018-03-27_RDA_GEA.html), [(@capblancq2021redundancy)](#refs)  |
+|   LFMM / LFMM2 (R, LEA Package)   | Univariate LFMM (MCMC or least-squares) | Latent factors (K) estimate neutral structure | Low, but higher in complex population structure; use FDR or Bonferroni | [Bay Lab Tutorial](https://baylab.github.io/MarineGenomicsSemester/genome-environment-association.html) [(@frichot2013testing)](#refs) |
+|   Gradient Forest (GF)   | Random forest; multivariate machine learning | Requires prior filtering (e.g., Bayenv) or spatial predictors (MEMs) | Thought to be lower; avoids multiple testing directly | [GF Tutorial](https://github.com/pgugger/LandscapeGenomics/blob/master/2018_China/Exercise3.md) [(@ellis2012gradient)](#refs) |
+|   Generalized Dissimilarity Modeling (GDM)   | Distance-based multivariate; nonlinear relationships | Can include population structure if provided | Fewer assumptions; less affected by marker independence | [(@ferrier2007using)](#refs) [(@fitzpatrick2015ecological](#refs) |
+|   Weighted-Z Analysis (WZA)   | Window-based GEA; aggregated signal across SNPs | Does not explicitly model structure; looks at patterns across windows | Reported lower than BayPass and LFMM | [WZA GitHub Tutorial](https://github.com/CoAdapTree/WZA), [(@booker2024wza)](#refs) |
+
+### TABLE 3A. Oceanographic databases {#Table3A}
+A summary (not an exhaustive list) of freely available oceanographic databases to find environmental data for your GEAs
 
@@
 
-### TABLE 3B: Environmental data reanalyses
+### TABLE 3B: Environmental data reanalyses {#Table3B}
 Climate reanalyses combine past observations with models and so they tend to have wide coverage and can provide data for variables and locations where we don't have observations (i.e [here](https://www.youtube.com/watch?v=FAGobvUGl24))
@@
 
-### TABLE 3C: Regional projections/ Future Climate Scenarios
+### TABLE 3C: Regional projections/ Future Climate Scenarios {#Table3C}
 Examples of some present and future climate change predictions (regional, typically modeled with reanalysis products).
 
-| Database                                            | Description                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                 | Link/ Reference                                                           |
-|-----------------------------------------------------|-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------|---------------------------------------------------------------------------|
-| Copernicus Marine                                   | A database hosting various re-analysis products.
-Includes For example Global Ocean Physics Reanalysis (GLORYS12V1 [(@lellouche2018recent)](#refs)) includes GLOBAL_MULTIYEAR_PHY_001_030 which specifically provides an eddy resolving (1/12°) global ocean simulation, covering the recent period,  at 1/12 degree (approx.
-8km) and on 50 standard levels, in four dimensions.
-ERA5 provides a snapshot of the atmosphere, land surface and ocean waves for each hour from 1940 onwards (i.e  reanalysis ERA5)  Many code examples for accessing data using your fave language (R, python, Julia) here.
-| [Copernicus Link](https://data.marine.copernicus.eu/products) [(@jean2021copernicus)](#refs) |
-| Bluelink ReANalysis (BRAN) (Australsia region)      | A series of ocean reanalyses; the most recent being BRAN2020 [(@chamberlain2021multiscale)](#refs)  BRAN is closely related to OceanMAPS, an operational ocean reanalysis system used by the Australian Bureau of Meteorology [(@brassington2007bluelink)](#refs)                                                                                                                                                                                                                                                                                                                                           | [(@chamberlain2021multiscale)](#refs)                                     |
-| European Centre for Medium-Range Weather Forecasts  | ERA5 provides a snapshot of the atmosphere, land surface and ocean waves for each hour from 1940 onwards (i.e  reanalysis ERA5)                                                                                                                                                                                                                                                                                                                                                                                                                                                                             | [ECMWF Link](https://www.ecmwf.int/en/research/climate-reanalysis)                      |
-
+| Database                                                                              | Description                                                                                                                                                                                                                                                                                                                                                                                                                      | Link/ Reference                                                                             |   |
+|---------------------------------------------------------------------------------------|----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------|---------------------------------------------------------------------------------------------|---|
+| Bedford Institute of Oceanography North Atlantic Model (BNAM)                         | An eddy-resolving, Nucleus for European Modelling of the Ocean (NEMO) based ice–ocean coupled model.
+Includes past, present day and future projections.
+| [(@wang2021life)](#refs), Contact the authors for the data, .nc files.
+|   |
+| The Gulf of Maine (GoM) Projections                                                   | Includes two high resolution numerical ocean models used to downscale global climate projections to produce four estimates of ocean physical properties in the GoM in 2050 for the “business as usual” carbon emission scenario.
+| [(@brickman2021projections)](#refs) model outputs are available upon request as .nc files.
+|   |
+| Northeastern Pacific Canadian Ocean Ecosystem Model (NEP36-CanOE) Climate Projections | Three simulations from the Northeastern Pacific Canadian Ocean Ecosystem Model (NEP36-CanOE), a configuration of the Nucleus for European Modelling of the Ocean (NEMO) V3.6.
+Includes a  historical simulation of the 1986-2005 mean climate and two future simulations project the 2046-2065 mean climate for representative concentration pathways (RCP) 4.5 (moderate mitigation scenario) and 8.5 (no mitigation scenario).
+| Contact authors                                              |
 # Genome-Wide Association Study (GWAS)
@@
 In addition, these two types of GWAS could include a GxE interaction where traits are measured under different experimental treatment exposures and genotype information is collected at the end of the study.
-While both approaches employ similar statistical analyses, such as a linear or logistic regression depending on whether the trait is continuous (e.g., body size), or binary (alive/dead) (See Table 4 below), the model matrix design may require a different set up.
+While both approaches employ similar statistical analyses, such as a linear or logistic regression depending on whether the trait is continuous (e.g., body size), or binary (alive/dead) (See [Table 4](#Table4)), the model matrix design may require a different set up.
 For example, if sampling spans across a large and variable landscape, significant genotype-phenotype associations may actually be false positives due to population structure [(@tam2019benefits)](#refs).
 Demographic history should be accounted for as a covariate in the model.
-Different methods outlined in Table 4 below will estimate population structure differently and the reader may want to choose a method that is best suited to the biology of their system.
+Different methods outlined in [Table 4](#Table4) will estimate population structure differently and the reader may want to choose a method that is best suited to the biology of their system.
 
@@
 Even if researchers were to use this method for population-based GWAS, we recommend researchers use external programs to estimate genetic relatedness (i.e., COLONY, CERVUS, AlpaAssign, King’s relatedness matrix) to identify related individuals to remove them from the dataset before proceeding with GWAS.
-For more details, see Table 4 and our further recommended reading below.
+For more details, see [Table 4](#Table4) and our further recommended reading below.
 
@@
 
-### TABLE 4: Common methods for identifying candidate SNPs associated with phenotypic data (GWAS), the associated approach for calculating divergence, the biological assumptions, and how demographic history is accounted for in the model.
+### TABLE 4: Common methods for identifying candidate SNPs associated with phenotypic data (GWAS) {#Table4}
+Common methods for identifying candidate SNPs associated with phenotypic data (GWAS), the associated approach for calculating divergence, the biological assumptions, and how demographic history is accounted for in the model.
 We also provide a link to public tutorials for each method.
+
 | Program                              | Approach/ Method                                                                                                                                                                                                                               | Demographic History                                                                                                                                                                                                                                                                              | Tutorial                                                                                                                                                                                         | Citation                                                                                       |
@@
 | BEAGLE + ANGSD-asso                  | Impute missing genotypes with Beagle, and use beagle output as input for ANGSD -latent model (using GLM framework) that can account for genotype probabilities for low depth seq data                                                          | * Pop.
-structure detected with NGSadmix or ADMIXTURE or PCAngsd programs -changes allele freq prior if pop structure is present <br>* does not account for  closely related individuals–our suggestion would be to remove these individuals from the dataset                                            | [Bay Lab Tutorial](https://baylab.github.io/MarineGenomicsSemester/week-10-genome-wide-association-study-gwas.html) <br>[ANSG Tutorial](https://www.popgen.dk/angsd/index.php/Association)                                              | [(@jorsboe2022efficient)](#refs)                                                               |
+structure detected with NGSadmix or ADMIXTURE or PCAngsd programs -changes allele freq prior if pop structure is present <br>* does not account for  closely related individuals–our suggestion would be to remove these individuals from the dataset                                            | [Bay Lab Tutorial](https://baylab.github.io/MarineGenomicsSemester/week-10-genome-wide-association-study-gwas.html) <br>[ANGSD Tutorial](https://www.popgen.dk/angsd/index.php/Association)                                              | [(@jorsboe2022efficient)](#refs)                                                               |
 | GEMMA (command line)                 | Univariate and multivariate Linear mixed models (LMMs)                                                                                                                                                                                         | * user accounts for pop structure independently by calculating eigenvalues from user-generated PCA as covariate option <br>* relatedness matrices calculated using either the centered or standardized genotype methods                                                                                | [U of Chicago Workshop Tutorial](https://github.com/rcc-uchicago/genetic-data-analysis-2) <br>[Northwestern Tutorial](http://htmlpreview.github.io/?https://github.com/northeastern-rc/training_GEMMA/blob/main/GWAS_2023.html)                                  | [(@zhou2012genome)](#refs) [(@zhou2014efficient)](#refs)                                       |
@@
 You might want to investigate whether loci under selection overlap any genes or are in linkage disequilibrium (LD) with genes, consider using the program LDannot if you have adequate genomic resources [(@prunier2019ld)](#refs).
-You need an annotated genome, but hot tip: If you don’t have an annotated genome, you can email NCBI to annotate it for you, or you can do it yourself with a program such as [Liftoff](https://https://github.com/agshumate/Liftoff) (more details in see Table 5).
+You need an annotated genome, but hot tip: If you don’t have an annotated genome, you can email NCBI to annotate it for you, or you can do it yourself with a program such as [Liftoff](https://github.com/agshumate/Liftoff) (more details in see [Table 5](#Table5)).
 The program calculates linkage disequilibrium information in your VCF file using known gene regions from your annotation file to identify which loci under selection are in LD with genes.
@@
 
-### TABLE 5: A short list of a few resources for annotating your candidate SNPs under selection. This is not an exhaustive list!. We also provide a link to public tutorials for each method.
+### TABLE 5: Annotating your candidate SNPs under selection {#Table5}
+A short list of a few resources for annotating your candidate SNPs under selection. This is not an exhaustive list!. We also provide a link to public tutorials for each method.
 
 |   Program          |   Program Function              |   Tutorial                                                                                                                                        |   Reference                      |
-|---------------|---------------------------------|-----------------------------------------------------------------------------------------------------------------------------------------|--------------------------------|
-| TopGO (topGO) | Functional Enrichment           | [TopGo Tutorial](https://datacatz.wordpress.com/2018/01/19/gene-set-enrichment-analysis-with-topgo-part-1/)                                               | [(@sayols2023rrvgo)](#refs)    |
-| REVIGO (rrgo) | Functional Enrichment           | [GUI web app](http://revigo.irb.hr/), [rrvgo manual](https://www.bioconductor.org/packages/release/bioc/vignettes/rrvgo/inst/doc/rrvgo.html) |  [(@supek2011revigo)](#refs)   |
-| LD-annot      | Find overlap of SNPs with genes | [GitHub Link](https://github.com/ArnaudDroitLab/LD-annot)                                                                                      | [(@prunier2019ld)](#refs)      |
-| Liftoff       | Annotate your genome            | [GitHub Link](https://github.com/agshumate/Liftoff)                                                                                            | [(@shumate2021liftoff)](#refs) |
+|--------------------|----------------------------------|---------------------------------------------------------------------------------------------------------------------------------------------------|----------------------------------|
+|   topGO            | Functional enrichment            | [TopGO Tutorial](https://datacatz.wordpress.com/2018/01/19/gene-set-enrichment-analysis-with-topgo-part-1/)                                      | [@sayols2023rrvgo](#refs)       |
+|   REVIGO / rrvgo   | Functional enrichment            | [GUI Web App](http://revigo.irb.hr/), [rrvgo Manual](https://www.bioconductor.org/packages/release/bioc/vignettes/rrvgo/inst/doc/rrvgo.html)    | [@supek2011revigo](#refs)       |
+|   LD-annot         | Find overlap of SNPs with genes  | [GitHub Repository](https://github.com/ArnaudDroitLab/LD-annot)                                                                                  | [@prunier2019ld](#refs)         |
+|   Liftoff          | Annotate your genome             | [GitHub Repository](https://github.com/agshumate/Liftoff)                                                                                        | [@shumate2021liftoff](#refs)    |
 

### 2025-04-17 POP_10_Signatures_of_Selection.Rmd
publishing sigs of selection page

--- 2025-04-15 DRAFT_Signatures_Selection/DRAFT_Signatures_of_Selection.Rmd
@@
       collapsed: true
-bibliography: "common-bib_01.bib"
+bibliography: "POP_10_Signatures_Selection/common-bib_01.bib"
 ---
@@
 [Source](https://htmlpreview.github.io/?https://github.com/whitlock/OutFLANK/blob/master/inst/doc/OutFLANKAnalysis.html)"}
-knitr::include_graphics("images/Fig1.png")
+knitr::include_graphics("POP_10_Signatures_Selection/images/Fig1.png")
 
@@
 [Source](https://bcm-uga.github.io/pcadapt/articles/pcadapt.html)"}
-knitr::include_graphics("images/Fig2.png")
+knitr::include_graphics("POP_10_Signatures_Selection/images/Fig2.png")
 
@@
 [Source](https://popgen.nescent.org/2018-03-27_RDA_GEA.html)"}
-knitr::include_graphics("images/Fig3.png")
+knitr::include_graphics("POP_10_Signatures_Selection/images/Fig3.png")
 
@@
 We can see neutral (a, c, e) or outlier (b, d, f) loci where the coloured map cells indicate similar allele frequencies and show that the genomic turnover patterns of putative neutral and outlier loci generally followed the biogeographical breaks of the marine environment in the region."}
-knitr::include_graphics("images/Fig4.png")
+knitr::include_graphics("POP_10_Signatures_Selection/images/Fig4.png")
 
@@
 (A) Allele frequencies and standardized minimum winter temperatures of the SNP, rs6001912 (B) shows the geographical distribution of SNP rs6001912 [(@fitzpatrick2021experimental)](#refs)"}
-knitr::include_graphics("images/Fig5.png")
+knitr::include_graphics("POP_10_Signatures_Selection/images/Fig5.png")
 
@@
 Very few SNPs passed the significance threshold and they are spread across multiple scaffolds, suggesting there is weak evidence for local adaptation to these environmental gradients."}
-knitr::include_graphics("images/Fig6.png")
+knitr::include_graphics("POP_10_Signatures_Selection/images/Fig6.png")
 
@@
 There are three main sampling design strategies that can be used in landscape genetic studies: (A) a sampling design that doesn’t rely on environmental or genetic data, (B) a sampling design that considers environmental data only, and (C) a sampling design that takes into account genetic and environmental information (Circles in box C represent sampling locations, with shadings referring to a priori known neutral genetic differences)."}
-knitr::include_graphics("images/Fig7.jpg")
+knitr::include_graphics("POP_10_Signatures_Selection/images/Fig7.jpg")
 
@@
 Source: [(@fuller2020population)](#refs)"}
-knitr::include_graphics("images/Fig8.png")
+knitr::include_graphics("POP_10_Signatures_Selection/images/Fig8.png")
 

### 2025-04-17 index.Rmd
publishing sigs of selection page

--- 2024-08-16 index.Rmd
@@
 * [PCA](POP_08_PCA.html)
+* [Ancient & Degraded DNA](POP_09_aDNA.html)
+* [Methods to detect signatures of selection](POP_10_Signatures_of_Selection.html)
 

### 2025-04-18 POP_10_Signatures_of_Selection.Rmd
removing duplicated table

@@
 | BayPass XTX statistics | Assumes HWE for PoolSeq data [(@gautier2015genome)](#refs) | * Uses a covariance matrix of allele frequencies among populations [(@pickrell2012inference)](#refs) <br> * Can be used with or without environmental data | Low | [Barley-Agroclimatic-Association Tutorial](https://eead-csic-compbio.github.io/barley-agroclimatic-association/HOWTOXtX_BAYPASS.html), [Erica Nielson Tutorial](https://esnielsen.github.io/post/pool-seq-analyses-poolfstat-baypass/) | [(@gautier2015genome)](#refs) |
-
-| Model & Approach | Demographic History | Assumptions / Known issues | False Positive Rate | Tutorial | Citation |
-|------------------|---------------------|-----------------------------|----------------------|----------|----------|
-| fdist & fdist2 (Fst; using Arlequin software) | * Accounts for loci with different heterozygosities and with finite sample sizes per subpopulation | * Assumes populations are equal in size and migration rates  <br> * Non-independence among populations | High [@lotterhos2014evaluation](#refs); [@narum2011comparison](#refs); [@perez2010comparing](#refs); [@whitlock2015reliable](#refs) | Outdated due to high false positive rates | [@beaumont1996evaluating](#refs) |
-| Bayescan (Fst) | * Populations defined ahead of time  <br> * Dirichlet-multinomial distribution: each population has its own parameter for neutral model | * Does not allow for non-random correlations among pairs of populations <br> * Slow run time [@luu2017pcadapt](#refs) | Low to high depending on the demography | [Benestan tutorial (2020)](https://rpubs.com/lbenestan/outlier) | [@foll2008genome](#refs) |
-| PCAdapt (PC loadings) | * Latent factors estimate values of K <br> * Uses PC loadings; less reliant on assumptions about the underlying demography [@duforet2015haplopop](#refs) <br> * Important to use a set of SNPs thinned for linkage disequilibrium [@lotterhos2019effect](#refs) | * Individuals not sorted into populations ahead of time <br> * Performs well in populations with high admixture | Low to medium [@capblancq2018evaluation](#refs); [@luu2017pcadapt](#refs) <br> Might not be suitable for populations with complex demographic histories | [Luu, Privé and Blum tutorial](https://bcm-uga.github.io/pcadapt/articles/pcadapt.html), [UC Davis Marine Genomics tutorial](https://baylab.github.io/MarineGenomics/week-7--fst-and-outlier-analysis.html#finding-outliers-using-pcadapt) | [@duforet2015haplopop](#refs); [@silliman2019population](#refs) |
-| Outflank (Weir & Cockerham’s Fst) | * Populations defined ahead of time | * Estimates the number of K clusters to inform a null Fst model <br> * Performs poorly in populations with admixture and PoolSeq <br> * Affected by finite sample sizes | Low, but suffers from low power | [Lotterhos Tutorial](https://github.com/whitlock/OutFLANK/blob/master/inst/doc/OutFLANKAnalysis.html), [UC Davis tutorial](https://baylab.github.io/MarineGenomics/week-7--fst-and-outlier-analysis.html#using-fst-to-find-outliers) | [@lotterhos2014evaluation](#refs); [@schlotterer2014sequencing](#refs) |
-| FLK / TF-LK | * Sensitive to complex migration patterns and sharp differences in migration rates [@bonhomme2010detecting](#refs) | * Uses a population tree (without migration) to describe the covariance among populations to build a null Fst model | Low [@bonhomme2010detecting](#refs) | [UoF Bioinformatics Hub Tutorial](https://uofabioinformaticshub.github.io/GumCreek2012/scripts/S1_FLK) | See Supplementary File 1 from [@bonhomme2010detecting](#refs) |
-| BayPass (XTX statistics) | * Assumes HWE for PoolSeq data [@gautier2015genome](#refs) | * Uses a covariance matrix of allele frequencies among populations [@pickrell2012inference](#refs) <br> * Can be used with or without environmental data | Low | [Barley Agroclimatic Tutorial](https://eead-csic-compbio.github.io/barley-agroclimatic-association/HOWTOXtX_BAYPASS.html), [Erica Nielsen Tutorial](https://esnielsen.github.io/post/pool-seq-analyses-poolfstat-baypass/) | [@gautier2015genome](#refs) |
 

### 2025-10-16 RDAtraitPredictionTutorial.Rmd
modified RDA tutorial function to be compatible with newer version of vegan:

@@
 ```{r function}
+#### old function from Katie with older version of vegan
+# rda_trait_pred <- function(rdaobj, env_row, K){
+#   #rdaobj is RDA object
+#   #env_row is the row of the environment in the biplot output
+#   #K is the number of RDA axes
+#   scores <- scores(rdaobj, choices=1:K)
+#   ind.sc <- scores$sites
+#   pred <- matrix(NA, nrow=nrow(ind.sc), ncol=K)
+#   for (k in 1:K){
+#     pred[,k] <- ind.sc[,k]*eigenvals(rdaobj)[k]*summary(rdaobj)$biplot[env_row,k]
+#   }
+#   trait_pred <- scale(rowSums(pred))
+#  return(trait_pred) 
+# }
+
+##### updated function that works with newer version of vegan
 rda_trait_pred <- function(rdaobj, env_row, K){
-  #rdaobj is RDA object
-  #env_row is the row of the environment in the biplot output
-  #K is the number of RDA axes
-  scores <- scores(rdaobj, choices=1:K)
-  ind.sc <- scores$sites
-  pred <- matrix(NA, nrow=nrow(ind.sc), ncol=K)
+  # clamp K to available constrained axes
+  K <- min(K, length(vegan::eigenvals(rdaobj)))
+
+  # site scores and biplot loadings for the first K axes
+  ind.sc <- vegan::scores(rdaobj, display = "sites", choices = 1:K)
+  bp     <- vegan::scores(rdaobj, display = "bp",    choices = 1:K)
+
+  pred <- matrix(NA, nrow = nrow(ind.sc), ncol = K)
   for (k in 1:K){
-    pred[,k] <- ind.sc[,k]*eigenvals(rdaobj)[k]*summary(rdaobj)$biplot[env_row,k]
+    pred[, k] <- ind.sc[, k] * vegan::eigenvals(rdaobj)[k] * bp[env_row, k]
   }
